### 2024-04-09 Manuscript.qmd
Draft of 'Manuscript.qmd' for special issue

---
title: "The cost of school consolidations: impacts of reduced accessibility"
bibliography: mybibfile.bib
linenumbers: true
numbersections: true
format: pdf
editor: source
author:
  - name: "Anastasia Soukhov"
    email: "soukhoa@mcmaster.ca"
    affiliation: McMaster University
    footnote: "Corresponding Author"
  - name: "Antonio Paez"
    email: "paezha@mcmaster.ca"
    affiliation: McMaster University
address:
  - code: McMaster University
    address: "School of Earth, Environment and Society, McMaster University, Hamilton, ON, L8S 4K1, Canada"
abstract: | 
    | The reduction of mobility-induced GHG emissions is a top priority for many communities around the world. From a different standpoint, communities are consistently under pressure to increase operational cost-efficiency savings, including the consideration of spatially consolidating public resource provision. This is the case for public schools, where governments are opting to close under-capacity schools and expand existing schools to accommodate lost capacity. This study quantifies the travel burdens of school consolidation/closure policy through the case of Hamilton, a mid-sized city in Ontario, Canada. Between the years 2011 and 2015, 7% of the elementary schools closed in the city. Specifically, we quantify how these closures changed the accessibility landscape for the elementary school aged population by using spatial availability, a novel singly-constrained accessibility measure. Spatial availability allows for an interpretable comparison between ‘before’ and ‘after’ policy implementation because of its unique proportional allocation property. Analysis is conducted at the parcel-level, all school-seats being eligible and travel behaviour based on observved home-to-school patterns for motorzied and non-motorized modes. Analysis on system-wide impacts such as potential GHG emission increases and walk minutes decreases are detailed. Overall, spatial availability decreased in the city. Though the school conslidation policy resulted in short-term operational cost-savings, families, students, and the environment now pay the price. This paper showcases how spatial availability can be used by decision- makers to quantify spatial and travel implications in policy.
keywords: [accessibility; spatial availability; carbon emissions; journey-to-school; walkability; active transportation; school consolidation policy; service provision thresholds]
---

```{r knitr-setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  cache = FALSE,
  warning = FALSE,
  message = FALSE,
  comment = '', 
  out.width = "1\\linewidth")
```

```{r, include=FALSE}
rm(list=ls())
```

<!--- 2016 and 2011 sp_avail is calculated in 'test-doc-2016run', LOAD THE BELOW! -->
```{r, include=FALSE}
library(dplyr)
library(sf)
library(ggplot2)
library(tidyverse)
library(units)
library(patchwork)
library(ggspatial)
library(scales)
library(stplanr)
library(tmap)
```

```{r, include=FALSE}
load(file="data-prep/data-products/cen_data_2011_hamiltononly.Rdata")
cen_data_2011_hamiltononly <- cen_data_2011_1 |> filter(Region.Name == "Hamilton") #NOTE: GEOUID 35250242, 35250133, 35250457 do not have any child population
rm("cen_data_2011_1")
load(file="data-prep/data-products/cen_data_2016_hamiltononly.Rdata")
cen_data_2016_hamiltononly <- cen_data_2016_1 |> filter(Region.Name == "Hamilton") #NOTE: GEOUID 35250133, 35250398, 35250427 do not have any child population
rm("cen_data_2016_1")

ham_city_bound <- st_read("data-prep/data-inputs/Boundaries/City_Boundary.shp") |> st_transform(st_crs(cen_data_2011_hamiltononly))
ham_community_bounds <- st_read("data-prep/data-inputs/Boundaries/Community_Boundaries.shp") |> st_transform(st_crs(cen_data_2011_hamiltononly))
```

```{r, include=FALSE}
cen_data_2011_hamiltononly <- 
  cen_data_2011_hamiltononly %>% dplyr::select( #removing the variables we don't need for this analysis
  -c("Shape.Area", "Quality.Flags", "Type", "CD_UID", "NHS.Non.Return.Rate", "CSD_UID", "CT_UID", "CT_UID","NHS.Non.Return.Rate.1", "v_CA11N_2564..Median.after.tax.household.income..", "v_CA11F_17..15.to.19.years")) %>%
  rename( #renaming the variables
    "DA_households" = "Households",
    "DA_dwellings" = "Dwellings",
    "DA_tot_pop" = "Population",
    "DA_med_AT_family_income" = "v_CA11N_2458..Median.after.tax.family.income..",
    "DA_med_AT_lone-family_income" = "v_CA11N_2476..Median.after.tax.family.income..",
    "DA_prev_LIMAT_under18" = "v_CA11N_2609..Less.than.18.years..",
    "DA_pop_5to9" = "v_CA11F_11..5.to.9.years",
    "DA_pop_10to14" = "v_CA11F_14..10.to.14.years") %>%
  mutate(DA_pop_5to14 = DA_pop_5to9+(DA_pop_10to14)) |> #*0.9)) %>%
  dplyr::select(-c("DA_pop_5to9","DA_pop_10to14"))

cen_data_2016_hamiltononly <- 
  cen_data_2016_hamiltononly %>% dplyr::select( #removing the variables we don't need for this analysis
  -c("Shape.Area",  "Type", "CD_UID", "CSD_UID", "CT_UID", "CT_UID", "v_CA16_2398..Median.after.tax.income.of.households.in.2015....", "v_CA16_64..15.to.19.years")) %>%
  rename( #renaming the variables
    "DA_households" = "Households",
    "DA_dwellings" = "Dwellings",
    "DA_tot_pop" = "Population",
    "DA_med_AT_family_income" = "v_CA16_2448..Median.after.tax.income.of.economic.families.in.2015....",
         "DA_med_AT_lone-family_income" = "v_CA16_2460..Median.after.tax.income.of.lone.parent.economic.families.in.2015....",
         "DA_prev_LIMAT_under18" = "v_CA16_2543..0.to.17.years....",
         "DA_pop_5to9" = "v_CA16_25..5.to.9.years",
         "DA_pop_10to14" = "v_CA16_43..10.to.14.years") %>% mutate(DA_pop_5to14 = DA_pop_5to9+(DA_pop_10to14)) |> #*0.9)) %>%
         dplyr::select(-c("DA_pop_5to9","DA_pop_10to14"))
```

```{r, include=FALSE}
library(hamONdests)
data(Schools_2011_2016)
ALL_SCHOOLS_Elem <- Schools_2011_2016 %>% filter(Level == "Elementary")
rm("Schools_2011_2016")
data(School_Catchments_2011_2016)
ALL_SCHOOLS_CATCH_Elem <- School_Catchments_2011_2016 %>% filter(Level == "Elementary")
rm("School_Catchments_2011_2016")

load(file="data-prep/data-products/TTS_gamma2011_motor.Rdata")
load(file="data-prep/data-products/TTS_gamma2016_motor.Rdata")
load(file="data-prep/data-products/TTS_exp2011_nonmotor.Rdata")
load(file="data-prep/data-products/TTS_exp2016_nonmotor.Rdata")
```

```{r}
# # maybe consider using year 2021 as a comparision? 2011 vs 2016 vs 2021?
# SCHOOLS_2023 <- st_read("data-prep/data-inputs/Educational_Institutions-2023/Educational_Institutions.shp") |> filter(CATEGORY == "Elementary School" & (SCHOOL_BOA == "Hamilton-Wentworth Catholic District School Board" | SCHOOL_BOA == "Hamilton-Wentworth District School Board" )) |> st_transform(st_crs(cen_data_2011_hamiltononly))
```

```{r augmenting-catchment}
ALL_SCHOOLS_CATCH_Elem_extra2011 <- ALL_SCHOOLS_CATCH_Elem |>
  left_join(ALL_SCHOOLS_Elem |> st_drop_geometry() |> group_by(CID2011) |> 
              summarize(OTGC = sum(OTGC2011, na.rm = T)), 
            by = c("CID" = "CID2011")) |> filter(!is.na(OTGC))

ALL_SCHOOLS_CATCH_Elem_extra2016 <- ALL_SCHOOLS_CATCH_Elem |>
  left_join(ALL_SCHOOLS_Elem |> st_drop_geometry() |> group_by(CID2016) |> 
              summarize(OTGC = sum(OTGC2016, na.rm = T)), 
            by = c("CID" = "CID2016")) |> filter(!is.na(OTGC))

ALL_SCHOOLS_CATCH_Elem_extra <- rbind(ALL_SCHOOLS_CATCH_Elem_extra2011, ALL_SCHOOLS_CATCH_Elem_extra2016)
rm(ALL_SCHOOLS_CATCH_Elem_extra2011,ALL_SCHOOLS_CATCH_Elem_extra2016)

#checks
sum(ALL_SCHOOLS_Elem$OTGC2011, na.rm=T) + sum(ALL_SCHOOLS_Elem$OTGC2016, na.rm=T)
ALL_SCHOOLS_CATCH_Elem_extra$OTGC |> sum(na.rm=T)

#reformating schools for mapping
ALL_SCHOOLS_Elem_nochangeorexpanded <- ALL_SCHOOLS_Elem |> filter(Year == "2011 and 2016") |> mutate(Year = ifelse(Year == "2011 and 2016", "2011", Year))
ALL_SCHOOLS_Elem_formapping <- rbind(ALL_SCHOOLS_Elem, ALL_SCHOOLS_Elem_nochangeorexpanded)
rm(ALL_SCHOOLS_Elem_nochangeorexpanded)
ALL_SCHOOLS_Elem_formapping <- ALL_SCHOOLS_Elem_formapping |> mutate(Year = ifelse(Year == "2011 and 2016", "2016", Year))
```

```{r descriptive-catchment-schools-plot, eval=FALSE}
tmap_defaults <-
  tm_scale_bar(position = c("LEFT", "BOTTOM"), breaks = c(0, 5, 10, 15))+
  tm_compass(size = 1.5, position = c("right", 'top')) +
  tm_layout(legend.position = c("left", "top")) #legend.hist.size = 10,

catchment_schools_plot <-
  tm_shape(ALL_SCHOOLS_CATCH_Elem_extra) +
  tm_polygons(col = "OTGC", border.alpha=0.1, palette = "Greys",
              labels = c("[181-318)", "[318-406)", "[406-566)","[566-1417]"),
              breaks = c(181, 318, 406, 566, 1417),
              title = "On the ground\ncapacity (OTGC)")+
  tm_facets(by=c("System","Year"))+
  tm_shape(ALL_SCHOOLS_Elem_formapping) +
  tm_dots(col = "Status", size=0.1, title = "School status between\n2011 and 2016",
          palette = c("#ffd14d", "black", "darkgreen","#0E86D4","#90EE90"),
          labels = c("No change","Closed before 2016", "Expanded before 2016","Moved before 2016","New after 2011")          
          ) +
    tm_facets(by=c("Year","System"))+
  tmap_defaults +
  tm_layout(panel.labels = list(c('HWDCSB (Catholic Public)','HWDSB (Public)'),c('2011','2016')))

tmap_save(catchment_schools_plot, "Figures/catchment_schools_plot.png", height=7, width=12, dpi=600)
```

```{r,include=FALSE}
load(file="data-prep/data-products/avail_2011.RData")
load(file="data-prep/data-products/avail_2016.RData")
```

```{r}
community_car_perc_2011 <- avail_2011 |>
  group_by(COMMUNITY) |> 
  summarize(TAZOrig_car_perc = mean(TAZOrig_car_perc, na.rm=T))

community_car_perc_2016 <- avail_2016 |> group_by(COMMUNITY) |> summarize(TAZOrig_car_perc = mean(TAZOrig_car_perc, na.rm=T))
```

```{r, eval=FALSE}
#check to make sure that V_i is equal to the total capacity in Hamilton. It is!
avail_2011$V_ij %>% sum()
ALL_SCHOOLS_Elem$OTGC2011 %>% sum(na.rm = TRUE)

avail_2016$V_ij %>% sum()
ALL_SCHOOLS_Elem$OTGC2016 %>% sum(na.rm = TRUE)

#this should equal 58,265, still does -- great!
avail_2011 %>%
  group_by(ID) %>%
  summarize(kid_popadj = sum(kid_popadj)) |> dplyr::select(kid_popadj) %>% sum()

#this should equal 58,865, still does -- great!
avail_2016 %>%
  group_by(ID) %>%
  summarize(kid_popadj = sum(kid_popadj)) |> dplyr::select(kid_popadj) %>% sum()
```

```{r}
SAvail_im_DA2011 <- avail_2011 %>%
  group_by(GeoUID, 
           mode) %>%
  summarize(COMMUNITY = first(COMMUNITY),
            total_tt = sum(travel_time_p50*kid_popadj),
            V_im = sum(V_ij),
            kid_popadj_m = sum(kid_popadj),
            DA_pop_5to14 = mean(DA_pop_5to14, na.rm=T),
            DA_prev_LIMAT_under18 = mean(DA_prev_LIMAT_under18),
            .groups = "drop") |>
  mutate(emissions_total_car_tt = ifelse(mode == "c", 0.1*total_tt, 0),
         v_im_kid_popadj = V_im/kid_popadj_m,
         v_im_kid_popadj_LIMAT = v_im_kid_popadj*(DA_prev_LIMAT_under18/100),
         v_im_kid_popadj_LIMAT = ifelse(is.na(v_im_kid_popadj_LIMAT) | is.nan(v_im_kid_popadj_LIMAT), 0, v_im_kid_popadj_LIMAT),
         V_i_kid_popadj_LIMAT = V_im*(DA_prev_LIMAT_under18/100),
         V_i_kid_popadj_LIMAT = ifelse(is.na(V_i_kid_popadj_LIMAT) | is.nan(V_i_kid_popadj_LIMAT), 0, V_i_kid_popadj_LIMAT)) |>
  filter(GeoUID != 35250242 & GeoUID != 35250133 & GeoUID != 35250457 ) #these GeoUID's have 0 kid population, so let's drop so we don't have zeros

SAvail_im_DA2016 <- avail_2016 %>%
  group_by(GeoUID, 
           mode) %>%
  summarize(COMMUNITY = first(COMMUNITY),
            total_tt = sum(travel_time_p50*kid_popadj),
            V_im = sum(V_ij),
            kid_popadj_m = sum(kid_popadj),
            DA_pop_5to14 = mean(DA_pop_5to14, na.rm=T),
            DA_prev_LIMAT_under18 = mean(DA_prev_LIMAT_under18),
            .groups = "drop") |>
  mutate(emissions_total_car_tt = ifelse(mode == "c", 0.1*total_tt, 0),
         v_im_kid_popadj = V_im/kid_popadj_m,
         v_im_kid_popadj = ifelse(is.na(v_im_kid_popadj) | is.nan(v_im_kid_popadj), 0, v_im_kid_popadj),
         v_im_kid_popadj_LIMAT = v_im_kid_popadj*(DA_prev_LIMAT_under18/100),
         v_im_kid_popadj_LIMAT = ifelse(is.na(v_im_kid_popadj_LIMAT) | is.nan(v_im_kid_popadj_LIMAT), 0, v_im_kid_popadj_LIMAT),
         V_i_kid_popadj_LIMAT = V_im*(DA_prev_LIMAT_under18/100),
         V_i_kid_popadj_LIMAT = ifelse(is.na(V_i_kid_popadj_LIMAT) | is.nan(V_i_kid_popadj_LIMAT), 0, V_i_kid_popadj_LIMAT)) |>
  filter(GeoUID != 35250398 & GeoUID != 35250133 & GeoUID != 35250427 ) #these GeoUID's have 0 kid population, so let's drop so we don't have zeros

#add the geometries -- census DAs
SAvail_im_DA2011 <- SAvail_im_DA2011 |>
  left_join(cen_data_2011_hamiltononly |> filter(Region.Name == "Hamilton") |> dplyr::select(GeoUID), by="GeoUID") |> st_set_geometry("geometry")

SAvail_im_DA2016 <- SAvail_im_DA2016 |>
  left_join(cen_data_2016_hamiltononly |> filter(Region.Name == "Hamilton") |> dplyr::select(GeoUID), by="GeoUID") |> st_set_geometry("geometry")
```
```{r}
#Create 1 sf with data joined, no changes
SAvail_im_DA <-
  rbind(SAvail_im_DA2011 |> mutate(year ="2011"),
        SAvail_im_DA2016 |> mutate(year ="2016")) |>
  mutate(mode = case_when(mode == "c" ~ "motorized",
                          mode == "w" ~ "non-motorized"))

#calc changes between the years - make that 1 sf
SAvail_im_DA_changes <- full_join(SAvail_im_DA2011|>st_drop_geometry(),
                                          SAvail_im_DA2016, by=c("GeoUID", "mode")) |>
  transmute(GeoUID,
            mode = case_when(mode == "c" ~ "motorized",
                          mode == "w" ~ "non-motorized"),
            COMMUNITY = dplyr::coalesce(COMMUNITY.x, COMMUNITY.y),
            c_total_tt = total_tt.y - total_tt.x, #2016 minus 2011 tt
            c_V_im = V_im.y -V_im.x,
            c_V_im_perc = (V_im.y -V_im.x)/V_im.x,
            c_kid_popadj_m = kid_popadj_m.y - kid_popadj_m.x,
            c_DA_pop_5to14 = DA_pop_5to14.y - DA_pop_5to14.x,
            c_DA_prev_LIMAT_under18 = DA_prev_LIMAT_under18.y - DA_prev_LIMAT_under18.x,
            c_emissions_total_car_tt = emissions_total_car_tt.y - emissions_total_car_tt.x,
            c_v_im_kid_popadj = v_im_kid_popadj.y - v_im_kid_popadj.x,
            c_v_im_kid_popadj_perc = (v_im_kid_popadj.y - v_im_kid_popadj.x)/v_im_kid_popadj.x,
            c_v_im_kid_popadj_LIMAT = v_im_kid_popadj_LIMAT.y - v_im_kid_popadj_LIMAT.x,
            c_V_i_kid_popadj_LIMAT = V_i_kid_popadj_LIMAT.y - V_i_kid_popadj_LIMAT.x,
            geometry) |> st_as_sf()

```

```{r}
#repeat but for all 1 mode
SAvail_i_DA2011 <- avail_2011 %>%
  group_by(GeoUID) %>%
  summarize(COMMUNITY = first(COMMUNITY),
            total_tt = sum(travel_time_p50*kid_popadj, na.rm = T),
            V_i = sum(V_ij, na.rm = T),
            DA_pop_5to14 = mean(DA_pop_5to14, na.rm=T),
            kid_pop_DAavg = mean(kid_pop, na.rm = T),
            DA_prev_LIMAT_under18 = mean(DA_prev_LIMAT_under18, na.rm = T),
            .groups = "drop") |>
  filter(GeoUID != 35250242 & GeoUID != 35250133 & GeoUID != 35250457 ) #these GeoUID's have 0 kid population, so let's drop so we don't have zeros

SAvail_i_DA2016 <- avail_2016 %>%
  group_by(GeoUID) %>%
  summarize(COMMUNITY = first(COMMUNITY),
            total_tt = sum(travel_time_p50*kid_popadj),
            V_i = sum(V_ij, na.rm = T),
            DA_pop_5to14 = mean(DA_pop_5to14, na.rm=T),
            kid_pop_DAavg = mean(kid_pop, na.rm = T),
            DA_prev_LIMAT_under18 = mean(DA_prev_LIMAT_under18, na.rm = T),
            .groups = "drop") |>
  filter(GeoUID != 35250398 & GeoUID != 35250133 & GeoUID != 35250427 ) #these GeoUID's have 0 kid population, so let's drop so we don't have zeros

#add the geometries -- census DAs
SAvail_i_DA2011 <- SAvail_i_DA2011 |>
  left_join(cen_data_2011_hamiltononly |> filter(Region.Name == "Hamilton") |> dplyr::select(GeoUID), by="GeoUID") |> mutate(Year = "2011") |> st_set_geometry("geometry")

SAvail_i_DA2016 <- SAvail_i_DA2016 |>
  left_join(cen_data_2016_hamiltononly |> filter(Region.Name == "Hamilton") |> dplyr::select(GeoUID), by="GeoUID") |> mutate(Year = "2016") |> st_set_geometry("geometry")

#join them together!
SAvail_i_DA_extra <- rbind(SAvail_i_DA2011, SAvail_i_DA2016)
rm(SAvail_i_DA2011,SAvail_i_DA2016)
```

```{r}
#now do this at the community level
ham_community_bounds <- ham_community_bounds |> 
  mutate(COMMUNITY = case_when(COMMUNITY_ == "Ancaster" ~ "AN",
                               COMMUNITY_ == "Dundas" ~ "DU",
                               COMMUNITY_ == "Flamborough" ~ "FL",
                               COMMUNITY_ == "Glanbrook" ~ "GL",
                               COMMUNITY_ == "Hamilton" ~ "HA",
                               COMMUNITY_ == "Stoney Creek" ~ "SC"))

SAvail_im_COMM2011 <- SAvail_im_DA2011 %>% st_drop_geometry() |>
  group_by(COMMUNITY, 
           mode) %>%
  summarize(total_tt = sum(total_tt, na.rm=T),
            V_im = sum(V_im, na.rm=T),
            kid_popadj_m = sum(kid_popadj_m, na.rm=T),
            COMM_prev_LIMAT_under18 = mean(DA_prev_LIMAT_under18, na.rm=T),
            emissions_total_car_tt = sum(emissions_total_car_tt, na.rm=T),
            .groups = "drop") |>
  mutate(v_im_kid_popadj = V_im/kid_popadj_m,
         v_im_kid_popadj_LIMAT = v_im_kid_popadj*(COMM_prev_LIMAT_under18/100),
         v_im_kid_popadj_LIMAT = ifelse(is.na(v_im_kid_popadj_LIMAT) | is.nan(v_im_kid_popadj_LIMAT), 0, v_im_kid_popadj_LIMAT),
         V_i_kid_popadj_LIMAT = V_im*(COMM_prev_LIMAT_under18/100),
         V_i_kid_popadj_LIMAT = ifelse(is.na(V_i_kid_popadj_LIMAT) | is.nan(V_i_kid_popadj_LIMAT), 0, V_i_kid_popadj_LIMAT))  |> left_join(ham_community_bounds |> dplyr::select(-c("OBJECTID")), by=c("COMMUNITY")) |> st_set_geometry("geometry")

SAvail_im_COMM2016 <- SAvail_im_DA2016 %>% st_drop_geometry() |>
  group_by(COMMUNITY, 
           mode) %>%
  summarize(total_tt = sum(total_tt, na.rm=T),
            V_im = sum(V_im, na.rm=T),
            kid_popadj_m = sum(kid_popadj_m, na.rm=T),
            COMM_prev_LIMAT_under18 = mean(DA_prev_LIMAT_under18, na.rm=T),
            emissions_total_car_tt = sum(emissions_total_car_tt, na.rm=T),
            .groups = "drop") |>
  mutate(v_im_kid_popadj = V_im/kid_popadj_m,
         v_im_kid_popadj_LIMAT = v_im_kid_popadj*(COMM_prev_LIMAT_under18/100),
         v_im_kid_popadj_LIMAT = ifelse(is.na(v_im_kid_popadj_LIMAT) | is.nan(v_im_kid_popadj_LIMAT), 0, v_im_kid_popadj_LIMAT),
         V_im_kid_popadj_LIMAT = V_im*(COMM_prev_LIMAT_under18/100),
         V_im_kid_popadj_LIMAT = ifelse(is.na(V_im_kid_popadj_LIMAT) | is.nan(V_im_kid_popadj_LIMAT), 0, V_im_kid_popadj_LIMAT))  |> left_join(ham_community_bounds |> dplyr::select(-c("OBJECTID")), by=c("COMMUNITY")) |> st_set_geometry("geometry")
```

```{r}
#repeat above but with just i, this is to get the correct LIM value.
SAvail_i_COMM2011 <- SAvail_im_DA2011 %>% st_drop_geometry() |>
  group_by(COMMUNITY) %>%
  summarize(COMM_prev_LIMAT_under18 = mean(DA_prev_LIMAT_under18, na.rm=T),
            .groups = "drop")

SAvail_i_COMM2016 <- SAvail_im_DA2016 %>% st_drop_geometry() |>
  group_by(COMMUNITY) %>%
  summarize(COMM_prev_LIMAT_under18 = mean(DA_prev_LIMAT_under18, na.rm=T),
            .groups = "drop")

#calc changes between the years - make that 1 sf
SAvail_i_COMM_changes <- full_join(SAvail_i_COMM2011 |> st_drop_geometry(),
                                   SAvail_i_COMM2016, by=c("COMMUNITY")) |>
  transmute(COMMUNITY,
            prev_LIMAT_2011 = COMM_prev_LIMAT_under18.x,
            prev_LIMAT_2016 = COMM_prev_LIMAT_under18.y,
            c_prev_LIMAT_perc = ((COMM_prev_LIMAT_under18.y - COMM_prev_LIMAT_under18.x)/
              COMM_prev_LIMAT_under18.x )|> scales::percent())
```

```{r}
#calc changes between the years - make that 1 sf
SAvail_im_COMM_changes <- full_join(SAvail_im_COMM2011|>st_drop_geometry(),
                                          SAvail_im_COMM2016, by=c("COMMUNITY", "mode")) |>
  transmute(COMMUNITY,
            mode = case_when(mode == "c" ~ "motorized",
                          mode == "w" ~ "non-motorized"),
            COMMUNITY_ = dplyr::coalesce(COMMUNITY_.x, COMMUNITY_.y),
            kid_2011 = kid_popadj_m.x,
            kid_2016 = kid_popadj_m.y,
            c_kid_perc = ((kid_popadj_m.y - kid_popadj_m.x)/kid_popadj_m.x) |> scales::percent(),
            kid_proportion_2011 = (kid_2011/sum(kid_2011)),
            kid_proportion_2016 = (kid_2016/sum(kid_2016)),
            c_kid_proportion_perc = ((kid_proportion_2016 - kid_proportion_2011)/kid_proportion_2011)
            |> scales::percent(),
            kid_proportion_2011_perc = (kid_proportion_2011) |> scales::percent(),
            kid_proportion_2016_perc = (kid_proportion_2016) |> scales::percent(),
            V_im_2011 = V_im.x,
            V_im_2016 = V_im.y,
            c_V_im_perc = ((V_im.y -V_im.x)/V_im.x) |> scales::percent(),
            V_im_proportion_2011 = (V_im.x/sum(V_im.x)),
            V_im_proportion_2016 = (V_im.y/sum(V_im.y)),
            c_V_im_proportion_perc = ((V_im_proportion_2016 - V_im_proportion_2011)/V_im_proportion_2011)
            |> scales::percent(),
            V_im_proportion_2011_perc = (V_im_proportion_2011) |> scales::percent(),
            V_im_proportion_2016_perc = (V_im_proportion_2016) |> scales::percent(),
            v_im_2011 = v_im_kid_popadj.x,
            v_im_2016 = v_im_kid_popadj.y,
            c_v_im_perc = ((v_im_kid_popadj.y - v_im_kid_popadj.x)/v_im_kid_popadj.x) |> scales::percent(),
            total_tt_2011 = total_tt.x,
            total_tt_2016 = total_tt.y,
            total_tt_proportion_2011 = (total_tt.x/sum(total_tt.x)),
            total_tt_proportion_2016 = (total_tt.y/sum(total_tt.y)),
            c_total_tt_proportion_perc = ((total_tt_proportion_2016 - total_tt_proportion_2011)/total_tt_proportion_2011)
            |> scales::percent(),
            total_tt_proportion_2011 = (total_tt_proportion_2011) |> scales::percent(),
            total_tt_proportion_2016 = (total_tt_proportion_2016) |> scales::percent(),
            c_total_tt_perc = ((total_tt.y - total_tt.x)/
                                  total_tt.x)|> scales::percent(),
            total_tt_perkid_2011 = total_tt.x/kid_2011,
            total_tt_perkid_2016 = total_tt.y/kid_2016,
            c_total_ttperkid_perc = ((total_tt_perkid_2016 - total_tt_perkid_2011)/
                                  total_tt_perkid_2011)|> scales::percent(),
            emissions_2011 = emissions_total_car_tt.x,
            emissions_2016 = emissions_total_car_tt.y,
            c_emissions_perc = ((emissions_total_car_tt.y - emissions_total_car_tt.x)/
                                  emissions_total_car_tt.x)|> scales::percent(),
            emissions_proportion_2011 = (emissions_2011/sum(emissions_2011)),
            emissions_proportion_2016 = (emissions_2016/sum(emissions_2016)),
            c_emissions_proportion = ((emissions_proportion_2016 - emissions_proportion_2011)/emissions_proportion_2011)
            |> scales::percent(),
            emissions_proportion_2011_perc = (emissions_proportion_2011) |> scales::percent(),
            emissions_proportion_2016_perc = (emissions_proportion_2016) |> scales::percent(),
            emissions_perkid_2011 = emissions_total_car_tt.x/kid_popadj_m.x,
            emissions_perkid_2016 = emissions_total_car_tt.y/kid_popadj_m.y,
            c_emissions_perkid_perc = ((emissions_perkid_2016 - emissions_perkid_2011)/
                                  emissions_perkid_2011)|> scales::percent(),
            geometry) |> 
  left_join(SAvail_i_COMM_changes, by="COMMUNITY") |> st_as_sf() 
```

```{r student-popabs-lowincperc-poprate-plot, eval=FALSE}
tmap_defaults <-
  tm_scale_bar(position = c("LEFT", "BOTTOM"), breaks = c(0, 5, 10, 15))+
  tm_compass(size = 1.5, position = c("right", 'top')) +
  tm_layout(legend.position = c("left", "top")) #legend.hist.size = 10,

student_pop_absolute_plot <-
  tm_shape(SAvail_i_DA_extra |> mutate(extra_col_for_panel = "HAMILTON")) +
  tm_polygons(col = "DA_pop_5to14", border.alpha=0, 
              palette = "Greys",
              labels = c("[0-40)", "[40-50)", "[50-75)","[75-835]"),
              breaks = c(0, 40, 50, 75, 835),
              title = "Population aged 5 to 14"
              )+
  tm_facets(by=c("Year"), ncol=2, nrow=1)+
  tmap_defaults +
  tm_layout(panel.label.height = 0.8)

student_pop_rate_plot <-
  tm_shape(SAvail_i_DA_extra) +
  tm_polygons(col = "kid_pop_DAavg", border.alpha=0, 
              palette = "Blues",
              labels = c("[0-0.0018)", "[0.0018-0.0025)", "[0.0025-0.0038)","[0.0038-0.59]"),
              breaks = c(0, 0.0018, 0.0025, 0.0038, 0.60),
              title = "Average pop. aged 5 to 14\nper residential unit"
              )+
  tm_facets(by=c("Year"), ncol=2, nrow=1)+
  tmap_defaults +
  tm_layout(panel.show=FALSE)

LIMAT_under18_DA_plot <-
  tm_shape(SAvail_i_DA_extra) +
  tm_polygons(col = "DA_prev_LIMAT_under18", border.alpha=0, 
              palette = "Reds",
              labels = c("[0-4.3)", "[4.3-13.0)", "[13.0-30.8)","[30.8-100)"),
              breaks = c(0, 4.30, 13.00, 30.8, 150),
              textNA = "Unavailable",
              title = "Proportion of LIM-AT households\n with children < 18"
              )+
  tm_facets(by=c("Year"), ncol=2, nrow=1)+
  tmap_defaults  +
  tm_layout(panel.show=FALSE)

stud_pop_rate_LIM_plot <- tmap_arrange(student_pop_absolute_plot, student_pop_rate_plot, LIMAT_under18_DA_plot, nrow = 3)

tmap_save(stud_pop_rate_LIM_plot, "Figures/stud_pop_rate_LIM_plot.png", height=7, width=12, dpi=600)
```

```{r importing-taz-and-associated-flows}
#Read the boundaries of the traffic analysis zones (retrieved from the TTS 2016):
ggh_taz <- st_read("data-prep/data-inputs/TTS/tts06_83_region.shp") %>% st_transform(crs = 32617)

# Prepare traffic analysis zones:
ggh_taz <- ggh_taz %>%
  transmute(GTA06 = as.character(GTA06),
            AREA = st_area(geometry) %>% 
              set_units(km^2) %>% 
              drop_units())
#Extract the IDs for all TAZ:
ghta_taz_id <- ggh_taz$GTA06

#Read the boundaries of the traffic analysis zones (retrieved from the TTS 2016):
ggh_pd <- st_read("data-prep/data-inputs/TTS/tts06_pd_83.shp") %>%
  st_transform(crs = 32617)

#Adding region names to the object as they are not included in the origin data: (http://dmg.utoronto.ca/pdf/tts/2016/2016TTS_Conduct.pdf)

ggh_pd <- ggh_pd %>%
  mutate(REGION_name = 
           ifelse(REGION == 1, "Toronto", 
                  ifelse(REGION == 2, "Durham", 
                         ifelse(REGION == 3, "York", 
                                ifelse(REGION == 4, "Peel", 
                                       ifelse(REGION == 5, "Halton", 
                                              ifelse(REGION == 6, "Hamilton", 
                                                     ifelse(REGION == 11, "Niagara", 
                                                            ifelse(REGION == 12, "Waterloo", 
                                                                   ifelse(REGION == 13, "Guelph", 
                                                                          ifelse(REGION == 14, "Wellington", 
                                                                                 ifelse(REGION == 15, "Orangeville", 
                                                                                        ifelse(REGION == 16, "Barrie", 
                                                                                               ifelse(REGION == 17, "Simcoe", 
                                                                                                      ifelse(REGION == 18, "Kawartha Lakes", 
                                                                                                             ifelse(REGION == 19, "Peterborough City", 
                                                                                                                    ifelse(REGION == 20, "Peterborough County", 
                                                                                                                           ifelse(REGION == 21, "Orillia", 
                                                                                                                                  ifelse(REGION == 22, "Dufferin", 
                                                                                                                                         ifelse(REGION == 23, "Brantford", 
                                                                                                                                                ifelse(REGION == 24, "Brant", "NOOOO")))))))))))))))))))))

# add 'REGION' from the planning distract object to the `ggh_taz`. 
#buffering to clean up the edges of these objects -> there wre issues using st_join. 
ggh_taz <- st_buffer(ggh_taz, dist=0)
ggh_pd <- st_buffer(ggh_pd, dist=0)

ggh_taz <- st_join(ggh_taz, ggh_pd %>% transmute(REGION),left = TRUE, largest = TRUE)
ham_taz <- ggh_taz %>% dplyr::filter(REGION==6) #Filter out everything other than Planning District 6 = Hamilton
```

```{r include=FALSE}
#the tts origin-destination trips, by mode
load(file="data-prep/data-inputs/TTS/od_trips_perc_2011.Rdata")
load(file="data-prep/data-inputs/TTS/od_trips_perc_2016.Rdata")

ham_taz_centroids <- ham_taz |> st_centroid() |> dplyr::select(GTA06)

od_trips_perc_car <- rbind(
  rbind(od_trips_perc_2011 |> group_by(Origin) |> 
  summarize(TAZOrig_car_perc = 
              sum(TAZOrig_car_trips, na.rm = T)/sum(TAZOrig_walk_trips+TAZOrig_car_trips, na.rm = T)) |>
  mutate(year = "2011")),
  od_trips_perc_2016 |> group_by(Origin) |> 
  summarize(TAZOrig_car_perc = 
              sum(TAZOrig_car_trips, na.rm = T)/sum(TAZOrig_walk_trips+TAZOrig_car_trips, na.rm = T)) |>
  mutate(year = "2016"))

od_trips_perc_car <- left_join(ham_taz,od_trips_perc_car, by=c("GTA06"="Origin"))

od_trips_perc_walk <- rbind(
  rbind(od_trips_perc_2011 |> group_by(Origin) |> 
  summarize(TAZOrig_walk_perc = 
              sum(TAZOrig_walk_trips)/sum(TAZOrig_walk_trips+TAZOrig_car_trips)) |>
  mutate(year = "2011")),
  od_trips_perc_2016 |> group_by(Origin) |> 
  summarize(TAZOrig_walk_perc = 
              sum(TAZOrig_walk_trips)/sum(TAZOrig_walk_trips+TAZOrig_car_trips)) |>
  mutate(year = "2016"))

od_trips_perc_walk <- left_join(ham_taz, od_trips_perc_walk, by=c("GTA06"="Origin"))

desire_lines_ham_2011 <- od2line(od_trips_perc_2011, ham_taz_centroids)
desire_lines_ham_2011 <- desire_lines_ham_2011 |> mutate(all = TAZOrig_car_trips + TAZOrig_walk_trips,
                                                         Year = "2011")

desire_lines_ham_2016 <- od2line(od_trips_perc_2016, ham_taz_centroids)
desire_lines_ham_2016 <- desire_lines_ham_2016 |> mutate(all = TAZOrig_car_trips + TAZOrig_walk_trips,
                                                         Year = "2016")

desire_lines_ham <- rbind(desire_lines_ham_2011, desire_lines_ham_2016)
rm(desire_lines_ham_2011, desire_lines_ham_2016)
```

```{r ploting-od-taz-flows, eval=FALSE}
car_OD_flow_plot <-  tm_shape(od_trips_perc_car |> mutate(mode = "motorized"), bbox=SAvail_i_DA_extra) +
    tm_polygons(col = "TAZOrig_car_perc", border.alpha = 0.4, border.col = "grey20", lwd=0.2,
              palette = "Greys",
              breaks = c(0, .25, .50, .75, 1),
              title = "Proportion of mode use",
              textNA = "Unavailable",
              colorNA = "white")+
  #tm_shape(SAvail_i_DA_extra) + tm_borders(alpha = 0.4, col = "grey20", lwd=0.2)+
  tm_shape(ham_taz) + tm_borders(alpha=0.4, col = "grey40", lwd=0.2) +
  tm_shape(ham_community_bounds) + tm_borders(alpha = 0.8, col = "lightblue", lwd=2)+
  tm_shape(ham_taz_centroids) + tm_dots(alpha=0.5, col="darkblue", size=0.01) +
  tm_shape(desire_lines_ham |> mutate(mode = "motorized")) +
  tm_lines(palette = "Reds", breaks = c(6, 22, 26, 49, 353),
           title.col = "Number of trips (motorized)",
           col = "TAZOrig_car_trips",
           alpha = 0.9,
           scale = 3,
           lwd = "all",
           legend.lwd.show = FALSE) +
  tm_facets(by = c("mode","Year"), nrow=1)+
  tmap_defaults +
  tm_layout(panel.label.height = 0.8)

walk_OD_flow <- tm_shape(od_trips_perc_walk |> mutate(mode = "non-motorized"), bbox=SAvail_i_DA_extra) +
      tm_polygons(col = "TAZOrig_walk_perc", border.alpha = 0.4, border.col = "grey20", lwd=0.2,
              palette = "Greys",
              breaks = c(0, .25, .50, .75, 1),
              title = "Proportion of mode use",
              textNA = "Unavailable",
              colorNA = "white",
              legend.show=FALSE)+
  #tm_shape(SAvail_i_DA_extra) + tm_borders(alpha = 0.4, col = "cornflowerblue", lwd=0.2)+
  tm_shape(ham_taz) + tm_borders(alpha=0.4, col = "grey40", lwd=0.2) +
  tm_shape(ham_community_bounds) + tm_borders(alpha = 0.8, col = "lightblue", lwd=2)+
  tm_shape(ham_taz_centroids) + tm_dots(alpha=0.5, col="darkblue", size=0.01) +
  tm_shape(desire_lines_ham |> filter(TAZOrig_walk_trips >= 1) 
           |> mutate(mode = "non-motorized")) +
  tm_lines(palette = "Greens", breaks = c(12, 23, 44, 67, 174),
           title.col = "Number of trips (non-motorized)",
           col = "TAZOrig_walk_trips",
           alpha = 0.9,
           scale = 4,
           lwd = "all",
           legend.lwd.show = FALSE) +
  tm_facets(by = c("mode","Year"), nrow=1)+
  tmap_defaults +
  tm_layout(panel.label.height = c(0.8,0,0)) +
  tm_add_legend(type = "line", 
                col = c("grey40","lightblue"),
                labels = c("TAZ boundary","Community boundary"))+
  tm_add_legend(type = "symbol",
                col = "darkblue",
                labels = "TAZ centroid")
OD_flow_plot <- tmap_arrange(car_OD_flow_plot, walk_OD_flow, nrow = 2)
tmap_save(OD_flow_plot, "Figures/OD_flow_plot.png", height=7, width=12, dpi=600)
```

```{r quartiles-of-tts-estimated-tt}
od_trips_perc_tt <- data.frame(x=c(gamma2011_motor$data, exp2011_nonmotor$data,
                                   gamma2016_motor$data, exp2016_nonmotor$data),
                               mode = c(rep("motorized", length(gamma2011_motor$data)),
                                        rep("non-motorized", length(exp2011_nonmotor$data)),
                                        rep("motorized", length(gamma2016_motor$data)),
                                        rep("non-motorized", length(exp2016_nonmotor$data))),
                               year = c(rep("2011", length(gamma2011_motor$data)),
                                        rep("2011", length(exp2011_nonmotor$data)),
                                        rep("2016", length(gamma2016_motor$data)),
                                        rep("2016", length(exp2016_nonmotor$data)))
                               )

q_motorized_2011 <- stats::quantile(od_trips_perc_tt |>
                                      filter(mode == "motorized", year == "2011") |>
                                      pull(x),
                                    probs = c(0.25, 0.5, 0.75), na.rm=T)

q_nonmotorized_2011 <- stats::quantile(od_trips_perc_tt  |>
                                         filter(mode == "non-motorized",
                                                year == "2011") |> 
                                         pull(x),
                     probs = c(0.25, 0.5, 0.75), na.rm=T)

q_motorized_2016 <- stats::quantile(od_trips_perc_tt  |>
                                      filter(mode == "motorized",
                                             year == "2016") |>
                                      pull(x),
                     probs = c(0.25, 0.5, 0.75), na.rm=T)

q_nonmotorized_2016 <- stats::quantile(od_trips_perc_tt |> 
                                         filter(mode == "non-motorized",
                                                year == "2016") |>
                                         pull(x),
                     probs = c(0.25, 0.5, 0.75), na.rm=T)

q_1 <- data.frame(
  label = paste0("Q1:\n", c(q_motorized_2011[1], q_nonmotorized_2011[1]|> round(),
                            q_motorized_2016[1], q_nonmotorized_2016[1]|> round())),
  mode   = c("motorized", "non-motorized", "motorized", "non-motorized"),
  year = c("2011", "2011", "2016", "2016"),
  x     = c(q_motorized_2011[1], q_nonmotorized_2011[1] |> round(),
            q_motorized_2016[1], q_nonmotorized_2016[1]|> round()),
  y     = c(0.11, 0.11, 0.11, 0.11)
)

q_2 <- data.frame(
  label = paste0("Q2:\n", c(q_motorized_2011[2], q_nonmotorized_2011[2]|> round(),
                            q_motorized_2016[2], q_nonmotorized_2016[2]|> round())),
  mode   = c("motorized", "non-motorized", "motorized", "non-motorized"),
  year = c("2011", "2011", "2016", "2016"),
  x     = c(q_motorized_2011[2], q_nonmotorized_2011[2] |> round(),
            q_motorized_2016[2], q_nonmotorized_2016[2]|> round()),
  y     = c(0.11, 0.11, 0.11, 0.11)
)

q_3 <- data.frame(
  label = paste0("Q3:\n", c(q_motorized_2011[3], q_nonmotorized_2011[3]|> round(),
                            q_motorized_2016[3], q_nonmotorized_2016[3]|> round())),
  mode   = c("motorized", "non-motorized", "motorized", "non-motorized"),
  year = c("2011", "2011", "2016", "2016"),
  x     = c(q_motorized_2011[3], q_nonmotorized_2011[3] |> round(),
            q_motorized_2016[3], q_nonmotorized_2016[3]|> round()),
  y     = c(0.11, 0.11, 0.11, 0.11)
)
```
```{r tts-theoretical-curve-creation}
x <- data.frame(t = seq(0, 45, 0.1))
X_shorter <-data.frame(t = seq(0, 27, 0.1))

x_motorized_2011 <- x %>%
  mutate(f = dgamma(t,
                    gamma2011_motor$estimate[1],
                    gamma2011_motor$estimate[2]),
         mode = "motorized",
         year = "2011")

x_nonmotorized_2011 <- X_shorter %>%
  mutate(f = dexp(t,
                    exp2011_nonmotor$estimate[1]),
         mode = "non-motorized",
         year = "2011")

x_motorized_2016 <- x %>%
  mutate(f = dgamma(t,
                    gamma2016_motor$estimate[1],
                    gamma2016_motor$estimate[2]),
         mode = "motorized",
         year = "2016")

x_nonmotorized_2016 <- X_shorter %>%
  mutate(f = dexp(t,
                    exp2016_nonmotor$estimate[1]),
         mode = "non-motorized",
         year = "2016")

tld_theoretical <- rbind(x_motorized_2011,
      x_nonmotorized_2011,
      x_motorized_2016,
      x_nonmotorized_2016) %>%
  mutate(mode = factor(mode, levels = c("motorized", "non-motorized")),
         year = factor(year, levels = c("2011", "2016")))
```

```{r ploting-empirical-and-theo-curve-TLD-tts, eval=FALSE}
TLD_empirical_wtheo_curve_plot <- ggplot() + 
  geom_histogram(data = od_trips_perc_tt,
                 aes(x = x, 
                     y = ..density..),
                 colour = "grey80",
                 alpha = 0.5,
                 boundary = 0,
                 binwidth = 3) +
  geom_vline(data = q_1, mapping = aes(xintercept = x), colour = "grey60") +
  geom_vline(data = q_2, mapping = aes(xintercept = x), colour = "grey60") +
  geom_vline(data = q_3, mapping = aes(xintercept = x), colour = "grey60") +
  geom_line(data = tld_theoretical, aes(x = t, y = f), colour = "blue", size=0.8) + 
  xlab("Estimated TTS travel time (minutes)") +
  scale_x_continuous(breaks = seq(0,40,10), expand=c(0,0))+
  scale_y_continuous(breaks = seq(0,0.12,0.04), limits = c(0,0.12))+
  facet_grid(~mode~year)+
  geom_text(data = q_1, 
            mapping = aes(x = x, y = y, label = label), colour = "grey20", size=3) +
  geom_text(data = q_2,
            mapping = aes(x = x, y = y, label = label), colour = "grey20", size=3) +
  geom_text(data = q_3,
            mapping = aes(x = x, y = y, label = label), colour = "grey20", size=3) +
  theme_classic()
TLD_empirical_wtheo_curve_plot
ggsave("Figures/TLD_empirical_wtheo_curve_plot.png", height=7, width=12, dpi=600)
```

```{r, eval=FALSE}
Vim_20112016_plot <-  tm_shape(SAvail_im_DA, bbox=SAvail_im_DA) +
    tm_polygons(col = "V_im", border.alpha = 0, border.col = "grey20", lwd=0.2,
              palette = "Greys",
              style="quantile", n=4,
              title = expression(V[i]^m*' (quartiles)'),
              textNA = "Unavailable",
              colorNA = "white")+
  tm_facets(by = c("mode","year"), nrow=1)+
  tm_shape(ALL_SCHOOLS_Elem_formapping) +
  tm_dots(col = "Status", size=0.03, title = "School status between\n2011 and 2016",
          palette = c("ivory", "black", "darkgreen","#0E86D4","#90EE90"),
          labels = c("No change","Closed before 2016", "Expanded before 2016","Moved before 2016","New after 2011"), legend.show=FALSE) +
  tm_shape(ham_community_bounds) + tm_borders(alpha = 0.8, col = "lightblue", lwd=1.5)+
  tmap_defaults +
  tm_layout(panel.label.height = 0.8)

vvim_20112016_plot <-  tm_shape(SAvail_im_DA, bbox=SAvail_im_DA) +
    tm_polygons(col = "v_im_kid_popadj", border.alpha = 0, border.col = "grey20", lwd=0.2,
              palette = "RdYlGn",
              midpoint = 1.0,
              title = expression(v[i]^m),
              breaks = c(0, 0.5, 1, 1.5, 2, 2.1), #note the highest value is ~12, I'm breaking the values early for the purpose of visualisation
              labels = c("[0-0.5)", "[0.5-1)", "[1-1.5)", "[1.5-2)", "[>2)"),
              textNA = "Unavailable",
              colorNA = "white")+
  tm_facets(by = c("mode","year"), nrow=1)+
  tm_shape(ALL_SCHOOLS_Elem_formapping) +
  tm_dots(col = "Status", size=0.03, title = "School status between\n2011 and 2016",
          palette = c("ivory", "black", "darkgreen","#0E86D4","#90EE90"),
          labels = c("No change","Closed before 2016", "Expanded before 2016","Moved before 2016","New after 2011")          ) +
  tm_shape(ham_community_bounds) + tm_borders(alpha = 0.8, col = "lightblue", lwd=1.5)+
  tmap_defaults +
  tm_layout(panel.label.height = 0.8)

Vim_vvim_20112016_plot <- tmap_arrange(Vim_20112016_plot, vvim_20112016_plot, nrow=2)

tmap_save(Vim_vvim_20112016_plot, "Figures/Vim_vvim_20112016_plot.png", height=7, width=12, dpi=600)
```

```{r, eval=FALSE}
Vim_20112016_CHANGE_plot <-  tm_shape(SAvail_im_DA_changes, bbox=SAvail_im_DA) +
    tm_polygons(col = "c_V_im", border.alpha = 0, border.col = "grey20", lwd=0.2,
              palette = "-PuOr", midpoint=0,
              breaks = c(-5, -4, -3, -2, -1, 0, 1, 2,3,4,5),
              labels = c("[>-5 to -4)", "[-4 to -3)", "[-3 to -2)","[-2 to -1)","[-1 to 0)", "[0 to 1)", 
                         "[1 to 2)", "[2 to 3)", "[3 to 4)", "[4 to <5)" ),
              title = expression("Change in "*V[i]^m),
              textNA = "Unavailable",
              colorNA = "white")+
  tm_facets(by = c("mode"), nrow=1)+
  tm_shape(ALL_SCHOOLS_Elem_formapping |> 
             filter(Status != "NoChange" & Status != "Moved")|>
             mutate(Status_change = case_when(Status == "Removed" ~ "Closed",
                                              (Status == "Expanded" | Status == "New" ) ~ "New or Expanded"))) +
  tm_dots(col = "Status_change", size=0.05, title = "Schools that changed status",
          palette = c("black", "#90EE90"), legend.show=FALSE) +
  tm_shape(ham_community_bounds) + tm_borders(alpha = 0.8, col = "lightblue", lwd=2)+
  tmap_defaults +
  tm_layout(panel.label.height = 0.8)

#tmap_save(Vim_20112016_CHANGE_plot, "Figures/Vim_20112016_CHANGE_plot.png", height=7, width=12, dpi=600)

vvim_20112016_CHANGE_plot <- tm_shape(SAvail_im_DA_changes, bbox=SAvail_im_DA) +
    tm_polygons(col = "c_v_im_kid_popadj", border.alpha = 0, border.col = "grey20", lwd=0.2,
              palette = "RdYlGn", midpoint=0,
              breaks = c(-0.1, -0.05, 0.00, 0.05, 0.1),
              labels = c("[>-0.1 to -0.05)", "[-0.05 to 0.00)", "[0.00 to 0.05)", "[0.05 to <0.1)"),
              title = expression("Change in "*v[i]^m),
              textNA = "Unavailable",
              colorNA = "white")+
  tm_facets(by = c("mode"), nrow=1)+
  tm_shape(ALL_SCHOOLS_Elem_formapping |> 
             filter(Status != "NoChange" & Status != "Moved")|>
             mutate(Status_change = case_when(Status == "Removed" ~ "Closed",
                                              (Status == "Expanded" | Status == "New" ) ~ "New or Expanded"))) +
  tm_dots(col = "Status_change", size=0.05, title = "Schools that changed status",
          palette = c("black", "#90EE90")) +
  tm_shape(ham_community_bounds) + tm_borders(alpha = 0.8, col = "lightblue", lwd=2)+
  tmap_defaults +
  tm_layout(panel.label.height = 0.8)

Vim_vvim_20112016_CHANGE_plot <- tmap_arrange(Vim_20112016_CHANGE_plot, vvim_20112016_CHANGE_plot, nrow=2)

tmap_save(Vim_vvim_20112016_CHANGE_plot, "Figures/Vim_vvim_20112016_CHANGE_plot.png", height=7, width=12, dpi=600)
```

```{r, eval=FALSE}
walk_OD_flow <- tm_shape(od_trips_perc_walk |> mutate(mode = "non-motorized"), bbox=SAvail_i_DA_extra) +
      tm_polygons(col = "TAZOrig_walk_perc", border.alpha = 0.4, border.col = "grey20", lwd=0.2,
              palette = "Greys",
              breaks = c(0, .25, .50, .75, 1),
              title = "Proportion of mode use",
              textNA = "Unavailable",
              colorNA = "white",
              legend.show=FALSE)+
  #tm_shape(SAvail_i_DA_extra) + tm_borders(alpha = 0.4, col = "cornflowerblue", lwd=0.2)+
  tm_shape(ham_taz) + tm_borders(alpha=0.4, col = "grey40", lwd=0.2) +
  tm_shape(ham_community_bounds) + tm_borders(alpha = 0.8, col = "lightblue", lwd=2)+
  tm_shape(ham_taz_centroids) + tm_dots(alpha=0.5, col="darkblue", size=0.01) +
  tm_shape(desire_lines_ham |> filter(TAZOrig_walk_trips >= 1) 
           |> mutate(mode = "non-motorized")) +
  tm_lines(palette = "Greens", breaks = c(12, 23, 44, 67, 174),
           title.col = "Number of trips (non-motorized)",
           col = "TAZOrig_walk_trips",
           alpha = 0.9,
           scale = 4,
           lwd = "all",
           legend.lwd.show = FALSE) +
  tm_facets(by = c("mode","Year"), nrow=1)+
  tmap_defaults +
  tm_layout(panel.label.height = c(0.8,0,0)) +
  tm_add_legend(type = "line", 
                col = c("grey40","lightblue"),
                labels = c("TAZ boundary","Community boundary"))+
  tm_add_legend(type = "symbol",
                col = "darkblue",
                labels = "TAZ centroid")
OD_flow_plot <- tmap_arrange(car_OD_flow_plot, walk_OD_flow, nrow = 2)
tmap_save(OD_flow_plot, "Figures/SA-and-SAperStud-20112016.png", height=7, width=12, dpi=600)
```

```{r, eval=FALSE}
SA_Vim_2011_plot <- ggplot() +
  geom_sf(data = SAvail_im_DA2011, aes(fill= V_im), color = NA) +
  scale_fill_binned(type = "viridis", na.value = "grey60",
                       breaks = c(min(SAvail_im_DA2011$V_im),
                                  quantile(SAvail_im_DA2011$V_im,0.25)%>%as.numeric(),
                                  median(SAvail_im_DA2011$V_im, na.rm=T),
                                  quantile(SAvail_im_DA2011$V_im,0.75)%>%as.numeric(),
                                  max(SAvail_im_DA2011$V_im, na.rm = T)))+
  geom_sf(data = ALL_SCHOOLS_Elem |> filter(Year != 2016), 
          size=0.5, colour = "Red", alpha=0.7)+
  geom_sf(data = ham_community_bounds, fill=NA, color = "grey30", alpha=0.7)+
  theme_minimal() +
  theme(legend.title = element_text(size=10),
                axis.text = element_blank(), #removes lat/long from the axises
                panel.grid.major=element_blank())+
  facet_wrap(~mode)

SA_Vim_2016_plot <- ggplot() +
  geom_sf(data = SAvail_im_DA2016, aes(fill= V_im), color = NA) +
  scale_fill_binned(type = "viridis", na.value = "grey60",
                       breaks = c(min(SAvail_im_DA2016$V_im),
                                  quantile(SAvail_im_DA2016$V_im,0.25)%>%as.numeric(),
                                  median(SAvail_im_DA2016$V_im, na.rm=T),
                                  quantile(SAvail_im_DA2016$V_im,0.75)%>%as.numeric(),
                                  max(SAvail_im_DA2016$V_im, na.rm = T)))+
  geom_sf(data = ALL_SCHOOLS_Elem |> filter(Year != 2011), 
          size=0.5, colour = "Red", alpha=0.7)+
  geom_sf(data = ham_community_bounds, fill=NA, color = "grey30", alpha=0.7)+
  theme_minimal() +
  theme(legend.title = element_text(size=10),
                axis.text = element_blank(), #removes lat/long from the axises
                panel.grid.major=element_blank())+
  facet_wrap(~mode)
```
```{r, eval=FALSE}
SA_vim_2011_plot <- ggplot() +
  geom_sf(data = SAvail_im_DA2011, aes(fill= v_im_kid_popadj), color = NA) +
   scale_fill_binned(type = "viridis",
                     breaks = c(min(SAvail_im_DA2011$v_im_kid_popadj, na.rm = T),
                                  quantile(SAvail_im_DA2011$v_im_kid_popadj,0.25)%>%as.numeric(),
                                  median(SAvail_im_DA2011$v_im_kid_popadj, na.rm=T),
                                  quantile(SAvail_im_DA2011$v_im_kid_popadj,0.75)%>%as.numeric(),
                                  max(SAvail_im_DA2011$v_im_kid_popadj, na.rm = T)))+
  geom_sf(data = ALL_SCHOOLS_Elem |> filter(Year != 2016), 
          size=0.5, colour = "Red", alpha=0.7)+
  geom_sf(data = ham_community_bounds, fill=NA, color = "grey30", alpha=0.7)+
  theme_minimal() +
  theme(legend.title = element_text(size=10),
                axis.text = element_blank(), #removes lat/long from the axises
                panel.grid.major=element_blank())+
  facet_wrap(~mode)

SA_vim_2016_plot <- ggplot() +
  geom_sf(data = SAvail_im_DA2016, aes(fill= v_im_kid_popadj), color = NA) +
   scale_fill_binned(type = "viridis",
                     breaks = c(min(SAvail_im_DA2016$v_im_kid_popadj, na.rm = T),
                                  quantile(SAvail_im_DA2016$v_im_kid_popadj,0.25)%>%as.numeric(),
                                  median(SAvail_im_DA2016$v_im_kid_popadj, na.rm=T),
                                  quantile(SAvail_im_DA2016$v_im_kid_popadj,0.75)%>%as.numeric(),
                                  max(SAvail_im_DA2016$v_im_kid_popadj, na.rm = T)))+
  geom_sf(data = ALL_SCHOOLS_Elem |> filter(Year != 2011), 
          size=0.5, colour = "Red", alpha=0.7)+
  geom_sf(data = ham_community_bounds, fill=NA, color = "grey30", alpha=0.7)+
  theme_minimal() +
  theme(legend.title = element_text(size=10),
                axis.text = element_blank(), #removes lat/long from the axises
                panel.grid.major=element_blank())+
  facet_wrap(~mode)
```

```{r, eval=FALSE}
SA_vim_LIMAT_2011_plot <- ggplot() +
  geom_sf(data = SAvail_im_DA2011, aes(fill= v_im_kid_popadj_LIMAT), color = NA) +
  scale_fill_binned(type = "viridis", 
                       breaks = c(min(SAvail_im_DA2011$v_im_kid_popadj_LIMAT, na.rm = T),
                                  quantile(SAvail_im_DA2011$v_im_kid_popadj_LIMAT,0.25)%>%as.numeric()+0.000000001,
                                  median(SAvail_im_DA2011$v_im_kid_popadj_LIMAT, na.rm=T),
                                  quantile(SAvail_im_DA2011$v_im_kid_popadj_LIMAT,0.75)%>%as.numeric(),
                                  max(SAvail_im_DA2011$v_im_kid_popadj_LIMAT, na.rm = T)))+
  geom_sf(data = ALL_SCHOOLS_Elem |> filter(Year != 2016), 
          size=0.5, colour = "Red", alpha=0.7)+
  geom_sf(data = ham_community_bounds, fill=NA, color = "grey30", alpha=0.7)+
  theme_minimal() +
  theme(legend.title = element_text(size=10),
                axis.text = element_blank(), #removes lat/long from the axises
                panel.grid.major=element_blank())+
  facet_wrap(~mode)

SA_vim_LIMAT_2016_plot <- ggplot() +
  geom_sf(data = SAvail_im_DA2016, aes(fill= v_im_kid_popadj_LIMAT), color = NA) +
  scale_fill_binned(type = "viridis", 
                       breaks = c(min(SAvail_im_DA2016$v_im_kid_popadj_LIMAT, na.rm = T),
                                  quantile(SAvail_im_DA2016$v_im_kid_popadj_LIMAT,0.25)%>%as.numeric()+0.000000001,
                                  median(SAvail_im_DA2016$v_im_kid_popadj_LIMAT, na.rm=T),
                                  quantile(SAvail_im_DA2016$v_im_kid_popadj_LIMAT,0.75)%>%as.numeric(),
                                  max(SAvail_im_DA2016$v_im_kid_popadj_LIMAT, na.rm = T)))+
  geom_sf(data = ALL_SCHOOLS_Elem |> filter(Year != 2011), 
          size=0.5, colour = "Red", alpha=0.7)+
  geom_sf(data = ham_community_bounds, fill=NA, color = "grey30", alpha=0.7)+
  theme_minimal() +
  theme(legend.title = element_text(size=10),
                axis.text = element_blank(), #removes lat/long from the axises
                panel.grid.major=element_blank())+
  facet_wrap(~mode)

SA_Vim_LIMAT_2011_plot <- ggplot() +
  geom_sf(data = SAvail_im_DA2011, aes(fill= V_i_kid_popadj_LIMAT), color = NA) +
  scale_fill_binned(type = "viridis", 
                       breaks = c(min(SAvail_im_DA2011$V_i_kid_popadj_LIMAT, na.rm = T),
                                  quantile(SAvail_im_DA2011$V_i_kid_popadj_LIMAT,0.25)%>%as.numeric()+0.000000001,
                                  median(SAvail_im_DA2011$V_i_kid_popadj_LIMAT, na.rm=T),
                                  quantile(SAvail_im_DA2011$V_i_kid_popadj_LIMAT,0.75)%>%as.numeric(),
                                  max(SAvail_im_DA2011$V_i_kid_popadj_LIMAT, na.rm = T)))+
  geom_sf(data = ALL_SCHOOLS_Elem |> filter(Year != 2016), 
          size=0.5, colour = "Red", alpha=0.7)+
  geom_sf(data = ham_community_bounds, fill=NA, color = "grey30", alpha=0.7)+
  theme_minimal() +
  theme(legend.title = element_text(size=10),
                axis.text = element_blank(), #removes lat/long from the axises
                panel.grid.major=element_blank())+
  facet_wrap(~mode)

SA_Vim_LIMAT_2016_plot <- ggplot() +
  geom_sf(data = SAvail_im_DA2016, aes(fill= V_i_kid_popadj_LIMAT), color = NA) +
  scale_fill_binned(type = "viridis", 
                       breaks = c(min(SAvail_im_DA2016$V_i_kid_popadj_LIMAT, na.rm = T),
                                  quantile(SAvail_im_DA2016$V_i_kid_popadj_LIMAT,0.25)%>%as.numeric()+0.000000001,
                                  median(SAvail_im_DA2016$V_i_kid_popadj_LIMAT, na.rm=T),
                                  quantile(SAvail_im_DA2016$V_i_kid_popadj_LIMAT,0.75)%>%as.numeric(),
                                  max(SAvail_im_DA2016$V_i_kid_popadj_LIMAT, na.rm = T)))+
  geom_sf(data = ALL_SCHOOLS_Elem |> filter(Year != 2011), 
          size=0.5, colour = "Red", alpha=0.7)+
  geom_sf(data = ham_community_bounds, fill=NA, color = "grey30", alpha=0.7)+
  theme_minimal() +
  theme(legend.title = element_text(size=10),
                axis.text = element_blank(), #removes lat/long from the axises
                panel.grid.major=element_blank())+
  facet_wrap(~mode)
```
```{r, eval=FALSE}
SA_Vim_2011_plot
SA_vim_2011_plot
SA_Vim_LIMAT_2011_plot
SA_vim_LIMAT_2011_plot

SA_Vim_2016_plot
SA_vim_2016_plot
SA_Vim_LIMAT_2016_plot
SA_vim_LIMAT_2016_plot
#*summarize LIM at city level
```

```{r, eval=FALSE}
SAvail_im_DA2011  |> st_drop_geometry() |> group_by(COMMUNITY, mode) |> summarise(
  LIM = mean(DA_prev_LIMAT_under18, na.rm = T)*mean(kid_popadj_m, na.rm = T),
  sum_total_tt = sum(total_tt),
  emis_perkid = sum(emissions_total_car_tt)/sum(kid_popadj_m),
  sum_V_im = sum(V_im),
  mn_v_im = mean(v_im_kid_popadj),
  mn_v_im_LIM = mean(v_im_kid_popadj_LIMAT))

SAvail_im_DA2016  |> st_drop_geometry() |> group_by(COMMUNITY, mode) |> summarise(
  LIM = mean(DA_prev_LIMAT_under18, na.rm = T)*mean(kid_popadj_m, na.rm = T),
  sum_total_tt = sum(total_tt),
  emis_perkid = sum(emissions_total_car_tt)/sum(kid_popadj_m),
  sum_V_im = sum(V_im),
  mn_v_im = mean(v_im_kid_popadj),
  mn_v_im_LIM = mean(v_im_kid_popadj_LIMAT))                                                
```

```{r creating-desc-stats-table-tt, echo=FALSE}
#forming a complete descriptive statistic table
# Statistics <- data.frame("Statistic" = c("Parcel pts", "Trips",
#                                          "TT min.", "TT 1st qu.", "TT mean", "TT 3rd qu.", "TT max."))
# `2011` <- data.frame(`Y2011` = c(ALL_nearest_network_point %>% filter(Year != "2016") %>% pull(ROLL_NUM) %>% length() %>% round() ,
#                                 ALL_PARCELS %>% filter(ParYear != "2016" & Level == "Elementary") %>% pull(ROLL_NUM) %>% length()%>% round(),
#                                 ALL_PARCELS %>% filter(ParYear != "2016" & Level == "Elementary") %>% pull(car_tt) %>% min()%>% round(),
#                                 ALL_PARCELS %>% filter(ParYear != "2016" & Level == "Elementary") %>% pull(car_tt) %>% quantile(0.25) %>% as.numeric()%>% round(), 
#                                 ALL_PARCELS %>% filter(ParYear != "2016" & Level == "Elementary") %>% pull(car_tt) %>% mean()%>% round(),
#                                 ALL_PARCELS %>% filter(ParYear != "2016" & Level == "Elementary") %>% pull(car_tt) %>% quantile(0.75) %>% as.numeric()%>% round(),
#                                 ALL_PARCELS %>% filter(ParYear != "2016" & Level == "Elementary") %>% pull(car_tt) %>% max()%>% round()
#                                 ))
# `2016` <- data.frame(`Y2016` = c(ALL_nearest_network_point %>% filter(Year != "2011") %>% pull(ROLL_NUM) %>% length()%>% round(),
#                                 ALL_PARCELS %>% filter(ParYear != "2011" & Level == "Elementary") %>% pull(ROLL_NUM) %>% length()%>% round(),
#                                 ALL_PARCELS %>% filter(ParYear != "2011" & Level == "Elementary") %>% pull(car_tt) %>% min()%>% round(),
#                                 ALL_PARCELS %>% filter(ParYear != "2011" & Level == "Elementary") %>% pull(car_tt) %>% quantile(0.25) %>% as.numeric()%>% round(), 
#                                 ALL_PARCELS %>% filter(ParYear != "2011" & Level == "Elementary") %>% pull(car_tt) %>% mean()%>% round(),
#                                 ALL_PARCELS %>% filter(ParYear != "2011" & Level == "Elementary") %>% pull(car_tt) %>% quantile(0.75) %>% as.numeric()%>% round(),
#                                 ALL_PARCELS %>% filter(ParYear != "2011" & Level == "Elementary") %>% pull(car_tt) %>% max()%>% round()
#                                 ))
# desc_stats_tt <- cbind(Statistics,`2011`, `2016`)
# 
# #kable tabling
# desc_stats_tt %>%
#   kableExtra::kable(format = "latex",
#         align="lrrrrrr",
#         booktabs = T,
#         col.names = c("Statistic", "2011", "2016"),
#         caption = "\\label{tab:desc-stats-table-tt}Descriptive statistics of the parcels, trips, and calculated origin-destination car travel times.") %>%
#   kableExtra::kable_styling(full_width = "T",
#                 latex_options = c("scale_down"),
#                 position = "center")
```

<!--
Additional methods considerations: circuity of trip by mode? Schoolbus is not as direct of a trip as car/taxi/bikeORwalk. Maybe we consider 3 modes instead of 2? The r5r travel time for car/taxi as well as walk/cycle can be used as is, but the travel time by school bus can be increased based on a trip distance factor (from TTS)?

RESULTS an alternative layout idea?:
1. Multimodal spatial availability aggregated at the DA for each year and for each mode. Each DA is ~200 parcel points with their low-income and other-income spatial availability per mode; both these are summed together along with the rest of the parcel points per da. (6 plots - 3 modes per 2011 and 3 modes per 2016). 
2. Income-specific spatial availability aggregated at the DA for each population type. Similarly, each DA is ~200 parcel points with 3 mode-specific spatial availability per income-group; all three are summed together along with the rest of the parcel points per da.  (4 plots - 2 population-groups per 2011 and 2 population-groups per 2016).
3. EXAMINING THE INTERMEDIATES: the change in spatial availability per focused communities:
  3.a. Hamilton-Central+Dundas experienced the majority of school closures, also have the highest LICO-AT as well as motorized transport. Glanbrook+StoneyCreek experienced the majority of new schools (two), also have low LICO-AT as well as non-motorized transport. So. 2 columns.  First column SA By Mode. SA By Mode in 2016 for HC+Dundas, then SA By Mode in 2016 for Glan+SC, then SA By Mode in 2016 Overall. Second column SA By Pop. SAY by LICO or not LICO in 2016 for HC+Dundas, then for GLAN+SC, then for overall. Showing SA for 2016 with brackets that say (+ or - X% to indicate difference from 2011).  
  3.b. Examining the change in travel time per focused communities: same bar graph style as in 3.a. Will show have non-motorized travel time decreased in Ham-central, esp for LICO(?)
  3.c. Change in emissions: same bar graph style as in 3.a. but for emissions. Will show have emissions increased, esp in Ham-central and LICO(?)
4. USING SPATIAL AVAILABILITY FOR PLANNING. Equity? We want to plan for school-seat availability _for_ active modes. Especially in areas with currently high LICO-AT. So lets plot spatial availability per non-motorized-LICO-AT capita. We'd like to see 1.2 SA per capita everywhere. We need improvements to the population, opportunities, and non motorized transport network (relative to all other modes/population groups) is the key point. -->

# Introduction

Communities are by nature in flux: economic and demographic shifts, rural and urban configurations, and political pressures stoke changes in community characteristics. These forces spur the re-organization and consolidation of public facilities in communities around the world where changes to service provision may improve the quality of service for some but not for all [@Christiaanse_2020; @Rosik_Pulawska-Obiedowska_Goliszek_2021]. Oftentimes, top-down consolidation decisions that are driven by operational cost-savings do not taken into consideration the full extent of benefit loss.

From this perspective, journey-to-school trips are tremendously important: this trip purpose is the largest contributor of carbon emissions from personal travel after journey-to-work trips [@rongReviewResearchLowcarbon2022; @pantelakiImpactHomeschoolCommuting2024]. In many communities globally, school locations are closing [@sageman2022] and trips made to school are increasingly motorized [@rongReviewResearchLowcarbon2022]. Nonetheless, school closure policies have been pursued by governments to optimize per-student operational cost-efficiency [@rongReviewResearchLowcarbon2022] but spatial and travel implications have often been overlooked by decision makers [@bierbaumMobilityJusticeLinking2021; @Lee_Lubienski_2017]. For instance, 65% of comprehensive schools (e.g. grades 1 through 9) in rural Finland since 1990 closed [@Autti_Hyry-Beihammer_2014], 10% of elementary schools (kindergarten through grade 8) in Chicago, USA in 2013 [@Lee_Lubienski_2017], and 7% of primary and lower secondary schools (kindergarten through grade 9) were closed in Denmark between 2010-2011 [@Beuchert_Humlum_Nielsen_Smith_2018].

In this study we focus on the spatial travel implications of school closure and consolidation policies. The quantification of lost potential active transportation trips and associated carbon emissions, due to school closure, and the unequal impact attributed to the increased reliance on a motorized commute is scarce in the literature. In this way, our contribution is both methodological and empirical. We quantify the accessibility changes before- and after the first wave of school closure/consolidations implemented in Hamilton, a mid-size city (\~570,000 pop) with rural, suburban and urban characteristics in Ontario, Canada [@governmentofcanadaProfileTableCensus2022].

In 2013, the City of Hamilton's largest English school board the Hamilton-Wentworth District School Board (HWDSB), serving the majority of students in the city, released their Long Term Facilities Master Plan [@hwdsbLongTermFacilities2013]. To ensure "equitable, affordable and sustainable learning facilities", the Master Plan indicated that 80% of it's elementary schools (i.e., schools attended by students aged 5 to 13) would be subject to accommodation strategies, including a "Accommodation Review" whose outcome is possibly closure/consolidation [@hwdsbLongTermFacilities2013; @craggsBoardEvaluatingFuture2013; @seasonsSchoolClosurePolicy2014]. These decisions were in part motivated by operational savings based on a projected reduction in student population and under-utilized school capacity [@craggsAxeCouldFall2012]. However, the case of Hamilton and the HWDSB is unlike others in the province: the HWDSB underwent an unprecedented number of Accommodation Reviews relative to other school boards [@seasons2014ARCCatalogues2014]. Though the Accommodation Review guidelines outline a community-focused process to deliberate a school's future viability to students, the community and the economy [@seasonsSchoolClosurePolicy2014; @ministryofeducationPupilAccommodationReview2006], concerned residences felt a fulsome breakdown of costs to repair schools were not made consistently available [@kleinhuisSeriousConcernsSchool2013]. Further, others pointed to flaud incentive structures motivating the HWDSB: "By closing three schools the HWDSB would have a strong case with the Ministry of Education to receive full funding for a new school" [@kleinhuisSeriousConcernsSchool2013]. Along with insufficient funding to maintain schools in a state of good repair [@auditorgeneralofontarioAnnualReport20152015], the operational savings as a case for school closures is a long-standing critique of the school "funding formula" established by the conservative provincial government in 1998 [@mackenzieBlueprintFixOntario2018]. In a backdrop of budget cuts, the funding formula pits financing of school maintenance and student needs against the financial compensation of staff and powers of local school boards (as they no longer have the ability to levee property taxes) [@mackenzieBlueprintFixOntario2018].

In the first wave of school accommodation reviews between 2013 and 2016, `r ALL_SCHOOLS_Elem %>% filter(Year == "2011" & Level == "Elementary" & Status == "Removed") %>% pull(SchoolID) %>% length()` elementary schools closed. From between 2011 (before the school closures) to 2016, these closures represented a `r ((ALL_SCHOOLS_Elem %>% filter(Year != "2016") %>% pull(SchoolID) %>% length() - ALL_SCHOOLS_Elem %>% filter(Year != "2011") %>% pull(SchoolID) %>% length())/ALL_SCHOOLS_Elem %>% filter(Year != "2016" & Level == "Elementary") %>% pull(SchoolID) %>% length()) %>% scales::percent()` and `r  (( ((avail_2011 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2011)) |> select(OTGC) |> sum())) - (avail_2016 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2016)) |> select(OTGC) |> sum())) / (avail_2016 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2016)) |> select(OTGC) |> sum()) ) |> scales::percent()` decline in elementary school locations and capacity, respectively, across the city though overall student population and households in the city did not significantly change between 2011 and 2016 ( `r avail_2011$kid_popadj %>% sum(na.rm=T) %>% prettyNum(big.mark=",")` (2011) and `r avail_2016$kid_popadj %>% sum(na.rm=T) %>% prettyNum(big.mark=",")` (2016)).

From the perspective of public money no longer spent as a result of reduced school operational costs, who and in what way does the Hamilton community pay for these savings by means of accessibility and accessibility-related impacts? In this paper, we hypothesize that that this policy negatively impacted the trips-to-school for many students in addition to decreasing potential active-trip opportunities and increasing emissions. Spatial availability, a singly-constrained competitive accessibility measure [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024], is used to quantify the before- and after- accessibility and associated impacts for the city of Hamilton and the elementary school-aged student population as a result of a province-wide school closure and consolidation policy.

This paper contributes the following: how does the (1) spatial availability, (2) loss in walk minutes and (3) associated carbon emissions change for home-to-school trips after a city closes and consolidates schools? Further, this work examines what communities, based on low-income prevalence, experienced the greatest losses as a result of this policy. The methods introduced in this paper can be used by decision-makers to evaluate the spatial and transport-related impacts of public service provision policy.

The remainder of the paper is organized as follows. Section 2 details the data and methods, particularly multimodal spatial availability, a singly-constrained competitive accessibility measure that considers demand for school-seats by students and friction of distance, as well associated data sources for the inputs into spatial availability. Section 3 showcases the results of the impact that the school consolidation policy had on Hamilton's school spatial availability landscape. Section 4 discusses the results and estimates potential walk minutes lost, where trip-related GHG emissions may have increased, and equity implications for the policy. Section 5 concludes the paper with key remarks. Furthermore, to increase the diffusion of this analysis, all work is completed in a R environment and all associated code and data will be open and reproducible within the lead author's [GitHub repository](https://github.com/soukhova).

# Data and Methods

Our focus is to evaluate the impact that the school consolidations between 2011 and 2016 had on the multimodal school spatial availability and travel-related GHG emission landscape of the city of Hamilton.

To facilitate comparison, two scenarios are generated. The first and second reflect estimated as-is conditions in 2011 and 2016 assuming the respective school configurations, average student population and low-income prevalence of households, residential parcel locations, and estimated modal travel times in both years. The difference between these scenarios are quantified in terms of accessibility, emission and equity implications and discussed in context with the school consolidation policy.

Data is retrieved at different spatial granularity from the following six data sources.

## Schools

**Firstly**, schools and associated school catchments are retrieve from the school boards. In Hamilton, the majority of student-aged population attends a school in one of the two English public school boards. One board is referred to as *public* (i.e., Hamilton Wentworth District School Board (HWDSB)) in our work and the other is referred to as *public-catholic* (Hamilton Wentworth Catholic District School Board (HWCDSB)). These catchments indicate what residential property is assigned to what school, by default. Families can decide to attend schools out of catchment, in fact 21%-23% <!--calculated in 03-travel-times-TTS.qmd--> of motorized Hamilton-based Hamilton school trips are out of catchment according to the regional travel survey, the Transportation Tomorrow Survey (TTS) in 2011 and 2016 [@data_management_group_tts_2018]. The elementary school locations and catchments are provided by the HWDSB and HWCDSB for the 2010-2011 and 2015-2016 academic years and visualised in @fig-Fig1.

Formally, elementary schools are defined as schools that provide instruction to any combination of grades between kindergarten to grade 8 (i.e. typically children aged 5 to 14). As such, elementary includes middle schools that only instruct grades 6 to 8 (*Mid*), primary schools which only instruct kindergarten to grade 5 or 6 (*JrElem*), and all grade elementary schools (*Elem*) which instruct all grades from kindergarten to grade 8.  The maximum number of students that may be enrolled in a school is calculated by the provincial Ministry of Education; this value is referred to as the on-the-ground capacity (*OTGC*) and is assigned to each school. The current/historic OTGC province-wide has been made publicly unavailable [@ontarioSchoolFacilityInventory2017]. However, for schools that underwent the full Accommodation Review process, the OTGC for a given year can be scraped from the generated publicly available School Information Profiles. A regression model was derived to estimate the OTGC for schools with no OTGC based on 1) the school's footprint (*F*) in units of (`m^2`), 2) school's shortest Euclidean distance from the centroid of Hamilton's central business district area (*DistCBD*) defined in units of (m), and 3) coefficients associated with a binary variable indicating the type of school grade instruction (*Mid*, *JrElem*, *Elem*). The building footprint from an archived spatial data set [@DMTI_Spatial_2015] and the 2016 footprints were retrieved from OSM [@OpenStreetMap_2021]. The center of the CBD is defined as a significant intersection in Hamilton Downtown Business Improvement Area [@Hamilton_2014]. This point serves as a pseudo-proxy for age of school construction as the oldest buildings are located within the Hamilton Downtown boundary [@Merrall_2021] and the age of schools is unavailable.

Equation (\ref{eq:OTGC-public}) and (\ref{eq:OTGC-catholic}) describes the estimated OTGC for public schools and catholic schools respectively. School status and OTGC summed by catchment is also visualised in @fig-Fig1. See the Table \ref{Tab:TabA1} in the Appendix for the associated regression results.

```{=tex}
\begin{equation}
\label{eq:OTGC-public}
OTGC_{Public} = F^{0.346}-e^{0.00003*DistCBD}+e^{3.123*JrElem_b}+e^{3.752*Elem_b}+ e^{3.068*Mid_b}
\end{equation}
```
```{=tex}
\begin{equation}
\label{eq:OTGC-catholic}
OTGC_{Catholic} =F^{0.471}-e^{0.00003*DistCBD}+e^{2.333*Elem_b}
\end{equation}
```

```{r catchments-2011-2016, fig.cap="\\label{fig:Fig1}The school catchments, estimated on the ground capacities of schools summed per catchment, and school status for both public and public catholic school boards for year 2011 and 2016. OTGC is presented by quartiles. ", out.width="100%", echo=FALSE}

knitr::include_graphics("Figures/catchment_schools_plot.png")
```

## Students, residential locations and low-income prevelance
Secondly, student population and proportion of household low-income after tax (LIM-AT) prevalence is retrieved from the 2011 and 2016 Canadian Census [@Statistics_Canada_2011; @Statistics_Canada_2016]. The census releases population data by age group category and as pertinent to this paper, 5 to 9 year olds and 10 to 14 year olds. As more relevant to this paper, the LIM-AT prevelance for households with children under 18 (there is no LIM-AT prevelance tabulated for households with exclusively elementary aged children) is retrained. LIM-AT is refers to the proportion of private households that are below the median after-tax income [@governmentofcanadaDictionaryCensusPopulation2017]. These census variables are taken at the finest spatial unit of analysis disseminated by the Canadian census publicly available, the spatial dissemination area (DA). They are retrieved using the {cancensus} R package [@von_Bergmann_Shkolnik_Jacobs_2021]. DAs are designed by Statistics Canada with the aim of population uniformity hence DAs greatly vary in area but represent between approximately 400 and 700 (1st and 3rd quartile) in total population. The population aged 5 to 14 and the proportion of LIM-AT are visualised in the first and last row in @fig-Fig2.

Thirdly, centroids of residential parcels in 2011 and 2016 are used to represent residential locations [@Teranet_2009]. There are 134,340 and 139,467 residential locations in 2011 and 2016 respectively, representing `r cen_data_2011_hamiltononly$DA_households |> sum() |> prettyNum(big.mark = ",")` and `r cen_data_2016_hamiltononly$DA_households |> sum() |> prettyNum(big.mark = ",")` households (according to the Canadian Census [@Statistics_Canada_2011; @Statistics_Canada_2016]). As the number of children per residential location is unavailable publicly, the average number of 5 to 14 year olds for each DA is divided by the number of residential households in that DA. Then the potential number of children at each residential location is calculated by multiplying this DA rate. As the [@Teranet_2009] dataset is proprietary, the the middle row in @fig-Fig2 visualises an aggregation of the information: the average rate of 5-14 year olds per residential household at the DA level.

```{r elem-res-rate-2016, fig.cap="\\label{fig:Fig2}The magnitude (top row) and rate per residential unit (middle row) of elementary aged student population per DA in 2011 and 2016. The proportion of low-income households with children under 18 years old is also shown (bottom row). All scales are represented in quartiles.", out.width="100%", echo=FALSE}
knitr::include_graphics("Figures/stud_pop_rate_LIM_plot.png")
```

## Estimated travel times and emissions

The fourth data source is an estimated travel time matrices from all residential locations to all elementary schools for both motorized (car) and non-motorized (walking) travel. The travel times are estimated using the R-package {r5r} [@Pereira2021r5r], a package that interfaces with the java-based R5 routing engine developed separately by Conveyal [@conveyalConveyalR5Routing2022]. For both modes, the inputs assume a maximum travel time threshold of 60 minutes and the free-flow OpenStreetMap road network of Hamilton (retrieved using Geofabrick [@geofabrikOntarioOpenStreetMapGeofabrik2022] and manually edited to only include the Hamilton boundary). Further, walk travel times that are only less than 27 min minutes were retained. This travel time is approximately equivalent to 1.6 km which is the minimum distance that is required from a school within-catchment to qualify for a school bus [@HWDSB_2019]. <!--put in statistics here-->

It is assumed that children can go to any elementary school, but certain schools are more likely to be attended depending on their proximity to their residential location. Based on this assumption, the fifth data source is empirical data regarding the mode-used and origin-destination locations of home-to-school trips from the 2011 and 2016 TTS [@data_management_group_tts_2018]. The TTS is representative travel survey conducted in the Greater Golden Horseshoe Area in Ontario every 5 years. The trip-level travel data extract for this paper represent `r od_trips_perc_2011$TAZOrig_car_trips |> sum() |> prettyNum(big.mark = ",")` and `r od_trips_perc_2016$TAZOrig_car_trips |> sum() |> prettyNum(big.mark = ",")` motorized trips (mode used includes private care passenger, school bus, taxi, and transit) and `r od_trips_perc_2011$TAZOrig_walk_trips |> sum() |> prettyNum(big.mark = ",")` and `r od_trips_perc_2016$TAZOrig_walk_trips |> sum() |> prettyNum(big.mark = ",")` non-motorized trips (mode used includes walk and cycling) for 5 to 13 year olds from home to school in 2011 and 2016 respectively. Data is available at the spatial resolution of Traffic Analysis Zones (TAZ). TAZ are spatial units created for the purpose of the TTS; a few DAs typically nest within each TAZ. {r5r} is used to estimated motorized and non-motorized travel from all TAZ centroids to all TAZ centroids and these estimated travel times are tied to the OD trips. The travel times and the intensity of modal home-to-school flows are visualised in @fig-Fig3 along with the boundaries of the TAZ and DAs. As can be seen in @fig-Fig3, not all TAZ capture an elementary school trip by both modes. For this reason, the proportion of modal-split is aggregated at the community level. Hamilton (Ancaster `r community_car_perc_2011$TAZOrig_car_perc[1] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[1] |> percent()`, Dundas `r community_car_perc_2011$TAZOrig_car_perc[2] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[2] |> percent()`, Flamborough `r community_car_perc_2011$TAZOrig_car_perc[3] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[3] |> percent()`, Glanbrook `r community_car_perc_2011$TAZOrig_car_perc[4] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[4] |> percent()`, Hamilton Central `r community_car_perc_2011$TAZOrig_car_perc[5] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[5] |> percent()`, and Stoney Creek `r community_car_perc_2011$TAZOrig_car_perc[6] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[6] |> percent()`).

```{r imped-20112016, fig.cap="\\label{fig:Fig3}The origin-destination flows from home-based motorized and non-motorized school trips for students between 5-13 years old as retrieved from TTS 2011 and 2016 mapped atop proportion of TAZ non/motorized modal share, DA boundaries, and community boundaries. ", out.width="100%", echo=FALSE}
knitr::include_graphics("Figures/OD_flow_plot.png")
```
Furthermore, as a sixth data source, an estimated travel time matrix for walk and car travel modes is calculated using {r5r} and the geographical centroids of the TTS origin-destination flows.  For both modes, the inputs assume a maximum travel time threshold of 60 minutes and the free-flow OpenStreetMap road network of Hamilton [@geofabrikOntarioOpenStreetMapGeofabrik2022]. The walk travel times are filtered to only include travel times under 27 minutes. These travel time matrices along with the origin-destination flows are used to create modal trip length distribution (TLD) functions. A TLD is a useful technique to calibrate impedance functions as they are a representation of the likelihood that a proportion of trips are taken at a specific travel cost [@horbachov_theoretical_2018; @batista_estimation_2019]. The empirical TLD is fit to a density distribution using maximum likelihood and moment matching techniques and the Nelder-Mead and Brent methods for direct optimization available within the {fitdistrplus} R package [@fitdistrplus_2015]. Based on goodness-of-fit criteria and diagnostics (included in Appendix), the gamma and exponential distribution is selected for the motorized and non-motorized modal distributions respectively. The gamma distribution is defined by the shape ($\alpha$) parameter of `r round(gamma2011_motor$estimate[1], 3)` (2011) and `r round(gamma2016_motor$estimate[1], 3)` (2016) and the rate ($\beta$) of `r round(gamma2011_motor$estimate[2], 3)` (2011) and `r round(gamma2016_motor$estimate[2], 3)` (2016). The exponential distribution is defined by the rate ($\beta$) parameter of `r round(exp2011_nonmotor$estimate[1], 3)` (2011) and `r round(exp2016_nonmotor$estimate[1], 3)` (2016). The TLDs for the motorized and non-motorized trips are shown in @fig-Fig4.

```{r imped-20112016-2, fig.cap="\\label{fig:Fig4}The empirical and theoretical motorized and non-motorized impedance functions. ", out.width="100%", echo=FALSE}
knitr::include_graphics("Figures/TLD_empirical_wtheo_curve_plot.png")
```

The sixth and final data source relates to motorized GHG well-to-wheel estimations. It assumed that all non-motorized trips produce no emissions and those that are motorized take the average emission factor of 0.1 kg CO2e per minute travel. This emission factor is assumed base on an average speed of 40 km/h, the 2,380 g $CO_{2}e$ produced when 1 L of gasoline is combusted and an internal combustion engine passenger vehicle fuel efficiency of 7 g/100km as discussed in @soukhovOccupancyGHGEmissions2022.

## Putting it all together: multimodal spatial availability

Accessibility is defined as "access to opportunities" or the "potential for spatial interaction". It is classically presented as the gravity-based measure popularized by @hansenHowAccessibilityShapes1959 and takes the following general multi-modal formulation (Equation (\ref{eq:multimodal-conventional-accessibility})):
```{=tex}
\begin{equation}
\label{eq:multi-modal-conventional-accessibility}
S_i^m = \sum_{j=1}^JO_j \cdot f^m(c_{ij}^m)
\end{equation}
```
\noindent where:

-   $m$ is a set of modes.
-   $c_{ij}^m$ is a measure of the cost of moving between $i$ and $j$ for each $m$.
-   $f^m(\cdot)$ is an impedance function of $c_{ij}^m$ for each $m$; it can take the form of any monotonically decreasing function chosen based on positive or normative criteria [@paez2012measuring].
-   $i$ is a set of origin locations ($i = 1,\cdots,N$).
-   $j$ is a set of destination locations ($j = 1,\cdots,J$).
-   $O_j$ is the number of opportunities at location $j$.
-   $S$ is Hansen-type accessibility as weighted sum of opportunities.

As indicators of urban structure, Hansen-type measures are informative, but the meaning of 10,000 accessible school-seats is harder to pin down: how many opportunities must any single student have access to? Furthermore, in scenario comparisons, the interpretation of the percentage change in accessible school-seats is unclear. The interpretability of Hansen-type accessibility has been discussed in numerous studies, including recently by @hu_2019_measuring, @kelobonye2020measuring, and in greater depth by @merlin2017competition along with @soukhovIntroducingSpatialAvailability2023 and @soukhovMultimodalSpatialAvailability2024. The interpretation of accessibility is suggested to depend on how many people demand the opportunity, especially for exclusive opportunity-types like schools-seats (i.e., one school-seat is for one student).

In this paper, we benefit from new development in accessibility research, particularly the multimodal spatial availability measure [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024]. Spatial availability is a singly-constrained accessibility measure that accounts for competition by students using different modes for exclusive opportunities, such as school seats. The measure’s single constraint ensures that the marginals at the destination are met and thus the number of estimated school seats (opportunities) are preserved and allocated proportionally to the mode-using student population. This proportional allocation of opportunities yields an interpretable and meaningful measure of opportunity access, particularly when comparing across modes, at the spatial resolution of a residential parcel, and multiple time periods. See @soukhovMultimodalSpatialAvailability2024 for further discussion, multimodal spatial availability  $V_{i}^m$ is defined as given by Equation (\ref{eq:spatial-availability}).

```{=tex}
\begin{equation}
\label{eq:spatial-availability}
V_{i}^{m} = \sum_{j=1}^J O_j\ F^{tm}_{ij}
\end{equation}
```
\noindent where:

-   $F^{tm}_{ij}$ is a balancing factor that depends on the population and cost of movement in the system as part of the gravity modelling framework and is captured in Equation (\ref{eq:balancing-factors}) for mode $m$.
-   $O_j$ is the number of opportunities at $j$.
-   $V_i^m$ is the number of spatially available opportunities from the perspective of $i$ for mode $m$.

$F^{tm}_{ij}$ can be understood as the joint probability of allocating opportunities, where $F^{pm}_{i}$ is the population-based balancing factor that grants a larger share of opportunities to larger $m$ population spatial units and $F^{cm}_{ij}$ is the impedance-based balancing factor that grants a larger share of the opportunities to less $m$-travel costly centers. Together $F^{tm}_{ij}$ ensures proportional allocation such that opportunities $O$ (like school-seats) are preserved for the whole region (i.e., $O = \sum_{j} O_j = \sum_{i} V_i = \sum_{m}\sum_{i} V_i^m$ ) and is reflected in Equation (\ref{eq:balancing-factors}):

```{=tex}
\begin{equation}
\label{eq:balancing-factors}
F^{tm}_{ij} = \frac{F^{pm}_{i} \cdot F^{cm}_{ij}}{\sum_{i} F^{pm}_{i} \cdot F^{cm}_{ij}}
\end{equation}
```
\noindent where:

- The factor for allocation by population for each $m$ at each $i$ is $F^{pm}_{i} = \frac{P_{i}^m}{\sum_{m}\sum_{i} P_{i}^m}$
- The factor for allocation by travel cost for each $m$ at each $i$ and $j$ is $F^{cm}_{ij} = \frac{f^m(c^m_{ij})}{\sum_{m}\sum_{i} f^m(c^m_{ij})}$

It should be noted, that the population-based allocation factors $F^{pm}_{i}$ when summed over all spatial units in the region always equals to 1 (i.e., $\sum_{m}\sum_{i} F^{pm}_{i}= 1$), likewise for impedance-based allocation factors $F^{cm}_{i}$ (i.e., $\sum_{m}\sum_{i} F^{cm}_{ij} = 1$).

Hansen-type accessibility is not designed to preserve the number of opportunities in the region, it simply counts the potential opportunities that those in a zone can interact with (weighted by the friction of distance). Also as discussed in @soukhovIntroducingSpatialAvailability2023, popular competitive accessibility measure such as the 2SFCA [@shenLocationCharacteristicsInnercity1998;@luoMeasuresSpatialAccessibility2003] ends up preserving the number of opportunities in the analysis region but the definitions of variables are internally obscured; the only way it preserves the number of opportunities is if the effect of the impedance function is ignored when expanding the values of opportunities per capita to obtain the total number of opportunities. On the other hand, the proportional allocation procedure associated with calculating multi-modal spatial availability $V_i^m$ consistently returns a number of opportunities available to all mode-using populations that matches the total number of opportunities in the region when summed. As a result of this consistency, it is possible to define a measure of multimodal spatial availability per capita ($P$) as presented in Equation (\ref{eq:SA-per-capita}) for use as a benchmark to compare against the regional opportunities per capita ($\frac{\sum_{j} O_j}{\sum_{i} P_i}$).

```{=tex}
\begin{equation}
\label{eq:SA-per-capita}
v_i^m = \frac{V_i^m}{P_i^m}
\end{equation}
```

For the reasons outlined, this paper uses the multimodal spatial availability $V_i^m$ and spatial availability per student $v_i^m$ measures to quantify accessibility changes.

- First, all residential locations are associated to their respective census DA, TTS TAZ and community boundary (@fig-Fig3) based on spatial location. Each residential location is assigned a number of 'potential' motorized and non-motorized student aged population from these different data sources to calculate a motorized and non-motorized population balancing factor $F_{i}^{pm}$. For each residential location the following is retained: the 2011 and 2016 census average elementary-aged student population per residential location (@fig-Fig2 second row) and the average motorized and non-motorized split as informed by the TTS aggregated by community (@fig-Fig3).
- Second, the school capacity $O_j$ is estimated and associated with each school (@fig-Fig1). All residential locations are assumed to be able to access all schools by either motorized or non-motorized mode. All origins (residential locations) can reach all schools by motorized mode, but few residential locations can reach schools within a 27 minute non-motorized trip. 
- Third, the motorized and non-motorized impedance-balancing factors $F_{ij}^{cm}$ are calculated using the modal theoretical TLDs (@fig-Fig4) based on the respective travel times (data source four) for each origin (residential location) to school combination. 
- Finally, outputs from all three stages are joined together. $F_{i}^{pm}$ joined with $F_{ij}^{cm}$ yields the total balancing factor $F^{tm}_{ij}$ which serves to proportionally allocates the school capacity $O_j$ to each residential location. The value is multimodal spatial availability $V_i^m$ which can be interpreted as the number of school-seats that are available to the mode-using population at that residential location. $V_i^m$ is then summed and represented at the DA and Community level along with the calculated $v_i^m$ for interpretation.

# Results

```{r Vim-vim-4-plots, fig.cap="\\label{fig:Fig6}Multimodal spatial availability (greys) and spatial availability per mode-using student (diverging reds and greens) aggregated at the level of DA for 2011 and 2016.", out.width="100%", echo=FALSE}
knitr::include_graphics("Figures/Vim_vvim_20112016_plot.png")
```

The motorized and non-motorized spatial availability $V_i$ for both 2011 and 2016 is represented in the top half of @fig-Fig6. Though at different magnitudes, both populations using non- and motorized modes have higher spatial availability when more proximate to schools. Hence, all DAs have higher values within Hamilton Central and those more proximate to schools in less-urban communities. Trends appear similar for both 2011 and 2016: proximate populations using motorized modes benefit from a wider-range in high spatial availability than non-motorized modes. When $V_im$ is divided by the number of mode-using students in that DA, $v_i^m$ is derived (bottom half of @fig-Fig6). A similar spatial distribution seen in $V_i^m$ can be seen in $v_i^m$ plots, but as a rate of student population. If our aim is to provide sufficient availability to schools for both motorized and non-motorized populations (i.e., greater than 1.0 $v_i^m$ school-seat availability per student), this rate is only achieved in the core of the city for motorized modes, and notably only for non-motorized populations in DAs that are in less-densely populated rural areas that are school proximate and certain pockets of Hamilton Central. The majority of schools closed were within Hamilton Central, so non-motorized populations in DAs within Hamilton Central were the most negatively, and certain motorized populations captured these losses as gains in availability.

```{r change-Vim-vim-plot, fig.cap="\\label{fig:Fig7}The change in multimodal spatial availability and spatial availability per mode-using student between 2016 and 2011 at the level of DA of the 2016 Canadian Census.", out.width="100%", echo=FALSE}
knitr::include_graphics("Figures/Vim_vvim_20112016_CHANGE_plot.png")
```

To emphasize how spatial availability changed between 2016 and 2011, the change in $V_i^m$ and $v_i^m$ is visualised for both modes in @fig-Fig7. Overall, a decrease in both $V_i^m$ and $v_i^m$ can be seen both modes modes around closed schools (black points). An increase near new or expanded schools (green points) is not as noteable as overall fewer school seats were added relative to population increase. The non-motorized mode-using populations are most impacted by the school closures/consolidations along with other competition factors such as increased children population in less urban areas and an increase in motorized mode use. For non-motorzied mode-using populations, $V_i^m$ decreases across the city. However, some motorized-mode using populations in DAs do experience an increase in $V_i^m$: these DAs are often relatively proximate to schools but more immediately outside walking-range. Under a competitive and constraint calculation, these DAs benefit from the losses in spatial availability that are predominately experienced by the non-motorized populations within Hamilton Central.

<!--Maybe worth adding a plot of GHG emissions or change in tt?-->

# Discussion

<!--This will be a table with summaries of LIMAT, Kid pop, V_im change, v_im change, total travel time change, and emission changed by community and mode? We shoudl discuss to decide what's the best presentation.-->
```{r}
# SAvail_im_COMM_changes |> filter(mode == "motorized") |> st_drop_geometry() |> 
#   dplyr::select(c("COMMUNITY",
#                   "prev_LIMAT_2016", 
#                   "c_kid_perc", 
#                   "c_V_im_perc", 
#                   "c_v_im_perc", 
#                   "emissions_2011",
#                   "emissions_2016"
#                   #"c_emissions_perc",
#                   # "emissions_perkid_2011",
#                   # "emissions_perkid_2016",
#                   #"c_emissions_perkid_perc" 
#                   )) |> arrange(desc(prev_LIMAT_2016)) 
```

```{r}
SAvail_im_COMM_changes |> st_drop_geometry() |> 
  dplyr::select(c("COMMUNITY",
                  "prev_LIMAT_2016", #prevalance of LIM after-tax for households with children under 18
                  "mode", 
                  "kid_proportion_2016", #the proportion of children 5-14 age, sums to 100%
                  "c_kid_proportion_perc", #the % change in proportion between 2016-2011 i.e. (2016-2011)/2011, negative means decrease
                  "V_im_proportion_2016",  #the proportion of V_im, sums to 100%
                  "c_V_im_proportion_perc", #the % change in proportion between 2016-2011 i.e. (2016-2011)/2011, negative means decrease
                  "v_im_2016", #the proportion of V_im (in community) /kid_population (in community) = v_im, sums to 100%
                  "c_v_im_perc", #the % change in proportion between 2016-2011 i.e. (2016-2011)/2011, negative means decrease
                  "total_tt_2016", #summed travel time (kid_pop at each household * travel time of mode, summed for the community). Values sums to 100%
                  "c_total_tt_perc",  #the % change in proportion between 2016-2011 i.e. (2016-2011)/2011, negative means decrease
                  "emissions_perkid_2016", #summed motorized travel time * 0.1 g/min (kid_pop at each household * travel time of mode, summed for the community). Values sums to 100%
                  "c_emissions_perkid_perc"  #the % change in proportion between 2016-2011 i.e. (2016-2011)/2011, negative means decrease
                  ))
```

Spatial availability can also be summarized by community. The majority of schools were closed in Hamilton Center, this community represents `r SAvail_im_COMM_changes |> st_drop_geometry() |> filter(COMMUNITY == "HA") |> dplyr::select("kid_proportion_2016") |> sum() |> scales::percent()` of the 5-14 year old population in 2016. Though the population between 2011 and 2016 decreased, the rate of school-seat spatial availability decreased disproportionately more (population by `r SAvail_im_COMM_changes |> st_drop_geometry() |> filter(COMMUNITY == "HA") |> transmute(calc = ((sum(kid_2016)-sum(kid_2011))/sum(kid_2011)) |> scales::percent()) |> first()` and spatial availability by `r `r SAvail_im_COMM_changes |> st_drop_geometry() |> filter(COMMUNITY == "HA") |> transmute(calc = ((sum(V_im_2016)-sum(V_im_2011))/sum(V_im_2011)) |> scales::percent())  |> first()`). This out of proportion decrease is greatest in Hamilton Central than any other community. This may have grave equity concerns as Hamilton Central has the highest LIM-AT prevalence, with `r SAvail_im_COMM_changes |> st_drop_geometry() |> filter(COMMUNITY == "HA" & mode == "motorized") |> dplyr::select("prev_LIMAT_2016") |> round() |> paste0("%")` of households LIM-AT in 2016, 3.7 times greater than Glanbrook with the lowest level of LIM-AT prevalence. Between 2016 and 2011, Hamilton Center had the lowest levels of emissions per student and saw the highest per student increase in 2016. Hamilton Center had the highest non-motorized trip potential, and it was reduced in part by the school closure policies.

Conversely, communities with the lowest LIM-AT benefit the most, notably in gains to motorized spatial availability. Glanbrook, Flamborough and Ancaster have the lowest prevalence of LIM-AT of any Hamilton Community and together account for `r SAvail_im_COMM_changes |> st_drop_geometry() |> filter(COMMUNITY == "FL" | COMMUNITY == "GL" | COMMUNITY == "AN") |> dplyr::select("kid_proportion_2016") |> sum() |> scales::percent()` of the 5-14 year old population in 2016. These communities experienced a growth in student population (or only a small decrease) and proportionally benefited from an increase in motorized spatial availability.

Across the city, those who commute by non-motorized mode suffer from the most dramatic drop in student population and especially spatial availability across the city. In all communities, spatial availability decreases at greater levels than population. Notably, in Dundas (2nd highest LIM-AT), the only school in the community closed leaving the population in the community with no other option but to use motorized modes. Glanbrook (lowest LIM-AT), which experienced a sharp increase in student population along with non-motorized using population, suffered a lose in non-motorized spatial availability despite seeing the highest increase in motorized spatial availability. Overall, the gain in motorized spatial availability resulted in a further erosion of non-motorized spatial availability. This erosion results in a lose in active transport trips and an increase in emissions city-wide (`r SAvail_im_COMM_changes |> st_drop_geometry() |> transmute(em = (sum(emissions_2011))/(sum(kid_2011)) ) |> sum() |> round(digit=2)`kg CO2e per student in 2011 to `r SAvail_im_COMM_changes |> st_drop_geometry() |> transmute(em = (sum(emissions_2016))/(sum(kid_2016)) ) |> sum() |> round(digit=2)` in 2016).

<!-- develop further?-->
Study limitations:

- potential students per households, potential trips to all schools
- estimated travel times, direct car or walk trip (shortest path). Not necessarily car, could be transit or school bus. Walking shortest-paths are also optimistic for children's travel: could be unsafe and not taken in reality and/or alternative more pleasant paths are taken.
- estimated emissions. GHG emissions per minute of driving varies extensively depending on operating conditions and vehicle characteristics [@soukhovOccupancyGHGEmissions2022].

<!-- Add an additional paper: discuss the potential impact of mode-switching on GHG emissions reductions [@pantelakiImpactHomeschoolCommuting2024].

Also, access to schools has broader implications; it has a relationship to residential and job choices [@pietrabissaSchoolAccessCity2023] and housing prices [@merrallWhatSchoolWorth2023]-->

# Concluding remarks

In this paper, we estimate the change in multimodal spatial availability as a result of school consolidations/closures between the years of 2011 and 2016 in an empirical case study of Hamilton, a mid-size city (\~550,000 pop) in Ontario, Canada. The spatial availability and spatial availability per student are visualized on a DA level; spatial availability for the city declined overall. This decline is then discussed by measuring potential walk minutes and GHG emissions (produced by additional car minutes that result since walk minutes are removed). Overall, the closure of `r ((ALL_SCHOOLS_Elem %>% filter(Year != "2016" & Level == "Elementary") %>% pull(SchoolID) %>% length() - ALL_SCHOOLS_Elem %>% filter(Year != "2011" & Level == "Elementary") %>% pull(SchoolID) %>% length())/ALL_SCHOOLS_Elem %>% filter(Year != "2016" & Level == "Elementary") %>% pull(SchoolID) %>% length()) %>% scales::percent()` of elementary schools in Hamilton resulted in: a `r (ALL_SCHOOLS_Elem %>% st_drop_geometry() %>% filter( (Year == "2011" | Year == "2011 and 2016") & Level == "Elementary") %>% dplyr::select(OTGC2011) %>% sum() - ALL_SCHOOLS_Elem %>% st_drop_geometry() %>% filter( (Year == "2016" | Year == "2011 and 2016") & Level == "Elementary") %>% dplyr::select(OTGC2016) %>% sum()) %>% prettyNum(big.mark=",", digits=1)` decline of school-seat capacity, a decline of `r SAvail_im_COMM_changes |> st_drop_geometry() |> filter(mode == "non-motorized") |> transmute(total_tt = total_tt_2016-total_tt_2011) |> sum() %>% abs() %>% prettyNum(big.mark=",", digits=1)` *potential* home-to-school walk-minutes (per one-way trips), an increase of `r SAvail_im_COMM_changes |> st_drop_geometry() |> filter(mode == "motorized") |> transmute(emissions = emissions_2016-emissions_2011) |> sum() %>% abs() %>% prettyNum(big.mark=",", digits=1)` kg CO2e as a result of increased car-minutes as a result of lost *potential* home-to-school walk-minutes (per one-way trips).

By quantifying the spatial availability, environmental, and active-travel implications of school closure/consolidation policies in Hamilton, we offer a methodology for spatial policy analysis scenario. This methodology can be used by researchers and decision-makers to plan and evaluate equitable and sustainable urban planning policies from a spatial perspective. We urge policy makers to view 'spatial availability' of public resources from a per-capita perspective, plan capacity and location of service provision with the cost of travel (and associated implications such as GHG emissions, active transportation mode, safety of active travel modes, etc.) that sufficiently serves the target population. In this paper we suggest and assume a benchmark of 1.0 school-seats per student which accounts for sufficy. Spatial availability can be used to obtain this benchmark since the total number of opportunities (i.e., school capacity) is preserved in the region of analysis unlike unconstrained forms of accessibility measurement e.g., Hansen-type measure [@hansenHowAccessibilityShapes1959]. As such, the assigned spatial availability can be divided by population at each zone to yield an interpretable per capita measurement.

Like many other localities, Hamilton is not immune to top-down financial efficient policy decisions to determine community infrastructure is underutilized and it be closed. From the perspective of the provincial government of Ontario, the school closures which occurred in Hamilton resulted in operational savings-per-student but we demonstrate how this resulted in both families and students paying a price through spatial availability, lost potential walk minutes, and increased GHG emissions. In the case of Hamilton Central, where households have the highest prevalence of LIM-AT in the city, those households suffer the most disproportionate decrease in motorized spatial availability and increase in percentage increase in emissions per capacity. As public schools should serve all students; the determination of spatial availability and spatial equity of impact, is critical to better inform decision-makers when evaluating school consolidation policies.

Ultimately, this case study furnishes evidence that consolidation of schools was particularly ill-timed: with the advent of the COVID-19 pandemic, the education system lacked the resiliency to accommodate reduced classroom capacities and left parents who once lived within active transportation distance to schools relying on motorized transportation for their children. These top-down austerity measures left the Hamilton local school system more vulnerable to the impacts of COVID-19 by undervaluing the societal role of schools in neighbourhoods. By identifying the impact of school consolidation policies in Ontario we anticipate that researchers and decision-makers will have better information to center the wellbeing of all residents, including students, and the planet in urban planning policies.

# References

### 2024-04-09 data-prep/00-Estimating-Impe2school-from-TTS.Rmd
Draft of 'Manuscript.qmd' for special issue

## Preliminaries

Load packages:
```{r load-packages}
library(readxl)
library(sf)
library(tidyverse)
library(units)
library(tmap)
library(ggplot2)
```

## Traffic Analysis Zones

Read the boundaries of the traffic analysis zones (retrieved from the TTS 2016):
```{r}
ggh_taz <- st_read("data-inputs/TTS/tts06_83_region.shp")
```

Project object:
```{r}
ggh_taz <- ggh_taz %>%
  st_transform(crs = 32617)
```

Prepare traffic analysis zones:
```{r}
ggh_taz <- ggh_taz %>%
  transmute(GTA06 = as.character(GTA06),
            AREA = st_area(geometry) %>% 
              set_units(km^2) %>% 
              drop_units())
```

Extract the IDs for all TAZ:
```{r}
ghta_taz_id <- ggh_taz$GTA06
```

## Planning boundaries

Read the boundaries of the traffic analysis zones (retrieved from the TTS 2016):
```{r}
ggh_pd <- st_read("data-inputs/TTS/tts06_pd_83.shp")
```

Project object:
```{r}
ggh_pd <- ggh_pd %>%
  st_transform(crs = 32617)
```

Adding region names to the object as they are not included in the origin data:
(http://dmg.utoronto.ca/pdf/tts/2016/2016TTS_Conduct.pdf):
```{r}
ggh_pd <- ggh_pd %>%
  mutate(REGION_name = 
           ifelse(REGION == 1, "Toronto", 
                  ifelse(REGION == 2, "Durham", 
                         ifelse(REGION == 3, "York", 
                                ifelse(REGION == 4, "Peel", 
                                       ifelse(REGION == 5, "Halton", 
                                              ifelse(REGION == 6, "Hamilton", 
                                                     ifelse(REGION == 11, "Niagara", 
                                                            ifelse(REGION == 12, "Waterloo", 
                                                                   ifelse(REGION == 13, "Guelph", 
                                                                          ifelse(REGION == 14, "Wellington", 
                                                                                 ifelse(REGION == 15, "Orangeville", 
                                                                                        ifelse(REGION == 16, "Barrie", 
                                                                                               ifelse(REGION == 17, "Simcoe", 
                                                                                                      ifelse(REGION == 18, "Kawartha Lakes", 
                                                                                                             ifelse(REGION == 19, "Peterborough City", 
                                                                                                                    ifelse(REGION == 20, "Peterborough County", 
                                                                                                                           ifelse(REGION == 21, "Orillia", 
                                                                                                                                  ifelse(REGION == 22, "Dufferin", 
                                                                                                                                         ifelse(REGION == 23, "Brantford", 
                                                                                                                                                ifelse(REGION == 24, "Brant", "NOOOO")))))))))))))))))))))
```

add 'REGION' from the planning distract object to the `ggh_taz`. 
```{r}
#buffering to clean up the edges of these objects -> there wre issues using st_join. 
ggh_taz <- st_buffer(ggh_taz, dist=0)
ggh_pd <- st_buffer(ggh_pd, dist=0)
```

```{r}
ggh_taz <- st_join(ggh_taz, ggh_pd %>% transmute(REGION),left = TRUE, largest = TRUE)
ham_taz <- ggh_taz %>% dplyr::filter(REGION==6) #Filter out everything other than Planning District 6 = Hamilton
```

## Origin destination matrices for place of residence and school destination (by person)
Read exported TTS 2011 and 2016 tables. The query was a cross-tabulation of <!--"persons"--> "trips" by 2006 zone of residence and 2006 zone of school using motorized modes ("mode_prime In B, D, G, J, M, O, P, S, T, U, 9" -- this is only one of four options, school bus, auto care passenger, taxi passenger, or transit) or non-motorized ("mode_prime In C, W" -- cycling or walking). The data was filtered to only include Hamilton residential households, "Trip purpose of destination" as school, the student status as full/part-time student, and age In 5-14:

```{r}
od_trips_2011_motor <- read_delim(file = "data-inputs/TTS/Schooltrip_trips_2011_motorized.txt", 
                      delim = "\t", 
                      col_names = FALSE)
od_trips_2016_motor <- read_delim(file = "data-inputs/TTS/Schooltrip_trips_2016_motorized.txt", 
                      delim = "\t", 
                      col_names = FALSE)

od_trips_2011_nonmotor <- read_delim(file = "data-inputs/TTS/Schooltrip_trips_2011_nonmotorized.txt", 
                      delim = "\t", 
                      col_names = FALSE)

od_trips_2016_nonmotor <- read_delim(file = "data-inputs/TTS/Schooltrip_trips_2016_nonmotorized.txt", 
                      delim = "\t", 
                      col_names = FALSE)
# od_trips_2011 <- read_delim(file = "data-inputs/TTS/Schooltrip_persons_2011.txt", 
#                       delim = "\t", 
#                       col_names = FALSE)
# od_trips_2016 <- read_delim(file = "data-inputs/TTS/Schooltrip_persons_2016.txt", 
#                       delim = "\t", 
#                       col_names = FALSE)

```

Find the position in the table where the information on work trips begins:
```{r}
# idx <- which(od_trips_2011$X1 == "TABLE    : stu_stat (Full time)")
idx <- which(od_trips_2011_motor$X1 == "COLUMN : gta06_sch")
```

Slice the table to obtain the work trips (remove heading information):
```{r}
od_trips_2011_motor <- od_trips_2011_motor %>% 
  slice((idx+2):n())
```

Separate the zone identifiers and the trips, and convert to numeric:
```{r}
od_trips_2011_motor <- od_trips_2011_motor %>%
  separate(X1, into = c("Zones", "Trips"), sep = " (?=[^ ]+$)") %>%
  mutate(Zones = str_trim(Zones, side = "both")) %>%
  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") %>%
  mutate(Origin = str_trim(Origin),
         Destination = str_trim(Destination),
         Trips = as.numeric(Trips))
```

```{r}
idx <- which(od_trips_2011_nonmotor$X1 == "COLUMN : gta06_sch")

od_trips_2011_nonmotor <- od_trips_2011_nonmotor %>% 
  slice((idx+2):n())

od_trips_2011_nonmotor <- od_trips_2011_nonmotor %>%
  separate(X1, into = c("Zones", "Trips"), sep = " (?=[^ ]+$)") %>%
  mutate(Zones = str_trim(Zones, side = "both")) %>%
  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") %>%
  mutate(Origin = str_trim(Origin),
         Destination = str_trim(Destination),
         Trips = as.numeric(Trips))
```

Repeat the same steps for 2016:
```{r}
idx <- which(od_trips_2016_motor$X1 == "COLUMN : gta06_sch")

od_trips_2016_motor <- od_trips_2016_motor %>% 
  slice((idx+2):n())

od_trips_2016_motor <- od_trips_2016_motor %>%
  separate(X1, into = c("Zones", "Trips"), sep = " (?=[^ ]+$)") %>%
  mutate(Zones = str_trim(Zones, side = "both")) %>%
  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") %>%
  mutate(Origin = str_trim(Origin),
         Destination = str_trim(Destination),
         Trips = as.numeric(Trips))

idx <- which(od_trips_2016_nonmotor$X1 == "COLUMN : gta06_sch")

od_trips_2016_nonmotor <- od_trips_2016_nonmotor %>% 
  slice((idx+2):n())

od_trips_2016_nonmotor <- od_trips_2016_nonmotor %>%
  separate(X1, into = c("Zones", "Trips"), sep = " (?=[^ ]+$)") %>%
  mutate(Zones = str_trim(Zones, side = "both")) %>%
  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") %>%
  mutate(Origin = str_trim(Origin),
         Destination = str_trim(Destination),
         Trips = as.numeric(Trips))
```

The resulting objects `od_trips_2011`.. and  `od_trips_2016`.. has the number of trips made by students aged 5-14 who are full-time/part-time students by place of residence (Origin) and place of school (Destination) using the GTA06 zoning system.

Now let's see how many students travel outside of their catchment. 
```{r}
#load catchments
load(file="data-products/ALL_CATCHS.RData")
CATCH_2011_Public <- ALL_CATCHS%>%filter(Year == 2011 & Level == "Elementary" & System == "Public") %>% st_set_crs(st_crs(ham_taz))
CATCH_2011_Cath <- ALL_CATCHS%>%filter(Year == 2011 & Level == "Elementary" & System == "Catholic")%>% st_set_crs(st_crs(ham_taz))
CATCH_2016_Public <- ALL_CATCHS%>%filter(Year == 2016 & Level == "Elementary" & System == "Public")%>% st_set_crs(st_crs(ham_taz))
CATCH_2016_Cath <- ALL_CATCHS%>%filter(Year == 2016 & Level == "Elementary" & System == "Catholic")%>% st_set_crs(st_crs(ham_taz))

#assign a catchment id (from elem, 2011) to od_trips_2011 taz ids, what I'm doing here is creating a column of lists -- the lists contain all the GTA06 within a single catchment. I do this for all 4 catchments (2011, 2016 and pub, cath). GTA06s that are within  multiple catchments are repeated i.e., a GTA06 that is in 3 catchments spatially is withinin 3 lists. 
ham_taz_catch_2011_public <-st_join(ham_taz,CATCH_2011_Public, join=st_intersects) %>% select(c("GTA06", "CID")) %>% st_drop_geometry() %>% group_by(CID) %>% summarize(GTA06_in_catch = list(GTA06))
ham_taz_catch_2011_cath <-st_join(ham_taz,CATCH_2011_Cath, join=st_intersects)%>% select(c("GTA06", "CID"))%>% st_drop_geometry()%>% group_by(CID) %>% summarize(GTA06_in_catch = list(GTA06))
ham_taz_catch_2016_public <-st_join(ham_taz,CATCH_2016_Public, join=st_intersects)%>% select(c("GTA06", "CID"))%>% st_drop_geometry()%>% group_by(CID) %>% summarize(GTA06_in_catch = list(GTA06))
ham_taz_catch_2016_cath <-st_join(ham_taz,CATCH_2016_Cath, join=st_intersects)%>% select(c("GTA06", "CID"))%>% st_drop_geometry()%>% group_by(CID) %>% summarize(GTA06_in_catch = list(GTA06))
```

```{r}
## THIS FUNCTION WAS CREATED USING CHATGDPT on DEC 6 2023

# Function to check if values in a row of df1 are within any list in df2
check_within_lists <- function(df1, df2) {
  # Initialize an empty list to store the results
  result_list <- list()

  # Check if values in a row of df1 are within any list in df2
  for (i in 1:nrow(df1)) {
    matching_ids <- which(sapply(df2$check_list, function(lst) all(df1[i, ] %in% unlist(lst))))
    if (length(matching_ids) > 0) {
      result_list[[i]] <- data.frame(result = TRUE, df1[i, ], CID = matching_ids)
    } else {
      result_list[[i]] <- data.frame(result = FALSE, df1[i, ], CID = NA)
    }
  }

  # Combine the results into a single dataframe
  result_df <- do.call(rbind, result_list)

  return(result_df)
}

# # Example usage is commented out:
# # Assuming you have two data frames named 'df1' and 'df2'
# # df1 has columns 'column1' and 'column2' with numeric values
# # df2 has columns 'id' and 'check_list' with lists of numeric values created using I() function
# df1 <- data.frame(column1 = c(1, 2, 3, 4, 5), column2 = c(10, 20, 30, 40, 50))
# df2 <- data.frame(id = c(1, 2, 3), check_list = I(list(c(1, 10), c(3, 30), c(1, 6, 10, 4, 40))))
# 
# # Call the function
# check_within_lists(df1, df2)
```
```{r, warning=FALSE}
## FOR 2011
#the origin-destination trips made within hamilton to hamilton schools (all in TTS Region `6`).
ODs <- data.frame(Origin = od_trips_2011_motor$Origin,
                 Destination = od_trips_2011_motor$Destination)
#the list of GTA06 zones within each "CID" id)
LISTS <- data.frame(CID = ham_taz_catch_2011_public$CID,
                  check_list = I(ham_taz_catch_2011_public$GTA06_in_catch))

LISTS_catch_2011_public <- check_within_lists(df1 = ODs, df2= LISTS)

LISTS <- data.frame(CID = ham_taz_catch_2011_cath$CID,
                  check_list = I(ham_taz_catch_2011_cath$GTA06_in_catch))
LISTS_catch_2011_cath <- check_within_lists(df1 = ODs, df2= LISTS)

## FOR 2016
ODs <- data.frame(Origin = od_trips_2016_motor$Origin,
                 Destination = od_trips_2016_motor$Destination)

LISTS <- data.frame(CID = ham_taz_catch_2016_public$CID,
                  check_list = I(ham_taz_catch_2016_public$GTA06_in_catch))

LISTS_catch_2016_public <- check_within_lists(df1 = ODs, df2= LISTS)

LISTS <- data.frame(CID = ham_taz_catch_2016_cath$CID,
                  check_list = I(ham_taz_catch_2016_cath$GTA06_in_catch))
LISTS_catch_2016_cath <- check_within_lists(df1 = ODs, df2= LISTS)

rm(ODs,LISTS)
```

```{r}
LISTS_catch_2011_public <- LISTS_catch_2011_public %>% 
  merge(od_trips_2011_motor, by=c("Origin", "Destination")) %>% 
  mutate(Trips_perc = Trips/sum(Trips))

LISTS_catch_2011_cath <- LISTS_catch_2011_cath %>% 
  merge(od_trips_2011_motor, by=c("Origin", "Destination")) %>% 
  mutate(Trips_perc = Trips/sum(Trips))

LISTS_catch_2016_public <- LISTS_catch_2016_public %>% 
  merge(od_trips_2016_motor, by=c("Origin", "Destination")) %>% 
  mutate(Trips_perc = Trips/sum(Trips))

LISTS_catch_2016_cath <- LISTS_catch_2016_cath %>% 
  merge(od_trips_2016_motor, by=c("Origin", "Destination")) %>% 
  mutate(Trips_perc = Trips/sum(Trips))
```
```{r}
(LISTS_catch_2011_public$Trips_perc*LISTS_catch_2011_public$result) %>% sum()
(LISTS_catch_2011_cath$Trips_perc*LISTS_catch_2011_cath$result) %>% sum()
(LISTS_catch_2016_public$Trips_perc*LISTS_catch_2016_public$result) %>% sum()
(LISTS_catch_2016_cath$Trips_perc*LISTS_catch_2016_cath$result) %>% sum()

# these sums make sense, almost all CIDs (minus 1 small CID in the public catchments) are represented... e.g., checking the count of 'LISTS_catch_2011_public$CID %>% unique() %>% count()' against how many invidual CIDs there are. 
```
Okay, so here we see approximately ~80% of trips that are 'motorized' made by students in Hamilton (5-14 age) to hamilton schools are made within catchment.

Interestingly, if we repeat this process for the non-motorized trips -- almost 100% of trips are made within catchment! This is an interesting but intuitive finding. Using a wider-range modes allows you to reach larger distance. In a way, those who walk/cycle have a reduced luxury of school-seat choice.

## Calculating car travel time for OD of hamilton TAZs

Libraries:
```{r setup, include=FALSE, eval=FALSE}
library(disk.frame)
library(ggplot2)
library(gridExtra)
library(httr)
library(knitr)
#install.packages("kableExtra", dependencies = TRUE) #installing kableExtra and dependencies as I was receiving errors
library(kableExtra)
#install.packages("devtools")
#install.packages("Rtools")
#devtools::install_github("https://github.com/hrbrmstr/lodes.git", dependencies = TRUE) # if you need to download
library(lodes)
library(progress)
library(purrr)
library(r5r)
#library(stplanr)
library(tidyr)
library(tmap)
library(zoo) # for rollapplyr

# setup for disk.frame
setup_disk.frame()
options(scipen = 999)
options(java.parameters = "-Xmx512m")
options(future.globals.maxSize = Inf)
```

Format the destination and origin points (ID, Long, Lat):
```{r, eval=FALSE}
ham_taz_coords <-  
  cbind(ham_taz %>% st_drop_geometry()%>% transmute(GTA06), st_centroid(ham_taz) %>% st_transform(crs = 4326) %>% st_coordinates()) %>%
  rename(lon = "X", lat = "Y", id = "GTA06") 
```
Set-up directories and download the osm network:
```{r set up r5 path, include=FALSE, eval=FALSE}
# the r5r package requires Java Development Kit version 11, which can be downloaded from https://www.oracle.com/java/technologies/javase-jdk11-downloads.html
dir.create("data-inputs/TTS/r5_graph")
r5_path <- file.path("data-inputs/TTS/r5_graph")
```

```{r download data, include=FALSE, eval=FALSE}
# downloading ontario osm in the correct format
# download.file(url = paste0("https://download.geofabrik.de/north-america/canada/ontario-latest.osm.pbf"),
#               destfile = file.path(r5_path, "osm.pbf"),
#               mode = "wb")

```
Set Up R5 Routing:
```{r build graph, include = FALSE, eval=FALSE}
# NOTE: I had signficant issues getting this function to run. Since the most recent r5r update (before Nov ~25 2022 when I was working on this), I believe the setup_r5 requires smaller road networks. So putting the full ontario geofabrik extract no longer works for me. I used http://share.mapbbcode.org/ to export the Hamilton area, and then OSMCONVERT to convert the .osm file into a .pbf.
r5_HAM <- setup_r5(data_path = r5_path, verbose = TRUE, overwrite = TRUE, temp_dir = TRUE)

```

Now calculate the OD matrix, for car travel time from all origins (residential GTA06) to all destinations (school GTA06):
```{r, eval=FALSE}
# set up batching according to how many origin rows to process at one time
chunksize = 50 
num_chunks = ceiling(nrow(ham_taz_coords)/chunksize)

# create origin-destination pairs
origins_chunks <- as.disk.frame(ham_taz_coords,
                          outdir = "data-inputs/df/Orig",
                          nchunks = num_chunks,
                          overwrite = TRUE)

start.time <- Sys.time()
pb <- txtProgressBar(0, num_chunks, style = 3)

for (i in 1:num_chunks){
  Orig_chunk <- get_chunk(origins_chunks, i)
  ttm_chunk <- travel_time_matrix(r5r_core = r5_HAM,
                          origins = Orig_chunk,
                          destinations = ham_taz_coords,
                          mode = c("CAR"), 
                          departure_datetime = as.POSIXct(strptime("2022-11-16 08:00:00", "%Y-%m-%d %H:%M:%S", tz = "EST5EDT")),
                          max_trip_duration = 120)

  # export output as disk.frame
  ifelse(i == 1, output_df <- as.disk.frame(ttm_chunk,
                                            nchunks = 1,
                                            outdir = "data-inputs/df/output_ttm",
                                            compress = 50,
                                            overwrite = TRUE),
         add_chunk(output_df, ttm_chunk, chunk_id = i))
  setTxtProgressBar(pb, i)
}
end.time <- Sys.time()
print(paste0("OD matrix calculation took ", round(difftime(end.time, start.time, units = "mins"), digits = 2), " minutes..."))

output_OD_car <- as.data.frame(output_df)

save("output_OD_car", file = "data-inputs/TTS/output_OD_car.Rdata")
#I only made the cut off 60 minutes, so if some parcels do not have a travel time, rerun this for a 120 min cut off. The following are 120mins cutoff
```

```{r, eval=FALSE}
# set up batching according to how many origin rows to process at one time
chunksize = 50 
num_chunks = ceiling(nrow(ham_taz_coords)/chunksize)

# create origin-destination pairs
origins_chunks <- as.disk.frame(ham_taz_coords,
                          outdir = "data-inputs/df/Orig",
                          nchunks = num_chunks,
                          overwrite = TRUE)

start.time <- Sys.time()
pb <- txtProgressBar(0, num_chunks, style = 3)

for (i in 1:num_chunks){
  Orig_chunk <- get_chunk(origins_chunks, i)
  ttm_chunk <- travel_time_matrix(r5r_core = r5_HAM,
                          origins = Orig_chunk,
                          destinations = ham_taz_coords,
                          mode = c("WALK"), 
                          departure_datetime = as.POSIXct(strptime("2022-11-16 08:00:00", "%Y-%m-%d %H:%M:%S", tz = "EST5EDT")),
                          max_trip_duration = 120)

  # export output as disk.frame
  ifelse(i == 1, output_df <- as.disk.frame(ttm_chunk,
                                            nchunks = 1,
                                            outdir = "data-inputs/df/output_ttm",
                                            compress = 50,
                                            overwrite = TRUE),
         add_chunk(output_df, ttm_chunk, chunk_id = i))
  setTxtProgressBar(pb, i)
}
end.time <- Sys.time()
print(paste0("OD matrix calculation took ", round(difftime(end.time, start.time, units = "mins"), digits = 2), " minutes..."))

output_OD_walk <- as.data.frame(output_df)

save("output_OD_walk", file = "data-inputs/TTS/output_OD_walk.Rdata")
```
Java garbage collector:
```{r, eval=FALSE}
stop_r5(r5_HAM)
rJava::.jgc(R.gc = TRUE)
```

## Inspecting the travel times and calculating TLD
```{r}
load(file = "data-inputs/TTS/output_OD_car.Rdata")
summary(output_OD_car)
#we'll set 0 travel times to equal 7 min for inter-zonal travel, half of the 1st Qu. travel time.

load(file = "data-inputs/TTS/output_OD_walk.Rdata")
summary(output_OD_walk) #max trip length calculated is 27 mins -- this corresponds to 'within catchment'. lets make 0 min trips into half of 1st Q. so 4min.
```

Trip length distribution (cutting the travel_time into 150 intervals and summing all trips within each interval). The TLD is a probability density function for likelihood of travel informed by travel time:
```{r}
#join travel_time to 2011 and 2016 OD trips
od_trips_2011_motor <- od_trips_2011_motor %>% 
  merge(output_OD_car, by.x = c("Origin", "Destination"), by.y = c("from_id", "to_id")) %>%
  merge(ham_taz %>% st_drop_geometry(), by.x="Origin", by.y="GTA06")

od_trips_2011_motor <- od_trips_2011_motor %>% 
  mutate(travel_time = ifelse(travel_time_p50 == 0, (7)*AREA/max(od_trips_2011_motor %>% filter(travel_time_p50 == 0) %>% dplyr::select(AREA)),
                                  travel_time_p50)) #~7 mins is Q1 for 'tl_2016_motor.

od_trips_2016_motor <- od_trips_2016_motor %>% 
  merge(output_OD_car, by.x = c("Origin", "Destination"), by.y = c("from_id", "to_id")) %>%
  merge(ham_taz %>% st_drop_geometry(), by.x="Origin", by.y="GTA06")

od_trips_2016_motor <- od_trips_2016_motor %>% 
  mutate(travel_time = ifelse(travel_time_p50 == 0, (7)*AREA/max(od_trips_2016_motor %>% filter(travel_time_p50 == 0) %>% dplyr::select(AREA)),
                                  travel_time_p50)) ##~7 mins is Q1 for 'tl_2016_motor.

od_trips_2011_nonmotor <- od_trips_2011_nonmotor %>%
  merge(output_OD_walk, by.x = c("Origin", "Destination"), by.y = c("from_id", "to_id")) %>%
  merge(ham_taz %>% st_drop_geometry(), by.x="Origin", by.y="GTA06")

od_trips_2011_nonmotor <- od_trips_2011_nonmotor %>% 
  mutate(travel_time = ifelse(travel_time_p50 == 0, (15)*AREA/max(od_trips_2011_nonmotor %>% filter(travel_time_p50 == 0) %>% dplyr::select(AREA)),
                                  travel_time_p50)) ##~15 mins is Q1 for 'tld_2016_nonmotor.

od_trips_2016_nonmotor <- od_trips_2016_nonmotor %>% 
  merge(output_OD_walk, by.x = c("Origin", "Destination"), by.y = c("from_id", "to_id"))  %>%
  merge(ham_taz %>% st_drop_geometry(), by.x="Origin", by.y="GTA06")

od_trips_2016_nonmotor <- od_trips_2016_nonmotor %>% 
  mutate(travel_time = ifelse(travel_time_p50 == 0, (15)*AREA/max(od_trips_2016_nonmotor %>% filter(travel_time_p50 == 0) %>% dplyr::select(AREA)),
                                  travel_time_p50)) #~15 mins is Q1 for 'tld_2016_nonmotor.
```

```{r}
tld_2011_motor <- od_trips_2011_motor %>%
  mutate(tt_classes = cut(travel_time, 
                          150,
                          ordered_result = TRUE)) %>%
  group_by(tt_classes) %>%
  summarize(Trips = sum(Trips),
            travel_time = mean(travel_time))

tld_2016_motor <- od_trips_2016_motor %>%
  mutate(tt_classes = cut(travel_time, 
                          150,
                          ordered_result = TRUE)) %>%
  group_by(tt_classes) %>%
  summarize(Trips = sum(Trips),
            travel_time = mean(travel_time))

tld_2011_nonmotor <- od_trips_2011_nonmotor %>%
  mutate(tt_classes = cut(travel_time, 
                          150,
                          ordered_result = TRUE)) %>%
  group_by(tt_classes) %>%
  summarize(Trips = sum(Trips),
            travel_time = mean(travel_time))

tld_2016_nonmotor <- od_trips_2016_nonmotor %>%
  mutate(tt_classes = cut(travel_time, 
                          150,
                          ordered_result = TRUE)) %>%
  group_by(tt_classes) %>%
  summarize(Trips = sum(Trips),
            travel_time = mean(travel_time))
```

Plot the TLD, motorized:
```{r}
ggplot(data = tld_2011_motor, 
       aes(x =travel_time, y = Trips)) +
  geom_point()

ggplot(data = tld_2016_motor, 
       aes(x =travel_time, y = Trips)) +
  geom_point()

summary(tld_2011_motor)
summary(tld_2016_motor)
```

Plot the TLD, nonmotorized:
```{r}
ggplot(data = tld_2011_nonmotor, 
       aes(x =travel_time, y = Trips)) +
  geom_point()

ggplot(data = tld_2016_nonmotor, 
       aes(x =travel_time, y = Trips)) +
  geom_point()

summary(tld_2011_nonmotor)
summary(tld_2016_nonmotor)
```

Lets fit a curve.

First, let's transform the data so each travel time for each OD trip is a row
```{r data-for-Hamilton-trips, include=FALSE}
#for 2011
all_tt_2011_motor <- od_trips_2011_motor  %>% 
  dplyr::select(Trips, travel_time)
sum((od_trips_2011_motor$Trips))
all_tt_2011_motor <- all_tt_2011_motor[rep(seq_len(dim(all_tt_2011_motor)[1]), all_tt_2011_motor$Trips), 2]

all_tt_2011_nonmotor <- od_trips_2011_nonmotor  %>% 
  dplyr::select(Trips, travel_time)
sum((all_tt_2011_nonmotor$Trips))
all_tt_2011_nonmotor <- all_tt_2011_nonmotor[rep(seq_len(dim(all_tt_2011_nonmotor)[1]), all_tt_2011_nonmotor$Trips), 2]

#for 2016
all_tt_2016_motor <- od_trips_2016_motor  %>% 
  dplyr::select(Trips, travel_time)
sum((all_tt_2016_motor$Trips))
all_tt_2016_motor <- all_tt_2016_motor[rep(seq_len(dim(all_tt_2016_motor)[1]), all_tt_2016_motor$Trips), 2]

all_tt_2016_nonmotor <- od_trips_2016_nonmotor  %>% 
  dplyr::select(Trips, travel_time)
sum((all_tt_2016_nonmotor$Trips))
all_tt_2016_nonmotor <- all_tt_2016_nonmotor[rep(seq_len(dim(all_tt_2016_nonmotor)[1]), all_tt_2016_nonmotor$Trips), 2]

```

```{r fitting-impedance-function-2011-2016-1, cache=TRUE}
library(fitdistrplus)
descdist(all_tt_2011_motor, discrete=FALSE)
descdist(all_tt_2016_motor, discrete=FALSE)

descdist(all_tt_2011_nonmotor, discrete=FALSE)
descdist(all_tt_2016_nonmotor, discrete=FALSE)
```

```{r fitting-impedance-function-2011-2016-2, cache=TRUE}
gamma2011_motor <- fitdistrplus::fitdist(data=all_tt_2011_motor, "gamma", method="mle", optim.method="Nelder-Mead")
exp2011_motor <- fitdistrplus::fitdist(data=all_tt_2011_motor, "exp", method="mme", optim.method="Brent")

gamma2016_motor <- fitdistrplus::fitdist(data=all_tt_2016_motor, "gamma", method="mle", optim.method="Nelder-Mead")
exp2016_motor <- fitdistrplus::fitdist(data=all_tt_2016_motor, "exp", method="mme", optim.method="Brent")

plot(gamma2011_motor)
plot(exp2011_motor)

plot(gamma2016_motor)
plot(exp2016_motor)

gamma2011_motor$aic #gamma has the lowest aic and bic, thus we pick gamma distribution. 
exp2011_motor$aic

gamma2016_motor$aic
exp2016_motor$aic
```

```{r fitting-impedance-function-2011-2016-2, cache=TRUE}
gamma2011_nonmotor <- fitdistrplus::fitdist(data=all_tt_2011_nonmotor, "gamma", method="mle", optim.method="Nelder-Mead")
exp2011_nonmotor <- fitdistrplus::fitdist(data=all_tt_2011_nonmotor, "exp", method="mme", optim.method="Brent")

gamma2016_nonmotor <- fitdistrplus::fitdist(data=all_tt_2016_nonmotor, "gamma", method="mle", optim.method="Nelder-Mead")
exp2016_nonmotor <- fitdistrplus::fitdist(data=all_tt_2016_nonmotor, "exp", method="mme", optim.method="Brent")

plot(gamma2011_nonmotor)
plot(exp2011_nonmotor)

plot(gamma2016_nonmotor)
plot(exp2016_nonmotor)

gamma2011_nonmotor$aic #gamma and exp have similar aic and bic. empirically exp is often selected, so we pick it here too.
exp2011_nonmotor$aic

gamma2016_nonmotor$aic
exp2016_nonmotor$aic
```
```{r save-impedance-plot, include=FALSE, cache=TRUE}
# For some reason plot(norm_) does not play well with knitr, so instead we save the figure and then include it as a graphic in the following chunk
png("Fig-imped2011-motor.png")
plot(gamma2011_motor)
dev.off()
```
```{r save-impedance-plot-2, include=FALSE, cache=TRUE}
# For some reason plot(norm_) does not play well with knitr, so instead we save the figure and then include it as a graphic in the following chunk
png("Fig-imped2016-motor.png")
plot(gamma2016_motor)
dev.off()
```
```{r save-impedance-plot-3, include=FALSE, cache=TRUE}
# For some reason plot(norm_) does not play well with knitr, so instead we save the figure and then include it as a graphic in the following chunk
png("Fig-imped2011-nonmotor.png")
plot(exp2011_nonmotor)
dev.off()
```
```{r save-impedance-plot-4, include=FALSE, cache=TRUE}
# For some reason plot(norm_) does not play well with knitr, so instead we save the figure and then include it as a graphic in the following chunk
png("Fig-imped2016-nonmotor.png")
plot(exp2016_nonmotor)
dev.off()
```

And save the gamma functions, to use in the analysis
```{r}
save(gamma2011_motor, file="data-inputs/TTS/gamma2011_motor.Rdata")
save(gamma2016_motor, file="data-inputs/TTS/gamma2016_motor.Rdata")
save(exp2011_nonmotor, file="data-inputs/TTS/exp2011_nonmotor.Rdata")
save(exp2016_nonmotor, file="data-inputs/TTS/exp2016_nonmotor.Rdata")
```

### 2024-04-09 data-prep/00-Testing-scenarios.Rmd
Draft of 'Manuscript.qmd' for special issue

---
title: "Testing-doc"
output: html_document
date: "2024-02-14"
---
```{r}
rm(list=ls())
```

```{r}
library(dplyr)
library(sf)
library(ggplot2)
library(tidyverse)
library(units)
library(patchwork)
library(ggspatial)
```

# Scenario 1 (WE ASSUME THIS FOR NOW): What if, all parcels are occupied with kids. And the average DA kid pop divided by parcels in the DA is assigned to each parcel and divided up by the number of OD trips. Effectively: each parcel has occupied with ~0.002 children that go to ALL schools by either walking (if travel time less than 27) AND/or driving. walking children are avg ~0.2% and driving children are on avg ~0.8%. 
```{r}
#test <- OD_2011 |> filter(GeoUID == 35250032 | GeoUID == 35250034)
load(file="test2.RData")
load(file="data-prep/data-products/cen_data_2011_hamiltononly.Rdata")
cen_data_2011_hamiltononly <- cen_data_2011_1
rm("cen_data_2011_1")
```

```{r}
test2 |> group_by(GeoUID) |> summarize(DA_pop_5to14 = mean(DA_pop_5to14)) |> select(DA_pop_5to14) |> sum()
```

```{r}
test_ID_count <- test2 |> group_by(ID) |> summarise(ID_count = n())
test2 <- test2 |> left_join(test_ID_count, by="ID")
test_IDsinDA_count <- test2 |> group_by(GeoUID) |> summarise(DA_ID_count = sum(ID_count))
test2 <- test2 |> left_join(test_IDsinDA_count, by="GeoUID")

test2 <- test2 |> mutate(kid_pop = (DA_pop_5to14/DA_ID_count.y)*ID_count, #SEE NOTE
               kid_pop = ifelse(is.na(kid_pop) | is.nan(kid_pop) | kid_pop == Inf, 0, kid_pop),
               kid_popadj = case_when(mode == "w" & mode_count == 2 ~ (1.0000000-TAZOrig_car_perc)*2*kid_pop,
                                         mode == "c" & mode_count == 2 ~ (TAZOrig_car_perc)*2*kid_pop,
                                         mode == "c" & mode_count == 1 ~ kid_pop),
                  kid_popadj = ifelse(is.na(kid_popadj) | is.nan(kid_popadj) | kid_popadj == Inf, 0, kid_popadj))

test2$kid_pop %>% sum()
test2$kid_popadj %>% sum()

#NOTE: ID_Count sum is equal to the number of rows (28475 -- the number that OD pairs (that originate in the 1 DA) appears in this 1 DA (each parcel goes to 148 schools by 1 mode (and sometimes another mode)). The "sum(ID_count)) should be the total number of rows in a DA if the numerator is the population. The *ID_count is in our case the number of times that ID appears. So the numerator (# of children in the DA) divided by the number of OD pairs (originated in the DA) times the number of parcels is the assumed number of children at each parcel. IN this case, it is assumed that all parcels have equal number of children -- but likely some households in fact do NOT have children. Maybe we should do a random assignment of full integer children (either 0, 1, 2, 3) to parcels instead of averages?
```

# Scenario 2: just integer children. Each parcel is randomly assigned to have either 0,1,2,3 children. All parcels are still assumed to make trips to ALL schools at the same ~%80:%20 drive/walk proportion (i.e., those 'integer' children are divided evenly to have an equal proportion going to all schools). We also assume each household uses only 1 mode.

```{r}
#GENERATED USING CHATGPT Feb 15 - the last number of vector is always biased
generate_vector_binomial <- function(length, total_sum) {
  vec <- integer(length)  # Initialize an empty integer vector

  # Loop until the sum matches the desired total_sum
  while (sum(vec) != total_sum) {
    # Generate a vector of length 'length' with random numbers from binomial distribution
    vec <-  rbinom(length, size = 15, prob = 0.03) #rgamma(length, shape = 2, rate = 0.2)

    #  # Scale the vector to integers between 0 and 5
    # vec <- as.integer(5 * (vec / max(vec)))

    # Adjust the last element to ensure the total sum is correct
    vec[length] <- total_sum - sum(vec[-length])

    # Ensure the adjusted last element is within the range [0, 5]
    vec[length] <- pmin(vec[length], 5)
    vec[length] <- pmax(vec[length], 0)

    # Shuffle the vector to randomize the order
    vec <- sample(vec)
  }

  return(vec)
}

# Example usage
length <-  160  # Change this to your desired length
total_sum <- 50  # Change this to your desired sum
result_vector  <- generate_vector_binomial(length, total_sum)
result_vector  
result_vector |> sum()
```

```{r}
test_scen2_ID_kid <- test |> group_by(GeoUID) |> 
  summarise(ID_unique_count = n_distinct(ID),
            DA_pop_5to13 = mean(DA_pop_5to13)) |>
  mutate(kid_pop_int = map2(ID_unique_count, DA_pop_5to13, generate_vector_gamma))

ID_vector <- data.frame(ID = test$ID |> unique(),
                        test_scen2_ID_kid |> select(kid_pop_int) |> unnest(kid_pop_int))

test <- test |> left_join(ID_vector, by="ID") 
```

```{r}
test <- test |> mutate(kid_pop_int_adj = case_when(mode == "w" & mode_count == 2 
                                           ~ (1.0000000-TAZOrig_car_perc)*2*kid_pop_int/ID_count,
                                         mode == "c" & mode_count == 2 
                                         ~ (TAZOrig_car_perc)*2*kid_pop_int/ID_count,
                                         mode == "c" & mode_count == 1 
                                         ~ kid_pop_int/ID_count),
               kid_pop_int_adj = ifelse(is.na(kid_pop_int_adj) | 
                                          is.nan(kid_pop_int_adj) | 
                                          kid_pop_int_adj == Inf, 
                                        0, kid_pop_int_adj))
#checking to make sure they all add up to the number of children in this DA
test$kid_pop %>% sum()
test$kid_popadj %>% sum()
test$kid_pop_int_adj %>% sum()
```

### 2024-04-09 data-prep/01-processing-schools-parcels-census.qmd
Draft of 'Manuscript.qmd' for special issue

---
title: "01-processing-schools-parcels-census"
format: html
---
Processing canadian census at the DA level, schools,  school catchments, and then household parcels (this data is not included in the repo, it is strictly available only through TERANET INC.).

```{r library-setup, message=FALSE}
library(tmap)
library(tidyverse)
library(cancensus)
library(sf)
```

```{r}
#import boundaries downloaded from Hamilton Open Data portal
ham_city_bound<- st_read("data-inputs/Boundaries/City_Boundary.shp")
ham_community_bounds<- st_read("data-inputs/Boundaries/Community_Boundaries.shp")
```

# Census
```{r import census data 2011}
set_cancensus_api_key("CensusMapper_53e5295237d9965a06425f1138ee6db7", install=TRUE, overwrite = TRUE) # this is my key, create your own account and get a key here: https://censusmapper.ca/

#explore census data for hamilton, you see here that the Hamilton is available at the level of CMA, CD, and CSD. Do this for both 2011 and 2016 census.

list_census_regions('CA11') %>%
   filter(name == c("Hamilton")) 
list_census_regions('CA16') %>%
   filter(name == c("Hamilton")) #from this result we see the code for the Hamilton CMA is 35537
```

```{r}
find_census_vectors("prevalence of low income", dataset = "CA16", type = "total", query_type = "keyword", interactive = FALSE)
```

```{r import census data 2011}
# Let's call the DA-level data by their vector IDs. Note, to retrieve all vector IDs you need, visit https://censusmapper.ca/, go to the API tab, and search for Variables and their corresponding IDs for your region and aggergation level of interest. Very user-friendly GUI! 
 cen_data_2011_1 <- get_census(dataset='CA11', regions=list(CMA="35537"),
                           vectors=c("v_CA11N_2458", #family median after-tax income in 2010, 
                                     "v_CA11N_2476", #lone-parent median after-tax family income in 2010
                                     "v_CA11N_2564", #private household Median after-tax income of households in 2010,
                                     "v_CA11N_2609", #CA 2011 NHS; Income; Status; Prevalence of low income in 2010 based on after-tax low-income measure %; Less than 18 years %
                                     "v_CA11F_11", # population age 5-9
                                     "v_CA11F_14", #population age 10-14
                                     "v_CA11F_17"), #population age 15-19
                           level='DA', use_cache = FALSE, geo_format = 'sf')

save(cen_data_2011_1, file="data-inputs/Census-data/cen_data_2011_1.RData")
```

Repeat for 2016 incomes:
```{r}
cen_data_2016_1 <- get_census(dataset='CA16', regions=list(CMA="35537"),
                           vectors=c("v_CA16_2448", #family median after-tax income in 2015,
                                     "v_CA16_2460", #lone-parent median after-tax family income in 2015
                                     "v_CA16_2398", #private household Median after-tax income of households in 2015,
                                     "v_CA16_2543", #	Income; Low Income Measures; Prevalence of low income based on the Low-income measure, after tax (LIM-AT) (%); 0 to 17 years (%)
                                     "v_CA16_25", # population age 5-9
                                     "v_CA16_43", #population age 10-14
                                     "v_CA16_64"), #population age 15-19
                           level='DA', use_cache = FALSE, geo_format = 'sf')

save(cen_data_2016_1, file="data-inputs/Census-data/cen_data_2016_1.RData")
```

Hamilton CMA is larger than the hamilton boundary. Clip the census data and resave:
```{r}
ham_city_bound <- ham_city_bound |> st_transform(st_crs(cen_data_2011_1))
ham_community_bounds <- ham_community_bounds |> st_transform(st_crs(cen_data_2011_1))

cen_data_2011_1 <- st_intersection(ham_city_bound |> select(geometry), cen_data_2011_1)
cen_data_2016_1 <- st_intersection(ham_city_bound |> select(geometry), cen_data_2016_1)

save(cen_data_2011_1, file="data-products/cen_data_2011_hamiltononly.RData")
save(cen_data_2016_1, file="data-products/cen_data_2016_hamiltononly.RData")
```

# Schools and catchments:
Available through the hamONdests package (https://github.com/soukhova/hamONdests):
```{r}
library(hamONdests)
data(Schools_2011_2016)
data(School_Catchments_2011_2016)
```

```{r}
schools_map <- tm_shape(cen_data_2011_1 %>% rename("LIM-ATperc_0to17" = "v_CA11N_2609..Less.than.18.years..")) +
  tm_polygons("LIM-ATperc_0to17", palette = 'Reds')+

  tm_shape(Schools_2011_2016 |> filter(Level == "Elementary" & (Year == "2011" | Year == "2011 and 2016"))) +
	tm_bubbles("OTGC2011", col = "grey",alpha=0.5, contrast=1, title.size="School OTGC") +

	tm_layout(legend.bg.color = "grey90", legend.bg.alpha=.5, legend.frame=TRUE)

tmap_leaflet(schools_map, add.titles = TRUE)
```

```{r}
school_catchments_map <- tm_shape(cen_data_2011_1 %>% rename("LIM-ATperc_0to17" = "v_CA11N_2609..Less.than.18.years..")) +
  tm_polygons("LIM-ATperc_0to17", palette = 'Reds')+

  tm_shape(School_Catchments_2011_2016 |> filter(Level == "Elementary" & System == "Public" & (Year == "2011" | Year == "2011 and 2016"))) +
	tm_polygons(col = "yellow",alpha=0.3, contrast=1) +

	tm_layout(legend.bg.color = "grey90", legend.bg.alpha=.5, legend.frame=TRUE)

tmap_leaflet(school_catchments_map, add.titles = TRUE)
```

# Parcels

Read shapefile as "sf" and drop all non-residential parcels:
```{r}
load(file="data-inputs/Census-data/cen_data_2011_1.RData")
load(file="data-inputs/Census-data/cen_data_2016_1.RData")
RED_LU_2011 <- st_read("data-inputs/Parcels/PED_LANDUSE_2011/PED_LANDUSE.shp") |> 
  filter(TYPE == "Residential") |> st_transform(st_crs(cen_data_2011_1)) # set the CRS to epsg 4326, same as the census data

RED_LU_2016 <- st_read("data-inputs/Parcels/PED_Landuse_2016/PED_LANDUSE.shp") |>
  mutate(TYPE = ifelse(LUC1_DESC == "Residential:Detached House" | 
                         LUC1_DESC == "Residential:Row/Town House"| 
                         LUC1_DESC == "Residential:Semi-Detached House" |
                         LUC1_DESC == "Residential:Apartment (7 or more units)"|
                         LUC1_DESC == "Residential:Multiplex Dwelling (6 units or less)",
                       "Residential", 0)) |>
  filter(TYPE != 0) |> #i.e. filter out non-residential
  st_transform(st_crs(cen_data_2016_1))
RED_LU_2016 <- st_make_valid(RED_LU_2016)
```

Transfer the GeoUID from the DAs (I.e., their DA IDs) that intersect with the residential land use parcel centroids to the RED_LU sf object. (we use 'over' here instead of st_intersection because it is way faster)
```{r}
RED_LU_centroids_2011 <- sp::over(as_Spatial(st_centroid(RED_LU_2011)),as_Spatial(cen_data_2011_1 |> select(GeoUID)))
RED_LU_centroids_2011<- cbind(RED_LU_centroids_2011,st_centroid(RED_LU_2011))
```
```{r}
RED_LU_centroids_2016 <- sp::over(as_Spatial(st_centroid(RED_LU_2016)),as_Spatial(cen_data_2016_1 |> select(GeoUID)))

RED_LU_centroids_2016<- cbind(RED_LU_centroids_2016,st_centroid(RED_LU_2016))
```

Now transfer the TAZ ids to the RED centroids -- this is so we can pull the proportion of motor/non-motorized trips. 
```{r}
ggh_taz <- st_read("data-inputs/TTS/tts06_83_region.shp") |> st_transform(st_crs(cen_data_2011_hamiltononly))
```
Now add the TAZ IDs column to the residential parcel id
```{r}
cols_TAZIDs <-sp::over(as_Spatial(st_centroid(RED_LU_2011)),
                              as_Spatial(ggh_taz |> select(GTA06) |> rename("GTA06_Orig" = "GTA06")))
RED_LU_centroids_2011<- cbind(RED_LU_centroids_2011,cols_TAZIDs)

cols_TAZIDs <- sp::over(as_Spatial(st_centroid(RED_LU_2016)),
                              as_Spatial(ggh_taz |> select(GTA06) |> rename("GTA06_Orig" = "GTA06")))

RED_LU_centroids_2016<- cbind(RED_LU_centroids_2016,cols_TAZIDs)
rm(cols_TAZIDs)

# and save
save(RED_LU_centroids_2011, file="data-inputs/Parcels/RED_LU_centroids_2011.RData")
save(RED_LU_centroids_2016, file="data-inputs/Parcels/RED_LU_centroids_2016.RData")
```

We also want to make a list of origin-destination TAZ ids and their percentage of car-use:
```{r}
od_trips_2011_motor <- read_delim(file = "data-inputs/TTS/Schooltrip_trips_2011_motorized.txt", 
                      delim = "\t", 
                      col_names = FALSE)
od_trips_2016_motor <- read_delim(file = "data-inputs/TTS/Schooltrip_trips_2016_motorized.txt", 
                      delim = "\t", 
                      col_names = FALSE)

od_trips_2011_nonmotor <- read_delim(file = "data-inputs/TTS/Schooltrip_trips_2011_nonmotorized.txt", 
                      delim = "\t", 
                      col_names = FALSE)

od_trips_2016_nonmotor <- read_delim(file = "data-inputs/TTS/Schooltrip_trips_2016_nonmotorized.txt", 
                      delim = "\t", 
                      col_names = FALSE)

#Find the position in the table where the information on work trips begins:
idx <- which(od_trips_2011_motor$X1 == "COLUMN : gta06_sch")

#Slice the table to obtain the work trips (remove heading information):
od_trips_2011_motor <- od_trips_2011_motor %>% 
  slice((idx+2):n())

#Separate the zone identifiers and the trips, and convert to numeric:
od_trips_2011_motor <- od_trips_2011_motor %>%
  separate(X1, into = c("Zones", "Trips"), sep = " (?=[^ ]+$)") %>%
  mutate(Zones = str_trim(Zones, side = "both")) %>%
  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") %>%
  mutate(Origin = str_trim(Origin),
         Destination = str_trim(Destination),
         Trips = as.numeric(Trips))

idx <- which(od_trips_2011_nonmotor$X1 == "COLUMN : gta06_sch")

od_trips_2011_nonmotor <- od_trips_2011_nonmotor %>% 
  slice((idx+2):n())

od_trips_2011_nonmotor <- od_trips_2011_nonmotor %>%
  separate(X1, into = c("Zones", "Trips"), sep = " (?=[^ ]+$)") %>%
  mutate(Zones = str_trim(Zones, side = "both")) %>%
  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") %>%
  mutate(Origin = str_trim(Origin),
         Destination = str_trim(Destination),
         Trips = as.numeric(Trips))
```

Repeat the same steps for 2016:
```{r}
idx <- which(od_trips_2016_motor$X1 == "COLUMN : gta06_sch")

od_trips_2016_motor <- od_trips_2016_motor %>% 
  slice((idx+2):n())

od_trips_2016_motor <- od_trips_2016_motor %>%
  separate(X1, into = c("Zones", "Trips"), sep = " (?=[^ ]+$)") %>%
  mutate(Zones = str_trim(Zones, side = "both")) %>%
  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") %>%
  mutate(Origin = str_trim(Origin),
         Destination = str_trim(Destination),
         Trips = as.numeric(Trips))

idx <- which(od_trips_2016_nonmotor$X1 == "COLUMN : gta06_sch")

od_trips_2016_nonmotor <- od_trips_2016_nonmotor %>% 
  slice((idx+2):n())

od_trips_2016_nonmotor <- od_trips_2016_nonmotor %>%
  separate(X1, into = c("Zones", "Trips"), sep = " (?=[^ ]+$)") %>%
  mutate(Zones = str_trim(Zones, side = "both")) %>%
  separate(Zones, into = c("Origin", "Destination"), sep = " (?=[^ ]+$)") %>%
  mutate(Origin = str_trim(Origin),
         Destination = str_trim(Destination),
         Trips = as.numeric(Trips))
```

Merge the the motor/non-motorized objects together:
```{r}
od_trips_perc_2011 <- od_trips_2011_motor %>% left_join(od_trips_2011_nonmotor, by=c("Origin", "Destination")) %>%
  group_by(Origin, Destination) |>
  summarize(TAZOrig_car_trips= sum(Trips.x, na.rm=TRUE),
            TAZOrig_walk_trips = sum(Trips.y, na.rm=TRUE)) |>
  mutate(TAZOrig_car_trip_perc = TAZOrig_car_trips/(TAZOrig_car_trips + TAZOrig_walk_trips))

od_trips_perc_2016 <- od_trips_2016_motor %>% left_join(od_trips_2016_nonmotor, by=c("Origin", "Destination")) %>%
  group_by(Origin,Destination) |>
  summarize(TAZOrig_car_trips = sum(Trips.x, na.rm=TRUE),
            TAZOrig_walk_trips = sum(Trips.y, na.rm=TRUE)) |>
  mutate(TAZOrig_car_trip_perc = TAZOrig_car_trips/(TAZOrig_car_trips + TAZOrig_walk_trips))
```

```{r}
save(od_trips_perc_2011, file="data-inputs/TTS/od_trips_perc_2011.RData")
save(od_trips_perc_2016, file="data-inputs/TTS/od_trips_perc_2016.RData")
```

### 2024-04-09 data-prep/02-travel-times-ODs.qmd
Draft of 'Manuscript.qmd' for special issue

---
title: "02-travel-times"
format: html
---
```{r}
library(disk.frame)
library(tidyverse)
library(r5r)
library(sf)
library(tmap)

# setup for disk.frame
setup_disk.frame()
options(scipen = 999)
options(java.parameters = "-Xmx6G")
```

# Prepare Input Data
Loading the destination points (schools, available through the hamONdests package (https://github.com/soukhova/hamONdests)) and origins parcel points:

```{r}
library(hamONdests)
data(Schools_2011_2016)
data(School_Catchments_2011_2016)
load(file = "data-inputs/Parcels/RED_LU_centroids_2011.Rdata")
load(file = "data-inputs/Parcels/RED_LU_centroids_2016.Rdata")
```

Format the destination points (ID, Long, Lat):
```{r}
Dest_all_schools_elem_2011 <- cbind(as.data.frame(Schools_2011_2016), st_coordinates(st_transform(Schools_2011_2016, crs = 4326))) %>% filter(Level == "Elementary" & (Year == "2011" | Year == "2011 and 2016")) %>%
  rename(lon = "X", lat = "Y", id = "SchoolID") %>%
  select(id, lon, lat)

Dest_all_schools_elem_2016 <- cbind(as.data.frame(Schools_2011_2016), st_coordinates(st_transform(Schools_2011_2016, crs = 4326))) %>% filter(Level == "Elementary" & (Year == "2016" | Year == "2011 and 2016")) %>%
  rename(lon = "X", lat = "Y", id = "SchoolID") %>%
  select(id, lon, lat)
```

Format the origin points (ID, Long, Lat):
```{r}
Orig_all_2011 <- cbind(RED_LU_centroids_2011,
                       st_coordinates(st_transform(RED_LU_centroids_2011 |> st_as_sf(), crs = 4326))) %>%
  rename(lon = X, lat = Y, id = ID) %>%
  select(id, lon, lat)

Orig_all_2016 <- cbind(RED_LU_centroids_2016,
                       st_coordinates(st_transform(RED_LU_centroids_2016 |> st_as_sf(), crs = 4326))) %>%
  rename(lon = X, lat = Y, id = ID) %>%
  select(id, lon, lat)
```

Subset data into 4 chunks(ID, Long, Lat):
```{r}
Orig_all_2011_1 <- Orig_all_2011[1:36903,]
Orig_all_2011_2 <- Orig_all_2011[36904:73807,]
Orig_all_2011_3 <- Orig_all_2011[73808:110710,]
Orig_all_2011_4 <- Orig_all_2011[110711:134340,]

Orig_all_2016_1 <- Orig_all_2016[1:36903,]
Orig_all_2016_2 <- Orig_all_2016[36904:73807,]
Orig_all_2016_3 <- Orig_all_2016[73808:110710,]
Orig_all_2016_4 <- Orig_all_2016[110711:139467,]
```

# Downloading OSM Network and GTFS Data

```{r set up r5 path, include=FALSE}
# the r5r package requires Java Development Kit version 11, which can be downloaded from https://www.oracle.com/java/technologies/javase-jdk11-downloads.html
dir.create("data-inputs/Travel-times")
r5_path <- file.path("data-inputs/Travel-times")
```

```{r download data, include=FALSE, eval=FALSE}
# downloading ontario osm in the correct format - this was done in 2021.
# download.file(url = paste0("https://download.geofabrik.de/north-america/canada/ontario-latest.osm.pbf"),
#               destfile = file.path(r5_path, "osm.pbf"),
#               mode = "wb")

#download and import GTFS (Hamilton transit data) - this was done in 2021 - but transit is not used in this analysis, so not inculded.
# download.file(url = "https://transitfeeds.com/p/hamilton-street-railway/31/latest/download", destfile = file.path(r5_path, "HSR_transit.zip"), mode = "wb")
```

# Set Up R5 Routing
```{r build graph, include = FALSE}
r5_HAM <- setup_r5(data_path = r5_path, verbose = TRUE, overwrite = TRUE, temp_dir = TRUE)
```

# Calculate OD Matrices

2011 Car travel times:
```{r}
# set up batching according to how many origin rows to process at one time
chunksize = 50 
num_chunks = ceiling(nrow(Orig_all_2011_1)/chunksize)

# create origin-destination pairs
origins_chunks <- as.disk.frame(Orig_all_2011_1,
                          outdir = "data-inputs/Travel-times/Processing/Orig",
                          nchunks = num_chunks,
                          overwrite = TRUE)

start.time <- Sys.time()
pb <- txtProgressBar(0, num_chunks, style = 3)

for (i in 1:num_chunks){
  Orig_chunk <- get_chunk(origins_chunks, i)
  ttm_chunk <- travel_time_matrix(r5r_core = r5_HAM,
                          origins = Orig_chunk,
                          destinations = Dest_all_schools_elem_2011,
                          mode = c("CAR"), 
                          departure_datetime = as.POSIXct(strptime("2021-03-01 08:00:00", "%Y-%m-%d %H:%M:%S", tz = "EST5EDT")),
                          max_trip_duration = 90)

  # export output as disk.frame
  ifelse(i == 1, output_df <- as.disk.frame(ttm_chunk,
                                            nchunks = 1,
                                            outdir = "data-inputs/Travel-Times/Processing/output_ttm",
                                            compress = 50,
                                            overwrite = TRUE),
         add_chunk(output_df, ttm_chunk, chunk_id = i))
  setTxtProgressBar(pb, i)
}
end.time <- Sys.time()
print(paste0("OD matrix calculation took ", round(difftime(end.time, start.time, units = "mins"), digits = 2), " minutes..."))

output_OD_car1 <- as.data.frame(output_df)

save("output_OD_car1", file = "data-products/OD_tt_car_2011_1.Rdata")
rm("output_OD_car1")
```

```{r}
# set up batching according to how many origin rows to process at one time
chunksize = 50 
num_chunks = ceiling(nrow(Orig_all_2011_2)/chunksize)

# create origin-destination pairs
origins_chunks <- as.disk.frame(Orig_all_2011_2,
                          outdir = "data-inputs/Travel-times/Processing/Orig",
                          nchunks = num_chunks,
                          overwrite = TRUE)

start.time <- Sys.time()
pb <- txtProgressBar(0, num_chunks, style = 3)

for (i in 1:num_chunks){
  Orig_chunk <- get_chunk(origins_chunks, i)
  ttm_chunk <- travel_time_matrix(r5r_core = r5_HAM,
                          origins = Orig_chunk,
                          destinations = Dest_all_schools_elem_2011,
                          mode = c("CAR"), 
                          departure_datetime = as.POSIXct(strptime("2021-03-01 08:00:00", "%Y-%m-%d %H:%M:%S", tz = "EST5EDT")),
                          max_trip_duration = 90)

  # export output as disk.frame
  ifelse(i == 1, output_df <- as.disk.frame(ttm_chunk,
                                            nchunks = 1,
                                            outdir = "data-inputs/Travel-Times/Processing/output_ttm",
                                            compress = 50,
                                            overwrite = TRUE),
         add_chunk(output_df, ttm_chunk, chunk_id = i))
  setTxtProgressBar(pb, i)
}
end.time <- Sys.time()
print(paste0("OD matrix calculation took ", round(difftime(end.time, start.time, units = "mins"), digits = 2), " minutes..."))

output_OD_car2 <- as.data.frame(output_df)

save("output_OD_car2", file = "data-products/OD_tt_car_2011_2.Rdata")

rm("output_OD_car2")
```

```{r}
# set up batching according to how many origin rows to process at one time
chunksize = 50 
num_chunks = ceiling(nrow(Orig_all_2011_3)/chunksize)

# create origin-destination pairs
origins_chunks <- as.disk.frame(Orig_all_2011_3,
                          outdir = "data-inputs/Travel-times/Processing/Orig",
                          nchunks = num_chunks,
                          overwrite = TRUE)

start.time <- Sys.time()
pb <- txtProgressBar(0, num_chunks, style = 3)

for (i in 1:num_chunks){
  Orig_chunk <- get_chunk(origins_chunks, i)
  ttm_chunk <- travel_time_matrix(r5r_core = r5_HAM,
                          origins = Orig_chunk,
                          destinations = Dest_all_schools_elem_2011,
                          mode = c("CAR"), 
                          departure_datetime = as.POSIXct(strptime("2021-03-01 08:00:00", "%Y-%m-%d %H:%M:%S", tz = "EST5EDT")),
                          max_trip_duration = 90)

  # export output as disk.frame
  ifelse(i == 1, output_df <- as.disk.frame(ttm_chunk,
                                            nchunks = 1,
                                            outdir = "data-inputs/Travel-Times/Processing/output_ttm",
                                            compress = 50,
                                            overwrite = TRUE),
         add_chunk(output_df, ttm_chunk, chunk_id = i))
  setTxtProgressBar(pb, i)
}
end.time <- Sys.time()
print(paste0("OD matrix calculation took ", round(difftime(end.time, start.time, units = "mins"), digits = 2), " minutes..."))

output_OD_car3 <- as.data.frame(output_df)

save("output_OD_car3", file = "data-products/OD_tt_car_2011_3.Rdata")
rm("output_OD_car3")
```

```{r}
# set up batching according to how many origin rows to process at one time
chunksize = 50 
num_chunks = ceiling(nrow(Orig_all_2011_4)/chunksize)

# create origin-destination pairs
origins_chunks <- as.disk.frame(Orig_all_2011_4,
                          outdir = "data-inputs/Travel-times/Processing/Orig",
                          nchunks = num_chunks,
                          overwrite = TRUE)

start.time <- Sys.time()
pb <- txtProgressBar(0, num_chunks, style = 3)

for (i in 1:num_chunks){
  Orig_chunk <- get_chunk(origins_chunks, i)
  ttm_chunk <- travel_time_matrix(r5r_core = r5_HAM,
                          origins = Orig_chunk,
                          destinations = Dest_all_schools_elem_2011,
                          mode = c("CAR"), 
                          departure_datetime = as.POSIXct(strptime("2021-03-01 08:00:00", "%Y-%m-%d %H:%M:%S", tz = "EST5EDT")),
                          max_trip_duration = 90)

  # export output as disk.frame
  ifelse(i == 1, output_df <- as.disk.frame(ttm_chunk,
                                            nchunks = 1,
                                            outdir = "data-inputs/Travel-Times/Processing/output_ttm",
                                            compress = 50,
                                            overwrite = TRUE),
         add_chunk(output_df, ttm_chunk, chunk_id = i))
  setTxtProgressBar(pb, i)
}
end.time <- Sys.time()
print(paste0("OD matrix calculation took ", round(difftime(end.time, start.time, units = "mins"), digits = 2), " minutes..."))

output_OD_car4 <- as.data.frame(output_df)

save("output_OD_car4", file = "data-products/OD_tt_car_2011_4.Rdata")
rm("output_OD_car4")
```

2016 car travel times:
```{r}
# set up batching according to how many origin rows to process at one time
chunksize = 50 
num_chunks = ceiling(nrow(Orig_all_2016_1)/chunksize)

# create origin-destination pairs
origins_chunks <- as.disk.frame(Orig_all_2016_1,
                          outdir = "data-inputs/Travel-times/Processing/Orig",
                          nchunks = num_chunks,
                          overwrite = TRUE)

start.time <- Sys.time()
pb <- txtProgressBar(0, num_chunks, style = 3)

for (i in 1:num_chunks){
  Orig_chunk <- get_chunk(origins_chunks, i)
  ttm_chunk <- travel_time_matrix(r5r_core = r5_HAM,
                          origins = Orig_chunk,
                          destinations = Dest_all_schools_elem_2016,
                          mode = c("CAR"), 
                          departure_datetime = as.POSIXct(strptime("2021-03-01 08:00:00", "%Y-%m-%d %H:%M:%S", tz = "EST5EDT")),
                          max_trip_duration = 90)

  # export output as disk.frame
  ifelse(i == 1, output_df <- as.disk.frame(ttm_chunk,
                                            nchunks = 1,
                                            outdir = "data-inputs/Travel-Times/Processing/output_ttm",
                                            compress = 50,
                                            overwrite = TRUE),
         add_chunk(output_df, ttm_chunk, chunk_id = i))
  setTxtProgressBar(pb, i)
}
end.time <- Sys.time()
print(paste0("OD matrix calculation took ", round(difftime(end.time, start.time, units = "mins"), digits = 2), " minutes..."))

output_OD_car1 <- as.data.frame(output_df)

save("output_OD_car1", file = "data-products/OD_tt_car_2016_1.Rdata")
rm("output_OD_car1")
```

```{r}
# set up batching according to how many origin rows to process at one time
chunksize = 50 
num_chunks = ceiling(nrow(Orig_all_2016_2)/chunksize)

# create origin-destination pairs
origins_chunks <- as.disk.frame(Orig_all_2016_2,
                          outdir = "data-inputs/Travel-times/Processing/Orig",
                          nchunks = num_chunks,
                          overwrite = TRUE)

start.time <- Sys.time()
pb <- txtProgressBar(0, num_chunks, style = 3)

for (i in 1:num_chunks){
  Orig_chunk <- get_chunk(origins_chunks, i)
  ttm_chunk <- travel_time_matrix(r5r_core = r5_HAM,
                          origins = Orig_chunk,
                          destinations = Dest_all_schools_elem_2016,
                          mode = c("CAR"), 
                          departure_datetime = as.POSIXct(strptime("2021-03-01 08:00:00", "%Y-%m-%d %H:%M:%S", tz = "EST5EDT")),
                          max_trip_duration = 90)

  # export output as disk.frame
  ifelse(i == 1, output_df <- as.disk.frame(ttm_chunk,
                                            nchunks = 1,
                                            outdir = "data-inputs/Travel-Times/Processing/output_ttm",
                                            compress = 50,
                                            overwrite = TRUE),
         add_chunk(output_df, ttm_chunk, chunk_id = i))
  setTxtProgressBar(pb, i)
}
end.time <- Sys.time()
print(paste0("OD matrix calculation took ", round(difftime(end.time, start.time, units = "mins"), digits = 2), " minutes..."))

output_OD_car2 <- as.data.frame(output_df)

save("output_OD_car2", file = "data-products/OD_tt_car_2016_2.Rdata")
rm("output_OD_car2")
```

```{r}
# set up batching according to how many origin rows to process at one time
chunksize = 50 
num_chunks = ceiling(nrow(Orig_all_2016_3)/chunksize)

# create origin-destination pairs
origins_chunks <- as.disk.frame(Orig_all_2016_3,
                          outdir = "data-inputs/Travel-times/Processing/Orig",
                          nchunks = num_chunks,
                          overwrite = TRUE)

start.time <- Sys.time()
pb <- txtProgressBar(0, num_chunks, style = 3)

for (i in 1:num_chunks){
  Orig_chunk <- get_chunk(origins_chunks, i)
  ttm_chunk <- travel_time_matrix(r5r_core = r5_HAM,
                          origins = Orig_chunk,
                          destinations = Dest_all_schools_elem_2016,
                          mode = c("CAR"), 
                          departure_datetime = as.POSIXct(strptime("2021-03-01 08:00:00", "%Y-%m-%d %H:%M:%S", tz = "EST5EDT")),
                          max_trip_duration = 90)

  # export output as disk.frame
  ifelse(i == 1, output_df <- as.disk.frame(ttm_chunk,
                                            nchunks = 1,
                                            outdir = "data-inputs/Travel-Times/Processing/output_ttm",
                                            compress = 50,
                                            overwrite = TRUE),
         add_chunk(output_df, ttm_chunk, chunk_id = i))
  setTxtProgressBar(pb, i)
}
end.time <- Sys.time()
print(paste0("OD matrix calculation took ", round(difftime(end.time, start.time, units = "mins"), digits = 2), " minutes..."))

output_OD_car3 <- as.data.frame(output_df)

save("output_OD_car3", file = "data-products/OD_tt_car_2016_3.Rdata")
rm("output_OD_car3")
```

```{r}
# set up batching according to how many origin rows to process at one time
chunksize = 50 
num_chunks = ceiling(nrow(Orig_all_2016_4)/chunksize)

# create origin-destination pairs
origins_chunks <- as.disk.frame(Orig_all_2016_4,
                          outdir = "data-inputs/Travel-times/Processing/Orig",
                          nchunks = num_chunks,
                          overwrite = TRUE)

start.time <- Sys.time()
pb <- txtProgressBar(0, num_chunks, style = 3)

for (i in 1:num_chunks){
  Orig_chunk <- get_chunk(origins_chunks, i)
  ttm_chunk <- travel_time_matrix(r5r_core = r5_HAM,
                          origins = Orig_chunk,
                          destinations = Dest_all_schools_elem_2016,
                          mode = c("CAR"), 
                          departure_datetime = as.POSIXct(strptime("2021-03-01 08:00:00", "%Y-%m-%d %H:%M:%S", tz = "EST5EDT")),
                          max_trip_duration = 90)

  # export output as disk.frame
  ifelse(i == 1, output_df <- as.disk.frame(ttm_chunk,
                                            nchunks = 1,
                                            outdir = "data-inputs/Travel-Times/Processing/output_ttm",
                                            compress = 50,
                                            overwrite = TRUE),
         add_chunk(output_df, ttm_chunk, chunk_id = i))
  setTxtProgressBar(pb, i)
}
end.time <- Sys.time()
print(paste0("OD matrix calculation took ", round(difftime(end.time, start.time, units = "mins"), digits = 2), " minutes..."))

output_OD_car4 <- as.data.frame(output_df)

save("output_OD_car4", file = "data-products/OD_tt_car_2016_4.Rdata")
rm("output_OD_car4")
```

2011 walk travel times:
```{r}
# set up batching according to how many origin rows to process at one time
chunksize = 50 
num_chunks = ceiling(nrow(Orig_all_2011_1)/chunksize)

# create origin-destination pairs
origins_chunks <- as.disk.frame(Orig_all_2011_1,
                          outdir = "data-inputs/Travel-times/Processing/Orig",
                          nchunks = num_chunks,
                          overwrite = TRUE)

start.time <- Sys.time()
pb <- txtProgressBar(0, num_chunks, style = 3)

for (i in 1:num_chunks){
  Orig_chunk <- get_chunk(origins_chunks, i)
  ttm_chunk <- travel_time_matrix(r5r_core = r5_HAM,
                          origins = Orig_chunk,
                          destinations = Dest_all_schools_elem_2011,
                          mode = c("WALK"),
                          departure_datetime = as.POSIXct(strptime("2021-03-01 08:00:00", "%Y-%m-%d %H:%M:%S", tz = "EST5EDT")),
                          walk_speed = 3.6,
                          max_trip_duration = 60)

  # export output as disk.frame
  ifelse(i == 1, output_df <- as.disk.frame(ttm_chunk,
                                            nchunks = 1,
                                            outdir = "data-inputs/Travel-Times/Processing/output_ttm",
                                            compress = 50,
                                            overwrite = TRUE),
         add_chunk(output_df, ttm_chunk, chunk_id = i))
  setTxtProgressBar(pb, i)
}
end.time <- Sys.time()
print(paste0("OD matrix calculation took ", round(difftime(end.time, start.time, units = "mins"), digits = 2), " minutes..."))

output_OD_walk_elem1 <- as.data.frame(output_df)

save("output_OD_walk_elem1", file = "data-products/OD_tt_walk_2011_1.Rdata")
rm("output_OD_walk_elem1")
```

```{r}
# set up batching according to how many origin rows to process at one time
chunksize = 50 
num_chunks = ceiling(nrow(Orig_all_2011_2)/chunksize)

# create origin-destination pairs
origins_chunks <- as.disk.frame(Orig_all_2011_2,
                          outdir = "data-inputs/Travel-times/Processing/Orig",
                          nchunks = num_chunks,
                          overwrite = TRUE)

start.time <- Sys.time()
pb <- txtProgressBar(0, num_chunks, style = 3)

for (i in 1:num_chunks){
  Orig_chunk <- get_chunk(origins_chunks, i)
  ttm_chunk <- travel_time_matrix(r5r_core = r5_HAM,
                          origins = Orig_chunk,
                          destinations = Dest_all_schools_elem_2011,
                          mode = c("WALK"),
                          departure_datetime = as.POSIXct(strptime("2021-03-01 08:00:00", "%Y-%m-%d %H:%M:%S", tz = "EST5EDT")),
                          walk_speed = 3.6,
                          max_trip_duration = 60)

  # export output as disk.frame
  ifelse(i == 1, output_df <- as.disk.frame(ttm_chunk,
                                            nchunks = 1,
                                            outdir = "data-inputs/Travel-Times/Processing/output_ttm",
                                            compress = 50,
                                            overwrite = TRUE),
         add_chunk(output_df, ttm_chunk, chunk_id = i))
  setTxtProgressBar(pb, i)
}
end.time <- Sys.time()
print(paste0("OD matrix calculation took ", round(difftime(end.time, start.time, units = "mins"), digits = 2), " minutes..."))

output_OD_walk_elem2 <- as.data.frame(output_df)

save("output_OD_walk_elem2", file = "data-products/OD_tt_walk_2011_2.Rdata")
rm("output_OD_walk_elem2")
```

```{r}
# set up batching according to how many origin rows to process at one time
chunksize = 50 
num_chunks = ceiling(nrow(Orig_all_2011_3)/chunksize)

# create origin-destination pairs
origins_chunks <- as.disk.frame(Orig_all_2011_3,
                          outdir = "data-inputs/Travel-times/Processing/Orig",
                          nchunks = num_chunks,
                          overwrite = TRUE)

start.time <- Sys.time()
pb <- txtProgressBar(0, num_chunks, style = 3)

for (i in 1:num_chunks){
  Orig_chunk <- get_chunk(origins_chunks, i)
  ttm_chunk <- travel_time_matrix(r5r_core = r5_HAM,
                          origins = Orig_chunk,
                          destinations = Dest_all_schools_elem_2011,
                          mode = c("WALK"),
                          departure_datetime = as.POSIXct(strptime("2021-03-01 08:00:00", "%Y-%m-%d %H:%M:%S", tz = "EST5EDT")),
                          walk_speed = 3.6,
                          max_trip_duration = 60)

  # export output as disk.frame
  ifelse(i == 1, output_df <- as.disk.frame(ttm_chunk,
                                            nchunks = 1,
                                            outdir = "data-inputs/Travel-Times/Processing/output_ttm",
                                            compress = 50,
                                            overwrite = TRUE),
         add_chunk(output_df, ttm_chunk, chunk_id = i))
  setTxtProgressBar(pb, i)
}
end.time <- Sys.time()
print(paste0("OD matrix calculation took ", round(difftime(end.time, start.time, units = "mins"), digits = 2), " minutes..."))

output_OD_walk_elem3 <- as.data.frame(output_df)

save("output_OD_walk_elem3", file = "data-products/OD_tt_walk_2011_3.Rdata")
rm("output_OD_walk_elem3")
```

```{r}
# set up batching according to how many origin rows to process at one time
chunksize = 50 
num_chunks = ceiling(nrow(Orig_all_2011_4)/chunksize)

# create origin-destination pairs
origins_chunks <- as.disk.frame(Orig_all_2011_4,
                          outdir = "data-inputs/Travel-times/Processing/Orig",
                          nchunks = num_chunks,
                          overwrite = TRUE)

start.time <- Sys.time()
pb <- txtProgressBar(0, num_chunks, style = 3)

for (i in 1:num_chunks){
  Orig_chunk <- get_chunk(origins_chunks, i)
  ttm_chunk <- travel_time_matrix(r5r_core = r5_HAM,
                          origins = Orig_chunk,
                          destinations = Dest_all_schools_elem_2011,
                          mode = c("WALK"),
                          departure_datetime = as.POSIXct(strptime("2021-03-01 08:00:00", "%Y-%m-%d %H:%M:%S", tz = "EST5EDT")),
                          walk_speed = 3.6,
                          max_trip_duration = 60)

  # export output as disk.frame
  ifelse(i == 1, output_df <- as.disk.frame(ttm_chunk,
                                            nchunks = 1,
                                            outdir = "data-inputs/Travel-Times/Processing/output_ttm",
                                            compress = 50,
                                            overwrite = TRUE),
         add_chunk(output_df, ttm_chunk, chunk_id = i))
  setTxtProgressBar(pb, i)
}
end.time <- Sys.time()
print(paste0("OD matrix calculation took ", round(difftime(end.time, start.time, units = "mins"), digits = 2), " minutes..."))

output_OD_walk_elem4 <- as.data.frame(output_df)

save("output_OD_walk_elem4", file = "data-products/OD_tt_walk_2011_4.Rdata")
rm("output_OD_walk_elem4")
```

2016 walk travel times:
```{r}
# set up batching according to how many origin rows to process at one time
chunksize = 50 
num_chunks = ceiling(nrow(Orig_all_2016_1)/chunksize)

# create origin-destination pairs
origins_chunks <- as.disk.frame(Orig_all_2016_1,
                          outdir = "data-inputs/Travel-times/Processing/Orig",
                          nchunks = num_chunks,
                          overwrite = TRUE)

start.time <- Sys.time()
pb <- txtProgressBar(0, num_chunks, style = 3)

for (i in 1:num_chunks){
  Orig_chunk <- get_chunk(origins_chunks, i)
  ttm_chunk <- travel_time_matrix(r5r_core = r5_HAM,
                          origins = Orig_chunk,
                          destinations = Dest_all_schools_elem_2011,
                          mode = c("WALK"),
                          departure_datetime = as.POSIXct(strptime("2021-03-01 08:00:00", "%Y-%m-%d %H:%M:%S", tz = "EST5EDT")),
                          walk_speed = 3.6,
                          max_trip_duration = 60)

  # export output as disk.frame
  ifelse(i == 1, output_df <- as.disk.frame(ttm_chunk,
                                            nchunks = 1,
                                            outdir = "data-inputs/Travel-Times/Processing/output_ttm",
                                            compress = 50,
                                            overwrite = TRUE),
         add_chunk(output_df, ttm_chunk, chunk_id = i))
  setTxtProgressBar(pb, i)
}
end.time <- Sys.time()
print(paste0("OD matrix calculation took ", round(difftime(end.time, start.time, units = "mins"), digits = 2), " minutes..."))

output_OD_walk_elem1 <- as.data.frame(output_df)

save("output_OD_walk_elem1", file = "data-products/OD_tt_walk_2016_1.Rdata")
rm("output_OD_walk_elem1")
```

```{r}
# set up batching according to how many origin rows to process at one time
chunksize = 50 
num_chunks = ceiling(nrow(Orig_all_2016_2)/chunksize)

# create origin-destination pairs
origins_chunks <- as.disk.frame(Orig_all_2016_2,
                          outdir = "data-inputs/Travel-times/Processing/Orig",
                          nchunks = num_chunks,
                          overwrite = TRUE)

start.time <- Sys.time()
pb <- txtProgressBar(0, num_chunks, style = 3)

for (i in 1:num_chunks){
  Orig_chunk <- get_chunk(origins_chunks, i)
  ttm_chunk <- travel_time_matrix(r5r_core = r5_HAM,
                          origins = Orig_chunk,
                          destinations = Dest_all_schools_elem_2011,
                          mode = c("WALK"),
                          departure_datetime = as.POSIXct(strptime("2021-03-01 08:00:00", "%Y-%m-%d %H:%M:%S", tz = "EST5EDT")),
                          walk_speed = 3.6,
                          max_trip_duration = 60)

  # export output as disk.frame
  ifelse(i == 1, output_df <- as.disk.frame(ttm_chunk,
                                            nchunks = 1,
                                            outdir = "data-inputs/Travel-Times/Processing/output_ttm",
                                            compress = 50,
                                            overwrite = TRUE),
         add_chunk(output_df, ttm_chunk, chunk_id = i))
  setTxtProgressBar(pb, i)
}
end.time <- Sys.time()
print(paste0("OD matrix calculation took ", round(difftime(end.time, start.time, units = "mins"), digits = 2), " minutes..."))

output_OD_walk_elem2 <- as.data.frame(output_df)

save("output_OD_walk_elem2", file = "data-products/OD_tt_walk_2016_2.Rdata")
rm("output_OD_walk_elem2")
```

```{r}
# set up batching according to how many origin rows to process at one time
chunksize = 50 
num_chunks = ceiling(nrow(Orig_all_2016_3)/chunksize)

# create origin-destination pairs
origins_chunks <- as.disk.frame(Orig_all_2016_3,
                          outdir = "data-inputs/Travel-times/Processing/Orig",
                          nchunks = num_chunks,
                          overwrite = TRUE)

start.time <- Sys.time()
pb <- txtProgressBar(0, num_chunks, style = 3)

for (i in 1:num_chunks){
  Orig_chunk <- get_chunk(origins_chunks, i)
  ttm_chunk <- travel_time_matrix(r5r_core = r5_HAM,
                          origins = Orig_chunk,
                          destinations = Dest_all_schools_elem_2011,
                          mode = c("WALK"),
                          departure_datetime = as.POSIXct(strptime("2021-03-01 08:00:00", "%Y-%m-%d %H:%M:%S", tz = "EST5EDT")),
                          walk_speed = 3.6,
                          max_trip_duration = 60)

  # export output as disk.frame
  ifelse(i == 1, output_df <- as.disk.frame(ttm_chunk,
                                            nchunks = 1,
                                            outdir = "data-inputs/Travel-Times/Processing/output_ttm",
                                            compress = 50,
                                            overwrite = TRUE),
         add_chunk(output_df, ttm_chunk, chunk_id = i))
  setTxtProgressBar(pb, i)
}
end.time <- Sys.time()
print(paste0("OD matrix calculation took ", round(difftime(end.time, start.time, units = "mins"), digits = 2), " minutes..."))

output_OD_walk_elem3 <- as.data.frame(output_df)

save("output_OD_walk_elem3", file = "data-products/OD_tt_walk_2016_3.Rdata")
rm("output_OD_walk_elem3")
```

```{r}
# set up batching according to how many origin rows to process at one time
chunksize = 50 
num_chunks = ceiling(nrow(Orig_all_2016_4)/chunksize)

# create origin-destination pairs
origins_chunks <- as.disk.frame(Orig_all_2016_4,
                          outdir = "data-inputs/Travel-times/Processing/Orig",
                          nchunks = num_chunks,
                          overwrite = TRUE)

start.time <- Sys.time()
pb <- txtProgressBar(0, num_chunks, style = 3)

for (i in 1:num_chunks){
  Orig_chunk <- get_chunk(origins_chunks, i)
  ttm_chunk <- travel_time_matrix(r5r_core = r5_HAM,
                          origins = Orig_chunk,
                          destinations = Dest_all_schools_elem_2011,
                          mode = c("WALK"),
                          departure_datetime = as.POSIXct(strptime("2021-03-01 08:00:00", "%Y-%m-%d %H:%M:%S", tz = "EST5EDT")),
                          walk_speed = 3.6,
                          max_trip_duration = 60)

  # export output as disk.frame
  ifelse(i == 1, output_df <- as.disk.frame(ttm_chunk,
                                            nchunks = 1,
                                            outdir = "data-inputs/Travel-Times/Processing/output_ttm",
                                            compress = 50,
                                            overwrite = TRUE),
         add_chunk(output_df, ttm_chunk, chunk_id = i))
  setTxtProgressBar(pb, i)
}
end.time <- Sys.time()
print(paste0("OD matrix calculation took ", round(difftime(end.time, start.time, units = "mins"), digits = 2), " minutes..."))

output_OD_walk_elem4 <- as.data.frame(output_df)

save("output_OD_walk_elem4", file = "data-products/OD_tt_walk_2016_4.Rdata")
rm("output_OD_walk_elem4")
```

```{r}
# stop_r5(r5_HAM)
# rJava::.jgc(R.gc = TRUE)
```

### 2024-04-09 data-prep/03-travel-times-TTS.qmd
Draft of 'Manuscript.qmd' for special issue

--- 2024-04-09 data-prep/00-Estimating-Impe2school-from-TTS.Rmd
@@
-## Preliminaries
-
+---
+title: "03-TTS-travel-to-school-impedance"
+format: html
+---
 Load packages:
@@
 ```{r}
-ggh_taz <- st_read("data-inputs/TTS/tts06_83_region.shp")
-```
-
-Project object:
-```{r}
-ggh_taz <- ggh_taz %>%
-  st_transform(crs = 32617)
+ggh_taz <- st_read("data-inputs/TTS/tts06_83_region.shp") %>%   st_transform(crs = 32617)
 ```
@@
               drop_units())
-```
-
-Extract the IDs for all TAZ:
-```{r}
+#Extract the IDs for all TAZ:
 ghta_taz_id <- ggh_taz$GTA06
@@
 ```{r}
-ggh_pd <- st_read("data-inputs/TTS/tts06_pd_83.shp")
-```
-
-Project object:
-```{r}
-ggh_pd <- ggh_pd %>%
+ggh_pd <- st_read("data-inputs/TTS/tts06_pd_83.shp") %>%
   st_transform(crs = 32617)
@@
                       col_names = FALSE)
-# od_trips_2011 <- read_delim(file = "data-inputs/TTS/Schooltrip_persons_2011.txt", 
-#                       delim = "\t", 
-#                       col_names = FALSE)
-# od_trips_2016 <- read_delim(file = "data-inputs/TTS/Schooltrip_persons_2016.txt", 
-#                       delim = "\t", 
-#                       col_names = FALSE)
-
 ```
@@
 ```{r}
-# idx <- which(od_trips_2011$X1 == "TABLE    : stu_stat (Full time)")
 idx <- which(od_trips_2011_motor$X1 == "COLUMN : gta06_sch")
@@
 ```{r} 
-#load catchments
-load(file="data-products/ALL_CATCHS.RData")
-CATCH_2011_Public <- ALL_CATCHS%>%filter(Year == 2011 & Level == "Elementary" & System == "Public") %>% st_set_crs(st_crs(ham_taz))
-CATCH_2011_Cath <- ALL_CATCHS%>%filter(Year == 2011 & Level == "Elementary" & System == "Catholic")%>% st_set_crs(st_crs(ham_taz))
-CATCH_2016_Public <- ALL_CATCHS%>%filter(Year == 2016 & Level == "Elementary" & System == "Public")%>% st_set_crs(st_crs(ham_taz))
-CATCH_2016_Cath <- ALL_CATCHS%>%filter(Year == 2016 & Level == "Elementary" & System == "Catholic")%>% st_set_crs(st_crs(ham_taz))
+#load catchments - available through the hamONdests package (https://github.com/soukhova/hamONdests):
+library(hamONdests)
+data(School_Catchments_2011_2016)
+
+CATCH_2011_Public <- School_Catchments_2011_2016%>%filter(Year == 2011 & Level == "Elementary" & System == "Public") %>% st_set_crs(st_crs(ham_taz))
+CATCH_2011_Cath <- School_Catchments_2011_2016%>%filter(Year == 2011 & Level == "Elementary" & System == "Catholic")%>% st_set_crs(st_crs(ham_taz))
+CATCH_2016_Public <- School_Catchments_2011_2016%>%filter(Year == 2016 & Level == "Elementary" & System == "Public")%>% st_set_crs(st_crs(ham_taz))
+CATCH_2016_Cath <- School_Catchments_2011_2016%>%filter(Year == 2016 & Level == "Elementary" & System == "Catholic")%>% st_set_crs(st_crs(ham_taz))
 
@@
 ```{r}
-## THIS FUNCTION WAS CREATED USING CHATGDPT on DEC 6 2023
-
-# Function to check if values in a row of df1 are within any list in df2
+## THIS FUNCTION WAS CREATED USING CHATGDPT on DEC 6 2023 -- this is a function to check if values in a row of df1 are within any list in df2
 check_within_lists <- function(df1, df2) {
@@
 library(disk.frame)
-library(ggplot2)
-library(gridExtra)
-library(httr)
-library(knitr)
-#install.packages("kableExtra", dependencies = TRUE) #installing kableExtra and dependencies as I was receiving errors
-library(kableExtra)
-#install.packages("devtools")
-#install.packages("Rtools")
-#devtools::install_github("https://github.com/hrbrmstr/lodes.git", dependencies = TRUE) # if you need to download
-library(lodes)
-library(progress)
-library(purrr)
+library(tidyverse)
 library(r5r)
-#library(stplanr)
-library(tidyr)
+library(sf)
 library(tmap)
-library(zoo) # for rollapplyr
 
@@
 options(scipen = 999)
-options(java.parameters = "-Xmx512m")
-options(future.globals.maxSize = Inf)
+options(java.parameters = "-Xmx6G")
 ```
@@
 # the r5r package requires Java Development Kit version 11, which can be downloaded from https://www.oracle.com/java/technologies/javase-jdk11-downloads.html
-dir.create("data-inputs/TTS/r5_graph")
-r5_path <- file.path("data-inputs/TTS/r5_graph")
+dir.create("data-inputs/Travel-times")
+r5_path <- file.path("data-inputs/Travel-times")
 ```
@@
 r5_HAM <- setup_r5(data_path = r5_path, verbose = TRUE, overwrite = TRUE, temp_dir = TRUE)
-
 ```
@@
 origins_chunks <- as.disk.frame(ham_taz_coords,
-                          outdir = "data-inputs/df/Orig",
+                          outdir = "data-inputs/Travel-times/Processing/Orig",
                           nchunks = num_chunks,
@@
                                             nchunks = 1,
-                                            outdir = "data-inputs/df/output_ttm",
+                                            outdir = "data-inputs/Travel-Times/Processing/output_ttm",
                                             compress = 50,
@@
 output_OD_car <- as.data.frame(output_df)
-
-save("output_OD_car", file = "data-inputs/TTS/output_OD_car.Rdata")
+save("output_OD_car", file = "data-inputs/TTS/TTS_tt_car.Rdata")
 #I only made the cut off 60 minutes, so if some parcels do not have a travel time, rerun this for a 120 min cut off. The following are 120mins cutoff
@@
 output_OD_walk <- as.data.frame(output_df)
-
-save("output_OD_walk", file = "data-inputs/TTS/output_OD_walk.Rdata")
+save("output_OD_walk", file = "data-inputs/TTS/TTS_tt_walk.Rdata")
 ```
@@
 ```{r}
-load(file = "data-inputs/TTS/output_OD_car.Rdata")
-summary(output_OD_car)
+load(file = "data-inputs/TTS/TTS_tt_car.Rdata")
+TTS_tt_car <- output_OD_car
+rm("output_OD_car")
+summary(TTS_tt_car)
 #we'll set 0 travel times to equal 7 min for inter-zonal travel, half of the 1st Qu. travel time.
 
-load(file = "data-inputs/TTS/output_OD_walk.Rdata")
-summary(output_OD_walk) #max trip length calculated is 27 mins -- this corresponds to 'within catchment'. lets make 0 min trips into half of 1st Q. so 4min.
+load(file = "data-inputs/TTS/TTS_tt_walk.Rdata")
+TTS_tt_walk <- output_OD_walk
+rm("output_OD_walk")
+summary(TTS_tt_walk) #max trip length calculated is 27 mins -- this corresponds to 'within catchment'. lets make 0 min trips into half of 1st Q. so 4min.
 ```
@@
 od_trips_2011_motor <- od_trips_2011_motor %>% 
-  merge(output_OD_car, by.x = c("Origin", "Destination"), by.y = c("from_id", "to_id")) %>%
+  merge(TTS_tt_car, by.x = c("Origin", "Destination"), by.y = c("from_id", "to_id")) %>%
   merge(ham_taz %>% st_drop_geometry(), by.x="Origin", by.y="GTA06")
@@
 od_trips_2016_motor <- od_trips_2016_motor %>% 
-  merge(output_OD_car, by.x = c("Origin", "Destination"), by.y = c("from_id", "to_id")) %>%
+  merge(TTS_tt_car, by.x = c("Origin", "Destination"), by.y = c("from_id", "to_id")) %>%
   merge(ham_taz %>% st_drop_geometry(), by.x="Origin", by.y="GTA06")
@@
 od_trips_2011_nonmotor <- od_trips_2011_nonmotor %>%
-  merge(output_OD_walk, by.x = c("Origin", "Destination"), by.y = c("from_id", "to_id")) %>%
+  merge(TTS_tt_walk, by.x = c("Origin", "Destination"), by.y = c("from_id", "to_id")) %>%
   merge(ham_taz %>% st_drop_geometry(), by.x="Origin", by.y="GTA06")
@@
 od_trips_2016_nonmotor <- od_trips_2016_nonmotor %>% 
-  merge(output_OD_walk, by.x = c("Origin", "Destination"), by.y = c("from_id", "to_id"))  %>%
+  merge(TTS_tt_walk, by.x = c("Origin", "Destination"), by.y = c("from_id", "to_id"))  %>%
   merge(ham_taz %>% st_drop_geometry(), by.x="Origin", by.y="GTA06")
@@
 
-tld_2011_nonmotor <- od_trips_2011_nonmotor %>%
+tld_2011_nonmotor <- od_trips_2011_nonmotor |> filter(travel_time <= 27) |>
   mutate(tt_classes = cut(travel_time, 
@@
 
-tld_2016_nonmotor <- od_trips_2016_nonmotor %>%
+tld_2016_nonmotor <- od_trips_2016_nonmotor |> filter(travel_time <= 27) |>
   mutate(tt_classes = cut(travel_time, 
@@
 
-all_tt_2011_nonmotor <- od_trips_2011_nonmotor  %>% 
+all_tt_2011_nonmotor <- od_trips_2011_nonmotor |> filter(travel_time <= 27) |>
   dplyr::select(Trips, travel_time)
@@
 
-all_tt_2016_nonmotor <- od_trips_2016_nonmotor  %>% 
+all_tt_2016_nonmotor <- od_trips_2016_nonmotor  |> filter(travel_time <= 27) |>
   dplyr::select(Trips, travel_time)
@@
 ```{r}
-save(gamma2011_motor, file="data-inputs/TTS/gamma2011_motor.Rdata")
-save(gamma2016_motor, file="data-inputs/TTS/gamma2016_motor.Rdata")
-save(exp2011_nonmotor, file="data-inputs/TTS/exp2011_nonmotor.Rdata")
-save(exp2016_nonmotor, file="data-inputs/TTS/exp2016_nonmotor.Rdata")
+save(gamma2011_motor, file="data-products/TTS_gamma2011_motor.Rdata")
+save(gamma2016_motor, file="data-products/TTS_gamma2016_motor.Rdata")
+save(exp2011_nonmotor, file="data-products/TTS_exp2011_nonmotor.Rdata")
+save(exp2016_nonmotor, file="data-products/TTS_exp2016_nonmotor.Rdata")
 ```

### 2024-04-09 data-prep/04-Creating-avail_2011-16.Rmd
Draft of 'Manuscript.qmd' for special issue

---
title: "Testing-doc-2016run"
output: html_document
date: "2024-02-16"
---
# 2011 calcs

```{r load-packages, cache=FALSE}
library(dplyr)
library(sf)
library(ggplot2)
library(tidyverse)
library(units)
library(patchwork)
library(ggspatial)
options(scipen = 999)
```

```{r sp_avail-function}
#defining the spatial availability function, this prints all the intermediates
sp_avail_detailed <- function(x, o_id, d_id, pop, opp, pop_type, r, f, alpha = 1){

  o_id <- rlang::enquo(o_id)
  d_id <- rlang::enquo(d_id)
  pop <- rlang::enquo(pop)
  opp <- rlang::enquo(opp)
  pop_type <- rlang::enquo(pop_type)
  r <- rlang::enquo(r)
  f <- rlang::enquo(f)

  sum_pop <- x %>%
    dplyr::distinct(!!o_id, !!pop_type,
                    .keep_all = TRUE) %>%
    dplyr::mutate(sum_pop = !!r*(!!pop)^alpha) %>%
    dplyr::pull(sum_pop) %>%
    sum()

  f_p <- dplyr::pull(x, !!r) * dplyr::pull(x, !!pop)^alpha / sum_pop

  sum_impedance <- x %>%
    dplyr::group_by(!!d_id) %>%
    dplyr::summarize(sum_impedance = sum(!!f))

  x <- x %>%
    dplyr::left_join(sum_impedance,
                     by = rlang::as_name(d_id))

  f_c <- dplyr::pull(x, !!f) / x$sum_impedance

  x$f_c <- f_c
  x$f_p <- f_p

  sum_pa <- x %>%
    dplyr::group_by(!!d_id) %>%
    dplyr::summarize(sum_pa= sum(f_p * f_c))

  x <- x %>%
    dplyr::left_join(sum_pa,
                     by = rlang::as_name(d_id))
  f_t <- (f_p * f_c) / dplyr::pull(x, sum_pa)

  x %>%
    dplyr::mutate(V_ij = !!opp * !!f_t)
}
```

```{r}
load(file="data-products/cen_data_2011_hamiltononly.Rdata")
cen_data_2011_hamiltononly <- cen_data_2011_1 |> filter(Region.Name == "Hamilton") #NOTE: GEOUID 35250242, 35250133, 35250457 do not have any child population
rm("cen_data_2011_1")
load(file="data-products/cen_data_2016_hamiltononly.Rdata")
cen_data_2016_hamiltononly <- cen_data_2016_1 |> filter(Region.Name == "Hamilton") #NOTE: GEOUID 35250133, 35250398, 35250427 do not have any child population
rm("cen_data_2016_1")

ham_city_bound <- st_read("data-inputs/Boundaries/City_Boundary.shp") |> st_transform(st_crs(cen_data_2011_hamiltononly))
ham_community_bounds <- st_read("data-inputs/Boundaries/Community_Boundaries.shp") |> st_transform(st_crs(cen_data_2011_hamiltononly))
ggh_taz <- st_read("data-inputs/TTS/tts06_83_region.shp") |> st_transform(st_crs(cen_data_2011_hamiltononly))
```

```{r}
cen_data_2011_hamiltononly <- 
  cen_data_2011_hamiltononly %>% dplyr::select( #removing the variables we don't need for this analysis
  -c("Shape.Area", "Quality.Flags", "Type", "CD_UID", "NHS.Non.Return.Rate", "CSD_UID", "CT_UID", "CT_UID","NHS.Non.Return.Rate.1", "v_CA11N_2564..Median.after.tax.household.income..", "v_CA11F_17..15.to.19.years")) %>%
  rename( #renaming the variables
    "DA_households" = "Households",
    "DA_dwellings" = "Dwellings",
    "DA_tot_pop" = "Population",
    "DA_med_AT_family_income" = "v_CA11N_2458..Median.after.tax.family.income..",
    "DA_med_AT_lone-family_income" = "v_CA11N_2476..Median.after.tax.family.income..",
    "DA_prev_LIMAT_under18" = "v_CA11N_2609..Less.than.18.years..",
    "DA_pop_5to9" = "v_CA11F_11..5.to.9.years",
    "DA_pop_10to14" = "v_CA11F_14..10.to.14.years") %>%
  mutate(DA_pop_5to14 = DA_pop_5to9+(DA_pop_10to14)) |> #*0.9)) %>%
  dplyr::select(-c("DA_pop_5to9","DA_pop_10to14"))

cen_data_2016_hamiltononly <- 
  cen_data_2016_hamiltononly %>% dplyr::select( #removing the variables we don't need for this analysis
  -c("Shape.Area",  "Type", "CD_UID", "CSD_UID", "CT_UID", "CT_UID", "v_CA16_2398..Median.after.tax.income.of.households.in.2015....", "v_CA16_64..15.to.19.years")) %>%
  rename( #renaming the variables
    "DA_households" = "Households",
    "DA_dwellings" = "Dwellings",
    "DA_tot_pop" = "Population",
    "DA_med_AT_family_income" = "v_CA16_2448..Median.after.tax.income.of.economic.families.in.2015....",
         "DA_med_AT_lone-family_income" = "v_CA16_2460..Median.after.tax.income.of.lone.parent.economic.families.in.2015....",
         "DA_prev_LIMAT_under18" = "v_CA16_2543..0.to.17.years....",
         "DA_pop_5to9" = "v_CA16_25..5.to.9.years",
         "DA_pop_10to14" = "v_CA16_43..10.to.14.years") %>% mutate(DA_pop_5to14 = DA_pop_5to9+(DA_pop_10to14)) |> #*0.9)) %>%
         dplyr::select(-c("DA_pop_5to9","DA_pop_10to14"))
```

```{r}
library(hamONdests)
data(Schools_2011_2016)
ALL_SCHOOLS_Elem <- Schools_2011_2016 %>% filter(Level == "Elementary")
rm("Schools_2011_2016")
#data(School_Catchments_2011_2016)
load(file="data-products/TTS_gamma2011_motor.Rdata")
load(file="data-products/TTS_gamma2016_motor.Rdata")
load(file="data-products/TTS_exp2011_nonmotor.Rdata")
load(file="data-products/TTS_exp2016_nonmotor.Rdata")
```

```{r}
#add GTA06 IDs to schools - "GTA06_Dest"}
cols_TAZIDs <- sp::over(as_Spatial(ALL_SCHOOLS_Elem |> st_transform(st_crs(ggh_taz))),
                              as_Spatial(ggh_taz |> dplyr::select(GTA06) |> rename("GTA06_Dest" = "GTA06")))

ALL_SCHOOLS_Elem <- cbind(ALL_SCHOOLS_Elem,cols_TAZIDs)
rm(cols_TAZIDs)
```

```{r}
load(file="data-products/OD_tt_car_2011_1.Rdata")
load(file="data-products/OD_tt_car_2011_2.Rdata")
load(file="data-products/OD_tt_car_2011_3.Rdata")
load(file="data-products/OD_tt_car_2011_4.Rdata")

load(file="data-products/OD_tt_walk_2011_1.Rdata")
load(file="data-products/OD_tt_walk_2011_2.Rdata")
load(file="data-products/OD_tt_walk_2011_3.Rdata")
load(file="data-products/OD_tt_walk_2011_4.Rdata")

load(file="data-inputs/TTS/od_trips_perc_2011.Rdata")

#renaming travel time matrices 
ttm_car_2011 <- bind_rows(output_OD_car1, output_OD_car2, output_OD_car3, output_OD_car4) %>% mutate(mode = "c")
rm(output_OD_car1, output_OD_car2, output_OD_car3, output_OD_car4)

ttm_walk_2011 <- bind_rows(output_OD_walk_elem1,output_OD_walk_elem2,output_OD_walk_elem3,output_OD_walk_elem4) %>% mutate(mode = "w")
rm(output_OD_walk_elem1,output_OD_walk_elem2,output_OD_walk_elem3,output_OD_walk_elem4)
```

```{r}
load(file="data-inputs/Parcels/RED_LU_centroids_2011.RData") #parcels

#drop all parcel points that do not have a GeoUID (i.e., a DA identifier -- this means they fall outside of Hamilton). It's 10 for 2011 and 11 for 2016
RED_LU_centroids_2011 <- RED_LU_centroids_2011 %>% drop_na(GeoUID) %>%
  group_by(ID) %>%
  mutate(res_units = sum(SINGLE_FAM,SEMI_DET,ROW_THOUSE,APARTMENT)) %>%
  dplyr::select(-c("SINGLE_FAM","SEMI_DET","ROW_THOUSE","APARTMENT",
                   "TERANET", "LOCATION", "LUC1", "LUC2", "TYPE", "ROLL_NUM"))
```

```{r}
#add census data to the parcel centroids:
RED_LU_centroids_2011 <- RED_LU_centroids_2011 %>%
  merge(cen_data_2011_hamiltononly %>% st_drop_geometry(), by="GeoUID") 
```

```{r}
#create OD matrix -- each parcel to each school, and then drop all OD pairs that don't have a travel time ()
OD_2011 <- RED_LU_centroids_2011 %>% 
  cross_join(ALL_SCHOOLS_Elem %>% st_drop_geometry() |> filter(Year != 2016) %>% dplyr::select(c("SchoolID", "OTGC2011", "GTA06_Dest")))

#add travel time, per mode
test <- OD_2011 %>% mutate(ID = as.character(ID),
                           SchoolID = as.character(SchoolID)) %>%
  left_join(ttm_car_2011, by=c("ID" = "from_id",
                               "SchoolID" = "to_id")) %>% 
  drop_na(travel_time_p50)

OD_2011 <- OD_2011 %>% mutate(ID = as.character(ID),
                           SchoolID = as.character(SchoolID)) %>%
  left_join(ttm_walk_2011 |> filter(travel_time_p50 <= 27), by=c("ID" = "from_id",
                                "SchoolID" = "to_id")) %>% 
  drop_na(travel_time_p50)

OD_2011 <- rbind(test,OD_2011) 
rm(test, ttm_car_2011, ttm_walk_2011)
```
```{r}
#reduce dimensions
OD_2011 <- OD_2011 %>% dplyr::select(-c("LUC1_DESC", "LUC2_DESC", "res_units", "DA_dwellings", "CMA_UID", "Region.Name", "geometry","PROP_AREA", "Area..sq.km."))

# Check the mean and median travel times by mode, the first is at the level of zn1259 and the second is at zn208
OD_2011 %>% 
  group_by(mode) %>% 
  summarize(mean_tt = mean(travel_time_p50),
            median_tt = median(travel_time_p50),
            min_tt = min(travel_time_p50),
            max_tt = max(travel_time_p50))
```

```{r}
#now let's add the communtiy names to the TAZ and summarize the % of students who use what mode by community. 
ggh_taz_communities <- st_intersection(st_centroid(ham_taz) |> st_transform(st_crs(ggh_taz)), 
                                       ham_community_bounds |> st_transform(st_crs(ggh_taz))) |>
  dplyr::select(c("GTA06", "REGION", "COMMUNITY_"))

od_trips_perc_2011 <- od_trips_perc_2011 %>% left_join(ggh_taz_communities |> st_drop_geometry() |> mutate(GTA06 = as.character(GTA06)), by=c("Origin"="GTA06"))

od_trips_perc_comm_2011 <- od_trips_perc_2011 %>% group_by(COMMUNITY_) %>%
  summarize(TAZOrig_car_trips = sum(TAZOrig_car_trips),
            TAZOrig_walk_trips = sum(TAZOrig_walk_trips)) %>% 
  mutate(TAZOrig_car_perc = TAZOrig_car_trips/(TAZOrig_car_trips + TAZOrig_walk_trips),
         COMMUNITY = case_when(COMMUNITY_ == "Ancaster" ~ "AN",
                                                         COMMUNITY_ == "Dundas" ~ "DU",
                                                         COMMUNITY_ == "Flamborough" ~ "FL",
                                                         COMMUNITY_ == "Glanbrook" ~ "GL",
                                                         COMMUNITY_ == "Hamilton" ~ "HA",
                                                         COMMUNITY_ == "Stoney Creek" ~ "SC")) |>
  dplyr::select(-c("COMMUNITY_"))

od_trips_perc_comm_2011
```

```{r}
# Lets add the % of car-mode used to the full OD list and then start transforming the frame for Spatial Availability calc.
OD_2011 <- OD_2011 |> left_join(od_trips_perc_comm_2011 |> dplyr::select(c("TAZOrig_car_perc", "COMMUNITY")), 
                     by = "COMMUNITY") %>%
  mutate(
  DA_pop_5to14 = ifelse(is.na(DA_pop_5to14), 0, DA_pop_5to14),
  DA_households = ifelse(is.na(DA_households), 0, DA_households))

#'mode_count' identifies if a OD pair (ID to SchoolID) is either just car (1) or car and walk (2). Join this to the full OD list.
OD_2011_mode_count <- OD_2011 |>
  group_by(ID, SchoolID) %>% 
  summarise(mode_count = n())

OD_2011 <- OD_2011 %>% left_join(OD_2011_mode_count, by = c("ID", "SchoolID"))
rm(OD_2011_mode_count)
```

```{r}
#now, here we make the assumption on how many students live in each parcel. So since we assume that any parcel could potentially go to any school - I think it makes sense to also assume that the most granular level of population data we have (DA-level) is equally assigned to each parcel and then equally assigned to each OD flow based on it's mode-split (on average ~80/20 car/walk).

#Count the number of ID's for each OD pair -- this can be seen as the number of 'trips' being made from a parcel. There are 147 elementary schools in 2011 -- all parcels can reach all schools by car in 60 mins. So if a parcel only has 1 mode_count (car) it will have a "DA_ID_count" of 147 (the number of OD home-to-all-school trips). This is the number that the rate of Students in DA per OD flows in the DA ("IDs_in_a_DA_count") is multipled by.
DA_ID_count <- OD_2011 |> group_by(ID) |> summarise(DA_ID_count = n()) 
OD_2011 <- OD_2011 |> left_join(DA_ID_count, by="ID")

#Here we calculate the number of OD trips per O (ID) that occur in each DA. This is how much the # of students in a DA is divded by for each OD. 
IDs_in_a_DA_count <- OD_2011 |> group_by(GeoUID) |> summarise(IDs_in_a_DA_count = sum(DA_ID_count)) 
OD_2011 <- OD_2011 |> left_join(IDs_in_a_DA_count, by="GeoUID")

#We calculate the _kid-pop_ per parcel and then we adjusted it based on if the OD flow is 'walking' or 'car'
OD_2011 <- OD_2011 |> mutate(kid_pop = DA_pop_5to14/IDs_in_a_DA_count*DA_ID_count, 
               kid_pop = ifelse(is.na(kid_pop) | is.nan(kid_pop) | kid_pop == Inf, 0, kid_pop),
               kid_popadj = case_when(mode == "w" & mode_count == 2 ~ (1.0000000-TAZOrig_car_perc)*2*kid_pop,
                                         mode == "c" & mode_count == 2 ~ (TAZOrig_car_perc)*2*kid_pop,
                                         mode == "c" & mode_count == 1 ~ kid_pop),
                  kid_popadj = ifelse(is.na(kid_popadj) | is.nan(kid_popadj) | kid_popadj == Inf, 0, kid_popadj))
```

```{r}
#checks!, these should be the total number of kids (^the code above) -- 58265.
OD_2011$kid_pop %>% sum()
OD_2011$kid_popadj %>% sum()
OD_2011 |> group_by(GeoUID) |> summarize(DA_pop_5to14 = mean(DA_pop_5to14)) |> dplyr::select(DA_pop_5to14) |> sum()
```

```{r}
#now add the impedance function, etc to the full OD 
OD_2011 <- OD_2011 |> mutate(
  travel_time_p50 = ifelse(travel_time_p50 == 0, 0.5, travel_time_p50),
  r = 1,
  f = ifelse(mode == "c", stats::dgamma(
    travel_time_p50, gamma2011_motor$estimate["shape"], gamma2011_motor$estimate["rate"]), 
    stats::dexp(
      travel_time_p50, exp2011_nonmotor$estimate["rate"])))
```

```{r}
#calculate spatial availability. 
avail_2011 <- sp_avail_detailed(OD_2011, ID, SchoolID, kid_popadj, OTGC2011, mode, r, f, alpha = 1)
save(avail_2011, file="data-products/avail_2011.RData")
save(OD_2011, file="data-products/OD_2011.RData")
rm(OD_2011)
```

# 2016 calcs
```{r}
rm(list=ls())
```

```{r load-packages, include=FALSE, cache=FALSE}
library(dplyr)
library(sf)
library(ggplot2)
library(tidyverse)
library(units)
library(patchwork)
library(ggspatial)
options(scipen = 999)
```

```{r sp_avail-function,include=FALSE}
#defining the spatial availability function, this prints all the intermediates
sp_avail_detailed <- function(x, o_id, d_id, pop, opp, pop_type, r, f, alpha = 1){

  o_id <- rlang::enquo(o_id)
  d_id <- rlang::enquo(d_id)
  pop <- rlang::enquo(pop)
  opp <- rlang::enquo(opp)
  pop_type <- rlang::enquo(pop_type)
  r <- rlang::enquo(r)
  f <- rlang::enquo(f)

  sum_pop <- x %>%
    dplyr::distinct(!!o_id, !!pop_type,
                    .keep_all = TRUE) %>%
    dplyr::mutate(sum_pop = !!r*(!!pop)^alpha) %>%
    dplyr::pull(sum_pop) %>%
    sum()

  f_p <- dplyr::pull(x, !!r) * dplyr::pull(x, !!pop)^alpha / sum_pop

  sum_impedance <- x %>%
    dplyr::group_by(!!d_id) %>%
    dplyr::summarize(sum_impedance = sum(!!f))

  x <- x %>%
    dplyr::left_join(sum_impedance,
                     by = rlang::as_name(d_id))

  f_c <- dplyr::pull(x, !!f) / x$sum_impedance

  x$f_c <- f_c
  x$f_p <- f_p

  sum_pa <- x %>%
    dplyr::group_by(!!d_id) %>%
    dplyr::summarize(sum_pa= sum(f_p * f_c))

  x <- x %>%
    dplyr::left_join(sum_pa,
                     by = rlang::as_name(d_id))
  f_t <- (f_p * f_c) / dplyr::pull(x, sum_pa)

  x %>%
    dplyr::mutate(V_ij = !!opp * !!f_t)
}
```

```{r}
load(file="data-products/cen_data_2016_hamiltononly.Rdata")
cen_data_2016_hamiltononly <- cen_data_2016_1 |> filter(Region.Name == "Hamilton") #NOTE: GEOUID 35250133, 35250398, 35250427 do not have any child population
rm("cen_data_2016_1")

ham_city_bound <- st_read("data-inputs/Boundaries/City_Boundary.shp") |> st_transform(st_crs(cen_data_2016_hamiltononly))
ham_community_bounds <- st_read("data-inputs/Boundaries/Community_Boundaries.shp") |> st_transform(st_crs(cen_data_2016_hamiltononly))
ggh_taz <- st_read("data-inputs/TTS/tts06_83_region.shp") 
# |> st_transform(st_crs(cen_data_2011_hamiltononly))
```

```{r}
cen_data_2016_hamiltononly <- 
  cen_data_2016_hamiltononly %>% dplyr::select( #removing the variables we don't need for this analysis
  -c("Shape.Area",  "Type", "CD_UID", "CSD_UID", "CT_UID", "CT_UID", "v_CA16_2398..Median.after.tax.income.of.households.in.2015....", "v_CA16_64..15.to.19.years")) %>%
  rename( #renaming the variables
    "DA_households" = "Households",
    "DA_dwellings" = "Dwellings",
    "DA_tot_pop" = "Population",
    "DA_med_AT_family_income" = "v_CA16_2448..Median.after.tax.income.of.economic.families.in.2015....",
         "DA_med_AT_lone-family_income" = "v_CA16_2460..Median.after.tax.income.of.lone.parent.economic.families.in.2015....",
         "DA_prev_LIMAT_under18" = "v_CA16_2543..0.to.17.years....",
         "DA_pop_5to9" = "v_CA16_25..5.to.9.years",
         "DA_pop_10to14" = "v_CA16_43..10.to.14.years") %>% mutate(DA_pop_5to14 = DA_pop_5to9+(DA_pop_10to14)) |> #*0.9)) %>%
         dplyr::select(-c("DA_pop_5to9","DA_pop_10to14"))
```

```{r, include=FALSE}
library(hamONdests)
data(Schools_2011_2016)
ALL_SCHOOLS_Elem <- Schools_2011_2016 %>% filter(Level == "Elementary")
rm("Schools_2011_2016")
#data(School_Catchments_2011_2016)
load(file="data-products/TTS_gamma2016_motor.Rdata")
load(file="data-products/TTS_exp2016_nonmotor.Rdata")
```

```{r, include=FALSE}
#add GTA06 IDs to schools - "GTA06_Dest"}
cols_TAZIDs <- sp::over(as_Spatial(ALL_SCHOOLS_Elem |> st_transform(st_crs(ggh_taz))),
                              as_Spatial(ggh_taz |> dplyr::select(GTA06) |> rename("GTA06_Dest" = "GTA06")))
ALL_SCHOOLS_Elem <- cbind(ALL_SCHOOLS_Elem,cols_TAZIDs)
rm(cols_TAZIDs)
```

```{r}
load(file="data-products/OD_tt_car_2016_1.Rdata")
load(file="data-products/OD_tt_car_2016_2.Rdata")
load(file="data-products/OD_tt_car_2016_3.Rdata")
load(file="data-products/OD_tt_car_2016_4.Rdata")

load(file="data-products/OD_tt_walk_2016_1.Rdata")
load(file="data-products/OD_tt_walk_2016_2.Rdata")
load(file="data-products/OD_tt_walk_2016_3.Rdata")
load(file="data-products/OD_tt_walk_2016_4.Rdata")

load(file="data-inputs/TTS/od_trips_perc_2016.Rdata")

#renaming travel time matrices 
ttm_car_2016 <- bind_rows(output_OD_car1, output_OD_car2, output_OD_car3, output_OD_car4) %>% mutate(mode = "c")
rm(output_OD_car1, output_OD_car2, output_OD_car3, output_OD_car4)

ttm_walk_2016 <- bind_rows(output_OD_walk_elem1,output_OD_walk_elem2,output_OD_walk_elem3,output_OD_walk_elem4) %>% mutate(mode = "w")
rm(output_OD_walk_elem1,output_OD_walk_elem2,output_OD_walk_elem3,output_OD_walk_elem4)
```

```{r}
load(file="data-inputs/Parcels/RED_LU_centroids_2016.RData") #parcels

#drop all parcel points that do not have a GeoUID (i.e., a DA identifier -- this means they fall outside of Hamilton). It's 10 for 2011 and 11 for 2016
RED_LU_centroids_2016 <- RED_LU_centroids_2016 %>% drop_na(GeoUID) %>%
  group_by(ID) %>%
  mutate(res_units = sum(SINGLE_FAM,SEMI_DET,ROW_THOUSE,APARTMENT)) %>%
  dplyr::select(-c("SINGLE_FAM","SEMI_DET","ROW_THOUSE","APARTMENT",
                   "TERANET", "LOCATION", "LUC1", "LUC2", "TYPE", "ROLL_NUM"))
```

```{r}
#add census data to the parcel centroids:
RED_LU_centroids_2016 <- RED_LU_centroids_2016 %>%
  merge(cen_data_2016_hamiltononly %>% st_drop_geometry(), by="GeoUID") 
```

```{r}
#create OD matrix -- each parcel to each school, and then drop all OD pairs that don't have a travel time ()
OD_2016 <- RED_LU_centroids_2016 %>% 
  cross_join(ALL_SCHOOLS_Elem %>% st_drop_geometry() |> filter(Year != 2011) %>% dplyr::select(c("SchoolID", "OTGC2016", "GTA06_Dest")))

#add travel time, per mode
test <- OD_2016 %>% mutate(ID = as.character(ID),
                           SchoolID = as.character(SchoolID)) %>%
  left_join(ttm_car_2016, by=c("ID" = "from_id",
                               "SchoolID" = "to_id")) %>% 
  drop_na(travel_time_p50)

OD_2016 <- OD_2016 %>% mutate(ID = as.character(ID),
                           SchoolID = as.character(SchoolID)) %>%
  left_join(ttm_walk_2016 |> filter(travel_time_p50 <= 27), by=c("ID" = "from_id",
                                "SchoolID" = "to_id")) %>% 
  drop_na(travel_time_p50)

OD_2016 <- rbind(test,OD_2016) 
rm(test, ttm_car_2016, ttm_walk_2016)
```
```{r}
#reduce dimensions
OD_2016 <- OD_2016 %>% dplyr::select(-c("LUC1_DESC", "LUC2_DESC", "res_units", "DA_dwellings", "CMA_UID", "Region.Name", "geometry","PROP_AREA", "Area..sq.km."))

# Check the mean and median travel times by mode, the first is at the level of zn1259 and the second is at zn208
OD_2016 %>% 
  group_by(mode) %>% 
  summarize(mean_tt = mean(travel_time_p50),
            median_tt = median(travel_time_p50),
            min_tt = min(travel_time_p50),
            max_tt = max(travel_time_p50))
```

```{r}
#now let's add the communtiy names to the TAZ and summarize the % of students who use what mode by community. 
ggh_taz_communities <- st_intersection(st_centroid(ggh_taz) |> st_transform((st_crs(ggh_taz))), ham_community_bounds |> st_transform(st_crs(ggh_taz))) |> dplyr::select(c("GTA06", "REGION", "COMMUNITY_"))

od_trips_perc_2016 <- od_trips_perc_2016 %>% left_join(ggh_taz_communities |> st_drop_geometry() |> mutate(GTA06 = as.character(GTA06)), by=c("Origin"="GTA06"))

od_trips_perc_comm_2016 <- od_trips_perc_2016 %>% group_by(COMMUNITY_) %>%
  summarize(TAZOrig_car_trips = sum(TAZOrig_car_trips),
            TAZOrig_walk_trips = sum(TAZOrig_walk_trips)) %>% 
  mutate(TAZOrig_car_perc = TAZOrig_car_trips/(TAZOrig_car_trips + TAZOrig_walk_trips),
         COMMUNITY = case_when(COMMUNITY_ == "Ancaster" ~ "AN",
                                                         COMMUNITY_ == "Dundas" ~ "DU",
                                                         COMMUNITY_ == "Flamborough" ~ "FL",
                                                         COMMUNITY_ == "Glanbrook" ~ "GL",
                                                         COMMUNITY_ == "Hamilton" ~ "HA",
                                                         COMMUNITY_ == "Stoney Creek" ~ "SC")) |>
  dplyr::select(-c("COMMUNITY_"))

od_trips_perc_comm_2016
```

```{r}
# Lets add the % of car-mode used to the full OD list and then start transforming the frame for Spatial Availability calc.
OD_2016 <- OD_2016 |> left_join(od_trips_perc_comm_2016 |> dplyr::select(c("TAZOrig_car_perc", "COMMUNITY")), 
                     by = "COMMUNITY") %>%
  mutate(
  DA_pop_5to14 = ifelse(is.na(DA_pop_5to14), 0, DA_pop_5to14),
  DA_households = ifelse(is.na(DA_households), 0, DA_households))

#'mode_count' identifies if a OD pair (ID to SchoolID) is either just car (1) or car and walk (2). Join this to the full OD list.
OD_2016_mode_count <- OD_2016 |>
  group_by(ID, SchoolID) %>% 
  summarise(mode_count = n())

OD_2016 <- OD_2016 %>% left_join(OD_2016_mode_count, by = c("ID", "SchoolID"))
rm(OD_2016_mode_count)
```

```{r}
#now, here we make the assumption on how many students live in each parcel. So since we assume that any parcel could potentially go to any school - I think it makes sense to also assume that the most granular level of population data we have (DA-level) is equally assigned to each parcel and then equally assigned to each OD flow based on it's mode-split (on average ~80/20 car/walk).

#Count the number of ID's for each OD pair -- this can be seen as the number of 'trips' being made from a parcel. There are 147 elementary schools in 2011 -- all parcels can reach all schools by car in 60 mins. So if a parcel only has 1 mode_count (car) it will have a "DA_ID_count" of 147 (the number of OD home-to-all-school trips). This is the number that the rate of Students in DA per OD flows in the DA ("IDs_in_a_DA_count") is multipled by.
DA_ID_count <- OD_2016 |> group_by(ID) |> summarise(DA_ID_count = n()) 
OD_2016 <- OD_2016 |> left_join(DA_ID_count, by="ID")

#Here we calculate the number of OD trips per O (ID) that occur in each DA. This is how much the # of students in a DA is divded by for each OD. 
IDs_in_a_DA_count <- OD_2016 |> group_by(GeoUID) |> summarise(IDs_in_a_DA_count = sum(DA_ID_count)) 
OD_2016 <- OD_2016 |> left_join(IDs_in_a_DA_count, by="GeoUID")

#We calculate the _kid-pop_ per parcel and then we adjusted it based on if the OD flow is 'walking' or 'car'
OD_2016 <- OD_2016 |> mutate(kid_pop = DA_pop_5to14/IDs_in_a_DA_count*DA_ID_count, 
               kid_pop = ifelse(is.na(kid_pop) | is.nan(kid_pop) | kid_pop == Inf, 0, kid_pop),
               kid_popadj = case_when(mode == "w" & mode_count == 2 ~ (1.0000000-TAZOrig_car_perc)*2*kid_pop,
                                         mode == "c" & mode_count == 2 ~ (TAZOrig_car_perc)*2*kid_pop,
                                         mode == "c" & mode_count == 1 ~ kid_pop),
                  kid_popadj = ifelse(is.na(kid_popadj) | is.nan(kid_popadj) | kid_popadj == Inf, 0, kid_popadj))
```

```{r}
#checks!, these should be the total number of kids (^the code above) -- 58265.
OD_2016$kid_pop %>% sum()
OD_2016$kid_popadj %>% sum()
OD_2016 |> group_by(GeoUID) |> summarize(DA_pop_5to14 = mean(DA_pop_5to14)) |> dplyr::select(DA_pop_5to14) |> sum()
```

```{r}
#now add the impedance function, etc to the full OD 
OD_2016 <- OD_2016 |> mutate(
  travel_time_p50 = ifelse(travel_time_p50 == 0, 0.5, travel_time_p50),
  r = 1,
  f = ifelse(mode == "c", stats::dgamma(
    travel_time_p50, gamma2016_motor$estimate["shape"], gamma2016_motor$estimate["rate"]), 
    stats::dexp(
      travel_time_p50, exp2016_nonmotor$estimate["rate"])))
```

```{r}
#calculate spatial availability. 
avail_2016 <- sp_avail_detailed(OD_2016, ID, SchoolID, kid_popadj, OTGC2016, mode, r, f, alpha = 1)
save(avail_2016, file="data-products/avail_2016.RData")
save(OD_2016, file="data-products/OD_2016.RData")
rm(OD_2016)
```

### 2024-04-10 paper/Manuscript.qmd
Initialized renv and reorganized the repo: the paper, bibliography, and figures now have their own folders

--- 2024-04-09 Manuscript.qmd
@@
 title: "The cost of school consolidations: impacts of reduced accessibility"
-bibliography: mybibfile.bib
+bibliography: 
+  - "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
 linenumbers: true
@@
 ```{r, include=FALSE}
-library(dplyr)
-library(sf)
-library(ggplot2)
-library(tidyverse)
-library(units)
-library(patchwork)
-library(ggspatial)
-library(scales)
-library(stplanr)
-library(tmap)
+library(dplyr) # A Grammar of Data Manipulation
+library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
+library(ggspatial) # Spatial Data Framework for ggplot2
+library(here) # A Simpler Way to Find Your Files
+library(patchwork) # The Composer of Plots
+library(scales) # Scale Functions for Visualization
+library(sf) # Simple Features for R
+library(stplanr) # Sustainable Transport Planning
+library(tidyverse) # Easily Install and Load the 'Tidyverse'
+library(tmap) # Thematic Maps
+library(units) # Measurement Units for R Vectors
 ```
@@
 ```{r, include=FALSE}
-load(file="data-prep/data-products/cen_data_2011_hamiltononly.Rdata")
+load(file = paste0(here::here(),"/data-prep/data-products/cen_data_2011_hamiltononly.Rdata"))
 cen_data_2011_hamiltononly <- cen_data_2011_1 |> filter(Region.Name == "Hamilton") #NOTE: GEOUID 35250242, 35250133, 35250457 do not have any child population
 rm("cen_data_2011_1")
-load(file="data-prep/data-products/cen_data_2016_hamiltononly.Rdata")
+load(file=paste0(here::here(),"/data-prep/data-products/cen_data_2016_hamiltononly.Rdata"))
 cen_data_2016_hamiltononly <- cen_data_2016_1 |> filter(Region.Name == "Hamilton") #NOTE: GEOUID 35250133, 35250398, 35250427 do not have any child population
@@
 
-ham_city_bound <- st_read("data-prep/data-inputs/Boundaries/City_Boundary.shp") |> st_transform(st_crs(cen_data_2011_hamiltononly))
-ham_community_bounds <- st_read("data-prep/data-inputs/Boundaries/Community_Boundaries.shp") |> st_transform(st_crs(cen_data_2011_hamiltononly))
+ham_city_bound <- st_read(paste0(here::here(),"/data-prep/data-inputs/Boundaries/City_Boundary.shp")) |> st_transform(st_crs(cen_data_2011_hamiltononly))
+ham_community_bounds <- st_read(paste0(here::here(),"/data-prep/data-inputs/Boundaries/Community_Boundaries.shp")) |> st_transform(st_crs(cen_data_2011_hamiltononly))
 ```
@@
 ```{r, include=FALSE}
-library(hamONdests)
+library(hamONdests) # A R package of Hamilton, Ontario destinations
 data(Schools_2011_2016)
@@
 
-load(file="data-prep/data-products/TTS_gamma2011_motor.Rdata")
-load(file="data-prep/data-products/TTS_gamma2016_motor.Rdata")
-load(file="data-prep/data-products/TTS_exp2011_nonmotor.Rdata")
-load(file="data-prep/data-products/TTS_exp2016_nonmotor.Rdata")
+load(file = paste0(here::here(),"/data-prep/data-products/TTS_gamma2011_motor.Rdata"))
+load(file = paste0(here::here(),"/data-prep/data-products/TTS_gamma2016_motor.Rdata"))
+load(file = paste0(here::here(),"/data-prep/data-products/TTS_exp2011_nonmotor.Rdata"))
+load(file = paste0(here::here(),"/data-prep/data-products/TTS_exp2016_nonmotor.Rdata"))
 ```
@@
 
-tmap_save(catchment_schools_plot, "Figures/catchment_schools_plot.png", height=7, width=12, dpi=600)
+tmap_save(catchment_schools_plot, paste0(here::here(),"/Figures/catchment_schools_plot.png"), height=7, width=12, dpi=600)
 ```
@@
 ```{r,include=FALSE}
-load(file="data-prep/data-products/avail_2011.RData")
-load(file="data-prep/data-products/avail_2016.RData")
+load(file = paste0(here::here(),"/data-prep/data-products/avail_2011.RData"))
+load(file = paste0(here::here(),"/data-prep/data-products/avail_2016.RData"))
 ```
@@
 
-tmap_save(stud_pop_rate_LIM_plot, "Figures/stud_pop_rate_LIM_plot.png", height=7, width=12, dpi=600)
+tmap_save(stud_pop_rate_LIM_plot, paste0(here::here(),"/Figures/stud_pop_rate_LIM_plot.png"), height=7, width=12, dpi=600)
 ```
@@
 #Read the boundaries of the traffic analysis zones (retrieved from the TTS 2016):
-ggh_taz <- st_read("data-prep/data-inputs/TTS/tts06_83_region.shp") %>% st_transform(crs = 32617)
+ggh_taz <- st_read(paste0(here::here(), "/data-prep/data-inputs/TTS/tts06_83_region.shp")) %>% st_transform(crs = 32617)
 
@@
 #Read the boundaries of the traffic analysis zones (retrieved from the TTS 2016):
-ggh_pd <- st_read("data-prep/data-inputs/TTS/tts06_pd_83.shp") %>%
+ggh_pd <- st_read(paste0(here::here(), "/data-prep/data-inputs/TTS/tts06_pd_83.shp")) %>%
   st_transform(crs = 32617)
@@
 #the tts origin-destination trips, by mode
-load(file="data-prep/data-inputs/TTS/od_trips_perc_2011.Rdata")
-load(file="data-prep/data-inputs/TTS/od_trips_perc_2016.Rdata")
+load(file = paste0(here::here(),"/data-prep/data-inputs/TTS/od_trips_perc_2011.Rdata"))
+load(file = paste0(here::here(),"/data-prep/data-inputs/TTS/od_trips_perc_2016.Rdata"))
 
@@
 OD_flow_plot <- tmap_arrange(car_OD_flow_plot, walk_OD_flow, nrow = 2)
-tmap_save(OD_flow_plot, "Figures/OD_flow_plot.png", height=7, width=12, dpi=600)
+tmap_save(OD_flow_plot, paste0(here::here(),"/Figures/OD_flow_plot.png"), height=7, width=12, dpi=600)
 ```
@@
 TLD_empirical_wtheo_curve_plot
-ggsave("Figures/TLD_empirical_wtheo_curve_plot.png", height=7, width=12, dpi=600)
+ggsave(paste0(here::here(),"/Figures/TLD_empirical_wtheo_curve_plot.png"), height=7, width=12, dpi=600)
 ```
@@
 
-tmap_save(Vim_vvim_20112016_plot, "Figures/Vim_vvim_20112016_plot.png", height=7, width=12, dpi=600)
+tmap_save(Vim_vvim_20112016_plot, paste0(here::here(),"/Figures/Vim_vvim_20112016_plot.png"), height=7, width=12, dpi=600)
 ```
@@
 
-tmap_save(Vim_vvim_20112016_CHANGE_plot, "Figures/Vim_vvim_20112016_CHANGE_plot.png", height=7, width=12, dpi=600)
+tmap_save(Vim_vvim_20112016_CHANGE_plot, 
+          paste0(here::here(),"/Figures/Vim_vvim_20112016_CHANGE_plot.png"),
+          height=7,
+          width=12, 
+          dpi=600)
 ```
@@
 OD_flow_plot <- tmap_arrange(car_OD_flow_plot, walk_OD_flow, nrow = 2)
-tmap_save(OD_flow_plot, "Figures/SA-and-SAperStud-20112016.png", height=7, width=12, dpi=600)
+tmap_save(OD_flow_plot, 
+          paste0(here::here(),"/Figures/SA-and-SAperStud-20112016.png"),
+          height=7, 
+          width=12,
+          dpi=600)
 ```
@@
 
-knitr::include_graphics("Figures/catchment_schools_plot.png")
+knitr::include_graphics(paste0(here::here(),"/Figures/catchment_schools_plot.png"))
 ```
@@
 All scales are represented in quartiles.", out.width="100%", echo=FALSE}
-knitr::include_graphics("Figures/stud_pop_rate_LIM_plot.png")
+knitr::include_graphics(paste0(here::here(),"/Figures/stud_pop_rate_LIM_plot.png"))
 ```
@@
 ", out.width="100%", echo=FALSE}
-knitr::include_graphics("Figures/OD_flow_plot.png")
+knitr::include_graphics(paste0(here::here(),"/Figures/OD_flow_plot.png"))
 ```
@@
 ```{r imped-20112016-2, fig.cap="\\label{fig:Fig4}The empirical and theoretical motorized and non-motorized impedance functions. ", out.width="100%", echo=FALSE}
-knitr::include_graphics("Figures/TLD_empirical_wtheo_curve_plot.png")
+knitr::include_graphics(paste0(here::here(),"/Figures/TLD_empirical_wtheo_curve_plot.png"))
 ```
@@
 ```{r Vim-vim-4-plots, fig.cap="\\label{fig:Fig6}Multimodal spatial availability (greys) and spatial availability per mode-using student (diverging reds and greens) aggregated at the level of DA for 2011 and 2016.", out.width="100%", echo=FALSE}
-knitr::include_graphics("Figures/Vim_vvim_20112016_plot.png")
+knitr::include_graphics(paste0(here::here(),"/Figures/Vim_vvim_20112016_plot.png"))
 ```
@@
 ```{r change-Vim-vim-plot, fig.cap="\\label{fig:Fig7}The change in multimodal spatial availability and spatial availability per mode-using student between 2016 and 2011 at the level of DA of the 2016 Canadian Census.", out.width="100%", echo=FALSE}
-knitr::include_graphics("Figures/Vim_vvim_20112016_CHANGE_plot.png")
+knitr::include_graphics(paste0(here::here(),"/Figures/Vim_vvim_20112016_CHANGE_plot.png"))
 ```

### 2024-04-10 paper/Manuscript.qmd
Fixed cross-references before sharing with Chris and Moataz

@@
     email: "soukhoa@mcmaster.ca"
-    affiliation: McMaster University
+    affiliation: McMaster University SEES
     footnote: "Corresponding Author"
+  - name: "Christopher D. Higgins"
+    email: "cd.higgins@utoronto.ca"
+    affiliation: University of Toronto
+  - name: "Moataz Mohamed"
+    email: "mmohame@mcmaster.ca "
+    affiliation: McMaster University CE
   - name: "Antonio Paez"
     email: "paezha@mcmaster.ca"
-    affiliation: McMaster University
+    affiliation: McMaster University SEES
 address:
-  - code: McMaster University
-    address: "School of Earth, Environment and Society, McMaster University, Hamilton, ON, L8S 4K1, Canada"
+  - code: McMaster University SEES
+    address: "School of Earth, Environment and Society, McMaster University, 1241 Main St. West, Hamilton, ON, L8S 4K1, Canada"
+  - code: McMaster University CE
+    address: "Department of Civil Engineering, McMaster University, 1241 Main St. West, Hamilton, ON, L8S 4K1, Canada"
+  - code: University of Toronto
+    address: "Department of Human Geography, University of Toronto Scarborough, 1265 Military Trail, Toronto, ON, M1C 1A4, Canada"  
 abstract: | 
@@
 
-ham_city_bound <- st_read(paste0(here::here(),"/data-prep/data-inputs/Boundaries/City_Boundary.shp")) |> st_transform(st_crs(cen_data_2011_hamiltononly))
-ham_community_bounds <- st_read(paste0(here::here(),"/data-prep/data-inputs/Boundaries/Community_Boundaries.shp")) |> st_transform(st_crs(cen_data_2011_hamiltononly))
+ham_city_bound <- st_read(paste0(here::here(),"/data-prep/data-inputs/Boundaries/City_Boundary.shp"),
+                          quiet = TRUE) |> st_transform(st_crs(cen_data_2011_hamiltononly))
+ham_community_bounds <- st_read(paste0(here::here(),"/data-prep/data-inputs/Boundaries/Community_Boundaries.shp"),
+                                quiet = TRUE) |> st_transform(st_crs(cen_data_2011_hamiltononly))
 ```
@@
 
-```{r augmenting-catchment}
+```{r augmenting-catchment, include=FALSE}
 ALL_SCHOOLS_CATCH_Elem_extra2011 <- ALL_SCHOOLS_CATCH_Elem |>
@@
 
-```{r importing-taz-and-associated-flows}
+```{r importing-taz-and-associated-flows, include=FALSE}
 #Read the boundaries of the traffic analysis zones (retrieved from the TTS 2016):
-ggh_taz <- st_read(paste0(here::here(), "/data-prep/data-inputs/TTS/tts06_83_region.shp")) %>% st_transform(crs = 32617)
+ggh_taz <- st_read(paste0(here::here(), "/data-prep/data-inputs/TTS/tts06_83_region.shp"),
+                   quiet = TRUE) %>% 
+  st_transform(crs = 32617)
 
@@
 #Read the boundaries of the traffic analysis zones (retrieved from the TTS 2016):
-ggh_pd <- st_read(paste0(here::here(), "/data-prep/data-inputs/TTS/tts06_pd_83.shp")) %>%
+ggh_pd <- st_read(paste0(here::here(), "/data-prep/data-inputs/TTS/tts06_pd_83.shp"),
+                  quiet = TRUE) %>%
   st_transform(crs = 32617)
@@
 ```
+
 ```{r tts-theoretical-curve-creation}
@@
 
-```{r catchments-2011-2016, fig.cap="\\label{fig:Fig1}The school catchments, estimated on the ground capacities of schools summed per catchment, and school status for both public and public catholic school boards for year 2011 and 2016.
-OTGC is presented by quartiles.
-", out.width="100%", echo=FALSE}
+```{r}
+#| label: fig-Fig1
+#| fig-cap: "The school catchments, estimated on the ground capacities of schools summed per catchment, and school status for both public and public catholic school boards for year 2011 and 2016.
+OTGC is presented by quartiles."
+#| echo: false
 
@@
 
-```{r elem-res-rate-2016, fig.cap="\\label{fig:Fig2}The magnitude (top row) and rate per residential unit (middle row) of elementary aged student population per DA in 2011 and 2016.
+```{r}
+#| label: fig-Fig2
+#| fig-cap: "The magnitude (top row) and rate per residential unit (middle row) of elementary aged student population per DA in 2011 and 2016.
 The proportion of low-income households with children under 18 years old is also shown (bottom row).
-All scales are represented in quartiles.", out.width="100%", echo=FALSE}
+All scales are represented in quartiles."
+#| echo: false
+
 knitr::include_graphics(paste0(here::here(),"/Figures/stud_pop_rate_LIM_plot.png"))
@@
 
-```{r imped-20112016, fig.cap="\\label{fig:Fig3}The origin-destination flows from home-based motorized and non-motorized school trips for students between 5-13 years old as retrieved from TTS 2011 and 2016 mapped atop proportion of TAZ non/motorized modal share, DA boundaries, and community boundaries.
-", out.width="100%", echo=FALSE}
+```{r}
+#| label: fig-Fig3
+#| fig-cap: "The origin-destination flows from home-based motorized and non-motorized school trips for students between 5-13 years old as retrieved from TTS 2011 and 2016 mapped atop proportion of TAZ non/motorized modal share, DA boundaries, and community boundaries."
+#| echo: false
+
 knitr::include_graphics(paste0(here::here(),"/Figures/OD_flow_plot.png"))
@@
 
-```{r imped-20112016-2, fig.cap="\\label{fig:Fig4}The empirical and theoretical motorized and non-motorized impedance functions. ", out.width="100%", echo=FALSE}
+```{r}
+#| label: fig-Fig4
+#| fig-cap: "Empirical and theoretical motorized and non-motorized impedance functions."
+#| echo: false
+
 knitr::include_graphics(paste0(here::here(),"/Figures/TLD_empirical_wtheo_curve_plot.png"))
@@
 Accessibility is defined as "access to opportunities" or the "potential for spatial interaction".
-It is classically presented as the gravity-based measure popularized by @hansenHowAccessibilityShapes1959 and takes the following general multi-modal formulation (Equation (\ref{eq:multimodal-conventional-accessibility})):
+It is classically presented as the gravity-based measure popularized by @hansenHowAccessibilityShapes1959 and takes the following general multi-modal formulation (Equation (\ref{eq:multi-modal-conventional-accessibility})):
 ```{=tex}
@@
 
-```{r Vim-vim-4-plots, fig.cap="\\label{fig:Fig6}Multimodal spatial availability (greys) and spatial availability per mode-using student (diverging reds and greens) aggregated at the level of DA for 2011 and 2016.", out.width="100%", echo=FALSE}
+```{r}
+#| label: fig-Fig6
+#| fig-cap: "Multimodal spatial availability (greys) and spatial availability per mode-using student (diverging reds and greens) aggregated at the level of DA for 2011 and 2016."
+#| echo: false
+
 knitr::include_graphics(paste0(here::here(),"/Figures/Vim_vvim_20112016_plot.png"))
@@
 
-```{r change-Vim-vim-plot, fig.cap="\\label{fig:Fig7}The change in multimodal spatial availability and spatial availability per mode-using student between 2016 and 2011 at the level of DA of the 2016 Canadian Census.", out.width="100%", echo=FALSE}
+```{r}
+#| label: fig-Fig7
+#| fig-cap: "Change in multimodal spatial availability and spatial availability per mode-using student between 2016 and 2011 at the level of DA of the 2016 Canadian Census."
+#| echo: false
+
 knitr::include_graphics(paste0(here::here(),"/Figures/Vim_vvim_20112016_CHANGE_plot.png"))
@@
 
-```{r}
+```{r include=FALSE}
 SAvail_im_COMM_changes |> st_drop_geometry() |> 

### 2024-04-10 paper/Manuscript.qmd
The paper still needs the appendix. I cached the code for speed.

@@
   echo = FALSE,
-  cache = FALSE,
+  cache = TRUE,
   warning = FALSE,
@@
 
+# Appendix
+
+The regression model for the OTGC for schools goes here.
+
 # References

### 2024-04-10 paper/Manuscript.qmd
Removed the cache which was creating problems, and used freeze: auto instead

--- 2024-04-10 paper/Manuscript.qmd
@@
 editor: source
+execute: 
+  freeze: auto
 author:

### 2024-04-18 Manuscript.qmd
added manuscript in docx version

@@
 numbersections: true
-format: pdf
+format: docx
 editor: source

### 2024-04-22 paper/Manuscript.qmd
cdh edits to first half of paper

@@
 Spatial availability allows for an interpretable comparison between ‘before’ and ‘after’ policy implementation because of its unique proportional allocation property.
-Analysis is conducted at the parcel-level, all school-seats being eligible and travel behaviour based on observved home-to-school patterns for motorzied and non-motorized modes.
-Analysis on system-wide impacts such as potential GHG emission increases and walk minutes decreases are detailed.
+The analysis is conducted at the parcel-level with all school seats being eligible and travel behaviour based on observed home-to-school patterns for motorzied and non-motorized modes.
+System-wide impacts resulting from the closures, such as potential GHG emission increases and walk minutes decreases, are detailed.
 Overall, spatial availability decreased in the city.
 Though the school conslidation policy resulted in short-term operational cost-savings, families, students, and the environment now pay the price.
-This paper showcases how spatial availability can be used by decision- makers to quantify spatial and travel implications in policy.
+This paper showcases how spatial availability can be used by decision- makers to quantify spatial and travel implications associated with policy changes.
 keywords: [accessibility; spatial availability; carbon emissions; journey-to-school; walkability; active transportation; school consolidation policy; service provision thresholds]
@@
 These forces spur the re-organization and consolidation of public facilities in communities around the world where changes to service provision may improve the quality of service for some but not for all [@Christiaanse_2020; @Rosik_Pulawska-Obiedowska_Goliszek_2021].
-Oftentimes, top-down consolidation decisions that are driven by operational cost-savings do not taken into consideration the full extent of benefit loss.
+Oftentimes, top-down consolidation decisions that are driven by operational cost-savings do not take into consideration the full extent of benefit loss.
 
@@
 
-In this study we focus on the spatial travel implications of school closure and consolidation policies.
-The quantification of lost potential active transportation trips and associated carbon emissions, due to school closure, and the unequal impact attributed to the increased reliance on a motorized commute is scarce in the literature.
+In this study, we focus on the spatial travel implications of school closure and consolidation policies.
+The quantification of lost potential active transportation trips and associated carbon emissions due to school closures and the unequal impacts attributed to the increased reliance on a motorized commute is scarce in the literature.
 In this way, our contribution is both methodological and empirical.
@@
 
-In 2013, the City of Hamilton's largest English school board the Hamilton-Wentworth District School Board (HWDSB), serving the majority of students in the city, released their Long Term Facilities Master Plan [@hwdsbLongTermFacilities2013].
+In 2013, the City of Hamilton's largest English school board the Hamilton-Wentworth District School Board (HWDSB), which is the largest English school board in the city, released their Long Term Facilities Master Plan [@hwdsbLongTermFacilities2013].
 To ensure "equitable, affordable and sustainable learning facilities", the Master Plan indicated that 80% of it's elementary schools (i.e., schools attended by students aged 5 to 13) would be subject to accommodation strategies, including a "Accommodation Review" whose outcome is possibly closure/consolidation [@hwdsbLongTermFacilities2013; @craggsBoardEvaluatingFuture2013; @seasonsSchoolClosurePolicy2014].
-These decisions were in part motivated by operational savings based on a projected reduction in student population and under-utilized school capacity [@craggsAxeCouldFall2012].
+These decisions were in part motivated by operational savings based on projected reductions in student population and under-utilized school capacity [@craggsAxeCouldFall2012].
 However, the case of Hamilton and the HWDSB is unlike others in the province: the HWDSB underwent an unprecedented number of Accommodation Reviews relative to other school boards [@seasons2014ARCCatalogues2014].
 Though the Accommodation Review guidelines outline a community-focused process to deliberate a school's future viability to students, the community and the economy [@seasonsSchoolClosurePolicy2014; @ministryofeducationPupilAccommodationReview2006], concerned residences felt a fulsome breakdown of costs to repair schools were not made consistently available [@kleinhuisSeriousConcernsSchool2013].
-Further, others pointed to flaud incentive structures motivating the HWDSB: "By closing three schools the HWDSB would have a strong case with the Ministry of Education to receive full funding for a new school" [@kleinhuisSeriousConcernsSchool2013].
+Further, others pointed to flawed incentive structures motivating the HWDSB: "By closing three schools the HWDSB would have a strong case with the Ministry of Education to receive full funding for a new school" [@kleinhuisSeriousConcernsSchool2013].
 Along with insufficient funding to maintain schools in a state of good repair [@auditorgeneralofontarioAnnualReport20152015], the operational savings as a case for school closures is a long-standing critique of the school "funding formula" established by the conservative provincial government in 1998 [@mackenzieBlueprintFixOntario2018].
@@
 In the first wave of school accommodation reviews between 2013 and 2016, `r ALL_SCHOOLS_Elem %>% filter(Year == "2011" & Level == "Elementary" & Status == "Removed") %>% pull(SchoolID) %>% length()` elementary schools closed.
-From between 2011 (before the school closures) to 2016, these closures represented a `r ((ALL_SCHOOLS_Elem %>% filter(Year != "2016") %>% pull(SchoolID) %>% length() - ALL_SCHOOLS_Elem %>% filter(Year != "2011") %>% pull(SchoolID) %>% length())/ALL_SCHOOLS_Elem %>% filter(Year != "2016" & Level == "Elementary") %>% pull(SchoolID) %>% length()) %>% scales::percent()` and `r  (( ((avail_2011 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2011)) |> select(OTGC) |> sum())) - (avail_2016 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2016)) |> select(OTGC) |> sum())) / (avail_2016 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2016)) |> select(OTGC) |> sum()) ) |> scales::percent()` decline in elementary school locations and capacity, respectively, across the city though overall student population and households in the city did not significantly change between 2011 and 2016 ( `r avail_2011$kid_popadj %>% sum(na.rm=T) %>% prettyNum(big.mark=",")` (2011) and `r avail_2016$kid_popadj %>% sum(na.rm=T) %>% prettyNum(big.mark=",")` (2016)).
-
-From the perspective of public money no longer spent as a result of reduced school operational costs, who and in what way does the Hamilton community pay for these savings by means of accessibility and accessibility-related impacts?
-In this paper, we hypothesize that that this policy negatively impacted the trips-to-school for many students in addition to decreasing potential active-trip opportunities and increasing emissions.
-Spatial availability, a singly-constrained competitive accessibility measure [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024], is used to quantify the before- and after- accessibility and associated impacts for the city of Hamilton and the elementary school-aged student population as a result of a province-wide school closure and consolidation policy.
-
-This paper contributes the following: how does the (1) spatial availability, (2) loss in walk minutes and (3) associated carbon emissions change for home-to-school trips after a city closes and consolidates schools?
-Further, this work examines what communities, based on low-income prevalence, experienced the greatest losses as a result of this policy.
+From between 2011 (before the school closures) to 2016, these closures represented a `r ((ALL_SCHOOLS_Elem %>% filter(Year != "2016") %>% pull(SchoolID) %>% length() - ALL_SCHOOLS_Elem %>% filter(Year != "2011") %>% pull(SchoolID) %>% length())/ALL_SCHOOLS_Elem %>% filter(Year != "2016" & Level == "Elementary") %>% pull(SchoolID) %>% length()) %>% scales::percent()` and `r  (( ((avail_2011 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2011)) |> select(OTGC) |> sum())) - (avail_2016 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2016)) |> select(OTGC) |> sum())) / (avail_2016 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2016)) |> select(OTGC) |> sum()) ) |> scales::percent()` decline in elementary school locations and capacity, respectively, across the city though overall student population in the city did not significantly change between 2011 and 2016 ( `r avail_2011$kid_popadj %>% sum(na.rm=T) %>% prettyNum(big.mark=",")` (2011) and `r avail_2016$kid_popadj %>% sum(na.rm=T) %>% prettyNum(big.mark=",")` (2016)).
+
+While accommodation reviews were undertaken to reduce operational costs, how might the closure of schools impact local communities?
+In this paper, we hypothesize that the school consolidation policy reduced accessibility to schools for many students.
+By extension, this loss in access resulted in a displacement of trips to school made by active modes that increased transportation-related emissions.
+To test these hypotheses, the paper examines the following: how does the (1) spatial availability, (2) loss in walk minutes and (3) associated carbon emissions change for home-to-school trips after a city closes and consolidates schools?
+Further, this work examines (4) what communities, based on low-income prevalence, experienced the greatest losses as a result of this policy.
+Spatial availability, a singly-constrained competitive accessibility measure [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024], is used to quantify the before- and after- accessibility and associated impacts for the city of Hamilton and the elementary school-aged student population as a result of the school closure and consolidation policy.
 The methods introduced in this paper can be used by decision-makers to evaluate the spatial and transport-related impacts of public service provision policy.
-
+Furthermore, to increase the diffusion of this analysis, all work is completed in a R environment and all associated code and data will be open and reproducible within the lead author's [GitHub repository](https://github.com/soukhova)
+
+ >CUT?
 The remainder of the paper is organized as follows.
-Section 2 details the data and methods, particularly multimodal spatial availability, a singly-constrained competitive accessibility measure that considers demand for school-seats by students and friction of distance, as well associated data sources for the inputs into spatial availability.
-Section 3 showcases the results of the impact that the school consolidation policy had on Hamilton's school spatial availability landscape.
+Section 2 details the data and methods and Section 3 showcases the results of the impact that the school consolidation policy had on Hamilton's school spatial availability landscape.
 Section 4 discusses the results and estimates potential walk minutes lost, where trip-related GHG emissions may have increased, and equity implications for the policy.
 Section 5 concludes the paper with key remarks.
-Furthermore, to increase the diffusion of this analysis, all work is completed in a R environment and all associated code and data will be open and reproducible within the lead author's [GitHub repository](https://github.com/soukhova).
 
@@
 Our focus is to evaluate the impact that the school consolidations between 2011 and 2016 had on the multimodal school spatial availability and travel-related GHG emission landscape of the city of Hamilton.
-
 To facilitate comparison, two scenarios are generated.
@@
 The difference between these scenarios are quantified in terms of accessibility, emission and equity implications and discussed in context with the school consolidation policy.
-
 Data is retrieved at different spatial granularity from the following six data sources.
@@
 
-**Firstly**, schools and associated school catchments are retrieve from the school boards.
+First, schools and associated school catchments are retrieved from the City's school boards.
 In Hamilton, the majority of student-aged population attends a school in one of the two English public school boards.
@@
 These catchments indicate what residential property is assigned to what school, by default.
-Families can decide to attend schools out of catchment, in fact 21%-23% <!--calculated in 03-travel-times-TTS.qmd--> of motorized Hamilton-based Hamilton school trips are out of catchment according to the regional travel survey, the Transportation Tomorrow Survey (TTS) in 2011 and 2016 [@data_management_group_tts_2018].
+Families can decide to attend schools out of catchment, in fact 21%-23% <!--calculated in 03-travel-times-TTS.qmd--> of motorized school trips are out of catchment according to the regional travel survey, the Transportation Tomorrow Survey (TTS) in 2011 and 2016 [@data_management_group_tts_2018].
 The elementary school locations and catchments are provided by the HWDSB and HWCDSB for the 2010-2011 and 2015-2016 academic years and visualised in @fig-Fig1.
@@
 typically children aged 5 to 14).
-As such, elementary includes middle schools that only instruct grades 6 to 8 (*Mid*), primary schools which only instruct kindergarten to grade 5 or 6 (*JrElem*), and all grade elementary schools (*Elem*) which instruct all grades from kindergarten to grade 8.
+As such, elementary includes middle schools that only instruct grades 6 to 8 (*Mid*), primary schools that only instruct kindergarten to grade 5 or 6 (*JrElem*), and all grade elementary schools (*Elem*) which instruct all grades from kindergarten to grade 8.
 The maximum number of students that may be enrolled in a school is calculated by the provincial Ministry of Education; this value is referred to as the on-the-ground capacity (*OTGC*) and is assigned to each school.
-The current/historic OTGC province-wide has been made publicly unavailable [@ontarioSchoolFacilityInventory2017].
-However, for schools that underwent the full Accommodation Review process, the OTGC for a given year can be scraped from the generated publicly available School Information Profiles.
-A regression model was derived to estimate the OTGC for schools with no OTGC based on 1) the school's footprint (*F*) in units of (`m^2`), 2) school's shortest Euclidean distance from the centroid of Hamilton's central business district area (*DistCBD*) defined in units of (m), and 3) coefficients associated with a binary variable indicating the type of school grade instruction (*Mid*, *JrElem*, *Elem*).
+The current/historic OTGC province-wide is not publicly available [@ontarioSchoolFacilityInventory2017].
+However, for schools that underwent the full Accommodation Review process, the OTGC for a given year can be obtained from publicly available School Information Profiles.
+A regression model was derived to estimate the OTGC for schools with no capacity information based on 1) the school's building footprint (*F*) in units of ($m^2$), 2) the school's shortest Euclidean distance from the centroid of Hamilton's central business district area (*DistCBD*) defined in units of (m), and 3) coefficients associated with a binary variable indicating the type of school grade instruction (*Mid*, *JrElem*, *Elem*).
 The building footprint from an archived spatial data set [@DMTI_Spatial_2015] and the 2016 footprints were retrieved from OSM [@OpenStreetMap_2021].
-The center of the CBD is defined as a significant intersection in Hamilton Downtown Business Improvement Area [@Hamilton_2014].
-This point serves as a pseudo-proxy for age of school construction as the oldest buildings are located within the Hamilton Downtown boundary [@Merrall_2021] and the age of schools is unavailable.
+Because the age of construction for different schools is not available, we use each school's distance from the centroid of the Downtown Business Improvement Area [@Hamilton_2014] as a proxy for school age as the oldest buildings are generally located closer to the city centre [@Merrall_2021].
 
@@
 ## Students, residential locations and low-income prevelance
-Secondly, student population and proportion of household low-income after tax (LIM-AT) prevalence is retrieved from the 2011 and 2016 Canadian Census [@Statistics_Canada_2011; @Statistics_Canada_2016].
-The census releases population data by age group category and as pertinent to this paper, 5 to 9 year olds and 10 to 14 year olds.
-As more relevant to this paper, the LIM-AT prevelance for households with children under 18 (there is no LIM-AT prevelance tabulated for households with exclusively elementary aged children) is retrained.
+Second, student population and proportion of household low-income after tax (LIM-AT) prevalence is retrieved from the 2011 and 2016 Canadian Census [@Statistics_Canada_2011; @Statistics_Canada_2016].
+The census releases population data by age group category in 5-year increments.
+The LIM-AT prevelance for households with children under 18 (there is no LIM-AT prevelance tabulated for households with exclusively elementary aged children) is retrained.
 LIM-AT is refers to the proportion of private households that are below the median after-tax income [@governmentofcanadaDictionaryCensusPopulation2017].
-These census variables are taken at the finest spatial unit of analysis disseminated by the Canadian census publicly available, the spatial dissemination area (DA).
-They are retrieved using the {cancensus} R package [@von_Bergmann_Shkolnik_Jacobs_2021].
+These census variables are taken at the Dissemination Area (DA) level, which is the finest geographic unit for which census data is made publicly available.
+Census data is retrieved using the {cancensus} R package [@von_Bergmann_Shkolnik_Jacobs_2021].
 DAs are designed by Statistics Canada with the aim of population uniformity hence DAs greatly vary in area but represent between approximately 400 and 700 (1st and 3rd quartile) in total population.
@@
 
-Thirdly, centroids of residential parcels in 2011 and 2016 are used to represent residential locations [@Teranet_2009].
+The third data source consists of centroids of residential parcels in 2011 and 2016 that are used to represent residential locations [@Teranet_2009].
 There are 134,340 and 139,467 residential locations in 2011 and 2016 respectively, representing `r cen_data_2011_hamiltononly$DA_households |> sum() |> prettyNum(big.mark = ",")` and `r cen_data_2016_hamiltononly$DA_households |> sum() |> prettyNum(big.mark = ",")` households (according to the Canadian Census [@Statistics_Canada_2011; @Statistics_Canada_2016]).
@@
 Then the potential number of children at each residential location is calculated by multiplying this DA rate.
-As the [@Teranet_2009] dataset is proprietary, the the middle row in @fig-Fig2 visualises an aggregation of the information: the average rate of 5-14 year olds per residential household at the DA level.
+Due to the proprietary nature of the parcel data, the middle row in @fig-Fig2 visualises an aggregation of the information: the average rate of 5-14 year olds per residential household at the DA level.
 
@@
 
-The fourth data source is an estimated travel time matrices from all residential locations to all elementary schools for both motorized (car) and non-motorized (walking) travel.
-The travel times are estimated using the R-package {r5r} [@Pereira2021r5r], a package that interfaces with the java-based R5 routing engine developed separately by Conveyal [@conveyalConveyalR5Routing2022].
-For both modes, the inputs assume a maximum travel time threshold of 60 minutes and the free-flow OpenStreetMap road network of Hamilton (retrieved using Geofabrick [@geofabrikOntarioOpenStreetMapGeofabrik2022] and manually edited to only include the Hamilton boundary).
-Further, walk travel times that are only less than 27 min minutes were retained.
-This travel time is approximately equivalent to 1.6 km which is the minimum distance that is required from a school within-catchment to qualify for a school bus [@HWDSB_2019].
-<!--put in statistics here-->
-
 It is assumed that children can go to any elementary school, but certain schools are more likely to be attended depending on their proximity to their residential location.
-Based on this assumption, the fifth data source is empirical data regarding the mode-used and origin-destination locations of home-to-school trips from the 2011 and 2016 TTS [@data_management_group_tts_2018].
-The TTS is representative travel survey conducted in the Greater Golden Horseshoe Area in Ontario every 5 years.
+Based on this assumption, the fourth data source is empirical data regarding the mode-used and origin-destination locations of home-to-school trips from the 2011 and 2016 TTS [@data_management_group_tts_2018].
+The TTS is a travel survey conducted in the Greater Golden Horseshoe Area in Ontario every 5 years.
+With a target 5% sampling rate, the survey is expanded to be representative at the traffic analysis zone (TAZ) level of geography.
+TAZ are spatial units created for the purpose of the TTS; a few DAs typically nest within each TAZ.
 The trip-level travel data extract for this paper represent `r od_trips_perc_2011$TAZOrig_car_trips |> sum() |> prettyNum(big.mark = ",")` and `r od_trips_perc_2016$TAZOrig_car_trips |> sum() |> prettyNum(big.mark = ",")` motorized trips (mode used includes private care passenger, school bus, taxi, and transit) and `r od_trips_perc_2011$TAZOrig_walk_trips |> sum() |> prettyNum(big.mark = ",")` and `r od_trips_perc_2016$TAZOrig_walk_trips |> sum() |> prettyNum(big.mark = ",")` non-motorized trips (mode used includes walk and cycling) for 5 to 13 year olds from home to school in 2011 and 2016 respectively.
-Data is available at the spatial resolution of Traffic Analysis Zones (TAZ).
-TAZ are spatial units created for the purpose of the TTS; a few DAs typically nest within each TAZ.
 {r5r} is used to estimated motorized and non-motorized travel from all TAZ centroids to all TAZ centroids and these estimated travel times are tied to the OD trips.
 The travel times and the intensity of modal home-to-school flows are visualised in @fig-Fig3 along with the boundaries of the TAZ and DAs.
-As can be seen in @fig-Fig3, not all TAZ capture an elementary school trip by both modes.
+As can be seen in @fig-Fig3, not all TAZs capture an elementary school trip by both modes.
 For this reason, the proportion of modal-split is aggregated at the community level.
@@
 ```
-Furthermore, as a sixth data source, an estimated travel time matrix for walk and car travel modes is calculated using {r5r} and the geographical centroids of the TTS origin-destination flows.
-For both modes, the inputs assume a maximum travel time threshold of 60 minutes and the free-flow OpenStreetMap road network of Hamilton [@geofabrikOntarioOpenStreetMapGeofabrik2022].
-The walk travel times are filtered to only include travel times under 27 minutes.
+
+The fifth data source consists of estimated travel time matrices from all residential parcel locations to all elementary schools for both motorized (car) and non-motorized (walking) travel.
+The travel times are estimated using the {r5r} [@Pereira2021r5r] package, which interfaces with the java-based R5 routing engine developed separately by Conveyal [@conveyalConveyalR5Routing2022].
+For both modes, the inputs assume a maximum travel time threshold of 60 minutes and the free-flow OpenStreetMap road network of Hamilton (retrieved using Geofabrik [@geofabrikOntarioOpenStreetMapGeofabrik2022] and manually edited to only include the Hamilton boundary).
+Further, walk travel times that are only less than 27 min minutes were retained.
+This travel time is approximately equivalent to 1.6 km which is the minimum distance that is required from a school within-catchment to qualify for a school bus [@HWDSB_2019].
+<!--put in statistics here-->
+
 These travel time matrices along with the origin-destination flows are used to create modal trip length distribution (TLD) functions.
@@
 The empirical TLD is fit to a density distribution using maximum likelihood and moment matching techniques and the Nelder-Mead and Brent methods for direct optimization available within the {fitdistrplus} R package [@fitdistrplus_2015].
-Based on goodness-of-fit criteria and diagnostics (included in Appendix), the gamma and exponential distribution is selected for the motorized and non-motorized modal distributions respectively.
+Based on goodness-of-fit criteria and diagnostics (included in the Appendix), the gamma and exponential distributions are selected for the motorized and non-motorized modal distributions respectively.
 The gamma distribution is defined by the shape ($\alpha$) parameter of `r round(gamma2011_motor$estimate[1], 3)` (2011) and `r round(gamma2016_motor$estimate[1], 3)` (2016) and the rate ($\beta$) of `r round(gamma2011_motor$estimate[2], 3)` (2011) and `r round(gamma2016_motor$estimate[2], 3)` (2016).
@@
 The TLDs for the motorized and non-motorized trips are shown in @fig-Fig4.
+<!--put in equations for the functions-->
 
@@
 It assumed that all non-motorized trips produce no emissions and those that are motorized take the average emission factor of 0.1 kg CO2e per minute travel.
-This emission factor is assumed base on an average speed of 40 km/h, the 2,380 g $CO_{2}e$ produced when 1 L of gasoline is combusted and an internal combustion engine passenger vehicle fuel efficiency of 7 g/100km as discussed in @soukhovOccupancyGHGEmissions2022.
+This emission factor is assumed based on an average speed of 40 km/h, the 2,380 g $CO_{2}e$ produced when 1 L of gasoline is combusted, and an internal combustion engine passenger vehicle fuel efficiency of 7 g/100km as discussed in @soukhovOccupancyGHGEmissions2022.
 
@@
 As indicators of urban structure, Hansen-type measures are informative, but the meaning of 10,000 accessible school-seats is harder to pin down: how many opportunities must any single student have access to?
-Furthermore, in scenario comparisons, the interpretation of the percentage change in accessible school-seats is unclear.
+Furthermore, in scenario comparisons, the interpretation of the percentage change in accessible school seats is unclear.
 The interpretability of Hansen-type accessibility has been discussed in numerous studies, including recently by @hu_2019_measuring, @kelobonye2020measuring, and in greater depth by @merlin2017competition along with @soukhovIntroducingSpatialAvailability2023 and @soukhovMultimodalSpatialAvailability2024.
@@
 
-In this paper, we benefit from new development in accessibility research, particularly the multimodal spatial availability measure [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024].
+In this paper, we benefit from new developments in accessibility research, particularly the multimodal spatial availability measure [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024].
 Spatial availability is a singly-constrained accessibility measure that accounts for competition by students using different modes for exclusive opportunities, such as school seats.
@@
 
-It should be noted, that the population-based allocation factors $F^{pm}_{i}$ when summed over all spatial units in the region always equals to 1 (i.e., $\sum_{m}\sum_{i} F^{pm}_{i}= 1$), likewise for impedance-based allocation factors $F^{cm}_{i}$ (i.e., $\sum_{m}\sum_{i} F^{cm}_{ij} = 1$).
-
-Hansen-type accessibility is not designed to preserve the number of opportunities in the region, it simply counts the potential opportunities that those in a zone can interact with (weighted by the friction of distance).
-Also as discussed in @soukhovIntroducingSpatialAvailability2023, popular competitive accessibility measure such as the 2SFCA [@shenLocationCharacteristicsInnercity1998;@luoMeasuresSpatialAccessibility2003] ends up preserving the number of opportunities in the analysis region but the definitions of variables are internally obscured; the only way it preserves the number of opportunities is if the effect of the impedance function is ignored when expanding the values of opportunities per capita to obtain the total number of opportunities.
-On the other hand, the proportional allocation procedure associated with calculating multi-modal spatial availability $V_i^m$ consistently returns a number of opportunities available to all mode-using populations that matches the total number of opportunities in the region when summed.
+It should be noted that, when summed over all spatial units in the region, the population-based allocation factors $F^{pm}_{i}$ always equal 1 (i.e., $\sum_{m}\sum_{i} F^{pm}_{i}= 1$), likewise for impedance-based allocation factors $F^{cm}_{i}$ (i.e., $\sum_{m}\sum_{i} F^{cm}_{ij} = 1$).
+
+Hansen-type accessibility is not designed to preserve the number of opportunities in the region, it simply counts the intensity of opportunities that those in a zone can potentially interact with (weighted by the friction of distance).
+Also, as discussed in @soukhovIntroducingSpatialAvailability2023, popular competitive accessibility measures such as the 2SFCA [@shenLocationCharacteristicsInnercity1998;@luoMeasuresSpatialAccessibility2003] end up preserving the number of opportunities in the analysis region but the definitions of variables are internally obscured; the only way it preserves the number of opportunities is if the effect of the impedance function is ignored when expanding the values of opportunities per capita to obtain the total number of opportunities.
+On the other hand, the proportional allocation procedure associated with calculating multimodal spatial availability $V_i^m$ consistently returns a number of opportunities available to all mode-using populations that matches the total number of opportunities in the region when summed.
 As a result of this consistency, it is possible to define a measure of multimodal spatial availability per capita ($P$) as presented in Equation (\ref{eq:SA-per-capita}) for use as a benchmark to compare against the regional opportunities per capita ($\frac{\sum_{j} O_j}{\sum_{i} P_i}$).
@@
 
-For the reasons outlined, this paper uses the multimodal spatial availability $V_i^m$ and spatial availability per student $v_i^m$ measures to quantify accessibility changes.
-
-- First, all residential locations are associated to their respective census DA, TTS TAZ and community boundary (@fig-Fig3) based on spatial location.
+For the reasons outlined, this paper uses the multimodal spatial availability $V_i^m$ and spatial availability per student $v_i^m$ measures to quantify accessibility changes for the following reasons.
+First, all residential locations are associated to their respective census DA, TTS TAZ and community boundary (@fig-Fig3) based on spatial location.
 Each residential location is assigned a number of 'potential' motorized and non-motorized student aged population from these different data sources to calculate a motorized and non-motorized population balancing factor $F_{i}^{pm}$.
 For each residential location the following is retained: the 2011 and 2016 census average elementary-aged student population per residential location (@fig-Fig2 second row) and the average motorized and non-motorized split as informed by the TTS aggregated by community (@fig-Fig3).
-- Second, the school capacity $O_j$ is estimated and associated with each school (@fig-Fig1).
+Second, the school capacity $O_j$ is estimated and associated with each school (@fig-Fig1).
 All residential locations are assumed to be able to access all schools by either motorized or non-motorized mode.
 All origins (residential locations) can reach all schools by motorized mode, but few residential locations can reach schools within a 27 minute non-motorized trip.
-- Third, the motorized and non-motorized impedance-balancing factors $F_{ij}^{cm}$ are calculated using the modal theoretical TLDs (@fig-Fig4) based on the respective travel times (data source four) for each origin (residential location) to school combination.
-- Finally, outputs from all three stages are joined together.
+
+Third, the motorized and non-motorized impedance-balancing factors $F_{ij}^{cm}$ are calculated using the modal theoretical TLDs (@fig-Fig4) based on the respective travel times (data source four) for each origin (residential location) to school combination.
+Finally, outputs from all three stages are joined together.
 $F_{i}^{pm}$ joined with $F_{ij}^{cm}$ yields the total balancing factor $F^{tm}_{ij}$ which serves to proportionally allocates the school capacity $O_j$ to each residential location.

### 2024-04-23 paper/Manuscript.qmd
Update Manuscript.qmd

@@
 It is classically presented as the gravity-based measure popularized by @hansenHowAccessibilityShapes1959 and takes the following general multi-modal formulation (Equation (\ref{eq:multi-modal-conventional-accessibility})):
+
 ```{=tex}
@@
 For the reasons outlined, this paper uses the multimodal spatial availability $V_i^m$ and spatial availability per student $v_i^m$ measures to quantify accessibility changes for the following reasons.
-First, all residential locations are associated to their respective census DA, TTS TAZ and community boundary (@fig-Fig3) based on spatial location.
+First, all residential locations are associated to their respective census DA, TTS TAZ, and community boundary (@fig-Fig3) based on spatial location.
 Each residential location is assigned a number of 'potential' motorized and non-motorized student aged population from these different data sources to calculate a motorized and non-motorized population balancing factor $F_{i}^{pm}$.
@@
 
-Third, the motorized and non-motorized impedance-balancing factors $F_{ij}^{cm}$ are calculated using the modal theoretical TLDs (@fig-Fig4) based on the respective travel times (data source four) for each origin (residential location) to school combination.
+Third, the motorized and non-motorized impedance-balancing factors $F_{ij}^{cm}$ are calculated using the modal theoretical TLDs (@fig-Fig4) based on the respective travel times for each origin (residential location) to school combination.
 Finally, outputs from all three stages are joined together.
-$F_{i}^{pm}$ joined with $F_{ij}^{cm}$ yields the total balancing factor $F^{tm}_{ij}$ which serves to proportionally allocates the school capacity $O_j$ to each residential location.
+$F_{i}^{pm}$ joined with $F_{ij}^{cm}$ yields the total balancing factor $F^{tm}_{ij}$ which serves to proportionally allocate the school capacity $O_j$ to each residential location.
 The value is multimodal spatial availability $V_i^m$ which can be interpreted as the number of school-seats that are available to the mode-using population at that residential location.
-$V_i^m$ is then summed and represented at the DA and Community level along with the calculated $v_i^m$ for interpretation.
+$V_i^m$ is then summed and represented at the DA and community level along with the calculated $v_i^m$ for interpretation.
 
@@
 Trends appear similar for both 2011 and 2016: proximate populations using motorized modes benefit from a wider-range in high spatial availability than non-motorized modes.
-When $V_im$ is divided by the number of mode-using students in that DA, $v_i^m$ is derived (bottom half of @fig-Fig6).
+When $V_i^m$ is divided by the number of mode-using students in that DA, $v_i^m$ is derived (bottom half of @fig-Fig6).
 A similar spatial distribution seen in $V_i^m$ can be seen in $v_i^m$ plots, but as a rate of student population.
 If our aim is to provide sufficient availability to schools for both motorized and non-motorized populations (i.e., greater than 1.0 $v_i^m$ school-seat availability per student), this rate is only achieved in the core of the city for motorized modes, and notably only for non-motorized populations in DAs that are in less-densely populated rural areas that are school proximate and certain pockets of Hamilton Central.
-The majority of schools closed were within Hamilton Central, so non-motorized populations in DAs within Hamilton Central were the most negatively, and certain motorized populations captured these losses as gains in availability.
+The majority of schools closed were within Hamilton Central, so non-motorized populations in DAs within Hamilton Central were the most negatively impacted by losses in availability while certain motorized populations captured these losses as availability gains.
 
@@
 To emphasize how spatial availability changed between 2016 and 2011, the change in $V_i^m$ and $v_i^m$ is visualised for both modes in @fig-Fig7.
-Overall, a decrease in both $V_i^m$ and $v_i^m$ can be seen both modes modes around closed schools (black points).
-An increase near new or expanded schools (green points) is not as noteable as overall fewer school seats were added relative to population increase.
-The non-motorized mode-using populations are most impacted by the school closures/consolidations along with other competition factors such as increased children population in less urban areas and an increase in motorized mode use.
-For non-motorzied mode-using populations, $V_i^m$ decreases across the city.
+Overall, a decrease in both $V_i^m$ and $v_i^m$ can be seen across both modes around closed schools (black points).
+An increase near new or expanded schools (green points) is not as notable as overall fewer school seats were added relative to increases in the school-aged population, leading to increased competition for these resources.
+For non-motorized mode-using populations, $V_i^m$ decreases across the city.
 However, some motorized-mode using populations in DAs do experience an increase in $V_i^m$: these DAs are often relatively proximate to schools but more immediately outside walking-range.
-Under a competitive and constraint calculation, these DAs benefit from the losses in spatial availability that are predominately experienced by the non-motorized populations within Hamilton Central.
+Under a competitive and constraint-based calculation, these DAs benefit from the losses in spatial availability that are predominately experienced by the non-motorized populations within Hamilton Central.
 
 <!--Maybe worth adding a plot of GHG emissions or change in tt?-->
-
-# Discussion
 
@@
 This out of proportion decrease is greatest in Hamilton Central than any other community.
-This may have grave equity concerns as Hamilton Central has the highest LIM-AT prevalence, with `r SAvail_im_COMM_changes |> st_drop_geometry() |> filter(COMMUNITY == "HA" & mode == "motorized") |> dplyr::select("prev_LIMAT_2016") |> round() |> paste0("%")` of households LIM-AT in 2016, 3.7 times greater than Glanbrook with the lowest level of LIM-AT prevalence.
+This may have equity concerns as Hamilton Central has the highest LIM-AT prevalence, with `r SAvail_im_COMM_changes |> st_drop_geometry() |> filter(COMMUNITY == "HA" & mode == "motorized") |> dplyr::select("prev_LIMAT_2016") |> round() |> paste0("%")` of households LIM-AT in 2016, 3.7 times greater than the community with the lowest level of LIM-AT prevalence in the city.
 Between 2016 and 2011, Hamilton Center had the lowest levels of emissions per student and saw the highest per student increase in 2016.
@@
 
-Across the city, those who commute by non-motorized mode suffer from the most dramatic drop in student population and especially spatial availability across the city.
+Across the city, those who commute by non-motorized mode see the most dramatic drop in spatial availability across the city.
 In all communities, spatial availability decreases at greater levels than population.
@@
 
-# Concluding remarks
+# Conclusion
 
@@
 We urge policy makers to view 'spatial availability' of public resources from a per-capita perspective, plan capacity and location of service provision with the cost of travel (and associated implications such as GHG emissions, active transportation mode, safety of active travel modes, etc.) that sufficiently serves the target population.
-In this paper we suggest and assume a benchmark of 1.0 school-seats per student which accounts for sufficy.
+In this paper we suggest and assume a benchmark of 1.0 school-seats per student which accounts for sufficiency.
 Spatial availability can be used to obtain this benchmark since the total number of opportunities (i.e., school capacity) is preserved in the region of analysis unlike unconstrained forms of accessibility measurement e.g., Hansen-type measure [@hansenHowAccessibilityShapes1959].

### 2024-04-26 paper/Manuscript.qmd
manuscript draft for AP review

@@
 numbersections: true
-format: docx
+format:
+  nature-pdf: default
 editor: source
@@
 Overall, spatial availability decreased in the city.
-Though the school conslidation policy resulted in short-term operational cost-savings, families, students, and the environment now pay the price.
+Though the school conslidation policy resulted in short-term operational cost-savings, families, students, and the environment now pays the price.
 This paper showcases how spatial availability can be used by decision- makers to quantify spatial and travel implications associated with policy changes.
@@
 
-<!--- 2016 and 2011 sp_avail is calculated in 'test-doc-2016run', LOAD THE BELOW! -->
 ```{r, include=FALSE}
@@
 
-<!--
-Additional methods considerations: circuity of trip by mode?
-Schoolbus is not as direct of a trip as car/taxi/bikeORwalk.
-Maybe we consider 3 modes instead of 2?
-The r5r travel time for car/taxi as well as walk/cycle can be used as is, but the travel time by school bus can be increased based on a trip distance factor (from TTS)?
-
-RESULTS an alternative layout idea?:
-1.
-Multimodal spatial availability aggregated at the DA for each year and for each mode.
-Each DA is ~200 parcel points with their low-income and other-income spatial availability per mode; both these are summed together along with the rest of the parcel points per da.
-(6 plots - 3 modes per 2011 and 3 modes per 2016).
-2.
-Income-specific spatial availability aggregated at the DA for each population type.
-Similarly, each DA is ~200 parcel points with 3 mode-specific spatial availability per income-group; all three are summed together along with the rest of the parcel points per da.
-(4 plots - 2 population-groups per 2011 and 2 population-groups per 2016).
-3. EXAMINING THE INTERMEDIATES: the change in spatial availability per focused communities:
-3.a.
-Hamilton-Central+Dundas experienced the majority of school closures, also have the highest LICO-AT as well as motorized transport.
-Glanbrook+StoneyCreek experienced the majority of new schools (two), also have low LICO-AT as well as non-motorized transport.
-So.
-2 columns.
-First column SA By Mode.
-SA By Mode in 2016 for HC+Dundas, then SA By Mode in 2016 for Glan+SC, then SA By Mode in 2016 Overall.
-Second column SA By Pop.
-SAY by LICO or not LICO in 2016 for HC+Dundas, then for GLAN+SC, then for overall.
-Showing SA for 2016 with brackets that say (+ or - X% to indicate difference from 2011).
-3.b. Examining the change in travel time per focused communities: same bar graph style as in 3.a. Will show have non-motorized travel time decreased in Ham-central, esp for LICO(?)
-3.c. Change in emissions: same bar graph style as in 3.a. but for emissions. Will show have emissions increased, esp in Ham-central and LICO(?)
-4.
-USING SPATIAL AVAILABILITY FOR PLANNING.
-Equity?
-We want to plan for school-seat availability _for_ active modes.
-Especially in areas with currently high LICO-AT.
-So lets plot spatial availability per non-motorized-LICO-AT capita.
-We'd like to see 1.2 SA per capita everywhere.
-We need improvements to the population, opportunities, and non motorized transport network (relative to all other modes/population groups) is the key point.
--->
-
 # Introduction
@@
 In this way, our contribution is both methodological and empirical.
-We quantify the accessibility changes before- and after the first wave of school closure/consolidations implemented in Hamilton, a mid-size city (\~570,000 pop) with rural, suburban and urban characteristics in Ontario, Canada [@governmentofcanadaProfileTableCensus2022].
+We use a novel singly-constrained accessibility measure to quantify changes before- and after the first wave of school closure/consolidations implemented in Hamilton, a mid-size city (\~570,000 pop) with rural, suburban and urban characteristics in Ontario, Canada [@governmentofcanadaProfileTableCensus2022].
 
@@
 Along with insufficient funding to maintain schools in a state of good repair [@auditorgeneralofontarioAnnualReport20152015], the operational savings as a case for school closures is a long-standing critique of the school "funding formula" established by the conservative provincial government in 1998 [@mackenzieBlueprintFixOntario2018].
-In a backdrop of budget cuts, the funding formula pits financing of school maintenance and student needs against the financial compensation of staff and powers of local school boards (as they no longer have the ability to levee property taxes) [@mackenzieBlueprintFixOntario2018].
+With the backdrop of budget cuts, the funding formula pits financing of school maintenance and student needs against the financial compensation of staff and powers of local school boards as they no longer have the ability to levee property taxes [@mackenzieBlueprintFixOntario2018].
 
@@
 Furthermore, to increase the diffusion of this analysis, all work is completed in a R environment and all associated code and data will be open and reproducible within the lead author's [GitHub repository](https://github.com/soukhova)
-
- >CUT?
-The remainder of the paper is organized as follows.
-Section 2 details the data and methods and Section 3 showcases the results of the impact that the school consolidation policy had on Hamilton's school spatial availability landscape.
-Section 4 discusses the results and estimates potential walk minutes lost, where trip-related GHG emissions may have increased, and equity implications for the policy.
-Section 5 concludes the paper with key remarks.
 
@@
 School status and OTGC summed by catchment is also visualised in @fig-Fig1.
-See the Table \ref{Tab:TabA1} in the Appendix for the associated regression results.
+See the Table <!--\ref{TabA1-OTGC}--> in the Appendix for the associated regression results.
+<!-- shall we explain further how we valuated this model?
+I don't think its entirely necessary.
+But the code is here: https://github.com/soukhova/hamONdests/blob/master/data-raw/01-school-data-preparation.Rmd (Antonio is a collaborator) -->
 
@@
 ```
+<!--perhaps in the legend of this map, I add the number of schools in brackets (for "school status between 2011 and 2016")?
+And the number of schools on the map in each box?
+what do you think?
+I'll also make the font on the panels bigger-->
 
@@
 The census releases population data by age group category in 5-year increments.
-The LIM-AT prevelance for households with children under 18 (there is no LIM-AT prevelance tabulated for households with exclusively elementary aged children) is retrained.
+The LIM-AT prevalence for households with children under 18 (there is no LIM-AT prevalence tabulated for households with exclusively elementary aged children) is retrained.
 LIM-AT is refers to the proportion of private households that are below the median after-tax income [@governmentofcanadaDictionaryCensusPopulation2017].
@@
 There are 134,340 and 139,467 residential locations in 2011 and 2016 respectively, representing `r cen_data_2011_hamiltononly$DA_households |> sum() |> prettyNum(big.mark = ",")` and `r cen_data_2016_hamiltononly$DA_households |> sum() |> prettyNum(big.mark = ",")` households (according to the Canadian Census [@Statistics_Canada_2011; @Statistics_Canada_2016]).
-As the number of children per residential location is unavailable publicly, the average number of 5 to 14 year olds for each DA is divided by the number of residential households in that DA.
+As the number of children per residential location are publicly unavailable, the average number of 5 to 14 year olds for each DA is divided by the number of residential households in that DA.
 Then the potential number of children at each residential location is calculated by multiplying this DA rate.
-Due to the proprietary nature of the parcel data, the middle row in @fig-Fig2 visualises an aggregation of the information: the average rate of 5-14 year olds per residential household at the DA level.
+<!-- NOTE: We consider the parcel centroids instead of DA centroids to more precisely calc.
+travel times.
+Since we do not know parcel student population, we assume a uniform spatial distribution of population across each DA.
+This assumption is a limitation, but it is better than what's typically done in studies that only consider the centroid of the census track.
+--> Due to the proprietary nature of the parcel data, the middle row in @fig-Fig2 visualises an aggregation of the information: the average rate of 5-14 year olds per residential household at the DA level.
 
@@
 The trip-level travel data extract for this paper represent `r od_trips_perc_2011$TAZOrig_car_trips |> sum() |> prettyNum(big.mark = ",")` and `r od_trips_perc_2016$TAZOrig_car_trips |> sum() |> prettyNum(big.mark = ",")` motorized trips (mode used includes private care passenger, school bus, taxi, and transit) and `r od_trips_perc_2011$TAZOrig_walk_trips |> sum() |> prettyNum(big.mark = ",")` and `r od_trips_perc_2016$TAZOrig_walk_trips |> sum() |> prettyNum(big.mark = ",")` non-motorized trips (mode used includes walk and cycling) for 5 to 13 year olds from home to school in 2011 and 2016 respectively.
-{r5r} is used to estimated motorized and non-motorized travel from all TAZ centroids to all TAZ centroids and these estimated travel times are tied to the OD trips.
+The travel times for motorized and non-motorized travel are estimated using the {r5r} [@Pereira2021r5r] package, which interfaces with the java-based R5 routing engine developed separately by Conveyal [@conveyalConveyalR5Routing2022].
+Motorized and non-motorized travel times are estimated for all TAZ centroids to all TAZ centroids and these estimated travel times are tied to the associated OD trips.
 The travel times and the intensity of modal home-to-school flows are visualised in @fig-Fig3 along with the boundaries of the TAZ and DAs.
@@
 Hamilton (Ancaster `r community_car_perc_2011$TAZOrig_car_perc[1] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[1] |> percent()`, Dundas `r community_car_perc_2011$TAZOrig_car_perc[2] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[2] |> percent()`, Flamborough `r community_car_perc_2011$TAZOrig_car_perc[3] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[3] |> percent()`, Glanbrook `r community_car_perc_2011$TAZOrig_car_perc[4] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[4] |> percent()`, Hamilton Central `r community_car_perc_2011$TAZOrig_car_perc[5] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[5] |> percent()`, and Stoney Creek `r community_car_perc_2011$TAZOrig_car_perc[6] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[6] |> percent()`).
+<!--Do we maybe need to discuss how this lack of data for each TAZ impacted the calculations?-->
 
@@
 The fifth data source consists of estimated travel time matrices from all residential parcel locations to all elementary schools for both motorized (car) and non-motorized (walking) travel.
-The travel times are estimated using the {r5r} [@Pereira2021r5r] package, which interfaces with the java-based R5 routing engine developed separately by Conveyal [@conveyalConveyalR5Routing2022].
+The travel times are estimated using the {r5r} package [@Pereira2021r5r].
 For both modes, the inputs assume a maximum travel time threshold of 60 minutes and the free-flow OpenStreetMap road network of Hamilton (retrieved using Geofabrik [@geofabrikOntarioOpenStreetMapGeofabrik2022] and manually edited to only include the Hamilton boundary).
@@
 This travel time is approximately equivalent to 1.6 km which is the minimum distance that is required from a school within-catchment to qualify for a school bus [@HWDSB_2019].
-<!--put in statistics here-->
+Estimated trip-weighted motorized travel times are on average 16.5 minutes in 2011 and 16.8 minutes in 2016 (summary statistics for both years are approximately min: 0.5, Q1: 11.0, Q2: 16.0, Q3: 21.0, max: 60.0).
+Estimated trip-weighted non-motorized travel times are on average 18.0 minutes in 2011 and 18.1 minutes in 2016 (summary statistics for both years are approximately min: 0.5, Q1: 13.0, Q2: 19.0, Q3: 24.0, max: 27.0).
+<!--to calc.
+the summary statistics i loaded "OD_2011" and "OD_2016" from data-products (big files) and ran: quantile(OD_2011|>filter(mode=="w" & travel_time_p50 <= 27)|>pull(travel_time_p50), weights=OD_2011|>filter(mode=="w")|>pull(kid_popadj))
+and 
+mean(OD_2011|>filter(mode=="w" & travel_time_p50 <= 27)|>pull(travel_time_p50), weights=OD_2011|>filter(mode=="w")|>pull(kid_popadj))
+-->
 
@@
 The empirical TLD is fit to a density distribution using maximum likelihood and moment matching techniques and the Nelder-Mead and Brent methods for direct optimization available within the {fitdistrplus} R package [@fitdistrplus_2015].
-Based on goodness-of-fit criteria and diagnostics (included in the Appendix), the gamma and exponential distributions are selected for the motorized and non-motorized modal distributions respectively.
+Based on goodness-of-fit criteria and diagnostics <!-- should I include the diagonostic plots in the appendix?
+q-q, etc..?
+I don't think its entirely necessary as the empirical is graphed over the theoretical line in the following figure-->, the gamma and exponential distributions are selected for the motorized and non-motorized modal distributions respectively.
 The gamma distribution is defined by the shape ($\alpha$) parameter of `r round(gamma2011_motor$estimate[1], 3)` (2011) and `r round(gamma2016_motor$estimate[1], 3)` (2016) and the rate ($\beta$) of `r round(gamma2011_motor$estimate[2], 3)` (2011) and `r round(gamma2016_motor$estimate[2], 3)` (2016).
@@
 The TLDs for the motorized and non-motorized trips are shown in @fig-Fig4.
-<!--put in equations for the functions-->
+For reference, the gamma distribution and the exponential distribution function are displayed in the following Equation (\ref{eq:exp-dist})) and (\ref{eq:gamma-dist})) where $x$ is $c_{ij}$:
+
+```{=tex}
+\begin{equation}
+\label{eq:exp-dist}
+f(x) = \beta e ^{-\beta x}
+\end{equation}
+```
+
+```{=tex}
+\begin{equation}
+\label{eq:gamma-dist}
+\begin{aligned}
+f(x, \alpha, \beta) = \frac {x^{\alpha-1}e^{-\frac{x}{\beta}}}{ \beta^{\alpha}\Gamma(\alpha)} \quad \text{for } 0 \leq x \leq \infty \\
+\Gamma(\alpha) =  \int_{0}^{\infty} x^{\alpha-1}e^{-x} \,dx
+\end{aligned}
+\end{equation}
+```
 
@@
 #| label: fig-Fig4
-#| fig-cap: "Empirical and theoretical motorized and non-motorized impedance functions."
+#| fig-cap: "Empirical (grey bars) and theoretical (blue curves) motorized and non-motorized impedance functions.
+Motorized theoretical impedance function is based on the gamma distribution function and non-motorized on the expoential distribution function"
 #| echo: false
@@
 ```
+<!--maybe bigger font here?-->
 
@@
 ```
-
+<!--I'll make this plot longer-->
 The motorized and non-motorized spatial availability $V_i$ for both 2011 and 2016 is represented in the top half of @fig-Fig6.
@@
 ```
+<!--The colours combos I choose here are absolutely ugly ;( but besides that, do you agree that V_im and v_im should use different colour palettes?-->
 
@@
 
-<!--Maybe worth adding a plot of GHG emissions or change in tt?-->
-
-<!--This will be a table with summaries of LIMAT, Kid pop, V_im change, v_im change, total travel time change, and emission changed by community and mode?
-We shoudl discuss to decide what's the best presentation.-->
+<!--Antonio, what do you think, maybe worth adding a plot of GHG emissions or change in tt?
+This will provide us some higher level summary stats.
+Moataz asked about this as well regarding the figures in the Results section-->
+
+<!--Antonio, what do you think, add a table with summaries of LIMAT, Kid pop, V_im change, v_im change, total travel time change, and emission changed by community and mode?
+We should discuss to decide what's the best presentation.-->
 ```{r}
@@
   dplyr::select(c("COMMUNITY",
-                  "prev_LIMAT_2016", #prevalance of LIM after-tax for households with children under 18
+                  "prev_LIMAT_2016", #prevalence of LIM after-tax for households with children under 18
                   "mode", 
@@
 Overall, the gain in motorized spatial availability resulted in a further erosion of non-motorized spatial availability.
-This erosion results in a lose in active transport trips and an increase in emissions city-wide (`r SAvail_im_COMM_changes |> st_drop_geometry() |> transmute(em = (sum(emissions_2011))/(sum(kid_2011)) ) |> sum() |> round(digit=2)`kg CO2e per student in 2011 to `r SAvail_im_COMM_changes |> st_drop_geometry() |> transmute(em = (sum(emissions_2016))/(sum(kid_2016)) ) |> sum() |> round(digit=2)` in 2016).
-
-<!-- develop further?-->
+This erosion results in a lose in active transport trips and an increase in emissions city-wide (`r SAvail_im_COMM_changes |> st_drop_geometry() |> transmute(em = (sum(emissions_2011))/(sum(kid_2011)) ) |> sum() |> round(digit=2)`kg $CO_2e$ per home-to-school potential student-trip in 2011 to `r SAvail_im_COMM_changes |> st_drop_geometry() |> transmute(em = (sum(emissions_2016))/(sum(kid_2016)) ) |> sum() |> round(digit=2)` in 2016).
+
+<!-- Is this worth further developing or perhaps just remove, what do you think? I think further. -->
 Study limitations:
@@
 
-<!-- Add an additional paper: discuss the potential impact of mode-switching on GHG emissions reductions [@pantelakiImpactHomeschoolCommuting2024].
-
-Also, access to schools has broader implications; it has a relationship to residential and job choices [@pietrabissaSchoolAccessCity2023] and housing prices [@merrallWhatSchoolWorth2023]-->
+<!-- Add this additional paper: it discusses the potential impact of mode-switching on GHG emissions reductions [@pantelakiImpactHomeschoolCommuting2024].
+
+Also, access to schools has broader implications; it has a relationship to residential and job choices [@pietrabissaSchoolAccessCity2023] and housing prices [@merrallWhatSchoolWorth2023] ?
+Shall I further elaborate in the discussion?-->
 
@@
 
+# References
+
+::: {#refs}
+:::
+
 # Appendix
 
-The regression model for the OTGC for schools goes here.
-
-# References
+\input{TabA1-OTGC-Reg.tex} 
+<!--made in the hamONdests repo-->

### 2024-04-26 paper/Manuscript.qmd
Edits by Antonio

@@
 ---
-title: "The cost of school consolidations: impacts of reduced accessibility"
+title: "School closures and consolidations: the active travel and emission implications of reduced accessibility"
 bibliography: 
@@
 abstract: | 
-  | The reduction of mobility-induced GHG emissions is a top priority for many communities around the world.
-From a different standpoint, communities are consistently under pressure to increase operational cost-efficiency savings, including the consideration of spatially consolidating public resource provision.
-This is the case for public schools, where governments are opting to close under-capacity schools and expand existing schools to accommodate lost capacity.
-This study quantifies the travel burdens of school consolidation/closure policy through the case of Hamilton, a mid-sized city in Ontario, Canada.
+  | Reducing emissions and increasing active travel are often top priorities for many communities.
+These priorities, though, often are confronted with a pressure to reduce the fiscal burden of public services, including schools.
+Public schools have for some time now been pressured to do more with less, and governments are opting to close under-capacity schools and expand existing schools to accommodate lost capacity.
+The objective of this study is to quantifies the travel burdens of school consolidation/closure policy.
+We do this through the case of Hamilton, a mid-sized city in Ontario, Canada.
 Between the years 2011 and 2015, 7% of the elementary schools closed in the city.
-Specifically, we quantify how these closures changed the accessibility landscape for the elementary school aged population by using spatial availability, a novel singly-constrained accessibility measure.
-Spatial availability allows for an interpretable comparison between ‘before’ and ‘after’ policy implementation because of its unique proportional allocation property.
-The analysis is conducted at the parcel-level with all school seats being eligible and travel behaviour based on observed home-to-school patterns for motorzied and non-motorized modes.
-System-wide impacts resulting from the closures, such as potential GHG emission increases and walk minutes decreases, are detailed.
-Overall, spatial availability decreased in the city.
-Though the school conslidation policy resulted in short-term operational cost-savings, families, students, and the environment now pays the price.
-This paper showcases how spatial availability can be used by decision- makers to quantify spatial and travel implications associated with policy changes.
+For this research, we quantify how these closures changed the accessibility landscape for the elementary school aged population, and the active travel and emission implications.
+In terms of methods, we use spatial availability, a singly-constrained multi-modal measure of spatial accessibility.
+Thanks to its proportional allocation mechanism, spatial availability allows for an interpretable comparison between 'before' and 'after' the implementation of school closure/consolidation policy.
+Analysis is conducted at the parcel-level with all school seats being eligible and travel behaviour based on observed home-to-school patterns for motorized and non-motorized modes.
+The system-wide impacts that result from the closures, such as potential GHG emission increases and walk minutes decreases, are estimated Overall, spatial availability decreased in the city.
+Though the school consolidation policy may have resulted in cost-savings in operation and maintenance of facilities, families, students, and the environment now pay the price.
 keywords: [accessibility; spatial availability; carbon emissions; journey-to-school; walkability; active transportation; school consolidation policy; service provision thresholds]
@@
 These forces spur the re-organization and consolidation of public facilities in communities around the world where changes to service provision may improve the quality of service for some but not for all [@Christiaanse_2020; @Rosik_Pulawska-Obiedowska_Goliszek_2021].
-Oftentimes, top-down consolidation decisions that are driven by operational cost-savings do not take into consideration the full extent of benefit loss.
-
-From this perspective, journey-to-school trips are tremendously important: this trip purpose is the largest contributor of carbon emissions from personal travel after journey-to-work trips [@rongReviewResearchLowcarbon2022; @pantelakiImpactHomeschoolCommuting2024].
-In many communities globally, school locations are closing [@sageman2022] and trips made to school are increasingly motorized [@rongReviewResearchLowcarbon2022].
-Nonetheless, school closure policies have been pursued by governments to optimize per-student operational cost-efficiency [@rongReviewResearchLowcarbon2022] but spatial and travel implications have often been overlooked by decision makers [@bierbaumMobilityJusticeLinking2021; @Lee_Lubienski_2017].
+While change is natural, however, oftentimes top-down consolidation decisions that are driven by operational cost-savings do not take into consideration the full extent of losses in benefits.
+
+From this perspective, school closure policies have been pursued by governments to optimize per-student operational cost-efficiency [@rongReviewResearchLowcarbon2022] but other implications, such as the impact of closures on property values [@MerrallWhatsASchool2024], the effects on the social infrastructure of communities [@butlerClosureRideau2019a; @irwinSchoolClosure2012], and particularly the travel implications of school closures have often been overlooked by decision makers [@bierbaumMobilityJusticeLinking2021; @Lee_Lubienski_2017].
+This seems a mistake: journey-to-school trips are tremendously important, and travel for this purpose is the largest contributor of carbon emissions from personal travel after the journey-to-work [@rongReviewResearchLowcarbon2022; @pantelakiImpactHomeschoolCommuting2024].
+In many communities globally, schools are closing [@sageman2022].
 For instance, 65% of comprehensive schools (e.g.
 grades 1 through 9) in rural Finland since 1990 closed [@Autti_Hyry-Beihammer_2014], 10% of elementary schools (kindergarten through grade 8) in Chicago, USA in 2013 [@Lee_Lubienski_2017], and 7% of primary and lower secondary schools (kindergarten through grade 9) were closed in Denmark between 2010-2011 [@Beuchert_Humlum_Nielsen_Smith_2018].
-
-In this study, we focus on the spatial travel implications of school closure and consolidation policies.
-The quantification of lost potential active transportation trips and associated carbon emissions due to school closures and the unequal impacts attributed to the increased reliance on a motorized commute is scarce in the literature.
-In this way, our contribution is both methodological and empirical.
-We use a novel singly-constrained accessibility measure to quantify changes before- and after the first wave of school closure/consolidations implemented in Hamilton, a mid-size city (\~570,000 pop) with rural, suburban and urban characteristics in Ontario, Canada [@governmentofcanadaProfileTableCensus2022].
-
-In 2013, the City of Hamilton's largest English school board the Hamilton-Wentworth District School Board (HWDSB), which is the largest English school board in the city, released their Long Term Facilities Master Plan [@hwdsbLongTermFacilities2013].
-To ensure "equitable, affordable and sustainable learning facilities", the Master Plan indicated that 80% of it's elementary schools (i.e., schools attended by students aged 5 to 13) would be subject to accommodation strategies, including a "Accommodation Review" whose outcome is possibly closure/consolidation [@hwdsbLongTermFacilities2013; @craggsBoardEvaluatingFuture2013; @seasonsSchoolClosurePolicy2014].
-These decisions were in part motivated by operational savings based on projected reductions in student population and under-utilized school capacity [@craggsAxeCouldFall2012].
-However, the case of Hamilton and the HWDSB is unlike others in the province: the HWDSB underwent an unprecedented number of Accommodation Reviews relative to other school boards [@seasons2014ARCCatalogues2014].
-Though the Accommodation Review guidelines outline a community-focused process to deliberate a school's future viability to students, the community and the economy [@seasonsSchoolClosurePolicy2014; @ministryofeducationPupilAccommodationReview2006], concerned residences felt a fulsome breakdown of costs to repair schools were not made consistently available [@kleinhuisSeriousConcernsSchool2013].
+School closures often increase the distance to school for many students and thus tend to induce motorized trips [@rongReviewResearchLowcarbon2022].
+
+In this study, we focus on the transportation implications of school closure and consolidation policies.
+While some research has considered the "steps in reserve", that is, the potential for active travel hidden behind motorized trips [@morencyStepsReserve2009; @morencyHowMany2007], school closures potentially make actual steps vanish, sent to the reserve for the foreseeable future.
+In exchange for less physical activity, communities get more motorized travel, with the concomitant costs and burdens, including emissions.
+Quantification of losses in potential active travel and associated growith in carbon emissions is scarce in the literature, and the focus of this paper, which aims to make both methodological and empirical contributions.
+We take a novel multi-modal and singly-constrained accessibility measure, called spatial availability [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024], and use it to estimate the changes in accessibility and travel before and after a recent wave of school closure/consolidations implemented in Hamilton, Ontario.
+
+The results of the research indicate that, overall, school seats became less spatially available in the city.
+Though the school consolidation policy pursued by the province and implemente by the local school board may have resulted in cost-savings in operation and maintenance of facilities, families, students, and the environment now pay the price in terms of lost opportunities for healthier active travel, as well as increased emissions from motorized travel.
+
+# Background
+
+Hamilton is a mid-size city (\~570,000 pop) with rural, suburban and urban characteristics in the province of Ontario, Canada [@governmentofcanadaProfileTableCensus2022].
+In 2013, the Hamilton-Wentworth District School Board (HWDSB),, the largest public English school board in the city, released its Long Term Facilities Master Plan [@hwdsbLongTermFacilities2013].
+To ensure "equitable, affordable and sustainable learning facilities", the Master Plan indicated that 80% of it's elementary schools (i.e., schools attended by students aged 5 to 13) would be subjected to accommodation evaluations, including an "Accommodation Review" whose outcome historically has led to school closures/consolidation [@hwdsbLongTermFacilities2013; @craggsBoardEvaluatingFuture2013; @seasonsSchoolClosurePolicy2014].
+These decisions by the school board were in part motivated by a search for operational savings based on projected reductions in student population and under-utilized school capacity [@craggsAxeCouldFall2012].
+However, the case of Hamilton and the HWDSB was unlike others in the province: the HWDSB underwent an unprecedented number of Accommodation Reviews relative to other school boards [@seasons2014ARCCatalogues2014].
+
+Though the Accommodation Review guidelines outline a community-focused process to deliberate a school's future viability to students, the community and the economy [@seasonsSchoolClosurePolicy2014; @ministryofeducationPupilAccommodationReview2006], concerned residents felt a wholesome breakdown of costs to repair schools was not made consistently available [@kleinhuisSeriousConcernsSchool2013].
 Further, others pointed to flawed incentive structures motivating the HWDSB: "By closing three schools the HWDSB would have a strong case with the Ministry of Education to receive full funding for a new school" [@kleinhuisSeriousConcernsSchool2013].
-Along with insufficient funding to maintain schools in a state of good repair [@auditorgeneralofontarioAnnualReport20152015], the operational savings as a case for school closures is a long-standing critique of the school "funding formula" established by the conservative provincial government in 1998 [@mackenzieBlueprintFixOntario2018].
-With the backdrop of budget cuts, the funding formula pits financing of school maintenance and student needs against the financial compensation of staff and powers of local school boards as they no longer have the ability to levee property taxes [@mackenzieBlueprintFixOntario2018].
-
-In the first wave of school accommodation reviews between 2013 and 2016, `r ALL_SCHOOLS_Elem %>% filter(Year == "2011" & Level == "Elementary" & Status == "Removed") %>% pull(SchoolID) %>% length()` elementary schools closed.
-From between 2011 (before the school closures) to 2016, these closures represented a `r ((ALL_SCHOOLS_Elem %>% filter(Year != "2016") %>% pull(SchoolID) %>% length() - ALL_SCHOOLS_Elem %>% filter(Year != "2011") %>% pull(SchoolID) %>% length())/ALL_SCHOOLS_Elem %>% filter(Year != "2016" & Level == "Elementary") %>% pull(SchoolID) %>% length()) %>% scales::percent()` and `r  (( ((avail_2011 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2011)) |> select(OTGC) |> sum())) - (avail_2016 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2016)) |> select(OTGC) |> sum())) / (avail_2016 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2016)) |> select(OTGC) |> sum()) ) |> scales::percent()` decline in elementary school locations and capacity, respectively, across the city though overall student population in the city did not significantly change between 2011 and 2016 ( `r avail_2011$kid_popadj %>% sum(na.rm=T) %>% prettyNum(big.mark=",")` (2011) and `r avail_2016$kid_popadj %>% sum(na.rm=T) %>% prettyNum(big.mark=",")` (2016)).
+Along with insufficient funding to maintain schools in a state of good repair [@auditorgeneralofontarioAnnualReport20152015], the operational savings as a case for school closures is a long-standing critique of the school "funding formula" established by the conservative provincial government in 1998 [@mackenzieBlueprintFixOntario2018; @irwinSchoolClosure2012].
+With the backdrop of budget cuts, the funding formula pits financing of school maintenance and student needs against the financial compensation of staff and powers of local school boards as they depend on the province for the allocation of resources [@mackenzieBlueprintFixOntario2018].
+
+In the wave of school accommodation reviews between 2013 and 2016, `r ALL_SCHOOLS_Elem %>% filter(Year == "2011" & Level == "Elementary" & Status == "Removed") %>% pull(SchoolID) %>% length()` elementary schools closed.
+These closures represented a `r ((ALL_SCHOOLS_Elem %>% filter(Year != "2016") %>% pull(SchoolID) %>% length() - ALL_SCHOOLS_Elem %>% filter(Year != "2011") %>% pull(SchoolID) %>% length())/ALL_SCHOOLS_Elem %>% filter(Year != "2016" & Level == "Elementary") %>% pull(SchoolID) %>% length()) %>% scales::percent()` and `r  (( ((avail_2011 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2011)) |> select(OTGC) |> sum())) - (avail_2016 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2016)) |> select(OTGC) |> sum())) / (avail_2016 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2016)) |> select(OTGC) |> sum()) ) |> scales::percent()` decline in elementary school locations and capacity between 2011 (before the school closures) and 2016, despite the fact that the overall student population in the city did not decline, and in fact increased slightly between 2011 (`r avail_2011$kid_popadj %>% sum(na.rm=T) %>% prettyNum(big.mark=",")` students) and 2016 (`r avail_2016$kid_popadj %>% sum(na.rm=T) %>% prettyNum(big.mark=",")` students).
 
@@
 In this paper, we hypothesize that the school consolidation policy reduced accessibility to schools for many students.
-By extension, this loss in access resulted in a displacement of trips to school made by active modes that increased transportation-related emissions.
+By extension, this loss in access and the consequent displacement of active trips to school increased transportation-related emissions.
 To test these hypotheses, the paper examines the following: how does the (1) spatial availability, (2) loss in walk minutes and (3) associated carbon emissions change for home-to-school trips after a city closes and consolidates schools?
 Further, this work examines (4) what communities, based on low-income prevalence, experienced the greatest losses as a result of this policy.
-Spatial availability, a singly-constrained competitive accessibility measure [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024], is used to quantify the before- and after- accessibility and associated impacts for the city of Hamilton and the elementary school-aged student population as a result of the school closure and consolidation policy.
-The methods introduced in this paper can be used by decision-makers to evaluate the spatial and transport-related impacts of public service provision policy.
-Furthermore, to increase the diffusion of this analysis, all work is completed in a R environment and all associated code and data will be open and reproducible within the lead author's [GitHub repository](https://github.com/soukhova)
-
-# Data and Methods
-
-Our focus is to evaluate the impact that the school consolidations between 2011 and 2016 had on the multimodal school spatial availability and travel-related GHG emission landscape of the city of Hamilton.
+
+Spatial availability, a singly-constrained measure of spatial accessibility [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024], is used to quantify accessibility before- and after- school closures, and the associated impacts for the city of Hamilton and the elementary school-aged student population as a result of a policy with narrowly defined benefits.
+The methods used in this paper can be applied by decision-makers to evaluate the spatial and transport-related impacts of public service provision policy.
+Furthermore, to increase the diffusion of this analysis, all work is transparent and reproducible, following best practices in the spatial sciences [@brunsdonOpeningPractice2021; @paezOpenSpatial2021c].
+Accordingly, all associated code and data will be publicly available within the lead author's [GitHub repository](https://github.com/soukhova)
+
+# Data
+
+Our focus is to evaluate the impact of school consolidations between 2011 and 2016 on multi-modal school spatial availability and travel-related GHG emission in Hamilton.
 To facilitate comparison, two scenarios are generated.
-The first and second reflect estimated as-is conditions in 2011 and 2016 assuming the respective school configurations, average student population and low-income prevalence of households, residential parcel locations, and estimated modal travel times in both years.
-The difference between these scenarios are quantified in terms of accessibility, emission and equity implications and discussed in context with the school consolidation policy.
-Data is retrieved at different spatial granularity from the following six data sources.
+Each scenario reflects the conditions in 2011 and 2016 respectively, assuming the school configurations in each year, as well as average student population and low-income prevalence of households, residential parcel locations, and estimated modal travel times.
+The difference between the two scenarios is quantified in terms of accessibility, emissions and equity implications and discussed in context with the school consolidation policy.
+Data were retrieved at different levels of spatial granularity from the following sources.
 
@@
 
-First, schools and associated school catchments are retrieved from the City's school boards.
+First, schools and associated school catchments were retrieved from the City's school boards.
 In Hamilton, the majority of student-aged population attends a school in one of the two English public school boards.
@@
 These catchments indicate what residential property is assigned to what school, by default.
-Families can decide to attend schools out of catchment, in fact 21%-23% <!--calculated in 03-travel-times-TTS.qmd--> of motorized school trips are out of catchment according to the regional travel survey, the Transportation Tomorrow Survey (TTS) in 2011 and 2016 [@data_management_group_tts_2018].
-The elementary school locations and catchments are provided by the HWDSB and HWCDSB for the 2010-2011 and 2015-2016 academic years and visualised in @fig-Fig1.
+Families can decide to attend schools out of catchment; in fact, 21%-23% <!--calculated in 03-travel-times-TTS.qmd--> of motorized school trips are out of catchment according to the regional travel survey, the Transportation Tomorrow Survey (TTS) in 2011 and 2016 [@data_management_group_tts_2018].
+The elementary school locations and catchments were provided by the HWDSB and HWCDSB for the 2010-2011 and 2015-2016 academic years, and are shown in @fig-Fig1.
 
@@
 As such, elementary includes middle schools that only instruct grades 6 to 8 (*Mid*), primary schools that only instruct kindergarten to grade 5 or 6 (*JrElem*), and all grade elementary schools (*Elem*) which instruct all grades from kindergarten to grade 8.
-The maximum number of students that may be enrolled in a school is calculated by the provincial Ministry of Education; this value is referred to as the on-the-ground capacity (*OTGC*) and is assigned to each school.
-The current/historic OTGC province-wide is not publicly available [@ontarioSchoolFacilityInventory2017].
-However, for schools that underwent the full Accommodation Review process, the OTGC for a given year can be obtained from publicly available School Information Profiles.
-A regression model was derived to estimate the OTGC for schools with no capacity information based on 1) the school's building footprint (*F*) in units of ($m^2$), 2) the school's shortest Euclidean distance from the centroid of Hamilton's central business district area (*DistCBD*) defined in units of (m), and 3) coefficients associated with a binary variable indicating the type of school grade instruction (*Mid*, *JrElem*, *Elem*).
+The maximum number of students that may be enrolled in a school is calculated by the provincial Ministry of Education; this value is referred to as the on-the-ground capacity (*OTGC*) and is assigned to each school <!-- [AP]: what are the units of OTGC?
+Students?
+-->.
+The current/historic OTGCs are not publicly available [@ontarioSchoolFacilityInventory2017].
+However, for schools that underwent the full Accommodation Review process, OTGCs for a given year could be obtained from publicly available School Information Profiles.
+A regression model was used to estimate the OTGCs for schools that lacked capacity information.
+The independent variables in the model were 1) the school's building footprint (*F*) in units of ($m^2$); 2) the school's shortest Euclidean distance from the centroid of Hamilton's central business district area (*DistCBD*) defined in units of (m); and 3) coefficients associated with a binary variable indicating the type of school grade instruction (*Mid*, *JrElem*, *Elem*).
 The building footprint from an archived spatial data set [@DMTI_Spatial_2015] and the 2016 footprints were retrieved from OSM [@OpenStreetMap_2021].
-Because the age of construction for different schools is not available, we use each school's distance from the centroid of the Downtown Business Improvement Area [@Hamilton_2014] as a proxy for school age as the oldest buildings are generally located closer to the city centre [@Merrall_2021].
-
-Equation (\ref{eq:OTGC-public}) and (\ref{eq:OTGC-catholic}) describes the estimated OTGC for public schools and catholic schools respectively.
+Because the age of construction for different schools is not available, we used each school's distance from the centroid of the Downtown Business Improvement Area [@Hamilton_2014] as a proxy for school age as the oldest buildings are generally located closer to the city centre [@Merrall_2021].
+
+Equations (\ref{eq:OTGC-public}) and (\ref{eq:OTGC-catholic}) describe the estimated OTGC for public schools and catholic schools respectively.
 School status and OTGC summed by catchment is also visualised in @fig-Fig1.
-See the Table <!--\ref{TabA1-OTGC}--> in the Appendix for the associated regression results.
-<!-- shall we explain further how we valuated this model?
+As seen in Table <!--\ref{TabA1-OTGC}--> in the Appendix, the coefficient of determination for these models was $R^2 = $ and $R^2 = $ for the Public and Catholic Boards respectively, and the standard deviation was $\sigma = $ and $sigma = $.
+
+<!-- [AS]: shall we explain further how we valuated this model?
 I don't think its entirely necessary.
 But the code is here: https://github.com/soukhova/hamONdests/blob/master/data-raw/01-school-data-preparation.Rmd (Antonio is a collaborator) -->
+
+<!-- [AP]: Just summarize the results here: what was the R2 and standard deviations of these models? -->
 
@@
 ```
+
 ```{=tex}
@@
 #| label: fig-Fig1
-#| fig-cap: "The school catchments, estimated on the ground capacities of schools summed per catchment, and school status for both public and public catholic school boards for year 2011 and 2016.
-OTGC is presented by quartiles."
+#| fig-cap: "The school catchments, estimated on the ground capacities (OTGCs) of schools summed per catchment, and school status for both public and public catholic school boards for year 2011 and 2016.
+OTGCs is presented by quartiles."
 #| echo: false
+#| out-width: "140%"
 
@@
 
+<!-- Number of schools?
+Yes Also, make the symbols that map the schools bigger, and use different shapes in addition to different colors.
+As an alternative, the OTGC could be the color, and the school category the shape.
+-->
+
 ## Students, residential locations and low-income prevelance
-Second, student population and proportion of household low-income after tax (LIM-AT) prevalence is retrieved from the 2011 and 2016 Canadian Census [@Statistics_Canada_2011; @Statistics_Canada_2016].
+
+In addition to OTGC for schools, statistics about the student population and proportion of low-income after tax (LIM-AT) housholds were retrieved from the 2011 and 2016 Canadian Census [@Statistics_Canada_2011; @Statistics_Canada_2016] using the {cancensus} `R` package [@von_Bergmann_Shkolnik_Jacobs_2021].
 The census releases population data by age group category in 5-year increments.
@@
 LIM-AT is refers to the proportion of private households that are below the median after-tax income [@governmentofcanadaDictionaryCensusPopulation2017].
-These census variables are taken at the Dissemination Area (DA) level, which is the finest geographic unit for which census data is made publicly available.
-Census data is retrieved using the {cancensus} R package [@von_Bergmann_Shkolnik_Jacobs_2021].
+These census variables are taken at the Dissemination Area (DA) level, which is the finest geographic unit for which census data are made publicly available.
 DAs are designed by Statistics Canada with the aim of population uniformity hence DAs greatly vary in area but represent between approximately 400 and 700 (1st and 3rd quartile) in total population.
@@
 
-The third data source consists of centroids of residential parcels in 2011 and 2016 that are used to represent residential locations [@Teranet_2009].
-There are 134,340 and 139,467 residential locations in 2011 and 2016 respectively, representing `r cen_data_2011_hamiltononly$DA_households |> sum() |> prettyNum(big.mark = ",")` and `r cen_data_2016_hamiltononly$DA_households |> sum() |> prettyNum(big.mark = ",")` households (according to the Canadian Census [@Statistics_Canada_2011; @Statistics_Canada_2016]).
-As the number of children per residential location are publicly unavailable, the average number of 5 to 14 year olds for each DA is divided by the number of residential households in that DA.
-Then the potential number of children at each residential location is calculated by multiplying this DA rate.
+The third source of data consists of centroids of residential parcels in 2011 and 2016 that are used to represent residential locations [@Teranet_2009].
+There are 134,340 and 139,467 residential locations in 2011 and 2016 respectively, representing `r cen_data_2011_hamiltononly$DA_households |> sum() |> prettyNum(big.mark = ",")` and `r cen_data_2016_hamiltononly$DA_households |> sum() |> prettyNum(big.mark = ",")` dwellings (according to the Canadian Census [@Statistics_Canada_2011; @Statistics_Canada_2016]).
+As the number of children per parcel are publicly unavailable, the average number of 5 to 14 year olds for each DA is divided by the number of residential households in that DA.
+Then the potential number of children at each dwelling is calculated by multiplying this DA rate by the number of dwellings in each parcel.
 <!-- NOTE: We consider the parcel centroids instead of DA centroids to more precisely calc.
@@
 #| echo: false
+#| out-width: "140%"
 
@@
 
-It is assumed that children can go to any elementary school, but certain schools are more likely to be attended depending on their proximity to their residential location.
-Based on this assumption, the fourth data source is empirical data regarding the mode-used and origin-destination locations of home-to-school trips from the 2011 and 2016 TTS [@data_management_group_tts_2018].
+It is assumed that children can go to any elementary school, but there is a preference for facilities that are closer.
+Based on this assumption, the fourth source of data is information regarding the mode-used and origin-destination locations of home-to-school trips from the 2011 and 2016 TTS [@data_management_group_tts_2018].
 The TTS is a travel survey conducted in the Greater Golden Horseshoe Area in Ontario every 5 years.
@@
 Hamilton (Ancaster `r community_car_perc_2011$TAZOrig_car_perc[1] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[1] |> percent()`, Dundas `r community_car_perc_2011$TAZOrig_car_perc[2] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[2] |> percent()`, Flamborough `r community_car_perc_2011$TAZOrig_car_perc[3] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[3] |> percent()`, Glanbrook `r community_car_perc_2011$TAZOrig_car_perc[4] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[4] |> percent()`, Hamilton Central `r community_car_perc_2011$TAZOrig_car_perc[5] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[5] |> percent()`, and Stoney Creek `r community_car_perc_2011$TAZOrig_car_perc[6] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[6] |> percent()`).
-<!--Do we maybe need to discuss how this lack of data for each TAZ impacted the calculations?-->
+<!--Do we maybe need to discuss how this lack of data for each TAZ impacted the calculations?
+[AP]: No.
+Unrelated: did you use information from TTS2016R?
+Or can you add it to the data package?
+And then cite your data paper.-->
 
@@
 
-The fifth data source consists of estimated travel time matrices from all residential parcel locations to all elementary schools for both motorized (car) and non-motorized (walking) travel.
+The fifth source of data consists of estimated travel time matrices from all residential parcel locations to all elementary schools for both motorized (car) and non-motorized (walking) travel.
 The travel times are estimated using the {r5r} package [@Pereira2021r5r].
 For both modes, the inputs assume a maximum travel time threshold of 60 minutes and the free-flow OpenStreetMap road network of Hamilton (retrieved using Geofabrik [@geofabrikOntarioOpenStreetMapGeofabrik2022] and manually edited to only include the Hamilton boundary).
-Further, walk travel times that are only less than 27 min minutes were retained.
-This travel time is approximately equivalent to 1.6 km which is the minimum distance that is required from a school within-catchment to qualify for a school bus [@HWDSB_2019].
+Further, only walking trips shorter than 27 min minutes were retained.
+This travel time is approximately equivalent to 1.6 km which is the point at which within-catchment students qualify for transport provided by the school board [@HWDSB_2019].
 Estimated trip-weighted motorized travel times are on average 16.5 minutes in 2011 and 16.8 minutes in 2016 (summary statistics for both years are approximately min: 0.5, Q1: 11.0, Q2: 16.0, Q3: 21.0, max: 60.0).
@@
 
-These travel time matrices along with the origin-destination flows are used to create modal trip length distribution (TLD) functions.
-A TLD is a useful technique to calibrate impedance functions as they are a representation of the likelihood that a proportion of trips are taken at a specific travel cost [@horbachov_theoretical_2018; @batista_estimation_2019].
-The empirical TLD is fit to a density distribution using maximum likelihood and moment matching techniques and the Nelder-Mead and Brent methods for direct optimization available within the {fitdistrplus} R package [@fitdistrplus_2015].
+These travel time matrices along with the origin-destination flows were used to create modal trip length distribution (TLD) functions.
+Analysis of the TLD is a useful technique to calibrate impedance functions as they are a representation of the propensity of travel by trip length [@horbachov_theoretical_2018; @batista_estimation_2019].
+The empirical TLD is used to fit a density distribution using maximum likelihood and moment matching techniques and the Nelder-Mead and Brent methods for direct optimization available within the {fitdistrplus} `R` package [@fitdistrplus_2015].
 Based on goodness-of-fit criteria and diagnostics <!-- should I include the diagonostic plots in the appendix?
 q-q, etc..?
-I don't think its entirely necessary as the empirical is graphed over the theoretical line in the following figure-->, the gamma and exponential distributions are selected for the motorized and non-motorized modal distributions respectively.
+I don't think its entirely necessary as the empirical is graphed over the theoretical line in the following figure.
+[AP] Agreed-->, the gamma and exponential distributions were selected for the motorized and non-motorized modal distributions respectively.
 The gamma distribution is defined by the shape ($\alpha$) parameter of `r round(gamma2011_motor$estimate[1], 3)` (2011) and `r round(gamma2016_motor$estimate[1], 3)` (2016) and the rate ($\beta$) of `r round(gamma2011_motor$estimate[2], 3)` (2011) and `r round(gamma2016_motor$estimate[2], 3)` (2016).
@@
 
-## Putting it all together: multimodal spatial availability
-
-Accessibility is defined as "access to opportunities" or the "potential for spatial interaction".
+# Methods: Multi-modal spatial availability
+
+Accessibility is defined as the "potential for spatial interaction".
 It is classically presented as the gravity-based measure popularized by @hansenHowAccessibilityShapes1959 and takes the following general multi-modal formulation (Equation (\ref{eq:multi-modal-conventional-accessibility})):
@@
 
-In this paper, we benefit from new developments in accessibility research, particularly the multimodal spatial availability measure [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024].
+In this paper, we benefit from new developments in accessibility research, particularly the multi-modal spatial availability measure [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024].
 Spatial availability is a singly-constrained accessibility measure that accounts for competition by students using different modes for exclusive opportunities, such as school seats.
@@
 This proportional allocation of opportunities yields an interpretable and meaningful measure of opportunity access, particularly when comparing across modes, at the spatial resolution of a residential parcel, and multiple time periods.
-See @soukhovMultimodalSpatialAvailability2024 for further discussion, multimodal spatial availability $V_{i}^m$ is defined as given by Equation (\ref{eq:spatial-availability}).
+See @soukhovMultimodalSpatialAvailability2024 for further discussion, multi-modal spatial availability $V_{i}^m$ is defined as given by Equation (\ref{eq:spatial-availability}).
 
@@
 Hansen-type accessibility is not designed to preserve the number of opportunities in the region, it simply counts the intensity of opportunities that those in a zone can potentially interact with (weighted by the friction of distance).
-Also, as discussed in @soukhovIntroducingSpatialAvailability2023, popular competitive accessibility measures such as the 2SFCA [@shenLocationCharacteristicsInnercity1998;@luoMeasuresSpatialAccessibility2003] end up preserving the number of opportunities in the analysis region but the definitions of variables are internally obscured; the only way it preserves the number of opportunities is if the effect of the impedance function is ignored when expanding the values of opportunities per capita to obtain the total number of opportunities.
-On the other hand, the proportional allocation procedure associated with calculating multimodal spatial availability $V_i^m$ consistently returns a number of opportunities available to all mode-using populations that matches the total number of opportunities in the region when summed.
-As a result of this consistency, it is possible to define a measure of multimodal spatial availability per capita ($P$) as presented in Equation (\ref{eq:SA-per-capita}) for use as a benchmark to compare against the regional opportunities per capita ($\frac{\sum_{j} O_j}{\sum_{i} P_i}$).
+Also, as discussed in @soukhovIntroducingSpatialAvailability2023, popular competitive accessibility measures such as the 2SFCA [@shenLocationCharacteristicsInnercity1998; @luoMeasuresSpatialAccessibility2003] are internally inconsistent, and the only way it preserves the number of opportunities is if the effect of the impedance function is ignored when expanding the values of opportunities per capita to obtain the total number of opportunities.
+On the other hand, the proportional allocation procedure associated with calculating multi-modal spatial availability $V_i^m$ consistently returns a number of opportunities available to populations by mode that matches the total number of opportunities in the region when summed.
+By doing this consistently, it is possible to define a measure of multi-modal spatial availability per capita ($P$) as presented in Equation (\ref{eq:SA-per-capita}) for use as a benchmark to compare against the regional opportunities per capita ($\frac{\sum_{j} O_j}{\sum_{i} P_i}$).
 
@@
 
-For the reasons outlined, this paper uses the multimodal spatial availability $V_i^m$ and spatial availability per student $v_i^m$ measures to quantify accessibility changes for the following reasons.
+This paper uses the multi-modal spatial availability $V_i^m$ and spatial availability per student $v_i^m$ measures to quantify accessibility changes for the following reasons.
 First, all residential locations are associated to their respective census DA, TTS TAZ, and community boundary (@fig-Fig3) based on spatial location.
-Each residential location is assigned a number of 'potential' motorized and non-motorized student aged population from these different data sources to calculate a motorized and non-motorized population balancing factor $F_{i}^{pm}$.
-For each residential location the following is retained: the 2011 and 2016 census average elementary-aged student population per residential location (@fig-Fig2 second row) and the average motorized and non-motorized split as informed by the TTS aggregated by community (@fig-Fig3).
+Each parcel is assigned a number of 'potential' motorized and non-motorized student aged population from these different data sources to calculate a motorized and non-motorized population balancing factor $F_{i}^{pm}$.
+For each parcel the following is retained: the 2011 and 2016 census average elementary-aged student population per parcel (@fig-Fig2 second row) and the average motorized and non-motorized split as informed by the TTS aggregated by community (@fig-Fig3).
 Second, the school capacity $O_j$ is estimated and associated with each school (@fig-Fig1).
@@
 
-Third, the motorized and non-motorized impedance-balancing factors $F_{ij}^{cm}$ are calculated using the modal theoretical TLDs (@fig-Fig4) based on the respective travel times for each origin (residential location) to school combination.
+Third, the motorized and non-motorized impedance-balancing factors $F_{ij}^{cm}$ are calculated using a mode-specific impedance function (@fig-Fig4) based on the respective travel times for each origin (parcel) to school combination.
 Finally, outputs from all three stages are joined together.
-$F_{i}^{pm}$ joined with $F_{ij}^{cm}$ yields the total balancing factor $F^{tm}_{ij}$ which serves to proportionally allocate the school capacity $O_j$ to each residential location.
-The value is multimodal spatial availability $V_i^m$ which can be interpreted as the number of school-seats that are available to the mode-using population at that residential location.
+$F_{i}^{pm}$ joined with $F_{ij}^{cm}$ yields the total balancing factor $F^{tm}_{ij}$ which serves to proportionally allocate the school capacity $O_j$ to each parcel.
+The value is multi-modal spatial availability $V_i^m$ which can be interpreted as the number of school-seats that are available to the mode-using population at that parcel.
 $V_i^m$ is then summed and represented at the DA and community level along with the calculated $v_i^m$ for interpretation.
@@
 #| label: fig-Fig6
-#| fig-cap: "Multimodal spatial availability (greys) and spatial availability per mode-using student (diverging reds and greens) aggregated at the level of DA for 2011 and 2016."
+#| fig-cap: "Multi-modal spatial availability (greys) and spatial availability per mode-using student (diverging reds and greens) aggregated at the level of DA for 2011 and 2016."
 #| echo: false
-
+#| out-width: "140%"
+#| 
 knitr::include_graphics(paste0(here::here(),"/Figures/Vim_vvim_20112016_plot.png"))
 ```
+
 <!--I'll make this plot longer-->
-The motorized and non-motorized spatial availability $V_i$ for both 2011 and 2016 is represented in the top half of @fig-Fig6.
-Though at different magnitudes, both populations using non- and motorized modes have higher spatial availability when more proximate to schools.
+
+The motorized and non-motorized spatial availabilities of school seats $V_i$ for both 2011 and 2016 are presented in the top half of @fig-Fig6.
+Though they show different magnitudes, both populations using non- and motorized modes have higher spatial availability when they are more proximate to schools.
 Hence, all DAs have higher values within Hamilton Central and those more proximate to schools in less-urban communities.
-Trends appear similar for both 2011 and 2016: proximate populations using motorized modes benefit from a wider-range in high spatial availability than non-motorized modes.
-When $V_i^m$ is divided by the number of mode-using students in that DA, $v_i^m$ is derived (bottom half of @fig-Fig6).
+Trends appear similar for both 2011 and 2016: proximate populations using motorized modes benefit from the ability to reach more distant destinations than non-motorized modes.
+When $V_i^m$ is divided by the number of students by mode in that DA, $v_i^m$ is derived (bottom half of @fig-Fig6).
 A similar spatial distribution seen in $V_i^m$ can be seen in $v_i^m$ plots, but as a rate of student population.
@@
 #| label: fig-Fig7
-#| fig-cap: "Change in multimodal spatial availability and spatial availability per mode-using student between 2016 and 2011 at the level of DA of the 2016 Canadian Census."
+#| fig-cap: "Change in multi-modal spatial availability and spatial availability per mode-using student between 2016 and 2011 at the level of DA of the 2016 Canadian Census."
 #| echo: false
@@
 ```
+
 <!--The colours combos I choose here are absolutely ugly ;( but besides that, do you agree that V_im and v_im should use different colour palettes?-->
@@
 However, some motorized-mode using populations in DAs do experience an increase in $V_i^m$: these DAs are often relatively proximate to schools but more immediately outside walking-range.
-Under a competitive and constraint-based calculation, these DAs benefit from the losses in spatial availability that are predominately experienced by the non-motorized populations within Hamilton Central.
+Under competitive and constrained calculations, these DAs benefit from the losses in spatial availability that are predominately experienced by the non-motorized populations within Hamilton Central.
 
@@
 This will provide us some higher level summary stats.
-Moataz asked about this as well regarding the figures in the Results section-->
+Moataz asked about this as well regarding the figures in the Results section.
+[AP] Yes, please add those statistics to support the discussion.
+-->
 
 <!--Antonio, what do you think, add a table with summaries of LIMAT, Kid pop, V_im change, v_im change, total travel time change, and emission changed by community and mode?
-We should discuss to decide what's the best presentation.-->
+We should discuss to decide what's the best presentation.
+[AP]: Best presentation: spatial availability by quartiles of the relevant dimensions (e.g., LIMAT) -->
+
 ```{r}
@@
 Though the population between 2011 and 2016 decreased, the rate of school-seat spatial availability decreased disproportionately more (population by `r SAvail_im_COMM_changes |> st_drop_geometry() |> filter(COMMUNITY == "HA") |> transmute(calc = ((sum(kid_2016)-sum(kid_2011))/sum(kid_2011)) |> scales::percent()) |> first()` and spatial availability by `r `r SAvail_im_COMM_changes |> st_drop_geometry() |> filter(COMMUNITY == "HA") |> transmute(calc = ((sum(V_im_2016)-sum(V_im_2011))/sum(V_im_2011)) |> scales::percent())  |> first()`).
-This out of proportion decrease is greatest in Hamilton Central than any other community.
+This disproportionate decrease is greatest in Hamilton Central than any other community.
 This may have equity concerns as Hamilton Central has the highest LIM-AT prevalence, with `r SAvail_im_COMM_changes |> st_drop_geometry() |> filter(COMMUNITY == "HA" & mode == "motorized") |> dplyr::select("prev_LIMAT_2016") |> round() |> paste0("%")` of households LIM-AT in 2016, 3.7 times greater than the community with the lowest level of LIM-AT prevalence in the city.
@@
 
-<!-- Is this worth further developing or perhaps just remove, what do you think? I think further. -->
+<!-- Is this worth further developing or perhaps just remove, what do you think?
+I think further.
+[AP]: tables to show availability by mode in 2011 and 2016 with changes, as well as lost minutes of active travel time and emissions.
+Check Elise's equity in accessibility to bike share paper, see table 1 there.
+-->
 Study limitations:
@@
 
-In this paper, we estimate the change in multimodal spatial availability as a result of school consolidations/closures between the years of 2011 and 2016 in an empirical case study of Hamilton, a mid-size city (\~550,000 pop) in Ontario, Canada.
+In this paper, we estimate the change in multi-modal spatial availability as a result of school consolidations/closures between the years of 2011 and 2016 in an empirical case study of Hamilton, a mid-size city (\~550,000 pop) in Ontario, Canada.
 The spatial availability and spatial availability per student are visualized on a DA level; spatial availability for the city declined overall.

### 2024-04-29 data-prep/01-processing-schools-parcels-census.qmd
Draft up until Results done

@@
 ```{r}
-RED_LU_centroids_2011 <- sp::over(as_Spatial(st_centroid(RED_LU_2011)),as_Spatial(cen_data_2011_1 |> select(GeoUID)))
-RED_LU_centroids_2011<- cbind(RED_LU_centroids_2011,st_centroid(RED_LU_2011))
+RED_LU_centroids_2011 <- sp::over(as_Spatial(st_centroid(RED_LU_2011)),as_Spatial(cen_data_2011_1 |> dplyr::select(GeoUID)))
+
+RED_LU_centroids_2011_w2016ID <- sp::over(as_Spatial(st_centroid(RED_LU_2011)),as_Spatial(cen_data_2016_1 |> dplyr::select(GeoUID) |> rename("GeoUID_2016" = "GeoUID")))
+
+RED_LU_centroids_2011 <- cbind(RED_LU_centroids_2011_w2016ID,
+                              RED_LU_centroids_2011,
+                              st_centroid(RED_LU_2011))
 ```
+
 ```{r}
@@
 ```{r}
-ggh_taz <- st_read("data-inputs/TTS/tts06_83_region.shp") |> st_transform(st_crs(cen_data_2011_hamiltononly))
+ggh_taz <- st_read("data-inputs/TTS/tts06_83_region.shp") |> st_transform(st_crs(cen_data_2011_1))
 ```

### 2024-04-29 data-prep/04-Creating-avail_2011-16.Rmd
Draft up until Results done

@@
 #now let's add the communtiy names to the TAZ and summarize the % of students who use what mode by community. 
+#NOTE: ham_taz is from 03-travel-times-TTS.qmd... add
 ggh_taz_communities <- st_intersection(st_centroid(ham_taz) |> st_transform(st_crs(ggh_taz)), 

### 2024-04-29 paper/Manuscript.qmd
Draft up until Results done

@@
 ```{r, include=FALSE}
+#boundaries for mapping
+hamilton_communities <- st_read(paste0(here::here(),"/data-prep/data-inputs/Boundaries/Community_Boundaries.shp")) |> mutate(COMMUNITY_ = ifelse(COMMUNITY_ == "Hamilton", "Hamilton-Central", COMMUNITY_))
+
+ham_bound <- st_read(paste0(here::here(),"/data-prep/data-inputs/Boundaries/City_Boundary.shp"))
+
+hydro_p_LakeOntario <- st_read(paste0(here::here(),"/data-prep/data-inputs/Boundaries/hydro_p_LakeOntario.shp")) |> st_transform(crs=4326)
+
+ham_bay <- st_read(paste0(here::here(),"/data-prep/data-inputs/Boundaries/Waterbodies.shp")) |> st_transform(crs=4326) |> filter(FEATURE_TY == "Lake")
+
+ham_streets <- st_read(paste0(here::here(),"/data-prep/data-inputs/Boundaries/Street_Centreline.shp")) |> st_transform(crs=4326) |> filter(ROAD_CLASS %in% c("Provincial Highway", "Parkway", "Arterial")) #no collectors, arterial, Local
+```
+
+```{r, include=FALSE}
 load(file = paste0(here::here(),"/data-prep/data-products/cen_data_2011_hamiltononly.Rdata"))
@@
 catchment_schools_plot <-
+   tm_shape(ham_bay, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0) +
+  tm_shape(hydro_p_LakeOntario, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0)+
   tm_shape(ALL_SCHOOLS_CATCH_Elem_extra) +
-  tm_polygons(col = "OTGC", border.alpha=0.1, palette = "Greys",
+  tm_polygons(col = "OTGC", border.alpha=0.1, palette = "RdPu",
               labels = c("[181-318)", "[318-406)", "[406-566)","[566-1417]"),
@@
   tm_shape(ALL_SCHOOLS_Elem_formapping) +
-  tm_dots(col = "Status", size=0.1, title = "School status between\n2011 and 2016",
-          palette = c("#ffd14d", "black", "darkgreen","#0E86D4","#90EE90"),
-          labels = c("No change","Closed before 2016", "Expanded before 2016","Moved before 2016","New after 2011")          
+  tm_symbols(shape = "Status",
+             size=0.15,
+          title.shape = "School status",
+          shapes = c(24,25,22,21,23),
+          shapes.labels = c(
+          "Expanded after 2011",
+          "Moved before 2016",
+          "New after 2011",
+          "No change btwn 2011-16",
+          "Closed before 2016")
   ) +
-  tm_facets(by=c("Year","System"))+
+  tm_facets(by=c("System","Year"))+
+  tm_credits(c("52 schools\n(18,771 OTGC)","49 schools\n(18,039 OTGC)",
+                "95 schools\n(41,796 OTGC)","88 schools\n(40,249 OTGC)"),
+             position = c("LEFT", "BOTTOM")) +
   tmap_defaults +
-  tm_layout(panel.labels = list(c('HWDCSB (Catholic Public)','HWDSB (Public)'),c('2011','2016')))
-
+  tm_layout(bg.color = "grey",
+            panel.labels = list(c('HWDCSB (Catholic Public)','HWDSB (Public)'),
+                                c('2011','2016')),
+            panel.label.bg.color = "white")
+
+catchment_schools_plot
+```
+
+```{r descriptive-catchment-schools-plot2, eval=FALSE}
 tmap_save(catchment_schools_plot, paste0(here::here(),"/Figures/catchment_schools_plot.png"), height=7, width=12, dpi=600)
 ```
+
+<!-- Number of schools?
+Yes Also, make the symbols that map the schools bigger, and use different shapes in addition to different colors.
+As an alternative, the OTGC could be the color, and the school category the shape.
+-->
 
@@
 ```{r}
+#calc spatial availability and other associated variables at the DA 2011 level
 SAvail_im_DA2011 <- avail_2011 %>%
@@
 
+#calc spatial availability and other associated variables at the DA 2016 level
 SAvail_im_DA2016 <- avail_2016 %>%
@@
   left_join(cen_data_2016_hamiltononly |> filter(Region.Name == "Hamilton") |> dplyr::select(GeoUID), by="GeoUID") |> st_set_geometry("geometry")
-```
-```{r}
-#Create 1 sf with data joined, no changes
+
+#Create 1 sf with data joined (no changes, just 2011 and 2016)
 SAvail_im_DA <-
@@
                           mode == "w" ~ "non-motorized"))
-
-#calc changes between the years - make that 1 sf
-SAvail_im_DA_changes <- full_join(SAvail_im_DA2011|>st_drop_geometry(),
-                                  SAvail_im_DA2016, by=c("GeoUID", "mode")) |>
-  transmute(GeoUID,
+```
+
+```{r}
+#calc changes between the years - but at the 2016 DA level 
+SAvail_im_DA2011_forGeoUID2016 <- avail_2011 %>%
+  group_by(GeoUID_2016, 
+           mode) %>%
+  summarize(COMMUNITY = first(COMMUNITY),
+            total_tt = sum(travel_time_p50*kid_popadj),
+            V_im = sum(V_ij),
+            kid_popadj_m = sum(kid_popadj),
+            DA_pop_5to14 = mean(DA_pop_5to14, na.rm=T),
+            DA_prev_LIMAT_under18 = mean(DA_prev_LIMAT_under18),
+            .groups = "drop") |>
+  mutate(emissions_total_car_tt = ifelse(mode == "c", 0.1*total_tt, 0),
+         v_im_kid_popadj = V_im/kid_popadj_m,
+         v_im_kid_popadj_LIMAT = v_im_kid_popadj*(DA_prev_LIMAT_under18/100),
+         v_im_kid_popadj_LIMAT = ifelse(is.na(v_im_kid_popadj_LIMAT) | is.nan(v_im_kid_popadj_LIMAT), 0, v_im_kid_popadj_LIMAT),
+         V_i_kid_popadj_LIMAT = V_im*(DA_prev_LIMAT_under18/100),
+         V_i_kid_popadj_LIMAT = ifelse(is.na(V_i_kid_popadj_LIMAT) | is.nan(V_i_kid_popadj_LIMAT), 0, V_i_kid_popadj_LIMAT))
+
+SAvail_im_DA_changes <- full_join(SAvail_im_DA2011_forGeoUID2016,
+                                  SAvail_im_DA2016, by=c("GeoUID_2016"="GeoUID", 
+                                                         "mode")) %>%
+  transmute(GeoUID_2016,
             mode = case_when(mode == "c" ~ "motorized",
@@
             c_total_tt = total_tt.y - total_tt.x, #2016 minus 2011 tt
+            c_total_ttperkid = (total_tt.y/kid_popadj_m.y) - (total_tt.x/kid_popadj_m.x),
+            c_V_im = V_im.y -V_im.x,
+            c_V_im_perc = (V_im.y -V_im.x)/V_im.x,
+            c_kid_popadj_m = kid_popadj_m.y - kid_popadj_m.x,
+            c_DA_pop_5to14 = DA_pop_5to14.y - DA_pop_5to14.x,
+            c_DA_prev_LIMAT_under18 = DA_prev_LIMAT_under18.y - DA_prev_LIMAT_under18.x,
+            c_emissions_total_car_tt = emissions_total_car_tt.y - emissions_total_car_tt.x,
+            c_v_im_kid_popadj = v_im_kid_popadj.y - v_im_kid_popadj.x,
+            c_v_im_kid_popadj_perc = (v_im_kid_popadj.y - v_im_kid_popadj.x)/v_im_kid_popadj.x,
+            c_v_im_kid_popadj_LIMAT = v_im_kid_popadj_LIMAT.y - v_im_kid_popadj_LIMAT.x,
+            c_V_i_kid_popadj_LIMAT = V_i_kid_popadj_LIMAT.y - V_i_kid_popadj_LIMAT.x) |>
+  left_join(cen_data_2016_hamiltononly|>dplyr::select(c("GeoUID")), 
+            by=c("GeoUID_2016" = "GeoUID")) %>% st_as_sf
+
+```
+
+```{r}
+#calc changes between the years - but between communities with geom for mapping
+ham_community_bounds <- ham_community_bounds |> 
+  mutate(COMMUNITY = case_when(COMMUNITY_ == "Ancaster" ~ "AN",
+                               COMMUNITY_ == "Dundas" ~ "DU",
+                               COMMUNITY_ == "Flamborough" ~ "FL",
+                               COMMUNITY_ == "Glanbrook" ~ "GL",
+                               COMMUNITY_ == "Hamilton" ~ "HA",
+                               COMMUNITY_ == "Stoney Creek" ~ "SC"))
+
+SAvail_im_COMM2011 <- avail_2011 %>% 
+  group_by(COMMUNITY, 
+           mode) %>%
+  summarize(total_tt = sum(travel_time_p50*kid_popadj),
+            V_im = sum(V_ij),
+            kid_popadj_m = sum(kid_popadj),
+            DA_pop_5to14 = mean(DA_pop_5to14, na.rm=T),
+            DA_prev_LIMAT_under18 = mean(DA_prev_LIMAT_under18),
+            .groups = "drop") |>
+  mutate(emissions_total_car_tt = ifelse(mode == "c", 0.1*total_tt, 0),
+         v_im_kid_popadj = V_im/kid_popadj_m,
+         v_im_kid_popadj_LIMAT = v_im_kid_popadj*(DA_prev_LIMAT_under18/100),
+         v_im_kid_popadj_LIMAT = ifelse(is.na(v_im_kid_popadj_LIMAT) | is.nan(v_im_kid_popadj_LIMAT), 0, v_im_kid_popadj_LIMAT),
+         V_i_kid_popadj_LIMAT = V_im*(DA_prev_LIMAT_under18/100),
+         V_i_kid_popadj_LIMAT = ifelse(is.na(V_i_kid_popadj_LIMAT) | is.nan(V_i_kid_popadj_LIMAT), 0, V_i_kid_popadj_LIMAT))
+
+SAvail_im_COMM2016 <- avail_2016 %>%
+  group_by(COMMUNITY, 
+           mode) %>%
+  summarize(total_tt = sum(travel_time_p50*kid_popadj),
+            V_im = sum(V_ij),
+            kid_popadj_m = sum(kid_popadj),
+            DA_pop_5to14 = mean(DA_pop_5to14, na.rm=T),
+            DA_prev_LIMAT_under18 = mean(DA_prev_LIMAT_under18),
+            .groups = "drop") |>
+  mutate(emissions_total_car_tt = ifelse(mode == "c", 0.1*total_tt, 0),
+         v_im_kid_popadj = V_im/kid_popadj_m,
+         v_im_kid_popadj = ifelse(is.na(v_im_kid_popadj) | is.nan(v_im_kid_popadj), 0, v_im_kid_popadj),
+         v_im_kid_popadj_LIMAT = v_im_kid_popadj*(DA_prev_LIMAT_under18/100),
+         v_im_kid_popadj_LIMAT = ifelse(is.na(v_im_kid_popadj_LIMAT) | is.nan(v_im_kid_popadj_LIMAT), 0, v_im_kid_popadj_LIMAT),
+         V_i_kid_popadj_LIMAT = V_im*(DA_prev_LIMAT_under18/100),
+         V_i_kid_popadj_LIMAT = ifelse(is.na(V_i_kid_popadj_LIMAT) | is.nan(V_i_kid_popadj_LIMAT), 0, V_i_kid_popadj_LIMAT))
+
+#add the geometries -- census DAs
+SAvail_im_COMM2011 <- SAvail_im_COMM2011 |>
+  left_join(ham_community_bounds |> dplyr::select(c("COMMUNITY_","COMMUNITY")), by=c("COMMUNITY"))|> st_set_geometry("geometry")
+
+SAvail_im_COMM2016 <- SAvail_im_COMM2016 |>
+  left_join(ham_community_bounds |> dplyr::select(c("COMMUNITY_","COMMUNITY")), by=c("COMMUNITY"))|> st_set_geometry("geometry")
+
+#calc changes between the years - make that 1 sf
+SAvail_im_COMM_changes <- full_join(SAvail_im_COMM2011|>st_drop_geometry(),
+                                  SAvail_im_COMM2016, by=c("COMMUNITY", "mode")) |>
+  transmute(COMMUNITY,
+            mode = case_when(mode == "c" ~ "motorized",
+                             mode == "w" ~ "non-motorized"),
+            c_total_tt = total_tt.y - total_tt.x, #2016 minus 2011 tt
+            c_total_ttperkid = (total_tt.y/kid_popadj_m.y) - (total_tt.x/kid_popadj_m.x),
             c_V_im = V_im.y -V_im.x,
@@
             c_V_i_kid_popadj_LIMAT = V_i_kid_popadj_LIMAT.y - V_i_kid_popadj_LIMAT.x,
-            geometry) |> st_as_sf()
-
-```
-
-```{r}
-#repeat but for all 1 mode
-SAvail_i_DA2011 <- avail_2011 %>%
-  group_by(GeoUID) %>%
-  summarize(COMMUNITY = first(COMMUNITY),
-            total_tt = sum(travel_time_p50*kid_popadj, na.rm = T),
-            V_i = sum(V_ij, na.rm = T),
-            DA_pop_5to14 = mean(DA_pop_5to14, na.rm=T),
-            kid_pop_DAavg = mean(kid_pop, na.rm = T),
-            DA_prev_LIMAT_under18 = mean(DA_prev_LIMAT_under18, na.rm = T),
-            .groups = "drop") |>
-  filter(GeoUID != 35250242 & GeoUID != 35250133 & GeoUID != 35250457 ) #these GeoUID's have 0 kid population, so let's drop so we don't have zeros
-
-SAvail_i_DA2016 <- avail_2016 %>%
-  group_by(GeoUID) %>%
-  summarize(COMMUNITY = first(COMMUNITY),
-            total_tt = sum(travel_time_p50*kid_popadj),
-            V_i = sum(V_ij, na.rm = T),
-            DA_pop_5to14 = mean(DA_pop_5to14, na.rm=T),
-            kid_pop_DAavg = mean(kid_pop, na.rm = T),
-            DA_prev_LIMAT_under18 = mean(DA_prev_LIMAT_under18, na.rm = T),
-            .groups = "drop") |>
-  filter(GeoUID != 35250398 & GeoUID != 35250133 & GeoUID != 35250427 ) #these GeoUID's have 0 kid population, so let's drop so we don't have zeros
-
-#add the geometries -- census DAs
-SAvail_i_DA2011 <- SAvail_i_DA2011 |>
-  left_join(cen_data_2011_hamiltononly |> filter(Region.Name == "Hamilton") |> dplyr::select(GeoUID), by="GeoUID") |> mutate(Year = "2011") |> st_set_geometry("geometry")
-
-SAvail_i_DA2016 <- SAvail_i_DA2016 |>
-  left_join(cen_data_2016_hamiltononly |> filter(Region.Name == "Hamilton") |> dplyr::select(GeoUID), by="GeoUID") |> mutate(Year = "2016") |> st_set_geometry("geometry")
-
-#join them together!
-SAvail_i_DA_extra <- rbind(SAvail_i_DA2011, SAvail_i_DA2016)
-rm(SAvail_i_DA2011,SAvail_i_DA2016)
-```
-
-```{r}
-#now do this at the community level
-ham_community_bounds <- ham_community_bounds |> 
-  mutate(COMMUNITY = case_when(COMMUNITY_ == "Ancaster" ~ "AN",
-                               COMMUNITY_ == "Dundas" ~ "DU",
-                               COMMUNITY_ == "Flamborough" ~ "FL",
-                               COMMUNITY_ == "Glanbrook" ~ "GL",
-                               COMMUNITY_ == "Hamilton" ~ "HA",
-                               COMMUNITY_ == "Stoney Creek" ~ "SC"))
-
-SAvail_im_COMM2011 <- SAvail_im_DA2011 %>% st_drop_geometry() |>
-  group_by(COMMUNITY, 
-           mode) %>%
-  summarize(total_tt = sum(total_tt, na.rm=T),
-            V_im = sum(V_im, na.rm=T),
-            kid_popadj_m = sum(kid_popadj_m, na.rm=T),
-            COMM_prev_LIMAT_under18 = mean(DA_prev_LIMAT_under18, na.rm=T),
-            emissions_total_car_tt = sum(emissions_total_car_tt, na.rm=T),
-            .groups = "drop") |>
-  mutate(v_im_kid_popadj = V_im/kid_popadj_m,
-         v_im_kid_popadj_LIMAT = v_im_kid_popadj*(COMM_prev_LIMAT_under18/100),
-         v_im_kid_popadj_LIMAT = ifelse(is.na(v_im_kid_popadj_LIMAT) | is.nan(v_im_kid_popadj_LIMAT), 0, v_im_kid_popadj_LIMAT),
-         V_i_kid_popadj_LIMAT = V_im*(COMM_prev_LIMAT_under18/100),
-         V_i_kid_popadj_LIMAT = ifelse(is.na(V_i_kid_popadj_LIMAT) | is.nan(V_i_kid_popadj_LIMAT), 0, V_i_kid_popadj_LIMAT))  |> left_join(ham_community_bounds |> dplyr::select(-c("OBJECTID")), by=c("COMMUNITY")) |> st_set_geometry("geometry")
-
-SAvail_im_COMM2016 <- SAvail_im_DA2016 %>% st_drop_geometry() |>
-  group_by(COMMUNITY, 
-           mode) %>%
-  summarize(total_tt = sum(total_tt, na.rm=T),
-            V_im = sum(V_im, na.rm=T),
-            kid_popadj_m = sum(kid_popadj_m, na.rm=T),
-            COMM_prev_LIMAT_under18 = mean(DA_prev_LIMAT_under18, na.rm=T),
-            emissions_total_car_tt = sum(emissions_total_car_tt, na.rm=T),
-            .groups = "drop") |>
-  mutate(v_im_kid_popadj = V_im/kid_popadj_m,
-         v_im_kid_popadj_LIMAT = v_im_kid_popadj*(COMM_prev_LIMAT_under18/100),
-         v_im_kid_popadj_LIMAT = ifelse(is.na(v_im_kid_popadj_LIMAT) | is.nan(v_im_kid_popadj_LIMAT), 0, v_im_kid_popadj_LIMAT),
-         V_im_kid_popadj_LIMAT = V_im*(COMM_prev_LIMAT_under18/100),
-         V_im_kid_popadj_LIMAT = ifelse(is.na(V_im_kid_popadj_LIMAT) | is.nan(V_im_kid_popadj_LIMAT), 0, V_im_kid_popadj_LIMAT))  |> left_join(ham_community_bounds |> dplyr::select(-c("OBJECTID")), by=c("COMMUNITY")) |> st_set_geometry("geometry")
-```
-
-```{r}
-#repeat above but with just i, this is to get the correct LIM value.
-SAvail_i_COMM2011 <- SAvail_im_DA2011 %>% st_drop_geometry() |>
-  group_by(COMMUNITY) %>%
-  summarize(COMM_prev_LIMAT_under18 = mean(DA_prev_LIMAT_under18, na.rm=T),
-            .groups = "drop")
-
-SAvail_i_COMM2016 <- SAvail_im_DA2016 %>% st_drop_geometry() |>
-  group_by(COMMUNITY) %>%
-  summarize(COMM_prev_LIMAT_under18 = mean(DA_prev_LIMAT_under18, na.rm=T),
-            .groups = "drop")
-
-#calc changes between the years - make that 1 sf
-SAvail_i_COMM_changes <- full_join(SAvail_i_COMM2011 |> st_drop_geometry(),
-                                   SAvail_i_COMM2016, by=c("COMMUNITY")) |>
-  transmute(COMMUNITY,
-            prev_LIMAT_2011 = COMM_prev_LIMAT_under18.x,
-            prev_LIMAT_2016 = COMM_prev_LIMAT_under18.y,
-            c_prev_LIMAT_perc = ((COMM_prev_LIMAT_under18.y - COMM_prev_LIMAT_under18.x)/
-                                   COMM_prev_LIMAT_under18.x )|> scales::percent())
-```
-
-```{r}
-#calc changes between the years - make that 1 sf
-SAvail_im_COMM_changes <- full_join(SAvail_im_COMM2011|>st_drop_geometry(),
-                                    SAvail_im_COMM2016, by=c("COMMUNITY", "mode")) |>
-  transmute(COMMUNITY,
-            mode = case_when(mode == "c" ~ "motorized",
-                             mode == "w" ~ "non-motorized"),
-            COMMUNITY_ = dplyr::coalesce(COMMUNITY_.x, COMMUNITY_.y),
-            kid_2011 = kid_popadj_m.x,
-            kid_2016 = kid_popadj_m.y,
-            c_kid_perc = ((kid_popadj_m.y - kid_popadj_m.x)/kid_popadj_m.x) |> scales::percent(),
-            kid_proportion_2011 = (kid_2011/sum(kid_2011)),
-            kid_proportion_2016 = (kid_2016/sum(kid_2016)),
-            c_kid_proportion_perc = ((kid_proportion_2016 - kid_proportion_2011)/kid_proportion_2011)
-            |> scales::percent(),
-            kid_proportion_2011_perc = (kid_proportion_2011) |> scales::percent(),
-            kid_proportion_2016_perc = (kid_proportion_2016) |> scales::percent(),
-            V_im_2011 = V_im.x,
-            V_im_2016 = V_im.y,
-            c_V_im_perc = ((V_im.y -V_im.x)/V_im.x) |> scales::percent(),
-            V_im_proportion_2011 = (V_im.x/sum(V_im.x)),
-            V_im_proportion_2016 = (V_im.y/sum(V_im.y)),
-            c_V_im_proportion_perc = ((V_im_proportion_2016 - V_im_proportion_2011)/V_im_proportion_2011)
-            |> scales::percent(),
-            V_im_proportion_2011_perc = (V_im_proportion_2011) |> scales::percent(),
-            V_im_proportion_2016_perc = (V_im_proportion_2016) |> scales::percent(),
-            v_im_2011 = v_im_kid_popadj.x,
-            v_im_2016 = v_im_kid_popadj.y,
-            c_v_im_perc = ((v_im_kid_popadj.y - v_im_kid_popadj.x)/v_im_kid_popadj.x) |> scales::percent(),
-            total_tt_2011 = total_tt.x,
-            total_tt_2016 = total_tt.y,
-            total_tt_proportion_2011 = (total_tt.x/sum(total_tt.x)),
-            total_tt_proportion_2016 = (total_tt.y/sum(total_tt.y)),
-            c_total_tt_proportion_perc = ((total_tt_proportion_2016 - total_tt_proportion_2011)/total_tt_proportion_2011)
-            |> scales::percent(),
-            total_tt_proportion_2011 = (total_tt_proportion_2011) |> scales::percent(),
-            total_tt_proportion_2016 = (total_tt_proportion_2016) |> scales::percent(),
-            c_total_tt_perc = ((total_tt.y - total_tt.x)/
-                                 total_tt.x)|> scales::percent(),
-            total_tt_perkid_2011 = total_tt.x/kid_2011,
-            total_tt_perkid_2016 = total_tt.y/kid_2016,
-            c_total_ttperkid_perc = ((total_tt_perkid_2016 - total_tt_perkid_2011)/
-                                       total_tt_perkid_2011)|> scales::percent(),
-            emissions_2011 = emissions_total_car_tt.x,
-            emissions_2016 = emissions_total_car_tt.y,
-            c_emissions_perc = ((emissions_total_car_tt.y - emissions_total_car_tt.x)/
-                                  emissions_total_car_tt.x)|> scales::percent(),
-            emissions_proportion_2011 = (emissions_2011/sum(emissions_2011)),
-            emissions_proportion_2016 = (emissions_2016/sum(emissions_2016)),
-            c_emissions_proportion = ((emissions_proportion_2016 - emissions_proportion_2011)/emissions_proportion_2011)
-            |> scales::percent(),
-            emissions_proportion_2011_perc = (emissions_proportion_2011) |> scales::percent(),
-            emissions_proportion_2016_perc = (emissions_proportion_2016) |> scales::percent(),
-            emissions_perkid_2011 = emissions_total_car_tt.x/kid_popadj_m.x,
-            emissions_perkid_2016 = emissions_total_car_tt.y/kid_popadj_m.y,
-            c_emissions_perkid_perc = ((emissions_perkid_2016 - emissions_perkid_2011)/
-                                         emissions_perkid_2011)|> scales::percent(),
-            geometry) |> 
-  left_join(SAvail_i_COMM_changes, by="COMMUNITY") |> st_as_sf() 
+            geometry,
+            # kid_2011 = kid_popadj_m.x,
+            # kid_2016 = kid_popadj_m.y,
+            # c_kid_perc = ((kid_popadj_m.y - kid_popadj_m.x)/kid_popadj_m.x) |> scales::percent(),
+            # kid_proportion_2011 = (kid_2011/sum(kid_2011)),
+            # kid_proportion_2016 = (kid_2016/sum(kid_2016)),
+            # c_kid_proportion_perc = ((kid_proportion_2016 - kid_proportion_2011)/kid_proportion_2011)
+            # %>% scales::percent(),
+            # kid_proportion_2011_perc = (kid_proportion_2011) |> scales::percent(),
+            # kid_proportion_2016_perc = (kid_proportion_2016) |> scales::percent(),
+            # V_im_2011 = V_im.x,
+            # V_im_2016 = V_im.y,
+            # c_V_im_perc = ((V_im.y -V_im.x)/V_im.x) |> scales::percent(),
+            # V_im_proportion_2011 = (V_im.x/sum(V_im.x)),
+            # V_im_proportion_2016 = (V_im.y/sum(V_im.y)),
+            # c_V_im_proportion_perc = ((V_im_proportion_2016 - V_im_proportion_2011)/V_im_proportion_2011)
+            # %>% scales::percent(),
+            # V_im_proportion_2011_perc = (V_im_proportion_2011) |> scales::percent(),
+            # V_im_proportion_2016_perc = (V_im_proportion_2016) |> scales::percent(),
+            # v_im_2011 = v_im_kid_popadj.x,
+            # v_im_2016 = v_im_kid_popadj.y,
+            # c_v_im_perc = ((v_im_kid_popadj.y - v_im_kid_popadj.x)/v_im_kid_popadj.x) |> scales::percent(),
+            # total_tt_2011 = total_tt.x,
+            # total_tt_2016 = total_tt.y,
+            # total_tt_proportion_2011 = (total_tt.x/sum(total_tt.x)),
+            # total_tt_proportion_2016 = (total_tt.y/sum(total_tt.y)),
+            # c_total_tt_proportion_perc = ((total_tt_proportion_2016 - total_tt_proportion_2011)/total_tt_proportion_2011)
+            # %>% scales::percent(),
+            # total_tt_proportion_2011 = (total_tt_proportion_2011) |> scales::percent(),
+            # total_tt_proportion_2016 = (total_tt_proportion_2016) |> scales::percent(),
+            # c_total_tt_perc = ((total_tt.y - total_tt.x)/
+            #                      total_tt.x)|> scales::percent(),
+            # total_tt_perkid_2011 = total_tt.x/kid_2011,
+            # total_tt_perkid_2016 = total_tt.y/kid_2016,
+            # c_total_ttperkid_perc = ((total_tt_perkid_2016 - total_tt_perkid_2011)/
+            #                            total_tt_perkid_2011)|> scales::percent(),
+            # emissions_2011 = emissions_total_car_tt.x,
+            # emissions_2016 = emissions_total_car_tt.y,
+            # c_emissions_perc = ((emissions_total_car_tt.y - emissions_total_car_tt.x)/
+            #                       emissions_total_car_tt.x)|> scales::percent(),
+            # emissions_proportion_2011 = (emissions_2011/sum(emissions_2011)),
+            # emissions_proportion_2016 = (emissions_2016/sum(emissions_2016)),
+            # c_emissions_proportion = ((emissions_proportion_2016 - emissions_proportion_2011)/emissions_proportion_2011)
+            # %>% scales::percent(),
+            # emissions_proportion_2011_perc = (emissions_proportion_2011) |> scales::percent(),
+            # emissions_proportion_2016_perc = (emissions_proportion_2016) |> scales::percent(),
+            # emissions_perkid_2011 = emissions_total_car_tt.x/kid_popadj_m.x,
+            # emissions_perkid_2016 = emissions_total_car_tt.y/kid_popadj_m.y,
+            # c_emissions_perkid_perc = ((emissions_perkid_2016 - emissions_perkid_2011)/
+            #                              emissions_perkid_2011)|> scales::percent()
+            ) |> st_as_sf()
 ```
@@
 student_pop_absolute_plot <-
+    tm_shape(ham_bay, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0) +
+  tm_shape(hydro_p_LakeOntario, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0)+
   tm_shape(SAvail_i_DA_extra |> mutate(extra_col_for_panel = "HAMILTON")) +
   tm_polygons(col = "DA_pop_5to14", border.alpha=0, 
-              palette = "Greys",
+              palette = "Greens",
               labels = c("[0-40)", "[40-50)", "[50-75)","[75-835]"),
               breaks = c(0, 40, 50, 75, 835),
-              title = "Population aged 5 to 14"
+              title = "Pop. aged 5 to 14"
   )+
   tm_facets(by=c("Year"), ncol=2, nrow=1)+
+    tm_credits(c("City total: 58,265","City total: 58,865"),
+             position = c("LEFT", "BOTTOM")) +
   tmap_defaults +
-  tm_layout(panel.label.height = 0.8)
+  tm_layout(bg.color = "grey",
+            panel.label.bg.color = "white",
+            panel.label.height = 0.8)
 
 student_pop_rate_plot <-
+  tm_shape(ham_bay, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0) +
+  tm_shape(hydro_p_LakeOntario, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0)+
   tm_shape(SAvail_i_DA_extra) +
@@
               breaks = c(0, 0.0018, 0.0025, 0.0038, 0.60),
-              title = "Average pop. aged 5 to 14\nper residential unit"
+              title = "Avg. pop. aged 5 to 14\nper residential unit"
   )+
   tm_facets(by=c("Year"), ncol=2, nrow=1)+
+
   tmap_defaults +
-  tm_layout(panel.show=FALSE)
+  tm_layout(bg.color = "grey",
+            panel.show=FALSE)
 
 LIMAT_under18_DA_plot <-
+  tm_shape(ham_bay, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0) +
+  tm_shape(hydro_p_LakeOntario, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0)+
   tm_shape(SAvail_i_DA_extra) +
@@
               breaks = c(0, 4.30, 13.00, 30.8, 150),
+              colorNA = "black",
               textNA = "Unavailable",
-              title = "Proportion of LIM-AT households\n with children < 18"
+              title = "Prop. of LIM-AT households\n with dependents < 18"
   )+
@@
   tmap_defaults  +
-  tm_layout(panel.show=FALSE)
+  tm_layout(bg.color = "grey",
+            panel.show=FALSE)
 
@@
 
-tmap_save(stud_pop_rate_LIM_plot, paste0(here::here(),"/Figures/stud_pop_rate_LIM_plot.png"), height=7, width=12, dpi=600)
+stud_pop_rate_LIM_plot
+```
+
+```{r student-popabs-lowincperc-poprate-plot2, eval=FALSE}
+tmap_save(stud_pop_rate_LIM_plot, paste0(here::here(),"/Figures/stud_pop_rate_LIM_plot.png"), height=10.5, width=12, dpi=600)
 ```
@@
 ```{r importing-taz-and-associated-flows, include=FALSE}
-#Read the boundaries of the traffic analysis zones (retrieved from the TTS 2016):
-ggh_taz <- st_read(paste0(here::here(), "/data-prep/data-inputs/TTS/tts06_83_region.shp"),
-                   quiet = TRUE) %>% 
-  st_transform(crs = 32617)
-
-# Prepare traffic analysis zones:
-ggh_taz <- ggh_taz %>%
-  transmute(GTA06 = as.character(GTA06),
-            AREA = st_area(geometry) %>% 
-              set_units(km^2) %>% 
-              drop_units())
-#Extract the IDs for all TAZ:
-ghta_taz_id <- ggh_taz$GTA06
-
-#Read the boundaries of the traffic analysis zones (retrieved from the TTS 2016):
-ggh_pd <- st_read(paste0(here::here(), "/data-prep/data-inputs/TTS/tts06_pd_83.shp"),
-                  quiet = TRUE) %>%
-  st_transform(crs = 32617)
-
-#Adding region names to the object as they are not included in the origin data: (http://dmg.utoronto.ca/pdf/tts/2016/2016TTS_Conduct.pdf)
-
-ggh_pd <- ggh_pd %>%
-  mutate(REGION_name = 
-           ifelse(REGION == 1, "Toronto", 
-                  ifelse(REGION == 2, "Durham", 
-                         ifelse(REGION == 3, "York", 
-                                ifelse(REGION == 4, "Peel", 
-                                       ifelse(REGION == 5, "Halton", 
-                                              ifelse(REGION == 6, "Hamilton", 
-                                                     ifelse(REGION == 11, "Niagara", 
-                                                            ifelse(REGION == 12, "Waterloo", 
-                                                                   ifelse(REGION == 13, "Guelph", 
-                                                                          ifelse(REGION == 14, "Wellington", 
-                                                                                 ifelse(REGION == 15, "Orangeville", 
-                                                                                        ifelse(REGION == 16, "Barrie", 
-                                                                                               ifelse(REGION == 17, "Simcoe", 
-                                                                                                      ifelse(REGION == 18, "Kawartha Lakes", 
-                                                                                                             ifelse(REGION == 19, "Peterborough City", 
-                                                                                                                    ifelse(REGION == 20, "Peterborough County", 
-                                                                                                                           ifelse(REGION == 21, "Orillia", 
-                                                                                                                                  ifelse(REGION == 22, "Dufferin", 
-                                                                                                                                         ifelse(REGION == 23, "Brantford", 
-                                                                                                                                                ifelse(REGION == 24, "Brant", "NOOOO")))))))))))))))))))))
-
-# add 'REGION' from the planning distract object to the `ggh_taz`. 
-#buffering to clean up the edges of these objects -> there wre issues using st_join. 
-ggh_taz <- st_buffer(ggh_taz, dist=0)
-ggh_pd <- st_buffer(ggh_pd, dist=0)
-
-ggh_taz <- st_join(ggh_taz, ggh_pd %>% transmute(REGION),left = TRUE, largest = TRUE)
+#reading boundaries of TAZs and PDs from the TTS2016R package (same TAZ for both 2011 and 2016)
+library(TTS2016R) # A R package of the 2016 TTS
+data(ggh_taz)
+data(ggh_pd)
+# 
+# #Read the boundaries of the traffic analysis zones (retrieved from the TTS 2016):
+# ggh_taz <- st_read(paste0(here::here(), "/data-prep/data-inputs/TTS/tts06_83_region.shp"),
+#                    quiet = TRUE) %>% 
+#   st_transform(crs = 32617)
+# 
+# # Prepare traffic analysis zones:
+# ggh_taz <- ggh_taz %>%
+#   transmute(GTA06 = as.character(GTA06),
+#             AREA = st_area(geometry) %>% 
+#               set_units(km^2) %>% 
+#               drop_units())
+# #Extract the IDs for all TAZ:
+# ghta_taz_id <- ggh_taz$GTA06
+# 
+# #Read the boundaries of the traffic analysis zones (retrieved from the TTS 2016):
+# ggh_pd <- st_read(paste0(here::here(), "/data-prep/data-inputs/TTS/tts06_pd_83.shp"),
+#                   quiet = TRUE) %>%
+#   st_transform(crs = 32617)
+# 
+# #Adding region names to the object as they are not included in the origin data: (http://dmg.utoronto.ca/pdf/tts/2016/2016TTS_Conduct.pdf)
+# 
+# ggh_pd <- ggh_pd %>%
+#   mutate(REGION_name = 
+#            ifelse(REGION == 1, "Toronto", 
+#                   ifelse(REGION == 2, "Durham", 
+#                          ifelse(REGION == 3, "York", 
+#                                 ifelse(REGION == 4, "Peel", 
+#                                        ifelse(REGION == 5, "Halton", 
+#                                               ifelse(REGION == 6, "Hamilton", 
+#                                                      ifelse(REGION == 11, "Niagara", 
+#                                                             ifelse(REGION == 12, "Waterloo", 
+#                                                                    ifelse(REGION == 13, "Guelph", 
+#                                                                           ifelse(REGION == 14, "Wellington", 
+#                                                                                  ifelse(REGION == 15, "Orangeville", 
+#                                                                                         ifelse(REGION == 16, "Barrie", 
+#                                                                                                ifelse(REGION == 17, "Simcoe", 
+#                                                                                                       ifelse(REGION == 18, "Kawartha Lakes", 
+#                                                                                                              ifelse(REGION == 19, "Peterborough City", 
+#                                                                                                                     ifelse(REGION == 20, "Peterborough County", 
+#                                                                                                                            ifelse(REGION == 21, "Orillia", 
+#                                                                                                                                   ifelse(REGION == 22, "Dufferin", 
+#                                                                                                                                          ifelse(REGION == 23, "Brantford", 
+#                                                                                                                                                 ifelse(REGION == 24, "Brant", "NOOOO")))))))))))))))))))))
+# 
+# # add 'REGION' from the planning distract object to the `ggh_taz`. 
+# #buffering to clean up the edges of these objects -> there wre issues using st_join. 
+# ggh_taz <- st_buffer(ggh_taz, dist=0)
+# ggh_pd <- st_buffer(ggh_pd, dist=0)
+# 
+# ggh_taz <- st_join(ggh_taz, ggh_pd %>% transmute(REGION),left = TRUE, largest = TRUE)
+
 ham_taz <- ggh_taz %>% dplyr::filter(REGION==6) #Filter out everything other than Planning District 6 = Hamilton
@@
 ```{r ploting-od-taz-flows, eval=FALSE}
-car_OD_flow_plot <-  tm_shape(od_trips_perc_car |> mutate(mode = "motorized"), bbox=SAvail_i_DA_extra) +
+tmap_defaults <-
+  tm_scale_bar(position = c("LEFT", "BOTTOM"), breaks = c(0, 5, 10, 15))+
+  tm_compass(size = 1.5, position = c("right", 'top')) +
+  tm_layout(legend.position = c("left", "top")) #legend.hist.size = 10,
+
+car_OD_flow_plot <-  
+    tm_shape(ham_bay, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0) +
+  tm_shape(hydro_p_LakeOntario, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0)+
+  tm_shape(od_trips_perc_car |> mutate(mode = "motorized"), bbox=SAvail_i_DA_extra) +
   tm_polygons(col = "TAZOrig_car_perc", border.alpha = 0.4, border.col = "grey20", lwd=0.2,
@@
               textNA = "Unavailable",
-              colorNA = "white")+
+              colorNA = "black")+
   #tm_shape(SAvail_i_DA_extra) + tm_borders(alpha = 0.4, col = "grey20", lwd=0.2)+
   tm_shape(ham_taz) + tm_borders(alpha=0.4, col = "grey40", lwd=0.2) +
-  tm_shape(ham_community_bounds) + tm_borders(alpha = 0.8, col = "lightblue", lwd=2)+
+  tm_shape(ham_community_bounds %>% st_cast("MULTILINESTRING")) + 
+  tm_lines(col = "COMMUNITY_", 
+           palette = c("#365C8DFF", "#277F8EFF", "#1FA187FF", "#4AC16DFF", "#9FDA3AFF", "#FDE725FF"),
+           title.col = "Hamilton communities", lwd = 1.5) + 
   tm_shape(ham_taz_centroids) + tm_dots(alpha=0.5, col="darkblue", size=0.01) +
   tm_shape(desire_lines_ham |> mutate(mode = "motorized")) +
-  tm_lines(palette = "Reds", breaks = c(6, 22, 26, 49, 353),
-           title.col = "Number of trips (motorized)",
+  tm_lines(palette = "Reds", breaks = c(6, 22, 26, 52, 353),
+           title.col = "Number of trips",
            col = "TAZOrig_car_trips",
@@
   tmap_defaults +
-  tm_layout(panel.label.height = 0.8)
-
-walk_OD_flow <- tm_shape(od_trips_perc_walk |> mutate(mode = "non-motorized"), bbox=SAvail_i_DA_extra) +
+  tm_layout(bg.color = "grey",
+            panel.label.bg.color = "white",
+            panel.label.height = 0.8)
+
+walk_OD_flow <- tm_shape(ham_bay, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0) +
+  tm_shape(hydro_p_LakeOntario, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0)+
+  tm_shape(od_trips_perc_walk |> mutate(mode = "non-motorized"), bbox=SAvail_i_DA_extra) +
   tm_polygons(col = "TAZOrig_walk_perc", border.alpha = 0.4, border.col = "grey20", lwd=0.2,
@@
               textNA = "Unavailable",
-              colorNA = "white",
+              colorNA = "black",
               legend.show=FALSE)+
@@
   tm_shape(ham_taz) + tm_borders(alpha=0.4, col = "grey40", lwd=0.2) +
-  tm_shape(ham_community_bounds) + tm_borders(alpha = 0.8, col = "lightblue", lwd=2)+
+  tm_shape(ham_community_bounds %>% st_cast("MULTILINESTRING")) + 
+  tm_lines(col = "COMMUNITY_", 
+           palette = c("#365C8DFF", "#277F8EFF", "#1FA187FF", "#4AC16DFF", "#9FDA3AFF", "#FDE725FF"),
+           title.col = "Hamilton communities", lwd = 1.5, legend.col.show=FALSE) + 
   tm_shape(ham_taz_centroids) + tm_dots(alpha=0.5, col="darkblue", size=0.01) +
@@
            |> mutate(mode = "non-motorized")) +
-  tm_lines(palette = "Greens", breaks = c(12, 23, 44, 67, 174),
+  tm_lines(palette = "Reds", breaks = c(6, 22, 26, 52, 353),
            title.col = "Number of trips (non-motorized)",
@@
            lwd = "all",
+           legend.col.show = FALSE,
            legend.lwd.show = FALSE) +
@@
   tmap_defaults +
-  tm_layout(panel.label.height = c(0.8,0,0)) +
+  tm_layout(bg.color = "grey",
+            panel.label.bg.color = "white",
+            panel.label.height = c(0.8,0,0)) +
   tm_add_legend(type = "line", 
@@
                 labels = "TAZ centroid")
+
 OD_flow_plot <- tmap_arrange(car_OD_flow_plot, walk_OD_flow, nrow = 2)
+OD_flow_plot
+
 tmap_save(OD_flow_plot, paste0(here::here(),"/Figures/OD_flow_plot.png"), height=7, width=12, dpi=600)
@@
                      y = ..density..),
-                 colour = "grey80",
-                 alpha = 0.5,
+                 colour = "grey70",
+                 fill = "grey70",
                  boundary = 0,
@@
   geom_vline(data = q_3, mapping = aes(xintercept = x), colour = "grey60") +
-  geom_line(data = tld_theoretical, aes(x = t, y = f), colour = "blue", size=0.8) + 
-  xlab("Estimated TTS travel time (minutes)") +
-  scale_x_continuous(breaks = seq(0,40,10), expand=c(0,0))+
+  geom_line(data = tld_theoretical, aes(x = t, y = f), colour = "blue", size=0.8, alpha=0.6) + 
+  xlab("Estimated travel time associated with TTS OD flows (minutes)") +
+  ylab("Density") +
+  scale_x_continuous(breaks = seq(0,45,10), expand=c(0,0))+
   scale_y_continuous(breaks = seq(0,0.12,0.04), limits = c(0,0.12))+
@@
   geom_text(data = q_1, 
-            mapping = aes(x = x, y = y, label = label), colour = "grey20", size=3) +
+            mapping = aes(x = x, y = y, label = label), colour = "grey20", size=4) +
   geom_text(data = q_2,
-            mapping = aes(x = x, y = y, label = label), colour = "grey20", size=3) +
+            mapping = aes(x = x, y = y, label = label), colour = "grey20", size=4) +
   geom_text(data = q_3,
-            mapping = aes(x = x, y = y, label = label), colour = "grey20", size=3) +
-  theme_classic()
+            mapping = aes(x = x, y = y, label = label), colour = "grey20", size=4) +
+  theme_classic() +
+  theme(axis.title = element_text(size = 11),
+        panel.border = element_rect(colour = "grey80", fill=NA),
+        strip.text = element_text(size=11))
 TLD_empirical_wtheo_curve_plot
+```
+
+```{r ploting-empirical-and-theo-curve-TLD-tts2, eval=FALSE}
 ggsave(paste0(here::here(),"/Figures/TLD_empirical_wtheo_curve_plot.png"), height=7, width=12, dpi=600)
 ```
+```{r}
+ALL_SCHOOLS_Elem_formapping2 <- ALL_SCHOOLS_Elem_formapping |> mutate(Mode = "motorized")
+ALL_SCHOOLS_Elem_formapping3 <- ALL_SCHOOLS_Elem_formapping |> mutate(Mode = "non-motorized")
+ALL_SCHOOLS_Elem_formapping2 <- rbind(ALL_SCHOOLS_Elem_formapping2, ALL_SCHOOLS_Elem_formapping3)
+rm(ALL_SCHOOLS_Elem_formapping3)
+```
 
 ```{r, eval=FALSE}
-Vim_20112016_plot <-  tm_shape(SAvail_im_DA, bbox=SAvail_im_DA) +
+Vim_20112016_plot <-  
+  tm_shape(ham_bay, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0) +
+  tm_shape(hydro_p_LakeOntario, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0)+
+  tm_shape(SAvail_im_DA, bbox=SAvail_im_DA) +
   tm_polygons(col = "V_im", border.alpha = 0, border.col = "grey20", lwd=0.2,
-              palette = "Greys",
+              palette = "Purples",
               style="quantile", n=4,
-              title = expression(V[i]^m*' (quartiles)'),
-              textNA = "Unavailable",
-              colorNA = "white")+
+              labels = c("[0-0.2)", "[0.2-12.8)", "[12.8-56.4)", "[56.4-672.9)"),
+              title = expression(V[i]^m),
+              textNULL = "Unavailable",
+              colorNULL = "black")+
   tm_facets(by = c("mode","year"), nrow=1)+
-  tm_shape(ALL_SCHOOLS_Elem_formapping) +
-  tm_dots(col = "Status", size=0.03, title = "School status between\n2011 and 2016",
-          palette = c("ivory", "black", "darkgreen","#0E86D4","#90EE90"),
-          labels = c("No change","Closed before 2016", "Expanded before 2016","Moved before 2016","New after 2011"), legend.show=FALSE) +
-  tm_shape(ham_community_bounds) + tm_borders(alpha = 0.8, col = "lightblue", lwd=1.5)+
+  tm_shape(ham_community_bounds %>% st_cast("MULTILINESTRING")) + 
+  tm_lines(col = "COMMUNITY_", 
+           palette = c("#365C8DFF", "#277F8EFF", "#1FA187FF", "#4AC16DFF", "#9FDA3AFF", "#FDE725FF"),
+           title.col = "Hamilton communities", lwd = 2, legend.col.show = FALSE) + 
+  tm_shape(ALL_SCHOOLS_Elem_formapping2) +
+  tm_symbols(shape = "Status",
+             size=0.1,
+             alpha=0.5,
+          title.shape = "School status",
+          shapes = c(24,25,22,21,23),
+          shapes.labels = c(
+          "Expanded after 2011",
+          "Moved before 2016",
+          "New after 2011",
+          "No change btwn 2011-16",
+          "Closed before 2016"),
+          legend.shape.show = FALSE
+  ) +
+  tm_facets(by=c("Mode","Year"))+
+  tm_credits(c("60,220 spatial availability \n(99.4% of the 60,567 city-wide OTGC)",
+               "58,101 spatial availability\n(99.7% of the 58,287 city-wide OTGC)",
+                "347 spatial availability \n(0.6% of the 60,567 city-wide OTGC)",
+               "186 spatial availability\n(0.3% of the 58,287 city-wide OTGC)"),
+             position = c("LEFT", "BOTTOM")) +
   tmap_defaults +
-  tm_layout(panel.label.height = 0.8)
-
-vvim_20112016_plot <-  tm_shape(SAvail_im_DA, bbox=SAvail_im_DA) +
+  tm_layout(bg.color = "grey",
+            panel.label.bg.color = "white",
+            panel.label.height = 0.8)
+
+vvim_20112016_plot <-  
+  tm_shape(ham_bay, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0) +
+  tm_shape(hydro_p_LakeOntario, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0)+
+  tm_shape(SAvail_im_DA, bbox=SAvail_im_DA) +
   tm_polygons(col = "v_im_kid_popadj", border.alpha = 0, border.col = "grey20", lwd=0.2,
@@
               title = expression(v[i]^m),
-              breaks = c(0, 0.5, 1, 1.5, 2, 2.1), #note the highest value is ~12, I'm breaking the values early for the purpose of visualisation
-              labels = c("[0-0.5)", "[0.5-1)", "[1-1.5)", "[1.5-2)", "[>2)"),
-              textNA = "Unavailable",
-              colorNA = "white")+
+              # style="quantile", n=4,
+              breaks = c(0, 0.62, 0.9, 1.22, 2.1), #note the highest value is ~10.29, I'm breaking the values early for the purpose of visualisation
+              labels = c("[0-0.62)", "[0.62-0.9)", "[0.9-1.22)", "[1.22-10.26)"),
+              textNULL = "Unavailable",
+              colorNULL = "black")+
   tm_facets(by = c("mode","year"), nrow=1)+
-  tm_shape(ALL_SCHOOLS_Elem_formapping) +
-  tm_dots(col = "Status", size=0.03, title = "School status between\n2011 and 2016",
-          palette = c("ivory", "black", "darkgreen","#0E86D4","#90EE90"),
-          labels = c("No change","Closed before 2016", "Expanded before 2016","Moved before 2016","New after 2011")          ) +
-  tm_shape(ham_community_bounds) + tm_borders(alpha = 0.8, col = "lightblue", lwd=1.5)+
+  tm_shape(ham_community_bounds %>% st_cast("MULTILINESTRING")) + 
+  tm_lines(col = "COMMUNITY_", 
+           palette = c("#365C8DFF", "#277F8EFF", "#1FA187FF", "#4AC16DFF", "#9FDA3AFF", "#FDE725FF"),
+           title.col = "Hamilton communities", lwd = 2) + 
+  tm_shape(ALL_SCHOOLS_Elem_formapping2) +
+  tm_symbols(shape = "Status",
+             size=0.1,
+             alpha=0.5,
+          title.shape = "School status",
+          shapes = c(24,25,22,21,23),
+          shapes.labels = c(
+          "Expanded after 2011",
+          "Moved before 2016",
+          "New after 2011",
+          "No change btwn 2011-16",
+          "Closed before 2016")
+  ) +
+  tm_facets(by=c("Mode","Year"))+
+  tm_credits(c("1.09 avg. spatial availability per capita",
+               "1.05 avg. spatial availability per capita",
+                "0.85 avg. spatial availability per capita",
+               "0.71 avg. spatial availability per capita"),
+             position = c("LEFT", "BOTTOM")) +
   tmap_defaults +
-  tm_layout(panel.label.height = 0.8)
+  tm_layout(bg.color = "grey",
+            panel.label.bg.color = "white",
+            panel.label.height = 0.8)
 
 Vim_vvim_20112016_plot <- tmap_arrange(Vim_20112016_plot, vvim_20112016_plot, nrow=2)
-
-tmap_save(Vim_vvim_20112016_plot, paste0(here::here(),"/Figures/Vim_vvim_20112016_plot.png"), height=7, width=12, dpi=600)
 ```
@@
 ```{r, eval=FALSE}
-Vim_20112016_CHANGE_plot <-  tm_shape(SAvail_im_DA_changes, bbox=SAvail_im_DA) +
+tmap_save(Vim_vvim_20112016_plot, paste0(here::here(),"/Figures/Vim_vvim_20112016_plot.png"), height=12, width=12, dpi=600)
+```
+
+```{r, eval=FALSE}
+tmap_defaults <-
+  tm_scale_bar(position = c("LEFT", "BOTTOM"), breaks = c(0, 5, 10, 15))+
+  tm_compass(size = 1.5, position = c("right", 'top')) +
+  tm_layout(legend.position = c("left", "top")) #legend.hist.size = 10,
+
+Vim_20112016_CHANGE_plot <-  
+  tm_shape(ham_bay, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0) +
+  tm_shape(hydro_p_LakeOntario, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0)+
+  tm_shape(SAvail_im_DA_changes, bbox=SAvail_im_DA) +
   tm_polygons(col = "c_V_im", border.alpha = 0, border.col = "grey20", lwd=0.2,
-              palette = "-PuOr", midpoint=0,
-              breaks = c(-5, -4, -3, -2, -1, 0, 1, 2,3,4,5),
-              labels = c("[>-5 to -4)", "[-4 to -3)", "[-3 to -2)","[-2 to -1)","[-1 to 0)", "[0 to 1)", 
-                         "[1 to 2)", "[2 to 3)", "[3 to 4)", "[4 to <5)" ),
+              palette = "RdYlBu", midpoint=0,
+              # style="quantile", n=4,
+              breaks = c(-5, -4.1, -0.2, 0, 5),
+              labels = c("[-208 to -4.1)", "[-4.1 to -0.2)", "[-0.2 to 0)","[0 to 283.6)"),
               title = expression("Change in "*V[i]^m),
-              textNA = "Unavailable",
+              textNA = "No V in 2016 or 2011",
               colorNA = "white")+
   tm_facets(by = c("mode"), nrow=1)+
+  tm_shape(ham_community_bounds %>% st_cast("MULTILINESTRING")) + 
+  tm_lines(col = "COMMUNITY_", 
+           palette = c("#365C8DFF", "#277F8EFF", "#1FA187FF", "#4AC16DFF", "#9FDA3AFF", "#FDE725FF"),
+           title.col = "Hamilton communities", lwd = 2, 
+           legend.col.show = FALSE) + 
   tm_shape(ALL_SCHOOLS_Elem_formapping |> 
@@
                                               (Status == "Expanded" | Status == "New" ) ~ "New or Expanded"))) +
-  tm_dots(col = "Status_change", size=0.05, title = "Schools that changed status",
-          palette = c("black", "#90EE90"), legend.show=FALSE) +
-  tm_shape(ham_community_bounds) + tm_borders(alpha = 0.8, col = "lightblue", lwd=2)+
+  tm_symbols(shape = "Status_change",
+             size=0.1,
+             alpha=0.5,
+          title.shape = "Schools that changed status",
+          shapes = c(22,23),
+          shapes.labels = c(
+          "New or expanded after 2011",
+          "Closed before 2016"),
+          legend.shape.show = FALSE) +
+  tm_facets(by=c("Year"))+
   tmap_defaults +
-  tm_layout(panel.label.height = 0.8)
-
-#tmap_save(Vim_20112016_CHANGE_plot, "Figures/Vim_20112016_CHANGE_plot.png", height=7, width=12, dpi=600)
-
-vvim_20112016_CHANGE_plot <- tm_shape(SAvail_im_DA_changes, bbox=SAvail_im_DA) +
+  tm_layout( bg.color = "grey",
+             panel.label.bg.color = "white",
+             panel.label.height = 0.8)
+
+vvim_20112016_CHANGE_plot <-
+  tm_shape(ham_bay, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0) +
+  tm_shape(hydro_p_LakeOntario, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0)+
+  tm_shape(SAvail_im_DA_changes, bbox=SAvail_im_DA) +
   tm_polygons(col = "c_v_im_kid_popadj", border.alpha = 0, border.col = "grey20", lwd=0.2,
-              palette = "RdYlGn", midpoint=0,
+              palette = "RdYlBu", midpoint=0,
               breaks = c(-0.1, -0.05, 0.00, 0.05, 0.1),
@@
   tm_facets(by = c("mode"), nrow=1)+
+  tm_shape(ham_community_bounds %>% st_cast("MULTILINESTRING")) + 
+  tm_lines(col = "COMMUNITY_", 
+           palette = c("#365C8DFF", "#277F8EFF", "#1FA187FF", "#4AC16DFF", "#9FDA3AFF", "#FDE725FF"),
+           title.col = "Hamilton communities", lwd = 2) + 
   tm_shape(ALL_SCHOOLS_Elem_formapping |> 
@@
                                               (Status == "Expanded" | Status == "New" ) ~ "New or Expanded"))) +
-  tm_dots(col = "Status_change", size=0.05, title = "Schools that changed status",
-          palette = c("black", "#90EE90")) +
-  tm_shape(ham_community_bounds) + tm_borders(alpha = 0.8, col = "lightblue", lwd=2)+
+  tm_symbols(shape = "Status_change",
+             size=0.1,
+             alpha=0.5,
+          title.shape = "Schools that changed status",
+          shapes = c(22,23),
+          shapes.labels = c(
+          "New or expanded after 2011",
+          "Closed before 2016")
+  ) +
+  tm_facets(by=c("Year"))+
   tmap_defaults +
-  tm_layout(panel.label.height = 0.8)
+  tm_layout( bg.color = "grey",
+             panel.label.bg.color = "white",
+             panel.label.height = 0.8,
+             panel.show = FALSE)
 
@@
 
+```
+
+```{r, eval=FALSE}
 tmap_save(Vim_vvim_20112016_CHANGE_plot, 
           paste0(here::here(),"/Figures/Vim_vvim_20112016_CHANGE_plot.png"),
+          height=7,
+          width=12, 
+          dpi=600)
+```
+
+```{r, eval=FALSE}
+tmap_defaults <-
+  tm_scale_bar(position = c("LEFT", "BOTTOM"), breaks = c(0, 5, 10, 15))+
+  tm_compass(size = 1.5, position = c("right", 'top')) +
+  tm_layout(legend.position = c("left", "top")) #legend.hist.size = 10,
+
+ttim_20112016_CHANGE_plot <-  
+  tm_shape(ham_bay, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0) +
+  tm_shape(hydro_p_LakeOntario, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0)+
+  tm_shape(SAvail_im_DA_changes, bbox=SAvail_im_DA) +
+  tm_polygons(col = "c_total_tt", border.alpha = 0, border.col = "grey20", lwd=0.2,
+              palette = "RdYlBu", midpoint=0,
+             # style="quantile", n=4,
+              breaks = c(-60, -47, -3.1, 0, 10),
+              labels = c("[-1971 to -47)", "[-47 to -3.1)", "[-3.1 to 0)","[0 to 10)"),
+              title = expression("Change in travel time minutes"),
+              textNA = "No V in 2016 or 2011",
+              colorNA = "white")+
+  tm_facets(by = c("mode"), nrow=1)+
+  tm_shape(ham_community_bounds %>% st_cast("MULTILINESTRING")) + 
+  tm_lines(col = "COMMUNITY_", 
+           palette = c("#365C8DFF", "#277F8EFF", "#1FA187FF", "#4AC16DFF", "#9FDA3AFF", "#FDE725FF"),
+           title.col = "Hamilton communities", lwd = 2, 
+           legend.col.show = FALSE) + 
+  tm_shape(ALL_SCHOOLS_Elem_formapping |> 
+             filter(Status != "NoChange" & Status != "Moved")|>
+             mutate(Status_change = case_when(Status == "Removed" ~ "Closed",
+                                              (Status == "Expanded" | Status == "New" ) ~ "New or Expanded"))) +
+  tm_symbols(shape = "Status_change",
+             size=0.1,
+             alpha=0.5,
+          title.shape = "Schools that changed status",
+          shapes = c(22,23),
+          shapes.labels = c(
+          "New or expanded after 2011",
+          "Closed before 2016"),
+          legend.shape.show = FALSE) +
+  tm_facets(by=c("Year"))+
+  tmap_defaults +
+  tm_layout( bg.color = "grey",
+             panel.label.bg.color = "white",
+             panel.label.height = 0.8)
+ttim_20112016_CHANGE_plot
+```
+
+```{r, eval=FALSE}
+GHGim_20112016_CHANGE_plot <-
+  tm_shape(ham_bay, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0) +
+  tm_shape(hydro_p_LakeOntario, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0)+
+  tm_shape(SAvail_im_DA_changes, bbox=SAvail_im_DA) +
+  tm_polygons(col = "c_emissions_total_car_tt", border.alpha = 0, border.col = "grey20", lwd=0.2,
+              palette = "RdYlBu", midpoint=0,
+              style="quantile", n=4,
+              #breaks = c(-0.1, -0.05, 0.00, 0.05, 0.1),
+              #labels = c("[>-0.1 to -0.05)", "[-0.05 to 0.00)", "[0.00 to 0.05)", "[0.05 to <0.1)"),
+              title = expression("Change in GHG emissions (kg CO2e for home-school trips)"),
+              textNA = "Unavailable",
+              colorNA = "white")+
+  tm_facets(by = c("mode"), nrow=1)+
+  tm_shape(ham_community_bounds %>% st_cast("MULTILINESTRING")) + 
+  tm_lines(col = "COMMUNITY_", 
+           palette = c("#365C8DFF", "#277F8EFF", "#1FA187FF", "#4AC16DFF", "#9FDA3AFF", "#FDE725FF"),
+           title.col = "Hamilton communities", lwd = 2) + 
+  tm_shape(ALL_SCHOOLS_Elem_formapping |> 
+             filter(Status != "NoChange" & Status != "Moved")|>
+             mutate(Status_change = case_when(Status == "Removed" ~ "Closed",
+                                              (Status == "Expanded" | Status == "New" ) ~ "New or Expanded"))) +
+  tm_symbols(shape = "Status_change",
+             size=0.1,
+             alpha=0.5,
+          title.shape = "Schools that changed status",
+          shapes = c(22,23),
+          shapes.labels = c(
+          "New or expanded after 2011",
+          "Closed before 2016")
+  ) +
+  tm_facets(by=c("Year"))+
+  tmap_defaults +
+  tm_layout( bg.color = "grey",
+             panel.label.bg.color = "white",
+             panel.label.height = 0.8,
+             panel.show = FALSE)
+
+ttim_GHGim_20112016_CHANGE_plot <- tmap_arrange(ttim_20112016_CHANGE_plot, GHGim_20112016_CHANGE_plot, nrow=2)
+ttim_GHGim_20112016_CHANGE_plot
+```
+
+```{r, eval=FALSE}
+tmap_save(ttim_GHGim_20112016_CHANGE_plot, 
+          paste0(here::here(),"/Figures/ttim_GHGim_20112016_CHANGE_plot.png"),
           height=7,
@@
 Hamilton is a mid-size city (\~570,000 pop) with rural, suburban and urban characteristics in the province of Ontario, Canada [@governmentofcanadaProfileTableCensus2022].
-In 2013, the Hamilton-Wentworth District School Board (HWDSB),, the largest public English school board in the city, released its Long Term Facilities Master Plan [@hwdsbLongTermFacilities2013].
+In 2013, the Hamilton-Wentworth District School Board (HWDSB), the largest public English school board in the city, released its Long Term Facilities Master Plan [@hwdsbLongTermFacilities2013].
 To ensure "equitable, affordable and sustainable learning facilities", the Master Plan indicated that 80% of it's elementary schools (i.e., schools attended by students aged 5 to 13) would be subjected to accommodation evaluations, including an "Accommodation Review" whose outcome historically has led to school closures/consolidation [@hwdsbLongTermFacilities2013; @craggsBoardEvaluatingFuture2013; @seasonsSchoolClosurePolicy2014].
@@
 #| label: fig-Fig1
-#| fig-cap: "The school catchments, estimated on the ground capacities (OTGCs) of schools summed per catchment, and school status for both public and public catholic school boards for year 2011 and 2016.
-OTGCs is presented by quartiles."
+#| fig-cap: "The school catchments, estimated On The Ground Capacity (OTGC) of schools summed per catchment, and school status for both public and public catholic school boards for year 2011 and 2016.
+OTGC scale is presented by quartiles."
 #| echo: false
-#| out-width: "140%"
+#| out-width: "120%"
 
@@
 #| fig-cap: "The magnitude (top row) and rate per residential unit (middle row) of elementary aged student population per DA in 2011 and 2016.
-The proportion of low-income households with children under 18 years old is also shown (bottom row).
+Bottom row: the proportion of low-income households (prevelance of low-income after-tax, LIM-AT, measured by the 2011 and 2016 Canadian Census) with dependents under 18 years old shown.
 All scales are represented in quartiles."
@@
 With a target 5% sampling rate, the survey is expanded to be representative at the traffic analysis zone (TAZ) level of geography.
-TAZ are spatial units created for the purpose of the TTS; a few DAs typically nest within each TAZ.
+TAZ are spatial units created for the purpose of the TTS and pulled from the R data package {TTS2016R} [@soukhovTTS2016RDataSet2023]; a few DAs typically nest within each TAZ.
 The trip-level travel data extract for this paper represent `r od_trips_perc_2011$TAZOrig_car_trips |> sum() |> prettyNum(big.mark = ",")` and `r od_trips_perc_2016$TAZOrig_car_trips |> sum() |> prettyNum(big.mark = ",")` motorized trips (mode used includes private care passenger, school bus, taxi, and transit) and `r od_trips_perc_2011$TAZOrig_walk_trips |> sum() |> prettyNum(big.mark = ",")` and `r od_trips_perc_2016$TAZOrig_walk_trips |> sum() |> prettyNum(big.mark = ",")` non-motorized trips (mode used includes walk and cycling) for 5 to 13 year olds from home to school in 2011 and 2016 respectively.
@@
 Or can you add it to the data package?
-And then cite your data paper.-->
+And then cite your data paper.
+[AS]: fantastic idea!
+Added!
+for now I just have the TAZ boundaries in TTS2016R.
+The package can be expanded to include school travel but I'd need to add the full GTHA region.
+For this analysis I only exported Hamilton (PD region 6) beginning/ending trips.-->
 
@@
 #| label: fig-Fig3
-#| fig-cap: "The origin-destination flows from home-based motorized and non-motorized school trips for students between 5-13 years old as retrieved from TTS 2011 and 2016 mapped atop proportion of TAZ non/motorized modal share, DA boundaries, and community boundaries."
+#| fig-cap: "The origin-destination flows from home-based motorized and non-motorized school trips for students between 5-13 years old as retrieved from TTS 2011 and 2016.
+Flows are mapped atop the proportion of TAZ non/motorized modal share from the TTS 2011 and 2016, DA boundaries, and Hamilton community boundaries."
 #| echo: false
-#| out-width: "140%"
+#| out-width: "120%"
 
@@
 The empirical TLD is used to fit a density distribution using maximum likelihood and moment matching techniques and the Nelder-Mead and Brent methods for direct optimization available within the {fitdistrplus} `R` package [@fitdistrplus_2015].
-Based on goodness-of-fit criteria and diagnostics <!-- should I include the diagonostic plots in the appendix?
+Based on goodness-of-fit criteria and diagnostics <!-- should I include the diagnostic plots in the appendix?
 q-q, etc..?
@@
 ```
-<!--maybe bigger font here?-->
 
@@
 Hansen-type accessibility is not designed to preserve the number of opportunities in the region, it simply counts the intensity of opportunities that those in a zone can potentially interact with (weighted by the friction of distance).
-Also, as discussed in @soukhovIntroducingSpatialAvailability2023, popular competitive accessibility measures such as the 2SFCA [@shenLocationCharacteristicsInnercity1998; @luoMeasuresSpatialAccessibility2003] are internally inconsistent, and the only way it preserves the number of opportunities is if the effect of the impedance function is ignored when expanding the values of opportunities per capita to obtain the total number of opportunities.
+Also, as discussed in @soukhovIntroducingSpatialAvailability2023, popular competitive accessibility measures such as the 2SFCA [@josephMeasuringPotentialPhysical1982; @weibullAxiomaticApproachMeasurement1976; @shenLocationCharacteristicsInnercity1998; @luoMeasuresSpatialAccessibility2003] are internally inconsistent, and the only way it preserves the number of opportunities is if the effect of the impedance function is ignored when expanding the values of opportunities per capita to obtain the total number of opportunities.
 On the other hand, the proportional allocation procedure associated with calculating multi-modal spatial availability $V_i^m$ consistently returns a number of opportunities available to populations by mode that matches the total number of opportunities in the region when summed.
-By doing this consistently, it is possible to define a measure of multi-modal spatial availability per capita ($P$) as presented in Equation (\ref{eq:SA-per-capita}) for use as a benchmark to compare against the regional opportunities per capita ($\frac{\sum_{j} O_j}{\sum_{i} P_i}$).
+By doing this consistently, it is possible to define a measure of multi-modal spatial availability per capita as presented in Equation (\ref{eq:SA-per-capita}) for use as a benchmark to compare against the regional opportunities per capita ($\frac{\sum_{j} O_j}{\sum_{i} P_i}$).
 
@@
 This paper uses the multi-modal spatial availability $V_i^m$ and spatial availability per student $v_i^m$ measures to quantify accessibility changes for the following reasons.
-First, all residential locations are associated to their respective census DA, TTS TAZ, and community boundary (@fig-Fig3) based on spatial location.
-Each parcel is assigned a number of 'potential' motorized and non-motorized student aged population from these different data sources to calculate a motorized and non-motorized population balancing factor $F_{i}^{pm}$.
+
+First, all residential parcels are associated to their respective census DA, TTS TAZ, and community boundary based on spatial location (aggregated by DA and shown as a per capita rate in the second row of @fig-Fig2).
+Each parcel is assigned a number of 'potential' motorized and non-motorized student aged population (from DA) and modal share (from TTS) to calculate a motorized and non-motorized population balancing factor $F_{i}^{pm}$.
 For each parcel the following is retained: the 2011 and 2016 census average elementary-aged student population per parcel (@fig-Fig2 second row) and the average motorized and non-motorized split as informed by the TTS aggregated by community (@fig-Fig3).
+
 Second, the school capacity $O_j$ is estimated and associated with each school (@fig-Fig1).
@@
 
-Third, the motorized and non-motorized impedance-balancing factors $F_{ij}^{cm}$ are calculated using a mode-specific impedance function (@fig-Fig4) based on the respective travel times for each origin (parcel) to school combination.
+Third, the motorized and non-motorized impedance-balancing factors $F_{ij}^{cm}$ are calculated using a mode-specific impedance function (@fig-Fig4) based on the respective estimated travel times for each origin (parcel) to school combination (OD flows retained from the TTS).
+Using network estimated travel time sensitive to observed parcel origin to school destination flows conceptually addresses the aggregation error that could result from using less granular zonal units to represent origin/destinations [@kaneParcelsPointsProximity2020; @kwanSpaceTimeIntegralMeasures1998; @hewkoMeasuringNeighbourhoodSpatial2002].
 Finally, outputs from all three stages are joined together.
@@
 ```{r}
-#| label: fig-Fig6
-#| fig-cap: "Multi-modal spatial availability (greys) and spatial availability per mode-using student (diverging reds and greens) aggregated at the level of DA for 2011 and 2016."
+#| label: fig-Fig5
+#| fig-cap: "Multi-modal spatial availability (purples) and spatial availability per mode-using student (diverging reds and greens) aggregated at the DA level for 2011 and 2016.
+Scales are represented in quartiles."
 #| echo: false
-#| out-width: "140%"
+#| out-width: "120%"
 #| 
@@
 
-<!--I'll make this plot longer-->
-
-The motorized and non-motorized spatial availabilities of school seats $V_i$ for both 2011 and 2016 are presented in the top half of @fig-Fig6.
-Though they show different magnitudes, both populations using non- and motorized modes have higher spatial availability when they are more proximate to schools.
+The motorized and non-motorized spatial availabilities of school seats $V_i$ for both 2011 and 2016 are presented in the top half of @fig-Fig5.
+It is worth noting that the sum of $V_i^m$ values in a year for both motorized and non-motorized population equals the city-wide OTGC (across both public and catholic school boards) in that year.
+This is a result of $V_i^m$ proportional allocation mechanism, singly-constraining opportunities through the proportion allocation balancing factors.
+In this way, $V_i^m$ results can be interpreted as a proportion of the total OTGC in that year.
+Though the top four plots of @fig-Fig5 demonstrate different magnitudes, both populations using non- and motorized modes have higher spatial availability when they are more proximate to schools.
 Hence, all DAs have higher values within Hamilton Central and those more proximate to schools in less-urban communities.
 Trends appear similar for both 2011 and 2016: proximate populations using motorized modes benefit from the ability to reach more distant destinations than non-motorized modes.
-When $V_i^m$ is divided by the number of students by mode in that DA, $v_i^m$ is derived (bottom half of @fig-Fig6).
-A similar spatial distribution seen in $V_i^m$ can be seen in $v_i^m$ plots, but as a rate of student population.
-If our aim is to provide sufficient availability to schools for both motorized and non-motorized populations (i.e., greater than 1.0 $v_i^m$ school-seat availability per student), this rate is only achieved in the core of the city for motorized modes, and notably only for non-motorized populations in DAs that are in less-densely populated rural areas that are school proximate and certain pockets of Hamilton Central.
+
+When $V_i^m$ is divided by the number of students by mode in that DA, $v_i^m$ is derived (bottom four plots in @fig-Fig5).
+The $v_i^m$ plots demonstrate different spatial distribution trends than $V_i^m$ plots.
+If our aim is to provide sufficient availability to schools for both motorized and non-motorized populations (i.e., greater than 1.0 $v_i^m$ school-seat availability per student), this rate is only achieved in the core of the city for motorized modes (greens and yellows).
+Notably, this rate only is available to non-motorized populations in DAs that are in less-densely populated rural areas that are school proximate and very few pockets of Hamilton Central.
 The majority of schools closed were within Hamilton Central, so non-motorized populations in DAs within Hamilton Central were the most negatively impacted by losses in availability while certain motorized populations captured these losses as availability gains.
@@
 ```{r}
-#| label: fig-Fig7
-#| fig-cap: "Change in multi-modal spatial availability and spatial availability per mode-using student between 2016 and 2011 at the level of DA of the 2016 Canadian Census."
+#| label: fig-Fig6
+#| fig-cap: "Change in multi-modal spatial availability and spatial availability per mode-using student between 2016 and 2011 at the 2016 DA Canadian Census aggregation.
+Scales are represented in quartiles."
 #| echo: false
@@
 ```
-
-<!--The colours combos I choose here are absolutely ugly ;( but besides that, do you agree that V_im and v_im should use different colour palettes?-->
-
-To emphasize how spatial availability changed between 2016 and 2011, the change in $V_i^m$ and $v_i^m$ is visualised for both modes in @fig-Fig7.
+<!-- TO DO: Fig6 will be 8 plots (1 for each mode) of 1) change in V_im, v_im, active transportation minutes, GHG emissions. The geometry will be in 2016 DAs. -->
+
+To emphasize how spatial availability changed between 2016 and 2011, the change in $V_i^m$ and $v_i^m$ is visualised for both modes in @fig-Fig6.
 Overall, a decrease in both $V_i^m$ and $v_i^m$ can be seen across both modes around closed schools (black points).
@@
 
-<!--Antonio, what do you think, maybe worth adding a plot of GHG emissions or change in tt?
-This will provide us some higher level summary stats.
-Moataz asked about this as well regarding the figures in the Results section.
-[AP] Yes, please add those statistics to support the discussion.
+<!-- TO DO: add tables to show availability by mode in 2011 and 2016 with changes, as well as lost minutes of active travel time, emissions, and LIM-AT.
+Tabulate by communtity and city-wide.
+[AP]: Check Elise's equity in accessibility to bike share paper, see table 1 there.
 -->
-
-<!--Antonio, what do you think, add a table with summaries of LIMAT, Kid pop, V_im change, v_im change, total travel time change, and emission changed by community and mode?
-We should discuss to decide what's the best presentation.
-[AP]: Best presentation: spatial availability by quartiles of the relevant dimensions (e.g., LIMAT) -->
 
@@
 
-<!-- Is this worth further developing or perhaps just remove, what do you think?
-I think further.
-[AP]: tables to show availability by mode in 2011 and 2016 with changes, as well as lost minutes of active travel time and emissions.
-Check Elise's equity in accessibility to bike share paper, see table 1 there.
--->
-Study limitations:
-
-- potential students per households, potential trips to all schools
-- estimated travel times, direct car or walk trip (shortest path).
-Not necessarily car, could be transit or school bus.
-Walking shortest-paths are also optimistic for children's travel: could be unsafe and not taken in reality and/or alternative more pleasant paths are taken.
-- estimated emissions. GHG emissions per minute of driving varies extensively depending on operating conditions and vehicle characteristics [@soukhovOccupancyGHGEmissions2022].
-
-<!-- Add this additional paper: it discusses the potential impact of mode-switching on GHG emissions reductions [@pantelakiImpactHomeschoolCommuting2024].
-
-Also, access to schools has broader implications; it has a relationship to residential and job choices [@pietrabissaSchoolAccessCity2023] and housing prices [@merrallWhatSchoolWorth2023] ?
-Shall I further elaborate in the discussion?-->
+The advantage of spatial availability is the constrained opportunities (singly constrained). This means we can:
+- calculate availability at the finest spatial resolution
+- we can aggregate it whichever spatial unit is most meaningful for interpretation.
+Here we demonstrate aggregations at the DA, then at neighbourhood community boundaries, and then city wide per mode.
+This can be used to communicate changes in spatial availability and other related potential mobility externalities like lose active transportation minutes and emissions.
 
@@
 
+Study limitations:
+
+- potential students per households, potential trips to all schools
+- estimated travel times, direct car or walk trip (shortest path).
+Not necessarily car, could be transit or school bus.
+Walking shortest-paths are also optimistic for children's travel: could be unsafe and not taken in reality and/or alternative more pleasant paths are taken.
+- estimated emissions. GHG emissions per minute of driving varies extensively depending on operating conditions and vehicle characteristics [@soukhovOccupancyGHGEmissions2022].
+
+<!-- Add this additional paper: it discusses the potential impact of mode-switching on GHG emissions reductions [@pantelakiImpactHomeschoolCommuting2024].
+
+Also, access to schools has broader implications; it has a relationship to residential and job choices [@pietrabissaSchoolAccessCity2023] and housing prices [@merrallWhatSchoolWorth2023] ?
+Shall I further elaborate in the discussion?-->
+
 # References

### 2024-05-02 paper/Manuscript.qmd
Final draft for AP review

@@
     email: "soukhoa@mcmaster.ca"
-    affiliation: McMaster University SEES
-    footnote: "Corresponding Author"
+    affiliations:
+      - id: 1
+        address: "School of Earth, Environment and Society, McMaster University, 1241 Main St. West, Hamilton, ON, L8S 4K1, Canada"
+    attributes:
+      corresponding: true
   - name: "Christopher D. Higgins"
     email: "cd.higgins@utoronto.ca"
-    affiliation: University of Toronto
+    affiliations:
+      - id: 2
+        address: "Department of Human Geography, University of Toronto Scarborough, 1265 Military Trail, Toronto, ON, M1C 1A4, Canada"
+  - name: "Antonio Paez"
+    email: "paezha@mcmaster.ca"
+    affiliations:
+      - id: 1
+        address: "School of Earth, Environment and Society, McMaster University, 1241 Main St. West, Hamilton, ON, L8S 4K1, Canada"
   - name: "Moataz Mohamed"
     email: "mmohame@mcmaster.ca "
-    affiliation: McMaster University CE
-  - name: "Antonio Paez"
-    email: "paezha@mcmaster.ca"
-    affiliation: McMaster University SEES
-address:
-  - code: McMaster University SEES
-    address: "School of Earth, Environment and Society, McMaster University, 1241 Main St. West, Hamilton, ON, L8S 4K1, Canada"
-  - code: McMaster University CE
+    affiliations:
+      - id: 3
         address: "Department of Civil Engineering, McMaster University, 1241 Main St. West, Hamilton, ON, L8S 4K1, Canada"
-  - code: University of Toronto
-    address: "Department of Human Geography, University of Toronto Scarborough, 1265 Military Trail, Toronto, ON, M1C 1A4, Canada"  
 abstract: | 
@@
 For this research, we quantify how these closures changed the accessibility landscape for the elementary school aged population, and the active travel and emission implications.
-In terms of methods, we use spatial availability, a singly-constrained multi-modal measure of spatial accessibility.
+In terms of methods, we use spatial availability, a singly-constrained multimodal measure of spatial accessibility.
 Thanks to its proportional allocation mechanism, spatial availability allows for an interpretable comparison between 'before' and 'after' the implementation of school closure/consolidation policy.
@@
 library(units) # Measurement Units for R Vectors
+library(miscTools) # Miscellaneous Tools and Utilities
+library(huxtable)
 ```
@@
 ```{r, include=FALSE}
-#boundaries for mapping
-hamilton_communities <- st_read(paste0(here::here(),"/data-prep/data-inputs/Boundaries/Community_Boundaries.shp")) |> mutate(COMMUNITY_ = ifelse(COMMUNITY_ == "Hamilton", "Hamilton-Central", COMMUNITY_))
-
-ham_bound <- st_read(paste0(here::here(),"/data-prep/data-inputs/Boundaries/City_Boundary.shp"))
-
+#lake and bay for mapping purposes
 hydro_p_LakeOntario <- st_read(paste0(here::here(),"/data-prep/data-inputs/Boundaries/hydro_p_LakeOntario.shp")) |> st_transform(crs=4326)
@@
 ham_bay <- st_read(paste0(here::here(),"/data-prep/data-inputs/Boundaries/Waterbodies.shp")) |> st_transform(crs=4326) |> filter(FEATURE_TY == "Lake")
-
-ham_streets <- st_read(paste0(here::here(),"/data-prep/data-inputs/Boundaries/Street_Centreline.shp")) |> st_transform(crs=4326) |> filter(ROAD_CLASS %in% c("Provincial Highway", "Parkway", "Arterial")) #no collectors, arterial, Local
 ```
@@
 ham_city_bound <- st_read(paste0(here::here(),"/data-prep/data-inputs/Boundaries/City_Boundary.shp"),
-                          quiet = TRUE) |> st_transform(st_crs(cen_data_2011_hamiltononly))
-ham_community_bounds <- st_read(paste0(here::here(),"/data-prep/data-inputs/Boundaries/Community_Boundaries.shp"),
-                                quiet = TRUE) |> st_transform(st_crs(cen_data_2011_hamiltononly))
+                          quiet = TRUE) |> st_transform(st_crs(cen_data_2016_hamiltononly))
+ham_community_bounds <- st_read(paste0(here::here(),"/data-prep/data-inputs/Boundaries/Community_Boundaries.shp"),quiet = TRUE) |> 
+  st_transform(st_crs(cen_data_2016_hamiltononly)) |> 
+  mutate(COMMUNITY = case_when(COMMUNITY_ == "Ancaster" ~ "AN",
+                               COMMUNITY_ == "Dundas" ~ "DU",
+                               COMMUNITY_ == "Flamborough" ~ "FL",
+                               COMMUNITY_ == "Glanbrook" ~ "GL",
+                               COMMUNITY_ == "Hamilton" ~ "HA",
+                               COMMUNITY_ == "Stoney Creek" ~ "SC"),
+         COMMUNITY_ = ifelse(COMMUNITY_ == "Hamilton", "Hamilton Central", COMMUNITY_)) |> st_transform(st_crs(cen_data_2016_hamiltononly))
 ```
@@
 library(hamONdests) # A R package of Hamilton, Ontario destinations
+
 data(Schools_2011_2016)
@@
 rm("Schools_2011_2016")
+
+# add community label to schools
+ALL_SCHOOLS_Elem <- st_intersection(ham_community_bounds,
+                        ALL_SCHOOLS_Elem |> st_transform(st_crs(cen_data_2016_hamiltononly)))
+
 data(School_Catchments_2011_2016)
@@
 catchment_schools_plot <-
-   tm_shape(ham_bay, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0) +
-  tm_shape(hydro_p_LakeOntario, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0)+
+   tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0) +
+  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0)+
   tm_shape(ALL_SCHOOLS_CATCH_Elem_extra) +
@@
           title.shape = "School status",
-          shapes = c(24,25,22,21,23),
+          shapes = c(24,25,21,23,22),
           shapes.labels = c(
@@
   tmap_defaults +
-  tm_layout(bg.color = "grey",
+  tm_layout(bg.color = "grey85",
             panel.labels = list(c('HWDCSB (Catholic Public)','HWDSB (Public)'),
@@
 ```
-
-<!-- Number of schools?
-Yes Also, make the symbols that map the schools bigger, and use different shapes in addition to different colors.
-As an alternative, the OTGC could be the color, and the school category the shape.
--->
 
@@
          v_im_kid_popadj = V_im/kid_popadj_m,
+         v_im_kid_popadj = ifelse(is.na(v_im_kid_popadj) | is.nan(v_im_kid_popadj), 0, v_im_kid_popadj),
          v_im_kid_popadj_LIMAT = v_im_kid_popadj*(DA_prev_LIMAT_under18/100),
@@
             c_total_tt = total_tt.y - total_tt.x, #2016 minus 2011 tt
+            c_total_tt = ifelse(is.na(c_total_tt), -total_tt.x, c_total_tt),
             c_total_ttperkid = (total_tt.y/kid_popadj_m.y) - (total_tt.x/kid_popadj_m.x),
+            c_total_ttperkid = ifelse(is.na(c_total_ttperkid), -(total_tt.x/kid_popadj_m.x), c_total_ttperkid),
             c_V_im = V_im.y -V_im.x,
+            c_V_im = ifelse(is.na(c_V_im), -V_im.x,c_V_im),
             c_V_im_perc = (V_im.y -V_im.x)/V_im.x,
@@
             c_DA_prev_LIMAT_under18 = DA_prev_LIMAT_under18.y - DA_prev_LIMAT_under18.x,
+            emissions_total_car_tt_2016 = emissions_total_car_tt.y,
+            emissions_total_car_tt_2011 = emissions_total_car_tt.x,
             c_emissions_total_car_tt = emissions_total_car_tt.y - emissions_total_car_tt.x,
+            c_emissions_total_car_tt = ifelse(is.na(c_emissions_total_car_tt), -emissions_total_car_tt.x, c_emissions_total_car_tt),
+            emissions_perkid_2011 = emissions_total_car_tt.x/kid_popadj_m.x,
+            emissions_perkid_2016 = emissions_total_car_tt.y/kid_popadj_m.y,
+            c_emissions_perkid = ((emissions_perkid_2016 - emissions_perkid_2011)/
+                                      emissions_perkid_2011),
+            c_emissions_perkid = ifelse(is.na(emissions_perkid_2016), -emissions_perkid_2011, c_emissions_perkid),
             c_v_im_kid_popadj = v_im_kid_popadj.y - v_im_kid_popadj.x,
+            c_v_im_kid_popadj = ifelse(is.na(c_v_im_kid_popadj), -v_im_kid_popadj.x, c_v_im_kid_popadj),
             c_v_im_kid_popadj_perc = (v_im_kid_popadj.y - v_im_kid_popadj.x)/v_im_kid_popadj.x,
@@
 ```{r}
+#repeat but for all 1 mode
+SAvail_i_DA2011 <- avail_2011 %>%
+  group_by(GeoUID) %>%
+  summarize(COMMUNITY = first(COMMUNITY),
+            total_tt = sum(travel_time_p50*kid_popadj, na.rm = T),
+            V_i = sum(V_ij, na.rm = T),
+            DA_pop_5to14 = mean(DA_pop_5to14, na.rm=T),
+            kid_pop_DAavg = mean(kid_pop, na.rm = T),
+            DA_prev_LIMAT_under18 = mean(DA_prev_LIMAT_under18, na.rm = T),
+            .groups = "drop") |>
+  filter(GeoUID != 35250242 & GeoUID != 35250133 & GeoUID != 35250457 ) #these GeoUID's have 0 kid population, so let's drop so we don't have zeros
+SAvail_i_DA2016 <- avail_2016 %>%
+  group_by(GeoUID) %>%
+  summarize(COMMUNITY = first(COMMUNITY),
+            total_tt = sum(travel_time_p50*kid_popadj),
+            V_i = sum(V_ij, na.rm = T),
+            DA_pop_5to14 = mean(DA_pop_5to14, na.rm=T),
+            kid_pop_DAavg = mean(kid_pop, na.rm = T),
+            DA_prev_LIMAT_under18 = mean(DA_prev_LIMAT_under18, na.rm = T),
+            .groups = "drop") |>
+  filter(GeoUID != 35250398 & GeoUID != 35250133 & GeoUID != 35250427 ) #these GeoUID's have 0 kid population, so let's drop so we don't have zeros
+#add the geometries -- census DAs
+SAvail_i_DA2011 <- SAvail_i_DA2011 |>
+  left_join(cen_data_2011_hamiltononly |> filter(Region.Name == "Hamilton") |> dplyr::select(GeoUID), by="GeoUID") |> mutate(Year = "2011") |> st_set_geometry("geometry")
+SAvail_i_DA2016 <- SAvail_i_DA2016 |>
+  left_join(cen_data_2016_hamiltononly |> filter(Region.Name == "Hamilton") |> dplyr::select(GeoUID), by="GeoUID") |> mutate(Year = "2016") |> st_set_geometry("geometry")
+#join them together!
+SAvail_i_DA_extra <- rbind(SAvail_i_DA2011, SAvail_i_DA2016)
+rm(SAvail_i_DA2011,SAvail_i_DA2016)
+```
+
+```{r}
 #calc changes between the years - but between communities with geom for mapping
-ham_community_bounds <- ham_community_bounds |> 
-  mutate(COMMUNITY = case_when(COMMUNITY_ == "Ancaster" ~ "AN",
-                               COMMUNITY_ == "Dundas" ~ "DU",
-                               COMMUNITY_ == "Flamborough" ~ "FL",
-                               COMMUNITY_ == "Glanbrook" ~ "GL",
-                               COMMUNITY_ == "Hamilton" ~ "HA",
-                               COMMUNITY_ == "Stoney Creek" ~ "SC"))
-
 SAvail_im_COMM2011 <- avail_2011 %>%
@@
             kid_popadj_m = sum(kid_popadj),
-            DA_pop_5to14 = mean(DA_pop_5to14, na.rm=T),
-            DA_prev_LIMAT_under18 = mean(DA_prev_LIMAT_under18),
             .groups = "drop") |>
   mutate(emissions_total_car_tt = ifelse(mode == "c", 0.1*total_tt, 0),
-         v_im_kid_popadj = V_im/kid_popadj_m,
-         v_im_kid_popadj_LIMAT = v_im_kid_popadj*(DA_prev_LIMAT_under18/100),
-         v_im_kid_popadj_LIMAT = ifelse(is.na(v_im_kid_popadj_LIMAT) | is.nan(v_im_kid_popadj_LIMAT), 0, v_im_kid_popadj_LIMAT),
-         V_i_kid_popadj_LIMAT = V_im*(DA_prev_LIMAT_under18/100),
-         V_i_kid_popadj_LIMAT = ifelse(is.na(V_i_kid_popadj_LIMAT) | is.nan(V_i_kid_popadj_LIMAT), 0, V_i_kid_popadj_LIMAT))
+         v_im_kid_popadj = V_im/kid_popadj_m)
 
@@
             kid_popadj_m = sum(kid_popadj),
-            DA_pop_5to14 = mean(DA_pop_5to14, na.rm=T),
-            DA_prev_LIMAT_under18 = mean(DA_prev_LIMAT_under18),
             .groups = "drop") |>
@@
          v_im_kid_popadj = V_im/kid_popadj_m,
-         v_im_kid_popadj = ifelse(is.na(v_im_kid_popadj) | is.nan(v_im_kid_popadj), 0, v_im_kid_popadj),
-         v_im_kid_popadj_LIMAT = v_im_kid_popadj*(DA_prev_LIMAT_under18/100),
-         v_im_kid_popadj_LIMAT = ifelse(is.na(v_im_kid_popadj_LIMAT) | is.nan(v_im_kid_popadj_LIMAT), 0, v_im_kid_popadj_LIMAT),
-         V_i_kid_popadj_LIMAT = V_im*(DA_prev_LIMAT_under18/100),
-         V_i_kid_popadj_LIMAT = ifelse(is.na(V_i_kid_popadj_LIMAT) | is.nan(V_i_kid_popadj_LIMAT), 0, V_i_kid_popadj_LIMAT))
+         v_im_kid_popadj = ifelse(is.na(v_im_kid_popadj) | is.nan(v_im_kid_popadj), 0, v_im_kid_popadj))
 
@@
             c_kid_popadj_m = kid_popadj_m.y - kid_popadj_m.x,
-            c_DA_pop_5to14 = DA_pop_5to14.y - DA_pop_5to14.x,
-            c_DA_prev_LIMAT_under18 = DA_prev_LIMAT_under18.y - DA_prev_LIMAT_under18.x,
             c_emissions_total_car_tt = emissions_total_car_tt.y - emissions_total_car_tt.x,
@@
             c_v_im_kid_popadj_perc = (v_im_kid_popadj.y - v_im_kid_popadj.x)/v_im_kid_popadj.x,
-            c_v_im_kid_popadj_LIMAT = v_im_kid_popadj_LIMAT.y - v_im_kid_popadj_LIMAT.x,
-            c_V_i_kid_popadj_LIMAT = V_i_kid_popadj_LIMAT.y - V_i_kid_popadj_LIMAT.x,
             geometry,
-            # kid_2011 = kid_popadj_m.x,
-            # kid_2016 = kid_popadj_m.y,
-            # c_kid_perc = ((kid_popadj_m.y - kid_popadj_m.x)/kid_popadj_m.x) |> scales::percent(),
-            # kid_proportion_2011 = (kid_2011/sum(kid_2011)),
-            # kid_proportion_2016 = (kid_2016/sum(kid_2016)),
-            # c_kid_proportion_perc = ((kid_proportion_2016 - kid_proportion_2011)/kid_proportion_2011)
-            # %>% scales::percent(),
-            # kid_proportion_2011_perc = (kid_proportion_2011) |> scales::percent(),
-            # kid_proportion_2016_perc = (kid_proportion_2016) |> scales::percent(),
-            # V_im_2011 = V_im.x,
-            # V_im_2016 = V_im.y,
-            # c_V_im_perc = ((V_im.y -V_im.x)/V_im.x) |> scales::percent(),
-            # V_im_proportion_2011 = (V_im.x/sum(V_im.x)),
-            # V_im_proportion_2016 = (V_im.y/sum(V_im.y)),
-            # c_V_im_proportion_perc = ((V_im_proportion_2016 - V_im_proportion_2011)/V_im_proportion_2011)
-            # %>% scales::percent(),
-            # V_im_proportion_2011_perc = (V_im_proportion_2011) |> scales::percent(),
-            # V_im_proportion_2016_perc = (V_im_proportion_2016) |> scales::percent(),
-            # v_im_2011 = v_im_kid_popadj.x,
-            # v_im_2016 = v_im_kid_popadj.y,
-            # c_v_im_perc = ((v_im_kid_popadj.y - v_im_kid_popadj.x)/v_im_kid_popadj.x) |> scales::percent(),
-            # total_tt_2011 = total_tt.x,
-            # total_tt_2016 = total_tt.y,
+            kid_2011 = kid_popadj_m.x,
+            kid_2016 = kid_popadj_m.y,
+            c_kid_perc = ((kid_popadj_m.y - kid_popadj_m.x)/kid_popadj_m.x) |> scales::percent(accuracy = 2L),
+            kid_proportion_2011 = (kid_2011/sum(kid_2011)),
+            kid_proportion_2016 = (kid_2016/sum(kid_2016)),
+            c_kid_proportion_perc = ((kid_proportion_2016 - kid_proportion_2011)/kid_proportion_2011)
+            %>% scales::percent(accuracy = 2L),
+            # kid_proportion_2011_perc = (kid_proportion_2011) |> scales::percent(accuracy = 2L),
+            # kid_proportion_2016_perc = (kid_proportion_2016) |> scales::percent(accuracy = 2L),
+            V_im_2011 = V_im.x,
+            V_im_2016 = V_im.y,
+            c_V_im_perc = ((V_im.y -V_im.x)/V_im.x) |> scales::percent(accuracy = 2L),
+            V_im_proportion_2011 = (V_im.x/sum(V_im.x)),
+            V_im_proportion_2016 = (V_im.y/sum(V_im.y)),
+            c_V_im_proportion_perc = ((V_im_proportion_2016 -V_im_proportion_2011)/V_im_proportion_2011) %>% scales::percent(accuracy = 2L),
+            # V_im_proportion_2011_perc = (V_im_proportion_2011) |> scales::percent(accuracy = 2L),
+            # V_im_proportion_2016_perc = (V_im_proportion_2016) |> scales::percent(accuracy = 2L),
+            v_im_2011 = v_im_kid_popadj.x,
+            v_im_2016 = v_im_kid_popadj.y,
+            c_v_im_perc = ((v_im_kid_popadj.y - v_im_kid_popadj.x)/v_im_kid_popadj.x) |> scales::percent(accuracy = 2L),
+            total_tt_2011 = total_tt.x,
+            total_tt_2016 = total_tt.y,
             # total_tt_proportion_2011 = (total_tt.x/sum(total_tt.x)),
@@
             # c_total_tt_proportion_perc = ((total_tt_proportion_2016 - total_tt_proportion_2011)/total_tt_proportion_2011)
-            # %>% scales::percent(),
-            # total_tt_proportion_2011 = (total_tt_proportion_2011) |> scales::percent(),
-            # total_tt_proportion_2016 = (total_tt_proportion_2016) |> scales::percent(),
-            # c_total_tt_perc = ((total_tt.y - total_tt.x)/
-            #                      total_tt.x)|> scales::percent(),
-            # total_tt_perkid_2011 = total_tt.x/kid_2011,
-            # total_tt_perkid_2016 = total_tt.y/kid_2016,
-            # c_total_ttperkid_perc = ((total_tt_perkid_2016 - total_tt_perkid_2011)/
-            #                            total_tt_perkid_2011)|> scales::percent(),
-            # emissions_2011 = emissions_total_car_tt.x,
-            # emissions_2016 = emissions_total_car_tt.y,
-            # c_emissions_perc = ((emissions_total_car_tt.y - emissions_total_car_tt.x)/
-            #                       emissions_total_car_tt.x)|> scales::percent(),
+            # %>% scales::percent(accuracy = 2L),
+            # total_tt_proportion_2011 = (total_tt_proportion_2011) |> scales::percent(accuracy = 2L),
+            # total_tt_proportion_2016 = (total_tt_proportion_2016) |> scales::percent(accuracy = 2L),
+            c_total_tt_perc = ((total_tt.y - total_tt.x)/
+                                  total_tt.x)|> scales::percent(accuracy = 2L),
+            total_tt_perkid_2011 = total_tt.x/kid_2011,
+            total_tt_perkid_2016 = total_tt.y/kid_2016,
+            c_total_ttperkid_perc = ((total_tt_perkid_2016 - total_tt_perkid_2011)/
+                                       total_tt_perkid_2011)|> scales::percent(accuracy = 2L),
+            emissions_2011 = emissions_total_car_tt.x,
+            emissions_2016 = emissions_total_car_tt.y,
+            c_emissions_perc = ((emissions_total_car_tt.y - emissions_total_car_tt.x)/emissions_total_car_tt.x)|> scales::percent(accuracy = 2L),
             # emissions_proportion_2011 = (emissions_2011/sum(emissions_2011)),
@@
             # c_emissions_proportion = ((emissions_proportion_2016 - emissions_proportion_2011)/emissions_proportion_2011)
-            # %>% scales::percent(),
-            # emissions_proportion_2011_perc = (emissions_proportion_2011) |> scales::percent(),
-            # emissions_proportion_2016_perc = (emissions_proportion_2016) |> scales::percent(),
-            # emissions_perkid_2011 = emissions_total_car_tt.x/kid_popadj_m.x,
-            # emissions_perkid_2016 = emissions_total_car_tt.y/kid_popadj_m.y,
-            # c_emissions_perkid_perc = ((emissions_perkid_2016 - emissions_perkid_2011)/
-            #                              emissions_perkid_2011)|> scales::percent()
-            ) |> st_as_sf()
+            # %>% scales::percent(accuracy = 2L),
+            # emissions_proportion_2011_perc = (emissions_proportion_2011) |> scales::percent(accuracy = 2L),
+            # emissions_proportion_2016_perc = (emissions_proportion_2016) |> scales::percent(accuracy = 2L),
+            emissions_perkid_2011 = emissions_total_car_tt.x/kid_popadj_m.x,
+            emissions_perkid_2016 = emissions_total_car_tt.y/kid_popadj_m.y,
+            c_emissions_perkid_perc = ((emissions_perkid_2016 - emissions_perkid_2011)/
+                                      emissions_perkid_2011)|> scales::percent(accuracy = 2L)
+            ) |> mutate_if(is.numeric, round, 1) |> st_as_sf()
 ```
@@
 student_pop_absolute_plot <-
-    tm_shape(ham_bay, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0) +
-  tm_shape(hydro_p_LakeOntario, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0)+
+    tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0) +
+  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0)+
   tm_shape(SAvail_i_DA_extra |> mutate(extra_col_for_panel = "HAMILTON")) +
@@
   tmap_defaults +
-  tm_layout(bg.color = "grey",
+  tm_layout(bg.color = "grey85",
             panel.label.bg.color = "white",
@@
 student_pop_rate_plot <-
-  tm_shape(ham_bay, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0) +
-  tm_shape(hydro_p_LakeOntario, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0)+
+  tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0) +
+  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0)+
   tm_shape(SAvail_i_DA_extra) +
@@
   tmap_defaults +
-  tm_layout(bg.color = "grey",
+  tm_layout(bg.color = "grey85",
             panel.show=FALSE)
@@
 LIMAT_under18_DA_plot <-
-  tm_shape(ham_bay, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0) +
-  tm_shape(hydro_p_LakeOntario, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0)+
+  tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0) +
+  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0)+
   tm_shape(SAvail_i_DA_extra) +
@@
   tmap_defaults  +
-  tm_layout(bg.color = "grey",
+  tm_layout(bg.color = "grey85",
             panel.show=FALSE)
@@
 ```{r student-popabs-lowincperc-poprate-plot2, eval=FALSE}
-tmap_save(stud_pop_rate_LIM_plot, paste0(here::here(),"/Figures/stud_pop_rate_LIM_plot.png"), height=10.5, width=12, dpi=600)
+tmap_save(stud_pop_rate_LIM_plot, paste0(here::here(),"/Figures/stud_pop_rate_LIM_plot.png"), height=10, width=12, dpi=600)
 ```
@@
 data(ggh_pd)
-# 
-# #Read the boundaries of the traffic analysis zones (retrieved from the TTS 2016):
-# ggh_taz <- st_read(paste0(here::here(), "/data-prep/data-inputs/TTS/tts06_83_region.shp"),
-#                    quiet = TRUE) %>% 
-#   st_transform(crs = 32617)
-# 
-# # Prepare traffic analysis zones:
-# ggh_taz <- ggh_taz %>%
-#   transmute(GTA06 = as.character(GTA06),
-#             AREA = st_area(geometry) %>% 
-#               set_units(km^2) %>% 
-#               drop_units())
-# #Extract the IDs for all TAZ:
-# ghta_taz_id <- ggh_taz$GTA06
-# 
-# #Read the boundaries of the traffic analysis zones (retrieved from the TTS 2016):
-# ggh_pd <- st_read(paste0(here::here(), "/data-prep/data-inputs/TTS/tts06_pd_83.shp"),
-#                   quiet = TRUE) %>%
-#   st_transform(crs = 32617)
-# 
-# #Adding region names to the object as they are not included in the origin data: (http://dmg.utoronto.ca/pdf/tts/2016/2016TTS_Conduct.pdf)
-# 
-# ggh_pd <- ggh_pd %>%
-#   mutate(REGION_name = 
-#            ifelse(REGION == 1, "Toronto", 
-#                   ifelse(REGION == 2, "Durham", 
-#                          ifelse(REGION == 3, "York", 
-#                                 ifelse(REGION == 4, "Peel", 
-#                                        ifelse(REGION == 5, "Halton", 
-#                                               ifelse(REGION == 6, "Hamilton", 
-#                                                      ifelse(REGION == 11, "Niagara", 
-#                                                             ifelse(REGION == 12, "Waterloo", 
-#                                                                    ifelse(REGION == 13, "Guelph", 
-#                                                                           ifelse(REGION == 14, "Wellington", 
-#                                                                                  ifelse(REGION == 15, "Orangeville", 
-#                                                                                         ifelse(REGION == 16, "Barrie", 
-#                                                                                                ifelse(REGION == 17, "Simcoe", 
-#                                                                                                       ifelse(REGION == 18, "Kawartha Lakes", 
-#                                                                                                              ifelse(REGION == 19, "Peterborough City", 
-#                                                                                                                     ifelse(REGION == 20, "Peterborough County", 
-#                                                                                                                            ifelse(REGION == 21, "Orillia", 
-#                                                                                                                                   ifelse(REGION == 22, "Dufferin", 
-#                                                                                                                                          ifelse(REGION == 23, "Brantford", 
-#                                                                                                                                                 ifelse(REGION == 24, "Brant", "NOOOO")))))))))))))))))))))
-# 
-# # add 'REGION' from the planning distract object to the `ggh_taz`. 
-# #buffering to clean up the edges of these objects -> there wre issues using st_join. 
-# ggh_taz <- st_buffer(ggh_taz, dist=0)
-# ggh_pd <- st_buffer(ggh_pd, dist=0)
-# 
-# ggh_taz <- st_join(ggh_taz, ggh_pd %>% transmute(REGION),left = TRUE, largest = TRUE)
 
@@
 car_OD_flow_plot <-  
-    tm_shape(ham_bay, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0) +
-  tm_shape(hydro_p_LakeOntario, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0)+
+    tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0) +
+  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0)+
   tm_shape(od_trips_perc_car |> mutate(mode = "motorized"), bbox=SAvail_i_DA_extra) +
@@
 
-walk_OD_flow <- tm_shape(ham_bay, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0) +
-  tm_shape(hydro_p_LakeOntario, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0)+
+walk_OD_flow <- tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0) +
+  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0)+
   tm_shape(od_trips_perc_walk |> mutate(mode = "non-motorized"), bbox=SAvail_i_DA_extra) +
@@
 ```{r ploting-empirical-and-theo-curve-TLD-tts2, eval=FALSE}
-ggsave(paste0(here::here(),"/Figures/TLD_empirical_wtheo_curve_plot.png"), height=7, width=12, dpi=600)
+ggsave(paste0(here::here(),"/Figures/TLD_empirical_wtheo_curve_plot.png"), height=7, width=10, dpi=600)
 ```
@@
 Vim_20112016_plot <-  
-  tm_shape(ham_bay, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0) +
-  tm_shape(hydro_p_LakeOntario, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0)+
+  tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0) +
+  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0)+
   tm_shape(SAvail_im_DA, bbox=SAvail_im_DA) +
@@
 vvim_20112016_plot <-  
-  tm_shape(ham_bay, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0) +
-  tm_shape(hydro_p_LakeOntario, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0)+
+  tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0) +
+  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0)+
   tm_shape(SAvail_im_DA, bbox=SAvail_im_DA) +
@@
 Vim_vvim_20112016_plot <- tmap_arrange(Vim_20112016_plot, vvim_20112016_plot, nrow=2)
-```
-
-```{r, eval=FALSE}
 tmap_save(Vim_vvim_20112016_plot, paste0(here::here(),"/Figures/Vim_vvim_20112016_plot.png"), height=12, width=12, dpi=600)
@@
 Vim_20112016_CHANGE_plot <-  
-  tm_shape(ham_bay, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0) +
-  tm_shape(hydro_p_LakeOntario, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0)+
+  tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0) +
+  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0)+
   tm_shape(SAvail_im_DA_changes, bbox=SAvail_im_DA) +
   tm_polygons(col = "c_V_im", border.alpha = 0, border.col = "grey20", lwd=0.2,
-              palette = "RdYlBu", midpoint=0,
+              palette = "RdBu", #midpoint=0,
               # style="quantile", n=4,
-              breaks = c(-5, -4.1, -0.2, 0, 5),
-              labels = c("[-208 to -4.1)", "[-4.1 to -0.2)", "[-0.2 to 0)","[0 to 283.6)"),
+              breaks = c(-20, -4.1, -0.2, 0, 20),
+              labels = c("[-208 to -4.1)", "[-4.1 to -0.2)", "[-0.2 to 0)","[0 to 283.6]"),
               title = expression("Change in "*V[i]^m),
@@
           title.shape = "Schools that changed status",
-          shapes = c(22,23),
-          shapes.labels = c(
-          "New or expanded after 2011",
-          "Closed before 2016"),
+          shapes = c(22,21),
+          # shapes.labels = c(
+          # "New or expanded after 2011",
+          # "Closed before 2016"),
           legend.shape.show = FALSE) +
-  tm_facets(by=c("Year"))+
   tmap_defaults +
-  tm_layout( bg.color = "grey",
+  tm_layout( bg.color = "grey85",
              panel.label.bg.color = "white",
@@
 vvim_20112016_CHANGE_plot <-
-  tm_shape(ham_bay, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0) +
-  tm_shape(hydro_p_LakeOntario, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0)+
+  tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0) +
+  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0)+
   tm_shape(SAvail_im_DA_changes, bbox=SAvail_im_DA) +
   tm_polygons(col = "c_v_im_kid_popadj", border.alpha = 0, border.col = "grey20", lwd=0.2,
-              palette = "RdYlBu", midpoint=0,
-              breaks = c(-0.1, -0.05, 0.00, 0.05, 0.1),
-              labels = c("[>-0.1 to -0.05)", "[-0.05 to 0.00)", "[0.00 to 0.05)", "[0.05 to <0.1)"),
+              palette = "RdBu", midpoint=0,
+              breaks = c(-0.1, -0.08, -0.05, 0, 0.1),
+              labels = c("[-3.77 to -0.08)", "[-0.08 to -0.05)", "[-0.05 to 0)", 
+                         "[0 to 1.46]"),
               title = expression("Change in "*v[i]^m),
-              textNA = "Unavailable",
+              textNA = "No V in 2016 or 2011",
               colorNA = "white")+
@@
           title.shape = "Schools that changed status",
-          shapes = c(22,23),
+          shapes = c(22,21),
           shapes.labels = c(
-          "New or expanded after 2011",
-          "Closed before 2016")
+          "Closed before 2016",
+          "New or expanded after 2011")
   ) +
-  tm_facets(by=c("Year"))+
   tmap_defaults +
-  tm_layout( bg.color = "grey",
+  tm_layout( bg.color = "grey85",
              panel.label.bg.color = "white",
-             panel.label.height = 0.8,
-             panel.show = FALSE)
+             panel.label.height = 0.8)
 
 Vim_vvim_20112016_CHANGE_plot <- tmap_arrange(Vim_20112016_CHANGE_plot, vvim_20112016_CHANGE_plot, nrow=2)
-
 ```
@@
 ttim_20112016_CHANGE_plot <-  
-  tm_shape(ham_bay, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0) +
-  tm_shape(hydro_p_LakeOntario, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0)+
-  tm_shape(SAvail_im_DA_changes, bbox=SAvail_im_DA) +
+  tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0) +
+  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0)+
+  tm_shape(SAvail_im_DA_changes |> filter(mode == "non-motorized"), 
+           bbox=SAvail_im_DA ) +
   tm_polygons(col = "c_total_tt", border.alpha = 0, border.col = "grey20", lwd=0.2,
-              palette = "RdYlBu", midpoint=0,
-             # style="quantile", n=4,
-              breaks = c(-60, -47, -3.1, 0, 10),
-              labels = c("[-1971 to -47)", "[-47 to -3.1)", "[-3.1 to 0)","[0 to 10)"),
-              title = expression("Change in travel time minutes"),
-              textNA = "No V in 2016 or 2011",
-              colorNA = "white")+
-  tm_facets(by = c("mode"), nrow=1)+
+              palette = "RdYlGn", midpoint=0,
+              #style="quantile", n=4, 
+              breaks = c(-10, -5.67, -2.94, 0, 10),
+              labels = c("[-69.0 to -5.7)", "[-5.7 to -2.94)", "[-2.94 to 0)","[0 to 20.81]"),
+              title = c("Non-motorized travel time (minutes)"))+
   tm_shape(ham_community_bounds %>% st_cast("MULTILINESTRING")) + 
@@
           title.shape = "Schools that changed status",
-          shapes = c(22,23),
-          shapes.labels = c(
-          "New or expanded after 2011",
-          "Closed before 2016"),
+          shapes = c(22,21),
+          # shapes.labels = c(
+          # "New or expanded after 2011",
+          # "Closed before 2016"),
           legend.shape.show = FALSE) +
-  tm_facets(by=c("Year"))+
   tmap_defaults +
-  tm_layout( bg.color = "grey",
+  tm_layout( bg.color = "grey85",
              panel.label.bg.color = "white",
-             panel.label.height = 0.8)
-ttim_20112016_CHANGE_plot
-```
-
-```{r, eval=FALSE}
+             panel.label.height = 0.9,
+             panel.labels = "Change in non-motorized minutes between 2011-16",
+             legend.outside = TRUE)
+
 GHGim_20112016_CHANGE_plot <-
-  tm_shape(ham_bay, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0) +
-  tm_shape(hydro_p_LakeOntario, bbox=hamilton_communities) + tm_polygons(col="skyblue", border.alpha = 0)+
-  tm_shape(SAvail_im_DA_changes, bbox=SAvail_im_DA) +
+  tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0) +
+  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0)+
+  tm_shape(SAvail_im_DA_changes |> filter(mode == "motorized"), bbox=SAvail_im_DA) +
   tm_polygons(col = "c_emissions_total_car_tt", border.alpha = 0, border.col = "grey20", lwd=0.2,
-              palette = "RdYlBu", midpoint=0,
-              style="quantile", n=4,
-              #breaks = c(-0.1, -0.05, 0.00, 0.05, 0.1),
-              #labels = c("[>-0.1 to -0.05)", "[-0.05 to 0.00)", "[0.00 to 0.05)", "[0.05 to <0.1)"),
-              title = expression("Change in GHG emissions (kg CO2e for home-school trips)"),
-              textNA = "Unavailable",
+              palette = "-RdYlGn", midpoint=0,
+              #style="quantile", n=4,
+              breaks = c(-200, -15, -4, 10, 200),
+              labels = c("[-197 to -15)", "[-15 to -4)", "[-4 to 0)", "[0 to 10)", "[10 to 1,220]"),
+              title = expression("kg CO"[2]*"e for potential home-school trips"),
+              textNA = "No V in 2016 or 2011",
               colorNA = "white")+
-  tm_facets(by = c("mode"), nrow=1)+
   tm_shape(ham_community_bounds %>% st_cast("MULTILINESTRING")) + 
@@
           title.shape = "Schools that changed status",
-          shapes = c(22,23),
+          shapes = c(22,21),
           shapes.labels = c(
-          "New or expanded after 2011",
-          "Closed before 2016")
+          "Closed before 2016",
+          "New or expanded after 2011")
   ) +
-  tm_facets(by=c("Year"))+
   tmap_defaults +
-  tm_layout( bg.color = "grey",
+  tm_layout( bg.color = "grey85",
              panel.label.bg.color = "white",
-             panel.label.height = 0.8,
-             panel.show = FALSE)
+             panel.label.height = 0.9,
+             panel.labels = "Change in GHG emissions between 2011-16",
+             legend.outside = TRUE)
 
 ttim_GHGim_20112016_CHANGE_plot <- tmap_arrange(ttim_20112016_CHANGE_plot, GHGim_20112016_CHANGE_plot, nrow=2)
-ttim_GHGim_20112016_CHANGE_plot
 ```
@@
           paste0(here::here(),"/Figures/ttim_GHGim_20112016_CHANGE_plot.png"),
-          height=7,
-          width=12, 
+          height=5.5,
+          width=7.5, 
           dpi=600)
@@
 ```{r, eval=FALSE}
-walk_OD_flow <- tm_shape(od_trips_perc_walk |> mutate(mode = "non-motorized"), bbox=SAvail_i_DA_extra) +
-  tm_polygons(col = "TAZOrig_walk_perc", border.alpha = 0.4, border.col = "grey20", lwd=0.2,
-              palette = "Greys",
-              breaks = c(0, .25, .50, .75, 1),
-              title = "Proportion of mode use",
-              textNA = "Unavailable",
-              colorNA = "white",
-              legend.show=FALSE)+
-  #tm_shape(SAvail_i_DA_extra) + tm_borders(alpha = 0.4, col = "cornflowerblue", lwd=0.2)+
-  tm_shape(ham_taz) + tm_borders(alpha=0.4, col = "grey40", lwd=0.2) +
-  tm_shape(ham_community_bounds) + tm_borders(alpha = 0.8, col = "lightblue", lwd=2)+
-  tm_shape(ham_taz_centroids) + tm_dots(alpha=0.5, col="darkblue", size=0.01) +
-  tm_shape(desire_lines_ham |> filter(TAZOrig_walk_trips >= 1) 
-           |> mutate(mode = "non-motorized")) +
-  tm_lines(palette = "Greens", breaks = c(12, 23, 44, 67, 174),
-           title.col = "Number of trips (non-motorized)",
-           col = "TAZOrig_walk_trips",
-           alpha = 0.9,
-           scale = 4,
-           lwd = "all",
-           legend.lwd.show = FALSE) +
-  tm_facets(by = c("mode","Year"), nrow=1)+
-  tmap_defaults +
-  tm_layout(panel.label.height = c(0.8,0,0)) +
-  tm_add_legend(type = "line", 
-                col = c("grey40","lightblue"),
-                labels = c("TAZ boundary","Community boundary"))+
-  tm_add_legend(type = "symbol",
-                col = "darkblue",
-                labels = "TAZ centroid")
-OD_flow_plot <- tmap_arrange(car_OD_flow_plot, walk_OD_flow, nrow = 2)
-tmap_save(OD_flow_plot, 
-          paste0(here::here(),"/Figures/SA-and-SAperStud-20112016.png"),
-          height=7, 
-          width=12,
-          dpi=600)
-```
-
-```{r, eval=FALSE}
-SA_Vim_2011_plot <- ggplot() +
-  geom_sf(data = SAvail_im_DA2011, aes(fill= V_im), color = NA) +
-  scale_fill_binned(type = "viridis", na.value = "grey60",
-                    breaks = c(min(SAvail_im_DA2011$V_im),
-                               quantile(SAvail_im_DA2011$V_im,0.25)%>%as.numeric(),
-                               median(SAvail_im_DA2011$V_im, na.rm=T),
-                               quantile(SAvail_im_DA2011$V_im,0.75)%>%as.numeric(),
-                               max(SAvail_im_DA2011$V_im, na.rm = T)))+
-  geom_sf(data = ALL_SCHOOLS_Elem |> filter(Year != 2016), 
-          size=0.5, colour = "Red", alpha=0.7)+
-  geom_sf(data = ham_community_bounds, fill=NA, color = "grey30", alpha=0.7)+
-  theme_minimal() +
-  theme(legend.title = element_text(size=10),
-        axis.text = element_blank(), #removes lat/long from the axises
-        panel.grid.major=element_blank())+
-  facet_wrap(~mode)
-
-SA_Vim_2016_plot <- ggplot() +
-  geom_sf(data = SAvail_im_DA2016, aes(fill= V_im), color = NA) +
-  scale_fill_binned(type = "viridis", na.value = "grey60",
-                    breaks = c(min(SAvail_im_DA2016$V_im),
-                               quantile(SAvail_im_DA2016$V_im,0.25)%>%as.numeric(),
-                               median(SAvail_im_DA2016$V_im, na.rm=T),
-                               quantile(SAvail_im_DA2016$V_im,0.75)%>%as.numeric(),
-                               max(SAvail_im_DA2016$V_im, na.rm = T)))+
-  geom_sf(data = ALL_SCHOOLS_Elem |> filter(Year != 2011), 
-          size=0.5, colour = "Red", alpha=0.7)+
-  geom_sf(data = ham_community_bounds, fill=NA, color = "grey30", alpha=0.7)+
-  theme_minimal() +
-  theme(legend.title = element_text(size=10),
-        axis.text = element_blank(), #removes lat/long from the axises
-        panel.grid.major=element_blank())+
-  facet_wrap(~mode)
-```
-```{r, eval=FALSE}
-SA_vim_2011_plot <- ggplot() +
-  geom_sf(data = SAvail_im_DA2011, aes(fill= v_im_kid_popadj), color = NA) +
-  scale_fill_binned(type = "viridis",
-                    breaks = c(min(SAvail_im_DA2011$v_im_kid_popadj, na.rm = T),
-                               quantile(SAvail_im_DA2011$v_im_kid_popadj,0.25)%>%as.numeric(),
-                               median(SAvail_im_DA2011$v_im_kid_popadj, na.rm=T),
-                               quantile(SAvail_im_DA2011$v_im_kid_popadj,0.75)%>%as.numeric(),
-                               max(SAvail_im_DA2011$v_im_kid_popadj, na.rm = T)))+
-  geom_sf(data = ALL_SCHOOLS_Elem |> filter(Year != 2016), 
-          size=0.5, colour = "Red", alpha=0.7)+
-  geom_sf(data = ham_community_bounds, fill=NA, color = "grey30", alpha=0.7)+
-  theme_minimal() +
-  theme(legend.title = element_text(size=10),
-        axis.text = element_blank(), #removes lat/long from the axises
-        panel.grid.major=element_blank())+
-  facet_wrap(~mode)
-
-SA_vim_2016_plot <- ggplot() +
-  geom_sf(data = SAvail_im_DA2016, aes(fill= v_im_kid_popadj), color = NA) +
-  scale_fill_binned(type = "viridis",
-                    breaks = c(min(SAvail_im_DA2016$v_im_kid_popadj, na.rm = T),
-                               quantile(SAvail_im_DA2016$v_im_kid_popadj,0.25)%>%as.numeric(),
-                               median(SAvail_im_DA2016$v_im_kid_popadj, na.rm=T),
-                               quantile(SAvail_im_DA2016$v_im_kid_popadj,0.75)%>%as.numeric(),
-                               max(SAvail_im_DA2016$v_im_kid_popadj, na.rm = T)))+
-  geom_sf(data = ALL_SCHOOLS_Elem |> filter(Year != 2011), 
-          size=0.5, colour = "Red", alpha=0.7)+
-  geom_sf(data = ham_community_bounds, fill=NA, color = "grey30", alpha=0.7)+
-  theme_minimal() +
-  theme(legend.title = element_text(size=10),
-        axis.text = element_blank(), #removes lat/long from the axises
-        panel.grid.major=element_blank())+
-  facet_wrap(~mode)
-```
-
-```{r, eval=FALSE}
-SA_vim_LIMAT_2011_plot <- ggplot() +
-  geom_sf(data = SAvail_im_DA2011, aes(fill= v_im_kid_popadj_LIMAT), color = NA) +
-  scale_fill_binned(type = "viridis", 
-                    breaks = c(min(SAvail_im_DA2011$v_im_kid_popadj_LIMAT, na.rm = T),
-                               quantile(SAvail_im_DA2011$v_im_kid_popadj_LIMAT,0.25)%>%as.numeric()+0.000000001,
-                               median(SAvail_im_DA2011$v_im_kid_popadj_LIMAT, na.rm=T),
-                               quantile(SAvail_im_DA2011$v_im_kid_popadj_LIMAT,0.75)%>%as.numeric(),
-                               max(SAvail_im_DA2011$v_im_kid_popadj_LIMAT, na.rm = T)))+
-  geom_sf(data = ALL_SCHOOLS_Elem |> filter(Year != 2016), 
-          size=0.5, colour = "Red", alpha=0.7)+
-  geom_sf(data = ham_community_bounds, fill=NA, color = "grey30", alpha=0.7)+
-  theme_minimal() +
-  theme(legend.title = element_text(size=10),
-        axis.text = element_blank(), #removes lat/long from the axises
-        panel.grid.major=element_blank())+
-  facet_wrap(~mode)
-
-SA_vim_LIMAT_2016_plot <- ggplot() +
-  geom_sf(data = SAvail_im_DA2016, aes(fill= v_im_kid_popadj_LIMAT), color = NA) +
-  scale_fill_binned(type = "viridis", 
-                    breaks = c(min(SAvail_im_DA2016$v_im_kid_popadj_LIMAT, na.rm = T),
-                               quantile(SAvail_im_DA2016$v_im_kid_popadj_LIMAT,0.25)%>%as.numeric()+0.000000001,
-                               median(SAvail_im_DA2016$v_im_kid_popadj_LIMAT, na.rm=T),
-                               quantile(SAvail_im_DA2016$v_im_kid_popadj_LIMAT,0.75)%>%as.numeric(),
-                               max(SAvail_im_DA2016$v_im_kid_popadj_LIMAT, na.rm = T)))+
-  geom_sf(data = ALL_SCHOOLS_Elem |> filter(Year != 2011), 
-          size=0.5, colour = "Red", alpha=0.7)+
-  geom_sf(data = ham_community_bounds, fill=NA, color = "grey30", alpha=0.7)+
-  theme_minimal() +
-  theme(legend.title = element_text(size=10),
-        axis.text = element_blank(), #removes lat/long from the axises
-        panel.grid.major=element_blank())+
-  facet_wrap(~mode)
-
-SA_Vim_LIMAT_2011_plot <- ggplot() +
-  geom_sf(data = SAvail_im_DA2011, aes(fill= V_i_kid_popadj_LIMAT), color = NA) +
-  scale_fill_binned(type = "viridis", 
-                    breaks = c(min(SAvail_im_DA2011$V_i_kid_popadj_LIMAT, na.rm = T),
-                               quantile(SAvail_im_DA2011$V_i_kid_popadj_LIMAT,0.25)%>%as.numeric()+0.000000001,
-                               median(SAvail_im_DA2011$V_i_kid_popadj_LIMAT, na.rm=T),
-                               quantile(SAvail_im_DA2011$V_i_kid_popadj_LIMAT,0.75)%>%as.numeric(),
-                               max(SAvail_im_DA2011$V_i_kid_popadj_LIMAT, na.rm = T)))+
-  geom_sf(data = ALL_SCHOOLS_Elem |> filter(Year != 2016), 
-          size=0.5, colour = "Red", alpha=0.7)+
-  geom_sf(data = ham_community_bounds, fill=NA, color = "grey30", alpha=0.7)+
-  theme_minimal() +
-  theme(legend.title = element_text(size=10),
-        axis.text = element_blank(), #removes lat/long from the axises
-        panel.grid.major=element_blank())+
-  facet_wrap(~mode)
-
-SA_Vim_LIMAT_2016_plot <- ggplot() +
-  geom_sf(data = SAvail_im_DA2016, aes(fill= V_i_kid_popadj_LIMAT), color = NA) +
-  scale_fill_binned(type = "viridis", 
-                    breaks = c(min(SAvail_im_DA2016$V_i_kid_popadj_LIMAT, na.rm = T),
-                               quantile(SAvail_im_DA2016$V_i_kid_popadj_LIMAT,0.25)%>%as.numeric()+0.000000001,
-                               median(SAvail_im_DA2016$V_i_kid_popadj_LIMAT, na.rm=T),
-                               quantile(SAvail_im_DA2016$V_i_kid_popadj_LIMAT,0.75)%>%as.numeric(),
-                               max(SAvail_im_DA2016$V_i_kid_popadj_LIMAT, na.rm = T)))+
-  geom_sf(data = ALL_SCHOOLS_Elem |> filter(Year != 2011), 
-          size=0.5, colour = "Red", alpha=0.7)+
-  geom_sf(data = ham_community_bounds, fill=NA, color = "grey30", alpha=0.7)+
-  theme_minimal() +
-  theme(legend.title = element_text(size=10),
-        axis.text = element_blank(), #removes lat/long from the axises
-        panel.grid.major=element_blank())+
-  facet_wrap(~mode)
-```
-```{r, eval=FALSE}
-SA_Vim_2011_plot
-SA_vim_2011_plot
-SA_Vim_LIMAT_2011_plot
-SA_vim_LIMAT_2011_plot
-
-SA_Vim_2016_plot
-SA_vim_2016_plot
-SA_Vim_LIMAT_2016_plot
-SA_vim_LIMAT_2016_plot
-#*summarize LIM at city level
-```
-
-```{r, eval=FALSE}
+#some community summary tables?
 SAvail_im_DA2011  |> st_drop_geometry() |> group_by(COMMUNITY, mode) |> summarise(
@@
 
-From this perspective, school closure policies have been pursued by governments to optimize per-student operational cost-efficiency [@rongReviewResearchLowcarbon2022] but other implications, such as the impact of closures on property values [@MerrallWhatsASchool2024], the effects on the social infrastructure of communities [@butlerClosureRideau2019a; @irwinSchoolClosure2012], and particularly the travel implications of school closures have often been overlooked by decision makers [@bierbaumMobilityJusticeLinking2021; @Lee_Lubienski_2017].
+From this perspective, school closure policies have been pursued by governments to optimize per-student operational cost-efficiency [@rongReviewResearchLowcarbon2022; @Dai_Wang_Zhang_Liao_Liu_2019] but other implications, such as the impact of closures on property values [@MerrallWhatsASchool2024], the effects on the social infrastructure of communities [@butlerClosureRideau2019a; @irwinSchoolClosure2012], and particularly the travel implications of school closures have often been overlooked by decision makers [@bierbaumMobilityJusticeLinking2021; @Lee_Lubienski_2017].
 This seems a mistake: journey-to-school trips are tremendously important, and travel for this purpose is the largest contributor of carbon emissions from personal travel after the journey-to-work [@rongReviewResearchLowcarbon2022; @pantelakiImpactHomeschoolCommuting2024].
@@
 Quantification of losses in potential active travel and associated growith in carbon emissions is scarce in the literature, and the focus of this paper, which aims to make both methodological and empirical contributions.
-We take a novel multi-modal and singly-constrained accessibility measure, called spatial availability [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024], and use it to estimate the changes in accessibility and travel before and after a recent wave of school closure/consolidations implemented in Hamilton, Ontario.
+We take a novel multimodal and singly-constrained accessibility measure, called spatial availability [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024], and use it to estimate the changes in accessibility and travel before and after a recent wave of school closure/consolidations implemented in Hamilton, Ontario.
 
@@
 
-Our focus is to evaluate the impact of school consolidations between 2011 and 2016 on multi-modal school spatial availability and travel-related GHG emission in Hamilton.
+Our focus is to evaluate the impact of school consolidations between 2011 and 2016 on multimodal school spatial availability and travel-related GHG emission in Hamilton.
 To facilitate comparison, two scenarios are generated.
@@
 typically children aged 5 to 14).
-As such, elementary includes middle schools that only instruct grades 6 to 8 (*Mid*), primary schools that only instruct kindergarten to grade 5 or 6 (*JrElem*), and all grade elementary schools (*Elem*) which instruct all grades from kindergarten to grade 8.
-The maximum number of students that may be enrolled in a school is calculated by the provincial Ministry of Education; this value is referred to as the on-the-ground capacity (*OTGC*) and is assigned to each school <!-- [AP]: what are the units of OTGC?
-Students?
--->.
+As such, elementary includes middle schools that only instruct grades 6 to 8 (*MidElem*), primary schools that only instruct kindergarten to grade 5 or 6 (*JrElem*), and all grade elementary schools (*Elem*) which instruct all grades from kindergarten to grade 8.
+The maximum number of students that may be enrolled in a school is calculated by the provincial Ministry of Education; this value is referred to as the on-the-ground capacity (*OTGC*), the number of school seats available in the school beyond which it will be over-capacity, and is assigned to each school.
 The current/historic OTGCs are not publicly available [@ontarioSchoolFacilityInventory2017].
@@
 A regression model was used to estimate the OTGCs for schools that lacked capacity information.
-The independent variables in the model were 1) the school's building footprint (*F*) in units of ($m^2$); 2) the school's shortest Euclidean distance from the centroid of Hamilton's central business district area (*DistCBD*) defined in units of (m); and 3) coefficients associated with a binary variable indicating the type of school grade instruction (*Mid*, *JrElem*, *Elem*).
+The independent variables in the model were 1) the school's building footprint (*F*) in units of ($m^2$); 2) the school's shortest Euclidean distance from the centroid of Hamilton's central business district area (*DistCBD*) defined in units of (m); and 3) coefficients associated with a binary variable indicating the type of school grade instruction (*MidElem*, *JrElem*, *Elem*).
+Of note, the models were also used to estimate OTGC of secondary schools (*Sec*) though not included in the scope of this work.
+These public institutions provide instruction for grades 9 to 12 (i.e., children aged 14 to 17).
 The building footprint from an archived spatial data set [@DMTI_Spatial_2015] and the 2016 footprints were retrieved from OSM [@OpenStreetMap_2021].
@@
 School status and OTGC summed by catchment is also visualised in @fig-Fig1.
-As seen in Table <!--\ref{TabA1-OTGC}--> in the Appendix, the coefficient of determination for these models was $R^2 = $ and $R^2 = $ for the Public and Catholic Boards respectively, and the standard deviation was $\sigma = $ and $sigma = $.
-
-<!-- [AS]: shall we explain further how we valuated this model?
-I don't think its entirely necessary.
-But the code is here: https://github.com/soukhova/hamONdests/blob/master/data-raw/01-school-data-preparation.Rmd (Antonio is a collaborator) -->
-
-<!-- [AP]: Just summarize the results here: what was the R2 and standard deviations of these models? -->
+As seen in Table \ref{TabA1-OTGC} in the Appendix, the coefficient of determination for these models is $R^2 = 0.999$ and $R^2 = 0.998$ for the Public $OTGC_{Public}$ and Catholic $OTGC_{Catholic}$ Boards respectively, and the residual standard deviation is $\sigma = 0.212$ and $sigma = 0.331$.
+<!-- [AS]: model estimation here: https://github.com/soukhova/hamONdests/blob/master/data-raw/01-school-data-preparation.Rmd -->
 
@@
 \label{eq:OTGC-public}
-OTGC_{Public} = F^{0.346}-e^{0.00003*DistCBD}+e^{3.123*JrElem_b}+e^{3.752*Elem_b}+ e^{3.068*Mid_b}
+OTGC_{Public} = F^{0.346}-e^{0.00003*DistCBD}+e^{3.123*JrElem}+e^{3.752*Elem}+ e^{3.068*MidElem} 
 \end{equation}
@@
 \label{eq:OTGC-catholic}
-OTGC_{Catholic} =F^{0.471}-e^{0.00003*DistCBD}+e^{2.333*Elem_b}
+OTGC_{Catholic} =F^{0.471}-e^{0.00003*DistCBD}+e^{2.333*Elem}
 \end{equation}
@@
 #| echo: false
-#| out-width: "140%"
+#| out-width: "120%"
 
@@
 Hamilton (Ancaster `r community_car_perc_2011$TAZOrig_car_perc[1] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[1] |> percent()`, Dundas `r community_car_perc_2011$TAZOrig_car_perc[2] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[2] |> percent()`, Flamborough `r community_car_perc_2011$TAZOrig_car_perc[3] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[3] |> percent()`, Glanbrook `r community_car_perc_2011$TAZOrig_car_perc[4] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[4] |> percent()`, Hamilton Central `r community_car_perc_2011$TAZOrig_car_perc[5] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[5] |> percent()`, and Stoney Creek `r community_car_perc_2011$TAZOrig_car_perc[6] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[6] |> percent()`).
-<!--Do we maybe need to discuss how this lack of data for each TAZ impacted the calculations?
-[AP]: No.
-Unrelated: did you use information from TTS2016R?
+<!--Unrelated: did you use information from TTS2016R?
 Or can you add it to the data package?
@@
 The empirical TLD is used to fit a density distribution using maximum likelihood and moment matching techniques and the Nelder-Mead and Brent methods for direct optimization available within the {fitdistrplus} `R` package [@fitdistrplus_2015].
-Based on goodness-of-fit criteria and diagnostics <!-- should I include the diagnostic plots in the appendix?
-q-q, etc..?
-I don't think its entirely necessary as the empirical is graphed over the theoretical line in the following figure.
-[AP] Agreed-->, the gamma and exponential distributions were selected for the motorized and non-motorized modal distributions respectively.
+Based on goodness-of-fit criteria and diagnostics, the gamma and exponential distributions were selected for the motorized and non-motorized modal distributions respectively.
 The gamma distribution is defined by the shape ($\alpha$) parameter of `r round(gamma2011_motor$estimate[1], 3)` (2011) and `r round(gamma2016_motor$estimate[1], 3)` (2016) and the rate ($\beta$) of `r round(gamma2011_motor$estimate[2], 3)` (2011) and `r round(gamma2016_motor$estimate[2], 3)` (2016).
@@
 #| echo: false
-#| out-width: "140%"
+#| out-width: "100%"
 
@@
 
-# Methods: Multi-modal spatial availability
+# Methods: Multimodal spatial availability
 
 Accessibility is defined as the "potential for spatial interaction".
-It is classically presented as the gravity-based measure popularized by @hansenHowAccessibilityShapes1959 and takes the following general multi-modal formulation (Equation (\ref{eq:multi-modal-conventional-accessibility})):
+It is classically presented as the gravity-based measure popularized by @hansenHowAccessibilityShapes1959 and takes the following general multimodal formulation (Equation (\ref{eq:multimodal-conventional-accessibility})):
 
@@
 \begin{equation}
-\label{eq:multi-modal-conventional-accessibility}
+\label{eq:multimodal-conventional-accessibility}
 S_i^m = \sum_{j=1}^JO_j \cdot f^m(c_{ij}^m)
@@
 
-In this paper, we benefit from new developments in accessibility research, particularly the multi-modal spatial availability measure [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024].
+In this paper, we benefit from new developments in accessibility research, particularly the multimodal spatial availability measure [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024].
 Spatial availability is a singly-constrained accessibility measure that accounts for competition by students using different modes for exclusive opportunities, such as school seats.
@@
 This proportional allocation of opportunities yields an interpretable and meaningful measure of opportunity access, particularly when comparing across modes, at the spatial resolution of a residential parcel, and multiple time periods.
-See @soukhovMultimodalSpatialAvailability2024 for further discussion, multi-modal spatial availability $V_{i}^m$ is defined as given by Equation (\ref{eq:spatial-availability}).
+See @soukhovMultimodalSpatialAvailability2024 for further discussion, multimodal spatial availability $V_{i}^m$ is defined as given by Equation (\ref{eq:spatial-availability}).
 
@@
 Also, as discussed in @soukhovIntroducingSpatialAvailability2023, popular competitive accessibility measures such as the 2SFCA [@josephMeasuringPotentialPhysical1982; @weibullAxiomaticApproachMeasurement1976; @shenLocationCharacteristicsInnercity1998; @luoMeasuresSpatialAccessibility2003] are internally inconsistent, and the only way it preserves the number of opportunities is if the effect of the impedance function is ignored when expanding the values of opportunities per capita to obtain the total number of opportunities.
-On the other hand, the proportional allocation procedure associated with calculating multi-modal spatial availability $V_i^m$ consistently returns a number of opportunities available to populations by mode that matches the total number of opportunities in the region when summed.
-By doing this consistently, it is possible to define a measure of multi-modal spatial availability per capita as presented in Equation (\ref{eq:SA-per-capita}) for use as a benchmark to compare against the regional opportunities per capita ($\frac{\sum_{j} O_j}{\sum_{i} P_i}$).
+On the other hand, the proportional allocation procedure associated with calculating multimodal spatial availability $V_i^m$ consistently returns a number of opportunities available to populations by mode that matches the total number of opportunities in the region when summed.
+By doing this consistently, it is possible to define a measure of multimodal spatial availability per capita as presented in Equation (\ref{eq:SA-per-capita}) for use as a benchmark to compare against the regional opportunities per capita ($\frac{\sum_{j} O_j}{\sum_{i} P_i}$).
 
@@
 
-This paper uses the multi-modal spatial availability $V_i^m$ and spatial availability per student $v_i^m$ measures to quantify accessibility changes for the following reasons.
+This paper uses the multimodal spatial availability $V_i^m$ and spatial availability per student $v_i^m$ measures to quantify accessibility changes for the following reasons.
 
@@
 $F_{i}^{pm}$ joined with $F_{ij}^{cm}$ yields the total balancing factor $F^{tm}_{ij}$ which serves to proportionally allocate the school capacity $O_j$ to each parcel.
-The value is multi-modal spatial availability $V_i^m$ which can be interpreted as the number of school-seats that are available to the mode-using population at that parcel.
+The value is multimodal spatial availability $V_i^m$ which can be interpreted as the number of school-seats that are available to the mode-using population at that parcel.
 $V_i^m$ is then summed and represented at the DA and community level along with the calculated $v_i^m$ for interpretation.
@@
 #| label: fig-Fig5
-#| fig-cap: "Multi-modal spatial availability (purples) and spatial availability per mode-using student (diverging reds and greens) aggregated at the DA level for 2011 and 2016.
+#| fig-cap: "Multimodal spatial availability (purples) and spatial availability per mode-using student (diverging reds and greens) aggregated at the DA level for 2011 and 2016.
 Scales are represented in quartiles."
@@
 ```
-
-The motorized and non-motorized spatial availabilities of school seats $V_i$ for both 2011 and 2016 are presented in the top half of @fig-Fig5.
-It is worth noting that the sum of $V_i^m$ values in a year for both motorized and non-motorized population equals the city-wide OTGC (across both public and catholic school boards) in that year.
-This is a result of $V_i^m$ proportional allocation mechanism, singly-constraining opportunities through the proportion allocation balancing factors.
-In this way, $V_i^m$ results can be interpreted as a proportion of the total OTGC in that year.
-Though the top four plots of @fig-Fig5 demonstrate different magnitudes, both populations using non- and motorized modes have higher spatial availability when they are more proximate to schools.
+<!--check to make sure schools are correct-->
+
+The motorized and non-motorized spatial availabilities of school seats $V_i^m$ for both 2011 and 2016 are presented in the top half of @fig-Fig5.
+The sum of $V_i^m$ values in a year for both motorized and non-motorized population equals the city-wide OTGC (across both public and catholic school boards) in that year.
+This is a result of $V_i^m$'s proportional allocation mechanism, singly-constraining opportunities through the proportion allocation balancing factors.
+In this way, $V_i^m$ results can be interpreted as a proportion of the total OTGC in that year as the sum of $V_i^m$ for both modes in a year is equal to the total OTGC in that year.
+In @fig-Fig5, $i$ is a DA, so $V_i^m$ is the sum of the school seat availability per parcel in each DA.
+Each parcel is assigned a potential student population based on the average rate of school-aged population per residential parcel in that DA and the travel impedance weighted travel time to all schools.
+The top four plots of @fig-Fig5 demonstrate different magnitudes in part due to significantly lower non-motorized modal share (refer to @fig-Fig3), however it is apparent that both mode using populations have higher spatial availability when they are more proximate to schools.
 Hence, all DAs have higher values within Hamilton Central and those more proximate to schools in less-urban communities.
-Trends appear similar for both 2011 and 2016: proximate populations using motorized modes benefit from the ability to reach more distant destinations than non-motorized modes.
+Trends appear similar for both 2011 and 2016: this aligns with intuition, proximate populations using motorized modes benefit from the ability to reach more distant destinations than non-motorized modes.
 
@@
 The $v_i^m$ plots demonstrate different spatial distribution trends than $V_i^m$ plots.
-If our aim is to provide sufficient availability to schools for both motorized and non-motorized populations (i.e., greater than 1.0 $v_i^m$ school-seat availability per student), this rate is only achieved in the core of the city for motorized modes (greens and yellows).
+If our aim is to provide sufficient availability to schools for both motorized and non-motorized populations (e.g., greater than 1.0 $v_i^m$ school-seat availability per student), this rate is only achieved in the core of the city for motorized modes (greens and yellows).
 Notably, this rate only is available to non-motorized populations in DAs that are in less-densely populated rural areas that are school proximate and very few pockets of Hamilton Central.
@@
 #| label: fig-Fig6
-#| fig-cap: "Change in multi-modal spatial availability and spatial availability per mode-using student between 2016 and 2011 at the 2016 DA Canadian Census aggregation.
-Scales are represented in quartiles."
+#| fig-cap: "Change in multimodal spatial availability and spatial availability per mode-using student between 2016 and 2011.
+Change is the subtraction of parcel-level results between 2016 and 2011 scenarios summed for each DA in the 2016 Canadian Census DA system.
+Scales are represented in quartiles, with Q4 rounded to 0."
 #| echo: false
-
+#| out-width: "120%"
+#| 
 knitr::include_graphics(paste0(here::here(),"/Figures/Vim_vvim_20112016_CHANGE_plot.png"))
 ```
-<!-- TO DO: Fig6 will be 8 plots (1 for each mode) of 1) change in V_im, v_im, active transportation minutes, GHG emissions. The geometry will be in 2016 DAs. -->
-
-To emphasize how spatial availability changed between 2016 and 2011, the change in $V_i^m$ and $v_i^m$ is visualised for both modes in @fig-Fig6.
-Overall, a decrease in both $V_i^m$ and $v_i^m$ can be seen across both modes around closed schools (black points).
-An increase near new or expanded schools (green points) is not as notable as overall fewer school seats were added relative to increases in the school-aged population, leading to increased competition for these resources.
-For non-motorized mode-using populations, $V_i^m$ decreases across the city.
-However, some motorized-mode using populations in DAs do experience an increase in $V_i^m$: these DAs are often relatively proximate to schools but more immediately outside walking-range.
+
+To highlight this point, @fig-Fig6 visualises the change in school spatial availability between 2016 and 2011 for both mode where $i$ are 2016 DAs.
+Though the 2011 and 2016 DAs system are very similar (see @fig-Fig2 and @fig-Fig3), they are not identical: `r (cen_data_2011_hamiltononly$GeoUID %in% cen_data_2016_hamiltononly$GeoUID |> length() - cen_data_2011_hamiltononly$GeoUID %in% cen_data_2016_hamiltononly$GeoUID |> sum()) + (cen_data_2016_hamiltononly$GeoUID %in% cen_data_2011_hamiltononly$GeoUID |> length() - cen_data_2016_hamiltononly$GeoUID %in% cen_data_2011_hamiltononly$GeoUID |> sum())` DAs changed in their spatial extent (i.e., added, removed, changed shape).
+The proportional allocation mechanism of spatial availability provides flexibility in aggregating parcel-level values at whichever spatial unit is meaningful.
+For this case, the change in $V_i^m$ and $v_i^m$ represented is the subtraction between the value aggregated at 2016 DA for the 2016 and 2011 scenario.
+
+From @fig-Fig6 (top plots), it is apparent that decreases (pinks and reds) in $V_i^m$ appear across the city for both mode-using populations.
+As the majority of closed schools are within the core of the city: the more urban Hamilton Central and urban-areas of other communities see a concentration of this decreased spatial availability to school seats.
+Evidently, the number of OTGC in the city (equal to the sum of $V_i^m$) decreased between the two years, so some sort of decrease is to be expected.
+However, even when spatial availability is divided by the mode-using population (which also decreased between the years) $v_i^m$ (bottom plots) continues to demonstrate decreases, especially those locally proximate to closed schools.
+
+This decrease follows different patterns depending on the mode.
+$v_i^{non-motorized}$ is most impacted for parcels within a 27 minute walk travel time to schools that closed: these DAs demonstrate the most dramatic decrease.
+Smaller-area DAs (in Hamilton-Central) and larger-area DAs (in rural communities) proximate to closed schools present Q1 decreases (red).
+For $v_i^{motorized}$, the range offered by motorized mode (estimated by shortest-path car travel times) is city-wide, so since the majority of schools were centrally located, the decrease is concentrated within the downtown core.
+Some motorized-mode using populations in DAs do experience an increase in $V_i^m$ and $v_i^m$: these DAs are often relatively proximate to schools but more immediately outside walking range.
 Under competitive and constrained calculations, these DAs benefit from the losses in spatial availability that are predominately experienced by the non-motorized populations within Hamilton Central.
 
-<!-- TO DO: add tables to show availability by mode in 2011 and 2016 with changes, as well as lost minutes of active travel time, emissions, and LIM-AT.
-Tabulate by communtity and city-wide.
-[AP]: Check Elise's equity in accessibility to bike share paper, see table 1 there.
--->
-
 ```{r}
-# SAvail_im_COMM_changes |> filter(mode == "motorized") |> st_drop_geometry() |> 
-#   dplyr::select(c("COMMUNITY",
-#                   "prev_LIMAT_2016", 
-#                   "c_kid_perc", 
-#                   "c_V_im_perc", 
-#                   "c_v_im_perc", 
-#                   "emissions_2011",
-#                   "emissions_2016"
-#                   #"c_emissions_perc",
-#                   # "emissions_perkid_2011",
-#                   # "emissions_perkid_2016",
-#                   #"c_emissions_perkid_perc" 
-#                   )) |> arrange(desc(prev_LIMAT_2016)) 
-```
+#| label: fig-Fig7
+#| fig-cap: "Change in active travel time minutes (top plot) and home-to-school GHG emissions (bottom plot) between 2011 and 2016 scenarios.
+Change is the subtraction of parcel-level results between 2016 and 2011 scenarios summed for each DA in the 2016 Canadian Census DA system.
+Scales are represented in quartiles, with Q4 rounded to 0."
+#| echo: false
+#| out-width: "120%"
+#| 
+knitr::include_graphics(paste0(here::here(),"/Figures/ttim_GHGim_20112016_CHANGE_plot.png"))
+```
+
+The decrease in school seat spatial availability can be further interpreted.
+How does potential non-motorized travel to school decrease?
+How does the increase in potential motorized travel minutes impact GHG emissions?
+@fig-Fig7 visualises these related intermediates for one-way home-to-school travel in a given day.
+The top plot of @fig-Fig7 demonstrates the sum of the change in non-motorized minutes per 2016 DA.
+These travel time minutes are the difference between the aggregated 2016 and 2011 scenarios for non-motorized travel times multiplied by the proportion of walking/cycling mode-use and rate of student population associated with the parcels in each 2016 DA.
+It can be seen that Q1 decreases are proximate to closed schools.
+Smaller decreases (and even a few increases) are apparent in a range further from closed schools, especially those also in proximity to the few opened/expanded schools.
+Intuitively, these trends mirror what is shown within the $V^{non-motorized}$ landscape in @fig-Fig6 (Pearson correlation coefficient of `r cor(SAvail_im_DA_changes |> filter(mode == "non-motorized") |> pull(c_total_tt), SAvail_im_DA_changes |> filter(mode == "non-motorized") |> pull(c_V_im)) |> round(digits=2)`) as travel time is an input into the cost balancing factor $F^{cm}_{ij}$.
+Similarly, the bottom plot (@fig-Fig7) visualises the change in estimated emissions from motorized travel minutes.
+The change in GHG emissions is also highly correlated with $V^{motorized}$ in @fig-Fig6 (Pearson correlation coefficient of `r cor(SAvail_im_DA_changes |> filter(mode == "motorized") |> pull(c_emissions_total_car_tt), SAvail_im_DA_changes |> filter(mode == "motorized") |> pull(c_V_im))`) as it is also a product of travel time.
+
+Notably, the mode type of spatial availability loss is important.
+The change in active travel minutes and GHG emissions highlights the negative externalities of _decreased_ non-motorized spatial availability.
+Normatively, our cities should be increasing spatial availability for non-motorized mode-users given the climate crisis.
+The wave of school closures in Hamilton increased motorized spatial availability and hence potential GHG emissions.
+The policy drastically reduced non-motorized spatial availability, and this is seen in the decrease in potential non-motorized travel minutes as well as increased GHG emissions in rural areas that did benefit from motorized spatial availability gains.
+The proportion of non-motorized mode use is low in both 2011 and 2016, the closure of schools that are more proximate to student-population density eliminates the potential of those trips ever becoming active.
+And who does the reduction of non-motorized spatial availability impact the most?
+A summary of spatial availability per Hamilton community and associated dimensions in 2016 along with the percentage change between 2011 and 2016 are summarized in @tbl-Tab1.
 
 ```{r include=FALSE}
-SAvail_im_COMM_changes |> st_drop_geometry() |> 
-  dplyr::select(c("COMMUNITY",
-                  "prev_LIMAT_2016", #prevalence of LIM after-tax for households with children under 18
+SAvail_im_COMM_changes_df <- SAvail_im_COMM_changes |> st_drop_geometry()
+
+LIMAT_COMM <- SAvail_im_DA |> st_drop_geometry() |> filter(mode == "motorized" & year == "2016") |>
+  group_by(COMMUNITY) |> 
+  summarise(COMM_prev_LIMAT_under18_2016 = mean(DA_prev_LIMAT_under18, na.rm=T))
+
+LIMAT_COMM_2011 <- SAvail_im_DA |> st_drop_geometry() |> 
+  filter(mode == "motorized" & year == "2011") |>
+  group_by(COMMUNITY) |> 
+  summarise(COMM_prev_LIMAT_under18_2011 = mean(DA_prev_LIMAT_under18, na.rm=T))
+
+LIMAT_COMM <- left_join(LIMAT_COMM, LIMAT_COMM_2011, by = "COMMUNITY") |>  
+  mutate(c_COMM_prev_LIMAT_under18 = ((COMM_prev_LIMAT_under18_2016 - COMM_prev_LIMAT_under18_2011)/COMM_prev_LIMAT_under18_2011) |>
+           scales::percent(accuracy = 2L)) 
+rm(LIMAT_COMM_2011)
+
+SCHOOL_SEATS_COMM <- ALL_SCHOOLS_Elem |> st_drop_geometry() |> group_by(COMMUNITY) |> 
+  summarise(OTGC2011 = sum(OTGC2011, na.rm=T),
+            OTGC2016 = sum(OTGC2016, na.rm=T)) |>
+  mutate(c_OTGC_perc = ((OTGC2016 - OTGC2011)/OTGC2011) |> scales::percent(accuracy = 2L))
+
+SAvail_im_COMM_changes_df <- SAvail_im_COMM_changes_df |> left_join(LIMAT_COMM, by="COMMUNITY") 
+SAvail_im_COMM_changes_df <- SAvail_im_COMM_changes_df |> left_join(SCHOOL_SEATS_COMM, by="COMMUNITY") |> mutate_if(is.numeric, round, 1) 
+```
+
+```{r include=FALSE}
+SAvail_im_COMM_changes_df <- SAvail_im_COMM_changes_df |> 
+  dplyr::select(c("COMMUNITY",
                   "mode",
-                  "kid_proportion_2016", #the proportion of children 5-14 age, sums to 100%
-                  "c_kid_proportion_perc", #the % change in proportion between 2016-2011 i.e. (2016-2011)/2011, negative means decrease
-                  "V_im_proportion_2016",  #the proportion of V_im, sums to 100%
-                  "c_V_im_proportion_perc", #the % change in proportion between 2016-2011 i.e. (2016-2011)/2011, negative means decrease
-                  "v_im_2016", #the proportion of V_im (in community) /kid_population (in community) = v_im, sums to 100%
-                  "c_v_im_perc", #the % change in proportion between 2016-2011 i.e. (2016-2011)/2011, negative means decrease
-                  "total_tt_2016", #summed travel time (kid_pop at each household * travel time of mode, summed for the community). Values sums to 100%
-                  "c_total_tt_perc",  #the % change in proportion between 2016-2011 i.e. (2016-2011)/2011, negative means decrease
-                  "emissions_perkid_2016", #summed motorized travel time * 0.1 g/min (kid_pop at each household * travel time of mode, summed for the community). Values sums to 100%
-                  "c_emissions_perkid_perc"  #the % change in proportion between 2016-2011 i.e. (2016-2011)/2011, negative means decrease
-  ))
-```
-
-Spatial availability can also be summarized by community.
-The majority of schools were closed in Hamilton Center, this community represents `r SAvail_im_COMM_changes |> st_drop_geometry() |> filter(COMMUNITY == "HA") |> dplyr::select("kid_proportion_2016") |> sum() |> scales::percent()` of the 5-14 year old population in 2016.
-Though the population between 2011 and 2016 decreased, the rate of school-seat spatial availability decreased disproportionately more (population by `r SAvail_im_COMM_changes |> st_drop_geometry() |> filter(COMMUNITY == "HA") |> transmute(calc = ((sum(kid_2016)-sum(kid_2011))/sum(kid_2011)) |> scales::percent()) |> first()` and spatial availability by `r `r SAvail_im_COMM_changes |> st_drop_geometry() |> filter(COMMUNITY == "HA") |> transmute(calc = ((sum(V_im_2016)-sum(V_im_2011))/sum(V_im_2011)) |> scales::percent())  |> first()`).
+                  "COMM_prev_LIMAT_under18_2016",
+                  "c_COMM_prev_LIMAT_under18",
+                  "kid_2016",
+                  "c_kid_perc",
+                  "OTGC2016",
+                  "c_OTGC_perc",
+
+                  "V_im_2016",
+                  "c_V_im_perc",
+                  "v_im_2016",
+                  "c_v_im_perc",
+
+                  "emissions_2016",
+                  "c_emissions_perc",
+                  "emissions_perkid_2016",
+                  "c_emissions_perkid_perc",
+
+                  "total_tt_2016",
+                  "c_total_tt_perc",
+                  "total_tt_perkid_2016",
+                  "c_total_ttperkid_perc")) |> arrange(desc(COMM_prev_LIMAT_under18_2016))
+```
+
+```{r include=FALSE}
+tdata_1 <- SAvail_im_COMM_changes_df |> 
+  filter(mode == "motorized")|> 
+  dplyr::select(c("COMMUNITY",
+                  "COMM_prev_LIMAT_under18_2016",
+                  "kid_2016",
+                  "OTGC2016"))
+
+tdata_1 <- insertRow(tdata_1 |> as.matrix(),2,
+          SAvail_im_COMM_changes_df |> 
+            filter(mode == "motorized" & COMMUNITY == "HA") |> 
+            dplyr::select(c("COMMUNITY", 
+                            "c_COMM_prev_LIMAT_under18", 
+                            "c_kid_perc",
+                            "c_OTGC_perc")) |> as.matrix())
+
+tdata_1 <- insertRow(tdata_1,4,
+          SAvail_im_COMM_changes_df |> 
+            filter(mode == "motorized" & COMMUNITY == "DU") |> 
+            dplyr::select(c("COMMUNITY", 
+                            "c_COMM_prev_LIMAT_under18", 
+                            "c_kid_perc",
+                            "c_OTGC_perc")) |> as.matrix())
+
+tdata_1 <- insertRow(tdata_1,6,
+          SAvail_im_COMM_changes_df |> 
+            filter(mode == "motorized" & COMMUNITY == "SC") |> 
+            dplyr::select(c("COMMUNITY", 
+                            "c_COMM_prev_LIMAT_under18", 
+                            "c_kid_perc",
+                            "c_OTGC_perc")) |> as.matrix())
+
+tdata_1 <- insertRow(tdata_1,8,
+          SAvail_im_COMM_changes_df |> 
+            filter(mode == "motorized" & COMMUNITY == "AN") |> 
+            dplyr::select(c("COMMUNITY", 
+                            "c_COMM_prev_LIMAT_under18", 
+                            "c_kid_perc",
+                            "c_OTGC_perc")) |> as.matrix())
+
+tdata_1 <- insertRow(tdata_1,10,
+          SAvail_im_COMM_changes_df |> 
+            filter(mode == "motorized" & COMMUNITY == "FL") |> 
+            dplyr::select(c("COMMUNITY", 
+                            "c_COMM_prev_LIMAT_under18", 
+                            "c_kid_perc",
+                            "c_OTGC_perc")) |> as.matrix())
+
+tdata_1 <- insertRow(tdata_1,12,
+          SAvail_im_COMM_changes_df |> 
+            filter(mode == "motorized" & COMMUNITY == "GL") |> 
+            dplyr::select(c("COMMUNITY", 
+                            "c_COMM_prev_LIMAT_under18", 
+                            "c_kid_perc",
+                            "c_OTGC_perc")) |> as.matrix())
+```
+
+```{r}
+tdata_2 <- SAvail_im_COMM_changes_df |> 
+  filter(mode == "motorized")|> 
+  dplyr::select(c("COMMUNITY",
+                  "V_im_2016",
+                  "v_im_2016",
+                  "emissions_2016",
+                  "emissions_perkid_2016"))
+
+tdata_2 <- insertRow(tdata_2 |> as.matrix(),2,
+          SAvail_im_COMM_changes_df |> 
+            filter(mode == "motorized" & COMMUNITY == "HA") |> 
+            dplyr::select(c("COMMUNITY", 
+                            "c_V_im_perc", 
+                            "c_v_im_perc",
+                            "c_emissions_perc",
+                            "c_emissions_perkid_perc")) |> as.matrix())
+
+tdata_2 <- insertRow(tdata_2,4,
+          SAvail_im_COMM_changes_df |> 
+            filter(mode == "motorized" & COMMUNITY == "DU") |> 
+            dplyr::select(c("COMMUNITY", 
+                            "c_V_im_perc", 
+                            "c_v_im_perc",
+                            "c_emissions_perc",
+                            "c_emissions_perkid_perc")) |> as.matrix())
+
+tdata_2 <- insertRow(tdata_2,6,
+          SAvail_im_COMM_changes_df |> 
+            filter(mode == "motorized" & COMMUNITY == "SC") |> 
+            dplyr::select(c("COMMUNITY", 
+                            "c_V_im_perc", 
+                            "c_v_im_perc",
+                            "c_emissions_perc",
+                            "c_emissions_perkid_perc")) |> as.matrix())
+
+tdata_2 <- insertRow(tdata_2,8,
+          SAvail_im_COMM_changes_df |> 
+            filter(mode == "motorized" & COMMUNITY == "AN") |> 
+            dplyr::select(c("COMMUNITY", 
+                            "c_V_im_perc", 
+                            "c_v_im_perc",
+                            "c_emissions_perc",
+                            "c_emissions_perkid_perc")) |> as.matrix())
+
+tdata_2 <- insertRow(tdata_2,10,
+          SAvail_im_COMM_changes_df |> 
+            filter(mode == "motorized" & COMMUNITY == "FL") |> 
+            dplyr::select(c("COMMUNITY", 
+                            "c_V_im_perc", 
+                            "c_v_im_perc",
+                            "c_emissions_perc",
+                            "c_emissions_perkid_perc")) |> as.matrix())
+
+tdata_2 <- insertRow(tdata_2,12,
+          SAvail_im_COMM_changes_df |> 
+            filter(mode == "motorized" & COMMUNITY == "GL") |> 
+            dplyr::select(c("COMMUNITY", 
+                            "c_V_im_perc", 
+                            "c_v_im_perc",
+                            "c_emissions_perc",
+                            "c_emissions_perkid_perc")) |> as.matrix())
+```
+
+```{r}
+tdata_3 <- SAvail_im_COMM_changes_df |> 
+  filter(mode == "non-motorized")|> 
+  dplyr::select(c("COMMUNITY",
+                  "V_im_2016",
+                  "v_im_2016",
+                  "total_tt_2016",
+                  "total_tt_perkid_2016"))
+
+tdata_3 <- insertRow(tdata_3 |> as.matrix(),2,
+          SAvail_im_COMM_changes_df |> 
+            filter(mode == "non-motorized" & COMMUNITY == "HA") |> 
+            dplyr::select(c("COMMUNITY", 
+                            "c_V_im_perc", 
+                            "c_v_im_perc",
+                            "c_total_tt_perc",
+                            "c_total_ttperkid_perc")) |> as.matrix())
+
+tdata_3 <- insertRow(tdata_3,4,
+          SAvail_im_COMM_changes_df |> 
+            filter(mode == "non-motorized" & COMMUNITY == "DU") |> 
+            dplyr::select(c("COMMUNITY", 
+                            "c_V_im_perc", 
+                            "c_v_im_perc",
+                            "c_total_tt_perc",
+                            "c_total_ttperkid_perc")) |> as.matrix())
+
+tdata_3 <- insertRow(tdata_3,6,
+          SAvail_im_COMM_changes_df |> 
+            filter(mode == "non-motorized" & COMMUNITY == "SC") |> 
+            dplyr::select(c("COMMUNITY", 
+                            "c_V_im_perc", 
+                            "c_v_im_perc",
+                            "c_total_tt_perc",
+                            "c_total_ttperkid_perc")) |> as.matrix())
+
+tdata_3 <- insertRow(tdata_3,8,
+          SAvail_im_COMM_changes_df |> 
+            filter(mode == "non-motorized" & COMMUNITY == "AN") |> 
+            dplyr::select(c("COMMUNITY", 
+                            "c_V_im_perc", 
+                            "c_v_im_perc",
+                            "c_total_tt_perc",
+                            "c_total_ttperkid_perc")) |> as.matrix())
+
+tdata_3 <- insertRow(tdata_3,10,
+          SAvail_im_COMM_changes_df |> 
+            filter(mode == "non-motorized" & COMMUNITY == "FL") |> 
+            dplyr::select(c("COMMUNITY", 
+                            "c_V_im_perc", 
+                            "c_v_im_perc",
+                            "c_total_tt_perc",
+                            "c_total_ttperkid_perc")) |> as.matrix())
+
+tdata_3 <- insertRow(tdata_3,12,
+          SAvail_im_COMM_changes_df |> 
+            filter(mode == "non-motorized" & COMMUNITY == "GL") |> 
+            dplyr::select(c("COMMUNITY", 
+                            "c_V_im_perc", 
+                            "c_v_im_perc",
+                            "c_total_tt_perc",
+                            "c_total_ttperkid_perc")) |> as.matrix())
+```
+
+```{r include=FALSE}
+tdata_1 <- t(tdata_1) |> data.frame() 
+tdata_2 <- t(tdata_2) |> data.frame() 
+tdata_3 <- t(tdata_3) |> data.frame()
+
+tdata <- rbind(tdata_1,tdata_2,tdata_3)
+tdata <- tdata[-c(5,10),]
+rownames(tdata) <- c("", "LIM-AT", "Pop.", "OTGC", 
+                     "V (mt)", "v (mt)", "GHG", "GHG/pop",
+                     "V (nmt)", "v (nmt)", "tt", "tt/pop")
+```
+
+```{r}
+#| label: tbl-Tab1
+#| tbl-cap: "Spatial availability and associated dimensions aggregated by communities"
+#| 
+tdata_hux <- as_hux(tdata) 
+tdata_hux[1,1:12] <- c("Hamilton C.", "", "Dundas" ,"", "Stoney Creek", "", "Ancaster", "", "Flamborough", "", "Glanbrook", "")
+# tdata_hux[1,1:12] <- c("HA", "", "DU" ,"", "SC", "", "AN", "", "FL", "", "GL", "")
+tdata_hux[2,1:12] <- c("", "% Δ", "" , "% Δ", "", "% Δ", "", "% Δ", "", "% Δ", "", "% Δ")
+
+options("huxtable.latex_use_fontspec" = TRUE) #use the same font as the document
+
+tdata_hux %>% 
+  insert_row("Motorized", fill = "", colspan=12, after=5) %>% 
+  insert_row("Non-motorized", fill = "", colspan=12, after=10) %>% 
+  merge_across(1,c(1:2)) %>%
+  merge_across(1,c(3:4)) %>%
+  merge_across(1,c(5:6)) %>%
+  merge_across(1,c(7:8)) %>%
+  merge_across(1,c(9:10)) %>%
+  merge_across(1,c(11:12)) %>%
+  add_rownames() |>
+
+  set_text_color(c(3,10), 3, "darkgreen") |>
+  set_text_color(c(4,5,7,8,9,12,13,14), 3, "darkred") |>
+
+  set_text_color(c(3), 5, "darkgreen") |>
+  set_text_color(c(4,7,9,12,13,14), 5, "darkred") |>
+
+  set_text_color(c(3,4,5,7,9,15), 7, "darkgreen") |>
+  set_text_color(c(12,13,14), 7, "darkred") |>
+
+  set_text_color(c(3,4,5,7,9,10,15), 9, "darkgreen") |>
+  set_text_color(c(8,12,13,14), 9, "darkred") |>
+
+  set_text_color(c(3,5,8,15), 11, "darkgreen") |>
+  set_text_color(c(4,7,9,12,13,14,15), 11, "darkred") |>
+
+  set_text_color(c(4,5,7,9,10), 13, "darkgreen") |>
+  set_text_color(c(3,12,13,14,15), 13, "darkred") |>
+  set_font_size(everywhere, everywhere, 6) |> 
+  set_all_padding(0) |>
+  set_bold(everywhere, 1, value = TRUE) |>
+  set_bold(1:2, everywhere, value = TRUE) |>
+  set_bold(c(6,11), everywhere, value = TRUE) |>
+  set_contents(c(1,6,11), 1, "") |>
+  set_right_border(c(3:15), c(3,5,7,9,11), brdr(1, "solid", "grey")) |>
+  #set_bottom_border(c(2,5,6,10,11), c(2:13), brdr(1, "solid", "grey")) |>
+  set_align(everywhere, everywhere, "left")|>
+  set_align(c(1,2,6,11), everywhere, "center")|>
+  set_caption("Spatial availability and associated dimensions aggregated by communities") |>
+  set_label("tbl-Tab1") %>% as_flextable()
+
+```
+
+Due to spatial availability's flexible proportional allocation mechanism and this cases' calculation at the parcel level, results can also be summarized by community (@tbl-Tab1).
+The majority of schools were closed in Hamilton Central.
+This community represents `r SAvail_im_COMM_changes |> st_drop_geometry() |> filter(COMMUNITY == "HA") |> dplyr::select("kid_proportion_2016") |> sum() |> scales::percent()` of the 5-14 year old population in 2016.
+Though the population between 2011 and 2016 decreased, the rate of school-seat spatial availability decreased disproportionately more (population by `r SAvail_im_COMM_changes |> st_drop_geometry() |> filter(COMMUNITY == "HA") |> transmute(calc = ((sum(kid_2016)-sum(kid_2011))/sum(kid_2011)) |> scales::percent()) |> pull() |> first()` and spatial availability by `r SAvail_im_COMM_changes |> st_drop_geometry() |> filter(COMMUNITY == "HA") |> transmute(calc = ((sum(V_im_2016)-sum(V_im_2011))/sum(V_im_2011)) |> scales::percent()) |> pull() |> first()`).
 This disproportionate decrease is greatest in Hamilton Central than any other community.
-This may have equity concerns as Hamilton Central has the highest LIM-AT prevalence, with `r SAvail_im_COMM_changes |> st_drop_geometry() |> filter(COMMUNITY == "HA" & mode == "motorized") |> dplyr::select("prev_LIMAT_2016") |> round() |> paste0("%")` of households LIM-AT in 2016, 3.7 times greater than the community with the lowest level of LIM-AT prevalence in the city.
-Between 2016 and 2011, Hamilton Center had the lowest levels of emissions per student and saw the highest per student increase in 2016.
+This may have equity concerns as Hamilton Central has the highest LIM-AT prevalence, with `r SAvail_im_COMM_changes_df |> filter(COMMUNITY == "HA" & mode == "motorized") |> dplyr::select("COMM_prev_LIMAT_under18_2016") |> paste0("%")` of households LIM-AT in 2016, 3.7 times greater than Glanbrook, a community with the lowest level of LIM-AT prevalence in the city.
+Between 2016 and 2011, Hamilton Central had the lowest levels of emissions per student and saw the highest per student increase in 2016.
 Hamilton Center had the highest non-motorized trip potential, and it was reduced in part by the school closure policies.
@@
 
-Across the city, those who commute by non-motorized mode see the most dramatic drop in spatial availability across the city.
+Across the city, those who commute by non-motorized mode see the most drastic lose in spatial availability.
 In all communities, spatial availability decreases at greater levels than population.
-Notably, in Dundas (2nd highest LIM-AT), the only school in the community closed leaving the population in the community with no other option but to use motorized modes.
-Glanbrook (lowest LIM-AT), which experienced a sharp increase in student population along with non-motorized using population, suffered a lose in non-motorized spatial availability despite seeing the highest increase in motorized spatial availability.
+Notably in Dundas (2nd highest LIM-AT), the only school in the community closed, leaving the population in the community with no other option but to use motorized modes.
+Glanbrook (lowest LIM-AT), which experienced a relatively sharp increase in student population along with non-motorized using population, suffered a lose in non-motorized spatial availability despite seeing the highest increase in motorized spatial availability.
 Overall, the gain in motorized spatial availability resulted in a further erosion of non-motorized spatial availability.
-This erosion results in a lose in active transport trips and an increase in emissions city-wide (`r SAvail_im_COMM_changes |> st_drop_geometry() |> transmute(em = (sum(emissions_2011))/(sum(kid_2011)) ) |> sum() |> round(digit=2)`kg $CO_2e$ per home-to-school potential student-trip in 2011 to `r SAvail_im_COMM_changes |> st_drop_geometry() |> transmute(em = (sum(emissions_2016))/(sum(kid_2016)) ) |> sum() |> round(digit=2)` in 2016).
-
-The advantage of spatial availability is the constrained opportunities (singly constrained). This means we can:
-- calculate availability at the finest spatial resolution
-- we can aggregate it whichever spatial unit is most meaningful for interpretation.
-Here we demonstrate aggregations at the DA, then at neighbourhood community boundaries, and then city wide per mode.
-This can be used to communicate changes in spatial availability and other related potential mobility externalities like lose active transportation minutes and emissions.
+This erosion is related to a lose in potential active home-to-school travel minutes (`r ((SAvail_im_COMM_changes |> filter(mode == "non-motorized") |> pull(total_tt_2011)  |> sum()) -  (SAvail_im_COMM_changes |> filter(mode == "non-motorized") |> pull(total_tt_2016)  |> sum())) |> round(digit=2)` fewer minutes than in 2011, decreasing the potential active travel time per (all mode using) student from `r ((SAvail_im_COMM_changes |> filter(mode == "non-motorized") |> pull(total_tt_2011)  |> sum()) / (SAvail_im_COMM_changes |> st_drop_geometry() |> pull(kid_2011) |> sum())) |> round(digit=2)` to `r ((SAvail_im_COMM_changes |> filter(mode == "non-motorized") |> pull(total_tt_2016)  |> sum()) / (SAvail_im_COMM_changes |> st_drop_geometry() |> pull(kid_2016) |> sum())) |> round(digit=2)` minutes), as well as an increase in potential GHG emissions city-wide (`r (SAvail_im_COMM_changes |> st_drop_geometry() |> pull(emissions_2016) |> sum()) -  (SAvail_im_COMM_changes |> st_drop_geometry() |> pull(emissions_2011) |> sum()) |> round(digit=2)` kg $CO_2e$ more emission than in 2011, increasing the kg $CO_2e$ per child rate for home-to-school trips from `r ((SAvail_im_COMM_changes |> st_drop_geometry() |> pull(emissions_2011) |> sum()) / (SAvail_im_COMM_changes |> st_drop_geometry() |> pull(kid_2011) |> sum())) |> round(digit=2)` to `r ((SAvail_im_COMM_changes |> st_drop_geometry() |> pull(emissions_2016) |> sum()) / (SAvail_im_COMM_changes |> st_drop_geometry() |> pull(kid_2016) |> sum())) |> round(digit=2)`).
 
@@
 
-In this paper, we estimate the change in multi-modal spatial availability as a result of school consolidations/closures between the years of 2011 and 2016 in an empirical case study of Hamilton, a mid-size city (\~550,000 pop) in Ontario, Canada.
-The spatial availability and spatial availability per student are visualized on a DA level; spatial availability for the city declined overall.
-This decline is then discussed by measuring potential walk minutes and GHG emissions (produced by additional car minutes that result since walk minutes are removed).
-Overall, the closure of `r ((ALL_SCHOOLS_Elem %>% filter(Year != "2016" & Level == "Elementary") %>% pull(SchoolID) %>% length() - ALL_SCHOOLS_Elem %>% filter(Year != "2011" & Level == "Elementary") %>% pull(SchoolID) %>% length())/ALL_SCHOOLS_Elem %>% filter(Year != "2016" & Level == "Elementary") %>% pull(SchoolID) %>% length()) %>% scales::percent()` of elementary schools in Hamilton resulted in: a `r (ALL_SCHOOLS_Elem %>% st_drop_geometry() %>% filter( (Year == "2011" | Year == "2011 and 2016") & Level == "Elementary") %>% dplyr::select(OTGC2011) %>% sum() - ALL_SCHOOLS_Elem %>% st_drop_geometry() %>% filter( (Year == "2016" | Year == "2011 and 2016") & Level == "Elementary") %>% dplyr::select(OTGC2016) %>% sum()) %>% prettyNum(big.mark=",", digits=1)` decline of school-seat capacity, a decline of `r SAvail_im_COMM_changes |> st_drop_geometry() |> filter(mode == "non-motorized") |> transmute(total_tt = total_tt_2016-total_tt_2011) |> sum() %>% abs() %>% prettyNum(big.mark=",", digits=1)` *potential* home-to-school walk-minutes (per one-way trips), an increase of `r SAvail_im_COMM_changes |> st_drop_geometry() |> filter(mode == "motorized") |> transmute(emissions = emissions_2016-emissions_2011) |> sum() %>% abs() %>% prettyNum(big.mark=",", digits=1)` kg CO2e as a result of increased car-minutes as a result of lost *potential* home-to-school walk-minutes (per one-way trips).
-
-By quantifying the spatial availability, environmental, and active-travel implications of school closure/consolidation policies in Hamilton, we offer a methodology for spatial policy analysis scenario.
-This methodology can be used by researchers and decision-makers to plan and evaluate equitable and sustainable urban planning policies from a spatial perspective.
-We urge policy makers to view 'spatial availability' of public resources from a per-capita perspective, plan capacity and location of service provision with the cost of travel (and associated implications such as GHG emissions, active transportation mode, safety of active travel modes, etc.) that sufficiently serves the target population.
-In this paper we suggest and assume a benchmark of 1.0 school-seats per student which accounts for sufficiency.
+In this paper, we estimate the change in multimodal spatial availability as a result of school consolidations/closures between the years of 2011 and 2016 in an empirical case study of Hamilton, a mid-size city (\~570,000 pop) in Ontario, Canada [@governmentofcanadaProfileTableCensus2022].
+The spatial availability and spatial availability per student are calculated at the parcel level and aggregated at the Canadian Census DA spatial unit and at the six Hamilton communities for comparison.
+Overall, schools closed and the number of school seats decreased so spatial availability declined overall: in this work, we also discuss where, for whom, and with what potential impacts.
+
+Overall, the closure of `r ((ALL_SCHOOLS_Elem %>% filter(Year != "2016" & Level == "Elementary") %>% pull(SchoolID) %>% length() - ALL_SCHOOLS_Elem %>% filter(Year != "2011" & Level == "Elementary") %>% pull(SchoolID) %>% length())/ALL_SCHOOLS_Elem %>% filter(Year != "2016" & Level == "Elementary") %>% pull(SchoolID) %>% length()) %>% scales::percent()` of elementary schools in Hamilton between 2011 and 2016 resulted in `r (((SAvail_im_COMM_changes$V_im_2016 |> sum()) - (SAvail_im_COMM_changes$V_im_2011 |> sum()))  / SAvail_im_COMM_changes$V_im_2011 |> sum()) |> scales::percent()` decline in school-seat availability city-wide (`r (ALL_SCHOOLS_Elem %>% st_drop_geometry() %>% filter( (Year == "2011" | Year == "2011 and 2016") & Level == "Elementary") %>% dplyr::select(OTGC2011) %>% sum() - ALL_SCHOOLS_Elem %>% st_drop_geometry() %>% filter( (Year == "2016" | Year == "2011 and 2016") & Level == "Elementary") %>% dplyr::select(OTGC2016) %>% sum()) %>% prettyNum(big.mark=",", digits=1)` fewer school-seats), but specifically a `r (((SAvail_im_COMM_changes |> filter(mode == "non-motorized") |> pull(V_im_2016) |> sum()) - (SAvail_im_COMM_changes  |> filter(mode == "non-motorized") |> pull(V_im_2011) |> sum()))  / SAvail_im_COMM_changes  |> filter(mode == "non-motorized") |> pull(V_im_2011) |> sum()) |> scales::percent()` decline in non-motorized spatial availability.
+Though non-motorized modal share was low in both 2011 and 2016, the _potential_ for active transport trips was significantly eroded as schools within a 27 minute walk catchment were closed in the more densely populated and central areas of the city.
+Further, the communities with the highest LIM-AT prevalence were the most impacted: communities like Hamilton Central and Dundas saw the most drastic loss in non-motorized spatial availability while communities with lower LIM-AT prevalence saw gains in motorized spatial availability (as a handful of rural schools opened).
+The motivation of operational savings associated with public service consolidation for the school boards did not account for the mobility-related burdens _certain_ families are now saddled with.
+
+In addition to spatial availability impacts, associated mobility measures are also assessed.
+City-wide, home-to-school non-motorized minutes (1 day) reduced by `r SAvail_im_COMM_changes |> st_drop_geometry() |> filter(mode == "non-motorized") |> transmute(total_tt = total_tt_2016-total_tt_2011) |> sum() %>% abs() %>% prettyNum(big.mark=",", digits=1)` and GHG emissions associated with motorized home-to-school travel (1 day)  increased by `r SAvail_im_COMM_changes |> st_drop_geometry() |> filter(mode == "motorized") |> transmute(emissions = emissions_2016-emissions_2011) |> sum() %>% abs() %>% prettyNum(big.mark=",", digits=1)` kg $CO_2e$.
+The decision to consolidate schools, increase the length of motorized travel and decrease the potential for non-motorized impacts will have daily impacts on students and their families [@mandicAdolescentsPerceptionsWalking2022; @pabayoImportanceActiveTransportation2012], the broader community [@pietrabissaSchoolAccessCity2023;@MerrallWhatsASchool2024; @bittencourtEvaluatingAccessibilityAvailability2023] and the environment [@pantelakiImpactHomeschoolCommuting2024; @rongReviewResearchLowcarbon2022].
+
+Like many other localities [@Lee_Lubienski_2017; @Autti_Hyry-Beihammer_2014; @Beuchert_Humlum_Nielsen_Smith_2018], Hamilton is not immune to top-down operational efficiency assessments that determine if community infrastructure is underutilized and closed.
+From the perspective of the provincial government of Ontario, the school closures which occurred in Hamilton resulted in operational savings-per-student.
+In this work, we demonstrate how these decisions resulted in both families and students paying a price through spatial availability, lost potential walk minutes, and increased GHG emissions.
+ <!-- In the case of Hamilton Central, where households have the highest prevalence of LIM-AT in the city, those households suffer the most disproportionate decrease in motorized spatial availability and increase in percentage increase in emissions per capacity.
+As public schools should serve all students; the determination of spatial availability and spatial equity of impact, is critical to better inform decision-makers when evaluating school consolidation policies -->
+
+By quantifying the spatial availability, environmental, and active-travel implications of school closure/consolidation policies in Hamilton, our paper offers a methodology for spatial policy analysis scenario.
+The presented methodology can be used by researchers and decision-makers to plan and evaluate equitable and sustainable urban planning policies from a spatial perspective.
+At the core of the method is spatial availability [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024], a singly-constrained accessibility measure that can be calculated at the finest spatial resolution (in our paper, parcel-level) and aggregated at whichever spatial unit is most meaningful for interpretation.
+We urge policy makers to view the 'spatial availability' of public resources from a per-capita perspective, plan capacity and location of service provision with the cost of travel (and associated implications such as GHG emissions, active transportation mode, safety of active travel modes, etc.) that sufficiently serves the target population.
+In this work, we suggest and assume a benchmark of 1.0 school-seats per student which accounts for sufficiency.
 Spatial availability can be used to obtain this benchmark since the total number of opportunities (i.e., school capacity) is preserved in the region of analysis unlike unconstrained forms of accessibility measurement e.g., Hansen-type measure [@hansenHowAccessibilityShapes1959].
 As such, the assigned spatial availability can be divided by population at each zone to yield an interpretable per capita measurement.
-
-Like many other localities, Hamilton is not immune to top-down financial efficient policy decisions to determine community infrastructure is underutilized and it be closed.
-From the perspective of the provincial government of Ontario, the school closures which occurred in Hamilton resulted in operational savings-per-student but we demonstrate how this resulted in both families and students paying a price through spatial availability, lost potential walk minutes, and increased GHG emissions.
-In the case of Hamilton Central, where households have the highest prevalence of LIM-AT in the city, those households suffer the most disproportionate decrease in motorized spatial availability and increase in percentage increase in emissions per capacity.
-As public schools should serve all students; the determination of spatial availability and spatial equity of impact, is critical to better inform decision-makers when evaluating school consolidation policies.
 
@@
 
-Study limitations:
-
-- potential students per households, potential trips to all schools
-- estimated travel times, direct car or walk trip (shortest path).
-Not necessarily car, could be transit or school bus.
-Walking shortest-paths are also optimistic for children's travel: could be unsafe and not taken in reality and/or alternative more pleasant paths are taken.
-- estimated emissions. GHG emissions per minute of driving varies extensively depending on operating conditions and vehicle characteristics [@soukhovOccupancyGHGEmissions2022].
-
-<!-- Add this additional paper: it discusses the potential impact of mode-switching on GHG emissions reductions [@pantelakiImpactHomeschoolCommuting2024].
-
-Also, access to schools has broader implications; it has a relationship to residential and job choices [@pietrabissaSchoolAccessCity2023] and housing prices [@merrallWhatSchoolWorth2023] ?
-Shall I further elaborate in the discussion?-->
+As with all studies, our work is associated with limitations on how results should be interpreted.
+The calculated spatial availability assumes any residential parcel can have a student (at the DA rate of students per residential unit) so all parcels are assigned a DA rate weighted by how many residential units are contained within the parcel.
+Each parcel can also visit any school, though travel times based on observed origin-destination behaviour (from the TTS) are more likely through the use of the empirical travel impedance functions.
+In this way, the spatial availability demonstrates what _potential interaction_ a student could have based on historic and average travel behaviour and shortest network-path travel times if residing in point in space.
+This representation is average, and by design obfuscates considerations that could make non-average and shortest-path travel unrealistic such as safety and built-environment concerns [@leeNeighborhoodEnvironmentsUtilitarian2021; @yumitaSchoolCommutingBarriers2021] nor considers significant non-transportation factors [@palmRolePublicTransit2020].
+Further, this study does not consider transit, school-bus, or cycling modes which have implications on how the GHG emissions estimation can be interpreted.
+Transit and school bus modes often taken more circuitous routes [@yumitaSchoolCommutingBarriers2021] but pool students thereby reducing emissions.
+Cycling reduces travel time over the same walking distance at no extra emitted GHG.
+Lastly, this study also ascribes a static emission factor to represent motorized travel minutes based on an average diesel passenger car.
+In reality, GHG emissions emitted per minute of driving varies extensively depending on operating conditions and vehicle characteristics [@soukhovOccupancyGHGEmissions2022].
+In these ways, the assumptions made are to illustrate _potential_ and _average_ changes that resulted in a backdrop of school consolidation/closures to illustrate accessibility-related impacts for the city of Hamilton.
 

### 2024-05-07 paper/Manuscript.qmd
AP edits before submission

@@
 Public schools have for some time now been pressured to do more with less, and governments are opting to close under-capacity schools and expand existing schools to accommodate lost capacity.
-The objective of this study is to quantifies the travel burdens of school consolidation/closure policy.
+The objective of this study is to quantify the travel burdens of school consolidation/closure policy.
 We do this through the case of Hamilton, a mid-sized city in Ontario, Canada.
 Between the years 2011 and 2015, 7% of the elementary schools closed in the city.
-For this research, we quantify how these closures changed the accessibility landscape for the elementary school aged population, and the active travel and emission implications.
+For this research, we study how these closures changed the accessibility landscape for the elementary school aged population, and the active travel and emission implications.
 In terms of methods, we use spatial availability, a singly-constrained multimodal measure of spatial accessibility.
@@
 Analysis is conducted at the parcel-level with all school seats being eligible and travel behaviour based on observed home-to-school patterns for motorized and non-motorized modes.
-The system-wide impacts that result from the closures, such as potential GHG emission increases and walk minutes decreases, are estimated Overall, spatial availability decreased in the city.
-Though the school consolidation policy may have resulted in cost-savings in operation and maintenance of facilities, families, students, and the environment now pay the price.
+We report estimates of system-wide impacts that result from the closures, such as potential growth in GHG emission and reduction of walking minutes.
+Overall, spatial availability decreased in the city.
+Though the school consolidation policy may have resulted in cost-savings in operation and maintenance of facilities, students, families, and communities pay the price in reduced opportunities for physical activity and greater pollution.
 keywords: [accessibility; spatial availability; carbon emissions; journey-to-school; walkability; active transportation; school consolidation policy; service provision thresholds]
@@
 
-Communities are by nature in flux: economic and demographic shifts, rural and urban configurations, and political pressures stoke changes in community characteristics.
+Communities are by nature in flux: economic and demographic shifts, rural and urban configurations, and political pressures all combine to produce change.
 These forces spur the re-organization and consolidation of public facilities in communities around the world where changes to service provision may improve the quality of service for some but not for all [@Christiaanse_2020; @Rosik_Pulawska-Obiedowska_Goliszek_2021].
-While change is natural, however, oftentimes top-down consolidation decisions that are driven by operational cost-savings do not take into consideration the full extent of losses in benefits.
-
-From this perspective, school closure policies have been pursued by governments to optimize per-student operational cost-efficiency [@rongReviewResearchLowcarbon2022; @Dai_Wang_Zhang_Liao_Liu_2019] but other implications, such as the impact of closures on property values [@MerrallWhatsASchool2024], the effects on the social infrastructure of communities [@butlerClosureRideau2019a; @irwinSchoolClosure2012], and particularly the travel implications of school closures have often been overlooked by decision makers [@bierbaumMobilityJusticeLinking2021; @Lee_Lubienski_2017].
-This seems a mistake: journey-to-school trips are tremendously important, and travel for this purpose is the largest contributor of carbon emissions from personal travel after the journey-to-work [@rongReviewResearchLowcarbon2022; @pantelakiImpactHomeschoolCommuting2024].
-In many communities globally, schools are closing [@sageman2022].
+While change is natural, oftentimes top-down consolidation decisions are driven by operational cost-savings that do not take into consideration the full extent of losses in benefits.
+
+From this perspective, school closure policies have been pursued by governments to optimize per-student operational cost-efficiency [@rongReviewResearchLowcarbon2022; @Dai_Wang_Zhang_Liao_Liu_2019] but other implications, such as the impact of closures on property values [@MerrallWhatsASchool2024], their effects on the social infrastructure of communities [@butlerClosureRideau2019a; @irwinSchoolClosure2012], and particularly their travel implications have often been overlooked by decision makers [@bierbaumMobilityJusticeLinking2021; @Lee_Lubienski_2017].
+This seems a mistake: journey-to-school trips are tremendously important, and travel for this purpose is currently the largest contributor of carbon emissions from personal travel after the journey-to-work [@rongReviewResearchLowcarbon2022; @pantelakiImpactHomeschoolCommuting2024], where it could be instead an excellent opportunity for physical activity [@desjardinsFramingActive2024].
+Alas, in many communities globally, schools are closing [@sageman2022].
 For instance, 65% of comprehensive schools (e.g.
@@
 In exchange for less physical activity, communities get more motorized travel, with the concomitant costs and burdens, including emissions.
-Quantification of losses in potential active travel and associated growith in carbon emissions is scarce in the literature, and the focus of this paper, which aims to make both methodological and empirical contributions.
-We take a novel multimodal and singly-constrained accessibility measure, called spatial availability [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024], and use it to estimate the changes in accessibility and travel before and after a recent wave of school closure/consolidations implemented in Hamilton, Ontario.
-
-The results of the research indicate that, overall, school seats became less spatially available in the city.
-Though the school consolidation policy pursued by the province and implemente by the local school board may have resulted in cost-savings in operation and maintenance of facilities, families, students, and the environment now pay the price in terms of lost opportunities for healthier active travel, as well as increased emissions from motorized travel.
+Quantification of losses in potential active travel and associated growith in carbon emissions is scarce in the literature, and the focus of this paper.
+In addition to its empirical contributions, the paper makes use of a novel multimodal and singly-constrained accessibility measure, called spatial availability [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024], which we use to estimate the changes in accessibility and travel after a recent wave of school closures/consolidations in Hamilton, Ontario.
+
+The results of the research indicate that, overall, school seats became less spatially available in the city, although this happened unevenly, with some communities being disproportionately affected.
+Though the school consolidation policy pursued by the province and implemented by the local school board may have resulted in cost-savings in operation and maintenance of facilities, families, students, and communities now pay the price in terms of lost opportunities for healthier active travel, as well as increased emissions from motorized travel.
 
@@
 In 2013, the Hamilton-Wentworth District School Board (HWDSB), the largest public English school board in the city, released its Long Term Facilities Master Plan [@hwdsbLongTermFacilities2013].
-To ensure "equitable, affordable and sustainable learning facilities", the Master Plan indicated that 80% of it's elementary schools (i.e., schools attended by students aged 5 to 13) would be subjected to accommodation evaluations, including an "Accommodation Review" whose outcome historically has led to school closures/consolidation [@hwdsbLongTermFacilities2013; @craggsBoardEvaluatingFuture2013; @seasonsSchoolClosurePolicy2014].
-These decisions by the school board were in part motivated by a search for operational savings based on projected reductions in student population and under-utilized school capacity [@craggsAxeCouldFall2012].
-However, the case of Hamilton and the HWDSB was unlike others in the province: the HWDSB underwent an unprecedented number of Accommodation Reviews relative to other school boards [@seasons2014ARCCatalogues2014].
-
-Though the Accommodation Review guidelines outline a community-focused process to deliberate a school's future viability to students, the community and the economy [@seasonsSchoolClosurePolicy2014; @ministryofeducationPupilAccommodationReview2006], concerned residents felt a wholesome breakdown of costs to repair schools was not made consistently available [@kleinhuisSeriousConcernsSchool2013].
+To ensure "equitable, affordable and sustainable learning facilities", the Master Plan indicated that 80% of it's elementary schools (i.e., schools attended by students aged 5 to 13) would be subjected to accommodation evaluations, including an "Accommodation Review", the outcome of which historically has been school closures/consolidations [@hwdsbLongTermFacilities2013; @craggsBoardEvaluatingFuture2013; @seasonsSchoolClosurePolicy2014].
+These decisions pushed by the province were in part motivated by a search for operational savings based on projected reductions in student population and under-utilized school capacity [@craggsAxeCouldFall2012], and Hamilton was particularly affected, as the HWDSB underwent an unprecedented number of Accommodation Reviews relative to other school boards [@seasons2014ARCCatalogues2014].
+
+Though the Accommodation Review guidelines outline a community-focused process to assess the value that a school provides to students, the community, and the local economy [@seasonsSchoolClosurePolicy2014; @ministryofeducationPupilAccommodationReview2006], in practice residents felt that a wholesome breakdown of costs to repair/maintain schools was not made consistently available [@kleinhuisSeriousConcernsSchool2013].
 Further, others pointed to flawed incentive structures motivating the HWDSB: "By closing three schools the HWDSB would have a strong case with the Ministry of Education to receive full funding for a new school" [@kleinhuisSeriousConcernsSchool2013].
@@
 
-In the wave of school accommodation reviews between 2013 and 2016, `r ALL_SCHOOLS_Elem %>% filter(Year == "2011" & Level == "Elementary" & Status == "Removed") %>% pull(SchoolID) %>% length()` elementary schools closed.
+In the wave of school accommodation reviews between 2013 and 2016, `r ALL_SCHOOLS_Elem %>% filter(Year == "2011" & Level == "Elementary" & Status == "Removed") %>% pull(SchoolID) %>% length()` elementary schools closed in Hamilton.
 These closures represented a `r ((ALL_SCHOOLS_Elem %>% filter(Year != "2016") %>% pull(SchoolID) %>% length() - ALL_SCHOOLS_Elem %>% filter(Year != "2011") %>% pull(SchoolID) %>% length())/ALL_SCHOOLS_Elem %>% filter(Year != "2016" & Level == "Elementary") %>% pull(SchoolID) %>% length()) %>% scales::percent()` and `r  (( ((avail_2011 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2011)) |> select(OTGC) |> sum())) - (avail_2016 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2016)) |> select(OTGC) |> sum())) / (avail_2016 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2016)) |> select(OTGC) |> sum()) ) |> scales::percent()` decline in elementary school locations and capacity between 2011 (before the school closures) and 2016, despite the fact that the overall student population in the city did not decline, and in fact increased slightly between 2011 (`r avail_2011$kid_popadj %>% sum(na.rm=T) %>% prettyNum(big.mark=",")` students) and 2016 (`r avail_2016$kid_popadj %>% sum(na.rm=T) %>% prettyNum(big.mark=",")` students).

### 2024-05-07 paper/Manuscript.qmd
Latest version

@@
 The TLDs for the motorized and non-motorized trips are shown in @fig-Fig4.
-For reference, the gamma distribution and the exponential distribution function are displayed in the following Equation (\ref{eq:exp-dist})) and (\ref{eq:gamma-dist})) where $x$ is $c_{ij}$:
+For reference, the gamma distribution and the exponential distribution function are displayed in Equations (\ref{eq:exp-dist}) and (\ref{eq:gamma-dist}) where $x$ is $c_{ij}$:
 

### 2024-05-09 paper/Manuscript.qmd
Version submitted to journal

@@
   - "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
-linenumbers: true
-numbersections: true
 format:
-  nature-pdf: default
+  nature-pdf: 
+    journal: "default"
+    keep-tex: true
 editor: source
@@
   freeze: auto
+
 author:
@@
         address: "Department of Human Geography, University of Toronto Scarborough, 1265 Military Trail, Toronto, ON, M1C 1A4, Canada"
-  - name: "Antonio Paez"
+  - name: "Antonio Páez"
     email: "paezha@mcmaster.ca"
@@
   - name: "Moataz Mohamed"
-    email: "mmohame@mcmaster.ca "
+    email: "mmohame@mcmaster.ca"
     affiliations:
@@
 abstract: | 
-  | Reducing emissions and increasing active travel are often top priorities for many communities.
-These priorities, though, often are confronted with a pressure to reduce the fiscal burden of public services, including schools.
-Public schools have for some time now been pressured to do more with less, and governments are opting to close under-capacity schools and expand existing schools to accommodate lost capacity.
-The objective of this study is to quantify the travel burdens of school consolidation/closure policy.
+  | Reducing emissions and increasing active travel are top priorities for many communities.
+Often, though, these priorities are confronted with pressures to reduce the fiscal burden of public services, including schools.
+Public schools have been strained to do more with less, and governments are opting to close under-capacity schools and expand existing schools to accommodate lost capacity.
+The objective of this study is to quantify the travel burdens of school consolidation/closure policies.
 We do this through the case of Hamilton, a mid-sized city in Ontario, Canada.
 Between the years 2011 and 2015, 7% of the elementary schools closed in the city.
-For this research, we study how these closures changed the accessibility landscape for the elementary school aged population, and the active travel and emission implications.
+We study how these closures changed the accessibility landscape for the elementary school aged population, as well as the active travel and emission implications.
 In terms of methods, we use spatial availability, a singly-constrained multimodal measure of spatial accessibility.
@@
 We report estimates of system-wide impacts that result from the closures, such as potential growth in GHG emission and reduction of walking minutes.
-Overall, spatial availability decreased in the city.
-Though the school consolidation policy may have resulted in cost-savings in operation and maintenance of facilities, students, families, and communities pay the price in reduced opportunities for physical activity and greater pollution.
+Findings demonstrate that overall, spatial availability decreased in the city.
+Though the school consolidation policy may have resulted in cost-savings in operation and maintenance of facilities, students, families, and communities pay the price in reduced opportunities for physical activity and greater motorized mobility-related pollution.
 keywords: [accessibility; spatial availability; carbon emissions; journey-to-school; walkability; active transportation; school consolidation policy; service provision thresholds]
 ---
+<!--NOTE WHEN PREPARING THE .TEX for journal submission I ran into errors as their system uses PDFlatex+Bibtex to compile. The following packages only work in lualatex, so do the following:
+- remove "  \usepackage[T1]{fontenc}" 
+- remove "\usepackage{fontspec}" 
+- add "\DeclareUnicodeCharacter{394}{$\Delta$}" to the tex preamble (this is because PHDtex does not recognize unicode? So need to explicitly define) (add to line 5) -->
 
@@
 library(huxtable)
+library(tiff)
+library(grid)
 ```
@@
 ```{r descriptive-catchment-schools-plot2, eval=FALSE}
-tmap_save(catchment_schools_plot, paste0(here::here(),"/Figures/catchment_schools_plot.png"), height=7, width=12, dpi=600)
+tmap_save(catchment_schools_plot, paste0(here::here(),"/Figures/Fig1.tiff"), height=7, width=12, dpi=600)
 ```
@@
 ```{r student-popabs-lowincperc-poprate-plot2, eval=FALSE}
-tmap_save(stud_pop_rate_LIM_plot, paste0(here::here(),"/Figures/stud_pop_rate_LIM_plot.png"), height=10, width=12, dpi=600)
+tmap_save(stud_pop_rate_LIM_plot, paste0(here::here(),"/Figures/Fig2.tiff"), height=10, width=12, dpi=500)
 ```
@@
 
-tmap_save(OD_flow_plot, paste0(here::here(),"/Figures/OD_flow_plot.png"), height=7, width=12, dpi=600)
+tmap_save(OD_flow_plot, paste0(here::here(),"/Figures/Fig3.tiff"), height=7, width=12, dpi=600)
 ```
@@
 TLD_empirical_wtheo_curve_plot
-```
-
-```{r ploting-empirical-and-theo-curve-TLD-tts2, eval=FALSE}
-ggsave(paste0(here::here(),"/Figures/TLD_empirical_wtheo_curve_plot.png"), height=7, width=10, dpi=600)
+ggsave(paste0(here::here(),"/Figures/Fig4.tiff"), height=7, width=10, dpi=600)
 ```
@@
 Vim_vvim_20112016_plot <- tmap_arrange(Vim_20112016_plot, vvim_20112016_plot, nrow=2)
-tmap_save(Vim_vvim_20112016_plot, paste0(here::here(),"/Figures/Vim_vvim_20112016_plot.png"), height=12, width=12, dpi=600)
+
+tmap_save(Vim_vvim_20112016_plot, paste0(here::here(),"/Figures/Fig5.tiff"), height=12, width=12, dpi=450)
 ```
@@
   tmap_defaults +
-  tm_layout( bg.color = "grey85",
+  tm_layout(bg.color = "grey85",
              panel.label.bg.color = "white",
@@
 tmap_save(Vim_vvim_20112016_CHANGE_plot, 
-          paste0(here::here(),"/Figures/Vim_vvim_20112016_CHANGE_plot.png"),
+          paste0(here::here(),"/Figures/Fig6.tiff"),
           height=7,
@@
 tmap_save(ttim_GHGim_20112016_CHANGE_plot, 
-          paste0(here::here(),"/Figures/ttim_GHGim_20112016_CHANGE_plot.png"),
+          paste0(here::here(),"/Figures/Fig7.tiff"),
           height=5.5,
@@
 
-In this study, we focus on the transportation implications of school closure and consolidation policies.
+We focus on the transportation implications of school closure and consolidation policies.
 While some research has considered the "steps in reserve", that is, the potential for active travel hidden behind motorized trips [@morencyStepsReserve2009; @morencyHowMany2007], school closures potentially make actual steps vanish, sent to the reserve for the foreseeable future.
 In exchange for less physical activity, communities get more motorized travel, with the concomitant costs and burdens, including emissions.
-Quantification of losses in potential active travel and associated growith in carbon emissions is scarce in the literature, and the focus of this paper.
-In addition to its empirical contributions, the paper makes use of a novel multimodal and singly-constrained accessibility measure, called spatial availability [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024], which we use to estimate the changes in accessibility and travel after a recent wave of school closures/consolidations in Hamilton, Ontario.
-
-The results of the research indicate that, overall, school seats became less spatially available in the city, although this happened unevenly, with some communities being disproportionately affected.
+Quantification of losses in potential active travel and associated growth in carbon emissions is scarce in the literature, and the focus of this paper.
+In addition to its empirical contributions, this work makes use of a novel multimodal and singly-constrained accessibility measure, called spatial availability [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024], which we use to estimate the changes in accessibility and travel after a recent wave of school closures/consolidations in Hamilton, Ontario.
+
+The results of the research indicate that, overall, school seats became less spatially available in the city.
+This change happened unevenly, with some communities being disproportionately affected.
 Though the school consolidation policy pursued by the province and implemented by the local school board may have resulted in cost-savings in operation and maintenance of facilities, families, students, and communities now pay the price in terms of lost opportunities for healthier active travel, as well as increased emissions from motorized travel.
@@
 In 2013, the Hamilton-Wentworth District School Board (HWDSB), the largest public English school board in the city, released its Long Term Facilities Master Plan [@hwdsbLongTermFacilities2013].
-To ensure "equitable, affordable and sustainable learning facilities", the Master Plan indicated that 80% of it's elementary schools (i.e., schools attended by students aged 5 to 13) would be subjected to accommodation evaluations, including an "Accommodation Review", the outcome of which historically has been school closures/consolidations [@hwdsbLongTermFacilities2013; @craggsBoardEvaluatingFuture2013; @seasonsSchoolClosurePolicy2014].
-These decisions pushed by the province were in part motivated by a search for operational savings based on projected reductions in student population and under-utilized school capacity [@craggsAxeCouldFall2012], and Hamilton was particularly affected, as the HWDSB underwent an unprecedented number of Accommodation Reviews relative to other school boards [@seasons2014ARCCatalogues2014].
+To ensure "equitable, affordable and sustainable learning facilities", the Master Plan indicated that 80% of its elementary schools (attended by students aged 5 to 13) would be subjected to accommodation evaluations, including an "Accommodation Review", the outcome of which historically has been school closures/consolidations [@hwdsbLongTermFacilities2013; @craggsBoardEvaluatingFuture2013; @seasonsSchoolClosurePolicy2014].
+These decisions, underhandedly supported by the province, were in part motivated by a search for operational savings based on projected reductions in student population and underutilized school capacity [@craggsAxeCouldFall2012].
+Hamilton was particularly affected, as the HWDSB underwent an unprecedented number of Accommodation Reviews relative to other school boards [@seasons2014ARCCatalogues2014].
 
@@
 Along with insufficient funding to maintain schools in a state of good repair [@auditorgeneralofontarioAnnualReport20152015], the operational savings as a case for school closures is a long-standing critique of the school "funding formula" established by the conservative provincial government in 1998 [@mackenzieBlueprintFixOntario2018; @irwinSchoolClosure2012].
-With the backdrop of budget cuts, the funding formula pits financing of school maintenance and student needs against the financial compensation of staff and powers of local school boards as they depend on the province for the allocation of resources [@mackenzieBlueprintFixOntario2018].
+Within a backdrop of budget cuts, the funding formula pits the financing of school maintenance and student needs against the financial compensation of staff and the powers of local school boards, which depend on the province for the allocation of resources [@mackenzieBlueprintFixOntario2018].
 
 In the wave of school accommodation reviews between 2013 and 2016, `r ALL_SCHOOLS_Elem %>% filter(Year == "2011" & Level == "Elementary" & Status == "Removed") %>% pull(SchoolID) %>% length()` elementary schools closed in Hamilton.
-These closures represented a `r ((ALL_SCHOOLS_Elem %>% filter(Year != "2016") %>% pull(SchoolID) %>% length() - ALL_SCHOOLS_Elem %>% filter(Year != "2011") %>% pull(SchoolID) %>% length())/ALL_SCHOOLS_Elem %>% filter(Year != "2016" & Level == "Elementary") %>% pull(SchoolID) %>% length()) %>% scales::percent()` and `r  (( ((avail_2011 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2011)) |> select(OTGC) |> sum())) - (avail_2016 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2016)) |> select(OTGC) |> sum())) / (avail_2016 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2016)) |> select(OTGC) |> sum()) ) |> scales::percent()` decline in elementary school locations and capacity between 2011 (before the school closures) and 2016, despite the fact that the overall student population in the city did not decline, and in fact increased slightly between 2011 (`r avail_2011$kid_popadj %>% sum(na.rm=T) %>% prettyNum(big.mark=",")` students) and 2016 (`r avail_2016$kid_popadj %>% sum(na.rm=T) %>% prettyNum(big.mark=",")` students).
-
-While accommodation reviews were undertaken to reduce operational costs, how might the closure of schools impact local communities?
+These closures represented a `r ((ALL_SCHOOLS_Elem %>% filter(Year != "2016") %>% pull(SchoolID) %>% length() - ALL_SCHOOLS_Elem %>% filter(Year != "2011") %>% pull(SchoolID) %>% length())/ALL_SCHOOLS_Elem %>% filter(Year != "2016" & Level == "Elementary") %>% pull(SchoolID) %>% length()) %>% scales::percent()` and `r  (( ((avail_2011 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2011)) |> select(OTGC) |> sum())) - (avail_2016 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2016)) |> select(OTGC) |> sum())) / (avail_2016 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2016)) |> select(OTGC) |> sum()) ) |> scales::percent()` decline in elementary school locations and capacity between 2011 (before the school closures) and 2016, despite the elementary-aged student population in the city slightly increasing between 2011 and 2016( from `r avail_2011$kid_popadj %>% sum(na.rm=T) %>% prettyNum(big.mark=",")` students to `r avail_2016$kid_popadj %>% sum(na.rm=T) %>% prettyNum(big.mark=",")` students).
+
+While accommodation reviews were undertaken to reduce operational costs, how might have the closure of schools impacted local communities?
 In this paper, we hypothesize that the school consolidation policy reduced accessibility to schools for many students.
@@
 Furthermore, to increase the diffusion of this analysis, all work is transparent and reproducible, following best practices in the spatial sciences [@brunsdonOpeningPractice2021; @paezOpenSpatial2021c].
-Accordingly, all associated code and data will be publicly available within the lead author's [GitHub repository](https://github.com/soukhova)
+Accordingly, all associated code and data will be publicly available within the lead author's [GitHub repository](https://github.com/soukhova/School-closures-accessibility-impacts)
 
@@
 The difference between the two scenarios is quantified in terms of accessibility, emissions and equity implications and discussed in context with the school consolidation policy.
-Data were retrieved at different levels of spatial granularity from the following sources.
+Data was retrieved at different levels of spatial granularity from the following sources.
 
@@
 First, schools and associated school catchments were retrieved from the City's school boards.
-In Hamilton, the majority of student-aged population attends a school in one of the two English public school boards.
+In Hamilton, the majority of student-aged population attends a school in one of the two English public school boards [@faoOntarioSchoolBoards2023].
 One board is referred to as *public* (i.e., Hamilton Wentworth District School Board (HWDSB)) in our work and the other is referred to as *public-catholic* (Hamilton Wentworth Catholic District School Board (HWCDSB)).
@@
 As such, elementary includes middle schools that only instruct grades 6 to 8 (*MidElem*), primary schools that only instruct kindergarten to grade 5 or 6 (*JrElem*), and all grade elementary schools (*Elem*) which instruct all grades from kindergarten to grade 8.
-The maximum number of students that may be enrolled in a school is calculated by the provincial Ministry of Education; this value is referred to as the on-the-ground capacity (*OTGC*), the number of school seats available in the school beyond which it will be over-capacity, and is assigned to each school.
+The number of 'seats' available in the school beyond or below which it will be over/under- capacity is calculated by the provincial Ministry of Education; this value is referred to as the on-the-ground capacity (*OTGC*), and is assigned to each school.
 The current/historic OTGCs are not publicly available [@ontarioSchoolFacilityInventory2017].
 However, for schools that underwent the full Accommodation Review process, OTGCs for a given year could be obtained from publicly available School Information Profiles.
-A regression model was used to estimate the OTGCs for schools that lacked capacity information.
+A regression model was used to estimate the OTGCs for schools that lacked OTGC information.
 The independent variables in the model were 1) the school's building footprint (*F*) in units of ($m^2$); 2) the school's shortest Euclidean distance from the centroid of Hamilton's central business district area (*DistCBD*) defined in units of (m); and 3) coefficients associated with a binary variable indicating the type of school grade instruction (*MidElem*, *JrElem*, *Elem*).
-Of note, the models were also used to estimate OTGC of secondary schools (*Sec*) though not included in the scope of this work.
-These public institutions provide instruction for grades 9 to 12 (i.e., children aged 14 to 17).
 The building footprint from an archived spatial data set [@DMTI_Spatial_2015] and the 2016 footprints were retrieved from OSM [@OpenStreetMap_2021].
 Because the age of construction for different schools is not available, we used each school's distance from the centroid of the Downtown Business Improvement Area [@Hamilton_2014] as a proxy for school age as the oldest buildings are generally located closer to the city centre [@Merrall_2021].
+Of note, the models were also used to estimate OTGC of secondary schools (*Sec*) though not included in the scope of this work; these public institutions provide instruction for grades 9 to 12 (i.e., children aged 14 to 17).
 
@@
 
-knitr::include_graphics(paste0(here::here(),"/Figures/catchment_schools_plot.png"))
+grid.raster(readTIFF(paste0(here::here(),"/Figures/Fig1.tiff")))
 ```
@@
 
-In addition to OTGC for schools, statistics about the student population and proportion of low-income after tax (LIM-AT) housholds were retrieved from the 2011 and 2016 Canadian Census [@Statistics_Canada_2011; @Statistics_Canada_2016] using the {cancensus} `R` package [@von_Bergmann_Shkolnik_Jacobs_2021].
-The census releases population data by age group category in 5-year increments.
-The LIM-AT prevalence for households with children under 18 (there is no LIM-AT prevalence tabulated for households with exclusively elementary aged children) is retrained.
-LIM-AT is refers to the proportion of private households that are below the median after-tax income [@governmentofcanadaDictionaryCensusPopulation2017].
-These census variables are taken at the Dissemination Area (DA) level, which is the finest geographic unit for which census data are made publicly available.
+In addition to OTGC for schools, statistics about the student population and proportion of low-income after tax (LIM-AT) households were retrieved from the 2011 and 2016 Canadian Census [@Statistics_Canada_2011; @Statistics_Canada_2016] using the {cancensus} `R` package [@von_Bergmann_Shkolnik_Jacobs_2021].
+The census releases population data by age group category, so the population aged between 5-9 years and 10-14 years were retained.
+LIM-AT refers to the proportion of private households that are below the median after-tax income [@governmentofcanadaDictionaryCensusPopulation2017].
+The LIM-AT prevalence for households with children under 18 (there is no LIM-AT prevalence tabulated for households with exclusively elementary aged children) was retrained.
+Population and LIM-AT variables were taken at the Dissemination Area (DA) level, which is the finest geographic unit publicly available.
 DAs are designed by Statistics Canada with the aim of population uniformity hence DAs greatly vary in area but represent between approximately 400 and 700 (1st and 3rd quartile) in total population.
-The population aged 5 to 14 and the proportion of LIM-AT are visualised in the first and last row in @fig-Fig2.
+The population aged 5 to 14 and the proportion of LIM-AT are visualised in the first and last rows in @fig-Fig2.
 
@@
 #| out-width: "120%"
-
-knitr::include_graphics(paste0(here::here(),"/Figures/stud_pop_rate_LIM_plot.png"))
+#| fig-pos: "H"
+
+grid.raster(readTIFF(paste0(here::here(),"/Figures/Fig2.tiff")))
 ```
@@
 
-It is assumed that children can go to any elementary school, but there is a preference for facilities that are closer.
+It is assumed that children can go to any elementary school, but there is a preference for facilities that are more proximate.
 Based on this assumption, the fourth source of data is information regarding the mode-used and origin-destination locations of home-to-school trips from the 2011 and 2016 TTS [@data_management_group_tts_2018].
-The TTS is a travel survey conducted in the Greater Golden Horseshoe Area in Ontario every 5 years.
+The TTS is a travel survey conducted in the Greater Golden Horseshoe Area in Ontario every five years.
 With a target 5% sampling rate, the survey is expanded to be representative at the traffic analysis zone (TAZ) level of geography.
 TAZ are spatial units created for the purpose of the TTS and pulled from the R data package {TTS2016R} [@soukhovTTS2016RDataSet2023]; a few DAs typically nest within each TAZ.
-The trip-level travel data extract for this paper represent `r od_trips_perc_2011$TAZOrig_car_trips |> sum() |> prettyNum(big.mark = ",")` and `r od_trips_perc_2016$TAZOrig_car_trips |> sum() |> prettyNum(big.mark = ",")` motorized trips (mode used includes private care passenger, school bus, taxi, and transit) and `r od_trips_perc_2011$TAZOrig_walk_trips |> sum() |> prettyNum(big.mark = ",")` and `r od_trips_perc_2016$TAZOrig_walk_trips |> sum() |> prettyNum(big.mark = ",")` non-motorized trips (mode used includes walk and cycling) for 5 to 13 year olds from home to school in 2011 and 2016 respectively.
-The travel times for motorized and non-motorized travel are estimated using the {r5r} [@Pereira2021r5r] package, which interfaces with the java-based R5 routing engine developed separately by Conveyal [@conveyalConveyalR5Routing2022].
-Motorized and non-motorized travel times are estimated for all TAZ centroids to all TAZ centroids and these estimated travel times are tied to the associated OD trips.
+The trip-level travel data extracted for this paper represent `r od_trips_perc_2011$TAZOrig_car_trips |> sum() |> prettyNum(big.mark = ",")` and `r od_trips_perc_2016$TAZOrig_car_trips |> sum() |> prettyNum(big.mark = ",")` motorized trips (mode used includes private car passenger, school bus, taxi, and transit) and `r od_trips_perc_2011$TAZOrig_walk_trips |> sum() |> prettyNum(big.mark = ",")` and `r od_trips_perc_2016$TAZOrig_walk_trips |> sum() |> prettyNum(big.mark = ",")` non-motorized trips (mode used includes walk and cycling) for 5 to 13 year olds from home-to-school in 2011 and 2016 respectively.
+The travel times for motorized and non-motorized travel are estimated using {r5r} [@Pereira2021r5r], which interfaces with the java-based R5 routing engine developed separately by Conveyal [@conveyalConveyalR5Routing2022].
+As travel times are not associated with the TTS, motorized and non-motorized travel times are estimated for all TAZ centroids to all TAZ centroids.
+We estimate motorized travel to be by car and non-motorized travel to be by walk as those are the most used modes in their respective categories.
 The travel times and the intensity of modal home-to-school flows are visualised in @fig-Fig3 along with the boundaries of the TAZ and DAs.
-As can be seen in @fig-Fig3, not all TAZs capture an elementary school trip by both modes.
-For this reason, the proportion of modal-split is aggregated at the community level.
-Hamilton (Ancaster `r community_car_perc_2011$TAZOrig_car_perc[1] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[1] |> percent()`, Dundas `r community_car_perc_2011$TAZOrig_car_perc[2] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[2] |> percent()`, Flamborough `r community_car_perc_2011$TAZOrig_car_perc[3] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[3] |> percent()`, Glanbrook `r community_car_perc_2011$TAZOrig_car_perc[4] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[4] |> percent()`, Hamilton Central `r community_car_perc_2011$TAZOrig_car_perc[5] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[5] |> percent()`, and Stoney Creek `r community_car_perc_2011$TAZOrig_car_perc[6] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[6] |> percent()`).
+In @fig-Fig3, not all TAZs capture an elementary school trip by both modes.
+For this reason, the proportion of motorized modal share is aggregated at the community level.
+Hamilton (Ancaster: `r community_car_perc_2011$TAZOrig_car_perc[1] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[1] |> percent()`, Dundas: `r community_car_perc_2011$TAZOrig_car_perc[2] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[2] |> percent()`, Flamborough: `r community_car_perc_2011$TAZOrig_car_perc[3] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[3] |> percent()`, Glanbrook: `r community_car_perc_2011$TAZOrig_car_perc[4] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[4] |> percent()`, Hamilton Central: `r community_car_perc_2011$TAZOrig_car_perc[5] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[5] |> percent()`, and Stoney Creek: `r community_car_perc_2011$TAZOrig_car_perc[6] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[6] |> percent()`).
 <!--Unrelated: did you use information from TTS2016R?
@@
 #| out-width: "120%"
-
-knitr::include_graphics(paste0(here::here(),"/Figures/OD_flow_plot.png"))
-```
-
-The fifth source of data consists of estimated travel time matrices from all residential parcel locations to all elementary schools for both motorized (car) and non-motorized (walking) travel.
-The travel times are estimated using the {r5r} package [@Pereira2021r5r].
-For both modes, the inputs assume a maximum travel time threshold of 60 minutes and the free-flow OpenStreetMap road network of Hamilton (retrieved using Geofabrik [@geofabrikOntarioOpenStreetMapGeofabrik2022] and manually edited to only include the Hamilton boundary).
+#| fig-pos: "H"
+
+grid.raster(readTIFF(paste0(here::here(),"/Figures/Fig3.tiff")))
+```
+
+The fifth source of data consists of estimated travel time matrices from all residential parcel locations to all elementary schools for both motorized and non-motorized travel.
+The travel times are also estimated using the {r5r} package [@Pereira2021r5r], with car mode and walking representing motorized and non-motorized travel respectively.
+For both motorized and non-motorized modes, the inputs assume a maximum travel time threshold of 60 minutes and the free-flow OpenStreetMap road network of Hamilton (retrieved using Geofabrik [@geofabrikOntarioOpenStreetMapGeofabrik2022] and manually edited to only include the network within Hamilton).
 Further, only walking trips shorter than 27 min minutes were retained.
-This travel time is approximately equivalent to 1.6 km which is the point at which within-catchment students qualify for transport provided by the school board [@HWDSB_2019].
-Estimated trip-weighted motorized travel times are on average 16.5 minutes in 2011 and 16.8 minutes in 2016 (summary statistics for both years are approximately min: 0.5, Q1: 11.0, Q2: 16.0, Q3: 21.0, max: 60.0).
-Estimated trip-weighted non-motorized travel times are on average 18.0 minutes in 2011 and 18.1 minutes in 2016 (summary statistics for both years are approximately min: 0.5, Q1: 13.0, Q2: 19.0, Q3: 24.0, max: 27.0).
+This travel time is approximately equivalent to 1.6 km which is the point within-catchment students qualify for transport provided by the school board [@HWDSB_2019].
+Estimated trip-weighted motorized travel times are on average 16.5 minutes in 2011 and 16.8 minutes in 2016 (summary statistics for both years are approximately; min: 0.5, Q1: 11.0, Q2: 16.0, Q3: 21.0, max: 60.0).
+Estimated trip-weighted non-motorized travel times are on average 18.0 minutes in 2011 and 18.1 minutes in 2016 (summary statistics for both years are approximately; min: 0.5, Q1: 13.0, Q2: 19.0, Q3: 24.0, max: 27.0).
 <!--to calc.
@@
 Analysis of the TLD is a useful technique to calibrate impedance functions as they are a representation of the propensity of travel by trip length [@horbachov_theoretical_2018; @batista_estimation_2019].
-The empirical TLD is used to fit a density distribution using maximum likelihood and moment matching techniques and the Nelder-Mead and Brent methods for direct optimization available within the {fitdistrplus} `R` package [@fitdistrplus_2015].
+The empirical TLD is used to fit a density distribution using maximum likelihood and moment matching techniques and the Nelder-Mead and Brent methods for direct optimization available within the {fitdistrplus} package [@fitdistrplus_2015].
 Based on goodness-of-fit criteria and diagnostics, the gamma and exponential distributions were selected for the motorized and non-motorized modal distributions respectively.
@@
 #| out-width: "100%"
-
-knitr::include_graphics(paste0(here::here(),"/Figures/TLD_empirical_wtheo_curve_plot.png"))
-```
-
-The sixth and final data source relates to motorized GHG well-to-wheel estimations.
-It assumed that all non-motorized trips produce no emissions and those that are motorized take the average emission factor of 0.1 kg CO2e per minute travel.
-This emission factor is assumed based on an average speed of 40 km/h, the 2,380 g $CO_{2}e$ produced when 1 L of gasoline is combusted, and an internal combustion engine passenger vehicle fuel efficiency of 7 g/100km as discussed in @soukhovOccupancyGHGEmissions2022.
+#| fig-pos: "H"
+
+grid.raster(readTIFF(paste0(here::here(),"/Figures/Fig4.tiff")))
+```
+
+The sixth and final data source relates to motorized GHG estimations.
+It assumed that all non-motorized trips produce no emissions and those that are motorized take the average emission factor of 0.1 kg $CO_{2}e$  per minute travel.
+This emission factor is assumed based on an average speed of 40 km/h, the 2,380 g $CO_{2}e$ produced when 1 L of gasoline (well-to-wheel) is combusted, and an internal combustion engine passenger vehicle fuel efficiency of 7 g/100km as discussed in @soukhovOccupancyGHGEmissions2022.
 
@@
 Accessibility is defined as the "potential for spatial interaction".
-It is classically presented as the gravity-based measure popularized by @hansenHowAccessibilityShapes1959 and takes the following general multimodal formulation (Equation (\ref{eq:multimodal-conventional-accessibility})):
+It is classically presented as the gravity-based measure popularized by @hansenHowAccessibilityShapes1959.
+It takes the following general multimodal formulation (Equation (\ref{eq:multimodal-conventional-accessibility})):
 
@@
 
-It should be noted that, when summed over all spatial units in the region, the population-based allocation factors $F^{pm}_{i}$ always equal 1 (i.e., $\sum_{m}\sum_{i} F^{pm}_{i}= 1$), likewise for impedance-based allocation factors $F^{cm}_{i}$ (i.e., $\sum_{m}\sum_{i} F^{cm}_{ij} = 1$).
+It should be noted that, when summed over all spatial units in the region, the population-based allocation factors $F^{pm}_{i}$ always equal 1 ($\sum_{m}\sum_{i} F^{pm}_{i}= 1$), likewise for impedance-based allocation factors $F^{cm}_{i}$ ($\sum_{m}\sum_{i} F^{cm}_{ij} = 1$).
 
@@
 Each parcel is assigned a number of 'potential' motorized and non-motorized student aged population (from DA) and modal share (from TTS) to calculate a motorized and non-motorized population balancing factor $F_{i}^{pm}$.
-For each parcel the following is retained: the 2011 and 2016 census average elementary-aged student population per parcel (@fig-Fig2 second row) and the average motorized and non-motorized split as informed by the TTS aggregated by community (@fig-Fig3).
+For each parcel, the following is retained: the 2011 and 2016 census average elementary-aged student population per parcel (@fig-Fig2 second row) and the average motorized and non-motorized share as informed by the TTS aggregated by community (@fig-Fig3).
 
 Second, the school capacity $O_j$ is estimated and associated with each school (@fig-Fig1).
-All residential locations are assumed to be able to access all schools by either motorized or non-motorized mode.
+All residential locations are assumed to be able to access all schools by non-motorized and/or motorized mode.
 All origins (residential locations) can reach all schools by motorized mode, but few residential locations can reach schools within a 27 minute non-motorized trip.
@@
 # Results
-
-```{r}
-#| label: fig-Fig5
-#| fig-cap: "Multimodal spatial availability (purples) and spatial availability per mode-using student (diverging reds and greens) aggregated at the DA level for 2011 and 2016.
-Scales are represented in quartiles."
-#| echo: false
-#| out-width: "120%"
-#| 
-knitr::include_graphics(paste0(here::here(),"/Figures/Vim_vvim_20112016_plot.png"))
-```
-<!--check to make sure schools are correct-->
 
@@
 If our aim is to provide sufficient availability to schools for both motorized and non-motorized populations (e.g., greater than 1.0 $v_i^m$ school-seat availability per student), this rate is only achieved in the core of the city for motorized modes (greens and yellows).
-Notably, this rate only is available to non-motorized populations in DAs that are in less-densely populated rural areas that are school proximate and very few pockets of Hamilton Central.
+Notably, this rate is only available to non-motorized populations in DAs that are in less-densely populated rural areas that are school proximate and very few pockets of Hamilton Central.
 The majority of schools closed were within Hamilton Central, so non-motorized populations in DAs within Hamilton Central were the most negatively impacted by losses in availability while certain motorized populations captured these losses as availability gains.
+
+```{r}
+#| label: fig-Fig5
+#| fig-cap: "Multimodal spatial availability (purples) and spatial availability per mode-using student (diverging reds and greens) aggregated at the DA level for 2011 and 2016.
+Scales are represented in quartiles."
+#| echo: false
+#| out-width: "120%"
+#| fig-pos: "H"
+
+grid.raster(readTIFF(paste0(here::here(),"/Figures/Fig5.tiff")))
+```
+
+To highlight this point, @fig-Fig6 visualises the change in school spatial availability between 2016 and 2011 for both mode where $i$ are 2016 DAs.
+Though the 2011 and 2016 DAs system are very similar (see @fig-Fig2 and @fig-Fig3), they are not identical: `r (cen_data_2011_hamiltononly$GeoUID %in% cen_data_2016_hamiltononly$GeoUID |> length() - cen_data_2011_hamiltononly$GeoUID %in% cen_data_2016_hamiltononly$GeoUID |> sum()) + (cen_data_2016_hamiltononly$GeoUID %in% cen_data_2011_hamiltononly$GeoUID |> length() - cen_data_2016_hamiltononly$GeoUID %in% cen_data_2011_hamiltononly$GeoUID |> sum())` DAs changed in their spatial extent (i.e., added, removed, changed boundaries).
+The proportional allocation mechanism of spatial availability provides flexibility in aggregating parcel-level values at whichever spatial unit is meaningful.
+For this case, the change in $V_i^m$ and $v_i^m$ represented is the subtraction between the value aggregated at 2016 DA for the 2016 and 2011 scenario.
 
@@
 #| out-width: "120%"
-#| 
-knitr::include_graphics(paste0(here::here(),"/Figures/Vim_vvim_20112016_CHANGE_plot.png"))
-```
-
-To highlight this point, @fig-Fig6 visualises the change in school spatial availability between 2016 and 2011 for both mode where $i$ are 2016 DAs.
-Though the 2011 and 2016 DAs system are very similar (see @fig-Fig2 and @fig-Fig3), they are not identical: `r (cen_data_2011_hamiltononly$GeoUID %in% cen_data_2016_hamiltononly$GeoUID |> length() - cen_data_2011_hamiltononly$GeoUID %in% cen_data_2016_hamiltononly$GeoUID |> sum()) + (cen_data_2016_hamiltononly$GeoUID %in% cen_data_2011_hamiltononly$GeoUID |> length() - cen_data_2016_hamiltononly$GeoUID %in% cen_data_2011_hamiltononly$GeoUID |> sum())` DAs changed in their spatial extent (i.e., added, removed, changed shape).
-The proportional allocation mechanism of spatial availability provides flexibility in aggregating parcel-level values at whichever spatial unit is meaningful.
-For this case, the change in $V_i^m$ and $v_i^m$ represented is the subtraction between the value aggregated at 2016 DA for the 2016 and 2011 scenario.
+#| fig-pos: "H"
+
+grid.raster(readTIFF(paste0(here::here(),"/Figures/Fig6.tiff")))
+```
 
@@
 Smaller-area DAs (in Hamilton-Central) and larger-area DAs (in rural communities) proximate to closed schools present Q1 decreases (red).
-For $v_i^{motorized}$, the range offered by motorized mode (estimated by shortest-path car travel times) is city-wide, so since the majority of schools were centrally located, the decrease is concentrated within the downtown core.
+For $v_i^{motorized}$, the range offered by motorized mode (estimated by shortest-path car travel times) is city-wide, so since the majority of schools were centrally located, the decrease is concentrated within the downtown core of Hamilton Central.
 Some motorized-mode using populations in DAs do experience an increase in $V_i^m$ and $v_i^m$: these DAs are often relatively proximate to schools but more immediately outside walking range.
 Under competitive and constrained calculations, these DAs benefit from the losses in spatial availability that are predominately experienced by the non-motorized populations within Hamilton Central.
-
-```{r}
-#| label: fig-Fig7
-#| fig-cap: "Change in active travel time minutes (top plot) and home-to-school GHG emissions (bottom plot) between 2011 and 2016 scenarios.
-Change is the subtraction of parcel-level results between 2016 and 2011 scenarios summed for each DA in the 2016 Canadian Census DA system.
-Scales are represented in quartiles, with Q4 rounded to 0."
-#| echo: false
-#| out-width: "120%"
-#| 
-knitr::include_graphics(paste0(here::here(),"/Figures/ttim_GHGim_20112016_CHANGE_plot.png"))
-```
 
@@
 Smaller decreases (and even a few increases) are apparent in a range further from closed schools, especially those also in proximity to the few opened/expanded schools.
-Intuitively, these trends mirror what is shown within the $V^{non-motorized}$ landscape in @fig-Fig6 (Pearson correlation coefficient of `r cor(SAvail_im_DA_changes |> filter(mode == "non-motorized") |> pull(c_total_tt), SAvail_im_DA_changes |> filter(mode == "non-motorized") |> pull(c_V_im)) |> round(digits=2)`) as travel time is an input into the cost balancing factor $F^{cm}_{ij}$.
+Intuitively, these trends mirror what is shown within the $V^{non-motorized}$ landscape in @fig-Fig6 (Pearson correlation coefficient of `r cor(SAvail_im_DA_changes |> filter(mode == "non-motorized") |> pull(c_total_tt), SAvail_im_DA_changes |> filter(mode == "non-motorized") |> pull(c_V_im)) |> round(digits=3)`) as travel time is an input into the cost balancing factor $F^{cm}_{ij}$.
 Similarly, the bottom plot (@fig-Fig7) visualises the change in estimated emissions from motorized travel minutes.
-The change in GHG emissions is also highly correlated with $V^{motorized}$ in @fig-Fig6 (Pearson correlation coefficient of `r cor(SAvail_im_DA_changes |> filter(mode == "motorized") |> pull(c_emissions_total_car_tt), SAvail_im_DA_changes |> filter(mode == "motorized") |> pull(c_V_im))`) as it is also a product of travel time.
+The change in GHG emissions is also highly correlated with $V^{motorized}$ in @fig-Fig6 (Pearson correlation coefficient of `r cor(SAvail_im_DA_changes |> filter(mode == "motorized") |> pull(c_emissions_total_car_tt), SAvail_im_DA_changes |> filter(mode == "motorized") |> pull(c_V_im)) |> round(digits=3)`) as it is also a product of travel time.
+
+```{r}
+#| label: fig-Fig7
+#| fig-cap: "Change in active travel time minutes (top plot) and home-to-school GHG emissions (bottom plot) between 2011 and 2016 scenarios.
+Change is the subtraction of parcel-level results between 2016 and 2011 scenarios summed for each DA in the 2016 Canadian Census DA system.
+Scales are represented in quartiles, with Q4 rounded to 0."
+#| echo: false
+#| out-width: "90%"
+#| fig-pos: "H"
+
+grid.raster(readTIFF(paste0(here::here(),"/Figures/Fig7.tiff")))
+```
 
@@
 The change in active travel minutes and GHG emissions highlights the negative externalities of _decreased_ non-motorized spatial availability.
-Normatively, our cities should be increasing spatial availability for non-motorized mode-users given the climate crisis.
-The wave of school closures in Hamilton increased motorized spatial availability and hence potential GHG emissions.
+Normatively, our cities should be increasing spatial availability for non-motorized mode users given the climate crisis.
+The wave of school closures in Hamilton increased motorized spatial availability and, hence, potential GHG emissions.
 The policy drastically reduced non-motorized spatial availability, and this is seen in the decrease in potential non-motorized travel minutes as well as increased GHG emissions in rural areas that did benefit from motorized spatial availability gains.
-The proportion of non-motorized mode use is low in both 2011 and 2016, the closure of schools that are more proximate to student-population density eliminates the potential of those trips ever becoming active.
-And who does the reduction of non-motorized spatial availability impact the most?
-A summary of spatial availability per Hamilton community and associated dimensions in 2016 along with the percentage change between 2011 and 2016 are summarized in @tbl-Tab1.
+The proportion of non-motorized mode use is low in both 2011 and 2016, however, the closure of schools that are more proximate to student-population density eliminates the potential of those trips ever becoming active.
+And who is most impacted by the reduction of non-motorized spatial availability?
+A summary of spatial availability per Hamilton community and associated dimensions in 2016, along with the percentage change between 2011 and 2016, is summarized in @tbl-Tab1.
+
+\pagebreak
 
@@
 #| tbl-cap: "Spatial availability and associated dimensions aggregated by communities"
-#| 
-tdata_hux <- as_hux(tdata) 
-tdata_hux[1,1:12] <- c("Hamilton C.", "", "Dundas" ,"", "Stoney Creek", "", "Ancaster", "", "Flamborough", "", "Glanbrook", "")
-# tdata_hux[1,1:12] <- c("HA", "", "DU" ,"", "SC", "", "AN", "", "FL", "", "GL", "")
-tdata_hux[2,1:12] <- c("", "% Δ", "" , "% Δ", "", "% Δ", "", "% Δ", "", "% Δ", "", "% Δ")
+#| out-width: "90%"
+
+tdata_hux <- as_hux(tdata) %>% set_escape_contents(FALSE)
+# tdata_hux[1,1:12] <- c("Hamilton C.", "", "Dundas" ,"", "Stoney Creek", "", "Ancaster", "", "Flamborough", "", "Glanbrook", "")
+# # tdata_hux[1,1:12] <- c("HA", "", "DU" ,"", "SC", "", "AN", "", "FL", "", "GL", "")
+# tdata_hux[2,1:12] <- c("", "% {\\Delta}", "" , "% {\\Delta}", "", "% {\\Delta}Δ", "", "% ${\\Delta}", "", "% {\\Delta}", "", "% {\\Delta}")
+
+tdata_hux <- set_contents(tdata_hux,
+                          row= c(1),
+                          col = c(1:12),
+                          value = c("Hamilton C.", "", "Dundas" ,"", "Stoney Creek", "", "Ancaster", "", "Flamborough", "", "Glanbrook", ""))
+tdata_hux <- set_contents(tdata_hux,
+                          row= c(2),
+                          col = c(1:12),
+                          value = c("", "% Δ", "", "% Δ", "", "% Δ", "", "% Δ",
+                                    "", "% Δ", "", "% Δ"))
 
@@
   set_caption("Spatial availability and associated dimensions aggregated by communities") |>
-  set_label("tbl-Tab1") %>% as_flextable()
+  set_label("tbl-Tab1") |> as_flextable()
 
@@
 The spatial availability and spatial availability per student are calculated at the parcel level and aggregated at the Canadian Census DA spatial unit and at the six Hamilton communities for comparison.
-Overall, schools closed and the number of school seats decreased so spatial availability declined overall: in this work, we also discuss where, for whom, and with what potential impacts.
+Schools closed and the number of school seats decreased so spatial availability declined overall: we discuss where, for whom, and with what potential impacts.
 
@@
 
-In addition to spatial availability impacts, associated mobility measures are also assessed.
-City-wide, home-to-school non-motorized minutes (1 day) reduced by `r SAvail_im_COMM_changes |> st_drop_geometry() |> filter(mode == "non-motorized") |> transmute(total_tt = total_tt_2016-total_tt_2011) |> sum() %>% abs() %>% prettyNum(big.mark=",", digits=1)` and GHG emissions associated with motorized home-to-school travel (1 day)  increased by `r SAvail_im_COMM_changes |> st_drop_geometry() |> filter(mode == "motorized") |> transmute(emissions = emissions_2016-emissions_2011) |> sum() %>% abs() %>% prettyNum(big.mark=",", digits=1)` kg $CO_2e$.
+In addition to spatial availability impacts, associated mobility measures are assessed.
+City-wide, home-to-school non-motorized minutes (for a one-way trip) reduced by `r SAvail_im_COMM_changes |> st_drop_geometry() |> filter(mode == "non-motorized") |> transmute(total_tt = total_tt_2016-total_tt_2011) |> sum() %>% abs() %>% prettyNum(big.mark=",", digits=1)` and GHG emissions associated with motorized home-to-school travel increased by `r SAvail_im_COMM_changes |> st_drop_geometry() |> filter(mode == "motorized") |> transmute(emissions = emissions_2016-emissions_2011) |> sum() %>% abs() %>% prettyNum(big.mark=",", digits=1)` kg $CO_2e$.
 The decision to consolidate schools, increase the length of motorized travel and decrease the potential for non-motorized impacts will have daily impacts on students and their families [@mandicAdolescentsPerceptionsWalking2022; @pabayoImportanceActiveTransportation2012], the broader community [@pietrabissaSchoolAccessCity2023;@MerrallWhatsASchool2024; @bittencourtEvaluatingAccessibilityAvailability2023] and the environment [@pantelakiImpactHomeschoolCommuting2024; @rongReviewResearchLowcarbon2022].
@@
 From the perspective of the provincial government of Ontario, the school closures which occurred in Hamilton resulted in operational savings-per-student.
-In this work, we demonstrate how these decisions resulted in both families and students paying a price through spatial availability, lost potential walk minutes, and increased GHG emissions.
- <!-- In the case of Hamilton Central, where households have the highest prevalence of LIM-AT in the city, those households suffer the most disproportionate decrease in motorized spatial availability and increase in percentage increase in emissions per capacity.
-As public schools should serve all students; the determination of spatial availability and spatial equity of impact, is critical to better inform decision-makers when evaluating school consolidation policies -->
+We demonstrate how these decisions resulted in both families and students paying a price through spatial availability, lost potential walk minutes, and increased GHG emissions.
 
@@
 At the core of the method is spatial availability [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024], a singly-constrained accessibility measure that can be calculated at the finest spatial resolution (in our paper, parcel-level) and aggregated at whichever spatial unit is most meaningful for interpretation.
-We urge policy makers to view the 'spatial availability' of public resources from a per-capita perspective, plan capacity and location of service provision with the cost of travel (and associated implications such as GHG emissions, active transportation mode, safety of active travel modes, etc.) that sufficiently serves the target population.
+We urge policy-makers to view the 'spatial availability' of public resources from a per-capita perspective, plan capacity and location of service provision with the cost of travel (and associated implications such as GHG emissions, active transportation mode, safety of active travel modes, etc.) that sufficiently serves the target population.
 In this work, we suggest and assume a benchmark of 1.0 school-seats per student which accounts for sufficiency.
@@
 
+\pagebreak
+
 # References
@@
 
+\pagebreak
+
 # Appendix
 
-\input{TabA1-OTGC-Reg.tex} 
+\input{TabA1.tex} 
 <!--made in the hamONdests repo-->

### 2024-05-09 paper/Manuscript.tex
Version submitted to journal

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
default
]{sn-jnl}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

%%%% Standard Packages

\usepackage{graphicx}%
\usepackage{multirow}%
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{amsthm}%
\usepackage{mathrsfs}%
\usepackage[title]{appendix}%
\usepackage{xcolor}%
\usepackage{textcomp}%
\usepackage{manyfoot}%
\usepackage{booktabs}%
\usepackage{algorithm}%
\usepackage{algorithmicx}%
\usepackage{algpseudocode}%
\usepackage{listings}%

%%%%

\raggedbottom
\usepackage{fontspec}
\usepackage{multirow}
\usepackage{multicol}
\usepackage{colortbl}
\usepackage{hhline}
\newlength\Oldarrayrulewidth
\newlength\Oldtabcolsep
\usepackage{longtable}
\usepackage{array}
\usepackage{hyperref}
\usepackage{float}
\usepackage{wrapfig}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={School closures and consolidations: the active travel and emission implications of reduced accessibility},
  pdfauthor={Anastasia Soukhov; Christopher D. Higgins; Antonio Páez; Moataz Mohamed},
  pdfkeywords={accessibility; spatial availability; carbon emissions;
journey-to-school; walkability; active transportation; school
consolidation policy; service provision thresholds},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title[School closures and consolidations: the active travel and
emission implications of reduced accessibility]{School closures and
consolidations: the active travel and emission implications of reduced
accessibility}

% author setup
\author*[1]{\fnm{Anastasia} \sur{Soukhov}}\email{soukhoa@mcmaster.ca}\author[2]{\fnm{Christopher D.} \sur{Higgins}}\email{cd.higgins@utoronto.ca}\author[1]{\fnm{Antonio} \sur{Páez}}\email{paezha@mcmaster.ca}\author[3]{\fnm{Moataz} \sur{Mohamed}}\email{mmohame@mcmaster.ca}
% affil setup
\affil[1]{, \orgaddress{\street{School of Earth, Environment and
Society, McMaster University, 1241 Main St.~West, Hamilton, ON, L8S 4K1,
Canada}}}
\affil[2]{, \orgaddress{\street{Department of Human Geography,
University of Toronto Scarborough, 1265 Military Trail, Toronto, ON, M1C
1A4, Canada}}}
\affil[3]{, \orgaddress{\street{Department of Civil Engineering,
McMaster University, 1241 Main St.~West, Hamilton, ON, L8S 4K1,
Canada}}}

% abstract

\abstract{Reducing emissions and increasing active travel are top
priorities for many communities. Often, though, these priorities are
confronted with pressures to reduce the fiscal burden of public
services, including schools. Public schools have been strained to do
more with less, and governments are opting to close under-capacity
schools and expand existing schools to accommodate lost capacity. The
objective of this study is to quantify the travel burdens of school
consolidation/closure policies. We do this through the case of Hamilton,
a mid-sized city in Ontario, Canada. Between the years 2011 and 2015,
7\% of the elementary schools closed in the city. We study how these
closures changed the accessibility landscape for the elementary school
aged population, as well as the active travel and emission implications.
In terms of methods, we use spatial availability, a singly-constrained
multimodal measure of spatial accessibility. Thanks to its proportional
allocation mechanism, spatial availability allows for an interpretable
comparison between `before' and `after' the implementation of school
closure/consolidation policy. Analysis is conducted at the parcel-level
with all school seats being eligible and travel behaviour based on
observed home-to-school patterns for motorized and non-motorized modes.
We report estimates of system-wide impacts that result from the
closures, such as potential growth in GHG emission and reduction of
walking minutes. Findings demonstrate that overall, spatial availability
decreased in the city. Though the school consolidation policy may have
resulted in cost-savings in operation and maintenance of facilities,
students, families, and communities pay the price in reduced
opportunities for physical activity and greater motorized
mobility-related pollution.}

% keywords
\keywords{accessibility; spatial availability; carbon emissions;
journey-to-school; walkability; active transportation; school
consolidation policy; service provision thresholds}

\begin{document}
\maketitle

\section{Introduction}\label{introduction}

Communities are by nature in flux: economic and demographic shifts,
rural and urban configurations, and political pressures all combine to
produce change. These forces spur the re-organization and consolidation
of public facilities in communities around the world where changes to
service provision may improve the quality of service for some but not
for all (Christiaanse 2020; Rosik, Puławska-Obiedowska, and Goliszek
2021). While change is natural, oftentimes top-down consolidation
decisions are driven by operational cost-savings that do not take into
consideration the full extent of losses in benefits.

From this perspective, school closure policies have been pursued by
governments to optimize per-student operational cost-efficiency (Rong et
al. 2022; Dai et al. 2019) but other implications, such as the impact of
closures on property values (Merrall, Higgins, and Páez 2024), their
effects on the social infrastructure of communities (Butler, Kane, and
Cooligan 2019; Irwin and Seasons 2012), and particularly their travel
implications have often been overlooked by decision makers (Bierbaum,
Karner, and Barajas 2021; J. Lee and Lubienski 2017). This seems a
mistake: journey-to-school trips are tremendously important, and travel
for this purpose is currently the largest contributor of carbon
emissions from personal travel after the journey-to-work (Rong et al.
2022; Pantelaki, Claudia Caspani, and Maggi 2024), where it could be
instead an excellent opportunity for physical activity (Desjardins et
al. 2024). Alas, in many communities globally, schools are closing
(Sageman 2022). For instance, 65\% of comprehensive schools (e.g.~grades
1 through 9) in rural Finland since 1990 closed (Autti and
Hyry-Beihammer 2014), 10\% of elementary schools (kindergarten through
grade 8) in Chicago, USA in 2013 (J. Lee and Lubienski 2017), and 7\% of
primary and lower secondary schools (kindergarten through grade 9) were
closed in Denmark between 2010-2011 (Beuchert et al. 2018). School
closures often increase the distance to school for many students and
thus tend to induce motorized trips (Rong et al. 2022).

We focus on the transportation implications of school closure and
consolidation policies. While some research has considered the ``steps
in reserve'', that is, the potential for active travel hidden behind
motorized trips (Morency, Roorda, and Demers 2009; Morency, Demers, and
Lapierre 2007), school closures potentially make actual steps vanish,
sent to the reserve for the foreseeable future. In exchange for less
physical activity, communities get more motorized travel, with the
concomitant costs and burdens, including emissions. Quantification of
losses in potential active travel and associated growth in carbon
emissions is scarce in the literature, and the focus of this paper. In
addition to its empirical contributions, this work makes use of a novel
multimodal and singly-constrained accessibility measure, called spatial
availability (Soukhov et al. 2023; Soukhov et al. 2024), which we use to
estimate the changes in accessibility and travel after a recent wave of
school closures/consolidations in Hamilton, Ontario.

The results of the research indicate that, overall, school seats became
less spatially available in the city. This change happened unevenly,
with some communities being disproportionately affected. Though the
school consolidation policy pursued by the province and implemented by
the local school board may have resulted in cost-savings in operation
and maintenance of facilities, families, students, and communities now
pay the price in terms of lost opportunities for healthier active
travel, as well as increased emissions from motorized travel.

\section{Background}\label{background}

Hamilton is a mid-size city (\textasciitilde570,000 pop) with rural,
suburban and urban characteristics in the province of Ontario, Canada
(Government of Canada 2022). In 2013, the Hamilton-Wentworth District
School Board (HWDSB), the largest public English school board in the
city, released its Long Term Facilities Master Plan (HWDSB 2013). To
ensure ``equitable, affordable and sustainable learning facilities'',
the Master Plan indicated that 80\% of its elementary schools (attended
by students aged 5 to 13) would be subjected to accommodation
evaluations, including an ``Accommodation Review'', the outcome of which
historically has been school closures/consolidations (HWDSB 2013; Craggs
2013; Seasons 2014b). These decisions, underhandedly supported by the
province, were in part motivated by a search for operational savings
based on projected reductions in student population and underutilized
school capacity (Craggs 2012). Hamilton was particularly affected, as
the HWDSB underwent an unprecedented number of Accommodation Reviews
relative to other school boards (Seasons 2014a).

Though the Accommodation Review guidelines outline a community-focused
process to assess the value that a school provides to students, the
community, and the local economy (Seasons 2014b; Ministry of Education
2006), in practice residents felt that a wholesome breakdown of costs to
repair/maintain schools was not made consistently available (Kleinhuis
2013). Further, others pointed to flawed incentive structures motivating
the HWDSB: ``By closing three schools the HWDSB would have a strong case
with the Ministry of Education to receive full funding for a new
school'' (Kleinhuis 2013). Along with insufficient funding to maintain
schools in a state of good repair (Auditor General of Ontario 2015), the
operational savings as a case for school closures is a long-standing
critique of the school ``funding formula'' established by the
conservative provincial government in 1998 (Mackenzie 2018; Irwin and
Seasons 2012). Within a backdrop of budget cuts, the funding formula
pits the financing of school maintenance and student needs against the
financial compensation of staff and the powers of local school boards,
which depend on the province for the allocation of resources (Mackenzie
2018).

In the wave of school accommodation reviews between 2013 and 2016, 12
elementary schools closed in Hamilton. These closures represented a 7\%
and 4\% decline in elementary school locations and capacity between 2011
(before the school closures) and 2016, despite the elementary-aged
student population in the city slightly increasing between 2011 and
2016( from 58,265 students to 58,865 students).

While accommodation reviews were undertaken to reduce operational costs,
how might have the closure of schools impacted local communities? In
this paper, we hypothesize that the school consolidation policy reduced
accessibility to schools for many students. By extension, this loss in
access and the consequent displacement of active trips to school
increased transportation-related emissions. To test these hypotheses,
the paper examines the following: how does the (1) spatial availability,
(2) loss in walk minutes and (3) associated carbon emissions change for
home-to-school trips after a city closes and consolidates schools?
Further, this work examines (4) what communities, based on low-income
prevalence, experienced the greatest losses as a result of this policy.

Spatial availability, a singly-constrained measure of spatial
accessibility (Soukhov et al. 2023; Soukhov et al. 2024), is used to
quantify accessibility before- and after- school closures, and the
associated impacts for the city of Hamilton and the elementary
school-aged student population as a result of a policy with narrowly
defined benefits. The methods used in this paper can be applied by
decision-makers to evaluate the spatial and transport-related impacts of
public service provision policy. Furthermore, to increase the diffusion
of this analysis, all work is transparent and reproducible, following
best practices in the spatial sciences (Brunsdon and Comber 2021;
Antonio Páez 2021). Accordingly, all associated code and data will be
publicly available within the lead author's
\href{https://github.com/soukhova/School-closures-accessibility-impacts}{GitHub
repository}

\section{Data}\label{data}

Our focus is to evaluate the impact of school consolidations between
2011 and 2016 on multimodal school spatial availability and
travel-related GHG emission in Hamilton. To facilitate comparison, two
scenarios are generated. Each scenario reflects the conditions in 2011
and 2016 respectively, assuming the school configurations in each year,
as well as average student population and low-income prevalence of
households, residential parcel locations, and estimated modal travel
times. The difference between the two scenarios is quantified in terms
of accessibility, emissions and equity implications and discussed in
context with the school consolidation policy. Data was retrieved at
different levels of spatial granularity from the following sources.

\subsection{Schools}\label{schools}

First, schools and associated school catchments were retrieved from the
City's school boards. In Hamilton, the majority of student-aged
population attends a school in one of the two English public school
boards (FAO 2023). One board is referred to as \emph{public} (i.e.,
Hamilton Wentworth District School Board (HWDSB)) in our work and the
other is referred to as \emph{public-catholic} (Hamilton Wentworth
Catholic District School Board (HWCDSB)). These catchments indicate what
residential property is assigned to what school, by default. Families
can decide to attend schools out of catchment; in fact, 21\%-23\% of
motorized school trips are out of catchment according to the regional
travel survey, the Transportation Tomorrow Survey (TTS) in 2011 and 2016
(Data Management Group 2018). The elementary school locations and
catchments were provided by the HWDSB and HWCDSB for the 2010-2011 and
2015-2016 academic years, and are shown in Figure~\ref{fig-Fig1}.

Formally, elementary schools are defined as schools that provide
instruction to any combination of grades between kindergarten to grade 8
(i.e.~typically children aged 5 to 14). As such, elementary includes
middle schools that only instruct grades 6 to 8 (\emph{MidElem}),
primary schools that only instruct kindergarten to grade 5 or 6
(\emph{JrElem}), and all grade elementary schools (\emph{Elem}) which
instruct all grades from kindergarten to grade 8. The number of `seats'
available in the school beyond or below which it will be over/under-
capacity is calculated by the provincial Ministry of Education; this
value is referred to as the on-the-ground capacity (\emph{OTGC}), and is
assigned to each school. The current/historic OTGCs are not publicly
available (Ontario 2017). However, for schools that underwent the full
Accommodation Review process, OTGCs for a given year could be obtained
from publicly available School Information Profiles. A regression model
was used to estimate the OTGCs for schools that lacked OTGC information.
The independent variables in the model were 1) the school's building
footprint (\emph{F}) in units of (\(m^2\)); 2) the school's shortest
Euclidean distance from the centroid of Hamilton's central business
district area (\emph{DistCBD}) defined in units of (m); and 3)
coefficients associated with a binary variable indicating the type of
school grade instruction (\emph{MidElem}, \emph{JrElem}, \emph{Elem}).
The building footprint from an archived spatial data set (Spatial 2015)
and the 2016 footprints were retrieved from OSM (OpenStreetMap 2021).
Because the age of construction for different schools is not available,
we used each school's distance from the centroid of the Downtown
Business Improvement Area (Hamilton 2014) as a proxy for school age as
the oldest buildings are generally located closer to the city centre
(Merrall 2021). Of note, the models were also used to estimate OTGC of
secondary schools (\emph{Sec}) though not included in the scope of this
work; these public institutions provide instruction for grades 9 to 12
(i.e., children aged 14 to 17).

Equations (\ref{eq:OTGC-public}) and (\ref{eq:OTGC-catholic}) describe
the estimated OTGC for public schools and catholic schools respectively.
School status and OTGC summed by catchment is also visualised in
Figure~\ref{fig-Fig1}. As seen in Table \ref{TabA1-OTGC} in the
Appendix, the coefficient of determination for these models is
\(R^2 = 0.999\) and \(R^2 = 0.998\) for the Public \(OTGC_{Public}\) and
Catholic \(OTGC_{Catholic}\) Boards respectively, and the residual
standard deviation is \(\sigma = 0.212\) and \(sigma = 0.331\).

\begin{equation}
\label{eq:OTGC-public}
OTGC_{Public} = F^{0.346}-e^{0.00003*DistCBD}+e^{3.123*JrElem}+e^{3.752*Elem}+ e^{3.068*MidElem} 
\end{equation}

\begin{equation}
\label{eq:OTGC-catholic}
OTGC_{Catholic} =F^{0.471}-e^{0.00003*DistCBD}+e^{2.333*Elem}
\end{equation}

\begin{figure}

\centering{

\includegraphics[width=1.2\textwidth,height=\textheight]{Manuscript_files/figure-pdf/fig-Fig1-1.pdf}

}

\caption{\label{fig-Fig1}The school catchments, estimated On The Ground
Capacity (OTGC) of schools summed per catchment, and school status for
both public and public catholic school boards for year 2011 and 2016.
OTGC scale is presented by quartiles.}

\end{figure}%

\subsection{Students, residential locations and low-income
prevelance}\label{students-residential-locations-and-low-income-prevelance}

In addition to OTGC for schools, statistics about the student population
and proportion of low-income after tax (LIM-AT) households were
retrieved from the 2011 and 2016 Canadian Census (Canada 2011, 2016)
using the \{cancensus\} \texttt{R} package (Bergmann, Shkolnik, and
Jacobs 2021). The census releases population data by age group category,
so the population aged between 5-9 years and 10-14 years were retained.
LIM-AT refers to the proportion of private households that are below the
median after-tax income (Government of Canada 2017). The LIM-AT
prevalence for households with children under 18 (there is no LIM-AT
prevalence tabulated for households with exclusively elementary aged
children) was retrained. Population and LIM-AT variables were taken at
the Dissemination Area (DA) level, which is the finest geographic unit
publicly available. DAs are designed by Statistics Canada with the aim
of population uniformity hence DAs greatly vary in area but represent
between approximately 400 and 700 (1st and 3rd quartile) in total
population. The population aged 5 to 14 and the proportion of LIM-AT are
visualised in the first and last rows in Figure~\ref{fig-Fig2}.

The third source of data consists of centroids of residential parcels in
2011 and 2016 that are used to represent residential locations (Teranet
2009). There are 134,340 and 139,467 residential locations in 2011 and
2016 respectively, representing 203,806 and 211,596 dwellings (according
to the Canadian Census (Canada 2011, 2016)). As the number of children
per parcel are publicly unavailable, the average number of 5 to 14 year
olds for each DA is divided by the number of residential households in
that DA. Then the potential number of children at each dwelling is
calculated by multiplying this DA rate by the number of dwellings in
each parcel. Due to the proprietary nature of the parcel data, the
middle row in Figure~\ref{fig-Fig2} visualises an aggregation of the
information: the average rate of 5-14 year olds per residential
household at the DA level.

\begin{figure}[H]

\centering{

\includegraphics[width=1.2\textwidth,height=\textheight]{Manuscript_files/figure-pdf/fig-Fig2-1.pdf}

}

\caption{\label{fig-Fig2}The magnitude (top row) and rate per
residential unit (middle row) of elementary aged student population per
DA in 2011 and 2016. Bottom row: the proportion of low-income households
(prevelance of low-income after-tax, LIM-AT, measured by the 2011 and
2016 Canadian Census) with dependents under 18 years old shown. All
scales are represented in quartiles.}

\end{figure}%

\subsection{Estimated travel times and
emissions}\label{estimated-travel-times-and-emissions}

It is assumed that children can go to any elementary school, but there
is a preference for facilities that are more proximate. Based on this
assumption, the fourth source of data is information regarding the
mode-used and origin-destination locations of home-to-school trips from
the 2011 and 2016 TTS (Data Management Group 2018). The TTS is a travel
survey conducted in the Greater Golden Horseshoe Area in Ontario every
five years. With a target 5\% sampling rate, the survey is expanded to
be representative at the traffic analysis zone (TAZ) level of geography.
TAZ are spatial units created for the purpose of the TTS and pulled from
the R data package \{TTS2016R\} (Soukhov and Páez 2023); a few DAs
typically nest within each TAZ. The trip-level travel data extracted for
this paper represent 13,715 and 12,878 motorized trips (mode used
includes private car passenger, school bus, taxi, and transit) and 2,736
and 1,430 non-motorized trips (mode used includes walk and cycling) for
5 to 13 year olds from home-to-school in 2011 and 2016 respectively. The
travel times for motorized and non-motorized travel are estimated using
\{r5r\} (Pereira et al. 2021), which interfaces with the java-based R5
routing engine developed separately by Conveyal (Conveyal {[}2015{]}
2022). As travel times are not associated with the TTS, motorized and
non-motorized travel times are estimated for all TAZ centroids to all
TAZ centroids. We estimate motorized travel to be by car and
non-motorized travel to be by walk as those are the most used modes in
their respective categories. The travel times and the intensity of modal
home-to-school flows are visualised in Figure~\ref{fig-Fig3} along with
the boundaries of the TAZ and DAs. In Figure~\ref{fig-Fig3}, not all
TAZs capture an elementary school trip by both modes. For this reason,
the proportion of motorized modal share is aggregated at the community
level. Hamilton (Ancaster: 85\%-95\%, Dundas: 89\%-100\%, Flamborough:
91\%-97\%, Glanbrook: 92\%-95\%, Hamilton Central: 80\%-87\%, and Stoney
Creek: 82\%-88\%).

\begin{figure}[H]

\centering{

\includegraphics[width=1.2\textwidth,height=\textheight]{Manuscript_files/figure-pdf/fig-Fig3-1.pdf}

}

\caption{\label{fig-Fig3}The origin-destination flows from home-based
motorized and non-motorized school trips for students between 5-13 years
old as retrieved from TTS 2011 and 2016. Flows are mapped atop the
proportion of TAZ non/motorized modal share from the TTS 2011 and 2016,
DA boundaries, and Hamilton community boundaries.}

\end{figure}%

The fifth source of data consists of estimated travel time matrices from
all residential parcel locations to all elementary schools for both
motorized and non-motorized travel. The travel times are also estimated
using the \{r5r\} package (Pereira et al. 2021), with car mode and
walking representing motorized and non-motorized travel respectively.
For both motorized and non-motorized modes, the inputs assume a maximum
travel time threshold of 60 minutes and the free-flow OpenStreetMap road
network of Hamilton (retrieved using Geofabrik (Geofabrik 2022) and
manually edited to only include the network within Hamilton). Further,
only walking trips shorter than 27 min minutes were retained. This
travel time is approximately equivalent to 1.6 km which is the point
within-catchment students qualify for transport provided by the school
board (HWDSB 2019). Estimated trip-weighted motorized travel times are
on average 16.5 minutes in 2011 and 16.8 minutes in 2016 (summary
statistics for both years are approximately; min: 0.5, Q1: 11.0, Q2:
16.0, Q3: 21.0, max: 60.0). Estimated trip-weighted non-motorized travel
times are on average 18.0 minutes in 2011 and 18.1 minutes in 2016
(summary statistics for both years are approximately; min: 0.5, Q1:
13.0, Q2: 19.0, Q3: 24.0, max: 27.0).

These travel time matrices along with the origin-destination flows were
used to create modal trip length distribution (TLD) functions. Analysis
of the TLD is a useful technique to calibrate impedance functions as
they are a representation of the propensity of travel by trip length
(Horbachov and Svichynskyi 2018; Batista, Leclercq, and Geroliminis
2019). The empirical TLD is used to fit a density distribution using
maximum likelihood and moment matching techniques and the Nelder-Mead
and Brent methods for direct optimization available within the
\{fitdistrplus\} package (Delignette-Muller and Dutang 2015). Based on
goodness-of-fit criteria and diagnostics, the gamma and exponential
distributions were selected for the motorized and non-motorized modal
distributions respectively. The gamma distribution is defined by the
shape (\(\alpha\)) parameter of 1.939 (2011) and 2.046 (2016) and the
rate (\(\beta\)) of 0.233 (2011) and 0.236 (2016). The exponential
distribution is defined by the rate (\(\beta\)) parameter of 0.092
(2011) and 0.1 (2016). The TLDs for the motorized and non-motorized
trips are shown in Figure~\ref{fig-Fig4}. For reference, the gamma
distribution and the exponential distribution function are displayed in
Equations (\ref{eq:exp-dist}) and (\ref{eq:gamma-dist}) where \(x\) is
\(c_{ij}\):

\begin{equation}
\label{eq:exp-dist}
f(x) = \beta e ^{-\beta x}
\end{equation}

\begin{equation}
\label{eq:gamma-dist}
\begin{aligned}
f(x, \alpha, \beta) = \frac {x^{\alpha-1}e^{-\frac{x}{\beta}}}{ \beta^{\alpha}\Gamma(\alpha)} \quad \text{for } 0 \leq x \leq \infty \\
\Gamma(\alpha) =  \int_{0}^{\infty} x^{\alpha-1}e^{-x} \,dx
\end{aligned}
\end{equation}

\begin{figure}[H]

\centering{

\includegraphics[width=1\textwidth,height=\textheight]{Manuscript_files/figure-pdf/fig-Fig4-1.pdf}

}

\caption{\label{fig-Fig4}Empirical (grey bars) and theoretical (blue
curves) motorized and non-motorized impedance functions. Motorized
theoretical impedance function is based on the gamma distribution
function and non-motorized on the expoential distribution function}

\end{figure}%

The sixth and final data source relates to motorized GHG estimations. It
assumed that all non-motorized trips produce no emissions and those that
are motorized take the average emission factor of 0.1 kg \(CO_{2}e\) per
minute travel. This emission factor is assumed based on an average speed
of 40 km/h, the 2,380 g \(CO_{2}e\) produced when 1 L of gasoline
(well-to-wheel) is combusted, and an internal combustion engine
passenger vehicle fuel efficiency of 7 g/100km as discussed in Soukhov
and Mohamed (2022).

\section{Methods: Multimodal spatial
availability}\label{methods-multimodal-spatial-availability}

Accessibility is defined as the ``potential for spatial interaction''.
It is classically presented as the gravity-based measure popularized by
Hansen (1959). It takes the following general multimodal formulation
(Equation (\ref{eq:multimodal-conventional-accessibility})):

\begin{equation}
\label{eq:multimodal-conventional-accessibility}
S_i^m = \sum_{j=1}^JO_j \cdot f^m(c_{ij}^m)
\end{equation}

\noindent where:

\begin{itemize}
\tightlist
\item
  \(m\) is a set of modes.
\item
  \(c_{ij}^m\) is a measure of the cost of moving between \(i\) and
  \(j\) for each \(m\).
\item
  \(f^m(\cdot)\) is an impedance function of \(c_{ij}^m\) for each
  \(m\); it can take the form of any monotonically decreasing function
  chosen based on positive or normative criteria (A. Páez, Scott, and
  Morency 2012).
\item
  \(i\) is a set of origin locations (\(i = 1,\cdots,N\)).
\item
  \(j\) is a set of destination locations (\(j = 1,\cdots,J\)).
\item
  \(O_j\) is the number of opportunities at location \(j\).
\item
  \(S\) is Hansen-type accessibility as weighted sum of opportunities.
\end{itemize}

As indicators of urban structure, Hansen-type measures are informative,
but the meaning of 10,000 accessible school-seats is harder to pin down:
how many opportunities must any single student have access to?
Furthermore, in scenario comparisons, the interpretation of the
percentage change in accessible school seats is unclear. The
interpretability of Hansen-type accessibility has been discussed in
numerous studies, including recently by Hu and Downs (2019), Kelobonye
et al. (2020), and in greater depth by Merlin and Hu (2017) along with
Soukhov et al. (2023) and Soukhov et al. (2024). The interpretation of
accessibility is suggested to depend on how many people demand the
opportunity, especially for exclusive opportunity-types like
schools-seats (i.e., one school-seat is for one student).

In this paper, we benefit from new developments in accessibility
research, particularly the multimodal spatial availability measure
(Soukhov et al. 2023; Soukhov et al. 2024). Spatial availability is a
singly-constrained accessibility measure that accounts for competition
by students using different modes for exclusive opportunities, such as
school seats. The measure's single constraint ensures that the marginals
at the destination are met and thus the number of estimated school seats
(opportunities) are preserved and allocated proportionally to the
mode-using student population. This proportional allocation of
opportunities yields an interpretable and meaningful measure of
opportunity access, particularly when comparing across modes, at the
spatial resolution of a residential parcel, and multiple time periods.
See Soukhov et al. (2024) for further discussion, multimodal spatial
availability \(V_{i}^m\) is defined as given by Equation
(\ref{eq:spatial-availability}).

\begin{equation}
\label{eq:spatial-availability}
V_{i}^{m} = \sum_{j=1}^J O_j\ F^{tm}_{ij}
\end{equation}

\noindent where:

\begin{itemize}
\tightlist
\item
  \(F^{tm}_{ij}\) is a balancing factor that depends on the population
  and cost of movement in the system as part of the gravity modelling
  framework and is captured in Equation (\ref{eq:balancing-factors}) for
  mode \(m\).
\item
  \(O_j\) is the number of opportunities at \(j\).
\item
  \(V_i^m\) is the number of spatially available opportunities from the
  perspective of \(i\) for mode \(m\).
\end{itemize}

\(F^{tm}_{ij}\) can be understood as the joint probability of allocating
opportunities, where \(F^{pm}_{i}\) is the population-based balancing
factor that grants a larger share of opportunities to larger \(m\)
population spatial units and \(F^{cm}_{ij}\) is the impedance-based
balancing factor that grants a larger share of the opportunities to less
\(m\)-travel costly centers. Together \(F^{tm}_{ij}\) ensures
proportional allocation such that opportunities \(O\) (like
school-seats) are preserved for the whole region (i.e.,
\(O = \sum_{j} O_j = \sum_{i} V_i = \sum_{m}\sum_{i} V_i^m\) ) and is
reflected in Equation (\ref{eq:balancing-factors}):

\begin{equation}
\label{eq:balancing-factors}
F^{tm}_{ij} = \frac{F^{pm}_{i} \cdot F^{cm}_{ij}}{\sum_{i} F^{pm}_{i} \cdot F^{cm}_{ij}}
\end{equation}

\noindent where:

\begin{itemize}
\tightlist
\item
  The factor for allocation by population for each \(m\) at each \(i\)
  is \(F^{pm}_{i} = \frac{P_{i}^m}{\sum_{m}\sum_{i} P_{i}^m}\)
\item
  The factor for allocation by travel cost for each \(m\) at each \(i\)
  and \(j\) is
  \(F^{cm}_{ij} = \frac{f^m(c^m_{ij})}{\sum_{m}\sum_{i} f^m(c^m_{ij})}\)
\end{itemize}

It should be noted that, when summed over all spatial units in the
region, the population-based allocation factors \(F^{pm}_{i}\) always
equal 1 (\(\sum_{m}\sum_{i} F^{pm}_{i}= 1\)), likewise for
impedance-based allocation factors \(F^{cm}_{i}\)
(\(\sum_{m}\sum_{i} F^{cm}_{ij} = 1\)).

Hansen-type accessibility is not designed to preserve the number of
opportunities in the region, it simply counts the intensity of
opportunities that those in a zone can potentially interact with
(weighted by the friction of distance). Also, as discussed in Soukhov et
al. (2023), popular competitive accessibility measures such as the 2SFCA
(Joseph and Bantock 1982; Weibull 1976; Shen 1998; Luo and Wang 2003)
are internally inconsistent, and the only way it preserves the number of
opportunities is if the effect of the impedance function is ignored when
expanding the values of opportunities per capita to obtain the total
number of opportunities. On the other hand, the proportional allocation
procedure associated with calculating multimodal spatial availability
\(V_i^m\) consistently returns a number of opportunities available to
populations by mode that matches the total number of opportunities in
the region when summed. By doing this consistently, it is possible to
define a measure of multimodal spatial availability per capita as
presented in Equation (\ref{eq:SA-per-capita}) for use as a benchmark to
compare against the regional opportunities per capita
(\(\frac{\sum_{j} O_j}{\sum_{i} P_i}\)).

\begin{equation}
\label{eq:SA-per-capita}
v_i^m = \frac{V_i^m}{P_i^m}
\end{equation}

This paper uses the multimodal spatial availability \(V_i^m\) and
spatial availability per student \(v_i^m\) measures to quantify
accessibility changes for the following reasons.

First, all residential parcels are associated to their respective census
DA, TTS TAZ, and community boundary based on spatial location
(aggregated by DA and shown as a per capita rate in the second row of
Figure~\ref{fig-Fig2}). Each parcel is assigned a number of `potential'
motorized and non-motorized student aged population (from DA) and modal
share (from TTS) to calculate a motorized and non-motorized population
balancing factor \(F_{i}^{pm}\). For each parcel, the following is
retained: the 2011 and 2016 census average elementary-aged student
population per parcel (Figure~\ref{fig-Fig2} second row) and the average
motorized and non-motorized share as informed by the TTS aggregated by
community (Figure~\ref{fig-Fig3}).

Second, the school capacity \(O_j\) is estimated and associated with
each school (Figure~\ref{fig-Fig1}). All residential locations are
assumed to be able to access all schools by non-motorized and/or
motorized mode. All origins (residential locations) can reach all
schools by motorized mode, but few residential locations can reach
schools within a 27 minute non-motorized trip.

Third, the motorized and non-motorized impedance-balancing factors
\(F_{ij}^{cm}\) are calculated using a mode-specific impedance function
(Figure~\ref{fig-Fig4}) based on the respective estimated travel times
for each origin (parcel) to school combination (OD flows retained from
the TTS). Using network estimated travel time sensitive to observed
parcel origin to school destination flows conceptually addresses the
aggregation error that could result from using less granular zonal units
to represent origin/destinations (Kane and Kim 2020; Kwan 1998; Hewko,
Smoyer-Tomic, and Hodgson 2002). Finally, outputs from all three stages
are joined together. \(F_{i}^{pm}\) joined with \(F_{ij}^{cm}\) yields
the total balancing factor \(F^{tm}_{ij}\) which serves to
proportionally allocate the school capacity \(O_j\) to each parcel. The
value is multimodal spatial availability \(V_i^m\) which can be
interpreted as the number of school-seats that are available to the
mode-using population at that parcel. \(V_i^m\) is then summed and
represented at the DA and community level along with the calculated
\(v_i^m\) for interpretation.

\section{Results}\label{results}

The motorized and non-motorized spatial availabilities of school seats
\(V_i^m\) for both 2011 and 2016 are presented in the top half of
Figure~\ref{fig-Fig5}. The sum of \(V_i^m\) values in a year for both
motorized and non-motorized population equals the city-wide OTGC (across
both public and catholic school boards) in that year. This is a result
of \(V_i^m\)'s proportional allocation mechanism, singly-constraining
opportunities through the proportion allocation balancing factors. In
this way, \(V_i^m\) results can be interpreted as a proportion of the
total OTGC in that year as the sum of \(V_i^m\) for both modes in a year
is equal to the total OTGC in that year. In Figure~\ref{fig-Fig5}, \(i\)
is a DA, so \(V_i^m\) is the sum of the school seat availability per
parcel in each DA. Each parcel is assigned a potential student
population based on the average rate of school-aged population per
residential parcel in that DA and the travel impedance weighted travel
time to all schools. The top four plots of Figure~\ref{fig-Fig5}
demonstrate different magnitudes in part due to significantly lower
non-motorized modal share (refer to Figure~\ref{fig-Fig3}), however it
is apparent that both mode using populations have higher spatial
availability when they are more proximate to schools. Hence, all DAs
have higher values within Hamilton Central and those more proximate to
schools in less-urban communities. Trends appear similar for both 2011
and 2016: this aligns with intuition, proximate populations using
motorized modes benefit from the ability to reach more distant
destinations than non-motorized modes.

When \(V_i^m\) is divided by the number of students by mode in that DA,
\(v_i^m\) is derived (bottom four plots in Figure~\ref{fig-Fig5}). The
\(v_i^m\) plots demonstrate different spatial distribution trends than
\(V_i^m\) plots. If our aim is to provide sufficient availability to
schools for both motorized and non-motorized populations (e.g., greater
than 1.0 \(v_i^m\) school-seat availability per student), this rate is
only achieved in the core of the city for motorized modes (greens and
yellows). Notably, this rate is only available to non-motorized
populations in DAs that are in less-densely populated rural areas that
are school proximate and very few pockets of Hamilton Central. The
majority of schools closed were within Hamilton Central, so
non-motorized populations in DAs within Hamilton Central were the most
negatively impacted by losses in availability while certain motorized
populations captured these losses as availability gains.

\begin{figure}[H]

\centering{

\includegraphics[width=1.2\textwidth,height=\textheight]{Manuscript_files/figure-pdf/fig-Fig5-1.pdf}

}

\caption{\label{fig-Fig5}Multimodal spatial availability (purples) and
spatial availability per mode-using student (diverging reds and greens)
aggregated at the DA level for 2011 and 2016. Scales are represented in
quartiles.}

\end{figure}%

To highlight this point, Figure~\ref{fig-Fig6} visualises the change in
school spatial availability between 2016 and 2011 for both mode where
\(i\) are 2016 DAs. Though the 2011 and 2016 DAs system are very similar
(see Figure~\ref{fig-Fig2} and Figure~\ref{fig-Fig3}), they are not
identical: 16 DAs changed in their spatial extent (i.e., added, removed,
changed boundaries). The proportional allocation mechanism of spatial
availability provides flexibility in aggregating parcel-level values at
whichever spatial unit is meaningful. For this case, the change in
\(V_i^m\) and \(v_i^m\) represented is the subtraction between the value
aggregated at 2016 DA for the 2016 and 2011 scenario.

\begin{figure}[H]

\centering{

\includegraphics[width=1.2\textwidth,height=\textheight]{Manuscript_files/figure-pdf/fig-Fig6-1.pdf}

}

\caption{\label{fig-Fig6}Change in multimodal spatial availability and
spatial availability per mode-using student between 2016 and 2011.
Change is the subtraction of parcel-level results between 2016 and 2011
scenarios summed for each DA in the 2016 Canadian Census DA system.
Scales are represented in quartiles, with Q4 rounded to 0.}

\end{figure}%

From Figure~\ref{fig-Fig6} (top plots), it is apparent that decreases
(pinks and reds) in \(V_i^m\) appear across the city for both mode-using
populations. As the majority of closed schools are within the core of
the city: the more urban Hamilton Central and urban-areas of other
communities see a concentration of this decreased spatial availability
to school seats. Evidently, the number of OTGC in the city (equal to the
sum of \(V_i^m\)) decreased between the two years, so some sort of
decrease is to be expected. However, even when spatial availability is
divided by the mode-using population (which also decreased between the
years) \(v_i^m\) (bottom plots) continues to demonstrate decreases,
especially those locally proximate to closed schools.

This decrease follows different patterns depending on the mode.
\(v_i^{non-motorized}\) is most impacted for parcels within a 27 minute
walk travel time to schools that closed: these DAs demonstrate the most
dramatic decrease. Smaller-area DAs (in Hamilton-Central) and
larger-area DAs (in rural communities) proximate to closed schools
present Q1 decreases (red). For \(v_i^{motorized}\), the range offered
by motorized mode (estimated by shortest-path car travel times) is
city-wide, so since the majority of schools were centrally located, the
decrease is concentrated within the downtown core of Hamilton Central.
Some motorized-mode using populations in DAs do experience an increase
in \(V_i^m\) and \(v_i^m\): these DAs are often relatively proximate to
schools but more immediately outside walking range. Under competitive
and constrained calculations, these DAs benefit from the losses in
spatial availability that are predominately experienced by the
non-motorized populations within Hamilton Central.

The decrease in school seat spatial availability can be further
interpreted. How does potential non-motorized travel to school decrease?
How does the increase in potential motorized travel minutes impact GHG
emissions? Figure~\ref{fig-Fig7} visualises these related intermediates
for one-way home-to-school travel in a given day. The top plot of
Figure~\ref{fig-Fig7} demonstrates the sum of the change in
non-motorized minutes per 2016 DA. These travel time minutes are the
difference between the aggregated 2016 and 2011 scenarios for
non-motorized travel times multiplied by the proportion of
walking/cycling mode-use and rate of student population associated with
the parcels in each 2016 DA. It can be seen that Q1 decreases are
proximate to closed schools. Smaller decreases (and even a few
increases) are apparent in a range further from closed schools,
especially those also in proximity to the few opened/expanded schools.
Intuitively, these trends mirror what is shown within the
\(V^{non-motorized}\) landscape in Figure~\ref{fig-Fig6} (Pearson
correlation coefficient of 0.796) as travel time is an input into the
cost balancing factor \(F^{cm}_{ij}\). Similarly, the bottom plot
(Figure~\ref{fig-Fig7}) visualises the change in estimated emissions
from motorized travel minutes. The change in GHG emissions is also
highly correlated with \(V^{motorized}\) in Figure~\ref{fig-Fig6}
(Pearson correlation coefficient of 0.838) as it is also a product of
travel time.

\begin{figure}[H]

\centering{

\includegraphics[width=0.9\textwidth,height=\textheight]{Manuscript_files/figure-pdf/fig-Fig7-1.pdf}

}

\caption{\label{fig-Fig7}Change in active travel time minutes (top plot)
and home-to-school GHG emissions (bottom plot) between 2011 and 2016
scenarios. Change is the subtraction of parcel-level results between
2016 and 2011 scenarios summed for each DA in the 2016 Canadian Census
DA system. Scales are represented in quartiles, with Q4 rounded to 0.}

\end{figure}%

Notably, the mode type of spatial availability loss is important. The
change in active travel minutes and GHG emissions highlights the
negative externalities of \emph{decreased} non-motorized spatial
availability. Normatively, our cities should be increasing spatial
availability for non-motorized mode users given the climate crisis. The
wave of school closures in Hamilton increased motorized spatial
availability and, hence, potential GHG emissions. The policy drastically
reduced non-motorized spatial availability, and this is seen in the
decrease in potential non-motorized travel minutes as well as increased
GHG emissions in rural areas that did benefit from motorized spatial
availability gains. The proportion of non-motorized mode use is low in
both 2011 and 2016, however, the closure of schools that are more
proximate to student-population density eliminates the potential of
those trips ever becoming active. And who is most impacted by the
reduction of non-motorized spatial availability? A summary of spatial
availability per Hamilton community and associated dimensions in 2016,
along with the percentage change between 2011 and 2016, is summarized in
Table~\ref{tbl-Tab1}.

\global\setlength{\Oldarrayrulewidth}{\arrayrulewidth}

\global\setlength{\Oldtabcolsep}{\tabcolsep}

\setlength{\tabcolsep}{2pt}

\renewcommand*{\arraystretch}{1.5}

\providecommand{\ascline}[3]{\noalign{\global\arrayrulewidth #1}\arrayrulecolor[HTML]{#2}\cline{#3}}

\begin{longtable}[c]{|p{0.47in}|p{0.40in}|p{0.29in}|p{0.38in}|p{0.34in}|p{0.40in}|p{0.29in}|p{0.38in}|p{0.29in}|p{0.40in}|p{0.29in}|p{0.38in}|p{0.29in}}

\caption{\label{tbl-Tab1}Spatial availability and associated dimensions
aggregated by communities}

\tabularnewline

\multicolumn{1}{>{\centering}p{\dimexpr 0.47in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{}}}} & \multicolumn{2}{>{\centering}p{\dimexpr 0.7in+2\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{Hamilton\ C.}}}} & \multicolumn{2}{>{\centering}p{\dimexpr 0.72in+2\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{Dundas}}}} & \multicolumn{2}{>{\centering}p{\dimexpr 0.7in+2\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{Stoney\ Creek}}}} & \multicolumn{2}{>{\centering}p{\dimexpr 0.67in+2\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{Ancaster}}}} & \multicolumn{2}{>{\centering}p{\dimexpr 0.7in+2\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{Flamborough}}}} & \multicolumn{2}{>{\centering}p{\dimexpr 0.67in+2\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{Glanbrook}}}} \\

\multicolumn{1}{>{\centering}p{\dimexpr 0.47in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{}}}} & \multicolumn{1}{>{\centering}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{}}}} & \multicolumn{1}{>{\centering}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{\%\ Δ}}}} & \multicolumn{1}{>{\centering}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{}}}} & \multicolumn{1}{>{\centering}p{\dimexpr 0.34in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{\%\ Δ}}}} & \multicolumn{1}{>{\centering}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{}}}} & \multicolumn{1}{>{\centering}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{\%\ Δ}}}} & \multicolumn{1}{>{\centering}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{}}}} & \multicolumn{1}{>{\centering}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{\%\ Δ}}}} & \multicolumn{1}{>{\centering}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{}}}} & \multicolumn{1}{>{\centering}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{\%\ Δ}}}} & \multicolumn{1}{>{\centering}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{}}}} & \multicolumn{1}{>{\centering}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{\%\ Δ}}}} \\

\multicolumn{1}{>{\raggedright}p{\dimexpr 0.47in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{LIM-AT}}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{27.0}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{14\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{12.5}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.34in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{10\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{11.6}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{10\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 8.9}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{24\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 7.6}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{4\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 7.5}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-14\%}}} \\

\multicolumn{1}{>{\raggedright}p{\dimexpr 0.47in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{Pop.}}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{33437.7}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-4\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 2365.3}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.34in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-8\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 8070.3}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{6\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 5387.5}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{12\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 5172.4}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-4\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 4150.0}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{58\%}}} \\

\multicolumn{1}{>{\raggedright}p{\dimexpr 0.47in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{OTGC}}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{36878.3}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-8\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 2377.0}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.34in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{0\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 9387.8}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{4\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 3404.7}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{0\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 4638.2}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{2\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 1601.8}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{22\%}}} \\

\multicolumn{1}{>{\centering}p{\dimexpr 0.47in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{}}}} & \multicolumn{12}{>{\centering}p{\dimexpr 4.15in+22\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{Motorized}}}} \\

\multicolumn{1}{>{\raggedright}p{\dimexpr 0.47in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{V\ (mt)}}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{39626.9}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-8\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 1817.7}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.34in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-8\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 6656.8}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{6\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 4338.0}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{8\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 3290.3}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-2\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 2371.7}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{58\%}}} \\

\multicolumn{1}{>{\raggedright}p{\dimexpr 0.47in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{v\ (mt)}}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{1.2}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-4\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{0.8}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.34in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{0\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{0.8}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{0\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{0.8}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-4\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{0.6}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{2\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{0.6}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{0\%}}} \\

\multicolumn{1}{>{\raggedright}p{\dimexpr 0.47in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{GHG}}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{47788.1}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-2\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 4851.7}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.34in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-8\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{15183.3}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{8\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 9470.9}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{14\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{12839.1}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-4\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 8798.4}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{60\%}}} \\

\multicolumn{1}{>{\raggedright}p{\dimexpr 0.47in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{GHG/pop}}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{1.4}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{2\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{2.1}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.34in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{0\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{1.9}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{0\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{1.8}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{2\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{2.5}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{0\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{2.1}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{2\%}}} \\

\multicolumn{1}{>{\centering}p{\dimexpr 0.47in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{}}}} & \multicolumn{12}{>{\centering}p{\dimexpr 4.15in+22\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{Non-motorized}}}} \\

\multicolumn{1}{>{\raggedright}p{\dimexpr 0.47in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{V\ (nmt)}}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{152.7}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-44\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ \ 0.0}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.34in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-100\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 23.9}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-42\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ \ 4.0}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-68\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ \ 3.9}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-64\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ \ 1.9}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-2\%}}} \\

\multicolumn{1}{>{\raggedright}p{\dimexpr 0.47in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{v\ (nmt)}}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{0.6}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-8\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{0.0}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.34in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-100\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{0.7}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-18\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{0.7}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-18\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{1.7}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-6\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{1.4}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-16\%}}} \\

\multicolumn{1}{>{\raggedright}p{\dimexpr 0.47in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{tt}}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{4365.7}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-38\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ \ \ 0.0}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.34in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-100\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 585.3}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-28\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 103.3}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-62\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ \ 40.4}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-62\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ \ 22.5}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{12\%}}} \\

\multicolumn{1}{>{\raggedright}p{\dimexpr 0.47in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{tt/pop}}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{18.2}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{0\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.34in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{17.8}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{2\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{18.3}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{2\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{17.9}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-2\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{16.4}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-4\%}}} \\

\end{longtable}

\arrayrulecolor[HTML]{000000}

\global\setlength{\arrayrulewidth}{\Oldarrayrulewidth}

\global\setlength{\tabcolsep}{\Oldtabcolsep}

\renewcommand*{\arraystretch}{1}

Due to spatial availability's flexible proportional allocation mechanism
and this cases' calculation at the parcel level, results can also be
summarized by community (Table~\ref{tbl-Tab1}). The majority of schools
were closed in Hamilton Central. This community represents 60\% of the
5-14 year old population in 2016. Though the population between 2011 and
2016 decreased, the rate of school-seat spatial availability decreased
disproportionately more (population by -4\% and spatial availability by
-8\%). This disproportionate decrease is greatest in Hamilton Central
than any other community. This may have equity concerns as Hamilton
Central has the highest LIM-AT prevalence, with 27\% of households
LIM-AT in 2016, 3.7 times greater than Glanbrook, a community with the
lowest level of LIM-AT prevalence in the city. Between 2016 and 2011,
Hamilton Central had the lowest levels of emissions per student and saw
the highest per student increase in 2016. Hamilton Center had the
highest non-motorized trip potential, and it was reduced in part by the
school closure policies.

Conversely, communities with the lowest LIM-AT benefit the most, notably
in gains to motorized spatial availability. Glanbrook, Flamborough and
Ancaster have the lowest prevalence of LIM-AT of any Hamilton Community
and together account for 30\% of the 5-14 year old population in 2016.
These communities experienced a growth in student population (or only a
small decrease) and proportionally benefited from an increase in
motorized spatial availability.

Across the city, those who commute by non-motorized mode see the most
drastic lose in spatial availability. In all communities, spatial
availability decreases at greater levels than population. Notably in
Dundas (2nd highest LIM-AT), the only school in the community closed,
leaving the population in the community with no other option but to use
motorized modes. Glanbrook (lowest LIM-AT), which experienced a
relatively sharp increase in student population along with non-motorized
using population, suffered a lose in non-motorized spatial availability
despite seeing the highest increase in motorized spatial availability.
Overall, the gain in motorized spatial availability resulted in a
further erosion of non-motorized spatial availability. This erosion is
related to a lose in potential active home-to-school travel minutes
(3309.3 fewer minutes than in 2011, decreasing the potential active
travel time per (all mode using) student from 0.14 to 0.09 minutes), as
well as an increase in potential GHG emissions city-wide (3416 kg
\(CO_2e\) more emission than in 2011, increasing the kg \(CO_2e\) per
child rate for home-to-school trips from 1.64 to 1.68).

\section{Conclusion}\label{conclusion}

In this paper, we estimate the change in multimodal spatial availability
as a result of school consolidations/closures between the years of 2011
and 2016 in an empirical case study of Hamilton, a mid-size city
(\textasciitilde570,000 pop) in Ontario, Canada (Government of Canada
2022). The spatial availability and spatial availability per student are
calculated at the parcel level and aggregated at the Canadian Census DA
spatial unit and at the six Hamilton communities for comparison. Schools
closed and the number of school seats decreased so spatial availability
declined overall: we discuss where, for whom, and with what potential
impacts.

Overall, the closure of 7\% of elementary schools in Hamilton between
2011 and 2016 resulted in -4\% decline in school-seat availability
city-wide (2,280 fewer school-seats), but specifically a -46\% decline
in non-motorized spatial availability. Though non-motorized modal share
was low in both 2011 and 2016, the \emph{potential} for active transport
trips was significantly eroded as schools within a 27 minute walk
catchment were closed in the more densely populated and central areas of
the city. Further, the communities with the highest LIM-AT prevalence
were the most impacted: communities like Hamilton Central and Dundas saw
the most drastic loss in non-motorized spatial availability while
communities with lower LIM-AT prevalence saw gains in motorized spatial
availability (as a handful of rural schools opened). The motivation of
operational savings associated with public service consolidation for the
school boards did not account for the mobility-related burdens
\emph{certain} families are now saddled with.

In addition to spatial availability impacts, associated mobility
measures are assessed. City-wide, home-to-school non-motorized minutes
(for a one-way trip) reduced by 3,309 and GHG emissions associated with
motorized home-to-school travel increased by 3,416 kg \(CO_2e\). The
decision to consolidate schools, increase the length of motorized travel
and decrease the potential for non-motorized impacts will have daily
impacts on students and their families (Mandic et al. 2022; Pabayo et
al. 2012), the broader community (Pietrabissa 2023; Merrall, Higgins,
and Páez 2024; Bittencourt and Giannotti 2023) and the environment
(Pantelaki, Claudia Caspani, and Maggi 2024; Rong et al. 2022).

Like many other localities (J. Lee and Lubienski 2017; Autti and
Hyry-Beihammer 2014; Beuchert et al. 2018), Hamilton is not immune to
top-down operational efficiency assessments that determine if community
infrastructure is underutilized and closed. From the perspective of the
provincial government of Ontario, the school closures which occurred in
Hamilton resulted in operational savings-per-student. We demonstrate how
these decisions resulted in both families and students paying a price
through spatial availability, lost potential walk minutes, and increased
GHG emissions.

By quantifying the spatial availability, environmental, and
active-travel implications of school closure/consolidation policies in
Hamilton, our paper offers a methodology for spatial policy analysis
scenario. The presented methodology can be used by researchers and
decision-makers to plan and evaluate equitable and sustainable urban
planning policies from a spatial perspective. At the core of the method
is spatial availability (Soukhov et al. 2023; Soukhov et al. 2024), a
singly-constrained accessibility measure that can be calculated at the
finest spatial resolution (in our paper, parcel-level) and aggregated at
whichever spatial unit is most meaningful for interpretation. We urge
policy-makers to view the `spatial availability' of public resources
from a per-capita perspective, plan capacity and location of service
provision with the cost of travel (and associated implications such as
GHG emissions, active transportation mode, safety of active travel
modes, etc.) that sufficiently serves the target population. In this
work, we suggest and assume a benchmark of 1.0 school-seats per student
which accounts for sufficiency. Spatial availability can be used to
obtain this benchmark since the total number of opportunities (i.e.,
school capacity) is preserved in the region of analysis unlike
unconstrained forms of accessibility measurement e.g., Hansen-type
measure (Hansen 1959). As such, the assigned spatial availability can be
divided by population at each zone to yield an interpretable per capita
measurement.

Ultimately, this case study furnishes evidence that consolidation of
schools was particularly ill-timed: with the advent of the COVID-19
pandemic, the education system lacked the resiliency to accommodate
reduced classroom capacities and left parents who once lived within
active transportation distance to schools relying on motorized
transportation for their children. These top-down austerity measures
left the Hamilton local school system more vulnerable to the impacts of
COVID-19 by undervaluing the societal role of schools in neighbourhoods.
By identifying the impact of school consolidation policies in Ontario we
anticipate that researchers and decision-makers will have better
information to center the wellbeing of all residents, including
students, and the planet in urban planning policies.

As with all studies, our work is associated with limitations on how
results should be interpreted. The calculated spatial availability
assumes any residential parcel can have a student (at the DA rate of
students per residential unit) so all parcels are assigned a DA rate
weighted by how many residential units are contained within the parcel.
Each parcel can also visit any school, though travel times based on
observed origin-destination behaviour (from the TTS) are more likely
through the use of the empirical travel impedance functions. In this
way, the spatial availability demonstrates what \emph{potential
interaction} a student could have based on historic and average travel
behaviour and shortest network-path travel times if residing in point in
space. This representation is average, and by design obfuscates
considerations that could make non-average and shortest-path travel
unrealistic such as safety and built-environment concerns (C. Lee et al.
2021; Yumita et al. 2021) nor considers significant non-transportation
factors (Palm and Farber 2020). Further, this study does not consider
transit, school-bus, or cycling modes which have implications on how the
GHG emissions estimation can be interpreted. Transit and school bus
modes often taken more circuitous routes (Yumita et al. 2021) but pool
students thereby reducing emissions. Cycling reduces travel time over
the same walking distance at no extra emitted GHG. Lastly, this study
also ascribes a static emission factor to represent motorized travel
minutes based on an average diesel passenger car. In reality, GHG
emissions emitted per minute of driving varies extensively depending on
operating conditions and vehicle characteristics (Soukhov and Mohamed
2022). In these ways, the assumptions made are to illustrate
\emph{potential} and \emph{average} changes that resulted in a backdrop
of school consolidation/closures to illustrate accessibility-related
impacts for the city of Hamilton.

\pagebreak

\section{References}\label{references}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-auditorgeneralofontarioAnnualReport20152015}
Auditor General of Ontario. 2015. {``Annual Report 2015.''} Office of
the Auditor General of Ontario.
\url{https://www.auditor.on.ca/en/content/annualreports/arreports/en15/2015AR_en_final.pdf}.

\bibitem[\citeproctext]{ref-Autti_Hyry-Beihammer_2014}
Autti, Outi, and Eeva Kaisa Hyry-Beihammer. 2014. {``School Closures in
Rural Finnish Communities.''} \emph{Journal of Research in Rural
Education} 29 (1): 1--17.

\bibitem[\citeproctext]{ref-batista_estimation_2019}
Batista, S. F. A., Ludovic Leclercq, and Nikolas Geroliminis. 2019.
{``Estimation of Regional Trip Length Distributions for the Calibration
of the Aggregated Network Traffic Models.''} \emph{Transportation
Research Part B: Methodological} 122 (April): 192--217.
\url{https://doi.org/10.1016/j.trb.2019.02.009}.

\bibitem[\citeproctext]{ref-von_Bergmann_Shkolnik_Jacobs_2021}
Bergmann, J von, D Shkolnik, and A Jacobs. 2021. \emph{Cancensus: R
Package to Access, Retrieve, and Work with Canadian Census Data and
Geography}.

\bibitem[\citeproctext]{ref-Beuchert_Humlum_Nielsen_Smith_2018}
Beuchert, Louise, Maria Knoth Humlum, Helena Skyt Nielsen, and Nina
Smith. 2018. {``The Short-Term Effects of School Consolidation on
Student Achievement: Evidence of Disruption?''} \emph{Economics of
Education Review} 65 (August): 31--47.
\url{https://doi.org/10.1016/j.econedurev.2018.05.004}.

\bibitem[\citeproctext]{ref-bierbaumMobilityJusticeLinking2021}
Bierbaum, Ariel H., Alex Karner, and Jesus M. Barajas. 2021. {``Toward
{Mobility Justice}: {Linking Transportation} and {Education Equity} in
the {Context} of {School Choice}.''} \emph{Journal of the American
Planning Association} 87 (2): 197--210.
\url{https://doi.org/10.1080/01944363.2020.1803104}.

\bibitem[\citeproctext]{ref-bittencourtEvaluatingAccessibilityAvailability2023}
Bittencourt, Tainá A., and Mariana Giannotti. 2023. {``Evaluating the
Accessibility and Availability of Public Services to Reduce Inequalities
in Everyday Mobility.''} \emph{Transportation Research Part A: Policy
and Practice} 177 (November): 103833.
\url{https://doi.org/10.1016/j.tra.2023.103833}.

\bibitem[\citeproctext]{ref-brunsdonOpeningPractice2021}
Brunsdon, Chris, and Alexis Comber. 2021. {``Opening Practice:
Supporting Reproducibility and Critical Spatial Data Science.''}
\emph{Journal of Geographical Systems} 23 (4): 477--96.
\url{https://doi.org/10.1007/s10109-020-00334-2}.

\bibitem[\citeproctext]{ref-butlerClosureRideau2019a}
Butler, Jesse K., Ruth G. Kane, and Fiona R. Cooligan. 2019. {``The
{Closure} of {Rideau High School}: {A Case Study} in the {Political
Economy} of {Urban Education} in {Ontario}.''} \emph{Canadian Journal of
Educational Administration and Policy}, no. 191.

\bibitem[\citeproctext]{ref-Statistics_Canada_2011}
Canada, Statistics. 2011. {``Census of Canda Census Profiles: Median
After-Tax Household Income, 5 to 9 Years, 10 to 14 Years, 15 to 19
Years, for Census Metroplian Areas and Dissimination Areas.''}

\bibitem[\citeproctext]{ref-Statistics_Canada_2016}
---------. 2016. {``Census of Canda Census Profiles: Median After-Tax
Household Income, 5 to 9 Years, 10 to 14 Years, 15 to 19 Years, for
Census Metroplian Areas and Dissimination Areas.''}

\bibitem[\citeproctext]{ref-Christiaanse_2020}
Christiaanse, Suzan. 2020. {``Rural Facility Decline: A Longitudinal
Accessibility Analysis Questioning the Focus of Dutch
Depopulation-Policy.''} \emph{Applied Geography} 121 (August): 102251.
\url{https://doi.org/10.1016/j.apgeog.2020.102251}.

\bibitem[\citeproctext]{ref-conveyalConveyalR5Routing2022}
Conveyal. (2015) 2022. {``Conveyal {R5 Routing Engine}.''} {Conveyal}.
\url{https://github.com/conveyal/r5}.

\bibitem[\citeproctext]{ref-craggsAxeCouldFall2012}
Craggs, Samantha. 2012. {``Axe Could Fall on More Hamilton Schools,
Projections Show.''} \emph{{CBC} News}, October.
\url{https://www.cbc.ca/news/canada/hamilton/headlines/axe-could-fall-on-more-hamilton-schools-projections-show-1.1183565}.

\bibitem[\citeproctext]{ref-craggsBoardEvaluatingFuture2013}
---------. 2013. {``Board Evaluating Future of 76 Public Elementary
Schools.''} \emph{{CBC} News}, February.
\url{https://www.cbc.ca/news/canada/hamilton/headlines/board-evaluating-future-of-76-public-elementary-schools-1.1347581}.

\bibitem[\citeproctext]{ref-Dai_Wang_Zhang_Liao_Liu_2019}
Dai, Te-qi, Liang Wang, Yu-chao Zhang, Cong Liao, and Zheng-bing Liu.
2019. {``The Cost of School Consolidation Policy: Implications from
Decomposing School Commuting Distances in Yanqing, Beijing.''}
\emph{Applied Spatial Analysis and Policy} 12 (2): 191--204.
\url{https://doi.org/10.1007/s12061-017-9238-2}.

\bibitem[\citeproctext]{ref-data_management_group_tts_2018}
Data Management Group. 2018. {``{TTS} - {Transportation} {Tomorrow}
{Survey} 2016.''}
\url{http://dmg.utoronto.ca/transportation-tomorrow-survey/tts-introduction}.

\bibitem[\citeproctext]{ref-fitdistrplus_2015}
Delignette-Muller, Marie Laure, and Christophe Dutang. 2015.
{``{fitdistrplus}: An {R} Package for Fitting Distributions.''}
\emph{Journal of Statistical Software} 64 (4): 1--34.
\url{https://www.jstatsoft.org/article/view/v064i04}.

\bibitem[\citeproctext]{ref-desjardinsFramingActive2024}
Desjardins, Elise, Jason Lam, Darcy Reynard, Damian Collins, E. Owen D.
Waygood, and Antonio Páez. 2024. {``Framing Active School Travel in
{Ontario}, or How Spinach Is Good for You.''} \emph{Transportation
Research Part A: Policy and Practice} 180 (February): 103953.
\url{https://doi.org/10.1016/j.tra.2024.103953}.

\bibitem[\citeproctext]{ref-faoOntarioSchoolBoards2023}
FAO. 2023. {``Ontario School Boards: Enrolment, Finances and Student
Outcomes.''} 978-1-4868-7600-6. Financial Accountability Office of
Ontario.
\url{https://www.fao-on.org/en/Blog/Publications/FA2207schoolboards}.

\bibitem[\citeproctext]{ref-geofabrikOntarioOpenStreetMapGeofabrik2022}
Geofabrik. 2022. {``Ontario {OpenStreetMap} - {Geofabrik Download
Server}.''} 2022.
\url{https://download.geofabrik.de/north-america/canada/ontario.html}.

\bibitem[\citeproctext]{ref-governmentofcanadaDictionaryCensusPopulation2017}
Government of Canada, Statistics Canada. 2017. {``Dictionary, Census of
Population, 2016 - Low-Income Measure, After Tax ({LIM}-{AT}).''} May 3,
2017.
\url{https://www12.statcan.gc.ca/census-recensement/2016/ref/dict/fam021-eng.cfm}.

\bibitem[\citeproctext]{ref-governmentofcanadaProfileTableCensus2022}
---------. 2022. {``Profile Table, Census Profile, 2021 Census of
Population - Hamilton, City (c) {[}Census Subdivision{]}, Ontario.''}
February 9, 2022.
\url{https://www12.statcan.gc.ca/census-recensement/2021/dp-pd/prof/index.cfm?Lang=E}.

\bibitem[\citeproctext]{ref-Hamilton_2014}
Hamilton. 2014. {``Map 5: Downtown Hamilton Buisness Improvement
Area.''} Hamilton.
\url{https://www.hamilton.ca/sites/default/files/media/browser/2014-11-27/downtown-hamilton-bia-map.pdf}.

\bibitem[\citeproctext]{ref-hansenHowAccessibilityShapes1959}
Hansen, Walter G. 1959. {``How {Accessibility Shapes Land Use}.''}
\emph{Journal of the American Institute of Planners} 25 (2): 73--76.
\url{https://doi.org/10.1080/01944365908978307}.

\bibitem[\citeproctext]{ref-hewkoMeasuringNeighbourhoodSpatial2002}
Hewko, Jared, Karen E Smoyer-Tomic, and M John Hodgson. 2002.
{``Measuring Neighbourhood Spatial Accessibility to Urban Amenities:
Does Aggregation Error Matter?''} \emph{Environment and Planning A:
Economy and Space} 34 (7): 1185--1206.
\url{https://doi.org/10.1068/a34171}.

\bibitem[\citeproctext]{ref-horbachov_theoretical_2018}
Horbachov, Peter, and Stanislav Svichynskyi. 2018. {``Theoretical
Substantiation of Trip Length Distribution for Home-Based Work Trips in
Urban Transit Systems.''} \emph{Journal of Transport and Land Use} 11
(1): 593--632. \url{https://www.jstor.org/stable/26622420}.

\bibitem[\citeproctext]{ref-hu_2019_measuring}
Hu, Yujie, and Joni Downs. 2019. {``Measuring and Visualizing
Place-Based Space-Time Job Accessibility.''} Journal Article.
\emph{Journal of Transport Geography} 74: 278--88.
\url{https://doi.org/10.1016/j.jtrangeo.2018.12.002}.

\bibitem[\citeproctext]{ref-hwdsbLongTermFacilities2013}
HWDSB. 2013. {``Long Term Facilities Master Plan - Accomodation Review
Schedule.''} Hamilton, Ontario: Hosted by {CBC}.
\url{https://www.documentcloud.org/documents/603032-hwdsb-schools-for-review}.

\bibitem[\citeproctext]{ref-HWDSB_2019}
HWDSB. 2019. {``Transportation: Policy No. 3.10.''}
\url{https://www.hamiltonschoolbus.ca/policies/hwdsb/HWDSB_Transportation_Policy.pdf}.

\bibitem[\citeproctext]{ref-irwinSchoolClosure2012}
Irwin, Bill, and Mark Seasons. 2012. {``School {Closure Decision-Making
Processes Problems} and {Prospects}.''} \emph{Canadian Journal of Urban
Research} 21 (2): 45--67.

\bibitem[\citeproctext]{ref-josephMeasuringPotentialPhysical1982}
Joseph, Alun E., and Peter R. Bantock. 1982. {``Measuring Potential
Physical Accessibility to General Practitioners in Rural Areas: A Method
and Case Study.''} \emph{Social Science \& Medicine} 16 (1): 85--90.
\url{https://doi.org/10.1016/0277-9536(82)90428-2}.

\bibitem[\citeproctext]{ref-kaneParcelsPointsProximity2020}
Kane, Kevin, and Young-An Kim. 2020. {``Parcels, Points, and Proximity:
Can Exhaustive Sources of Big Data Improve Measurement in Cities?''}
\emph{Environment and Planning B: Urban Analytics and City Science} 47
(4): 695--715. \url{https://doi.org/10.1177/2399808318797135}.

\bibitem[\citeproctext]{ref-kelobonye2020measuring}
Kelobonye, Keone, Heng Zhou, Gary McCarney, and Jianhong Xia. 2020.
{``Measuring the Accessibility and Spatial Equity of Urban Services
Under Competition Using the Cumulative Opportunities Measure.''} Journal
Article. \emph{Journal of Transport Geography} 85: 102706.
https://doi.org/\url{https://doi.org/10.1016/j.jtrangeo.2020.102706}.

\bibitem[\citeproctext]{ref-kleinhuisSeriousConcernsSchool2013}
Kleinhuis, Amanda. 2013. {``Serious Concerns about School Accommodation
Review Process - Raise the Hammer.''} 2013.
\url{https://www.raisethehammer.org/article/1979/serious_concerns_about_school_accommodation_review_process}.

\bibitem[\citeproctext]{ref-kwanSpaceTimeIntegralMeasures1998}
Kwan, Mei-Po. 1998. {``Space-Time and Integral Measures of Individual
Accessibility: A Comparative Analysis Using a Point-Based Framework.''}
\emph{Geographical Analysis} 30 (3): 191--216.
\url{https://doi.org/10.1111/j.1538-4632.1998.tb00396.x}.

\bibitem[\citeproctext]{ref-leeNeighborhoodEnvironmentsUtilitarian2021}
Lee, C, C Lee, OT Stewart, HA Carlos, A Adachi-Mejia, EM Berke, and MP
Doescher. 2021. {``Neighborhood Environments and Utilitarian Walking
Among Older Vs. Younger Rural Adults.''} \emph{{FRONTIERS} {IN} {PUBLIC}
{HEALTH}} 9 (June). \url{https://doi.org/10.3389/fpubh.2021.634751}.

\bibitem[\citeproctext]{ref-Lee_Lubienski_2017}
Lee, Jin, and Christopher Lubienski. 2017. {``The Impact of School
Closures on Equity of Access in Chicago.''} \emph{Education and Urban
Society} 49 (1): 53--80. \url{https://doi.org/10.1177/0013124516630601}.

\bibitem[\citeproctext]{ref-luoMeasuresSpatialAccessibility2003}
Luo, Wei, and Fahui Wang. 2003. {``Measures of {Spatial Accessibility}
to {Health Care} in a {GIS Environment}: {Synthesis} and a {Case Study}
in the {Chicago Region}.''} \emph{Environment and Planning B: Planning
and Design} 30 (6): 865--84. \url{https://doi.org/10.1068/b29120}.

\bibitem[\citeproctext]{ref-mackenzieBlueprintFixOntario2018}
Mackenzie, Hugh. 2018. {``A Blueprint to Fix Ontario's Education Funding
Formula.''} \emph{Canadian Centre for Policy Alternatives {\textbar}
Ontario}.
\url{https://policyalternatives.ca/sites/default/files/uploads/publications/Ontario\%20Office/2018/03/Course\%20Correction.pdf}.

\bibitem[\citeproctext]{ref-mandicAdolescentsPerceptionsWalking2022}
Mandic, Sandra, Enrique García Bengoechea, Debbie Hopkins, Kirsten
Coppell, and John C. Spence. 2022. {``Adolescents' Perceptions of
Walking and Cycling to School Differ Based on How Far They Live from
School.''} \emph{Journal of Transport \& Health} 24 (March): 101316.
\url{https://doi.org/10.1016/j.jth.2021.101316}.

\bibitem[\citeproctext]{ref-merlin2017competition}
Merlin, Louis A., and Lingqian Hu. 2017. {``Does Competition Matter in
Measures of Job Accessibility? Explaining Employment in Los Angeles.''}
Journal Article. \emph{Journal of Transport Geography} 64: 77--88.
\url{https://doi.org/10.1016/j.jtrangeo.2017.08.009}.

\bibitem[\citeproctext]{ref-Merrall_2021}
Merrall, John. 2021. {``The Effect of School Closures on House Prices in
Hamilton, Ontario.''} PhD thesis, McMaster University; School of Earth,
Environment,; Society.

\bibitem[\citeproctext]{ref-MerrallWhatsASchool2024}
Merrall, John, Christopher D. Higgins, and Antonio Páez. 2024. {``What's
a School Worth to a Neighborhood? {A} Spatial Hedonic Analysis of
Property Prices in the Context of Accommodation Reviews in Ontario.''}
\emph{Geographical Analysis} 56 (2): 217--43.
\url{https://doi.org/10.1111/gean.12377}.

\bibitem[\citeproctext]{ref-ministryofeducationPupilAccommodationReview2006}
Ministry of Education. 2006. {``Pupil Accommodation Review Guidelines -
Memo.''} Toronto: Business; Finance Divisio.
\url{https://efis.fma.csc.gov.on.ca/faab/Memos/B2006/B_12.pdf}.

\bibitem[\citeproctext]{ref-morencyHowMany2007}
Morency, C., M. Demers, and L. Lapierre. 2007. {``How Many Steps Do You
Have in Reserve? {Thoughts} and Measures about a Healthier Way to
Travel.''} \emph{Transportation Research Record} 2002: 1--6.
\url{https://doi.org/10.3141/2002-01}.

\bibitem[\citeproctext]{ref-morencyStepsReserve2009}
Morency, C., M. J. Roorda, and M. Demers. 2009. {``Steps in {Reserve
Comparing Latent Walk Trips} in {Toronto}, {Ontario}, and {Montreal},
{Quebec}, {Canada}.''} \emph{Transportation Research Record} 2140:
111--19. \url{https://doi.org/10.3141/2140-12}.

\bibitem[\citeproctext]{ref-ontarioSchoolFacilityInventory2017}
Ontario. 2017. {``School Facility Inventory System ({SFIS}) - Dataset -
Ontario Data Catalogue.''} February 2017.
\url{https://data.ontario.ca/dataset/school-facility-inventory-system-sfis}.

\bibitem[\citeproctext]{ref-OpenStreetMap_2021}
OpenStreetMap. 2021.

\bibitem[\citeproctext]{ref-pabayoImportanceActiveTransportation2012}
Pabayo, Roman, Katerina Maximova, John C. Spence, Kerry Vander Ploeg,
Biao Wu, and Paul J. Veugelers. 2012. {``The Importance of Active
Transportation to and from School for Daily Physical Activity Among
Children.''} \emph{Preventive Medicine} 55 (3): 196--200.
\url{https://doi.org/10.1016/j.ypmed.2012.06.008}.

\bibitem[\citeproctext]{ref-paezOpenSpatial2021c}
Páez, Antonio. 2021. {``Open Spatial Sciences: An Introduction.''}
\emph{Journal of Geographical Systems} 23 (4): 467--76.
\url{https://doi.org/10.1007/s10109-021-00364-4}.

\bibitem[\citeproctext]{ref-paez2012measuring}
Páez, A., D. M. Scott, and C. Morency. 2012. {``Measuring Accessibility:
Positive and Normative Implementations of Various Accessibility
Indicators.''} Journal Article. \emph{Journal of Transport Geography}
25: 141--53. \url{https://doi.org/10.1016/j.jtrangeo.2012.03.016}.

\bibitem[\citeproctext]{ref-palmRolePublicTransit2020}
Palm, Matthew, and Steven Farber. 2020. {``The Role of Public Transit in
School Choice and After-School Activity Participation Among Toronto High
School Students.''} \emph{Travel Behaviour and Society} 19 (April):
219--30. \url{https://doi.org/10.1016/j.tbs.2020.01.007}.

\bibitem[\citeproctext]{ref-pantelakiImpactHomeschoolCommuting2024}
Pantelaki, Evangelia, Anna Claudia Caspani, and Elena Maggi. 2024.
{``Impact of Home-School Commuting Mode Choice on Carbon Footprint and
Sustainable Transport Policy Scenarios.''} \emph{Case Studies on
Transport Policy} 15 (March): 101110.
\url{https://doi.org/10.1016/j.cstp.2023.101110}.

\bibitem[\citeproctext]{ref-Pereira2021r5r}
Pereira, Rafael H. M., Marcus Saraiva, Daniel Herszenhut, Carlos Kaue
Vieira Braga, and Matthew Wigginton Conway. 2021. {``R5r: Rapid
Realistic Routing on Multimodal Transport Networks with
r\textsuperscript{5} in r.''} \emph{Findings}, March.
\url{https://doi.org/10.32866/001c.21262}.

\bibitem[\citeproctext]{ref-pietrabissaSchoolAccessCity2023}
Pietrabissa, Giorgio. 2023. {``School Access and City Structure.''}
\url{https://giorgiopietrabissa.github.io/files/school_sorting.pdf}.

\bibitem[\citeproctext]{ref-rongReviewResearchLowcarbon2022}
Rong, Peijun, Mei-Po Kwan, Yaochen Qin, and Zhicheng Zheng. 2022. {``A
Review of Research on Low-Carbon School Trips and Their Implications for
Human-Environment Relationship.''} \emph{Journal of Transport Geography}
99 (February): 103306.
\url{https://doi.org/10.1016/j.jtrangeo.2022.103306}.

\bibitem[\citeproctext]{ref-Rosik_Pulawska-Obiedowska_Goliszek_2021}
Rosik, Piotr, Sabina Puławska-Obiedowska, and Sławomir Goliszek. 2021.
{``Public Transport Accessibility to Upper Secondary Schools Measured by
the Potential Quotient: The Case of Kraków.''} \emph{Moravian
Geographical Reports} 29 (1): 15--26.
\url{https://doi.org/10.2478/mgr-2021-0002}.

\bibitem[\citeproctext]{ref-sageman2022}
Sageman, Joseph. 2022. {``School Closures and Rural Population
Decline.''} \emph{Rural Sociology} 87 (3): 960--92.
\url{https://doi.org/10.1111/ruso.12437}.

\bibitem[\citeproctext]{ref-seasons2014ARCCatalogues2014}
Seasons, Mark. 2014a. {``2014 {ARC} Catalogues. 2014 English Public
{ARC} Inventory.''} 2014.
\url{https://uwaterloo.ca/school-closure-policy-research/research-findings/arc-data/2014-arc-catalogue}.

\bibitem[\citeproctext]{ref-seasonsSchoolClosurePolicy2014}
---------. 2014b. {``School Closure Policy {\textbar} School Closure
Policy Research.''} January 1, 2014.
\url{https://uwaterloo.ca/school-closure-policy-research/exploring-issue/school-closure-policy}.

\bibitem[\citeproctext]{ref-shenLocationCharacteristicsInnercity1998}
Shen, Q. 1998. {``Location Characteristics of Inner-City Neighborhoods
and Employment Accessibility of Low-Wage Workers.''} \emph{Environment
and Planning B: Planning and Design} 25 (3): 345--65.
\url{https://doi.org/10.1068/b250345}.

\bibitem[\citeproctext]{ref-soukhovOccupancyGHGEmissions2022}
Soukhov, Anastasia, and Moataz Mohamed. 2022. {``Occupancy and {GHG}
Emissions: Thresholds for Disruptive Transportation Modes and Emerging
Technologies.''} \emph{Transportation Research Part D: Transport and
Environment} 102 (January): 103127.
\url{https://doi.org/10.1016/j.trd.2021.103127}.

\bibitem[\citeproctext]{ref-soukhovTTS2016RDataSet2023}
Soukhov, Anastasia, and Antonio Páez. 2023. {``{TTS}2016R: A Data Set to
Study Population and Employment Patterns from the 2016 Transportation
Tomorrow Survey in the Greater Golden Horseshoe Area, Ontario,
Canada.''} \emph{Environment and Planning B: Urban Analytics and City
Science}, January, 23998083221146781.
\url{https://doi.org/10.1177/23998083221146781}.

\bibitem[\citeproctext]{ref-soukhovIntroducingSpatialAvailability2023}
Soukhov, Anastasia, Antonio Páez, Christopher D. Higgins, and Moataz
Mohamed. 2023. {``Introducing Spatial Availability, a Singly-Constrained
Measure of Competitive Accessibility \textbar{} {PLOS ONE}.''}
\emph{PLOS ONE}, 1--30.
https://doi.org/\href{https://\%20doi.org/10.1371/journal.pone.0278468}{https://
doi.org/10.1371/journal.pone.0278468}.

\bibitem[\citeproctext]{ref-soukhovMultimodalSpatialAvailability2024}
Soukhov, Anastasia, Javier Tarriño-Ortiz, Julio A. Soria-Lara, and
Antonio Páez. 2024. {``Multimodal Spatial Availability: A
Singly-Constrained Measure of Accessibility Considering Multiple
Modes.''} \emph{{PLOS} {ONE}} 19 (2): e0299077.
\url{https://doi.org/10.1371/journal.pone.0299077}.

\bibitem[\citeproctext]{ref-DMTI_Spatial_2015}
Spatial, DMTI. 2015. {``Building Footprints Region,''} September.
\url{http://geo.scholarsportal.info/\#r/details/_uri@=24}.

\bibitem[\citeproctext]{ref-Teranet_2009}
Teranet. 2009. {``Hamilton Parcel/Land Use Data.''} Dept. of Planning;
Economic Development.

\bibitem[\citeproctext]{ref-weibullAxiomaticApproachMeasurement1976}
Weibull, Jörgen W. 1976. {``An Axiomatic Approach to the Measurement of
Accessibility.''} \emph{Regional Science and Urban Economics} 6 (4):
357--79. \url{https://doi.org/10.1016/0166-0462(76)90031-4}.

\bibitem[\citeproctext]{ref-yumitaSchoolCommutingBarriers2021}
Yumita, FR, MZ Irawan, S Malkhamah, and MIH Kamal. 2021. {``School
Commuting: Barriers, Abilities and Strategies Toward Sustainable Public
Transport Systems in Yogyakarta, Indonesia.''} \emph{{SUSTAINABILITY}}
13 (16). \url{https://doi.org/10.3390/su13169372}.

\end{CSLReferences}

\pagebreak

\section{Appendix}\label{appendix}

\input{TabA1.tex}

\end{document}

### 2024-05-10 paper/Manuscript.qmd
Added declarations to manuscript (upon request by editorial office) and updated lock.file

@@
 
+# Declarations
+
+**Ethical approval**
+Not applicable
+
+**Participate consent**
+Not applicable
+
+**Competing interests**
+The authors declare there are no personal or financial conflicts of interest.
+
+**Authors' contributions**
+A.S., A.P., C.D.H., and M.M. all contributed to the study conception and design. Material preparation, data collection
+and formal analysis were performed by A.S., A.P, and C.D.H.
+All visuals and the first draft of the manuscript was written by A.S.
+A.S., A.P., C.D.H., and M.M.
+commented on earlier versions of the manuscript.
+A.S., A.P., C.D.H., and M.M.
+reviewed the final manuscript.
+
+**Funding declaration**
+This work was supported by the Mobilizing Justice Partnership and the Canada Graduate Scholarships—Doctoral Program from the Social Sciences and Humanities Research Council of Canada (SSHRC).
+
+**Availability of data and materials**
+The manuscript text, code, analysis and data supporting this work is available in the lead author's GitHub repository : https://github.com/soukhova/School-closures-accessibility-impacts
+
 \pagebreak

### 2024-05-10 paper/Manuscript.tex
Added declarations to manuscript (upon request by editorial office) and updated lock.file

@@
 
+\pagebreak
+
 \global\setlength{\Oldarrayrulewidth}{\arrayrulewidth}
@@
 
+\section{Declarations}\label{declarations}
+
+\textbf{Ethical approval} Not applicable
+
+\textbf{Participate consent} Not applicable
+
+\textbf{Competing interests} The authors declare there are no personal
+or financial conflicts of interest.
+
+\textbf{Authors' contributions} A.S., A.P., C.D.H., and M.M. all
+contributed to the study conception and design. Material preparation,
+data collection and formal analysis were performed by A.S., A.P, and
+C.D.H. All visuals and the first draft of the manuscript was written by
+A.S. A.S., A.P., C.D.H., and M.M. commented on earlier versions of the
+manuscript. A.S., A.P., C.D.H., and M.M. reviewed the final manuscript.
+
+\textbf{Funding declaration} This work was supported by the Mobilizing
+Justice Partnership and the Canada Graduate Scholarships---Doctoral
+Program from the Social Sciences and Humanities Research Council of
+Canada (SSHRC).
+
+\textbf{Availability of data and materials} The manuscript text, code,
+analysis and data supporting this work is available in the lead author's
+GitHub repository :
+https://github.com/soukhova/School-closures-accessibility-impacts
+
 \pagebreak

### 2024-11-21 data-prep/01-processing-schools-parcels-census.qmd
Added 2021 schools and census to the database. Started Manuscript changes.

@@
 list_census_regions('CA16') %>%
+   filter(name == c("Hamilton"))
+list_census_regions('CA21') %>%
    filter(name == c("Hamilton")) #from this result we see the code for the Hamilton CMA is 35537
-```
-
-```{r}
-find_census_vectors("prevalence of low income", dataset = "CA16", type = "total", query_type = "keyword", interactive = FALSE)
 ```
@@
 
-Hamilton CMA is larger than the hamilton boundary. Clip the census data and resave:
+Repeat for 2021 incomes:
+```{r}
+cen_data_2021_1 <- get_census(dataset='CA21', regions=list(CMA="35537"),
+                           vectors=c("v_CA21_566", #Median after-tax income in 2020 among recipients ($),
+                                     "v_CA21_966", #Median after-tax income of economic family in 2020 ($)
+                                     "v_CA21_978", #Median after-tax income of 1-parent economic families in 2020 ($)
+                                     "v_CA21_1043", #	Income; Low Income Measures; Prevalence of low income based on the Low-income measure, after tax (LIM-AT) (%); 0 to 17 years (%)
+                                     "v_CA21_32", # population age 5-9
+                                     "v_CA21_50", #population age 10-14
+                                     "v_CA21_71"), #population age 15-19
+                           level='DA', use_cache = FALSE, geo_format = 'sf')
+
+save(cen_data_2021_1, file="data-inputs/Census-data/cen_data_2021_1.RData")
+```
+
+Hamilton CMA includes Burlington and Grimbsy. Filter those out and re-save:
 ```{r}
@@
 
-cen_data_2011_1 <- st_intersection(ham_city_bound |> select(geometry), cen_data_2011_1)
-cen_data_2016_1 <- st_intersection(ham_city_bound |> select(geometry), cen_data_2016_1)
+cen_data_2011_1 <- cen_data_2011_1 |> filter(CD_UID == 3525) |> st_as_sf()
+cen_data_2016_1 <- cen_data_2016_1 |> filter(CD_UID == 3525)|> st_as_sf()
+cen_data_2021_1 <- cen_data_2021_1 |> filter(CD_UID == 3525)|> st_as_sf()
 
@@
 save(cen_data_2016_1, file="data-products/cen_data_2016_hamiltononly.RData")
+save(cen_data_2021_1, file="data-products/cen_data_2021_hamiltononly.RData")
 ```
@@
 # Schools and catchments:
-Available through the hamONdests package (https://github.com/soukhova/hamONdests):
+2011 and 2016 schools available through the hamONdests package (https://github.com/soukhova/hamONdests):
 ```{r}
@@
 
+```{r}
+ALL_SCHOOLS_Elem <- Schools_2011_2016 %>% filter(Level == "Elementary") %>%
+  st_transform(crs = 32617)
+rm("Schools_2011_2016")
+
+SCHOOL_Elem_2016 <- ALL_SCHOOLS_Elem |> filter(Year == "2011 and 2016" | Year == "2016")
+```
+
+Schools (as of 2024) are retrieved from Hamilton Open Data:
+```{r}
+SCHOOL_Elem_2024 <- st_read("data-inputs/Educational_Institutions-2024/Educational_Institutions.shp")
+SCHOOL_Elem_2024 <- SCHOOL_Elem_2024 |> filter(CATEGORY == "Elementary School" | CATEGORY == "Middle School" ) |> mutate(SCHOOL_ID = as.character(SCHOOL_ID)) %>%
+  st_transform(crs = 32617)
+```
+
+Identifying the schools (ID and Names) that are different between 2011-16 and 2024.
+To do so, I will spatially compare 2016 to 2024.
+We want to identify which schools were closed between 2016 and 2024, and which ones opened.
+```{r}
+#creating a 100m buffer around school points
+SCHOOL_Elem_2024_buff <- SCHOOL_Elem_2024 |> st_buffer(100)
+
+SCHOOLS_2024_and_2016 <- st_intersection(SCHOOL_Elem_2024_buff, SCHOOL_Elem_2016) |> select(c("SchoolID","OBJECTID","NAME","Name")) |> mutate(Year = "2016 and 2024")
+```
+
+There are apparently 123 schools that are common between 2024 and 2016. I manually check this list, and we're taking the 2016 names into the 2021 analysis. 
+```{r}
+SCHOOLS_2024_and_2016 |> st_drop_geometry()
+```
+NOTE: 
+- Ancaster Senior became Frank Panabaker Elementary School (South Campus) and Fessenden Elementary School became Frank Panabaker Elementary School (North Campus) in 2019 -- effectively they are the same school in 2019 but under the same 'management' now, so keep names as is (https://en.wikipedia.org/wiki/Frank_Panabaker_Elementary_School)
+
+- Glen Echo, Glen Brae, Sir Isaac Brock, and Elizabeth Bagshaw elementary schools closed -- and "Viola Desmond Elementary School" opened in September 2021 on the site of Glen Brae.
+Let's make Glen Brae as "Removed", and add new school named "Viola Desmond" as SchoolID 156.202021, ELEM 0, urban.dist 7948.096, footprint 3,902 m²
+(https://www.hwdsb.on.ca/sirwilfridlaurier/2021/05/19/viola-desmond-elementary-school-update/)
+
+- Ryerson was renamed "Kanétskare Elementary School" in September 2022, nothing else changed. For this analysis, keep Ryerson name. (https://www.hwdsb.on.ca/kanetskare/renaming/)
+- C.H.
+Bray was renamed "Spring Valley Elementary School" in September 2019, nothing else changed.
+For this analysis, keep C.H.
+Bray name.
+(https://www.hwdsb.on.ca/blog/new-school-names-for-the-schools-on-the-c-h-bray-and-beverly-community-centre-sites/)
+- G.R.
+Allan was expanded and renamed to "Cootes Paradise Elementary School" in 2014.
+For this analysis, keep G.R.
+Allan name.
+(https://www.hwdsb.on.ca/cootesparadise/files/2014/09/Agenda-insert-for-CP-2017-2018.pdf)
+
+Let's join add new columns, "Year_2" and "Status_2" and "footprint2021" corresponding to what happened between 2016-2021 for these 123 common schools.
+```{r}
+test_ALL_SCHOOLS_ELEM <- ALL_SCHOOLS_Elem |> mutate(Year_2 = ifelse(SchoolID %in% SCHOOLS_2024_and_2016$SchoolID, "2016 and 2021", NA),
+                           Status_2 = ifelse(!is.na(Year_2), "NoChange", NA),
+                           Status_2 = ifelse(SchoolID == "156", "Removed", Status_2),
+                           footprint2021 = ifelse(SchoolID =="156", 3902, footprint2016))
+
+test <- test_ALL_SCHOOLS_ELEM |> filter (SchoolID == "156") |> #takes the geometry and urban.dist of Glen Brae
+  mutate(SchoolID = "156.2021",
+         Name = "Viola Desmond",
+         ELEM = "ELEM",
+         MID = "MID",
+         Year_2 = "2021",
+         Status_2 = "New",
+         footprint2021 =3902)
+
+test[,c(3,4,7,8,9,10,11,12,18,19)] <- NA
+
+test_ALL_SCHOOLS_ELEM <- rbind(test_ALL_SCHOOLS_ELEM, test)
+```
+
+Now, what schools are NOT in common? Check each one.
+```{r}
+full_join(SCHOOL_Elem_2024,
+          SCHOOLS_2024_and_2016|> st_drop_geometry(), by="OBJECTID") |> filter(is.na(Year)) |> st_drop_geometry()
+```
+Update the schools that remained the same into 2021:
+```{r}
+test_ALL_SCHOOLS_ELEM <- test_ALL_SCHOOLS_ELEM |> 
+  mutate(Year_2 = ifelse(SchoolID %in% c("16","176","4362"), "2016 and 2021", Year_2),
+         Status_2 = ifelse(SchoolID %in% c("16","176","4362"), "NoChange", Status_2),
+         footprint2021 = ifelse(SchoolID %in% c("16","176","4362"), footprint2016, footprint2021))
+```
+
+Update the schools that changed locations or opened up between 2016-2021:
+```{r}
+test_ALL_SCHOOLS_ELEM <- test_ALL_SCHOOLS_ELEM |> 
+  mutate(Year_2 = ifelse(SchoolID %in% c("3564"), "2021", Year_2),
+         Status_2 = ifelse(SchoolID %in% c("3564"), "Moved", Status_2))
+
+test<- SCHOOL_Elem_2024 |> filter(NAME == "Our Lady of the Assumption Elementary School") |> select("geometry") |> st_as_sf()
+test <- test_ALL_SCHOOLS_ELEM |> filter (SchoolID == "3564") |>
+  mutate(SchoolID = "3564.2021",
+         Name = "Our Lady of the Assumption",
+         Year_2 = "2021",
+         Status_2 = "New",
+         MID = "MID",
+         footprint2021 = 4093,
+         geometry = test$geometry)
+
+test[,c(3,4,7,8,9,10,11,12,18,19)] <- NA
+
+test1<- SCHOOL_Elem_2024 |> filter(NAME == "Rockton Elementary School") |> select("geometry") |> st_as_sf()
+test1 <- test_ALL_SCHOOLS_ELEM |> filter (SchoolID == "10") |>
+  mutate(SchoolID = "374",
+         Name = "Rockton Elementary School",
+         Year_2 = "2021",
+         Status_2 = "New",
+         footprint2021 = (30.29*19.42 + 20.5*100 + 29.43*53.6),
+         geometry = test1$geometry)
+test1[,c(3,4,7,8,9,10,11,12,18,19)] <- NA
+
+test2<- SCHOOL_Elem_2024 |> filter(NAME == "South Meadow Elementary School") |> select("geometry") |> st_as_sf()
+test2 <- test_ALL_SCHOOLS_ELEM |> filter (SchoolID == "10") |>
+  mutate(SchoolID = "273",
+         Name = "South Meadow Elementary School",
+         Year_2 = "2021",
+         Status_2 = "New",
+         footprint2021 = 3639,
+         geometry = test2$geometry)
+test2[,c(3,4,7,8,9,10,11,12,18,19)] <- NA
+
+test3<- SCHOOL_Elem_2024 |> filter(NAME == "Tiffany Hills Elementary School") |> select("geometry") |> st_as_sf()
+test3 <- test_ALL_SCHOOLS_ELEM |> filter (SchoolID == "10") |>
+  mutate(SchoolID = "425",
+         Name = "Tiffany Hills Elementary School",
+         Year_2 = "2021",
+         Status_2 = "New",
+         footprint2021 = 3124,
+         geometry = test3$geometry)
+test3[,c(3,4,7,8,9,10,11,12,18,19)] <- NA
+
+test4<- SCHOOL_Elem_2024 |> filter(NAME == "Our Lady of Hope Elementary School") |> select("geometry") |> st_as_sf()
+test4 <- test_ALL_SCHOOLS_ELEM |> filter (SchoolID == "10") |>
+  mutate(SchoolID = "647",
+         Name = "Our Lady of Hope Elementary School",
+         Year_2 = "2021",
+         Status_2 = "New",
+         footprint2021 = 3157,
+         geometry = test4$geometry)
+test4[,c(3,4,7,8,9,10,11,12,18,19)] <- NA
+
+test5<- SCHOOL_Elem_2024 |> filter(NAME == "Shannen Koostachin Elementary School") |> select("geometry") |> st_as_sf()
+test5 <- test_ALL_SCHOOLS_ELEM |> filter (SchoolID == "10") |>
+  mutate(SchoolID = "416",
+         Name = "Shannen Koostachin Elementary School",
+         Year_2 = "2021",
+         Status_2 = "New",
+         footprint2021 = (29*119 + 24*20),
+         geometry = test5$geometry)
+test5[,c(3,4,7,8,9,10,11,12,18,19)] <- NA
+
+test_ALL_SCHOOLS_ELEM <- rbind(test_ALL_SCHOOLS_ELEM, test, test1, test2, test3, test4, test5)
+```
+
+Update schools that were removed:
+```{r}
+test_ALL_SCHOOLS_ELEM <- test_ALL_SCHOOLS_ELEM |> 
+  mutate(Year_2 = ifelse(SchoolID %in% c("3564","32","106","344","407",
+                                         "281","272",
+                                         "352","162", "405", "117"),
+                         "2016", Year_2),
+         Status_2 = ifelse(SchoolID %in% c("3564","32","106","344","407",
+                                         "281","272",
+                                         "352","162", "405", "117"),
+                           "Removed", Status_2),
+         footprint2021 = ifelse(SchoolID %in% c("3564","32","106","344","407",
+                                         "281","272",
+                                         "352","162", "405", "117"),
+                                NA, footprint2021))
+```
+(Our lady of the assumption closed its old location but moved to a new location (bigger) https://www.thespec.com/news/rebuilt-our-lady-of-the-assumption-catholic-elementary-school-opens/article_599fb9fe-fa59-5f85-bb1e-4d618e621bfb.html)
+(32,106,344,407 -- replaced with Rockton school https://12ft.io/proxy?q=https%3A%2F%2Fwww.thespec.com%2Fnews%2Fprescribed-process-through-the-ministry-4-former-flamborough-area-school-buildings-up-for-sale%2Farticle_f7983aeb-5f6c-5717-b875-fd3a13c5d5e1.html)
+(South Meadow and associated -- https://www.hwdsb.on.ca/blog/stoney-creeks-new-memorial-school-to-be-renamed-south-meadow/)
+(Tiffany falls opens -- https://www.hwdsb.on.ca/blog/hwdsb-to-open-tiffany-hills-school-on-january-9-2017-to-address-ancaster-growth/)
+(R.L. Hyslop is closed -- https://www.hwdsb.on.ca/wp-content/uploads/2021/07/Notice-for-Public-Information-R.L.-Hyslop.pdf)
+(Glen echo is closed -- https://www.thespec.com/news/hamilton-public-boards-sale-of-glen-echo-school-a-punch-in-the-gut/article_57915796-dab5-53af-a936-89d43aa427b5.html)
+(Sir Isaac Brook is closed -- https://www.thespec.com/news/new-policy-puts-sir-isaac-brock-school-name-in-crosshairs/article_c3ad138f-1e0f-5929-bdb9-c3d862ec5cd6.html)
+
+Update the urban.dist for the moved / new schools in 2021:
+```{r}
+# add a 'proxy' for age of construction.
+Assuming schools constructed closer to the CBD (the lat-long coordinate of "King St W and James St S: (43.256684, -79.869039)) is older and further away is newer.
+urban_pt <- st_sfc(st_point(c(-79.869039, 43.256684)), crs=4326)
+
+NEW_MOVED_SCHOOLS_2021_WSG84 <- st_transform(test_ALL_SCHOOLS_ELEM |> filter(Year_2 == "2021" & (Status_2 == "New" | Status_2 == "Moved")),crs=4326)
+NEW_MOVED_SCHOOLS_2021_WSG84$urban.dist <- st_distance(NEW_MOVED_SCHOOLS_2021_WSG84, y = urban_pt)
+```
+
+Now estimate OTGC2021 for these 7 locations. First load the models and reformat the dataframes:
+```{r}
+load("data-products/OTG_fit_DSB.rda") #created Nov. 19.2024 in the hamONdests package
+load("data-products/OTG_fit_CDSB.rda")
+
+#Create dummy variables corresponding to Grades (for Public) and for Type (for Catholic)
+NEW_MOVED_SCHOOLS_2021_WSG84 <- NEW_MOVED_SCHOOLS_2021_WSG84 %>%
+  mutate(Grades.JKto5_6 = ifelse((ELEM == "ELEM" & MID == "0" & HIGH == "0"), 1, 0),
+         Grades.JKto8 = ifelse((ELEM == "ELEM" & MID == "MID" & HIGH == "0"), 1, 0),
+         Grades.6to8 = ifelse((ELEM == "0" & MID == "MID" & HIGH == "0"), 1, 0),
+         Grades.9to12 = ifelse((ELEM == "0" & MID == "0" & HIGH == "HIGH"), 1, 0),
+         Type.Elementary = ifelse(Level == "Elementary", 1, 0),
+         Type.Secondary = ifelse(Level == "High", 1, 0))
+
+NEW_MOVED_SCHOOLS_2021_WSG84
+```
+
+Now predict the public schools:
+```{r}
+DSB_OTGC <- NEW_MOVED_SCHOOLS_2021_WSG84 |> filter(System == "Public")
+
+#add a field with index numbers for merging predicted OTGC 
+DSB_OTGC$rowID <- 1:nrow(DSB_OTGC)
+
+t <- as.data.frame(DSB_OTGC) %>%
+  transmute(footprint = footprint2021,Grades.JKto5_6,Grades.6to8,Grades.JKto8,Grades.9to12,urban.dist)
+t$urban.dist <- as.matrix(t$urban.dist)
+
+t2 <- predict(OTG_fit_DSB, newdata = t)
+t2 <- data.frame(OTGC2021 = exp(t2))
+t2$rowID <- 1:nrow(t2)
+
+DSB_OTGC <- merge(DSB_OTGC, t2, by= "rowID", all.x=TRUE ) |> select(c("SchoolID","OTGC2021")) |> st_drop_geometry()
+```
+
+Predict the 1 catholic school:
+```{r}
+CDSB_OTGC <- NEW_MOVED_SCHOOLS_2021_WSG84 |> filter(System == "Catholic")
+
+#add a field with index numbers for merging predicted OTGC 
+CDSB_OTGC$rowID <- 1:nrow(CDSB_OTGC)
+
+t <- as.data.frame(CDSB_OTGC) %>%
+  transmute(footprint = footprint2021,Type.Elementary,Type.Secondary, urban.dist)
+t$urban.dist <- as.matrix(t$urban.dist)
+
+t2 <- predict(OTG_fit_CDSB, newdata = t)
+t2 <- data.frame(OTGC2021 = exp(t2))
+t2$rowID <- 1:nrow(t2)
+
+CDSB_OTGC <- merge(CDSB_OTGC, t2, by= "rowID", all.x=TRUE ) |> select(c("SchoolID","OTGC2021")) |> st_drop_geometry()
+```
+
+Add the 2021 OTGCs to the schools database:
+```{r}
+SCHOOLS <- left_join(test_ALL_SCHOOLS_ELEM, rbind(DSB_OTGC,CDSB_OTGC), by="SchoolID")
+```
+Also, to the OTGC2021 column, for schools that existed in 2016 and exist in 2021, copy the OTGC of 2016. 
+```{r}
+SCHOOLS <- SCHOOLS |> mutate(OTGC2021 = ifelse(Year_2 == "2016 and 2021", OTGC2016, OTGC2021))
+
+save(SCHOOLS,file="data-products/SCHOOLS.Rdata") #for this paper, the schools under analysi
+```
+
 # Parcels
-
 Read shapefile as "sf" and drop all non-residential parcels:
@@
 load(file="data-inputs/Census-data/cen_data_2016_1.RData")
+load(file="data-inputs/Census-data/cen_data_2021_1.RData")
+
 RED_LU_2011 <- st_read("data-inputs/Parcels/PED_LANDUSE_2011/PED_LANDUSE.shp") |> 
@@
 RED_LU_2016 <- st_make_valid(RED_LU_2016)
+
+RED_LU_2020 <- st_read("data-inputs/Parcels/PED_LANDUSE_2020/PED_LANDUSE.shp") |>
+  mutate(TYPE = ifelse(LUC1_DESCR == "Residential:Detached House" | 
+                         LUC1_DESCR == "Residential:Row/Town House"| 
+                         LUC1_DESCR == "Residential:Semi-Detached House" |
+                         LUC1_DESCR == "Residential:Apartment (7 or more units)"|
+                         LUC1_DESCR == "Residential:Multiplex Dwelling (6 units or less)",
+                       "Residential", 0)) |>
+  filter(TYPE != 0) |> #i.e. filter out non-residential
+  st_transform(st_crs(cen_data_2021_1))
+RED_LU_2020 <- st_make_valid(RED_LU_2020)
+RED_LU_2020 <- RED_LU_2020 |> mutate(ID = row_number())
 ```
@@
 Transfer the GeoUID from the DAs (I.e., their DA IDs) that intersect with the residential land use parcel centroids to the RED_LU sf object.
-(we use 'over' here instead of st_intersection because it is way faster)
+(we use 'over' here instead of st_intersection because it is way faster).
+First assign 2016 DA IDs to the centroids of 2016 parcels:
+```{r}
+RED_LU_centroids_2016 <- sp::over(as_Spatial(st_centroid(RED_LU_2016)),as_Spatial(cen_data_2016_1 |> select(GeoUID)))
+
+RED_LU_centroids_2016 <- cbind(RED_LU_centroids_2016,st_centroid(RED_LU_2016))
+```
+
+Next assign 2016 DA IDs to the centroids of 2011 parcels:
 ```{r}
@@
 
+Next assign 2016 DA IDs to the centroids of 2021 parcels:
 ```{r}
-RED_LU_centroids_2016 <- sp::over(as_Spatial(st_centroid(RED_LU_2016)),as_Spatial(cen_data_2016_1 |> select(GeoUID)))
+RED_LU_centroids_2021 <- sp::over(as_Spatial(st_centroid(RED_LU_2020)),
+                                  as_Spatial(cen_data_2021_1 |> dplyr::select(GeoUID) |> st_sf()))
 
-RED_LU_centroids_2016<- cbind(RED_LU_centroids_2016,st_centroid(RED_LU_2016))
+RED_LU_centroids_2021_w2016ID <- sp::over(as_Spatial(st_centroid(RED_LU_2020)),
+                                          as_Spatial(cen_data_2016_1 |> dplyr::select(GeoUID) |>
+                                                       rename("GeoUID_2016" = "GeoUID")))
+
+RED_LU_centroids_2021 <- cbind(RED_LU_centroids_2021_w2016ID,
+                              RED_LU_centroids_2021,
+                              st_centroid(RED_LU_2020))
 ```
 
-Now transfer the TAZ ids to the RED centroids -- this is so we can pull the proportion of motor/non-motorized trips. 
+As another step, we need to transfer the TAZ ids to the RED centroids -- this is so we can pull the proportion of motor/non-motorized trips. First, load in TAZs:
 ```{r}
@@
 ```
-Now add the TAZ IDs column to the residential parcel id
+
+Now add the TAZ IDs column to the residential parcel id for 2011, 2016, and 2021
 ```{r}
@@
 RED_LU_centroids_2016<- cbind(RED_LU_centroids_2016,cols_TAZIDs)
+
+cols_TAZIDs <- sp::over(as_Spatial(st_centroid(RED_LU_2020)),
+                              as_Spatial(ggh_taz |> select(GTA06) |> rename("GTA06_Orig" = "GTA06")))
+
+RED_LU_centroids_2021<- cbind(RED_LU_centroids_2021, cols_TAZIDs)
+
 rm(cols_TAZIDs)
@@
 save(RED_LU_centroids_2016, file="data-inputs/Parcels/RED_LU_centroids_2016.RData")
+save(RED_LU_centroids_2021, file="data-inputs/Parcels/RED_LU_centroids_2021.RData")
 ```

### 2024-11-21 data-prep/02-travel-times-ODs.qmd
Added 2021 schools and census to the database. Started Manuscript changes.

@@
 ```{r}
-library(hamONdests)
-data(Schools_2011_2016)
-data(School_Catchments_2011_2016)
+# library(hamONdests)
+# data(SCHOOLS)
+# data(School_Catchments_2011_2016)
+load(file="data-products/SCHOOLS.Rdata") #for this paper, the schools under analysis
 load(file = "data-inputs/Parcels/RED_LU_centroids_2011.Rdata")
 load(file = "data-inputs/Parcels/RED_LU_centroids_2016.Rdata")
+load(file = "data-inputs/Parcels/RED_LU_centroids_2021.Rdata")
 ```
@@
 ```{r}
-Dest_all_schools_elem_2011 <- cbind(as.data.frame(Schools_2011_2016), st_coordinates(st_transform(Schools_2011_2016, crs = 4326))) %>% filter(Level == "Elementary" & (Year == "2011" | Year == "2011 and 2016")) %>%
+Dest_all_schools_elem_2011 <- cbind(as.data.frame(SCHOOLS), st_coordinates(st_transform(SCHOOLS, crs = 4326))) %>% filter(Level == "Elementary" & (Year == "2011" | Year == "2011 and 2016")) %>%
   rename(lon = "X", lat = "Y", id = "SchoolID") %>%
@@
 
-Dest_all_schools_elem_2016 <- cbind(as.data.frame(Schools_2011_2016), st_coordinates(st_transform(Schools_2011_2016, crs = 4326))) %>% filter(Level == "Elementary" & (Year == "2016" | Year == "2011 and 2016")) %>%
+Dest_all_schools_elem_2016 <- cbind(as.data.frame(SCHOOLS), st_coordinates(st_transform(SCHOOLS, crs = 4326))) %>% filter(Level == "Elementary" & (Year == "2016" | Year == "2011 and 2016")) %>%
+  rename(lon = "X", lat = "Y", id = "SchoolID") %>%
+  select(id, lon, lat)
+
+Dest_all_schools_elem_2021 <- cbind(as.data.frame(SCHOOLS), st_coordinates(st_transform(SCHOOLS, crs = 4326))) %>% filter(Level == "Elementary" & (Year_2 == "2021" | Year_2 == "2016 and 2021")) %>%
   rename(lon = "X", lat = "Y", id = "SchoolID") %>%
@@
   rename(lon = X, lat = Y, id = ID) %>%
-  select(id, lon, lat)
+  dplyr::select(c(id, lon, lat))
 
@@
                        st_coordinates(st_transform(RED_LU_centroids_2016 |> st_as_sf(), crs = 4326))) %>%
+  rename(lon = X, lat = Y, id = ID) %>%
+  select(id, lon, lat)
+
+Orig_all_2021 <- cbind(RED_LU_centroids_2021,
+                       st_coordinates(st_transform(RED_LU_centroids_2021 |> st_as_sf(), crs = 4326))) %>%
   rename(lon = X, lat = Y, id = ID) %>%
@@
 Orig_all_2016_4 <- Orig_all_2016[110711:139467,]
+```
+
+```{r}
+Orig_all_2021_1 <- Orig_all_2021[1:36903,]
+Orig_all_2021_2 <- Orig_all_2021[36904:73807,]
+Orig_all_2021_3 <- Orig_all_2021[73808:110710,]
+Orig_all_2021_4 <- Orig_all_2021[110711:143890,]
 ```
@@
 ```{r build graph, include = FALSE}
-r5_HAM <- setup_r5(data_path = r5_path, verbose = TRUE, overwrite = TRUE, temp_dir = TRUE)
+r5_HAM <- setup_r5(data_path = r5_path, verbose = TRUE,  temp_dir = TRUE)#, overwrite = TRUE)
 ```
@@
 
-2011 Car travel times:
+## 2011 Car travel times:
 ```{r}
@@
 
-2016 car travel times:
+## 2016 car travel times:
 ```{r}
@@
 
-2011 walk travel times:
+## 2021 car travel times:
+```{r}
+# set up batching according to how many origin rows to process at one time
+chunksize = 50 
+num_chunks = ceiling(nrow(Orig_all_2021_1)/chunksize)
+
+# create origin-destination pairs
+origins_chunks <- as.disk.frame(Orig_all_2021_1,
+                          outdir = "data-inputs/Travel-times/Processing/Orig",
+                          nchunks = num_chunks,
+                          overwrite = TRUE)
+
+start.time <- Sys.time()
+pb <- txtProgressBar(0, num_chunks, style = 3)
+
+for (i in 1:num_chunks){
+  Orig_chunk <- get_chunk(origins_chunks, i)
+  ttm_chunk <- travel_time_matrix(r5r_core = r5_HAM,
+                          origins = Orig_chunk,
+                          destinations = Dest_all_schools_elem_2021,
+                          mode = c("CAR"), 
+                          departure_datetime = as.POSIXct(strptime("2021-03-01 08:00:00", "%Y-%m-%d %H:%M:%S", tz = "EST5EDT")),
+                          max_trip_duration = 90)
+
+  # export output as disk.frame
+  ifelse(i == 1, output_df <- as.disk.frame(ttm_chunk,
+                                            nchunks = 1,
+                                            outdir = "data-inputs/Travel-Times/Processing/output_ttm",
+                                            compress = 50,
+                                            overwrite = TRUE),
+         add_chunk(output_df, ttm_chunk, chunk_id = i))
+  setTxtProgressBar(pb, i)
+}
+end.time <- Sys.time()
+print(paste0("OD matrix calculation took ", round(difftime(end.time, start.time, units = "mins"), digits = 2), " minutes..."))
+
+output_OD_car1 <- as.data.frame(output_df)
+
+save("output_OD_car1", file = "data-products/OD_tt_car_2021_1.Rdata")
+rm("output_OD_car1")
+```
+
+```{r}
+# set up batching according to how many origin rows to process at one time
+chunksize = 50 
+num_chunks = ceiling(nrow(Orig_all_2021_2)/chunksize)
+
+# create origin-destination pairs
+origins_chunks <- as.disk.frame(Orig_all_2021_2,
+                          outdir = "data-inputs/Travel-times/Processing/Orig",
+                          nchunks = num_chunks,
+                          overwrite = TRUE)
+
+start.time <- Sys.time()
+pb <- txtProgressBar(0, num_chunks, style = 3)
+
+for (i in 1:num_chunks){
+  Orig_chunk <- get_chunk(origins_chunks, i)
+  ttm_chunk <- travel_time_matrix(r5r_core = r5_HAM,
+                          origins = Orig_chunk,
+                          destinations = Dest_all_schools_elem_2021,
+                          mode = c("CAR"), 
+                          departure_datetime = as.POSIXct(strptime("2021-03-01 08:00:00", "%Y-%m-%d %H:%M:%S", tz = "EST5EDT")),
+                          max_trip_duration = 90)
+
+  # export output as disk.frame
+  ifelse(i == 1, output_df <- as.disk.frame(ttm_chunk,
+                                            nchunks = 1,
+                                            outdir = "data-inputs/Travel-Times/Processing/output_ttm",
+                                            compress = 50,
+                                            overwrite = TRUE),
+         add_chunk(output_df, ttm_chunk, chunk_id = i))
+  setTxtProgressBar(pb, i)
+}
+end.time <- Sys.time()
+print(paste0("OD matrix calculation took ", round(difftime(end.time, start.time, units = "mins"), digits = 2), " minutes..."))
+
+output_OD_car2 <- as.data.frame(output_df)
+
+save("output_OD_car2", file = "data-products/OD_tt_car_2021_2.Rdata")
+rm("output_OD_car2")
+```
+
+```{r}
+# set up batching according to how many origin rows to process at one time
+chunksize = 50 
+num_chunks = ceiling(nrow(Orig_all_2021_3)/chunksize)
+
+# create origin-destination pairs
+origins_chunks <- as.disk.frame(Orig_all_2021_3,
+                          outdir = "data-inputs/Travel-times/Processing/Orig",
+                          nchunks = num_chunks,
+                          overwrite = TRUE)
+
+start.time <- Sys.time()
+pb <- txtProgressBar(0, num_chunks, style = 3)
+
+for (i in 1:num_chunks){
+  Orig_chunk <- get_chunk(origins_chunks, i)
+  ttm_chunk <- travel_time_matrix(r5r_core = r5_HAM,
+                          origins = Orig_chunk,
+                          destinations = Dest_all_schools_elem_2021,
+                          mode = c("CAR"), 
+                          departure_datetime = as.POSIXct(strptime("2021-03-01 08:00:00", "%Y-%m-%d %H:%M:%S", tz = "EST5EDT")),
+                          max_trip_duration = 90)
+
+  # export output as disk.frame
+  ifelse(i == 1, output_df <- as.disk.frame(ttm_chunk,
+                                            nchunks = 1,
+                                            outdir = "data-inputs/Travel-Times/Processing/output_ttm",
+                                            compress = 50,
+                                            overwrite = TRUE),
+         add_chunk(output_df, ttm_chunk, chunk_id = i))
+  setTxtProgressBar(pb, i)
+}
+end.time <- Sys.time()
+print(paste0("OD matrix calculation took ", round(difftime(end.time, start.time, units = "mins"), digits = 2), " minutes..."))
+
+output_OD_car3 <- as.data.frame(output_df)
+
+save("output_OD_car3", file = "data-products/OD_tt_car_2021_3.Rdata")
+rm("output_OD_car3")
+```
+
+```{r}
+# set up batching according to how many origin rows to process at one time
+chunksize = 50 
+num_chunks = ceiling(nrow(Orig_all_2021_4)/chunksize)
+
+# create origin-destination pairs
+origins_chunks <- as.disk.frame(Orig_all_2021_4,
+                          outdir = "data-inputs/Travel-times/Processing/Orig",
+                          nchunks = num_chunks,
+                          overwrite = TRUE)
+
+start.time <- Sys.time()
+pb <- txtProgressBar(0, num_chunks, style = 3)
+
+for (i in 1:num_chunks){
+  Orig_chunk <- get_chunk(origins_chunks, i)
+  ttm_chunk <- travel_time_matrix(r5r_core = r5_HAM,
+                          origins = Orig_chunk,
+                          destinations = Dest_all_schools_elem_2021,
+                          mode = c("CAR"), 
+                          departure_datetime = as.POSIXct(strptime("2021-03-01 08:00:00", "%Y-%m-%d %H:%M:%S", tz = "EST5EDT")),
+                          max_trip_duration = 90)
+
+  # export output as disk.frame
+  ifelse(i == 1, output_df <- as.disk.frame(ttm_chunk,
+                                            nchunks = 1,
+                                            outdir = "data-inputs/Travel-Times/Processing/output_ttm",
+                                            compress = 50,
+                                            overwrite = TRUE),
+         add_chunk(output_df, ttm_chunk, chunk_id = i))
+  setTxtProgressBar(pb, i)
+}
+end.time <- Sys.time()
+print(paste0("OD matrix calculation took ", round(difftime(end.time, start.time, units = "mins"), digits = 2), " minutes..."))
+
+output_OD_car4 <- as.data.frame(output_df)
+
+save("output_OD_car4", file = "data-products/OD_tt_car_2021_4.Rdata")
+rm("output_OD_car4")
+```
+
+## 2011 walk travel times:
 ```{r}
@@
 
-2016 walk travel times:
+## 2016 walk travel times:
 ```{r}
@@
 
+## 2021 walk travel times:
+```{r}
+# set up batching according to how many origin rows to process at one time
+chunksize = 50 
+num_chunks = ceiling(nrow(Orig_all_2021_1)/chunksize)
+
+# create origin-destination pairs
+origins_chunks <- as.disk.frame(Orig_all_2021_1,
+                          outdir = "data-inputs/Travel-times/Processing/Orig",
+                          nchunks = num_chunks,
+                          overwrite = TRUE)
+
+start.time <- Sys.time()
+pb <- txtProgressBar(0, num_chunks, style = 3)
+
+for (i in 1:num_chunks){
+  Orig_chunk <- get_chunk(origins_chunks, i)
+  ttm_chunk <- travel_time_matrix(r5r_core = r5_HAM,
+                          origins = Orig_chunk,
+                          destinations = Dest_all_schools_elem_2021,
+                          mode = c("WALK"),
+                          departure_datetime = as.POSIXct(strptime("2021-03-01 08:00:00", "%Y-%m-%d %H:%M:%S", tz = "EST5EDT")),
+                          walk_speed = 3.6,
+                          max_trip_duration = 30)
+
+  # export output as disk.frame
+  ifelse(i == 1, output_df <- as.disk.frame(ttm_chunk,
+                                            nchunks = 1,
+                                            outdir = "data-inputs/Travel-Times/Processing/output_ttm",
+                                            compress = 50,
+                                            overwrite = TRUE),
+         add_chunk(output_df, ttm_chunk, chunk_id = i))
+  setTxtProgressBar(pb, i)
+}
+end.time <- Sys.time()
+print(paste0("OD matrix calculation took ", round(difftime(end.time, start.time, units = "mins"), digits = 2), " minutes..."))
+
+output_OD_walk_elem1 <- as.data.frame(output_df)
+
+save("output_OD_walk_elem1", file = "data-products/OD_tt_walk_2021_1.Rdata")
+rm("output_OD_walk_elem1")
+
+# set up batching according to how many origin rows to process at one time
+chunksize = 50 
+num_chunks = ceiling(nrow(Orig_all_2021_2)/chunksize)
+
+# create origin-destination pairs
+origins_chunks <- as.disk.frame(Orig_all_2021_2,
+                          outdir = "data-inputs/Travel-times/Processing/Orig",
+                          nchunks = num_chunks,
+                          overwrite = TRUE)
+
+start.time <- Sys.time()
+pb <- txtProgressBar(0, num_chunks, style = 3)
+
+for (i in 1:num_chunks){
+  Orig_chunk <- get_chunk(origins_chunks, i)
+  ttm_chunk <- travel_time_matrix(r5r_core = r5_HAM,
+                          origins = Orig_chunk,
+                          destinations = Dest_all_schools_elem_2021,
+                          mode = c("WALK"),
+                          departure_datetime = as.POSIXct(strptime("2021-03-01 08:00:00", "%Y-%m-%d %H:%M:%S", tz = "EST5EDT")),
+                          walk_speed = 3.6,
+                          max_trip_duration = 30)
+
+  # export output as disk.frame
+  ifelse(i == 1, output_df <- as.disk.frame(ttm_chunk,
+                                            nchunks = 1,
+                                            outdir = "data-inputs/Travel-Times/Processing/output_ttm",
+                                            compress = 50,
+                                            overwrite = TRUE),
+         add_chunk(output_df, ttm_chunk, chunk_id = i))
+  setTxtProgressBar(pb, i)
+}
+end.time <- Sys.time()
+print(paste0("OD matrix calculation took ", round(difftime(end.time, start.time, units = "mins"), digits = 2), " minutes..."))
+
+output_OD_walk_elem2 <- as.data.frame(output_df)
+
+save("output_OD_walk_elem2", file = "data-products/OD_tt_walk_2021_2.Rdata")
+rm("output_OD_walk_elem2")
+
+# set up batching according to how many origin rows to process at one time
+chunksize = 50 
+num_chunks = ceiling(nrow(Orig_all_2021_3)/chunksize)
+
+# create origin-destination pairs
+origins_chunks <- as.disk.frame(Orig_all_2021_3,
+                          outdir = "data-inputs/Travel-times/Processing/Orig",
+                          nchunks = num_chunks,
+                          overwrite = TRUE)
+
+start.time <- Sys.time()
+pb <- txtProgressBar(0, num_chunks, style = 3)
+
+for (i in 1:num_chunks){
+  Orig_chunk <- get_chunk(origins_chunks, i)
+  ttm_chunk <- travel_time_matrix(r5r_core = r5_HAM,
+                          origins = Orig_chunk,
+                          destinations = Dest_all_schools_elem_2021,
+                          mode = c("WALK"),
+                          departure_datetime = as.POSIXct(strptime("2021-03-01 08:00:00", "%Y-%m-%d %H:%M:%S", tz = "EST5EDT")),
+                          walk_speed = 3.6,
+                          max_trip_duration = 30)
+
+  # export output as disk.frame
+  ifelse(i == 1, output_df <- as.disk.frame(ttm_chunk,
+                                            nchunks = 1,
+                                            outdir = "data-inputs/Travel-Times/Processing/output_ttm",
+                                            compress = 50,
+                                            overwrite = TRUE),
+         add_chunk(output_df, ttm_chunk, chunk_id = i))
+  setTxtProgressBar(pb, i)
+}
+end.time <- Sys.time()
+print(paste0("OD matrix calculation took ", round(difftime(end.time, start.time, units = "mins"), digits = 2), " minutes..."))
+
+output_OD_walk_elem3 <- as.data.frame(output_df)
+
+save("output_OD_walk_elem3", file = "data-products/OD_tt_walk_2021_3.Rdata")
+rm("output_OD_walk_elem3")
+
+# set up batching according to how many origin rows to process at one time
+chunksize = 50 
+num_chunks = ceiling(nrow(Orig_all_2021_4)/chunksize)
+
+# create origin-destination pairs
+origins_chunks <- as.disk.frame(Orig_all_2021_4,
+                          outdir = "data-inputs/Travel-times/Processing/Orig",
+                          nchunks = num_chunks,
+                          overwrite = TRUE)
+
+start.time <- Sys.time()
+pb <- txtProgressBar(0, num_chunks, style = 3)
+
+for (i in 1:num_chunks){
+  Orig_chunk <- get_chunk(origins_chunks, i)
+  ttm_chunk <- travel_time_matrix(r5r_core = r5_HAM,
+                          origins = Orig_chunk,
+                          destinations = Dest_all_schools_elem_2021,
+                          mode = c("WALK"),
+                          departure_datetime = as.POSIXct(strptime("2021-03-01 08:00:00", "%Y-%m-%d %H:%M:%S", tz = "EST5EDT")),
+                          walk_speed = 3.6,
+                          max_trip_duration = 30)
+
+  # export output as disk.frame
+  ifelse(i == 1, output_df <- as.disk.frame(ttm_chunk,
+                                            nchunks = 1,
+                                            outdir = "data-inputs/Travel-Times/Processing/output_ttm",
+                                            compress = 50,
+                                            overwrite = TRUE),
+         add_chunk(output_df, ttm_chunk, chunk_id = i))
+  setTxtProgressBar(pb, i)
+}
+end.time <- Sys.time()
+print(paste0("OD matrix calculation took ", round(difftime(end.time, start.time, units = "mins"), digits = 2), " minutes..."))
+
+output_OD_walk_elem4 <- as.data.frame(output_df)
+
+save("output_OD_walk_elem4", file = "data-products/OD_tt_walk_2021_4.Rdata")
+rm("output_OD_walk_elem4")
+```
+
 ```{r}

### 2024-11-21 data-prep/05-Exploring-avail_2021.qmd
Added 2021 schools and census to the database. Started Manuscript changes.

---
title: "05-Exploring-avail_2021"
format: html
---
library:
```{r}
#install.packages("sf")
library(dplyr)
library(sf)
# if(!require(remotes)){
#     install.packages("remotes")
#     library(remotes)
# }
#remotes::install_github("soukhova/HamONdests")
library(hamONdests)
```

```{r sp_avail-function}
#defining the spatial availability function, this prints all the intermediates
sp_avail_detailed <- function(x, o_id, d_id, pop, opp, pop_type, r, f, alpha = 1){

  o_id <- rlang::enquo(o_id)
  d_id <- rlang::enquo(d_id)
  pop <- rlang::enquo(pop)
  opp <- rlang::enquo(opp)
  pop_type <- rlang::enquo(pop_type)
  r <- rlang::enquo(r)
  f <- rlang::enquo(f)

  sum_pop <- x %>%
    dplyr::distinct(!!o_id, !!pop_type,
                    .keep_all = TRUE) %>%
    dplyr::mutate(sum_pop = !!r*(!!pop)^alpha) %>%
    dplyr::pull(sum_pop) %>%
    sum()

  f_p <- dplyr::pull(x, !!r) * dplyr::pull(x, !!pop)^alpha / sum_pop

  sum_impedance <- x %>%
    dplyr::group_by(!!d_id) %>%
    dplyr::summarize(sum_impedance = sum(!!f))

  x <- x %>%
    dplyr::left_join(sum_impedance,
                     by = rlang::as_name(d_id))

  f_c <- dplyr::pull(x, !!f) / x$sum_impedance

  x$f_c <- f_c
  x$f_p <- f_p

  sum_pa <- x %>%
    dplyr::group_by(!!d_id) %>%
    dplyr::summarize(sum_pa= sum(f_p * f_c))

  x <- x %>%
    dplyr::left_join(sum_pa,
                     by = rlang::as_name(d_id))
  f_t <- (f_p * f_c) / dplyr::pull(x, sum_pa)

  x %>%
    dplyr::mutate(V_ij = !!opp * !!f_t)
}
```

All schools, including 2024 schools. Supercedes the hamonDESTs package:
```{r}
load(file="data-products/SCHOOLS.Rdata") #for this paper, the schools under analysis
```

Load census
```{r}
load(file="data-products/cen_data_2011_hamiltononly.Rdata")
cen_data_2011_hamiltononly <- cen_data_2011_1 |> filter(`Region Name` == "Hamilton") #NOTE: GEOUID 35250242, 35250133, 35250457 do not have any child population
rm("cen_data_2011_1")
load(file="data-products/cen_data_2016_hamiltononly.Rdata")
cen_data_2016_hamiltononly <- cen_data_2016_1 |> filter(`Region Name` == "Hamilton") #NOTE: GEOUID 35250133, 35250398, 35250427 do not have any child population
rm("cen_data_2016_1")
load(file="data-products/cen_data_2021_hamiltononly.Rdata")
cen_data_2021_hamiltononly <- cen_data_2021_1 
rm("cen_data_2021_1")

ham_city_bound <- st_read("data-inputs/Boundaries/City_Boundary.shp") |> st_transform(st_crs(cen_data_2011_hamiltononly))
ham_community_bounds <- st_read("data-inputs/Boundaries/Community_Boundaries.shp") |> st_transform(st_crs(cen_data_2011_hamiltononly))
ggh_taz <- st_read("data-inputs/TTS/tts06_83_region.shp") |> st_transform(st_crs(cen_data_2011_hamiltononly))
```

```{r}
cen_data_2011_hamiltononly <- 
  cen_data_2011_hamiltononly %>% dplyr::select( #removing the variables we don't need for this analysis
  -c("Shape Area", "Quality Flags", "Type", "CD_UID", "NHS Non-Return Rate", "CSD_UID", "CT_UID", "CT_UID","NHS Non Return Rate", "v_CA11N_2564: Median after-tax household income $", "v_CA11F_17: 15 to 19 years")) %>%
  rename( #renaming the variables
    "DA_households" = "Households",
    "DA_dwellings" = "Dwellings",
    "DA_tot_pop" = "Population",
    "DA_med_AT_family_income" = "v_CA11N_2458: Median after-tax family income $",
    "DA_med_AT_lone-family_income" = "v_CA11N_2476: Median after-tax family income $",
    "DA_prev_LIMAT_under18" = "v_CA11N_2609: Less than 18 years %",
    "DA_pop_5to9" = "v_CA11F_11: 5 to 9 years",
    "DA_pop_10to14" = "v_CA11F_14: 10 to 14 years") %>%
  mutate(DA_pop_5to14 = DA_pop_5to9+(DA_pop_10to14)) |> #*0.9)) %>%
  dplyr::select(-c("DA_pop_5to9","DA_pop_10to14"))
```

```{r}
cen_data_2016_hamiltononly <- 
  cen_data_2016_hamiltononly %>% dplyr::select( #removing the variables we don't need for this analysis
  -c("Shape Area",  "Type", "CD_UID", "CSD_UID", "CT_UID", "CT_UID", "v_CA16_2398: Median after-tax income of households in 2015 ($)",  "v_CA16_64: 15 to 19 years"  )) %>%
  rename( #renaming the variables
    "DA_households" = "Households",
    "DA_dwellings" = "Dwellings",
    "DA_tot_pop" = "Population",
    "DA_med_AT_family_income" = "v_CA16_2448: Median after-tax income of economic families in 2015 ($)",
    "DA_med_AT_lone-family_income" = "v_CA16_2460: Median after-tax income of lone-parent economic families in 2015 ($)",
         "DA_prev_LIMAT_under18" = "v_CA16_2543: 0 to 17 years (%)",
         "DA_pop_5to9" = "v_CA16_25: 5 to 9 years",
         "DA_pop_10to14" = "v_CA16_43: 10 to 14 years") %>% mutate(DA_pop_5to14 = DA_pop_5to9+(DA_pop_10to14)) |> #*0.9)) %>%
         dplyr::select(-c("DA_pop_5to9","DA_pop_10to14"))
```

```{r}
cen_data_2021_hamiltononly <- 
  cen_data_2021_hamiltononly %>% dplyr::select( #removing the variables we don't need for this analysis
  -c("Shape Area",  "Type", "CD_UID", "CSD_UID", "CT_UID", "CT_UID", "v_CA21_566: Median after-tax income in 2020 among recipients ($)" ,"v_CA21_71: 15 to 19 years")) %>%
  rename( #renaming the variables
    "DA_households" = "Households",
    "DA_dwellings" = "Dwellings",
    "DA_tot_pop" = "Population",
    "DA_med_AT_family_income" = "v_CA21_966: Median after-tax income of economic family in 2020 ($)",
    "DA_med_AT_lone-family_income" = "v_CA21_978: Median after-tax income of one-parent economic families in 2020 ($)",
    "DA_prev_LIMAT_under18" =  "v_CA21_1043: 0 to 17 years",
    "DA_pop_5to9" = "v_CA21_32: 5 to 9 years",
    "DA_pop_10to14" = "v_CA21_50: 10 to 14 years") %>%
    mutate(DA_pop_5to14 = DA_pop_5to9+(DA_pop_10to14)) |> #*0.9)) %>%
         dplyr::select(-c("DA_pop_5to9","DA_pop_10to14"))
```

load TTS travel impedance functions, and format SCHOOLS to refer to the TTS IDS (add GTA06 IDs to schools - "GTA06_Dest"):
```{r}
load(file="data-products/TTS_gamma2016_motor.Rdata")
load(file="data-products/TTS_exp2016_nonmotor.Rdata")

#formatting
cols_TAZIDs <- sp::over(as_Spatial(SCHOOLS |> st_transform(st_crs(ggh_taz))),
                              as_Spatial(ggh_taz |> dplyr::select(GTA06) |> rename("GTA06_Dest" = "GTA06")))

SCHOOLS <- cbind(SCHOOLS,cols_TAZIDs)
rm(cols_TAZIDs)
```

Loading travel times in 2021 and modal split in 2016:
```{r}
load(file="data-products/OD_tt_car_2021_1.Rdata")
load(file="data-products/OD_tt_car_2021_2.Rdata")
load(file="data-products/OD_tt_car_2021_3.Rdata")
load(file="data-products/OD_tt_car_2021_4.Rdata")

load(file="data-products/OD_tt_walk_2021_1.Rdata")
load(file="data-products/OD_tt_walk_2021_2.Rdata")
load(file="data-products/OD_tt_walk_2021_3.Rdata")
load(file="data-products/OD_tt_walk_2021_4.Rdata")

load(file="data-inputs/TTS/od_trips_perc_2016.Rdata")

#renaming travel time matrices 
ttm_car_2021 <- bind_rows(output_OD_car1, output_OD_car2, output_OD_car3, output_OD_car4) %>% mutate(mode = "c")
rm(output_OD_car1, output_OD_car2, output_OD_car3, output_OD_car4)

ttm_walk_2021 <- bind_rows(output_OD_walk_elem1,output_OD_walk_elem2,output_OD_walk_elem3,output_OD_walk_elem4) %>% mutate(mode = "w")
rm(output_OD_walk_elem1,output_OD_walk_elem2,output_OD_walk_elem3,output_OD_walk_elem4)
```

load centroids:
```{r}
load(file="data-inputs/Parcels/RED_LU_centroids_2021.RData") #parcels

#drop all parcel points that do not have a GeoUID (i.e., a DA identifier -- this means they fall outside of Hamilton). It's 10 for 2011,  11 for 2016, and 11 for 2021
RED_LU_centroids_2021 <- RED_LU_centroids_2021 %>% drop_na(GeoUID) %>%
  group_by(ID) %>%
  dplyr::select(c("ID","GeoUID_2016","GeoUID","GTA06_Orig","WARD","COMMUNITY","URBAN_RURA","LUC1","LUC2","LUC1_DESCR","TYPE","geometry"))
```

Add census data to the parcel centroids:
```{r}
RED_LU_centroids_2021 <- RED_LU_centroids_2021 %>%
  merge(cen_data_2021_hamiltononly %>% st_drop_geometry(), by="GeoUID") 
```

```{r}
#create OD matrix -- each parcel to each school, and then drop all OD pairs that don't have a travel time ()
OD_2021 <- RED_LU_centroids_2021 %>% 
  cross_join(SCHOOLS %>% st_drop_geometry() |> filter(Year_2 == "2016 and 2021" | Year_2 == "2021") %>%
               dplyr::select(c("SchoolID", "OTGC2021", "GTA06_Dest")))
#add travel time, per mode
test <- OD_2021 %>% mutate(ID = as.character(ID),
                           SchoolID = as.character(SchoolID)) %>%
  left_join(ttm_car_2021, by=c("ID" = "from_id",
                               "SchoolID" = "to_id")) %>% 
  drop_na(travel_time_p50)  ##CHECK!! USE test$OTGC2021 |> unique() to see if ~133 schools have unique capacities..
```

```{r}
OD_2021 <- OD_2021 %>% mutate(ID = as.character(ID),
                           SchoolID = as.character(SchoolID)) %>%
  left_join(ttm_walk_2021 |> filter(travel_time_p50 <= 27), 
            by=c("ID" = "from_id",
                 "SchoolID" = "to_id")) %>% 
  drop_na(travel_time_p50)

OD_2021 <- rbind(test,OD_2021)
rm(test, ttm_car_2021, ttm_walk_2021)
```

```{r}
#reduce dimensions
OD_2021 <- OD_2021 %>% dplyr::select(-c("LUC1","LUC1_DESCR", "LUC2", "DA_dwellings", "CMA_UID", "Region Name", "geometry","Area (sq km)"))

# Check the mean and median travel times by mode, the first is at the level of zn1259 and the second is at zn208
OD_2021 %>% 
  group_by(mode) %>% 
  summarize(mean_tt = mean(travel_time_p50),
            median_tt = median(travel_time_p50),
            min_tt = min(travel_time_p50),
            max_tt = max(travel_time_p50))
```
```{r}
#buffering to clean up the edges of these objects -> there wre issues using st_join. 
sf_use_s2(FALSE)
ggh_taz <- ggh_taz |> st_make_valid()
ggh_taz <- st_buffer(ggh_taz, dist=0)
ham_taz <- ggh_taz %>% dplyr::filter(REGION==6) #Filter out everything other than Planning District 6 = Hamilton
```

```{r}
#now let's add the communtiy names to the TAZ and summarize the % of students who use what mode by community. 
ggh_taz_communities <- st_intersection(st_centroid(ham_taz),
                                       ham_community_bounds) |>
  dplyr::select(c("GTA06", "REGION", "COMMUNITY_"))

od_trips_perc_2016 <- od_trips_perc_2016 %>%
  left_join(ggh_taz_communities |> st_drop_geometry() |>
              mutate(GTA06 = as.character(GTA06)), 
            by=c("Origin"="GTA06"))

od_trips_perc_comm_2016 <- od_trips_perc_2016 %>% group_by(COMMUNITY_) %>%
  summarize(TAZOrig_car_trips = sum(TAZOrig_car_trips),
            TAZOrig_walk_trips = sum(TAZOrig_walk_trips)) %>% 
  mutate(TAZOrig_car_perc = TAZOrig_car_trips/(TAZOrig_car_trips + TAZOrig_walk_trips),
         COMMUNITY = case_when(COMMUNITY_ == "Ancaster" ~ "AN",
                                                         COMMUNITY_ == "Dundas" ~ "DU",
                                                         COMMUNITY_ == "Flamborough" ~ "FL",
                                                         COMMUNITY_ == "Glanbrook" ~ "GL",
                                                         COMMUNITY_ == "Hamilton" ~ "HA",
                                                         COMMUNITY_ == "Stoney Creek" ~ "SC")) |>
  dplyr::select(-c("COMMUNITY_"))

od_trips_perc_comm_2016
```

```{r}
# Lets add the % of car-mode used to the full OD list and then start transforming the frame for Spatial Availability calc.
OD_2021 <- OD_2021 |> left_join(od_trips_perc_comm_2016 |>
                                  dplyr::select(c("TAZOrig_car_perc", "COMMUNITY")),
                                by = "COMMUNITY") %>%
  mutate(
  DA_pop_5to14 = ifelse(is.na(DA_pop_5to14), 0, DA_pop_5to14),
  DA_households = ifelse(is.na(DA_households), 0, DA_households))

#'mode_count' identifies if a OD pair (ID to SchoolID) is either just car (1) or car and walk (2). Join this to the full OD list.
OD_2021_mode_count <- OD_2021 |>
  group_by(ID, SchoolID) %>% 
  summarise(mode_count = n())

OD_2021 <- OD_2021 %>% left_join(OD_2021_mode_count, by = c("ID", "SchoolID"))
rm(OD_2021_mode_count)
```

```{r}
#now, here we make the assumption on how many students live in each parcel. So since we assume that any parcel could potentially go to any school - I think it makes sense to also assume that the most granular level of population data we have (DA-level) is equally assigned to each parcel and then equally assigned to each OD flow based on it's mode-split (on average ~80/20 car/walk).

#Count the number of ID's for each OD pair -- this can be seen as the number of 'trips' being made from a parcel. There are 147 elementary schools in 2011 -- all parcels can reach all schools by car in 60 mins. So if a parcel only has 1 mode_count (car) it will have a "DA_ID_count" of 147 (the number of OD home-to-all-school trips). This is the number that the rate of Students in DA per OD flows in the DA ("IDs_in_a_DA_count") is multipled by.
DA_ID_count <- OD_2021 |> group_by(ID) |> summarise(DA_ID_count = n()) 
OD_2021 <- OD_2021 |> left_join(DA_ID_count, by="ID")

#Here we calculate the number of OD trips per O (ID) that occur in each DA. This is how much the # of students in a DA is divded by for each OD. 
IDs_in_a_DA_count <- OD_2021 |> group_by(GeoUID) |> summarise(IDs_in_a_DA_count = sum(DA_ID_count)) 
OD_2021 <- OD_2021 |> left_join(IDs_in_a_DA_count, by="GeoUID")

#We calculate the _kid-pop_ per parcel and then we adjusted it based on if the OD flow is 'walking' or 'car'
OD_2021 <- OD_2021 |> mutate(kid_pop = DA_pop_5to14/IDs_in_a_DA_count*DA_ID_count, 
               kid_pop = ifelse(is.na(kid_pop) | is.nan(kid_pop) | kid_pop == Inf, 0, kid_pop),
               kid_popadj = case_when(mode == "w" & mode_count == 2 ~ (1.0000000-TAZOrig_car_perc)*2*kid_pop,
                                         mode == "c" & mode_count == 2 ~ (TAZOrig_car_perc)*2*kid_pop,
                                         mode == "c" & mode_count == 1 ~ kid_pop),
                  kid_popadj = ifelse(is.na(kid_popadj) | is.nan(kid_popadj) | kid_popadj == Inf, 0, kid_popadj))
```

```{r}
#checks!, these should be the total number of kids (^the code above) -- 62195
sum(cen_data_2021_hamiltononly$DA_pop_5to14, na.rm=TRUE)

OD_2021$kid_pop %>% sum()
OD_2021$kid_popadj %>% sum()
OD_2021 |> group_by(GeoUID) |> summarize(DA_pop_5to14 = mean(DA_pop_5to14)) |> dplyr::select(DA_pop_5to14) |> sum()
```

```{r}
#now add the impedance function, etc to the full OD 
OD_2021 <- OD_2021 |> mutate(
  travel_time_p50 = ifelse(travel_time_p50 == 0, 0.5, travel_time_p50),
  r = 1,
  f = ifelse(mode == "c", stats::dgamma(
    travel_time_p50, gamma2016_motor$estimate["shape"], gamma2016_motor$estimate["rate"]), 
    stats::dexp(
      travel_time_p50, exp2016_nonmotor$estimate["rate"])))
```

```{r}
#calculate spatial availability. 
avail_2021 <- sp_avail_detailed(OD_2021, ID, SchoolID, kid_popadj, OTGC2021, mode, r, f, alpha = 1)
save(avail_2021, file="data-products/avail_2021.RData")
save(OD_2021, file="data-products/OD_2021.RData")
```

```{r}
#checks, yes!
OD_2021$OTGC2021 |> unique() |> sum()
avail_2021$V_ij |> sum(na.rm=T)
```

```{r}
rm(OD_2021)
```

### 2024-11-21 paper/Manuscript.qmd
Added 2021 schools and census to the database. Started Manuscript changes.

@@
 
+<!--Chat with Antonio about the assumption regarding the 'student population' per parcel.
+As it stands, when the student pop for the OD matrix is summed, it equals the total student population in the city.
+Each parcel is assigned a DA-pop/(the number of potential parcel trips in a DA * the potential parcels trips from that parcel).
+This is highly correlated with we just weighted it by DA-pop/DA-dwellings (but need to double check).
+--->
+
 ```{r knitr-setup, include=FALSE}
@@
 ```{r, include=FALSE}
-load(file = paste0(here::here(),"/data-prep/data-products/cen_data_2011_hamiltononly.Rdata"))
-cen_data_2011_hamiltononly <- cen_data_2011_1 |> filter(Region.Name == "Hamilton") #NOTE: GEOUID 35250242, 35250133, 35250457 do not have any child population
+load(paste0(here::here(),file="/data-prep/data-products/cen_data_2011_hamiltononly.Rdata"))
+cen_data_2011_hamiltononly <- cen_data_2011_1 |> filter(`Region Name` == "Hamilton") #NOTE: GEOUID 35250242, 35250133, 35250457 do not have any child population
 rm("cen_data_2011_1")
-load(file=paste0(here::here(),"/data-prep/data-products/cen_data_2016_hamiltononly.Rdata"))
-cen_data_2016_hamiltononly <- cen_data_2016_1 |> filter(Region.Name == "Hamilton") #NOTE: GEOUID 35250133, 35250398, 35250427 do not have any child population
+load(paste0(here::here(),file="/data-prep/data-products/cen_data_2016_hamiltononly.Rdata"))
+cen_data_2016_hamiltononly <- cen_data_2016_1 |> filter(`Region Name` == "Hamilton") #NOTE: GEOUID 35250133, 35250398, 35250427 do not have any child population
 rm("cen_data_2016_1")
+load(paste0(here::here(),file="/data-prep/data-products/cen_data_2021_hamiltononly.Rdata"))
+cen_data_2021_hamiltononly <- cen_data_2021_1 
+rm("cen_data_2021_1")
 
@@
          COMMUNITY_ = ifelse(COMMUNITY_ == "Hamilton", "Hamilton Central", COMMUNITY_)) |> st_transform(st_crs(cen_data_2016_hamiltononly))
+
+ggh_taz <- st_read(paste0(here::here(),"/data-prep/data-inputs/TTS/tts06_83_region.shp")) |> st_transform(st_crs(cen_data_2011_hamiltononly))
 ```
@@
   cen_data_2011_hamiltononly %>% dplyr::select( #removing the variables we don't need for this analysis
-    -c("Shape.Area", "Quality.Flags", "Type", "CD_UID", "NHS.Non.Return.Rate", "CSD_UID", "CT_UID", "CT_UID","NHS.Non.Return.Rate.1", "v_CA11N_2564..Median.after.tax.household.income..", "v_CA11F_17..15.to.19.years")) %>%
+  -c("Shape Area", "Quality Flags", "Type", "CD_UID", "NHS Non-Return Rate", "CSD_UID", "CT_UID", "CT_UID","NHS Non Return Rate", "v_CA11N_2564: Median after-tax household income $", "v_CA11F_17: 15 to 19 years")) %>%
   rename( #renaming the variables
@@
     "DA_tot_pop" = "Population",
-    "DA_med_AT_family_income" = "v_CA11N_2458..Median.after.tax.family.income..",
-    "DA_med_AT_lone-family_income" = "v_CA11N_2476..Median.after.tax.family.income..",
-    "DA_prev_LIMAT_under18" = "v_CA11N_2609..Less.than.18.years..",
-    "DA_pop_5to9" = "v_CA11F_11..5.to.9.years",
-    "DA_pop_10to14" = "v_CA11F_14..10.to.14.years") %>%
+    "DA_med_AT_family_income" = "v_CA11N_2458: Median after-tax family income $",
+    "DA_med_AT_lone-family_income" = "v_CA11N_2476: Median after-tax family income $",
+    "DA_prev_LIMAT_under18" = "v_CA11N_2609: Less than 18 years %",
+    "DA_pop_5to9" = "v_CA11F_11: 5 to 9 years",
+    "DA_pop_10to14" = "v_CA11F_14: 10 to 14 years") %>%
   mutate(DA_pop_5to14 = DA_pop_5to9+(DA_pop_10to14)) |> #*0.9)) %>%
@@
   cen_data_2016_hamiltononly %>% dplyr::select( #removing the variables we don't need for this analysis
-    -c("Shape.Area",  "Type", "CD_UID", "CSD_UID", "CT_UID", "CT_UID", "v_CA16_2398..Median.after.tax.income.of.households.in.2015....", "v_CA16_64..15.to.19.years")) %>%
+  -c("Shape Area",  "Type", "CD_UID", "CSD_UID", "CT_UID", "CT_UID", "v_CA16_2398: Median after-tax income of households in 2015 ($)",  "v_CA16_64: 15 to 19 years"  )) %>%
   rename( #renaming the variables
@@
     "DA_tot_pop" = "Population",
-    "DA_med_AT_family_income" = "v_CA16_2448..Median.after.tax.income.of.economic.families.in.2015....",
-    "DA_med_AT_lone-family_income" = "v_CA16_2460..Median.after.tax.income.of.lone.parent.economic.families.in.2015....",
-    "DA_prev_LIMAT_under18" = "v_CA16_2543..0.to.17.years....",
-    "DA_pop_5to9" = "v_CA16_25..5.to.9.years",
-    "DA_pop_10to14" = "v_CA16_43..10.to.14.years") %>% mutate(DA_pop_5to14 = DA_pop_5to9+(DA_pop_10to14)) |> #*0.9)) %>%
+    "DA_med_AT_family_income" = "v_CA16_2448: Median after-tax income of economic families in 2015 ($)",
+    "DA_med_AT_lone-family_income" = "v_CA16_2460: Median after-tax income of lone-parent economic families in 2015 ($)",
+         "DA_prev_LIMAT_under18" = "v_CA16_2543: 0 to 17 years (%)",
+         "DA_pop_5to9" = "v_CA16_25: 5 to 9 years",
+         "DA_pop_10to14" = "v_CA16_43: 10 to 14 years") %>% mutate(DA_pop_5to14 = DA_pop_5to9+(DA_pop_10to14)) |> #*0.9)) %>%
          dplyr::select(-c("DA_pop_5to9","DA_pop_10to14"))
+
+cen_data_2021_hamiltononly <- 
+  cen_data_2021_hamiltononly %>% dplyr::select( #removing the variables we don't need for this analysis
+  -c("Shape Area",  "Type", "CD_UID", "CSD_UID", "CT_UID", "CT_UID", "v_CA21_566: Median after-tax income in 2020 among recipients ($)" ,"v_CA21_71: 15 to 19 years")) %>%
+  rename( #renaming the variables
+    "DA_households" = "Households",
+    "DA_dwellings" = "Dwellings",
+    "DA_tot_pop" = "Population",
+    "DA_med_AT_family_income" = "v_CA21_966: Median after-tax income of economic family in 2020 ($)",
+    "DA_med_AT_lone-family_income" = "v_CA21_978: Median after-tax income of one-parent economic families in 2020 ($)",
+    "DA_prev_LIMAT_under18" =  "v_CA21_1043: 0 to 17 years",
+    "DA_pop_5to9" = "v_CA21_32: 5 to 9 years",
+    "DA_pop_10to14" = "v_CA21_50: 10 to 14 years") %>%
+    mutate(DA_pop_5to14 = DA_pop_5to9+(DA_pop_10to14)) |> #*0.9)) %>%
+         dplyr::select(-c("DA_pop_5to9","DA_pop_10to14"))
 ```
@@
 ```{r, include=FALSE}
-library(hamONdests) # A R package of Hamilton, Ontario destinations
-
-data(Schools_2011_2016)
-ALL_SCHOOLS_Elem <- Schools_2011_2016 %>% filter(Level == "Elementary")
-rm("Schools_2011_2016")
+# library(hamONdests) # A R package of Hamilton, Ontario destinations
+load(file=paste0(here::here(),"/data-prep/data-products/SCHOOLS.Rdata")) #for this paper, the schools under analysis
 
 # add community label to schools
-ALL_SCHOOLS_Elem <- st_intersection(ham_community_bounds,
-                        ALL_SCHOOLS_Elem |> st_transform(st_crs(cen_data_2016_hamiltononly)))
-
-data(School_Catchments_2011_2016)
-ALL_SCHOOLS_CATCH_Elem <- School_Catchments_2011_2016 %>% filter(Level == "Elementary")
-rm("School_Catchments_2011_2016")
+SCHOOLS <- st_intersection(ham_community_bounds,
+                        SCHOOLS |> st_transform(st_crs(cen_data_2016_hamiltononly)))
+
+# data(School_Catchments_2011_2016)
+# ALL_SCHOOLS_CATCH_Elem <- School_Catchments_2011_2016 %>% filter(Level == "Elementary")
+# rm("School_Catchments_2011_2016")
 
@@
 
-```{r}
-# # maybe consider using year 2021 as a comparision? 2011 vs 2016 vs 2021?
-# SCHOOLS_2023 <- st_read("data-prep/data-inputs/Educational_Institutions-2023/Educational_Institutions.shp") |> filter(CATEGORY == "Elementary School" & (SCHOOL_BOA == "Hamilton-Wentworth Catholic District School Board" | SCHOOL_BOA == "Hamilton-Wentworth District School Board" )) |> st_transform(st_crs(cen_data_2011_hamiltononly))
-```
-
 ```{r augmenting-catchment, include=FALSE}
-ALL_SCHOOLS_CATCH_Elem_extra2011 <- ALL_SCHOOLS_CATCH_Elem |>
-  left_join(ALL_SCHOOLS_Elem |> st_drop_geometry() |> group_by(CID2011) |> 
-              summarize(OTGC = sum(OTGC2011, na.rm = T)), 
-            by = c("CID" = "CID2011")) |> filter(!is.na(OTGC))
-
-ALL_SCHOOLS_CATCH_Elem_extra2016 <- ALL_SCHOOLS_CATCH_Elem |>
-  left_join(ALL_SCHOOLS_Elem |> st_drop_geometry() |> group_by(CID2016) |> 
-              summarize(OTGC = sum(OTGC2016, na.rm = T)), 
-            by = c("CID" = "CID2016")) |> filter(!is.na(OTGC))
-
-ALL_SCHOOLS_CATCH_Elem_extra <- rbind(ALL_SCHOOLS_CATCH_Elem_extra2011, ALL_SCHOOLS_CATCH_Elem_extra2016)
-rm(ALL_SCHOOLS_CATCH_Elem_extra2011,ALL_SCHOOLS_CATCH_Elem_extra2016)
-
-#checks
-sum(ALL_SCHOOLS_Elem$OTGC2011, na.rm=T) + sum(ALL_SCHOOLS_Elem$OTGC2016, na.rm=T)
-ALL_SCHOOLS_CATCH_Elem_extra$OTGC |> sum(na.rm=T)
-
-#reformating schools for mapping
-ALL_SCHOOLS_Elem_nochangeorexpanded <- ALL_SCHOOLS_Elem |> filter(Year == "2011 and 2016") |> mutate(Year = ifelse(Year == "2011 and 2016", "2011", Year))
-ALL_SCHOOLS_Elem_formapping <- rbind(ALL_SCHOOLS_Elem, ALL_SCHOOLS_Elem_nochangeorexpanded)
-rm(ALL_SCHOOLS_Elem_nochangeorexpanded)
+# ALL_SCHOOLS_CATCH_Elem_extra2011 <- ALL_SCHOOLS_CATCH_Elem |>
+#   left_join(ALL_SCHOOLS_Elem |> st_drop_geometry() |> group_by(CID2011) |> 
+#               summarize(OTGC = sum(OTGC2011, na.rm = T)), 
+#             by = c("CID" = "CID2011")) |> filter(!is.na(OTGC))
+# 
+# ALL_SCHOOLS_CATCH_Elem_extra2016 <- ALL_SCHOOLS_CATCH_Elem |>
+#   left_join(ALL_SCHOOLS_Elem |> st_drop_geometry() |> group_by(CID2016) |> 
+#               summarize(OTGC = sum(OTGC2016, na.rm = T)), 
+#             by = c("CID" = "CID2016")) |> filter(!is.na(OTGC))
+# 
+# ALL_SCHOOLS_CATCH_Elem_extra <- rbind(ALL_SCHOOLS_CATCH_Elem_extra2011, ALL_SCHOOLS_CATCH_Elem_extra2016)
+# rm(ALL_SCHOOLS_CATCH_Elem_extra2011,ALL_SCHOOLS_CATCH_Elem_extra2016)
+# 
+# #checks
+# sum(ALL_SCHOOLS_Elem$OTGC2011, na.rm=T) + sum(ALL_SCHOOLS_Elem$OTGC2016, na.rm=T)
+# ALL_SCHOOLS_CATCH_Elem_extra$OTGC |> sum(na.rm=T)
+
+#reformatting schools for mapping
+SCHOOLS_nochangeorexpanded <- SCHOOLS |> filter(Year == "2011 and 2016" & Year_2 == "2016 and 2021") |> 
+  mutate(Year = ifelse(Year == "2011 and 2016", "2011", Year))
+ALL_SCHOOLS_Elem_formapping <- rbind(SCHOOLS, SCHOOLS_nochangeorexpanded)
+rm(SCHOOLS_nochangeorexpanded)
 ALL_SCHOOLS_Elem_formapping <- ALL_SCHOOLS_Elem_formapping |> mutate(Year = ifelse(Year == "2011 and 2016", "2016", Year))

### 2024-12-12 data-prep/01-processing-schools-parcels-census.qmd
changes to bib and analysis files after revision

@@
 cen_data_2021_1 <- get_census(dataset='CA21', regions=list(CMA="35537"),
-                           vectors=c("v_CA21_566", #Median after-tax income in 2020 among recipients ($),
+                           vectors=c(
+                                    "v_CA21_566", #Median after-tax income in 2020 among recipients ($),
                                      "v_CA21_966", #Median after-tax income of economic family in 2020 ($)
@@
 - Glen Echo, Glen Brae, Sir Isaac Brock, and Elizabeth Bagshaw elementary schools closed -- and "Viola Desmond Elementary School" opened in September 2021 on the site of Glen Brae.
-Let's make Glen Brae as "Removed", and add new school named "Viola Desmond" as SchoolID 156.202021, ELEM 0, urban.dist 7948.096, footprint 3,902 m²
+Let's make Glen Brae "Removed", and add new school named "Viola Desmond" as SchoolID 156.202021, ELEM 0, urban.dist 7948.096, footprint 3,902 m²
 (https://www.hwdsb.on.ca/sirwilfridlaurier/2021/05/19/viola-desmond-elementary-school-update/)
@@
 ```{r}
-test_ALL_SCHOOLS_ELEM <- ALL_SCHOOLS_Elem |> mutate(Year_2 = ifelse(SchoolID %in% SCHOOLS_2024_and_2016$SchoolID, "2016 and 2021", NA),
+test_ALL_SCHOOLS_ELEM <- ALL_SCHOOLS_Elem |>
+  mutate(Year_2 = ifelse(SchoolID %in% SCHOOLS_2024_and_2016$SchoolID, "2016 and 2021", NA),
                            Status_2 = ifelse(!is.na(Year_2), "NoChange", NA),
+
+                           Year_2 = ifelse(SchoolID == "156", "2016", Year_2),
                            Status_2 = ifelse(SchoolID == "156", "Removed", Status_2),
-                           footprint2021 = ifelse(SchoolID =="156", 3902, footprint2016))
+         footprint2021 = NA)
 
@@
          Status_2 = "New",
+         urban.dist = 7948.0958,
          footprint2021 =3902)
@@
 
-Next assign 2016 DA IDs to the centroids of 2011 parcels:
+Next assign 2016 DA IDs and 2021 DA IDs to the centroids of 2011 parcels:
 ```{r}
@@
 
+RED_LU_centroids_2011_w2021ID <- sp::over(as_Spatial(st_centroid(RED_LU_2011)),as_Spatial(cen_data_2021_1 |> dplyr::select(GeoUID) |> rename("GeoUID_2021" = "GeoUID")))
+
 RED_LU_centroids_2011 <- cbind(RED_LU_centroids_2011_w2016ID,
+                               RED_LU_centroids_2011_w2021ID,
                               RED_LU_centroids_2011,
@@
 
-Next assign 2016 DA IDs to the centroids of 2021 parcels:
+And assign 2016 DA and 2011 IDs to the centroids of 2021 parcels:
 ```{r}
@@
 
+RED_LU_centroids_2021_w2011ID <- sp::over(as_Spatial(st_centroid(RED_LU_2020)),
+                                          as_Spatial(cen_data_2011_1 |> dplyr::select(GeoUID) |> rename("GeoUID_2011" = "GeoUID")))
+
 RED_LU_centroids_2021 <- cbind(RED_LU_centroids_2021_w2016ID,
+                               RED_LU_centroids_2021_w2011ID,
                               RED_LU_centroids_2021,
                               st_centroid(RED_LU_2020))
+```
+
+For good measure, do the same for the 2016 centroids:
+```{r}
+RED_LU_centroids_2016 <- sp::over(as_Spatial(st_centroid(RED_LU_2016)),
+                                  as_Spatial(cen_data_2016_1 |> dplyr::select(GeoUID)))
+
+RED_LU_centroids_2016_w2011ID <- sp::over(as_Spatial(st_centroid(RED_LU_2016)),as_Spatial(cen_data_2011_1 |> dplyr::select(GeoUID) |> rename("GeoUID_2011" = "GeoUID")))
+
+RED_LU_centroids_2016_w2021ID <- sp::over(as_Spatial(st_centroid(RED_LU_2016)),as_Spatial(cen_data_2021_1 |> dplyr::select(GeoUID) |> rename("GeoUID_2021" = "GeoUID")))
+
+RED_LU_centroids_2016 <- cbind(RED_LU_centroids_2016_w2011ID,
+                               RED_LU_centroids_2016_w2021ID,
+                              RED_LU_centroids_2016,
+                              st_centroid(RED_LU_2016))
 ```
@@
                               as_Spatial(ggh_taz |> select(GTA06) |> rename("GTA06_Orig" = "GTA06")))
+
 RED_LU_centroids_2011<- cbind(RED_LU_centroids_2011,cols_TAZIDs)
 
-cols_TAZIDs <- sp::over(as_Spatial(st_centroid(RED_LU_2016)),
+cols_TAZIDs <-sp::over(as_Spatial(st_centroid(RED_LU_2016)),
                               as_Spatial(ggh_taz |> select(GTA06) |> rename("GTA06_Orig" = "GTA06")))
@@
 ```{r}
-od_trips_perc_2011 <- od_trips_2011_motor %>% left_join(od_trips_2011_nonmotor, by=c("Origin", "Destination")) %>%
+od_trips_perc_2011 <- od_trips_2011_motor %>% full_join(od_trips_2011_nonmotor, by=c("Origin", "Destination")) %>%
   group_by(Origin, Destination) |>
@@
 
-od_trips_perc_2016 <- od_trips_2016_motor %>% left_join(od_trips_2016_nonmotor, by=c("Origin", "Destination")) %>%
+od_trips_perc_2016 <- od_trips_2016_motor %>% full_join(od_trips_2016_nonmotor, by=c("Origin", "Destination"))%>%
   group_by(Origin,Destination) |>
@@
 ```
+
+```{r}
+##TEEESSST -- here we make sure that the full_join worked correctly for calculating the % of trips per mode.. it works!
+st_intersection(st_centroid(ham_taz) %>% st_transform(crs = 4326), 
+                ham_community_bounds%>% st_transform(crs = 4326)) |> filter(COMMUNITY_ == "Dundas")
+
+TEST <-c(5178, 5191, 5202, 5205, 5207, 5213, 5215, 5236, 5246, 5247, 5252)
+```
+
+```{r}
+#see, 29 rows --> it comes from 25 rows and 7 rows (added together = 32 + collapsing 3 rows as those ODs both have walk and car trips = 29 unique ODs!)
+od_trips_perc_2011 |> filter(Origin %in% TEST)
+od_trips_2011_motor |> filter(Origin %in% TEST)
+od_trips_2011_nonmotor |> filter(Origin %in% TEST)
+```

### 2024-12-12 data-prep/04-Creating-avail_2011-16.Rmd
Added 2021 schools and census to the database. Started Manuscript changes.
changes to bib and analysis files after revision

@@
 ---
-# 2011 calcs
 
@@
 
+All schools, including 2024 schools. Supercedes the hamonDESTs package:
+```{r}
+load(file="data-products/SCHOOLS.Rdata") #for this paper, the schools under analysis
+```
+
+Load census
 ```{r}
 load(file="data-products/cen_data_2011_hamiltononly.Rdata")
-cen_data_2011_hamiltononly <- cen_data_2011_1 |> filter(Region.Name == "Hamilton") #NOTE: GEOUID 35250242, 35250133, 35250457 do not have any child population
+cen_data_2011_hamiltononly <- cen_data_2011_1 |> filter(`Region Name` == "Hamilton") #NOTE: GEOUID 35250242, 35250133, 35250457 do not have any child population
 rm("cen_data_2011_1")
 load(file="data-products/cen_data_2016_hamiltononly.Rdata")
-cen_data_2016_hamiltononly <- cen_data_2016_1 |> filter(Region.Name == "Hamilton") #NOTE: GEOUID 35250133, 35250398, 35250427 do not have any child population
+cen_data_2016_hamiltononly <- cen_data_2016_1 |> filter(`Region Name` == "Hamilton") #NOTE: GEOUID 35250133, 35250398, 35250427 do not have any child population
 rm("cen_data_2016_1")
+load(file="data-products/cen_data_2021_hamiltononly.Rdata")
+cen_data_2021_hamiltononly <- cen_data_2021_1 
+rm("cen_data_2021_1")
 
@@
 ```
+```{r}
+#buffering to clean up the edges of these objects -> there wre issues using st_join. 
+sf_use_s2(FALSE)
+ggh_taz <- ggh_taz |> st_make_valid()
+ggh_taz <- st_buffer(ggh_taz, dist=0)
+ham_taz <- ggh_taz %>% dplyr::filter(REGION==6) #Filter out everything other than Planning District 6 = Hamilton
+```
 
@@
   cen_data_2011_hamiltononly %>% dplyr::select( #removing the variables we don't need for this analysis
-  -c("Shape.Area", "Quality.Flags", "Type", "CD_UID", "NHS.Non.Return.Rate", "CSD_UID", "CT_UID", "CT_UID","NHS.Non.Return.Rate.1", "v_CA11N_2564..Median.after.tax.household.income..", "v_CA11F_17..15.to.19.years")) %>%
+  -c("Shape Area", "Quality Flags", "Type", "CD_UID", "NHS Non-Return Rate", "CSD_UID", "CT_UID", "CT_UID","NHS Non Return Rate", "v_CA11N_2564: Median after-tax household income $", "v_CA11F_17: 15 to 19 years")) %>%
   rename( #renaming the variables
@@
     "DA_tot_pop" = "Population",
-    "DA_med_AT_family_income" = "v_CA11N_2458..Median.after.tax.family.income..",
-    "DA_med_AT_lone-family_income" = "v_CA11N_2476..Median.after.tax.family.income..",
-    "DA_prev_LIMAT_under18" = "v_CA11N_2609..Less.than.18.years..",
-    "DA_pop_5to9" = "v_CA11F_11..5.to.9.years",
-    "DA_pop_10to14" = "v_CA11F_14..10.to.14.years") %>%
+    "DA_med_AT_family_income" = "v_CA11N_2458: Median after-tax family income $",
+    "DA_med_AT_lone-family_income" = "v_CA11N_2476: Median after-tax family income $",
+    "DA_prev_LIMAT_under18" = "v_CA11N_2609: Less than 18 years %",
+    "DA_pop_5to9" = "v_CA11F_11: 5 to 9 years",
+    "DA_pop_10to14" = "v_CA11F_14: 10 to 14 years") %>%
   mutate(DA_pop_5to14 = DA_pop_5to9+(DA_pop_10to14)) |> #*0.9)) %>%
   dplyr::select(-c("DA_pop_5to9","DA_pop_10to14"))
+```
 
+```{r}
 cen_data_2016_hamiltononly <- 
   cen_data_2016_hamiltononly %>% dplyr::select( #removing the variables we don't need for this analysis
-  -c("Shape.Area",  "Type", "CD_UID", "CSD_UID", "CT_UID", "CT_UID", "v_CA16_2398..Median.after.tax.income.of.households.in.2015....", "v_CA16_64..15.to.19.years")) %>%
+  -c("Shape Area",  "Type", "CD_UID", "CSD_UID", "CT_UID", "CT_UID", "v_CA16_2398: Median after-tax income of households in 2015 ($)",  "v_CA16_64: 15 to 19 years"  )) %>%
   rename( #renaming the variables
@@
     "DA_tot_pop" = "Population",
-    "DA_med_AT_family_income" = "v_CA16_2448..Median.after.tax.income.of.economic.families.in.2015....",
-         "DA_med_AT_lone-family_income" = "v_CA16_2460..Median.after.tax.income.of.lone.parent.economic.families.in.2015....",
-         "DA_prev_LIMAT_under18" = "v_CA16_2543..0.to.17.years....",
-         "DA_pop_5to9" = "v_CA16_25..5.to.9.years",
-         "DA_pop_10to14" = "v_CA16_43..10.to.14.years") %>% mutate(DA_pop_5to14 = DA_pop_5to9+(DA_pop_10to14)) |> #*0.9)) %>%
+    "DA_med_AT_family_income" = "v_CA16_2448: Median after-tax income of economic families in 2015 ($)",
+    "DA_med_AT_lone-family_income" = "v_CA16_2460: Median after-tax income of lone-parent economic families in 2015 ($)",
+         "DA_prev_LIMAT_under18" = "v_CA16_2543: 0 to 17 years (%)",
+         "DA_pop_5to9" = "v_CA16_25: 5 to 9 years",
+         "DA_pop_10to14" = "v_CA16_43: 10 to 14 years") %>% mutate(DA_pop_5to14 = DA_pop_5to9+(DA_pop_10to14)) |> #*0.9)) %>%
          dplyr::select(-c("DA_pop_5to9","DA_pop_10to14"))
@@
 ```{r}
-library(hamONdests)
-data(Schools_2011_2016)
-ALL_SCHOOLS_Elem <- Schools_2011_2016 %>% filter(Level == "Elementary")
-rm("Schools_2011_2016")
-#data(School_Catchments_2011_2016)
-load(file="data-products/TTS_gamma2011_motor.Rdata")
-load(file="data-products/TTS_gamma2016_motor.Rdata")
-load(file="data-products/TTS_exp2011_nonmotor.Rdata")
-load(file="data-products/TTS_exp2016_nonmotor.Rdata")
+cen_data_2021_hamiltononly <- 
+  cen_data_2021_hamiltononly %>% dplyr::select( #removing the variables we don't need for this analysis
+  -c("Shape Area",  "Type", "CD_UID", "CSD_UID", "CT_UID", "CT_UID", "v_CA21_566: Median after-tax income in 2020 among recipients ($)" ,"v_CA21_71: 15 to 19 years")) %>%
+  rename( #renaming the variables
+    "DA_households" = "Households",
+    "DA_dwellings" = "Dwellings",
+    "DA_tot_pop" = "Population",
+    "DA_med_AT_family_income" = "v_CA21_966: Median after-tax income of economic family in 2020 ($)",
+    "DA_med_AT_lone-family_income" = "v_CA21_978: Median after-tax income of one-parent economic families in 2020 ($)",
+    "DA_prev_LIMAT_under18" =  "v_CA21_1043: 0 to 17 years",
+    "DA_pop_5to9" = "v_CA21_32: 5 to 9 years",
+    "DA_pop_10to14" = "v_CA21_50: 10 to 14 years") %>%
+    mutate(DA_pop_5to14 = DA_pop_5to9+(DA_pop_10to14)) |> #*0.9)) %>%
+         dplyr::select(-c("DA_pop_5to9","DA_pop_10to14"))
 ```
 
+load TTS travel impedance functions, and format SCHOOLS to refer to the TTS IDS (add GTA06 IDs to schools - "GTA06_Dest"):
 ```{r}
-#add GTA06 IDs to schools - "GTA06_Dest"}
-cols_TAZIDs <- sp::over(as_Spatial(ALL_SCHOOLS_Elem |> st_transform(st_crs(ggh_taz))),
+load(file="data-products/TTS_gamma2011_motor.Rdata")
+load(file="data-products/TTS_exp2011_nonmotor.Rdata")
+load(file="data-products/TTS_gamma2016_motor.Rdata")
+load(file="data-products/TTS_exp2016_nonmotor.Rdata")
+
+#formatting
+cols_TAZIDs <- sp::over(as_Spatial(SCHOOLS |> st_transform(st_crs(ggh_taz))),
                               as_Spatial(ggh_taz |> dplyr::select(GTA06) |> rename("GTA06_Dest" = "GTA06")))
 
-ALL_SCHOOLS_Elem <- cbind(ALL_SCHOOLS_Elem,cols_TAZIDs)
+SCHOOLS <- cbind(SCHOOLS,cols_TAZIDs)
 rm(cols_TAZIDs)
@@
 
+# 2011 calcs
 ```{r}
@@
 OD_2011 <- RED_LU_centroids_2011 %>% 
-  cross_join(ALL_SCHOOLS_Elem %>% st_drop_geometry() |> filter(Year != 2016) %>% dplyr::select(c("SchoolID", "OTGC2011", "GTA06_Dest")))
+  cross_join(SCHOOLS %>% st_drop_geometry() |> filter(Year != 2016) %>% dplyr::select(c("SchoolID", "OTGC2011", "GTA06_Dest")))
 
@@
 #reduce dimensions
-OD_2011 <- OD_2011 %>% dplyr::select(-c("LUC1_DESC", "LUC2_DESC", "res_units", "DA_dwellings", "CMA_UID", "Region.Name", "geometry","PROP_AREA", "Area..sq.km."))
+OD_2011 <- OD_2011 %>% dplyr::select(-c("LUC1_DESC", "LUC2_DESC", "DA_dwellings", "CMA_UID", "Region Name", "geometry","Area (sq km)"))
 
@@
 ```{r}
-rm(list=ls())
-```
-
-```{r load-packages, include=FALSE, cache=FALSE}
-library(dplyr)
-library(sf)
-library(ggplot2)
-library(tidyverse)
-library(units)
-library(patchwork)
-library(ggspatial)
-options(scipen = 999)
-```
-
-```{r sp_avail-function,include=FALSE}
-#defining the spatial availability function, this prints all the intermediates
-sp_avail_detailed <- function(x, o_id, d_id, pop, opp, pop_type, r, f, alpha = 1){
-
-  o_id <- rlang::enquo(o_id)
-  d_id <- rlang::enquo(d_id)
-  pop <- rlang::enquo(pop)
-  opp <- rlang::enquo(opp)
-  pop_type <- rlang::enquo(pop_type)
-  r <- rlang::enquo(r)
-  f <- rlang::enquo(f)
-
-  sum_pop <- x %>%
-    dplyr::distinct(!!o_id, !!pop_type,
-                    .keep_all = TRUE) %>%
-    dplyr::mutate(sum_pop = !!r*(!!pop)^alpha) %>%
-    dplyr::pull(sum_pop) %>%
-    sum()
-
-  f_p <- dplyr::pull(x, !!r) * dplyr::pull(x, !!pop)^alpha / sum_pop
-
-  sum_impedance <- x %>%
-    dplyr::group_by(!!d_id) %>%
-    dplyr::summarize(sum_impedance = sum(!!f))
-
-  x <- x %>%
-    dplyr::left_join(sum_impedance,
-                     by = rlang::as_name(d_id))
-
-  f_c <- dplyr::pull(x, !!f) / x$sum_impedance
-
-  x$f_c <- f_c
-  x$f_p <- f_p
-
-  sum_pa <- x %>%
-    dplyr::group_by(!!d_id) %>%
-    dplyr::summarize(sum_pa= sum(f_p * f_c))
-
-  x <- x %>%
-    dplyr::left_join(sum_pa,
-                     by = rlang::as_name(d_id))
-  f_t <- (f_p * f_c) / dplyr::pull(x, sum_pa)
-
-  x %>%
-    dplyr::mutate(V_ij = !!opp * !!f_t)
-}
-```
-
-```{r}
-load(file="data-products/cen_data_2016_hamiltononly.Rdata")
-cen_data_2016_hamiltononly <- cen_data_2016_1 |> filter(Region.Name == "Hamilton") #NOTE: GEOUID 35250133, 35250398, 35250427 do not have any child population
-rm("cen_data_2016_1")
-
-ham_city_bound <- st_read("data-inputs/Boundaries/City_Boundary.shp") |> st_transform(st_crs(cen_data_2016_hamiltononly))
-ham_community_bounds <- st_read("data-inputs/Boundaries/Community_Boundaries.shp") |> st_transform(st_crs(cen_data_2016_hamiltononly))
-ggh_taz <- st_read("data-inputs/TTS/tts06_83_region.shp") 
-# |> st_transform(st_crs(cen_data_2011_hamiltononly))
-```
-
-```{r}
-cen_data_2016_hamiltononly <- 
-  cen_data_2016_hamiltononly %>% dplyr::select( #removing the variables we don't need for this analysis
-  -c("Shape.Area",  "Type", "CD_UID", "CSD_UID", "CT_UID", "CT_UID", "v_CA16_2398..Median.after.tax.income.of.households.in.2015....", "v_CA16_64..15.to.19.years")) %>%
-  rename( #renaming the variables
-    "DA_households" = "Households",
-    "DA_dwellings" = "Dwellings",
-    "DA_tot_pop" = "Population",
-    "DA_med_AT_family_income" = "v_CA16_2448..Median.after.tax.income.of.economic.families.in.2015....",
-         "DA_med_AT_lone-family_income" = "v_CA16_2460..Median.after.tax.income.of.lone.parent.economic.families.in.2015....",
-         "DA_prev_LIMAT_under18" = "v_CA16_2543..0.to.17.years....",
-         "DA_pop_5to9" = "v_CA16_25..5.to.9.years",
-         "DA_pop_10to14" = "v_CA16_43..10.to.14.years") %>% mutate(DA_pop_5to14 = DA_pop_5to9+(DA_pop_10to14)) |> #*0.9)) %>%
-         dplyr::select(-c("DA_pop_5to9","DA_pop_10to14"))
-```
-
-```{r, include=FALSE}
-library(hamONdests)
-data(Schools_2011_2016)
-ALL_SCHOOLS_Elem <- Schools_2011_2016 %>% filter(Level == "Elementary")
-rm("Schools_2011_2016")
-#data(School_Catchments_2011_2016)
-load(file="data-products/TTS_gamma2016_motor.Rdata")
-load(file="data-products/TTS_exp2016_nonmotor.Rdata")
-```
-
-```{r, include=FALSE}
-#add GTA06 IDs to schools - "GTA06_Dest"}
-cols_TAZIDs <- sp::over(as_Spatial(ALL_SCHOOLS_Elem |> st_transform(st_crs(ggh_taz))),
-                              as_Spatial(ggh_taz |> dplyr::select(GTA06) |> rename("GTA06_Dest" = "GTA06")))
-ALL_SCHOOLS_Elem <- cbind(ALL_SCHOOLS_Elem,cols_TAZIDs)
-rm(cols_TAZIDs)
+rm(avail_2011,ttm_car_2011, ttm_walk_2011, od_trips_perc_2011, od_trips_perc_comm_2011,RED_LU_centroids_2011)
 ```
@@
 OD_2016 <- RED_LU_centroids_2016 %>% 
-  cross_join(ALL_SCHOOLS_Elem %>% st_drop_geometry() |> filter(Year != 2011) %>% dplyr::select(c("SchoolID", "OTGC2016", "GTA06_Dest")))
+  cross_join(SCHOOLS %>% st_drop_geometry() |> filter(Year != 2011) %>% dplyr::select(c("SchoolID", "OTGC2016", "GTA06_Dest")))
 
@@
 #reduce dimensions
-OD_2016 <- OD_2016 %>% dplyr::select(-c("LUC1_DESC", "LUC2_DESC", "res_units", "DA_dwellings", "CMA_UID", "Region.Name", "geometry","PROP_AREA", "Area..sq.km."))
+OD_2016 <- OD_2016 %>% dplyr::select(-c("LUC1_DESC", "LUC2_DESC", "res_units", "DA_dwellings", "CMA_UID", "Region Name", "geometry","PROP_AREA", "Area (sq km)"))
 
@@
 ```{r}
-#checks!, these should be the total number of kids (^the code above) -- 58265.
+#checks!, these should be the total number of kids (^the code above) -- 58865
 OD_2016$kid_pop %>% sum()

### 2024-12-12 data-prep/05-Exploring-avail_2021.qmd
changes to bib and analysis files after revision

@@
 library(sf)
+library(tidyr)
 # if(!require(remotes)){
@@
 # }
-#remotes::install_github("soukhova/HamONdests")
+# remotes::install_github("soukhova/TTS2016R")
 library(hamONdests)
@@
   group_by(ID) %>%
-  dplyr::select(c("ID","GeoUID_2016","GeoUID","GTA06_Orig","WARD","COMMUNITY","URBAN_RURA","LUC1","LUC2","LUC1_DESCR","TYPE","geometry"))
+  dplyr::select(c("ID","GeoUID", "GeoUID_2011","GeoUID_2016",
+                  "GTA06_Orig","WARD","COMMUNITY","URBAN_RURA","LUC1","LUC2","LUC1_DESCR","TYPE","geometry"))
 ```
@@
 avail_2021 <- sp_avail_detailed(OD_2021, ID, SchoolID, kid_popadj, OTGC2021, mode, r, f, alpha = 1)
+```
+
+```{r}
 save(avail_2021, file="data-products/avail_2021.RData")
@@
 avail_2021$V_ij |> sum(na.rm=T)
+#should be  57696.08!
 ```
@@
 ```
+
+Also interesting! The parcels have "urban vs. rural" in 2021... let's summarise this per DA and save in data_products.
+```{r}
+load(file="data-inputs/Parcels/RED_LU_centroids_2021.RData") #parcels
+
+DA_2021_URBvsRUR <- RED_LU_centroids_2021 |> st_drop_geometry() |>
+  mutate(URB = ifelse(URBAN_RURA == "U", 1, 0),
+         SUB = ifelse(URBAN_RURA == "S", 1, 0),
+         RUR = ifelse(URBAN_RURA == "R", 1, 0)) |>
+  group_by(GeoUID) |> summarise(count_urb= sum(URB),
+                                count_sub = sum(SUB),
+                                count_rur = sum(RUR))
+
+DA_2021_URBvsRUR <- DA_2021_URBvsRUR |>
+  transmute(GeoUID,
+            perc_urb = count_urb/rowSums(DA_2021_URBvsRUR[2:4]))
+
+save(DA_2021_URBvsRUR, file="data-products/DA_2021_URBvsRUR.RData")
+```

### 2024-12-12 paper/Manuscript.qmd
draft revised manuscript - for other author reviews

@@
 ---
-title: "School closures and consolidations: the active travel and emission implications of reduced accessibility"
+title: "10 years of school closures and consolidations: the impact on school accessibility"
 bibliography: 
@@
 abstract: | 
-  | Reducing emissions and increasing active travel are top priorities for many communities.
-Often, though, these priorities are confronted with pressures to reduce the fiscal burden of public services, including schools.
-Public schools have been strained to do more with less, and governments are opting to close under-capacity schools and expand existing schools to accommodate lost capacity.
-The objective of this study is to quantify the travel burdens of school consolidation/closure policies.
-We do this through the case of Hamilton, a mid-sized city in Ontario, Canada.
-Between the years 2011 and 2015, 7% of the elementary schools closed in the city.
-We study how these closures changed the accessibility landscape for the elementary school aged population, as well as the active travel and emission implications.
-In terms of methods, we use spatial availability, a singly-constrained multimodal measure of spatial accessibility.
-Thanks to its proportional allocation mechanism, spatial availability allows for an interpretable comparison between 'before' and 'after' the implementation of school closure/consolidation policy.
-Analysis is conducted at the parcel-level with all school seats being eligible and travel behaviour based on observed home-to-school patterns for motorized and non-motorized modes.
-We report estimates of system-wide impacts that result from the closures, such as potential growth in GHG emission and reduction of walking minutes.
-Findings demonstrate that overall, spatial availability decreased in the city.
-Though the school consolidation policy may have resulted in cost-savings in operation and maintenance of facilities, students, families, and communities pay the price in reduced opportunities for physical activity and greater motorized mobility-related pollution.
+  | Reducing motorized travel and increasing active travel are top priorities for many communities.
+However, these priorities are confronted with pressures to reduce the fiscal burden of public services, including schools.
+Public school boards have been strained to do more with less, and opting to close under-capacity schools to deliver educational services more cost-effectively.
+However, once a school is closed, students' proximity to schools decreases, and those who previously lived close enough to walk or cycle can no longer do so.
+In this paper, the City of Hamilton, a mid-sized city in Ontario, Canada, which experienced a 10% decrease in elementary school locations between 2011 and 2021, serves as a case study.
+We quanitfiy how these closures and consolidations have reshaped accessibility for the elementary school-aged population and discuss their implications for active and motorized travel.
+To measure accessibility, spatial availability, a singly-constrained multimodal measure of spatial accessibility is used.
+This method's proportional allocation mechanism enables interpretable comparison between accessibility 'before' and 'after' the implementation of school closure/consolidation policy considering changes in potential travel time and population.
+The analysis is conducted at the parcel level, considering all school seats as eligible and incorporating observed home-to-school travel patterns for motorized and non-motorized modes.
+We present system-wide impact estimates resulting from these closures.
+The findings reveal that overall spatial availability in the city declined.
+While the school consolidation policy may have achieved cost savings in facility operation and maintenance, the burden falls on students, families, and communities through _potentially_ reduced opportunities for physical activity through an active journey to school and increased motorized travel externalities.
+
 keywords: [accessibility; spatial availability; carbon emissions; journey-to-school; walkability; active transportation; school consolidation policy; service provision thresholds]
@@
 Each parcel is assigned a DA-pop/(the number of potential parcel trips in a DA * the potential parcels trips from that parcel).
-This is highly correlated with we just weighted it by DA-pop/DA-dwellings (but need to double check).
+This is highly correlated with the value just weighted by DA-pop/DA-dwellings (but need to double check).
 --->
@@
 library(miscTools) # Miscellaneous Tools and Utilities
-library(huxtable)
+# library(huxtable)
 library(tiff)
 library(grid)
+library(cowplot)
+library(flextable)
+library(gt)
 ```
@@
                           quiet = TRUE) |> st_transform(st_crs(cen_data_2016_hamiltononly))
+
+#Add % urbanization to the label for ham_community_bounds
+# SAvail_im_DA <- left_join(SAvail_im_DA, DA_2021_URBvsRUR, by="GeoUID")
+# 
+# SAvail_im_DA |> filter(mode == "Motorized") |> st_drop_geometry() |> group_by(COMMUNITY) |> 
+#   summarise(perc_urb = mean(perc_urb, na.rm=TRUE))
+
 ham_community_bounds <- st_read(paste0(here::here(),"/data-prep/data-inputs/Boundaries/Community_Boundaries.shp"),quiet = TRUE) |> 
@@
                                COMMUNITY_ == "Hamilton" ~ "HA",
-                               COMMUNITY_ == "Stoney Creek" ~ "SC"),
-         COMMUNITY_ = ifelse(COMMUNITY_ == "Hamilton", "Hamilton Central", COMMUNITY_)) |> st_transform(st_crs(cen_data_2016_hamiltononly))
+                               COMMUNITY_ == "Stoney Creek" ~ "SC")) |> st_transform(st_crs(cen_data_2016_hamiltononly)) |> 
+  mutate(COMMUNITY_URB = case_when(COMMUNITY == "AN" ~ "Ancaster (87%)",
+                               COMMUNITY == "DU" ~ "Dundas (96%)",
+                               COMMUNITY == "FL" ~ "Flamborough (37%)",
+                               COMMUNITY == "GL" ~ "Glanbrook (47%)",
+                               COMMUNITY == "HA" ~ "Hamilton Central (100%)" ,
+                               COMMUNITY == "SC" ~"Stoney Creek (94%)"),
+         COMMUNITY_ = ifelse(COMMUNITY=="HA", "Hamilton Central", COMMUNITY_))
 
@@
          dplyr::select(-c("DA_pop_5to9","DA_pop_10to14"))
+
+#adding $ URBAN to the 2021 census DA file... retrieved from parcel points. No other year of parcel points provide this information 
+load(paste0(here::here(),file="/data-prep/data-products/DA_2021_URBvsRUR.Rdata"))
+cen_data_2021_hamiltononly <- cen_data_2021_hamiltononly |> left_join(DA_2021_URBvsRUR, by="GeoUID")
 ```
@@
 
-```{r augmenting-catchment, include=FALSE}
-# ALL_SCHOOLS_CATCH_Elem_extra2011 <- ALL_SCHOOLS_CATCH_Elem |>
-#   left_join(ALL_SCHOOLS_Elem |> st_drop_geometry() |> group_by(CID2011) |> 
-#               summarize(OTGC = sum(OTGC2011, na.rm = T)), 
-#             by = c("CID" = "CID2011")) |> filter(!is.na(OTGC))
-# 
-# ALL_SCHOOLS_CATCH_Elem_extra2016 <- ALL_SCHOOLS_CATCH_Elem |>
-#   left_join(ALL_SCHOOLS_Elem |> st_drop_geometry() |> group_by(CID2016) |> 
-#               summarize(OTGC = sum(OTGC2016, na.rm = T)), 
-#             by = c("CID" = "CID2016")) |> filter(!is.na(OTGC))
-# 
-# ALL_SCHOOLS_CATCH_Elem_extra <- rbind(ALL_SCHOOLS_CATCH_Elem_extra2011, ALL_SCHOOLS_CATCH_Elem_extra2016)
-# rm(ALL_SCHOOLS_CATCH_Elem_extra2011,ALL_SCHOOLS_CATCH_Elem_extra2016)
-# 
-# #checks
-# sum(ALL_SCHOOLS_Elem$OTGC2011, na.rm=T) + sum(ALL_SCHOOLS_Elem$OTGC2016, na.rm=T)
-# ALL_SCHOOLS_CATCH_Elem_extra$OTGC |> sum(na.rm=T)
-
+```{r formatting-schools-for-mapping, include=FALSE}
 #reformatting schools for mapping
-SCHOOLS_nochangeorexpanded <- SCHOOLS |> filter(Year == "2011 and 2016" & Year_2 == "2016 and 2021") |> 
-  mutate(Year = ifelse(Year == "2011 and 2016", "2011", Year))
-ALL_SCHOOLS_Elem_formapping <- rbind(SCHOOLS, SCHOOLS_nochangeorexpanded)
-rm(SCHOOLS_nochangeorexpanded)
-ALL_SCHOOLS_Elem_formapping <- ALL_SCHOOLS_Elem_formapping |> mutate(Year = ifelse(Year == "2011 and 2016", "2016", Year))
-```
-
-```{r descriptive-catchment-schools-plot, eval=FALSE}
+ALL_SCHOOLS_Elem_formapping_11 <- SCHOOLS |>
+  filter(Year == "2011 and 2016" & Year_2 == "2016 and 2021" & Status == "NoChange" & Status_2 == "NoChange") |>
+  mutate(combo_status = "1) No change",
+         combo_year = "2011",
+         OTGC = OTGC2011) 
+ALL_SCHOOLS_Elem_formapping_12 <- SCHOOLS |>
+  filter(Year == "2011 and 2016" & Year_2 == "2016 and 2021" & Status == "NoChange" & Status_2 == "NoChange") |>
+  mutate(combo_status = "1) No change",
+         combo_year = "2016",
+         OTGC = OTGC2016) 
+ALL_SCHOOLS_Elem_formapping_13 <- SCHOOLS |>
+  filter(Year == "2011 and 2016" & Year_2 == "2016 and 2021" & Status == "NoChange" & Status_2 == "NoChange") |>
+  mutate(combo_status = "1) No change",
+         combo_year = "2021",
+         OTGC = OTGC2021)
+
+ALL_SCHOOLS_Elem_formapping_21 <- SCHOOLS |>
+  filter(Year == "2011 and 2016" & Year_2 == "2016 and 2021" & Status == "Expanded" & Status_2 == "NoChange") |>
+  mutate(combo_status = "2) Expanded between 2011-16",
+         combo_year = "2011",
+         OTGC = OTGC2011)
+ALL_SCHOOLS_Elem_formapping_22 <- SCHOOLS |>
+  filter(Year == "2011 and 2016" & Year_2 == "2016 and 2021" & Status == "Expanded" & Status_2 == "NoChange") |>
+  mutate(combo_status = "2) Expanded between 2011-16",
+         combo_year = "2016",
+         OTGC = OTGC2016)
+ALL_SCHOOLS_Elem_formapping_23 <- SCHOOLS |>
+  filter(Year == "2011 and 2016" & Year_2 == "2016 and 2021" & Status == "Expanded" & Status_2 == "NoChange") |>
+  mutate(combo_status = "2) Expanded between 2011-16",
+         combo_year = "2021",
+         OTGC = OTGC2021)
+
+ALL_SCHOOLS_Elem_formapping_31 <- SCHOOLS |>
+  filter(Year == "2016" & Year_2 == "2016 and 2021" & Status == "Moved" & Status_2 == "NoChange") |>
+  mutate(combo_status = "3) Relocated + expanded between 2011-16",
+         combo_year = "2016",
+         OTGC = OTGC2016)
+ALL_SCHOOLS_Elem_formapping_32 <- SCHOOLS |>
+  filter(Year == "2016" & Year_2 == "2016 and 2021" & Status == "Moved" & Status_2 == "NoChange") |>
+  mutate(combo_status = "3) Relocated + expanded between 2011-16",
+         combo_year = "2021",
+         OTGC = OTGC2021)
+
+ALL_SCHOOLS_Elem_formapping_41 <- SCHOOLS |>
+  filter(Year == "2011" & is.na(Year_2) & Status == "Moved" & is.na(Status_2)) |>
+  mutate(combo_status = "3) Relocated + expanded between 2011-16",
+         combo_year = "2011",
+         OTGC = OTGC2011)
+
+ALL_SCHOOLS_Elem_formapping_51 <- SCHOOLS |>
+  filter(Year == "2016" & Year_2 == "2016 and 2021" & Status == "New" & Status_2 == "NoChange") |>
+  mutate(combo_status = "6) Opened between 2011-16",
+         combo_year = "2016",
+         OTGC = OTGC2016) 
+ALL_SCHOOLS_Elem_formapping_52 <- SCHOOLS |>
+  filter(Year == "2016" & Year_2 == "2016 and 2021" & Status == "New" & Status_2 == "NoChange") |>
+  mutate(combo_status = "6) Opened between 2011-16",
+         combo_year = "2021",
+         OTGC = OTGC2021)
+
+ALL_SCHOOLS_Elem_formapping_81 <- SCHOOLS |>
+  filter(is.na(Year) & Year_2 == "2021" & is.na(Status) & Status_2 == "New") |>
+  mutate(combo_status = "7) Opened between 2016-21",
+         combo_year = "2021",
+         OTGC = OTGC2021)
+
+ALL_SCHOOLS_Elem_formapping_71 <- SCHOOLS |>
+  filter(Year == "2011" & is.na(Year_2) & Status == "Removed" & is.na(Status_2)) |>
+  mutate(combo_status = "4) Closed between 2011-16",
+         combo_year = "2011",
+         OTGC = OTGC2011)
+
+ALL_SCHOOLS_Elem_formapping_61 <- SCHOOLS |>
+  filter(Year == "2011 and 2016" & Year_2 == "2016" & Status == "NoChange" & Status_2 == "Removed") |>
+  mutate(combo_status = "5) Closed between 2016-21",
+         combo_year = "2011",
+         OTGC = OTGC2011)
+
+ALL_SCHOOLS_Elem_formapping_62 <- SCHOOLS |>
+  filter(Year == "2011 and 2016" & Year_2 == "2016" & Status == "NoChange" & Status_2 == "Removed") |>
+  mutate(combo_status = "5) Closed between 2016-21",
+         combo_year = "2016",
+         OTGC = OTGC2016)
+
+ALL_SCHOOLS_Elem_formapping <- rbind(
+  ALL_SCHOOLS_Elem_formapping_11,ALL_SCHOOLS_Elem_formapping_12, ALL_SCHOOLS_Elem_formapping_13,
+  ALL_SCHOOLS_Elem_formapping_21,ALL_SCHOOLS_Elem_formapping_22,ALL_SCHOOLS_Elem_formapping_23, 
+  ALL_SCHOOLS_Elem_formapping_31,ALL_SCHOOLS_Elem_formapping_32, 
+  ALL_SCHOOLS_Elem_formapping_41,
+  ALL_SCHOOLS_Elem_formapping_51,ALL_SCHOOLS_Elem_formapping_52,
+  ALL_SCHOOLS_Elem_formapping_61,ALL_SCHOOLS_Elem_formapping_62,
+  ALL_SCHOOLS_Elem_formapping_71,
+  ALL_SCHOOLS_Elem_formapping_81
+  )
+
+rm(ALL_SCHOOLS_Elem_formapping_11,ALL_SCHOOLS_Elem_formapping_12, ALL_SCHOOLS_Elem_formapping_13,
+  ALL_SCHOOLS_Elem_formapping_21,ALL_SCHOOLS_Elem_formapping_22,ALL_SCHOOLS_Elem_formapping_23, 
+  ALL_SCHOOLS_Elem_formapping_31,ALL_SCHOOLS_Elem_formapping_32, 
+  ALL_SCHOOLS_Elem_formapping_41,
+  ALL_SCHOOLS_Elem_formapping_51,ALL_SCHOOLS_Elem_formapping_52,
+  ALL_SCHOOLS_Elem_formapping_61,ALL_SCHOOLS_Elem_formapping_62,
+  ALL_SCHOOLS_Elem_formapping_71,
+  ALL_SCHOOLS_Elem_formapping_81)
+```
+
+```{r descriptive-schools-plot, eval=FALSE}
 tmap_defaults <-
@@
 
-catchment_schools_plot <-
-   tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0) +
-  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0)+
-  tm_shape(ALL_SCHOOLS_CATCH_Elem_extra) +
-  tm_polygons(col = "OTGC", border.alpha=0.1, palette = "RdPu",
-              labels = c("[181-318)", "[318-406)", "[406-566)","[566-1417]"),
-              breaks = c(181, 318, 406, 566, 1417),
-              title = "On the ground\ncapacity (OTGC)")+
-  tm_facets(by=c("System","Year"))+
+schools_plot <-
+  tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0) +
+  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0)+
+
+  tm_shape(cen_data_2021_hamiltononly |> filter(!is.na(perc_urb))) +
+  tm_polygons(col = "perc_urb", border.alpha=0, 
+              alph=0.8,
+              palette = c("yellow3","yellow4"),
+              #labels = c("[0-50%)", "[25-50%)", "[50-75%)","[75-100%]"),
+              breaks = c(0,  0.25, .5, .75,  1),
+              legend.is.portrait = FALSE,
+              style="cont",
+              title = "Proportion urban land-use") +
+
   tm_shape(ALL_SCHOOLS_Elem_formapping) +
-  tm_symbols(shape = "Status",
-             size=0.15,
+  tm_symbols(shape = "combo_status",
+             col="grey85",
+             size= "OTGC",
+             alpha=0.9,
+             scale = 0.4,
+             shapes = c(1,
+                        17,24,
+                        15,22,
+                        18, 23),
           title.shape = "School status",
-          shapes = c(24,25,21,23,22),
+          title.size = "OTGC",
           shapes.labels =  c(
-          "Expanded after 2011",
-          "Moved before 2016",
-          "New after 2011",
-          "No change btwn 2011-16",
-          "Closed before 2016")
-  ) +
-  tm_facets(by=c("System","Year"))+
-  tm_credits(c("52 schools\n(18,771 OTGC)","49 schools\n(18,039 OTGC)",
-                "95 schools\n(41,796 OTGC)","88 schools\n(40,249 OTGC)"),
+          "No change (btwn 2011-21)", 
+          "Expanded (btwn 2011-16)", 
+          "Relocated + expanded (btwn 2011-16)",
+          "Closed (btwn 2011-16)", 
+          "Closed (btwn 2016-21)",
+          "Opened (btwn 2011-16)", 
+          "Opened (btwn 2016-21)")) +
+  tm_facets(by=c("System","combo_year"), nrow = 3)+
+
+  tm_credits(c("52 schools open\n(18,771 OTGC)","49 schools open\n(18,039 OTGC)", "49 schools open\n(18,218 OTGC)",
+                "95 schools open\n(41,796 OTGC)","88 schools open\n(40,249 OTGC)","83 schools open\n(39,478 OTGC)"),
              position = c("LEFT", "BOTTOM")) +
+  tm_shape(ham_city_bound)+  tm_borders(col = "grey80", lwd=0.5)+
   tmap_defaults +
-  tm_layout(bg.color = "grey85",
+  tm_layout(bg.color = "grey99",
             panel.labels = list(c('HWDCSB (Catholic Public)','HWDSB (Public)'),
-                                c('2011','2016')),
+                                c('2011','2016','2021')),
             panel.label.bg.color = "white")
 
-catchment_schools_plot
-```
-
-```{r descriptive-catchment-schools-plot2, eval=FALSE}
-tmap_save(catchment_schools_plot, paste0(here::here(),"/Figures/Fig1.tiff"), height=7, width=12, dpi=600)
+tmap_save(schools_plot, paste0(here::here(),"/Figures/Fig2.tiff"), height=4.5, width=10, dpi=600)
 ```
@@
 load(file = paste0(here::here(),"/data-prep/data-products/avail_2016.RData"))
+load(file = paste0(here::here(),"/data-prep/data-products/avail_2021.RData"))
 ```
@@
 avail_2011$V_ij %>% sum()
-ALL_SCHOOLS_Elem$OTGC2011 %>% sum(na.rm = TRUE)
+SCHOOLS$OTGC2011 %>% sum(na.rm = TRUE)
 
 avail_2016$V_ij %>% sum()
-ALL_SCHOOLS_Elem$OTGC2016 %>% sum(na.rm = TRUE)
+SCHOOLS$OTGC2016 %>% sum(na.rm = TRUE)
+
+avail_2021$V_ij %>% sum()
+SCHOOLS$OTGC2021 %>% sum(na.rm = TRUE)
 
@@
 avail_2016 %>%
+  group_by(ID) %>%
+  summarize(kid_popadj = sum(kid_popadj)) |> dplyr::select(kid_popadj) %>% sum()
+
+#this should equal 62195, still does -- great!
+avail_2021 %>%
   group_by(ID) %>%
@@
 
+#calc spatial availability and other associated variables at the DA 2021 level
+SAvail_im_DA2021 <- avail_2021 %>%
+  group_by(GeoUID, 
+           mode) %>%
+  summarize(COMMUNITY = first(COMMUNITY),
+            total_tt = sum(travel_time_p50*kid_popadj),
+            V_im = sum(V_ij),
+            kid_popadj_m = sum(kid_popadj),
+            DA_pop_5to14 = mean(DA_pop_5to14, na.rm=T),
+            DA_prev_LIMAT_under18 = mean(DA_prev_LIMAT_under18),
+            .groups = "drop") |>
+  mutate(emissions_total_car_tt = ifelse(mode == "c", 0.1*total_tt, 0),
+         v_im_kid_popadj = V_im/kid_popadj_m,
+         v_im_kid_popadj = ifelse(is.na(v_im_kid_popadj) | is.nan(v_im_kid_popadj), 0, v_im_kid_popadj),
+         v_im_kid_popadj_LIMAT = v_im_kid_popadj*(DA_prev_LIMAT_under18/100),
+         v_im_kid_popadj_LIMAT = ifelse(is.na(v_im_kid_popadj_LIMAT) | is.nan(v_im_kid_popadj_LIMAT), 0, v_im_kid_popadj_LIMAT),
+         V_i_kid_popadj_LIMAT = V_im*(DA_prev_LIMAT_under18/100),
+         V_i_kid_popadj_LIMAT = ifelse(is.na(V_i_kid_popadj_LIMAT) | is.nan(V_i_kid_popadj_LIMAT), 0, V_i_kid_popadj_LIMAT)) |>
+  filter(GeoUID != 35250211 & GeoUID != 35250242 & GeoUID != 35250491 & GeoUID != 35250982 ) #these GeoUID's have 0 kid population, so let's drop so we don't have zeros
+```
+
+```{r}
 #add the geometries -- census DAs
 SAvail_im_DA2011 <- SAvail_im_DA2011 |>
-  left_join(cen_data_2011_hamiltononly |> filter(Region.Name == "Hamilton") |> dplyr::select(GeoUID), by="GeoUID") |> st_set_geometry("geometry")
+  left_join(cen_data_2011_hamiltononly |> dplyr::select(GeoUID), by="GeoUID") |> st_set_geometry("geometry")
 
 SAvail_im_DA2016 <- SAvail_im_DA2016 |>
-  left_join(cen_data_2016_hamiltononly |> filter(Region.Name == "Hamilton") |> dplyr::select(GeoUID), by="GeoUID") |> st_set_geometry("geometry")
+  left_join(cen_data_2016_hamiltononly |> dplyr::select(GeoUID), by="GeoUID") |> st_set_geometry("geometry")
+
+SAvail_im_DA2021 <- SAvail_im_DA2021 |>
+  left_join(cen_data_2021_hamiltononly |> dplyr::select(GeoUID), by="GeoUID") |> st_set_geometry("geometry")
 
@@
   rbind(SAvail_im_DA2011 |> mutate(year ="2011"),
-        SAvail_im_DA2016 |> mutate(year ="2016")) |>
-  mutate(mode = case_when(mode == "c" ~ "motorized",
-                          mode == "w" ~ "non-motorized"))
+        SAvail_im_DA2016 |> mutate(year ="2016"),
+        SAvail_im_DA2021 |> mutate(year ="2021")) |>
+  mutate(mode = case_when(mode == "c" ~ "Motorized",
+                          mode == "w" ~ "Non-motorized"))
 ```
@@
 ```{r}
-#calc changes between the years - but at the 2016 DA level 
-SAvail_im_DA2011_forGeoUID2016 <- avail_2011 %>%
-  group_by(GeoUID_2016, 
+#calc changes between the years - but do it all at the 2021 DA grid. So let's calculate all SAvails at the level of the 2021 grid.
+
+#first for 2011 @ 2021 grid
+SAvail_im_DA2011_forGeoUID2021 <- avail_2011 %>%
+  group_by(GeoUID_2021, 
            mode) %>%
@@
 
-SAvail_im_DA_changes <- full_join(SAvail_im_DA2011_forGeoUID2016,
-                                  SAvail_im_DA2016, by=c("GeoUID_2016"="GeoUID", 
+#then for 2016 @ 2021 grid
+SAvail_im_DA2016_forGeoUID2021 <- avail_2016 %>%
+  group_by(GeoUID_2021, 
+           mode) %>%
+  summarize(COMMUNITY = first(COMMUNITY),
+            total_tt = sum(travel_time_p50*kid_popadj),
+            V_im = sum(V_ij),
+            kid_popadj_m = sum(kid_popadj),
+            DA_pop_5to14 = mean(DA_pop_5to14, na.rm=T),
+            DA_prev_LIMAT_under18 = mean(DA_prev_LIMAT_under18),
+            .groups = "drop") |>
+  mutate(emissions_total_car_tt = ifelse(mode == "c", 0.1*total_tt, 0),
+         v_im_kid_popadj = V_im/kid_popadj_m,
+         v_im_kid_popadj = ifelse(is.na(v_im_kid_popadj) | is.nan(v_im_kid_popadj), 0, v_im_kid_popadj),
+         v_im_kid_popadj_LIMAT = v_im_kid_popadj*(DA_prev_LIMAT_under18/100),
+         v_im_kid_popadj_LIMAT = ifelse(is.na(v_im_kid_popadj_LIMAT) | is.nan(v_im_kid_popadj_LIMAT), 0, v_im_kid_popadj_LIMAT),
+         V_i_kid_popadj_LIMAT = V_im*(DA_prev_LIMAT_under18/100),
+         V_i_kid_popadj_LIMAT = ifelse(is.na(V_i_kid_popadj_LIMAT) | is.nan(V_i_kid_popadj_LIMAT), 0, V_i_kid_popadj_LIMAT))
+
+#then for 2021 @ 2021 grid
+SAvail_im_DA2021_forGeoUID2021 <- avail_2021 %>%
+  group_by(GeoUID, 
+           mode) %>%
+  summarize(COMMUNITY = first(COMMUNITY),
+            total_tt = sum(travel_time_p50*kid_popadj),
+            V_im = sum(V_ij),
+            kid_popadj_m = sum(kid_popadj),
+            DA_pop_5to14 = mean(DA_pop_5to14, na.rm=T),
+            DA_prev_LIMAT_under18 = mean(DA_prev_LIMAT_under18),
+            .groups = "drop") |>
+  mutate(emissions_total_car_tt = ifelse(mode == "c", 0.1*total_tt, 0),
+         v_im_kid_popadj = V_im/kid_popadj_m,
+         v_im_kid_popadj = ifelse(is.na(v_im_kid_popadj) | is.nan(v_im_kid_popadj), 0, v_im_kid_popadj),
+         v_im_kid_popadj_LIMAT = v_im_kid_popadj*(DA_prev_LIMAT_under18/100),
+         v_im_kid_popadj_LIMAT = ifelse(is.na(v_im_kid_popadj_LIMAT) | is.nan(v_im_kid_popadj_LIMAT), 0, v_im_kid_popadj_LIMAT),
+         V_i_kid_popadj_LIMAT = V_im*(DA_prev_LIMAT_under18/100),
+         V_i_kid_popadj_LIMAT = ifelse(is.na(V_i_kid_popadj_LIMAT) | is.nan(V_i_kid_popadj_LIMAT), 0, V_i_kid_popadj_LIMAT))
+```
+
+```{r}
+#Now calculate the changes.. first 2011 to 2016, then 2016 to 2021 (are the trends similar?), then 2011 to 2021 (are the trends similar)
+SAvail_im_DA_changes_2011_16 <- full_join(SAvail_im_DA2011_forGeoUID2021,
+                                  SAvail_im_DA2016_forGeoUID2021,
+                                  by=c("GeoUID_2021",
                                        "mode")) %>%
-  transmute(GeoUID_2016,
+  transmute(GeoUID_2021,
             mode = case_when(mode == "c" ~ "motorized",
@@
             c_V_i_kid_popadj_LIMAT = V_i_kid_popadj_LIMAT.y - V_i_kid_popadj_LIMAT.x) |>
-  left_join(cen_data_2016_hamiltononly|>dplyr::select(c("GeoUID")), 
-            by=c("GeoUID_2016" = "GeoUID")) %>% st_as_sf
-
+  left_join(cen_data_2021_hamiltononly|>dplyr::select(c("GeoUID")), 
+            by=c("GeoUID_2021" = "GeoUID")) %>% st_as_sf()
+
+# then 2016 to 2021
+SAvail_im_DA_changes_2016_21 <- full_join(SAvail_im_DA2016_forGeoUID2021,
+                                  SAvail_im_DA2021_forGeoUID2021,
+                                  by=c("GeoUID_2021"="GeoUID",
+                                       "mode")) %>%
+  transmute(GeoUID_2021,
+            mode = case_when(mode == "c" ~ "motorized",
+                             mode == "w" ~ "non-motorized"),
+            COMMUNITY = dplyr::coalesce(COMMUNITY.x, COMMUNITY.y),
+            c_total_tt = total_tt.y - total_tt.x, #2021 minus 2016 tt
+            c_total_tt = ifelse(is.na(c_total_tt), -total_tt.x, c_total_tt),
+            c_total_ttperkid = (total_tt.y/kid_popadj_m.y) - (total_tt.x/kid_popadj_m.x),
+            c_total_ttperkid = ifelse(is.na(c_total_ttperkid), -(total_tt.x/kid_popadj_m.x), c_total_ttperkid),
+            c_V_im = V_im.y -V_im.x,
+            c_V_im = ifelse(is.na(c_V_im), -V_im.x,c_V_im),
+            c_V_im_perc = (V_im.y -V_im.x)/V_im.x,
+            c_kid_popadj_m = kid_popadj_m.y - kid_popadj_m.x,
+            c_DA_pop_5to14 = DA_pop_5to14.y - DA_pop_5to14.x,
+            c_DA_prev_LIMAT_under18 = DA_prev_LIMAT_under18.y - DA_prev_LIMAT_under18.x,
+            emissions_total_car_tt_2021 = emissions_total_car_tt.y,
+            emissions_total_car_tt_2016 = emissions_total_car_tt.x,
+            c_emissions_total_car_tt = emissions_total_car_tt.y - emissions_total_car_tt.x,
+            c_emissions_total_car_tt = ifelse(is.na(c_emissions_total_car_tt), -emissions_total_car_tt.x, c_emissions_total_car_tt),
+            emissions_perkid_2016 = emissions_total_car_tt.x/kid_popadj_m.x,
+            emissions_perkid_2021 = emissions_total_car_tt.y/kid_popadj_m.y,
+            c_emissions_perkid = ((emissions_perkid_2021 - emissions_perkid_2016)/
+                                      emissions_perkid_2016),
+            c_emissions_perkid = ifelse(is.na(emissions_perkid_2021), -emissions_perkid_2016, c_emissions_perkid),
+            c_v_im_kid_popadj = v_im_kid_popadj.y - v_im_kid_popadj.x,
+            c_v_im_kid_popadj = ifelse(is.na(c_v_im_kid_popadj), -v_im_kid_popadj.x, c_v_im_kid_popadj),
+            c_v_im_kid_popadj_perc = (v_im_kid_popadj.y - v_im_kid_popadj.x)/v_im_kid_popadj.x,
+            c_v_im_kid_popadj_LIMAT = v_im_kid_popadj_LIMAT.y - v_im_kid_popadj_LIMAT.x,
+            c_V_i_kid_popadj_LIMAT = V_i_kid_popadj_LIMAT.y - V_i_kid_popadj_LIMAT.x) |>
+  left_join(cen_data_2021_hamiltononly|>dplyr::select(c("GeoUID")), 
+            by=c("GeoUID_2021" = "GeoUID")) %>% st_as_sf()
+
+#then, 2011 compared to 2021
+SAvail_im_DA_changes_2011_21 <- full_join(SAvail_im_DA2011_forGeoUID2021,
+                                  SAvail_im_DA2021_forGeoUID2021,
+                                  by=c("GeoUID_2021"="GeoUID",
+                                       "mode")) %>%
+  transmute(GeoUID_2021,
+            mode = case_when(mode == "c" ~ "motorized",
+                             mode == "w" ~ "non-motorized"),
+            COMMUNITY = dplyr::coalesce(COMMUNITY.x, COMMUNITY.y),
+            c_total_tt = total_tt.y - total_tt.x, #2021 minus 2011 tt
+            c_total_tt = ifelse(is.na(c_total_tt), -total_tt.x, c_total_tt),
+            c_total_ttperkid = (total_tt.y/kid_popadj_m.y) - (total_tt.x/kid_popadj_m.x),
+            c_total_ttperkid = ifelse(is.na(c_total_ttperkid), -(total_tt.x/kid_popadj_m.x), c_total_ttperkid),
+            c_V_im = V_im.y -V_im.x,
+            c_V_im = ifelse(is.na(c_V_im), -V_im.x,c_V_im),
+            c_V_im_perc = (V_im.y -V_im.x)/V_im.x,
+            c_kid_popadj_m = kid_popadj_m.y - kid_popadj_m.x,
+            c_DA_pop_5to14 = DA_pop_5to14.y - DA_pop_5to14.x,
+            c_DA_prev_LIMAT_under18 = DA_prev_LIMAT_under18.y - DA_prev_LIMAT_under18.x,
+            emissions_total_car_tt_2021 = emissions_total_car_tt.y,
+            emissions_total_car_tt_2011 = emissions_total_car_tt.x,
+            c_emissions_total_car_tt = emissions_total_car_tt.y - emissions_total_car_tt.x,
+            c_emissions_total_car_tt = ifelse(is.na(c_emissions_total_car_tt), -emissions_total_car_tt.x, c_emissions_total_car_tt),
+            emissions_perkid_2011 = emissions_total_car_tt.x/kid_popadj_m.x,
+            emissions_perkid_2021 = emissions_total_car_tt.y/kid_popadj_m.y,
+            c_emissions_perkid = ((emissions_perkid_2021 - emissions_perkid_2011)/
+                                      emissions_perkid_2011),
+            c_emissions_perkid = ifelse(is.na(emissions_perkid_2021), -emissions_perkid_2011, c_emissions_perkid),
+            c_v_im_kid_popadj = v_im_kid_popadj.y - v_im_kid_popadj.x,
+            c_v_im_kid_popadj = ifelse(is.na(c_v_im_kid_popadj), -v_im_kid_popadj.x, c_v_im_kid_popadj),
+            c_v_im_kid_popadj_perc = (v_im_kid_popadj.y - v_im_kid_popadj.x)/v_im_kid_popadj.x,
+            c_v_im_kid_popadj_LIMAT = v_im_kid_popadj_LIMAT.y - v_im_kid_popadj_LIMAT.x,
+            c_V_i_kid_popadj_LIMAT = V_i_kid_popadj_LIMAT.y - V_i_kid_popadj_LIMAT.x) |>
+  left_join(cen_data_2021_hamiltononly|>dplyr::select(c("GeoUID")), 
+            by=c("GeoUID_2021" = "GeoUID")) %>% st_as_sf()
 ```
@@
 ```{r}
-#repeat but for all 1 mode
+#repeat the above, but for all 1 mode
 SAvail_i_DA2011 <- avail_2011 %>%
@@
   filter(GeoUID != 35250398 & GeoUID != 35250133 & GeoUID != 35250427 ) #these GeoUID's have 0 kid population, so let's drop so we don't have zeros
+
+SAvail_i_DA2021 <- avail_2021 %>%
+  group_by(GeoUID) %>%
+  summarize(COMMUNITY = first(COMMUNITY),
+            total_tt = sum(travel_time_p50*kid_popadj),
+            V_i = sum(V_ij, na.rm = T),
+            DA_pop_5to14 = mean(DA_pop_5to14, na.rm=T),
+            kid_pop_DAavg = mean(kid_pop, na.rm = T),
+            DA_prev_LIMAT_under18 = mean(DA_prev_LIMAT_under18, na.rm = T),
+            .groups = "drop") |>
+  filter(GeoUID != 35250211 & GeoUID != 35250242 & GeoUID != 35250491 & GeoUID != 35250982 ) #these GeoUID's have 0 kid population, so let's drop so we don't have zeros
+
 #add the geometries -- census DAs
 SAvail_i_DA2011 <- SAvail_i_DA2011 |>
-  left_join(cen_data_2011_hamiltononly |> filter(Region.Name == "Hamilton") |> dplyr::select(GeoUID), by="GeoUID") |> mutate(Year = "2011") |> st_set_geometry("geometry")
+  left_join(cen_data_2011_hamiltononly  |> dplyr::select(GeoUID), by="GeoUID") |> mutate(Year = "2011") |> st_set_geometry("geometry")
+
 SAvail_i_DA2016 <- SAvail_i_DA2016 |>
-  left_join(cen_data_2016_hamiltononly |> filter(Region.Name == "Hamilton") |> dplyr::select(GeoUID), by="GeoUID") |> mutate(Year = "2016") |> st_set_geometry("geometry")
+  left_join(cen_data_2016_hamiltononly |> dplyr::select(GeoUID), by="GeoUID") |> mutate(Year = "2016") |> st_set_geometry("geometry")
+
+SAvail_i_DA2021 <- SAvail_i_DA2021 |>
+  left_join(cen_data_2021_hamiltononly |> dplyr::select(GeoUID), by="GeoUID") |> mutate(Year = "2021") |> st_set_geometry("geometry")
+
 #join them together!
-SAvail_i_DA_extra <- rbind(SAvail_i_DA2011, SAvail_i_DA2016)
-rm(SAvail_i_DA2011,SAvail_i_DA2016)
-```
-
-```{r}
-#calc changes between the years - but between communities with geom for mapping
-SAvail_im_COMM2011 <- avail_2011 %>%
-  group_by(COMMUNITY, 
-           mode) %>%
-  summarize(total_tt = sum(travel_time_p50*kid_popadj),
-            V_im = sum(V_ij),
-            kid_popadj_m = sum(kid_popadj),
-            .groups = "drop") |>
-  mutate(emissions_total_car_tt = ifelse(mode == "c", 0.1*total_tt, 0),
-         v_im_kid_popadj = V_im/kid_popadj_m)
-
-SAvail_im_COMM2016 <- avail_2016 %>%
-  group_by(COMMUNITY, 
-           mode) %>%
-  summarize(total_tt = sum(travel_time_p50*kid_popadj),
-            V_im = sum(V_ij),
-            kid_popadj_m = sum(kid_popadj),
-            .groups = "drop") |>
-  mutate(emissions_total_car_tt = ifelse(mode == "c", 0.1*total_tt, 0),
-         v_im_kid_popadj = V_im/kid_popadj_m,
-         v_im_kid_popadj = ifelse(is.na(v_im_kid_popadj) | is.nan(v_im_kid_popadj), 0, v_im_kid_popadj))
-
-#add the geometries -- census DAs
-SAvail_im_COMM2011 <- SAvail_im_COMM2011 |>
-  left_join(ham_community_bounds |> dplyr::select(c("COMMUNITY_","COMMUNITY")), by=c("COMMUNITY"))|> st_set_geometry("geometry")
-
-SAvail_im_COMM2016 <- SAvail_im_COMM2016 |>
-  left_join(ham_community_bounds |> dplyr::select(c("COMMUNITY_","COMMUNITY")), by=c("COMMUNITY"))|> st_set_geometry("geometry")
-
-#calc changes between the years - make that 1 sf
-SAvail_im_COMM_changes <- full_join(SAvail_im_COMM2011|>st_drop_geometry(),
-                                  SAvail_im_COMM2016, by=c("COMMUNITY", "mode")) |>
-  transmute(COMMUNITY,
-            mode = case_when(mode == "c" ~ "motorized",
-                             mode == "w" ~ "non-motorized"),
-            c_total_tt = total_tt.y - total_tt.x, #2016 minus 2011 tt
-            c_total_ttperkid = (total_tt.y/kid_popadj_m.y) - (total_tt.x/kid_popadj_m.x),
-            c_V_im = V_im.y -V_im.x,
-            c_V_im_perc = (V_im.y -V_im.x)/V_im.x,
-            c_kid_popadj_m = kid_popadj_m.y - kid_popadj_m.x,
-            c_emissions_total_car_tt = emissions_total_car_tt.y - emissions_total_car_tt.x,
-            c_v_im_kid_popadj = v_im_kid_popadj.y - v_im_kid_popadj.x,
-            c_v_im_kid_popadj_perc = (v_im_kid_popadj.y - v_im_kid_popadj.x)/v_im_kid_popadj.x,
-            geometry,
-            kid_2011 = kid_popadj_m.x,
-            kid_2016 = kid_popadj_m.y,
-            c_kid_perc = ((kid_popadj_m.y - kid_popadj_m.x)/kid_popadj_m.x) |> scales::percent(accuracy = 2L),
-            kid_proportion_2011 = (kid_2011/sum(kid_2011)),
-            kid_proportion_2016 = (kid_2016/sum(kid_2016)),
-            c_kid_proportion_perc = ((kid_proportion_2016 - kid_proportion_2011)/kid_proportion_2011)
-            %>% scales::percent(accuracy = 2L),
-            # kid_proportion_2011_perc = (kid_proportion_2011) |> scales::percent(accuracy = 2L),
-            # kid_proportion_2016_perc = (kid_proportion_2016) |> scales::percent(accuracy = 2L),
-            V_im_2011 = V_im.x,
-            V_im_2016 = V_im.y,
-            c_V_im_perc = ((V_im.y -V_im.x)/V_im.x) |> scales::percent(accuracy = 2L),
-            V_im_proportion_2011 = (V_im.x/sum(V_im.x)),
-            V_im_proportion_2016 = (V_im.y/sum(V_im.y)),
-            c_V_im_proportion_perc = ((V_im_proportion_2016 -V_im_proportion_2011)/V_im_proportion_2011) %>% scales::percent(accuracy = 2L),
-            # V_im_proportion_2011_perc = (V_im_proportion_2011) |> scales::percent(accuracy = 2L),
-            # V_im_proportion_2016_perc = (V_im_proportion_2016) |> scales::percent(accuracy = 2L),
-            v_im_2011 = v_im_kid_popadj.x,
-            v_im_2016 = v_im_kid_popadj.y,
-            c_v_im_perc = ((v_im_kid_popadj.y - v_im_kid_popadj.x)/v_im_kid_popadj.x) |> scales::percent(accuracy = 2L),
-            total_tt_2011 = total_tt.x,
-            total_tt_2016 = total_tt.y,
-            # total_tt_proportion_2011 = (total_tt.x/sum(total_tt.x)),
-            # total_tt_proportion_2016 = (total_tt.y/sum(total_tt.y)),
-            # c_total_tt_proportion_perc = ((total_tt_proportion_2016 - total_tt_proportion_2011)/total_tt_proportion_2011)
-            # %>% scales::percent(accuracy = 2L),
-            # total_tt_proportion_2011 = (total_tt_proportion_2011) |> scales::percent(accuracy = 2L),
-            # total_tt_proportion_2016 = (total_tt_proportion_2016) |> scales::percent(accuracy = 2L),
-            c_total_tt_perc = ((total_tt.y - total_tt.x)/
-                                  total_tt.x)|> scales::percent(accuracy = 2L),
-            total_tt_perkid_2011 = total_tt.x/kid_2011,
-            total_tt_perkid_2016 = total_tt.y/kid_2016,
-            c_total_ttperkid_perc = ((total_tt_perkid_2016 - total_tt_perkid_2011)/
-                                       total_tt_perkid_2011)|> scales::percent(accuracy = 2L),
-            emissions_2011 = emissions_total_car_tt.x,
-            emissions_2016 = emissions_total_car_tt.y,
-            c_emissions_perc = ((emissions_total_car_tt.y - emissions_total_car_tt.x)/emissions_total_car_tt.x)|> scales::percent(accuracy = 2L),
-            # emissions_proportion_2011 = (emissions_2011/sum(emissions_2011)),
-            # emissions_proportion_2016 = (emissions_2016/sum(emissions_2016)),
-            # c_emissions_proportion = ((emissions_proportion_2016 - emissions_proportion_2011)/emissions_proportion_2011)
-            # %>% scales::percent(accuracy = 2L),
-            # emissions_proportion_2011_perc = (emissions_proportion_2011) |> scales::percent(accuracy = 2L),
-            # emissions_proportion_2016_perc = (emissions_proportion_2016) |> scales::percent(accuracy = 2L),
-            emissions_perkid_2011 = emissions_total_car_tt.x/kid_popadj_m.x,
-            emissions_perkid_2016 = emissions_total_car_tt.y/kid_popadj_m.y,
-            c_emissions_perkid_perc = ((emissions_perkid_2016 - emissions_perkid_2011)/
-                                      emissions_perkid_2011)|> scales::percent(accuracy = 2L)
-            ) |> mutate_if(is.numeric, round, 1) |> st_as_sf()
+SAvail_i_DA_extra <- rbind(SAvail_i_DA2011, SAvail_i_DA2016, SAvail_i_DA2021)
+rm(SAvail_i_DA2011,SAvail_i_DA2016,SAvail_i_DA2021)
 ```
@@
 student_pop_absolute_plot <-
-    tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0) +
-  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0)+
-  tm_shape(SAvail_i_DA_extra |> mutate(extra_col_for_panel = "HAMILTON")) +
+  tm_shape(SAvail_i_DA_extra |> mutate(extra_col_for_panel = "HAMILTON") |> st_as_sf()) +
   tm_polygons(col = "DA_pop_5to14", border.alpha=0, 
               palette = "Greens",
-              labels = c("[0-40)", "[40-50)", "[50-75)","[75-835]"),
-              breaks = c(0, 40, 50, 75, 835),
-              title = "Pop. aged 5 to 14"
-  )+
-  tm_facets(by=c("Year"), ncol=2, nrow=1)+
-    tm_credits(c("City total: 58,265","City total: 58,865"),
+              labels = c("[0-40)", "[40-50)", "[50-75)",
+                         "[75-610 (2011 max.))","[610-835 (2016 max.))","[835-1345 (2021 max.)]"),
+              breaks = c(0,40,50,75, 610, 835, 1345),
+              title = "Pop. aged 5 to 14")+
+  tm_facets(by=c("Year"), nrow=1)+
+    tm_credits(c("City total: 58,265","City total: 58,865","City total: 62,195"),
              position = c("LEFT", "BOTTOM")) +
+  tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0) +
+  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0)+
+  tm_shape(ham_city_bound)+  tm_borders(col = "grey80", lwd=0.5)+
   tmap_defaults +
-  tm_layout(bg.color = "grey85",
+  tm_layout(bg.color = "grey99",
             panel.label.bg.color = "white",
@@
 student_pop_rate_plot <-
-  tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0) +
-  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0)+
   tm_shape(SAvail_i_DA_extra) +
   tm_polygons(col = "kid_pop_DAavg", border.alpha=0, 
-              palette = "Blues",
-              labels = c("[0-0.0018)", "[0.0018-0.0025)", "[0.0025-0.0038)","[0.0038-0.59]"),
-              breaks = c(0, 0.0018, 0.0025, 0.0038, 0.60),
+              palette = "Purples",
+              #style = "jenks",
+              labels = c("[0-0.0018)", "[0.0018-0.0025)", "[0.0025-0.0038)",
+                         "[0.0038-0.66]"),
+               breaks = c(0, 0.0018, 0.0025, 0.0038,0.66),
               title = "Avg. pop. aged 5 to 14\nper residential unit"
   )+
-  tm_facets(by=c("Year"), ncol=2, nrow=1)+
-
+  tm_facets(by=c("Year"), nrow=1)+
+  tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0) +
+  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0)+
+  tm_shape(ham_city_bound)+  tm_borders(col = "grey80", lwd=0.5)+
   tmap_defaults +
-  tm_layout(bg.color = "grey85",
+  tm_layout(bg.color = "grey99",
             panel.show=FALSE)
@@
 LIMAT_under18_DA_plot <-
-  tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0) +
-  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0)+
-  tm_shape(SAvail_i_DA_extra) +
+  tm_shape(SAvail_i_DA_extra |>
+             mutate(DA_prev_LIMAT_under18 = 
+                      ifelse(is.na(DA_prev_LIMAT_under18), 0, DA_prev_LIMAT_under18))) +
   tm_polygons(col = "DA_prev_LIMAT_under18", border.alpha=0, 
               palette = "Reds",
-              labels = c("[0-4.3)", "[4.3-13.0)", "[13.0-30.8)","[30.8-100)"),
-              breaks = c(0, 4.30, 13.00, 30.8, 150),
+              labels = c("[0-10.0)", "[10-18)", "[18-33)","[33-150)"),
+              breaks = c(0, 10, 18, 33, 150),
               colorNA = "black",
-              textNA = "Unavailable",
-              title = "Prop. of LIM-AT households\n with dependents < 18"
-  )+
-  tm_facets(by=c("Year"), ncol=2, nrow=1)+
+              title = "% LIM-AT households\n with dependents < 18")+
+  tm_facets(by=c("Year"), nrow=1)+
+  tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0) +
+  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0)+
+  tm_shape(ham_city_bound)+  tm_borders(col = "grey80", lwd=0.5)+
   tmap_defaults +
-  tm_layout(bg.color = "grey85",
+  tm_layout(bg.color = "grey99",
             panel.show=FALSE)
@@
 
-stud_pop_rate_LIM_plot
-```
-
-```{r student-popabs-lowincperc-poprate-plot2, eval=FALSE}
-tmap_save(stud_pop_rate_LIM_plot, paste0(here::here(),"/Figures/Fig2.tiff"), height=10, width=12, dpi=500)
+tmap_save(stud_pop_rate_LIM_plot, paste0(here::here(),"/Figures/Fig3.tiff"), height=7, width=10, dpi=600)
 ```
@@
 #reading boundaries of TAZs and PDs from the TTS2016R package (same TAZ for both 2011 and 2016)
+#remotes::install_github("soukhova/TTS2016R")
 library(TTS2016R) # A R package of the 2016 TTS
@@
 car_OD_flow_plot <-  
-    tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0) +
-  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0)+
+  tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0) +
+  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0)+
   tm_shape(od_trips_perc_car |> mutate(mode = "motorized"), bbox=SAvail_i_DA_extra) +
@@
               palette = "Greys",
-              breaks = c(0, .25, .50, .75, 1),
+              breaks = c(0, 0.1, 0.2, 0.3, 0.4,.50,.6,0.7,0.7,0.9, 1),
               title = "Proportion of mode use",
               textNA = "Unavailable",
-              colorNA = "black")+
+              colorNA = "white")+
   #tm_shape(SAvail_i_DA_extra) + tm_borders(alpha = 0.4, col = "grey20", lwd=0.2)+
@@
            palette = c("#365C8DFF", "#277F8EFF", "#1FA187FF", "#4AC16DFF", "#9FDA3AFF", "#FDE725FF"),
-           title.col = "Hamilton communities", lwd = 1.5) + 
+           title.col = "Hamilton communities", lwd = 1) + 
   tm_shape(ham_taz_centroids) + tm_dots(alpha=0.5, col="darkblue", size=0.01) +
@@
   tmap_defaults +
-  tm_layout(bg.color = "grey85",
+  tm_layout(bg.color = "grey99",
             panel.label.bg.color = "white",
@@
 
-walk_OD_flow <- tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0) +
-  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0)+
+walk_OD_flow <- tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0) +
+  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0)+
   tm_shape(od_trips_perc_walk |> mutate(mode = "non-motorized"), bbox=SAvail_i_DA_extra) +
@@
               textNA = "Unavailable",
-              colorNA = "black",
+              colorNA = "white",
               legend.show=FALSE)+
@@
            palette = c("#365C8DFF", "#277F8EFF", "#1FA187FF", "#4AC16DFF", "#9FDA3AFF", "#FDE725FF"),
-           title.col = "Hamilton communities", lwd = 1.5, legend.col.show=FALSE) + 
+           title.col = "Hamilton communities", lwd = 1, legend.col.show=FALSE) + 
   tm_shape(ham_taz_centroids) + tm_dots(alpha=0.5, col="darkblue", size=0.01) +
@@
   tmap_defaults +
-  tm_layout(bg.color = "grey85",
+  tm_layout(bg.color = "grey99",
             panel.label.bg.color = "white",
             panel.label.height = c(0.8,0,0)) +
-  tm_add_legend(type = "line", 
-                col = c("grey40","lightblue"),
-                labels = c("TAZ boundary","Community boundary"))+
+  # tm_add_legend(type = "line", 
+  #               col = c("grey40","lightblue"),
+  #               labels = c("TAZ boundary","Community boundary"))+
   tm_add_legend(type = "symbol",
@@
 OD_flow_plot <- tmap_arrange(car_OD_flow_plot, walk_OD_flow, nrow = 2)
-OD_flow_plot
-
-tmap_save(OD_flow_plot, paste0(here::here(),"/Figures/Fig3.tiff"), height=7, width=12, dpi=600)
+
+tmap_save(OD_flow_plot, paste0(here::here(),"/Figures/Fig4.tiff"), height=4.5, width=8, dpi=600)
 ```
@@
 TLD_empirical_wtheo_curve_plot
-ggsave(paste0(here::here(),"/Figures/Fig4.tiff"), height=7, width=10, dpi=600)
-```
+ggsave(paste0(here::here(),"/Figures/Fig5.tiff"), height=4.5, width=6, dpi=600)
+```
+
+```{r, eval=FALSE}
+#create 4 prob density plots to demonstrate the change in per parcel spatial availability per year per mode-using group
+df_Vij <- data.frame(
+    mode = c(avail_2011$mode, avail_2016$mode, avail_2021$mode),
+   Year= c(rep("2011", each=nrow(avail_2011)), rep("2016", each=nrow(avail_2016)), rep("2021", each=nrow(avail_2021))),
+   Vij = c(avail_2011$V_ij, avail_2016$V_ij, avail_2021$V_ij),
+   Measure = "Vij") |> 
+  mutate(mode = ifelse(mode == "c", "Motorized", "Non-motorized")) |>
+  mutate(Vij = ifelse(mode == "Motorized" & Vij > 0.01, NA, Vij)) |>
+  mutate(Vij = ifelse(mode == "Non-motorized" & Vij > 0.003, NA, Vij))  |> 
+           filter(!is.na(Vij))
+
+df_vij <- data.frame(
+    mode = c(avail_2011$mode, avail_2016$mode, avail_2021$mode),
+   Year= c(rep("2011", each=nrow(avail_2011)), rep("2016", each=nrow(avail_2016)), rep("2021", each=nrow(avail_2021))),
+   vij = c(avail_2011$V_ij/avail_2011$kid_popadj,
+           avail_2016$V_ij/avail_2016$kid_popadj,
+          avail_2021$V_ij/avail_2021$kid_popadj),
+   Measure = "vij") |> filter(!is.na(vij))|>
+  mutate(mode = ifelse(mode == "c", "Motorized", "Non-motorized"))
+
+density_theme <-  theme(legend.position = c(0.9,1),
+        legend.title = element_blank(),
+        axis.title = element_blank(),
+        legend.key.size = unit(0.25, "cm"),
+        legend.text = element_text(size=6))
+# Basic density plot for parcel-level 
+Vij_dens_plot <-
+  ggplot(df_Vij, aes(x=Vij, color=NULL, fill=Year)) +
+  #scale_x_log10() +
+  geom_density(alpha=0.4, linewidth=0.25) +
+  scale_fill_manual(values=viridis_pal()(3)) +
+  facet_wrap(~mode, nrow=2, scale="free") +
+  theme_minimal() +
+  density_theme
+
+vij_dens_plot <-
+  ggplot(df_vij, aes(x=vij, color=NULL, fill=Year)) +
+  #scale_x_log10() +
+  geom_density(alpha=0.4, adjust=3, linewidth=0.25) +
+  scale_fill_manual(values=viridis_pal()(3)) +
+  xlim(c(0,3))+
+  facet_wrap(~mode, nrow=2) +
+  theme_minimal()+
+  density_theme
+
+#repeat the above, but for the DA level scores
+Vij_dens_DA_plot <-
+  ggplot(SAvail_im_DA |> 
+           mutate(V_im = ifelse(mode == "Non-motorized" & V_im > 1.5, NA, V_im)) |>
+           mutate(V_im = ifelse(mode == "Motorized" & V_im > 250, NA, V_im))  |> 
+           filter(!is.na(V_im)),
+         aes(x=V_im, color=NULL, fill=year)) +
+  geom_density(alpha=0.4, linewidth=0.25) +
+  scale_fill_manual(values=viridis_pal()(3)) +
+  facet_wrap(~mode, nrow=2, scales = "free") +
+  theme_minimal() +
+  density_theme
+
+vij_dens_DA_plot <-
+  ggplot(SAvail_im_DA, aes(x=v_im_kid_popadj, color=NULL, fill=year)) +
+  geom_density(alpha=0.4, linewidth=0.25) +
+  scale_fill_manual(values=viridis_pal()(3)) +
+  xlim(c(0,3))+
+  facet_wrap(~mode, nrow=2) +
+  theme_minimal()+
+  density_theme
+
+Vijvi_dens_parcel_DA_plot <- plot_grid(
+  Vij_dens_plot, Vij_dens_DA_plot,
+  vij_dens_plot, vij_dens_DA_plot,
+  labels = c('Parcel Vi', 'DA Vi',
+              "Parcel vi", "DA vi"),
+  nrow=2, ncol=2, label_size = 9)
+
+ggsave(Vijvi_dens_parcel_DA_plot, file=paste0(here::here(),"/Figures/Fig6.tiff"),
+                                        height=4.5, width=6, dpi=600)
+```
+
 ```{r}
@@
 ```{r, eval=FALSE}
-Vim_20112016_plot <-  
-  tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0) +
-  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0)+
-  tm_shape(SAvail_im_DA, bbox=SAvail_im_DA) +
+Vim_201120162021_plot <-  
+  tm_shape(SAvail_im_DA |> filter(V_im >0), bbox=SAvail_im_DA) +
   tm_polygons(col = "V_im", border.alpha = 0, border.col = "grey20", lwd=0.2,
               palette = "Purples",
-              style="quantile", n=4,
-              labels = c("[0-0.2)", "[0.2-12.8)", "[12.8-56.4)", "[56.4-672.9)"),
+              alpha=0.7,
+              style="quantile", n=10,
+              #labels = c("[>0-0.2)", "[0.2-15.2)", "[15.2-54.8)", "[54.8-1200)"),
               title = expression(V[i]^m),
               textNULL = "Unavailable",
-              colorNULL = "black")+
+              colorNULL = "white")+
   tm_facets(by = c("mode","year"), nrow=1)+
   tm_shape(ham_community_bounds %>% st_cast("MULTILINESTRING")) + 
-  tm_lines(col = "COMMUNITY_", 
-           palette = c("#365C8DFF", "#277F8EFF", "#1FA187FF", "#4AC16DFF", "#9FDA3AFF", "#FDE725FF"),
-           title.col = "Hamilton communities", lwd = 2, legend.col.show = FALSE) + 
+  tm_lines(col = "COMMUNITY_URB", 
+  palette = c("grey0","grey20","grey40", "grey60",
+                       "grey80","grey100"),
+           title.col = "Hamilton communities", lwd = 1.5) + 
   tm_shape(ALL_SCHOOLS_Elem_formapping2) +
-  tm_symbols(shape = "Status",
-             size=0.1,
-             alpha=0.5,
+   tm_symbols(shape = "combo_status",
+             size= "OTGC",
+             alpha=0.6,
+             scale = 0.4,
+             shapes = c(1,
+                        17,24,
+                        15,22,
+                        18, 23),
           title.shape = "School status",
-          shapes = c(24,25,22,21,23),
+          title.size = "OTGC",
           shapes.labels =  c(
-          "Expanded after 2011",
-          "Moved before 2016",
-          "New after 2011",
-          "No change btwn 2011-16",
-          "Closed before 2016"),
-          legend.shape.show = FALSE
+          "No change (btwn 2011-21)", 
+          "Expanded (btwn 2011-16)", 
+          "Relocated + expanded (btwn 2011-16)",
+          "Closed (btwn 2011-16)", 
+          "Closed (btwn 2016-21)",
+          "Opened (btwn 2011-16)", 
+          "Opened (btwn 2016-21)")
   ) +
-  tm_facets(by=c("Mode","Year"))+
+  tm_facets(by=c("Mode","combo_year"))+
   tm_credits(c("60,220 spatial availability \n(99.4% of the 60,567 city-wide OTGC)",
                "58,101 spatial availability\n(99.7% of the 58,287 city-wide OTGC)",
+               "57,509 spatial availability\n(99.7% of the 58,697 city-wide OTGC)",
+
                 "347 spatial availability \n(0.6% of the 60,567 city-wide OTGC)",
-               "186 spatial availability\n(0.3% of the 58,287 city-wide OTGC)"),
+               "186 spatial availability\n(0.3% of the 58,287 city-wide OTGC)",
+               "187 spatial availability\n(0.3% of the 58,697 city-wide OTGC)"),
              position = c("LEFT", "BOTTOM")) +
+  tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0) +
+  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0)+
   tmap_defaults +
-  tm_layout(bg.color = "grey85",
+  tm_layout(bg.color = "grey99",
             panel.label.bg.color = "white",
@@
 
-vvim_20112016_plot <-  
-  tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0) +
-  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0)+
-  tm_shape(SAvail_im_DA, bbox=SAvail_im_DA) +
+vvim_201120162021_plot <-  
+  tm_shape(SAvail_im_DA  |> filter(v_im_kid_popadj >0), bbox=SAvail_im_DA) +
   tm_polygons(col = "v_im_kid_popadj", border.alpha = 0, border.col = "grey20", lwd=0.2,
@@
               title = expression(v[i]^m),
-              # style="quantile", n=4,
-              breaks = c(0, 0.62, 0.9, 1.22, 2.1), #note the highest value is ~10.29, I'm breaking the values early for the purpose of visualisation
-              labels = c("[0-0.62)", "[0.62-0.9)", "[0.9-1.22)", "[1.22-10.26)"),
+              style="quantile", n=10,
+              # breaks = c(0, 0.61, 0.88, 1.18, 2.2), #note the highest value is ~10.29, I'm breaking the values early for the purpose of visualisation
+              # labels = c("[0-0.6)", "[0.6-0.9)", "[0.9-1.2)", "[1.2-10.3)"),
               textNULL = "Unavailable",
-              colorNULL = "black")+
+              colorNULL = "white")+
   tm_facets(by = c("mode","year"), nrow=1)+
   tm_shape(ham_community_bounds %>% st_cast("MULTILINESTRING")) + 
-  tm_lines(col = "COMMUNITY_", 
-           palette = c("#365C8DFF", "#277F8EFF", "#1FA187FF", "#4AC16DFF", "#9FDA3AFF", "#FDE725FF"),
-           title.col = "Hamilton communities", lwd = 2) + 
+  tm_lines(col = "COMMUNITY_URB", 
+           palette = c("grey0","grey20","grey40", "grey60",
+                       "grey80","grey100"),
+           title.col = "Hamilton communities", lwd = 1.5, legend.col.show = FALSE) + 
 tm_shape(ALL_SCHOOLS_Elem_formapping2) +
-  tm_symbols(shape = "Status",
-             size=0.1,
-             alpha=0.5,
+  tm_symbols(shape = "combo_status",
+             size= "OTGC",
+             alpha=0.6,
+             scale = 0.4,
+             shapes = c(1,
+                        17,24,
+                        15,22,
+                        18, 23),
           title.shape = "School status",
-          shapes = c(24,25,21,23,22),
+          title.size = "OTGC",
           shapes.labels =  c(
-          "Expanded after 2011",
-          "Moved before 2016",
-          "New after 2011",
-          "No change btwn 2011-16",
-          "Closed before 2016")
-  ) +
-  tm_facets(by=c("Mode","Year"))+
+          "No change (btwn 2011-21)", 
+          "Expanded (btwn 2011-16)", 
+          "Relocated + expanded (btwn 2011-16)",
+          "Closed (btwn 2011-16)", 
+          "Closed (btwn 2016-21)",
+          "Opened (btwn 2011-16)", 
+          "Opened (btwn 2016-21)"),
+          legend.size.show = FALSE,
+          legend.shape.show = FALSE) +
+  tm_facets(by=c("Mode","combo_year"))+
   tm_credits(c("1.09 avg. spatial availability per capita",
                "1.05 avg. spatial availability per capita",
+               "0.99 avg. spatial availability per capita",
                 "0.85 avg. spatial availability per capita",
-               "0.71 avg. spatial availability per capita"),
+               "0.71 avg. spatial availability per capita",
+               "0.68 avg. spatial availability per capita"),
           position = c("LEFT", "BOTTOM")) +
+  tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0) +
+  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0)+
   tmap_defaults +
-  tm_layout(bg.color = "grey85",
+  tm_layout(bg.color = "grey99",
             panel.label.bg.color = "white",
@@
 
-Vim_vvim_20112016_plot <- tmap_arrange(Vim_20112016_plot, vvim_20112016_plot, nrow=2)
-
-tmap_save(Vim_vvim_20112016_plot, paste0(here::here(),"/Figures/Fig5.tiff"), height=12, width=12, dpi=450)
+Vim_vvim_201120162021_plot <- tmap_arrange(Vim_201120162021_plot, vvim_201120162021_plot, nrow=2)
+tmap_save(Vim_vvim_201120162021_plot, paste0(here::here(),"/Figures/Fig7.tiff"), height=12, width=10, dpi=300)
 ```
@@
 ```{r, eval=FALSE}
-tmap_defaults <-
-  tm_scale_bar(position = c("LEFT", "BOTTOM"), breaks = c(0, 5, 10, 15))+
-  tm_compass(size = 1.5, position = c("right", 'top')) +
-  tm_layout(legend.position = c("left", "top")) #legend.hist.size = 10,
-
-Vim_20112016_CHANGE_plot <-  
-  tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0) +
-  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0)+
-  tm_shape(SAvail_im_DA_changes, bbox=SAvail_im_DA) +
-  tm_polygons(col = "c_V_im", border.alpha = 0, border.col = "grey20", lwd=0.2,
-              palette = "RdBu", #midpoint=0,
-              # style="quantile", n=4,
-              breaks = c(-20, -4.1, -0.2, 0, 20),
-              labels = c("[-208 to -4.1)", "[-4.1 to -0.2)", "[-0.2 to 0)","[0 to 283.6]"),
-              title = expression("Change in "*V[i]^m),
-              textNA = "No V in 2016 or 2011",
+#comparison objects for mapping -- first only motorized
+comparision_mapping_motorized <- left_join(
+  SAvail_im_DA_changes_2011_16 |> st_drop_geometry()|> filter(mode == "motorized") |> select(c("GeoUID_2021", "c_V_im", "c_v_im_kid_popadj")) |> rename("c_V_im_2011_16"="c_V_im", "c_v_im_kid_popadj_2011_16"="c_v_im_kid_popadj"),
+  SAvail_im_DA_changes_2016_21 |> st_drop_geometry() |> filter(mode == "motorized") |> select(c("GeoUID_2021", "c_V_im", "c_v_im_kid_popadj")) |> rename("c_V_im_2016_21"="c_V_im", "c_v_im_kid_popadj_2016_21"="c_v_im_kid_popadj"),
+  by="GeoUID_2021")
+
+comparision_mapping_motorized <- left_join(
+  comparision_mapping_motorized, 
+  SAvail_im_DA_changes_2011_21 |> filter(!st_is_empty(geometry)) |> filter(mode == "motorized") |> select(c("GeoUID_2021", "c_V_im", "c_v_im_kid_popadj")) |> 
+    rename("c_V_im_2011_21"="c_V_im", "c_v_im_kid_popadj_2011_21"="c_v_im_kid_popadj"),
+  by="GeoUID_2021")
+
+comparision_mapping_motorized1 <- data.frame(
+  GeoUID_2021 = comparision_mapping_motorized$GeoUID_2021,
+  value_type = c(rep("1) c_Vi", each=length(comparision_mapping_motorized$c_V_im_2011_16)),
+                 rep("2) c_vi", each=length(comparision_mapping_motorized$c_v_im_kid_popadj_2011_16)),
+                 rep("1) c_Vi", each=length(comparision_mapping_motorized$c_V_im_2016_21)),
+                 rep("2) c_vi", each=length(comparision_mapping_motorized$c_v_im_kid_popadj_2016_21)),
+                 rep("1) c_Vi", each=length(comparision_mapping_motorized$c_V_im_2011_21)),
+                 rep("2) c_vi", each=length(comparision_mapping_motorized$c_v_im_kid_popadj_2011_21))),
+  year = c(rep("1) 2011-2016", each=length(comparision_mapping_motorized$c_V_im_2011_16)),
+                 rep("1) 2011-2016", each=length(comparision_mapping_motorized$c_v_im_kid_popadj_2011_16)),
+                 rep("2) 2016-2021", each=length(comparision_mapping_motorized$c_V_im_2016_21)),
+                 rep("2) 2016-2021", each=length(comparision_mapping_motorized$c_v_im_kid_popadj_2016_21)),
+                 rep("3) 2011-2021", each=length(comparision_mapping_motorized$c_V_im_2011_21)),
+                 rep("3) 2011-2021", each=length(comparision_mapping_motorized$c_v_im_kid_popadj_2011_21))),
+  value = c(comparision_mapping_motorized$c_V_im_2011_16, comparision_mapping_motorized$c_v_im_kid_popadj_2011_16,
+            comparision_mapping_motorized$c_V_im_2016_21, comparision_mapping_motorized$c_v_im_kid_popadj_2016_21,
+            comparision_mapping_motorized$c_V_im_2011_21, comparision_mapping_motorized$c_v_im_kid_popadj_2011_21)
+  )
+
+comparision_mapping_motorized_Vi <- left_join(comparision_mapping_motorized1, comparision_mapping_motorized |> select("GeoUID_2021", "geometry"), by = "GeoUID_2021") |> st_as_sf() |> filter(value_type == "1) c_Vi") |> mutate(mode="Motorized")
+comparision_mapping_motorized_vvi <- left_join(comparision_mapping_motorized1, comparision_mapping_motorized |> select("GeoUID_2021", "geometry"), by = "GeoUID_2021") |> st_as_sf() |> filter(value_type == "2) c_vi")|> mutate(mode="Motorized")
+
+#now, non-motorized:
+comparision_mapping_nonmotorized <- left_join(
+  SAvail_im_DA_changes_2011_16 |> st_drop_geometry()|> filter(mode == "non-motorized") |> select(c("GeoUID_2021", "c_V_im", "c_v_im_kid_popadj")) |> rename("c_V_im_2011_16"="c_V_im", "c_v_im_kid_popadj_2011_16"="c_v_im_kid_popadj"),
+  SAvail_im_DA_changes_2016_21 |> st_drop_geometry() |> filter(mode == "non-motorized") |> select(c("GeoUID_2021", "c_V_im", "c_v_im_kid_popadj")) |> rename("c_V_im_2016_21"="c_V_im", "c_v_im_kid_popadj_2016_21"="c_v_im_kid_popadj"),
+  by="GeoUID_2021")
+
+comparision_mapping_nonmotorized <- left_join(
+  comparision_mapping_nonmotorized, 
+  SAvail_im_DA_changes_2011_21 |> filter(!st_is_empty(geometry)) |> filter(mode == "non-motorized") |> select(c("GeoUID_2021", "c_V_im", "c_v_im_kid_popadj")) |> 
+    rename("c_V_im_2011_21"="c_V_im", "c_v_im_kid_popadj_2011_21"="c_v_im_kid_popadj"),
+  by="GeoUID_2021")
+
+comparision_mapping_nonmotorized1 <- data.frame(
+  GeoUID_2021 = comparision_mapping_nonmotorized$GeoUID_2021,
+  value_type = c(rep("1) c_Vi", each=length(comparision_mapping_nonmotorized$c_V_im_2011_16)),
+                 rep("2) c_vi", each=length(comparision_mapping_nonmotorized$c_v_im_kid_popadj_2011_16)),
+                 rep("1) c_Vi", each=length(comparision_mapping_nonmotorized$c_V_im_2016_21)),
+                 rep("2) c_vi", each=length(comparision_mapping_nonmotorized$c_v_im_kid_popadj_2016_21)),
+                 rep("1) c_Vi", each=length(comparision_mapping_nonmotorized$c_V_im_2011_21)),
+                 rep("2) c_vi", each=length(comparision_mapping_nonmotorized$c_v_im_kid_popadj_2011_21))),
+  year = c(rep("1) 2011-2016", each=length(comparision_mapping_nonmotorized$c_V_im_2011_16)),
+                 rep("1) 2011-2016", each=length(comparision_mapping_nonmotorized$c_v_im_kid_popadj_2011_16)),
+                 rep("2) 2016-2021", each=length(comparision_mapping_nonmotorized$c_V_im_2016_21)),
+                 rep("2) 2016-2021", each=length(comparision_mapping_nonmotorized$c_v_im_kid_popadj_2016_21)),
+                 rep("3) 2011-2021", each=length(comparision_mapping_nonmotorized$c_V_im_2011_21)),
+                 rep("3) 2011-2021", each=length(comparision_mapping_nonmotorized$c_v_im_kid_popadj_2011_21))),
+  value = c(comparision_mapping_nonmotorized$c_V_im_2011_16, comparision_mapping_nonmotorized$c_v_im_kid_popadj_2011_16,
+            comparision_mapping_nonmotorized$c_V_im_2016_21, comparision_mapping_nonmotorized$c_v_im_kid_popadj_2016_21,
+            comparision_mapping_nonmotorized$c_V_im_2011_21, comparision_mapping_nonmotorized$c_v_im_kid_popadj_2011_21)
+  )
+
+comparision_mapping_nonmotorized_Vi <- left_join(comparision_mapping_nonmotorized1, comparision_mapping_nonmotorized |> select("GeoUID_2021", "geometry"), by = "GeoUID_2021") |> st_as_sf() |> filter(value_type == "1) c_Vi") |> mutate(mode="Non-motorized")
+comparision_mapping_nonmotorized_vvi <- left_join(comparision_mapping_nonmotorized1, comparision_mapping_nonmotorized |> select("GeoUID_2021", "geometry"), by = "GeoUID_2021") |> st_as_sf() |> filter(value_type == "2) c_vi")|> mutate(mode="Non-motorized")
+```
+
+```{r}
+ALL_SCHOOLS_Elem_formapping_21 <- SCHOOLS |>
+  filter(Year == "2011 and 2016" & Year_2 == "2016 and 2021" & Status == "Expanded" & Status_2 == "NoChange") |>
+  mutate(combo_status = "2) Expanded between 2011-16",
+         combo_change_year = "2011-2016")
+ALL_SCHOOLS_Elem_formapping_22 <- SCHOOLS |>
+  filter(Year == "2011 and 2016" & Year_2 == "2016 and 2021" & Status == "Expanded" & Status_2 == "NoChange") |>
+  mutate(combo_status = "2) Expanded between 2011-16",
+         combo_change_year = "2011-2021")
+
+ALL_SCHOOLS_Elem_formapping_31 <- SCHOOLS |>
+  filter(Year == "2016" & Year_2 == "2016 and 2021" & Status == "Moved" & Status_2 == "NoChange") |>
+  mutate(combo_status = "3) Relocated + expanded between 2011-16",
+         combo_change_year = "2011-2016")
+ALL_SCHOOLS_Elem_formapping_32 <- SCHOOLS |>
+  filter(Year == "2016" & Year_2 == "2016 and 2021" & Status == "Moved" & Status_2 == "NoChange") |>
+  mutate(combo_status = "3) Relocated + expanded between 2011-16",
+         combo_change_year = "2011-2021")
+
+ALL_SCHOOLS_Elem_formapping_51 <- SCHOOLS |>
+  filter(Year == "2016" & Year_2 == "2016 and 2021" & Status == "New" & Status_2 == "NoChange") |>
+  mutate(combo_status = "6) Opened between 2011-16",
+         combo_change_year = "2011-2016")
+ALL_SCHOOLS_Elem_formapping_52 <- SCHOOLS |>
+  filter(Year == "2016" & Year_2 == "2016 and 2021" & Status == "New" & Status_2 == "NoChange") |>
+  mutate(combo_status = "6) Opened between 2011-16",
+         combo_change_year = "2011-2021")
+
+ALL_SCHOOLS_Elem_formapping_81 <- SCHOOLS |>
+  filter(is.na(Year) & Year_2 == "2021" & is.na(Status) & Status_2 == "New") |>
+  mutate(combo_status = "7) Opened between 2016-21",
+         combo_change_year = "2016-2021") 
+ALL_SCHOOLS_Elem_formapping_82 <- SCHOOLS |>
+  filter(is.na(Year) & Year_2 == "2021" & is.na(Status) & Status_2 == "New") |>
+  mutate(combo_status = "7) Opened between 2016-21",
+         combo_change_year = "2011-2021")
+
+ALL_SCHOOLS_Elem_formapping_71 <- SCHOOLS |>
+  filter(Year == "2011" & is.na(Year_2) & Status == "Removed" & is.na(Status_2)) |>
+  mutate(combo_status = "4) Closed between 2011-16",
+         combo_change_year = "2011-2016") 
+ALL_SCHOOLS_Elem_formapping_72 <- SCHOOLS |>
+  filter(Year == "2011" & is.na(Year_2) & Status == "Removed" & is.na(Status_2)) |>
+  mutate(combo_status = "4) Closed between 2011-16",
+         combo_change_year = "2011-2021")
+
+ALL_SCHOOLS_Elem_formapping_61 <- SCHOOLS |>
+  filter(Year == "2011 and 2016" & Year_2 == "2016" & Status == "NoChange" & Status_2 == "Removed") |>
+  mutate(combo_status = "5) Closed between 2016-21",
+         combo_change_year = "2016-2021") 
+ALL_SCHOOLS_Elem_formapping_62 <- SCHOOLS |>
+  filter(Year == "2011 and 2016" & Year_2 == "2016" & Status == "NoChange" & Status_2 == "Removed") |>
+  mutate(combo_status = "5) Closed between 2016-21",
+         combo_change_year = "2011-2021")
+
+ALL_SCHOOLS_Elem_formapping3 <- rbind(
+  ALL_SCHOOLS_Elem_formapping_21,ALL_SCHOOLS_Elem_formapping_22,
+  ALL_SCHOOLS_Elem_formapping_31,ALL_SCHOOLS_Elem_formapping_32,
+  ALL_SCHOOLS_Elem_formapping_51,ALL_SCHOOLS_Elem_formapping_52,
+  ALL_SCHOOLS_Elem_formapping_61,ALL_SCHOOLS_Elem_formapping_62,
+  ALL_SCHOOLS_Elem_formapping_71,ALL_SCHOOLS_Elem_formapping_72,
+  ALL_SCHOOLS_Elem_formapping_81, ALL_SCHOOLS_Elem_formapping_82
+  )
+
+rm(ALL_SCHOOLS_Elem_formapping_21,ALL_SCHOOLS_Elem_formapping_22,
+  ALL_SCHOOLS_Elem_formapping_31,ALL_SCHOOLS_Elem_formapping_32,
+  ALL_SCHOOLS_Elem_formapping_51,ALL_SCHOOLS_Elem_formapping_52,
+  ALL_SCHOOLS_Elem_formapping_61,ALL_SCHOOLS_Elem_formapping_62,
+  ALL_SCHOOLS_Elem_formapping_71,ALL_SCHOOLS_Elem_formapping_72,
+  ALL_SCHOOLS_Elem_formapping_81, ALL_SCHOOLS_Elem_formapping_82)
+
+ALL_SCHOOLS_Elem_formapping3 <- ALL_SCHOOLS_Elem_formapping3 |>
+  mutate(combo_change_year = case_when(combo_change_year== "2011-2016" ~ "1) 2011-2016",
+                                       combo_change_year== "2016-2021" ~ "2) 2016-2021",
+                                       combo_change_year== "2011-2021" ~ "3) 2011-2021"))
+```
+
+```{r, eval=FALSE}
+#New figure 7 showing the changes -- first motor only
+motor_Vi_CHANGE_plot <-  
+  tm_shape(comparision_mapping_motorized_Vi |> filter(!st_is_empty(geometry)),
+           bbox=SAvail_im_DA_changes_2011_21) +
+  tm_polygons(col = "value", border.alpha = 0, border.col = "grey20", lwd=0.2, alpha=0.6,
+              palette = "RdBu", midpoint=0,
+              style="quantile", n=10,
+              #breaks = c(-20, -4, -0.2, 0, 20),
+              midpoint = 0,
+              #labels = c("[-208 to -4)", "[-4 to -0.2)", "[-0.2 to 0)","[0 to 331]"),
+              title = expression("Change in "*V[i]^{Motorized}),
               colorNA = "white")+
-  tm_facets(by = c("mode"), nrow=1)+
+  tm_facets(by = c("year"), nrow=1)+
   tm_shape(ham_community_bounds %>% st_cast("MULTILINESTRING")) + 
-  tm_lines(col = "COMMUNITY_", 
-           palette = c("#365C8DFF", "#277F8EFF", "#1FA187FF", "#4AC16DFF", "#9FDA3AFF", "#FDE725FF"),
-           title.col = "Hamilton communities", lwd = 2, 
-           legend.col.show = FALSE) + 
-  tm_shape(ALL_SCHOOLS_Elem_formapping |> 
-             filter(Status != "NoChange" & Status != "Moved")|>
-             mutate(Status_change = case_when(Status == "Removed" ~ "Closed",
-                                              (Status == "Expanded" | Status == "New" ) ~ "New or Expanded"))) +
-  tm_symbols(shape = "Status_change",
-             size=0.1,
-             alpha=0.5,
-          title.shape = "Schools that changed status",
-          shapes = c(22,21),
-          # shapes.labels = c(
-          # "New or expanded after 2011",
-          # "Closed before 2016"),
-          legend.shape.show = FALSE) +
+  tm_lines(col = "COMMUNITY_URB", 
+           palette = c("grey0","grey20","grey40", "grey60","grey80","grey100"),
+           title.col = "Hamilton communities", lwd = 1.5
+           #legend.col.show = FALSE
+           ) + 
+  tm_shape(ALL_SCHOOLS_Elem_formapping3)+
+  tm_symbols(shape = "combo_status",
+             alpha=0.6,
+             scale = 0.3,
+             shapes = c(17,24,
+                        15,22,
+                        18, 23),
+          title.shape = "Schools that changed",
+          shapes.labels =  c(
+          "Expanded (btwn 2011-16)",
+          "Relocated + expanded (btwn 2011-16)",
+          "Closed (btwn 2011-16)",
+          "Closed (btwn 2016-21)",
+          "Opened (btwn 2011-16)",
+          "Opened (btwn 2016-21)")
+          #legend.size.show = FALSE,
+          #legend.shape.show = FALSE
+          ) +
+  tm_facets(by = c("combo_change_year"), nrow=1)+
+  tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0) + 
+  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0)+
   tmap_defaults +
-  tm_layout(bg.color = "grey85",
+  tm_layout( bg.color = "grey99",
              panel.label.bg.color = "white",
-            panel.label.height = 0.8)
-
-vvim_20112016_CHANGE_plot <-
-  tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0) + 
-  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0)+
-  tm_shape(SAvail_im_DA_changes, bbox=SAvail_im_DA) +
-  tm_polygons(col = "c_v_im_kid_popadj", border.alpha = 0, border.col = "grey20", lwd=0.2,
+             panel.label.height = 0.8,
+             legend.outside = TRUE,
+             panel.labels = c("2011-2016", "2016-2021", "2011-2021"))
+
+motor_vvi_CHANGE_plot <-  
+  tm_shape(comparision_mapping_motorized_vvi |> filter(!st_is_empty(geometry)),
+           bbox=SAvail_im_DA_changes_2011_21) +
+  tm_polygons(col = "value", border.alpha = 0, border.col = "grey20", lwd=0.2, alpha=0.6,
               palette = "RdBu", midpoint=0,
-              breaks = c(-0.1, -0.08, -0.05, 0, 0.1),
-              labels = c("[-3.77 to -0.08)", "[-0.08 to -0.05)", "[-0.05 to 0)", 
-                         "[0 to 1.46]"),
-              title = expression("Change in "*v[i]^m),
-              textNA = "No V in 2016 or 2011",
+              style="quantile", n=10,
+              #breaks = c(-20, -4, -0.2, 0, 20),
+              midpoint = 0,
+              #labels = c("[-208 to -4)", "[-4 to -0.2)", "[-0.2 to 0)","[0 to 331]"),
+              title = expression("Change in "*v[i]^{Motorized}),
               colorNA = "white")+
-  tm_facets(by = c("mode"), nrow=1)+
+  tm_facets(by = c("year"), nrow=1)+
   tm_shape(ham_community_bounds %>% st_cast("MULTILINESTRING")) + 
-  tm_lines(col = "COMMUNITY_", 
-           palette = c("#365C8DFF", "#277F8EFF", "#1FA187FF", "#4AC16DFF", "#9FDA3AFF", "#FDE725FF"),
-           title.col = "Hamilton communities", lwd = 2) + 
-  tm_shape(ALL_SCHOOLS_Elem_formapping |> 
-             filter(Status != "NoChange" & Status != "Moved")|>
-             mutate(Status_change = case_when(Status == "Removed" ~ "Closed",
-                                              (Status == "Expanded" | Status == "New" ) ~ "New or Expanded"))) +
-  tm_symbols(shape = "Status_change",
-             size=0.1,
-             alpha=0.5,
-          title.shape = "Schools that changed status",
-          shapes = c(22,21),
+  tm_lines(col = "COMMUNITY_URB", 
+           palette = c("grey0","grey20","grey40", "grey60","grey80","grey100"),
+           title.col = "Hamilton communities", lwd = 1.5,
+           legend.col.show = FALSE
+           ) + 
+  tm_shape(ALL_SCHOOLS_Elem_formapping3)+
+  tm_symbols(shape = "combo_status",
+             alpha=0.6,
+             scale = 0.3,
+             shapes = c(17,24,
+                        15,22,
+                        18, 23),
+          title.shape = "Schools that changed",
           shapes.labels =  c(
-          "Closed before 2016",
-          "New or expanded after 2011")
+          "Expanded (btwn 2011-16)",
+          "Relocated + expanded (btwn 2011-16)",
+          "Closed (btwn 2011-16)",
+          "Closed (btwn 2016-21)",
+          "Opened (btwn 2011-16)",
+          "Opened (btwn 2016-21)"),
+          legend.shape.show = FALSE
+          ) +
+  tm_facets(by = c("combo_change_year"), nrow=1)+
+  tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0) + 
+  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0)+
+  tmap_defaults +
+  tm_layout( bg.color = "grey99",
+             panel.label.bg.color = "white",
+             panel.label.height = 0.8,
+             legend.outside = TRUE,
+             panel.show = FALSE)
+
+#now repeat, but non-motorized - changes in 2011-2016, 2016 to 2021 and 2011 through 2021
+nonmotor_Vi_CHANGE_plot <-  
+  tm_shape(comparision_mapping_nonmotorized_Vi |> filter(!st_is_empty(geometry)),
+           bbox=SAvail_im_DA_changes_2011_21) +
+  tm_polygons(col = "value", border.alpha = 0, border.col = "grey20", lwd=0.2, alpha=0.6,
+              palette = "RdBu", midpoint=0,
+              style="quantile", n=10,
+              #breaks = c(-20, -4, -0.2, 0, 20),
+              midpoint = 0,
+              #labels = c("[-208 to -4)", "[-4 to -0.2)", "[-0.2 to 0)","[0 to 331]"),
+              title = expression("Change in "*V[i]^{Non-motorized}),
+              colorNA = "white")+
+  tm_facets(by = c("year"), nrow=1)+
+  tm_shape(ham_community_bounds %>% st_cast("MULTILINESTRING")) + 
+  tm_lines(col = "COMMUNITY_URB", 
+           palette = c("grey0","grey20","grey40", "grey60","grey80","grey100"),
+           title.col = "Hamilton communities", lwd = 1.5,
+           legend.col.show = FALSE
+           ) + 
+  tm_shape(ALL_SCHOOLS_Elem_formapping3)+
+  tm_symbols(shape = "combo_status",
+             alpha=0.6,
+             scale = 0.3,
+             shapes = c(17,24,
+                        15,22,
+                        18, 23),
+          title.shape = "Schools that changed",
+          shapes.labels =  c(
+          "Expanded (btwn 2011-16)",
+          "Relocated + expanded (btwn 2011-16)",
+          "Closed (btwn 2011-16)",
+          "Closed (btwn 2016-21)",
+          "Opened (btwn 2011-16)",
+          "Opened (btwn 2016-21)"),
+          legend.shape.show = FALSE
+          ) +
+  tm_facets(by = c("combo_change_year"), nrow=1)+
+  tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0) + 
+  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0)+
+  tmap_defaults +
+  tm_layout( bg.color = "grey99",
+             panel.label.bg.color = "white",
+             panel.label.height = 0.8,
+             legend.outside = TRUE,
+             panel.show = FALSE)
+
+nonmotor_vvi_CHANGE_plot <-  
+  tm_shape(comparision_mapping_nonmotorized_vvi |> filter(!st_is_empty(geometry)),
+           bbox=SAvail_im_DA_changes_2011_21) +
+  tm_polygons(col = "value", border.alpha = 0, border.col = "grey20", lwd=0.2, alpha=0.6,
+              palette = "RdBu", midpoint=0,
+              style="quantile", n=10,
+              #breaks = c(-20, -4, -0.2, 0, 20),
+              midpoint = 0,
+              #labels = c("[-208 to -4)", "[-4 to -0.2)", "[-0.2 to 0)","[0 to 331]"),
+              title = expression("Change in "*v[i]^{Non-motorized}),
+              colorNA = "white")+
+  tm_facets(by = c("year"), nrow=1)+
+  tm_shape(ham_community_bounds %>% st_cast("MULTILINESTRING")) + 
+  tm_lines(col = "COMMUNITY_URB", 
+           palette = c("grey0","grey20","grey40", "grey60","grey80","grey100"),
+           title.col = "Hamilton communities", lwd = 1.5,
+           legend.col.show = FALSE
+           ) + 
+  tm_shape(ALL_SCHOOLS_Elem_formapping3)+
+  tm_symbols(shape = "combo_status",
+             alpha=0.6,
+             scale = 0.3,
+             shapes = c(17,24,
+                        15,22,
+                        18, 23),
+          title.shape = "Schools that changed",
+          shapes.labels =  c(
+          "Expanded (btwn 2011-16)",
+          "Relocated + expanded (btwn 2011-16)",
+          "Closed (btwn 2011-16)",
+          "Closed (btwn 2016-21)",
+          "Opened (btwn 2011-16)",
+          "Opened (btwn 2016-21)"),
+          legend.shape.show = FALSE
+          ) +
+  tm_facets(by = c("combo_change_year"), nrow=1)+
+  tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0) + 
+  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0)+
+  tmap_defaults +
+  tm_layout(bg.color = "grey99",
+             panel.label.bg.color = "white",
+             panel.label.height = 0.8,
+             legend.outside = TRUE,
+             panel.show = FALSE)
+
+Vim_vvim_201120162021_CHANGE_plot <- tmap_arrange(motor_Vi_CHANGE_plot,
+                                                  nonmotor_Vi_CHANGE_plot,
+                                                  motor_vvi_CHANGE_plot,
+                                                  nonmotor_vvi_CHANGE_plot, nrow=4)
+
+tmap_save(Vim_vvim_201120162021_CHANGE_plot, paste0(here::here(),"/Figures/Fig8.tiff"), height=12, width=10, dpi=300)
+```
+
+```{r, eval=FALSE}
+motor_ttim_20112021_CHANGE_plot <-  
+  tm_shape(SAvail_im_DA_changes_2011_21 |> 
+             filter(!st_is_empty(geometry)) |>
+             filter(mode == "motorized"), bbox=SAvail_im_DA) +
+  tm_polygons(col = "c_total_ttperkid", border.alpha = 0, border.col = "grey20", lwd=0.2, alpha=0.8,
+              palette = "-RdYlGn", midpoint=0, 
+              style="quantile", n=10,
+              title = c("Motorized travel time minutes per pop-aged 5 to 14"))+
+  tm_shape(ham_community_bounds %>% st_cast("MULTILINESTRING")) +
+  tm_lines(col = "COMMUNITY_URB",
+           palette = c("grey0","grey20","grey40", "grey60","grey80","grey100"),
+           title.col = "Hamilton communities", lwd = 1.5,
+           legend.col.show = FALSE
+           ) +
+  tm_shape(ALL_SCHOOLS_Elem_formapping3 |> filter(combo_change_year == "3) 2011-2021"))+
+  tm_symbols(shape = "combo_status",
+             alpha=0.6,
+             scale = 0.3,
+             shapes = c(17,24,
+                        15,22,
+                        18, 23),
+          title.shape = "Schools that changed",
+          shapes.labels =  c(
+          "Expanded (btwn 2011-16)",
+          "Relocated + expanded (btwn 2011-16)",
+          "Closed (btwn 2011-16)",
+          "Closed (btwn 2016-21)",
+          "Opened (btwn 2011-16)",
+          "Opened (btwn 2016-21)")
+          #legend.shape.show = FALSE
           ) +
   tmap_defaults +
-  tm_layout(bg.color = "grey85",
-             panel.label.bg.color = "white",
-            panel.label.height = 0.8)
-
-Vim_vvim_20112016_CHANGE_plot <- tmap_arrange(Vim_20112016_CHANGE_plot, vvim_20112016_CHANGE_plot, nrow=2)
-```
-
-```{r, eval=FALSE}
-tmap_save(Vim_vvim_20112016_CHANGE_plot, 
-          paste0(here::here(),"/Figures/Fig6.tiff"),
-          height=7,
-          width=12, 
-          dpi=600)
-```
-
-```{r, eval=FALSE}
-tmap_defaults <-
-  tm_scale_bar(position = c("LEFT", "BOTTOM"), breaks = c(0, 5, 10, 15))+
-  tm_compass(size = 1.5, position = c("right", 'top')) +
-  tm_layout(legend.position = c("left", "top")) #legend.hist.size = 10,
-
-ttim_20112016_CHANGE_plot <-  
-  tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0) +
-  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0)+
-  tm_shape(SAvail_im_DA_changes |> filter(mode == "non-motorized"), 
-           bbox=SAvail_im_DA ) +
-  tm_polygons(col = "c_total_tt", border.alpha = 0, border.col = "grey20", lwd=0.2,
-              palette = "RdYlGn", midpoint=0,
-              #style="quantile", n=4, 
-              breaks = c(-10, -5.67, -2.94, 0, 10),
-              labels = c("[-69.0 to -5.7)", "[-5.7 to -2.94)", "[-2.94 to 0)","[0 to 20.81]"),
-              title = c("Non-motorized travel time (minutes)"))+
-  tm_shape(ham_community_bounds %>% st_cast("MULTILINESTRING")) + 
-  tm_lines(col = "COMMUNITY_", 
-           palette = c("#365C8DFF", "#277F8EFF", "#1FA187FF", "#4AC16DFF", "#9FDA3AFF", "#FDE725FF"),
-           title.col = "Hamilton communities", lwd = 2, 
-           legend.col.show = FALSE) + 
-  tm_shape(ALL_SCHOOLS_Elem_formapping |> 
-             filter(Status != "NoChange" & Status != "Moved")|>
-             mutate(Status_change = case_when(Status == "Removed" ~ "Closed",
-                                              (Status == "Expanded" | Status == "New" ) ~ "New or Expanded"))) +
-  tm_symbols(shape = "Status_change",
-             size=0.1,
-             alpha=0.5,
-          title.shape = "Schools that changed status",
-          shapes = c(22,21),
-          # shapes.labels = c(
-          # "New or expanded after 2011",
-          # "Closed before 2016"),
-          legend.shape.show = FALSE) +
-  tmap_defaults +
-  tm_layout( bg.color = "grey85",
+  tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0) + 
+  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0)+
+  tm_layout(bg.color = "grey99",
              panel.label.bg.color = "white",
              panel.label.height = 0.9,
-             panel.labels = "Change in non-motorized minutes between 2011-16",
+             panel.labels = "Motorized minutes",
              legend.outside = TRUE)
 
-GHGim_20112016_CHANGE_plot <-
-  tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0) +
-  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0)+
-  tm_shape(SAvail_im_DA_changes |> filter(mode == "motorized"), bbox=SAvail_im_DA) +
-  tm_polygons(col = "c_emissions_total_car_tt", border.alpha = 0, border.col = "grey20", lwd=0.2,
-              palette = "-RdYlGn", midpoint=0,
-              #style="quantile", n=4,
-              breaks = c(-200, -15, -4, 10, 200),
-              labels = c("[-197 to -15)", "[-15 to -4)", "[-4 to 0)", "[0 to 10)", "[10 to 1,220]"),
-              title = expression("kg CO"[2]*"e for potential home-school trips"),
-              textNA = "No V in 2016 or 2011",
-              colorNA = "white")+
+nonmotor_ttim_20112021_CHANGE_plot <-  
+  tm_shape(SAvail_im_DA_changes_2011_21 |> 
+             filter(!st_is_empty(geometry)) |>
+             filter(mode == "non-motorized"), bbox=SAvail_im_DA) +
+  tm_polygons(col = "c_total_ttperkid", border.alpha = 0, border.col = "grey20", lwd=0.2, alpha=0.8,
+              palette = "RdYlGn", midpoint=0,
+              style="quantile", n=10,
+              title = c("Non-motorized travel time minutes per pop-aged 5 to 14"))+
   tm_shape(ham_community_bounds %>% st_cast("MULTILINESTRING")) +
-  tm_lines(col = "COMMUNITY_", 
-           palette = c("#365C8DFF", "#277F8EFF", "#1FA187FF", "#4AC16DFF", "#9FDA3AFF", "#FDE725FF"),
-           title.col = "Hamilton communities", lwd = 2) + 
-  tm_shape(ALL_SCHOOLS_Elem_formapping |> 
-             filter(Status != "NoChange" & Status != "Moved")|>
-             mutate(Status_change = case_when(Status == "Removed" ~ "Closed",
-                                              (Status == "Expanded" | Status == "New" ) ~ "New or Expanded"))) +
-  tm_symbols(shape = "Status_change",
-             size=0.1,
-             alpha=0.5,
-          title.shape = "Schools that changed status",
-          shapes = c(22,21),
+  tm_lines(col = "COMMUNITY_URB",
+           palette = c("grey0","grey20","grey40", "grey60","grey80","grey100"),
+           title.col = "Hamilton communities", lwd = 1.5,
+           legend.col.show = FALSE
+           ) +
+  tm_shape(ALL_SCHOOLS_Elem_formapping3 |> filter(combo_change_year == "3) 2011-2021"))+
+  tm_symbols(shape = "combo_status",
+             alpha=0.6,
+             scale = 0.3,
+             shapes = c(17,24,
+                        15,22,
+                        18, 23),
+          title.shape = "Schools that changed",
           shapes.labels =  c(
-          "Closed before 2016",
-          "New or expanded after 2011")
+          "Expanded (btwn 2011-16)",
+          "Relocated + expanded (btwn 2011-16)",
+          "Closed (btwn 2011-16)",
+          "Closed (btwn 2016-21)",
+          "Opened (btwn 2011-16)",
+          "Opened (btwn 2016-21)"),
+          legend.shape.show = FALSE
           ) +
   tmap_defaults +
-  tm_layout( bg.color = "grey85",
+  tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0) + 
+  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0)+
+  tm_layout( bg.color = "grey99",
              panel.label.bg.color = "white",
              panel.label.height = 0.9,
-             panel.labels = "Change in GHG emissions between 2011-16",
+             panel.labels = "Non-motorized minutes",
              legend.outside = TRUE)
 
-ttim_GHGim_20112016_CHANGE_plot <- tmap_arrange(ttim_20112016_CHANGE_plot, GHGim_20112016_CHANGE_plot, nrow=2)
-```
-
-```{r, eval=FALSE}
-tmap_save(ttim_GHGim_20112016_CHANGE_plot, 
-          paste0(here::here(),"/Figures/Fig7.tiff"),
+ttim_20112021_CHANGE_plot <- tmap_arrange(motor_ttim_20112021_CHANGE_plot,
+                                                  nonmotor_ttim_20112021_CHANGE_plot, nrow=2)
+
+tmap_save(ttim_20112021_CHANGE_plot, 
+          paste0(here::here(),"/Figures/Fig9.tiff"),
           height=5.5,
@@
 
+```{r include=FALSE}
+#calc changes between the years - but between communities with geom for mapping
+
+#LIMAT: here we assume 2016 and 2011 LIMAT are only comparable... 2021 reflects one-time covid payments that temporary brought people out of poverty.
+t2011 <-avail_2011 |> group_by(GeoUID_2021) |>
+  summarise(COMMUNITY = first(COMMUNITY),
+            LIMAT_2011 =  mean(DA_prev_LIMAT_under18, na.rm=T),
+            kid_popadj_2011 = sum(kid_popadj, na.rm=T))
+
+t2016 <-avail_2016 |> group_by(GeoUID_2021) |>
+  summarise(COMMUNITY = first(COMMUNITY),
+            LIMAT_2016 =  mean(DA_prev_LIMAT_under18, na.rm=T),
+            kid_popadj_2016 = sum(kid_popadj, na.rm=T))
+
+t2021 <-avail_2021 |> group_by(GeoUID) |>
+  summarise(COMMUNITY = first(COMMUNITY),
+            LIMAT_2021 =  mean(DA_prev_LIMAT_under18, na.rm=T),
+            kid_popadj_2021 = sum(kid_popadj, na.rm=T))
+
+ct201621_2011 <- full_join(t2011, t2016, 
+                        by=c("GeoUID_2021"="GeoUID_2021", "COMMUNITY")) |>
+  full_join(t2021, 
+            by=c("GeoUID_2021"="GeoUID", "COMMUNITY")) |>
+  group_by(COMMUNITY) |> 
+  summarise(LIMAT_2011 = mean(LIMAT_2011, na.rm=T),
+            LIMAT_2016 = mean(LIMAT_2016, na.rm=T),
+            kid_popadj_2011 = sum(kid_popadj_2011, na.rm=T),
+            kid_popadj_2021 = sum(kid_popadj_2021, na.rm=T)) |>
+  mutate(cLIMAT = ((LIMAT_2016-LIMAT_2011)/LIMAT_2011) , #|> scales::percent(accuracy = 2L)
+         ckid_popadj  = ((kid_popadj_2021-kid_popadj_2011 )/kid_popadj_2011) )
+  #mutate_if(is.numeric, round, 1)
+
+#OTGC
+cs2021_2011 <- SCHOOLS |> st_drop_geometry() |> group_by(COMMUNITY) |> 
+  summarise(OTGC2011 = sum(OTGC2011, na.rm=T), 
+            OTGC2021 = sum(OTGC2021, na.rm=T)) |>
+  mutate(cOTGC = ((OTGC2021-OTGC2011)/OTGC2011) ) 
+#mutate_if(is.numeric, round, 1)
+
+#pop,
+# motorized: V, vi, tt per kid 
+# non-motorized: V, vi, tt per kid 
+SAvail_im_COMM2011 <- avail_2011 %>%
+  group_by(COMMUNITY, 
+           mode) %>%
+  summarize(total_tt_2011 = sum(travel_time_p50*kid_popadj),
+            V_im_2011 = sum(V_ij),
+            kid_popadj_m_2011 = sum(kid_popadj),
+            .groups = "drop") |>
+  mutate(total_tt_kid_2011 = total_tt_2011/kid_popadj_m_2011,
+         v_im_kid_popadj_2011 = V_im_2011/kid_popadj_m_2011)
+
+SAvail_im_COMM2021 <- avail_2021 %>%
+  group_by(COMMUNITY, 
+           mode) %>%
+  summarize(total_tt_2021 = sum(travel_time_p50*kid_popadj),
+            V_im_2021 = sum(V_ij),
+            kid_popadj_m_2021 = sum(kid_popadj),
+            .groups = "drop") |>
+  mutate(total_tt_kid_2021 = total_tt_2021/kid_popadj_m_2021,
+         v_im_kid_popadj_2021 = V_im_2021/kid_popadj_m_2021)
+
+cSAvail_im_COMM2011_2021 <- left_join(SAvail_im_COMM2011, SAvail_im_COMM2021, by =c("COMMUNITY","mode"))|>
+  mutate(c_total_tt = ((total_tt_2021-total_tt_2011)/total_tt_2011) ,
+         c_V_im = ((V_im_2021-V_im_2011)/V_im_2011) ,
+         c_kid_popadj_m = ((kid_popadj_m_2021-kid_popadj_m_2011)/kid_popadj_m_2011) ,
+         c_total_tt_kid = ((total_tt_kid_2021-total_tt_kid_2011)/total_tt_kid_2011) ,
+         cv_im_kid_popadj = ((v_im_kid_popadj_2021-v_im_kid_popadj_2011)/v_im_kid_popadj_2011) )
+
+ct201621_2011 #LIMAT and pop
+cs2021_2011 #OTGC
+cSAvail_im_COMM2011_2021
+```
+
+```{r include=FALSE}
+# ADD 2016 and 2021 data for each variable to this!! cSAvail_im_COMM2011_2021_df
+cSAvail_im_COMM2011_2021_df <- data.frame(
+  COMMUNITY = cSAvail_im_COMM2011_2021$COMMUNITY,
+  mode = cSAvail_im_COMM2011_2021$mode,
+  LIMAT_2011 = rep(ct201621_2011$LIMAT_2011, each=2),
+  cLIMAT = rep(ct201621_2011$cLIMAT, each=2),
+  kid_popadj_2011 = rep(ct201621_2011$kid_popadj_2011, each=2),
+  kid_popadj_2021 = rep(ct201621_2011$kid_popadj_2021, each=2),
+  ckid_popadj = rep(ct201621_2011$ckid_popadj, each=2),
+  OTGC2011 = rep(cs2021_2011$OTGC2011, each=2),
+  cOTGC = rep(cs2021_2011$cOTGC, each=2),
+
+  V_im_2011 = cSAvail_im_COMM2011_2021$V_im_2011,
+  V_im_2021 = cSAvail_im_COMM2011_2021$V_im_2021,
+  c_V_im_perc = cSAvail_im_COMM2011_2021$c_V_im,
+  v_im_kid_popadj_2011 = cSAvail_im_COMM2011_2021$v_im_kid_popadj_2011,
+  c_v_im_perc = cSAvail_im_COMM2011_2021$cv_im_kid_popadj,
+
+  total_tt_2011 = cSAvail_im_COMM2011_2021$total_tt_2011,
+  c_total_tt = cSAvail_im_COMM2011_2021$c_total_tt,
+  total_tt_kid_2011 = cSAvail_im_COMM2011_2021$total_tt_kid_2011,
+  c_total_tt_kid = cSAvail_im_COMM2011_2021$c_total_tt_kid )|>
+  arrange(desc(LIMAT_2011))
+```
+
+```{r include=FALSE}
+tdata_1 <- cSAvail_im_COMM2011_2021_df |> 
+  filter(mode == "c")|> 
+  dplyr::select(c("COMMUNITY",
+                  "LIMAT_2011",
+                  "kid_popadj_2011",
+                  "OTGC2011"))
+
+tdata_1 <- insertRow(tdata_1 |> as.matrix(),2,
+          cSAvail_im_COMM2011_2021_df |> 
+            filter(mode == "c" & COMMUNITY == "HA") |> 
+            dplyr::select(c("COMMUNITY", 
+                            "cLIMAT", 
+                            "ckid_popadj",
+                            "cOTGC")) |> as.matrix())
+
+tdata_1 <- insertRow(tdata_1,4,
+          cSAvail_im_COMM2011_2021_df |> 
+            filter(mode == "c" & COMMUNITY == "DU") |> 
+            dplyr::select(c("COMMUNITY", 
+                            "cLIMAT", 
+                            "ckid_popadj",
+                            "cOTGC")) |> as.matrix())
+
+tdata_1 <- insertRow(tdata_1,6,
+          cSAvail_im_COMM2011_2021_df |> 
+            filter(mode == "c" & COMMUNITY == "SC") |> 
+            dplyr::select(c("COMMUNITY", 
+                            "cLIMAT", 
+                            "ckid_popadj",
+                            "cOTGC")) |> as.matrix())
+
+tdata_1 <- insertRow(tdata_1,8,
+          cSAvail_im_COMM2011_2021_df |> 
+            filter(mode == "c" & COMMUNITY == "GL") |> 
+            dplyr::select(c("COMMUNITY", 
+                            "cLIMAT", 
+                            "ckid_popadj",
+                            "cOTGC")) |> as.matrix())
+
+tdata_1 <- insertRow(tdata_1,10,
+          cSAvail_im_COMM2011_2021_df |> 
+            filter(mode == "c" & COMMUNITY == "FL") |> 
+           dplyr::select(c("COMMUNITY", 
+                            "cLIMAT", 
+                            "ckid_popadj",
+                            "cOTGC")) |> as.matrix())
+
+tdata_1 <- insertRow(tdata_1,12,
+          cSAvail_im_COMM2011_2021_df |> 
+            filter(mode == "c" & COMMUNITY == "AN") |> 
+            dplyr::select(c("COMMUNITY", 
+                            "cLIMAT", 
+                            "ckid_popadj",
+                            "cOTGC")) |> as.matrix())
+```
+
+```{r include=FALSE}
+tdata_2 <- cSAvail_im_COMM2011_2021_df |> 
+  filter(mode == "c")|> 
+  dplyr::select(c("COMMUNITY",
+                  "V_im_2011",
+                  "v_im_kid_popadj_2011",
+                  "total_tt_kid_2011"))
+
+tdata_2 <- insertRow(tdata_2 |> as.matrix(),2,
+          cSAvail_im_COMM2011_2021_df |> 
+            filter(mode == "c" & COMMUNITY == "HA") |> 
+            dplyr::select(c("COMMUNITY", 
+                            "c_V_im_perc", 
+                            "c_v_im_perc",
+                            "c_total_tt_kid")) |> as.matrix())
+
+tdata_2 <- insertRow(tdata_2,4,
+          cSAvail_im_COMM2011_2021_df |> 
+            filter(mode == "c" & COMMUNITY == "DU") |> 
+            dplyr::select(c("COMMUNITY", 
+                            "c_V_im_perc", 
+                            "c_v_im_perc",
+                            "c_total_tt_kid")) |> as.matrix())
+
+tdata_2 <- insertRow(tdata_2,6,
+          cSAvail_im_COMM2011_2021_df |> 
+            filter(mode == "c" & COMMUNITY == "SC") |> 
+            dplyr::select(c("COMMUNITY", 
+                            "c_V_im_perc", 
+                            "c_v_im_perc",
+                            "c_total_tt_kid")) |> as.matrix())
+
+tdata_2 <- insertRow(tdata_2,8,
+          cSAvail_im_COMM2011_2021_df |> 
+            filter(mode == "c" & COMMUNITY == "GL") |> 
+            dplyr::select(c("COMMUNITY", 
+                            "c_V_im_perc", 
+                            "c_v_im_perc",
+                            "c_total_tt_kid")) |> as.matrix())
+
+tdata_2 <- insertRow(tdata_2,10,
+          cSAvail_im_COMM2011_2021_df |> 
+            filter(mode == "c" & COMMUNITY == "FL") |> 
+            dplyr::select(c("COMMUNITY", 
+                            "c_V_im_perc", 
+                            "c_v_im_perc",
+                            "c_total_tt_kid")) |> as.matrix())
+
+tdata_2 <- insertRow(tdata_2,12,
+          cSAvail_im_COMM2011_2021_df |> 
+            filter(mode == "c" & COMMUNITY == "AN") |> 
+            dplyr::select(c("COMMUNITY", 
+                            "c_V_im_perc", 
+                            "c_v_im_perc",
+                            "c_total_tt_kid")) |> as.matrix())
+
+tdata_3 <- cSAvail_im_COMM2011_2021_df |> 
+  filter(mode == "w")|> 
+  dplyr::select(c("COMMUNITY",
+                  "V_im_2011",
+                  "v_im_kid_popadj_2011",
+                  "total_tt_kid_2011"))
+
+tdata_3 <- insertRow(tdata_3 |> as.matrix(),2,
+          cSAvail_im_COMM2011_2021_df |> 
+            filter(mode == "w" & COMMUNITY == "HA") |> 
+            dplyr::select(c("COMMUNITY", 
+                            "c_V_im_perc", 
+                            "c_v_im_perc",
+                            "c_total_tt_kid")) |> as.matrix())
+
+tdata_3 <- insertRow(tdata_3,4,
+          cSAvail_im_COMM2011_2021_df |> 
+            filter(mode == "w" & COMMUNITY == "DU") |> 
+            dplyr::select(c("COMMUNITY", 
+                            "c_V_im_perc", 
+                            "c_v_im_perc",
+                            "c_total_tt_kid")) |> as.matrix())
+
+tdata_3 <- insertRow(tdata_3,6,
+          cSAvail_im_COMM2011_2021_df |> 
+            filter(mode == "w" & COMMUNITY == "SC") |> 
+            dplyr::select(c("COMMUNITY", 
+                            "c_V_im_perc", 
+                            "c_v_im_perc",
+                            "c_total_tt_kid")) |> as.matrix())
+
+tdata_3 <- insertRow(tdata_3,8,
+          cSAvail_im_COMM2011_2021_df |> 
+            filter(mode == "w" & COMMUNITY == "GL") |> 
+            dplyr::select(c("COMMUNITY", 
+                            "c_V_im_perc", 
+                            "c_v_im_perc",
+                            "c_total_tt_kid")) |> as.matrix())
+
+tdata_3 <- insertRow(tdata_3,10,
+          cSAvail_im_COMM2011_2021_df |> 
+            filter(mode == "w" & COMMUNITY == "FL") |> 
+            dplyr::select(c("COMMUNITY", 
+                            "c_V_im_perc", 
+                            "c_v_im_perc",
+                            "c_total_tt_kid")) |> as.matrix())
+
+tdata_3 <- insertRow(tdata_3,12,
+          cSAvail_im_COMM2011_2021_df |> 
+            filter(mode == "w" & COMMUNITY == "AN") |> 
+            dplyr::select(c("COMMUNITY", 
+                            "c_V_im_perc", 
+                            "c_v_im_perc",
+                            "c_total_tt_kid")) |> as.matrix())
+
+tdata_1 <- t(tdata_1) |> data.frame() 
+tdata_2 <- t(tdata_2) |> data.frame() 
+tdata_3 <- t(tdata_3) |> data.frame()
+```
+
+```{r}
+tdata <- rbind(tdata_1,tdata_2,tdata_3)
+tdata <- tdata[-c(1,5,9),]
+tdata <- tdata[-c(6,9),] #also removing tts.. not needed
+test <- data.frame(c("LIM-AT", "Kid pop.", "OTGC", "V (mt)", "v (mt)", "V (nmt)", "v (nmt)"))
+tdata <- tdata|>mutate_if(is.character,as.numeric)
+tdata <- cbind(test,tdata)
+colnames(tdata) <- c("type","HA" ,"HA.1", "DU", "DU.1", "SC", "SC.1", "GL","GL.1","FL", "FL.1","AN","AN.1")
+tdata <- tdata |> add_row(type = "Motorized") |> add_row(type = "Non-motorized")
+
+tdata <- tdata[c(1,2,3,8,4,5,9,6,7),]
+tdata <- tdata %>% replace(is.na(.), 0)
+
+#tdata <- tdata %>% mutate_if(is.numeric, round, 2) 
+```
+
 # Introduction
 
-Communities are by nature in flux: economic and demographic shifts, rural and urban configurations, and political pressures all combine to produce change.
-These forces spur the re-organization and consolidation of public facilities in communities around the world where changes to service provision may improve the quality of service for some but not for all [@Christiaanse_2020; @Rosik_Pulawska-Obiedowska_Goliszek_2021].
-While change is natural, oftentimes top-down consolidation decisions are driven by operational cost-savings that do not take into consideration the full extent of losses in benefits.
-
-From this perspective, school closure policies have been pursued by governments to optimize per-student operational cost-efficiency [@rongReviewResearchLowcarbon2022; @Dai_Wang_Zhang_Liao_Liu_2019] but other implications, such as the impact of closures on property values [@MerrallWhatsASchool2024], their effects on the social infrastructure of communities [@butlerClosureRideau2019a; @irwinSchoolClosure2012], and particularly their travel implications have often been overlooked by decision makers [@bierbaumMobilityJusticeLinking2021; @Lee_Lubienski_2017].
-This seems a mistake: journey-to-school trips are tremendously important, and travel for this purpose is currently the largest contributor of carbon emissions from personal travel after the journey-to-work [@rongReviewResearchLowcarbon2022; @pantelakiImpactHomeschoolCommuting2024], where it could be instead an excellent opportunity for physical activity [@desjardinsFramingActive2024].
-Alas, in many communities globally, schools are closing [@sageman2022].
+The amount of services accessible within communities oscillate: economic and demographic shifts, rural and urban configurations and political in/action all induce change.
+These forces spur pressures to reorganization and consolidation of public facilities that provide services, potentially improving access for some but not for all [@Christiaanse_2020; @Rosik_PulawskaObiedowska_Goliszek_2021].
+Schools are a compelling case study in this respect.
+In many communities globally, school closure and consolidation policies have been pursued by governments to optimize per-student operational cost-efficiency [@rongReviewResearchLowcarbon2022; @Dai_Wang_Zhang_Liao_Liu_2019].
+Research has highlighted that these polices have knock-on implications, including the weakening of communities' social infrastructure [@butlerClosureRideau2019a; @irwinSchoolClosure2012; @AuttiHyryBeihammer2014] and negative household-level fiscal impacts such as a reduction of property values [@MerrallWhatsASchool2024].
+Operationally, savings may be experienced but they are paid at different levels by students, families, communities and society.
+
+A school offers many things to a community: but one dimension is the travel to and from school, often overlooked by decision-makers considering school closure and consolidation policies [@bierbaumMobilityJusticeLinking2021; @Lee_Lubienski_2017].
+Schools and residential locations often proximately co-located, the journey-to-school is an excellent opportunity for physical activity [@desjardinsFramingActive2024].
+However, closures exchange the physical activity of students' realized active trips or "reserved" active trips [@morencyStepsReserve2009; @morencyHowMany2007] (e.g., trips that could be made active) for motorized travel [@rongReviewResearchLowcarbon2022].
+Indeed, motorized travel to school is the largest contributor of carbon emissions from personal daily travel after the journey-to-work [@rongReviewResearchLowcarbon2022; @pantelakiImpactHomeschoolCommuting2024].
+Despite these societal implications, many communities are closing and/or consolidating schools [@sageman2022].
 For instance, 65% of comprehensive schools (e.g.
-grades 1 through 9) in rural Finland since 1990 closed [@Autti_Hyry-Beihammer_2014], 10% of elementary schools (kindergarten through grade 8) in Chicago, USA in 2013 [@Lee_Lubienski_2017], and 7% of primary and lower secondary schools (kindergarten through grade 9) were closed in Denmark between 2010-2011 [@Beuchert_Humlum_Nielsen_Smith_2018].
-School closures often increase the distance to school for many students and thus tend to induce motorized trips [@rongReviewResearchLowcarbon2022].
-
-We focus on the transportation implications of school closure and consolidation policies.
-While some research has considered the "steps in reserve", that is, the potential for active travel hidden behind motorized trips [@morencyStepsReserve2009; @morencyHowMany2007], school closures potentially make actual steps vanish, sent to the reserve for the foreseeable future.
-In exchange for less physical activity, communities get more motorized travel, with the concomitant costs and burdens, including emissions.
-Quantification of losses in potential active travel and associated growth in carbon emissions is scarce in the literature, and the focus of this paper.
-In addition to its empirical contributions, this work makes use of a novel multimodal and singly-constrained accessibility measure, called spatial availability [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024], which we use to estimate the changes in accessibility and travel after a recent wave of school closures/consolidations in Hamilton, Ontario.
-
-The results of the research indicate that, overall, school seats became less spatially available in the city.
-This change happened unevenly, with some communities being disproportionately affected.
-Though the school consolidation policy pursued by the province and implemented by the local school board may have resulted in cost-savings in operation and maintenance of facilities, families, students, and communities now pay the price in terms of lost opportunities for healthier active travel, as well as increased emissions from motorized travel.
-
-# Background
-
-Hamilton is a mid-size city (\~570,000 pop) with rural, suburban and urban characteristics in the province of Ontario, Canada [@governmentofcanadaProfileTableCensus2022].
-In 2013, the Hamilton-Wentworth District School Board (HWDSB), the largest public English school board in the city, released its Long Term Facilities Master Plan [@hwdsbLongTermFacilities2013].
-To ensure "equitable, affordable and sustainable learning facilities", the Master Plan indicated that 80% of its elementary schools (attended by students aged 5 to 13) would be subjected to accommodation evaluations, including an "Accommodation Review", the outcome of which historically has been school closures/consolidations [@hwdsbLongTermFacilities2013; @craggsBoardEvaluatingFuture2013; @seasonsSchoolClosurePolicy2014].
-These decisions, underhandedly supported by the province, were in part motivated by a search for operational savings based on projected reductions in student population and underutilized school capacity [@craggsAxeCouldFall2012].
-Hamilton was particularly affected, as the HWDSB underwent an unprecedented number of Accommodation Reviews relative to other school boards [@seasons2014ARCCatalogues2014].
-
-Though the Accommodation Review guidelines outline a community-focused process to assess the value that a school provides to students, the community, and the local economy [@seasonsSchoolClosurePolicy2014; @ministryofeducationPupilAccommodationReview2006], in practice residents felt that a wholesome breakdown of costs to repair/maintain schools was not made consistently available [@kleinhuisSeriousConcernsSchool2013].
-Further, others pointed to flawed incentive structures motivating the HWDSB: "By closing three schools the HWDSB would have a strong case with the Ministry of Education to receive full funding for a new school" [@kleinhuisSeriousConcernsSchool2013].
-Along with insufficient funding to maintain schools in a state of good repair [@auditorgeneralofontarioAnnualReport20152015], the operational savings as a case for school closures is a long-standing critique of the school "funding formula" established by the conservative provincial government in 1998 [@mackenzieBlueprintFixOntario2018; @irwinSchoolClosure2012].
-Within a backdrop of budget cuts, the funding formula pits the financing of school maintenance and student needs against the financial compensation of staff and the powers of local school boards, which depend on the province for the allocation of resources [@mackenzieBlueprintFixOntario2018].
-
-In the wave of school accommodation reviews between 2013 and 2016, `r ALL_SCHOOLS_Elem %>% filter(Year == "2011" & Level == "Elementary" & Status == "Removed") %>% pull(SchoolID) %>% length()` elementary schools closed in Hamilton.
-These closures represented a `r ((ALL_SCHOOLS_Elem %>% filter(Year != "2016") %>% pull(SchoolID) %>% length() - ALL_SCHOOLS_Elem %>% filter(Year != "2011") %>% pull(SchoolID) %>% length())/ALL_SCHOOLS_Elem %>% filter(Year != "2016" & Level == "Elementary") %>% pull(SchoolID) %>% length()) %>% scales::percent()` and `r  (( ((avail_2011 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2011)) |> select(OTGC) |> sum())) - (avail_2016 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2016)) |> select(OTGC) |> sum())) / (avail_2016 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2016)) |> select(OTGC) |> sum()) ) |> scales::percent()` decline in elementary school locations and capacity between 2011 (before the school closures) and 2016, despite the elementary-aged student population in the city slightly increasing between 2011 and 2016( from `r avail_2011$kid_popadj %>% sum(na.rm=T) %>% prettyNum(big.mark=",")` students to `r avail_2016$kid_popadj %>% sum(na.rm=T) %>% prettyNum(big.mark=",")` students).
-
-While accommodation reviews were undertaken to reduce operational costs, how might have the closure of schools impacted local communities?
-In this paper, we hypothesize that the school consolidation policy reduced accessibility to schools for many students.
-By extension, this loss in access and the consequent displacement of active trips to school increased transportation-related emissions.
-To test these hypotheses, the paper examines the following: how does the (1) spatial availability, (2) loss in walk minutes and (3) associated carbon emissions change for home-to-school trips after a city closes and consolidates schools?
-Further, this work examines (4) what communities, based on low-income prevalence, experienced the greatest losses as a result of this policy.
-
-Spatial availability, a singly-constrained measure of spatial accessibility [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024], is used to quantify accessibility before- and after- school closures, and the associated impacts for the city of Hamilton and the elementary school-aged student population as a result of a policy with narrowly defined benefits.
+grades 1 through 9) in rural Finland since 1990 closed [@AuttiHyryBeihammer2014], 10% of elementary schools (kindergarten through grade 8) in Chicago, USA in 2013 [@Lee_Lubienski_2017], and 7% of primary and lower secondary schools (kindergarten through grade 9) were closed in Denmark between 2010-2011 [@Beuchert_Humlum_Nielsen_Smith_2018].
+
+In this paper, we explore the case study of Hamilton, a mid-sized city in Canada that has pursued similar school closure and consolidation policies.
+We detail the Canadian political context that incentived these policies and quantify their transportation implications using data that constructs where students may have lived and the capacity of schools in 2011, 2016 and 2021.
+To quantify transportation implications, we use a novel multimodal extension of a singly-constrained accessibility measure known as spatial availability [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024].
+Spatial availability is used to estimate active (i.e., walking) and motorized (i.e., car) accessibility considering the capacity of each school, and the residential location and quantity of school-aged population, and travel times from residential locations to schools for three time periods.
+In this way, this work contributes a empirical findings documenting how government austerity hurts public education, along with a methodology that results in interpretable accessibility values that can be compared across modes, the City, and three time periods.
+
+To foreshadow the paper's conclusions, our findings indicate that school seats have became less spatially available in the city.
+This change happened unevenly, with certain communities being disproportionately affected.
+The school closure and consolidation policy pursued may have achieved cost savings in facility operation and maintenance.
+However, it has come at a cost to families, students, and communities, resulting in lost opportunities for healthier active travel and a potential increase in motorized travel.
+We investigate: were the closures and consolidations worth the operational savings?
 The methods used in this paper can be applied by decision-makers to evaluate the spatial and transport-related impacts of public service provision policy.
 Furthermore, to increase the diffusion of this analysis, all work is transparent and reproducible, following best practices in the spatial sciences [@brunsdonOpeningPractice2021; @paezOpenSpatial2021c].
-Accordingly, all associated code and data will be publicly available within the lead author's [GitHub repository](https://github.com/soukhova/School-closures-accessibility-impacts)
+Accordingly, all associated code and data is publicly available within the lead author's [GitHub repository](https://github.com/soukhova/School-closures-accessibility-impacts).
+
+# Background
+
+## Overview of proximity and equity in the school accessibility literature
+
+The examination of access to public services is a well established area of the transportation literature: from parks [@elmurrWalkingAccessibilityParks2021] to healthcare [@pereiraGeographicAccessCOVID192021] and associated dimensions of inequities [@tiznadoaitkenPublicTransportAccessibility2021; @bittencourtEvaluatingAccessibilityAvailability2023; @kelobonye2020measuring], spatial accessibility has been explored in cases studies worldwide.
+
+Accessibility to schools has been a public service explored across several transportation dimensions.
+For instance, @marquesAccessibilityPrimarySchools2020 demonstrates that lower-income neighbourhoods, compounded by other intersecting dimensions of disadvantage, tend to have lower accessibility to schools compared to more advantaged neighbourhoods in Portugal.
+Similar conclusions are drawn in West Virginia, US by @Talen_2001, England by @burgess2011parental, Baton Rouge, US by @williamsDisparitiesAccessibilityPublic2014, São Paulo and Curitiba, Brazil by @morenomonroyPublicTransportSchool2018, @pizzolQualifyingAccessibilityEducation2021, and @bittencourtEvaluatingAccessibilityAvailability2023, and Santiago, Chile by @tiznadoaitkenPublicTransportAccessibility2021.
+Research has also demonstrated that when school closures occurs, what schools end up closing are not a politically neutral decision.
+For instance, these spatial trends are explored in the context of England by @pinchChangingGeographyPreschool1987, areas of declining school-aged populations in Dreseden, Germany by @mullerAssessmentSchoolClosures2011, rural Finnish communities by @AuttiHyryBeihammer2014, and Chicago, US by @Lee_Lubienski_2017.
+
+The methods applied to measure school accessibility have been varied, from simple indicators to complex metrics.
+Beginning from more simple, @marquesAccessibilityPrimarySchools2020 present indicators of school seats per student along with the shortest distance from origin to schools within a school catchment, methods also used in @Talen_2001 and @burgess2011parental.
+Authors make refinements to these methods; @mullerAssessmentSchoolClosures2011 considers travel time by public transit from origins to schools and @pizzolQualifyingAccessibilityEducation2021 considers travel times by different modes and multiple schools from a single origin along with other indicators related to school quality.
+By contrast, @williamsDisparitiesAccessibilityPublic2014, @Lee_Lubienski_2017, @morenomonroyPublicTransportSchool2018, @tiznadoaitkenPublicTransportAccessibility2021, and @bittencourtEvaluatingAccessibilityAvailability2023 use more complex approaches such as the two-step floating catchment area (2SFCA).
+The 2SFCA is popular in measure in evaluating access to healthcare services [@paez2019] and yields a supply-to-demand ratio that accounts for the population demanding a service and the quantity that is supplied within a certain distance range [@shenLocationCharacteristicsInnercity1998; @luoMeasuresSpatialAccessibility2003].
+The consideration of the supply of service is pertinent in the case of schools and is acknowledged to as a limitation to using simpler accessibility measures [@pizzolQualifyingAccessibilityEducation2021].
+
+In addition to competition for school-seats, theoretically, the advantage students using faster modes (e.g., motorized modes) offer in terms of a wider range in school-seat choice is also important.
+Though some of the reviewed works using 2SFCA approaches consider multiple modes, explicit attention to the modal _advantage_ has not been discussed.
+To this end, in this work we use spatial availability, a singly-constrained measure of spatial accessibility [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024].
+Spatial availability allows for the representation of how many 'school-seats' are spatially available to motorized- and active-trip taking elementary school-aged student population at a spatial-level, facilitating interpretable comparison across years.
+Further, spatial availability can be represented as school-seats per capita, mathematically equivalent to the supply-to-demand ratio of the 2SFCA (see appendix in @soukhovIntroducingSpatialAvailability2023).
+
+## Hamilton's school closures and consolidation
+
+Hamilton is a mid-size city (\~570,000 pop) with rural, suburban and urban characteristics in the province of Ontario, Canada [@governmentofcanadaProfileTableCensus2022].
+The city's public English education is covered by the Hamilton-Wentworth District School Board (HWDSB).
+In 2013, the HWDSB released its Long Term Facilities Master Plan [@hwdsbLongTermFacilities2013].
+To ensure "equitable, affordable and sustainable learning facilities", the Master Plan indicated that 80% of its elementary schools (attended by students aged approximately 5 to 14) would be subject to evaluations, including a "Pupil Accommodation Review" [@hwdsbLongTermFacilities2013].
+This Master Plan and the slated Pupil Accommodation Reviews prompted public outcry, as the outcome of Pupil Accommodation Reviews are typically schools closures and consolidations [@craggsBoardEvaluatingFuture2013; @seasonsSchoolClosurePolicy2014].
+
+Like all school boards in Ontario, HWDSB relies on funding from the province which is distributed through a formula based on pupil enrolment [@mackenzieBlueprintFixOntario2018; @irwinSchoolClosure2012].
+However, it has been demonstrated that this funding formula has become insufficient to maintain schools in a state of good repair [@auditorgeneralofontarioAnnualReport20152015].
+Further, existing "top-up" programs to assist in operation and maintenance costs (e.g., School Facilities Operation and Renewal Grant) were entirely phased out by 2018 [@tdsbLiftingMinistryEducations2024].
+Within this funding backdrop, the Pupil Accommodation Review guidelines outlined a "community-focused" process to assessing the value that a school provides to students, the community, and the local economy [@seasonsSchoolClosurePolicy2014; @ministryofeducationPupilAccommodationReview2006].
+However in practice, critical residents felt the outcomes of the Pupil Accommodation Reviews were predetermined [@thompsonItJustGalls2024]: older schools were slated to be closed, and a wholesome breakdown of costs to repair and maintain schools were not made available [@kleinhuisSeriousConcernsSchool2013].
+Other critics pointed to unbalanced incentives motivating the HWDSB: "By closing three schools, the HWDSB would have a strong case with the Ministry of Education to receive full funding for a new school" [@kleinhuisSeriousConcernsSchool2013].
+The Ministry's framework promotes the closing and consolidation of schools rather than cost-effective solutions for keeping older schools open [@kleinhuisSeriousConcernsSchool2013].
+
+The HWDSB and the Hamilton-Wentworth Catholic District School Board (HWCDSB), which together provide public English education to the majority of school-aged children in Hamilton [@faoOntarioSchoolBoards2023], were particularly impacted.
+While Accommodation Reviews were a province-wide trend during this period, Hamilton's school boards underwent an unprecedented number of reviews compared to other boards in other Ontario municipalities.
+[@seasons2014ARCCatalogues2014].
+In the wave of Pupil Accommodation Reviews between 2013 and 2016, `r SCHOOLS %>% filter(Status == "Removed") %>% pull(SchoolID) %>% length()` of the `r SCHOOLS %>% filter(Year == "2011 and 2016" | Year == "2011") %>% pull(SchoolID) %>% length()` elementary schools closed in Hamilton.
+Likely in part due to the public unpopularity of school closures, in 2017, the provincial government introduced a moratorium on beginning new Pupil Accommodation Reviews.
+In the following years, the HWDSB addressed its backlog and closed `r SCHOOLS %>% filter(Status_2 == "Removed") %>% pull(SchoolID) %>% length()` more schools between 2017 to 2021 [@hwdsb2023LongTermFacilities2023].
+Though elementary school locations declined by `r -((SCHOOLS %>% filter(Year_2 != "2016") %>% pull(SchoolID) %>% length() - SCHOOLS %>% filter(Year != "2016") %>% pull(SchoolID) %>% length())/SCHOOLS %>% filter(Year != "2016") %>% pull(SchoolID) %>% length()) %>% scales::percent()` between 2011-2021, HWDSB intends to conduct future Pupil Accommodation Reviews as initially scheduled, waiting until the end of the moratorium before dates are assigned [@hwdsb2023LongTermFacilities2023].
+
+A justification in these Pupil Accommodation Reviews are operational savings based on "projected" reductions in student population and "under-utilized" school capacity [@craggsAxeCouldFall2012; @tdsbLiftingMinistryEducations2024; @opsbaRightSchoolsRight2024].
+As later demonstrated in this study, these policies in fact did result in school capacity per student reducing overall.
+Indeed, school seats in the city declined `r  -(( ((avail_2021 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2021)) |> select(OTGC) |> sum())) - (avail_2011 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2011)) |> select(OTGC) |> sum())) / (avail_2011 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2011)) |> select(OTGC) |> sum()) ) |> scales::percent()`.
+It should be noted that though closures dominated the Pupil Accommodation Review process, some schools were "consolidated" e.g., a school or two were closed and another school was expanded or opened to accommodate students from closed schools.
+Taken that school locations decreased more than overall school-seats, elementary schools in Hamilton are on average larger than they once were.
+
+In 2024, the moratorium on Pupil Accommodation Reviews is still in place and school boards are motivated by the same funding conditions.
+In absence of new revenue-generating levers and under the same funding formula with even lower inflation adjusted per-pupil funding, many school boards are calling for the end of the moratorium [@opsbaRightSchoolsRight2024; @tdsbLiftingMinistryEducations2024; @draaismaTDSBConsideringCuts2024].
+In the meantime, some school boards are reducing specialized and community-wide programming [@tdsbLiftingMinistryEducations2024].
+Under the status-quo, the end of the moratorium will result in further school closures.
+These pertinent decisions, made under an austere funding formula and flawed school board incentives does not create an environment where solutions for a sustainable public education system can thrive.
+
+Though schools have a plethora of benefits not captured by their maintenance and operational costs [@butlerClosureRideau2019a; @irwinSchoolClosure2012; @AuttiHyryBeihammer2014], this paper's scope is limited to their travel implications.
+For students who live in proximity to schools, their closure or consolidation directly eliminate the realized or potential active trips that are, or could have been, made.
+Closing a school permanently reduces the potential for active trips and further entrenches reliance on motorized travel.
+Additionally, from the perspective of modal equity, school-aged populations with access to motorized transport gain a speed-time advantage, providing them with greater flexibility and choice in accessing school-seats.
+
+The spatial and population-level distribution of these impacts are of question for this paper.
+Specifically, the paper's aim is to demonstrate how these policies impacted the motorized and non-motorized (active) school-seat accessibility landscape.
+We hypothesize that the school closure and consolidation policies reduced accessibility to school-seats for many students.
+By extension, this loss in access and the displacement of active trips likely resulted in decreased potential active travel minutes and increased motorized travel minutes, with these impacts distributed inequitably across space and populations.
+To test these hypotheses, the paper examines the following:
+
+- How has spatial availability changed for home-to-school trips following school closures and consolidations?
+- What is the extent of the loss in walking minutes for these trips? 
+- How has motorized travel increased as a result? 
+- Which communities, particularly households with a high prevalence of low-income and hence more likely experiencing transport poverty, experienced the greatest losses due to this policy?
 
@@
 
-Our focus is to evaluate the impact of school consolidations between 2011 and 2016 on multimodal school spatial availability and travel-related GHG emission in Hamilton.
-To facilitate comparison, two scenarios are generated.
-Each scenario reflects the conditions in 2011 and 2016 respectively, assuming the school configurations in each year, as well as average student population and low-income prevalence of households, residential parcel locations, and estimated modal travel times.
-The difference between the two scenarios is quantified in terms of accessibility, emissions and equity implications and discussed in context with the school consolidation policy.
-Data was retrieved at different levels of spatial granularity from the following sources.
+Our focus is to evaluate the impact of school consolidations between 2011, 2016 and 2021 on multimodal school-seat spatial availability in Hamilton.
+To facilitate comparison, the following conditions are constructed for each year: 1) school configurations, 2) average student population, low-income prevalence of households, and residential parcel locations, and 3) estimated multimodal travel behaviour.
+The following sub-sections describe these three types of data.
 
@@
 
-First, schools and associated school catchments were retrieved from the City's school boards.
-In Hamilton, the majority of student-aged population attends a school in one of the two English public school boards [@faoOntarioSchoolBoards2023].
-One board is referred to as *public* (i.e., Hamilton Wentworth District School Board (HWDSB)) in our work and the other is referred to as *public-catholic* (Hamilton Wentworth Catholic District School Board (HWCDSB)).
-These catchments indicate what residential property is assigned to what school, by default.
-Families can decide to attend schools out of catchment; in fact, 21%-23% <!--calculated in 03-travel-times-TTS.qmd--> of motorized school trips are out of catchment according to the regional travel survey, the Transportation Tomorrow Survey (TTS) in 2011 and 2016 [@data_management_group_tts_2018].
-The elementary school locations and catchments were provided by the HWDSB and HWCDSB for the 2010-2011 and 2015-2016 academic years, and are shown in @fig-Fig1.
-
-Formally, elementary schools are defined as schools that provide instruction to any combination of grades between kindergarten to grade 8 (i.e.
+First, elementary schools locations, how the changed throughout the study period and their capacities are required for each studied year.
+
+Concerning school locations, the City's school boards provided the authors with locations and associated school catchments for the 2010-2011 and 2015-2016 academic years.
+Data for 2021 locations was retrieved from the City's open data portal [@hamiltonEducationalInstitutions2024].
+In Hamilton, the majority of student-aged population attend school in one of the two English public school boards [@faoOntarioSchoolBoards2023].
+One school board is referred as Hamilton Wentworth District School Board (HWDSB) and the other as Hamilton Wentworth Catholic District School Board (HWCDSB).
+School catchments for the public and public-catholic catchments indicate what residential property is assigned to what school by default.
+However, families can decide to attend schools out of catchment; in fact, 21%-23% <!--calculated in 03-travel-times-TTS.qmd--> of motorized school trips for populations aged 5 through 14 are out of catchment according to the regional travel survey (the Transportation Tomorrow Survey (TTS)) in 2011 and 2016, the most recent TTS available at the time of writing [@data_management_group_tts_2018].
+
+In addition to the school's location in 2011, 2016 and 2021, all schools were also tagged with a school status regarding their existence and/or change in location or school-seat capacity.
+A summary of the status of schools through 2011 to 2021 is illustrated in @fig-Fig1.
+@fig-Fig1 also displays a count of the number of schools with the status between years, the number of schools overall present in each year, and the symbolic representation of a school status that is used in Figures throughout the paper.
+Overall, a school can have one of seven school statuses in any given year.
+Either: 1) the school did not change throughout 2011 to 2021, 2) the school expanded (i.e., increased school-seat capacity) or was 3) relocated and expanded (N.B., schools expanded only between 2011 through 2016 as they were typically done _before_ school closures and no new Pupil Accommodation Review was commenced after 2017), 4) the school closed sometime after 2011 and before 2016 or 5) after 2016 and before 2021, and 6) the school opened sometime after 2011 and before 2016 or 7) after 2016 and before 2021.
+
+```{r}
+#| label: fig-Fig1
+#| fig-cap: "Overview of the number of schools in Hamilton in 2011, 2016 and 2021 along with the type of school status change.."
+#| echo: false
+#| out-width: "120%"
+
+knitr::include_graphics(paste0(here::here(),file="/Figures/Fig1.png"))
+#grid.raster(readTIFF(paste0(here::here(),"/Figures/Fig1.tiff")))
+```
+
+The capacity of each school is also required for each simulated year.
+This capacity is reflected as the number of 'seats' available in the school.
+Enrolment exceeding this capacity indicates over-enrolment, while enrolment below it indicates under-enrolment.
+This value is calculated by the provincial Ministry of Education and referred to as the on-the-ground capacity (OTGC).
+The current and historic OTGCs for schools are no longer publicly available [@ontarioSchoolFacilityInventory2017], so we estimated them using a multivariate regression model and validated the results with a set of OTGC values manually gathered.
+These values were retrieved for schools that had undergone a completed Pupil Accommodation Review and had publicly available School Information Profiles.
+
+For schools lacking OTGC values, two regression models were created: one for public schools and another for public catholic schools.
+The independent variables in the model were 1) a dummy variable indicating the type of school grade instruction (*MidElem*, *JrElem*, *Elem*), 2) the school's building footprint (*F*) in units of ($m^2$), and 3) the school's shortest Euclidean distance from the centroid of Hamilton's central business district area (*DistCBD*) defined in metres:
+
+- Formally, elementary schools are defined as schools that provide instruction to any combination of grades between kindergarten to grade 8 (i.e.
 typically children aged 5 to 14).
 As such, elementary includes middle schools that only instruct grades 6 to 8 (*MidElem*), primary schools that only instruct kindergarten to grade 5 or 6 (*JrElem*), and all grade elementary schools (*Elem*) which instruct all grades from kindergarten to grade 8.
-The number of 'seats' available in the school beyond or below which it will be over/under- capacity is calculated by the provincial Ministry of Education; this value is referred to as the on-the-ground capacity (*OTGC*), and is assigned to each school.
-The current/historic OTGCs are not publicly available [@ontarioSchoolFacilityInventory2017].
-However, for schools that underwent the full Accommodation Review process, OTGCs for a given year could be obtained from publicly available School Information Profiles.
-A regression model was used to estimate the OTGCs for schools that lacked OTGC information.
-The independent variables in the model were 1) the school's building footprint (*F*) in units of ($m^2$); 2) the school's shortest Euclidean distance from the centroid of Hamilton's central business district area (*DistCBD*) defined in units of (m); and 3) coefficients associated with a binary variable indicating the type of school grade instruction (*MidElem*, *JrElem*, *Elem*).
-The building footprint from an archived spatial data set [@DMTI_Spatial_2015] and the 2016 footprints were retrieved from OSM [@OpenStreetMap_2021].
-Because the age of construction for different schools is not available, we used each school's distance from the centroid of the Downtown Business Improvement Area [@Hamilton_2014] as a proxy for school age as the oldest buildings are generally located closer to the city centre [@Merrall_2021].
-Of note, the models were also used to estimate OTGC of secondary schools (*Sec*) though not included in the scope of this work; these public institutions provide instruction for grades 9 to 12 (i.e., children aged 14 to 17).
-
-Equations (\ref{eq:OTGC-public}) and (\ref{eq:OTGC-catholic}) describe the estimated OTGC for public schools and catholic schools respectively.
-School status and OTGC summed by catchment is also visualised in @fig-Fig1.
-As seen in Table \ref{TabA1-OTGC} in the Appendix, the coefficient of determination for these models is $R^2 = 0.999$ and $R^2 = 0.998$ for the Public $OTGC_{Public}$ and Catholic $OTGC_{Catholic}$ Boards respectively, and the residual standard deviation is $\sigma = 0.212$ and $sigma = 0.331$.
-<!-- [AS]: model estimation here: https://github.com/soukhova/hamONdests/blob/master/data-raw/01-school-data-preparation.Rmd -->
+*MidElem* and *JrElem* school grade instruction type is only present in the public school board.
+- The building footprint *F* was retrieved from an archived spatial data set [@DMTI_Spatial_2015] and footprints from newer schools in 2016 and 2021 from OSM [@OpenStreetMap_2021]. 
+- *DistCBD* was calculated 'as the crow flies' from each school to the centroid of the Hamilton CBD (43.256684°N, 79.869039°W).
+Notably, this variable can also be seen as a proxy for school construction age as older buildings in Hamilton are generally located closer to the CBD, the "old" Hamilton community [@Merrall_2021].
+
+Equations (\ref{eq:OTGC-public}) and (\ref{eq:OTGC-catholic}) summarise the OTGC regression model for public schools and public catholic schools respectively.
+As seen in Table \ref{TabA1-OTGC}, the coefficient of determination for these models are fantastic: $R^2 = 0.999$ and $R^2 = 0.998$ for the public $OTGC_{Public}$ and public catholic $OTGC_{Public Catholic}$ school boards respectively, and the residual standard deviation is $\sigma = 0.212$ and $sigma = 0.331$.
 
@@
 \label{eq:OTGC-catholic}
-OTGC_{Catholic} =F^{0.471}-e^{0.00003*DistCBD}+e^{2.333*Elem}
+OTGC_{Public Catholic} =F^{0.471}-e^{0.00003*DistCBD}+e^{2.333*Elem}
 \end{equation}
@@
 
+\input{Tab1.tex} <!--made in the hamONdests repo-->
+
+Taken together, @fig-Fig2 displays the city's elementary schools locations, status and their estimated or retrieved OTGCs for 2011, 2016 and 2021.
+For additional context on the degree of residential urbanization in Hamilton, the percentage of residential parcels considered 'urban' compared to 'suburban' or 'rural' is reflected at an aggregated spatial unit across all six plots in @fig-Fig2.
+This land-use classification is available within the 2021 residential parcel file described in the following sub-section.
+In @fig-Fig2, it can be seen that the majority of schools that changed status are in the HWDSB, with OTGC capacity decreasing in 2016 and 2021 compared to the previous year.
+Most school closures took place in the central and eastern parts of Hamilton's urban area (Hamilton Central), as well as in rural areas of western Hamilton (Flamborough).
+When schools did open or expand, it was primarily in the more recently urbanized southeastern area of Hamilton (Glanbrook), with a few also in Hamilton Central to offset some of the lost capacity.
+
 ```{r}
-#| label: fig-Fig1
-#| fig-cap: "The school catchments, estimated On The Ground Capacity (OTGC) of schools summed per catchment, and school status for both public and public catholic school boards for year 2011 and 2016.
-OTGC scale is presented by quartiles."
+#| label: fig-Fig2
+#| fig-cap: "The on-the-ground capacity (OTGC) of schools and school status for year 2011, 2016 and 2021 for both the HWDSB and HWCDSB.
+Schools are presented overtop a layer visualising the degree of residential urbanisation in 2021."
 #| echo: false
@@
 
-grid.raster(readTIFF(paste0(here::here(),"/Figures/Fig1.tiff")))
-```
-<!--perhaps in the legend of this map, I add the number of schools in brackets (for "school status between 2011 and 2016")?
-And the number of schools on the map in each box?
-what do you think?
-I'll also make the font on the panels bigger-->
-
-<!-- Number of schools?
-Yes Also, make the symbols that map the schools bigger, and use different shapes in addition to different colors.
-As an alternative, the OTGC could be the color, and the school category the shape.
--->
+grid.raster(readTIFF(paste0(here::here(),"/Figures/Fig2.tiff")))
+```
 
@@
 
-In addition to OTGC for schools, statistics about the student population and proportion of low-income after tax (LIM-AT) households were retrieved from the 2011 and 2016 Canadian Census [@Statistics_Canada_2011; @Statistics_Canada_2016] using the {cancensus} `R` package [@von_Bergmann_Shkolnik_Jacobs_2021].
+Secondly, a detailed account of the average student population, low-income prevalence of households, and where they may live is required for each studied year.
+
+Concerning the student population and poverty, information from the 2011, 2016 and 2021 Canadian Census [@Statistics_Canada_2011; @Statistics_Canada_2016; @Statistics_Canada_2021] was sourced using the {cancensus} R package [@von_Bergmann_Shkolnik_Jacobs_2021].
 The census releases population data by age group category, so the population aged between 5-9 years and 10-14 years were retained.
-LIM-AT refers to the proportion of private households that are below the median after-tax income [@governmentofcanadaDictionaryCensusPopulation2017].
-The LIM-AT prevalence for households with children under 18 (there is no LIM-AT prevalence tabulated for households with exclusively elementary aged children) was retrained.
-Population and LIM-AT variables were taken at the Dissemination Area (DA) level, which is the finest geographic unit publicly available.
+A common poverty measure across all three census years is the low-income after-tax measure (LIM-AT), which reflects the proportion of private households that are below the median after-tax income in the region [@governmentofcanadaDictionaryCensusPopulation2017].
+The LIM-AT prevalence for households with children under 18 was retrained for this paper as there is no LIM-AT prevalence tabulated for households with exclusively elementary aged children.
+Population and LIM-AT variables were taken at the Dissemination Area (DA) level, the finest geographic unit publicly available.
 DAs are designed by Statistics Canada with the aim of population uniformity hence DAs greatly vary in area but represent between approximately 400 and 700 (1st and 3rd quartile) in total population.
-The population aged 5 to 14 and the proportion of LIM-AT are visualised in the first and last rows in @fig-Fig2.
-
-The third source of data consists of centroids of residential parcels in 2011 and 2016 that are used to represent residential locations [@Teranet_2009].
-There are 134,340 and 139,467 residential locations in 2011 and 2016 respectively, representing `r cen_data_2011_hamiltononly$DA_households |> sum() |> prettyNum(big.mark = ",")` and `r cen_data_2016_hamiltononly$DA_households |> sum() |> prettyNum(big.mark = ",")` dwellings (according to the Canadian Census [@Statistics_Canada_2011; @Statistics_Canada_2016]).
-As the number of children per parcel are publicly unavailable, the average number of 5 to 14 year olds for each DA is divided by the number of residential households in that DA.
-Then the potential number of children at each dwelling is calculated by multiplying this DA rate by the number of dwellings in each parcel.
-<!-- NOTE: We consider the parcel centroids instead of DA centroids to more precisely calc.
-travel times.
-Since we do not know parcel student population, we assume a uniform spatial distribution of population across each DA.
-This assumption is a limitation, but it is better than what's typically done in studies that only consider the centroid of the census track.
---> Due to the proprietary nature of the parcel data, the middle row in @fig-Fig2 visualises an aggregation of the information: the average rate of 5-14 year olds per residential household at the DA level.
+The population aged 5 to 14 and the proportion of LIM-AT are visualised in the first and last rows in @fig-Fig3.
+LIM-AT prevalence is notably more concentrated within the centre of Hamilton (Hamilton Central) through 2011 to 2021, overlapping the most urbanised land-use and the largest amount of schools closed (@fig-Fig2).
+Also of note, LIM-AT prevalence drastically decreased in 2021 as a result of Pandemic benefits that reduced income inequality [@governmentofcanadaDailyPandemicBenefits2022].
+
+Where students live are estimated at the level of the residential parcel.
+Centroids for parcels in 2011, 2016 and 2021 are retrieved from @Teranet_2009, representing 134,340, 139,467 and 143,890 unique locations in those years, respectively.
+The datasets do not provide information on the number of households or children per parcel.
+To estimate this value, the average number of students per parcel is calculated using the number of school-aged population in each DA is divided by the number of households in each DA.
+This rate is assigned to all parcels within an associated DA.
+Due to the proprietary nature of the parcel data, the second row in @fig-Fig3 visualises an aggregation of the information: the average rate of 5-14 year old population per residential unit (parcel) at the DA level.
+Notably, though the rate of 5-14 year old population per parcel is higher in more peripheral (and rural) communities, there are many DAs within Hamilton Central that have high rates and populations that are similar to those in these more peripheral communities.
 
 ```{r}
-#| label: fig-Fig2
-#| fig-cap: "The magnitude (top row) and rate per residential unit (middle row) of elementary aged student population per DA in 2011 and 2016.
-Bottom row: the proportion of low-income households (prevelance of low-income after-tax, LIM-AT, measured by the 2011 and 2016 Canadian Census) with dependents under 18 years old shown.
+#| label: fig-Fig3
+#| fig-cap: "The magnitude (top row) and rate per residential unit (middle row) of elementary aged student population per DA in 2011, 2016 and 2021.
+Bottom row: the proportion of low-income households (prevelance of low-income after-tax, LIM-AT, measured by the 2011, 2016 and 2021 Canadian Census) with dependents under 18 years old shown.
 All scales are represented in quartiles."
@@
 
-grid.raster(readTIFF(paste0(here::here(),"/Figures/Fig2.tiff")))
-```
-
-## Estimated travel times and emissions
-
-It is assumed that children can go to any elementary school, but there is a preference for facilities that are more proximate.
-Based on this assumption, the fourth source of data is information regarding the mode-used and origin-destination locations of home-to-school trips from the 2011 and 2016 TTS [@data_management_group_tts_2018].
-The TTS is a travel survey conducted in the Greater Golden Horseshoe Area in Ontario every five years.
+grid.raster(readTIFF(paste0(here::here(),"/Figures/Fig3.tiff")))
+```
+
+## Multimodal travel behaviour
+
+To reflect multimodal travel behaviour for all studied years, two types of data are estimated and complied.
+First, observed home-to-school trips by mode are retained from travel surveys.
+Next, travel times are calculated assuming active and motorized modes at parcel-to-school level and at the level matching the travel survey.
+Then, travel impedance functions are calibrated using the observed travel data, and using the functions and the parcel-to-school calculated travel times as input, the travel impedance values for each residential location to each school (origin-destination pair) is estimated.
+These three steps are described as follows:
+
+First, we retrieve information about the mode used and origin-destination locations of home-to-school trips from the 2011 and 2016 Transportation Tomorrow Survey (TTS) [@data_management_group_tts_2018]; the 2021 TTS survey is unavailable at the time of writing so 2016 flows are assumed for the 2021 year.
+The TTS is a travel survey conducted in the Greater Golden Horseshoe Area in Ontario typically every five years.
 With a target 5% sampling rate, the survey is expanded to be representative at the traffic analysis zone (TAZ) level of geography.
 TAZ are spatial units created for the purpose of the TTS and pulled from the R data package {TTS2016R} [@soukhovTTS2016RDataSet2023]; a few DAs typically nest within each TAZ.
-The trip-level travel data extracted for this paper represent `r od_trips_perc_2011$TAZOrig_car_trips |> sum() |> prettyNum(big.mark = ",")` and `r od_trips_perc_2016$TAZOrig_car_trips |> sum() |> prettyNum(big.mark = ",")` motorized trips (mode used includes private car passenger, school bus, taxi, and transit) and `r od_trips_perc_2011$TAZOrig_walk_trips |> sum() |> prettyNum(big.mark = ",")` and `r od_trips_perc_2016$TAZOrig_walk_trips |> sum() |> prettyNum(big.mark = ",")` non-motorized trips (mode used includes walk and cycling) for 5 to 13 year olds from home-to-school in 2011 and 2016 respectively.
-The travel times for motorized and non-motorized travel are estimated using {r5r} [@Pereira2021r5r], which interfaces with the java-based R5 routing engine developed separately by Conveyal [@conveyalConveyalR5Routing2022].
-As travel times are not associated with the TTS, motorized and non-motorized travel times are estimated for all TAZ centroids to all TAZ centroids.
-We estimate motorized travel to be by car and non-motorized travel to be by walk as those are the most used modes in their respective categories.
-The travel times and the intensity of modal home-to-school flows are visualised in @fig-Fig3 along with the boundaries of the TAZ and DAs.
-In @fig-Fig3, not all TAZs capture an elementary school trip by both modes.
+The trip-level travel data extracted for this paper represent `r od_trips_perc_2011$TAZOrig_car_trips |> sum() |> prettyNum(big.mark = ",")` and `r od_trips_perc_2016$TAZOrig_car_trips |> sum() |> prettyNum(big.mark = ",")` motorized trips (mode used includes private car passenger, school bus, taxi, and transit) and `r od_trips_perc_2011$TAZOrig_walk_trips |> sum() |> prettyNum(big.mark = ",")` and `r od_trips_perc_2016$TAZOrig_walk_trips |> sum() |> prettyNum(big.mark = ",")` non-motorized trips (mode used includes walk and cycling) for 5 to 14 year olds from home-to-school in 2011 and 2016 respectively.
+The intensity of modal home-to-school flows are visualised in @fig-Fig4 along with the boundaries of the TAZ and DAs.
+Also in @fig-Fig4, not all TAZs capture an elementary school trip by both modes.
 For this reason, the proportion of motorized modal share is aggregated at the community level.
-Hamilton (Ancaster: `r community_car_perc_2011$TAZOrig_car_perc[1] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[1] |> percent()`, Dundas: `r community_car_perc_2011$TAZOrig_car_perc[2] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[2] |> percent()`, Flamborough: `r community_car_perc_2011$TAZOrig_car_perc[3] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[3] |> percent()`, Glanbrook: `r community_car_perc_2011$TAZOrig_car_perc[4] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[4] |> percent()`, Hamilton Central: `r community_car_perc_2011$TAZOrig_car_perc[5] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[5] |> percent()`, and Stoney Creek: `r community_car_perc_2011$TAZOrig_car_perc[6] |> percent()`-`r community_car_perc_2016$TAZOrig_car_perc[6] |> percent()`).
-<!--Unrelated: did you use information from TTS2016R?
-Or can you add it to the data package?
-And then cite your data paper.
-[AS]: fantastic idea!
-Added!
-for now I just have the TAZ boundaries in TTS2016R.
-The package can be expanded to include school travel but I'd need to add the full GTHA region.
-For this analysis I only exported Hamilton (PD region 6) beginning/ending trips.-->
+Hamilton (Ancaster - 2011: `r community_car_perc_2011$TAZOrig_car_perc[1] |> percent()`and 2016:`r community_car_perc_2016$TAZOrig_car_perc[1] |> percent()`, Dundas - 2011: `r community_car_perc_2011$TAZOrig_car_perc[2] |> percent()` and 2016: `r community_car_perc_2016$TAZOrig_car_perc[2] |> percent()`, Flamborough - 2011: `r community_car_perc_2011$TAZOrig_car_perc[3] |> percent()` and 2016: `r community_car_perc_2016$TAZOrig_car_perc[3] |> percent()`, Glanbrook - 2011: `r community_car_perc_2011$TAZOrig_car_perc[4] |> percent()` and 2016: `r community_car_perc_2016$TAZOrig_car_perc[4] |> percent()`, Hamilton Central - 2011: `r community_car_perc_2011$TAZOrig_car_perc[5] |> percent()` and 2016:`r community_car_perc_2016$TAZOrig_car_perc[5] |> percent()`, and Stoney Creek - 2011: `r community_car_perc_2011$TAZOrig_car_perc[6] |> percent()` and 2016: `r community_car_perc_2016$TAZOrig_car_perc[6] |> percent()`).
 
 ```{r}
-#| label: fig-Fig3
-#| fig-cap: "The origin-destination flows from home-based motorized and non-motorized school trips for students between 5-13 years old as retrieved from TTS 2011 and 2016.
+#| label: fig-Fig4
+#| fig-cap: "The origin-destination flows from home-based motorized and non-motorized school trips for students between 5-14 years old as retrieved from TTS 2011 and 2016.
 Flows are mapped atop the proportion of TAZ non/motorized modal share from the TTS 2011 and 2016, DA boundaries, and Hamilton community boundaries."
@@
 
-grid.raster(readTIFF(paste0(here::here(),"/Figures/Fig3.tiff")))
-```
-
-The fifth source of data consists of estimated travel time matrices from all residential parcel locations to all elementary schools for both motorized and non-motorized travel.
-The travel times are also estimated using the {r5r} package [@Pereira2021r5r], with car mode and walking representing motorized and non-motorized travel respectively.
-For both motorized and non-motorized modes, the inputs assume a maximum travel time threshold of 60 minutes and the free-flow OpenStreetMap road network of Hamilton (retrieved using Geofabrik [@geofabrikOntarioOpenStreetMapGeofabrik2022] and manually edited to only include the network within Hamilton).
-Further, only walking trips shorter than 27 min minutes were retained.
-This travel time is approximately equivalent to 1.6 km which is the point within-catchment students qualify for transport provided by the school board [@HWDSB_2019].
-Estimated trip-weighted motorized travel times are on average 16.5 minutes in 2011 and 16.8 minutes in 2016 (summary statistics for both years are approximately; min: 0.5, Q1: 11.0, Q2: 16.0, Q3: 21.0, max: 60.0).
-Estimated trip-weighted non-motorized travel times are on average 18.0 minutes in 2011 and 18.1 minutes in 2016 (summary statistics for both years are approximately; min: 0.5, Q1: 13.0, Q2: 19.0, Q3: 24.0, max: 27.0).
-<!--to calc.
-the summary statistics i loaded "OD_2011" and "OD_2016" from data-products (big files) and ran: quantile(OD_2011|>filter(mode=="w" & travel_time_p50 <= 27)|>pull(travel_time_p50), weights=OD_2011|>filter(mode=="w")|>pull(kid_popadj))
-and 
-mean(OD_2011|>filter(mode=="w" & travel_time_p50 <= 27)|>pull(travel_time_p50), weights=OD_2011|>filter(mode=="w")|>pull(kid_popadj))
--->
-
-These travel time matrices along with the origin-destination flows were used to create modal trip length distribution (TLD) functions.
-Analysis of the TLD is a useful technique to calibrate impedance functions as they are a representation of the propensity of travel by trip length [@horbachov_theoretical_2018; @batista_estimation_2019].
-The empirical TLD is used to fit a density distribution using maximum likelihood and moment matching techniques and the Nelder-Mead and Brent methods for direct optimization available within the {fitdistrplus} package [@fitdistrplus_2015].
+grid.raster(readTIFF(paste0(here::here(),"/Figures/Fig4.tiff")))
+```
+
+Next, the travel times matrices for motorized and active travel are estimated using the 'travel_time_matrix()' function part of the {r5r} package [@Pereira2021r5r].
+Travel time matrices are estimated for all TAZ centroids to TAZ centroids (matrices of size for 2011 and 2016: 234 x 234) as they ar enot reported in the TTS.
+Additionally matrices are estimated for all residential parcels to schools (matrices of size 134,340 x 147 for 2011, 139,467 x 137 for 2016, and 143,890 x 132 for 2021) for additional granularity.
+{r5r} is a R-based interface to the R5 routing engine developed separately by Conveyal [@conveyalConveyalR5Routing2022].
+For simplicity, we assume motorized travel is by car and non-motorized travel is by walking, as these are the most common modes in their respective categories.
+For travel time calculations, we set a maximum threshold of 60 minutes and use the free-flow OpenStreetMap road network of Hamilton [@geofabrikOntarioOpenStreetMapGeofabrik2022].
+While all car trips are retained, walking trips over 27 minutes are filtered out.
+This value approximately corresponds to a distance of 1.6 km at a walking speed of 3.6 km/h, which is the threshold for qualifying for motorized transport provided by the school board [@HWDSB_2019].
+
+In terms of the travel impedance function: we assume that children can go to any elementary school, however, there is a preference for facilities that are more proximate.
+Based on this assumption, we match the associated TAZ-to-TAZ travel times to all observed student-to-school travel flows from the TTS; these are visualised as grey bar columns in @fig-Fig5.
+Using these observed values, the  _theoretical_ trip length distribution (TLD) functions are calibrated.
+TLDs can be interpreted as travel impedance functions as they represent the propensity of realized travel, by trip length [@horbachov_theoretical_2018; @batista_estimation_2019].
+The TLDs are calibrated using the maximum likelihood and moment matching techniques and the Nelder-Mead and Brent methods for direct optimization available within the {fitdistrplus} R package [@fitdistrplus_2015].
+The theoretical TLD for each mode and available study year is visualised in blue in @fig-Fig5.
 Based on goodness-of-fit criteria and diagnostics, the gamma and exponential distributions were selected for the motorized and non-motorized modal distributions respectively.
@@
 The exponential distribution is defined by the rate ($\beta$) parameter of `r round(exp2011_nonmotor$estimate[1], 3)` (2011) and `r round(exp2016_nonmotor$estimate[1], 3)` (2016).
-The TLDs for the motorized and non-motorized trips are shown in @fig-Fig4.
 For reference, the gamma distribution and the exponential distribution function are displayed in Equations (\ref{eq:exp-dist}) and (\ref{eq:gamma-dist}) where $x$ is $c_{ij}$:
@@
 ```{r}
-#| label: fig-Fig4
-#| fig-cap: "Empirical (grey bars) and theoretical (blue curves) motorized and non-motorized impedance functions.
+#| label: fig-Fig5
+#| fig-cap: "Observed (grey bars) and theoretical (blue curves) motorized and non-motorized impedance functions.
 Motorized theoretical impedance function is based on the gamma distribution function and non-motorized on the expoential distribution function"
@@
 
-grid.raster(readTIFF(paste0(here::here(),"/Figures/Fig4.tiff")))
-```
-
-The sixth and final data source relates to motorized GHG estimations.
-It assumed that all non-motorized trips produce no emissions and those that are motorized take the average emission factor of 0.1 kg $CO_{2}e$  per minute travel.
-This emission factor is assumed based on an average speed of 40 km/h, the 2,380 g $CO_{2}e$ produced when 1 L of gasoline (well-to-wheel) is combusted, and an internal combustion engine passenger vehicle fuel efficiency of 7 g/100km as discussed in @soukhovOccupancyGHGEmissions2022.
+grid.raster(readTIFF(paste0(here::here(),"/Figures/Fig5.tiff")))
+```
+
+Lastly, to achieve greater granularity, the TLDs calibrated using TAZ-to-TAZ flows are used to estimate travel impedance values for each parcel-to-school flow based on calculated parcel-to-school travel times.
+These values represent the likelihood that students in a parcel will travel to a school, informed by observed home-to-school flows from the TTS.
 
@@
 Accessibility is defined as the "potential for spatial interaction".
-It is classically presented as the gravity-based measure popularized by @hansenHowAccessibilityShapes1959.
-It takes the following general multimodal formulation (Equation (\ref{eq:multimodal-conventional-accessibility})):
+It is classically presented as the gravity-based measure defined in @hansenHowAccessibilityShapes1959 and takes the following general multimodal formulation:
 
@@
 
-As indicators of urban structure, Hansen-type measures are informative, but the meaning of 10,000 accessible school-seats is harder to pin down: how many opportunities must any single student have access to?
-Furthermore, in scenario comparisons, the interpretation of the percentage change in accessible school seats is unclear.
+As indicators of urban structure, Hansen-type accessibility measures like Equation (\ref{eq:multimodal-conventional-accessibility}) are informative in reflecting the magnitude of access but meaning of the value itself is elusive.
+The significance of 10,000 accessible school-seats is hard to pin down: how many opportunities must any single student have access to?
+Furthermore, this opaque interpretation especially is complicated when comparing accessibility of school seats between years, such as different years as in this work.
 The interpretability of Hansen-type accessibility has been discussed in numerous studies, including recently by @hu_2019_measuring, @kelobonye2020measuring, and in greater depth by @merlin2017competition along with @soukhovIntroducingSpatialAvailability2023 and @soukhovMultimodalSpatialAvailability2024.
-The interpretation of accessibility is suggested to depend on how many people demand the opportunity, especially for exclusive opportunity-types like schools-seats (i.e., one school-seat is for one student).
-
-In this paper, we benefit from new developments in accessibility research, particularly the multimodal spatial availability measure [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024].
+The interpretation of accessibility depends on how many people demand for the opportunity, especially for exclusive opportunity-types like schools-seats (i.e., one school-seat is for one student).
+
+In this paper, our work benefits from new developments in accessibility research, particularly the multimodal spatial availability measure [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024].
 Spatial availability is a singly-constrained accessibility measure that accounts for competition by students using different modes for exclusive opportunities, such as school seats.
@@
 This proportional allocation of opportunities yields an interpretable and meaningful measure of opportunity access, particularly when comparing across modes, at the spatial resolution of a residential parcel, and multiple time periods.
-See @soukhovMultimodalSpatialAvailability2024 for further discussion, multimodal spatial availability $V_{i}^m$ is defined as given by Equation (\ref{eq:spatial-availability}).
+See @soukhovMultimodalSpatialAvailability2024 for further discussion, multimodal spatial availability $V_{i}^m$ is defined as given by Equation (\ref{eq:spatial-availability}):
 
@@
 Hansen-type accessibility is not designed to preserve the number of opportunities in the region, it simply counts the intensity of opportunities that those in a zone can potentially interact with (weighted by the friction of distance).
-Also, as discussed in @soukhovIntroducingSpatialAvailability2023, popular competitive accessibility measures such as the 2SFCA [@josephMeasuringPotentialPhysical1982; @weibullAxiomaticApproachMeasurement1976; @shenLocationCharacteristicsInnercity1998; @luoMeasuresSpatialAccessibility2003] are internally inconsistent, and the only way it preserves the number of opportunities is if the effect of the impedance function is ignored when expanding the values of opportunities per capita to obtain the total number of opportunities.
+Also, as discussed in @soukhovIntroducingSpatialAvailability2023, popular competitive accessibility measures such as the two-step floating catchment area (2SFCA) [@josephMeasuringPotentialPhysical1982; @weibullAxiomaticApproachMeasurement1976; @shenLocationCharacteristicsInnercity1998; @luoMeasuresSpatialAccessibility2003] are internally inconsistent, and the only way it preserves the number of opportunities is if the effect of the impedance function is ignored when expanding the values of opportunities per capita to obtain the total number of opportunities.
 On the other hand, the proportional allocation procedure associated with calculating multimodal spatial availability $V_i^m$ consistently returns a number of opportunities available to populations by mode that matches the total number of opportunities in the region when summed.
@@
 
-This paper uses the multimodal spatial availability $V_i^m$ and spatial availability per student $v_i^m$ measures to quantify accessibility changes for the following reasons.
-
-First, all residential parcels are associated to their respective census DA, TTS TAZ, and community boundary based on spatial location (aggregated by DA and shown as a per capita rate in the second row of @fig-Fig2).
+To summarise the methodology and the data as described in Section 3, multimodal spatial availability $V_i^m$ and spatial availability per student $v_i^m$ is calculated for each parcel for each studied year as follows:
+
+- First, all residential parcels are associated to their respective census DA, TTS TAZ, and community boundary based on spatial location.
 Each parcel is assigned a number of 'potential' motorized and non-motorized student aged population (from DA) and modal share (from TTS) to calculate a motorized and non-motorized population balancing factor $F_{i}^{pm}$.
-For each parcel, the following is retained: the 2011 and 2016 census average elementary-aged student population per parcel (@fig-Fig2 second row) and the average motorized and non-motorized share as informed by the TTS aggregated by community (@fig-Fig3).
-
-Second, the school capacity $O_j$ is estimated and associated with each school (@fig-Fig1).
+For each parcel, the elementary-aged student population per parcel (@fig-Fig3 second row) and the motorized and non-motorized share as informed by the TTS aggregated by community (@fig-Fig4) is retained.
+- Second, the school capacity $O_j$ is estimated and associated with each school (@fig-Fig2).
 All residential locations are assumed to be able to access all schools by non-motorized and/or motorized mode.
 All origins (residential locations) can reach all schools by motorized mode, but few residential locations can reach schools within a 27 minute non-motorized trip.
-
-Third, the motorized and non-motorized impedance-balancing factors $F_{ij}^{cm}$ are calculated using a mode-specific impedance function (@fig-Fig4) based on the respective estimated travel times for each origin (parcel) to school combination (OD flows retained from the TTS).
+- Third, the motorized and non-motorized impedance-balancing factors $F_{ij}^{cm}$ are calculated using a mode-specific impedance function (@fig-Fig5) based on the respective estimated travel times for each origin (parcel) to school combination.
 Using network estimated travel time sensitive to observed parcel origin to school destination flows conceptually addresses the aggregation error that could result from using less granular zonal units to represent origin/destinations [@kaneParcelsPointsProximity2020; @kwanSpaceTimeIntegralMeasures1998; @hewkoMeasuringNeighbourhoodSpatial2002].
-Finally, outputs from all three stages are joined together.
+- Finally, outputs from all three stages are joined together.
 $F_{i}^{pm}$ joined with $F_{ij}^{cm}$ yields the total balancing factor $F^{tm}_{ij}$ which serves to proportionally allocate the school capacity $O_j$ to each parcel.
@@
 
-The motorized and non-motorized spatial availabilities of school seats $V_i^m$ for both 2011 and 2016 are presented in the top half of @fig-Fig5.
-The sum of $V_i^m$ values in a year for both motorized and non-motorized population equals the city-wide OTGC (across both public and catholic school boards) in that year.
-This is a result of $V_i^m$'s proportional allocation mechanism, singly-constraining opportunities through the proportion allocation balancing factors.
-In this way, $V_i^m$ results can be interpreted as a proportion of the total OTGC in that year as the sum of $V_i^m$ for both modes in a year is equal to the total OTGC in that year.
-In @fig-Fig5, $i$ is a DA, so $V_i^m$ is the sum of the school seat availability per parcel in each DA.
-Each parcel is assigned a potential student population based on the average rate of school-aged population per residential parcel in that DA and the travel impedance weighted travel time to all schools.
-The top four plots of @fig-Fig5 demonstrate different magnitudes in part due to significantly lower non-motorized modal share (refer to @fig-Fig3), however it is apparent that both mode using populations have higher spatial availability when they are more proximate to schools.
-Hence, all DAs have higher values within Hamilton Central and those more proximate to schools in less-urban communities.
-Trends appear similar for both 2011 and 2016: this aligns with intuition, proximate populations using motorized modes benefit from the ability to reach more distant destinations than non-motorized modes.
-
-When $V_i^m$ is divided by the number of students by mode in that DA, $v_i^m$ is derived (bottom four plots in @fig-Fig5).
-The $v_i^m$ plots demonstrate different spatial distribution trends than $V_i^m$ plots.
-If our aim is to provide sufficient availability to schools for both motorized and non-motorized populations (e.g., greater than 1.0 $v_i^m$ school-seat availability per student), this rate is only achieved in the core of the city for motorized modes (greens and yellows).
-Notably, this rate is only available to non-motorized populations in DAs that are in less-densely populated rural areas that are school proximate and very few pockets of Hamilton Central.
-The majority of schools closed were within Hamilton Central, so non-motorized populations in DAs within Hamilton Central were the most negatively impacted by losses in availability while certain motorized populations captured these losses as availability gains.
+The probability density plots of the spatial availability values $V_i$ and the per capita spatial availability $v_i$ values for both modes and all three years are displayed in @fig-Fig6 as a city-wide overview.
+The left column displays parcel-level values, and the right reflects parcel-level values aggregated at the 2021 DA grid.
+Both the parcel and DA level values reflect the spatial unit at which they were calculated or aggregated.
+However, the values of the $V_i$ plots are not directly comparable since the x- and y- axes are set at different scale limits to highlight between year differences.
+However, overall trends in their distribution can be compared: in all four $V_i$ plots, the 2011 data is more right-skewed than more current years, indicating that spatial availability values used to be higher (e.g., more school-seat access).
+School closures and consolidations had an impact despite the background dynamics of assumed student population, mode use, travel time and parcel location.
+The DA level $V_i$ plots show the distribution of each DA value, which is the summation of all parcel $V_i$ values within a DA.
+In 2011, DAs can be observed to also have more right-skewed spatial availability values than previous years.
+In contrast, the 2016 and 2021 distributions are more intense and heavily skewed toward lower values.
+
+Although $V_i$ values cannot be directly compared, zonal per capita $v_i$ values can be as the school-age population per spatial unit (e.g., parcel or DA) is used to normalize the respective $V_i$ value.
+$v_i$ values are presented in the bottom four plots of @fig-Fig6.
+It can be observed that the same 2011 right-skewed trend persists across all modes.
+Motorized mode-using students have access to more spatially available school seats than non-motorized mode using students (in 2011, approximately 1.3 school-seats per student as compared to 0.7 and in 2021 approximately 1.2 compared to 0.5).
+Due to school-seats per student at the DA level being an interpretable aggregation, values at this level will be the discussion focus of the remainder of this paper.
 
 ```{r}
-#| label: fig-Fig5
-#| fig-cap: "Multimodal spatial availability (purples) and spatial availability per mode-using student (diverging reds and greens) aggregated at the DA level for 2011 and 2016.
-Scales are represented in quartiles."
+#| label: fig-Fig6
+#| fig-cap: "Spatial availability (Vi) and per capita spatial availability (vi) probability density distributions for the city of Hamilton in 2011, 2016 and 2021.
+Aggregated at the parcel level (left column) and DA level (right column)."
 #| echo: false
@@
 
-grid.raster(readTIFF(paste0(here::here(),"/Figures/Fig5.tiff")))
-```
-
-To highlight this point, @fig-Fig6 visualises the change in school spatial availability between 2016 and 2011 for both mode where $i$ are 2016 DAs.
-Though the 2011 and 2016 DAs system are very similar (see @fig-Fig2 and @fig-Fig3), they are not identical: `r (cen_data_2011_hamiltononly$GeoUID %in% cen_data_2016_hamiltononly$GeoUID |> length() - cen_data_2011_hamiltononly$GeoUID %in% cen_data_2016_hamiltononly$GeoUID |> sum()) + (cen_data_2016_hamiltononly$GeoUID %in% cen_data_2011_hamiltononly$GeoUID |> length() - cen_data_2016_hamiltononly$GeoUID %in% cen_data_2011_hamiltononly$GeoUID |> sum())` DAs changed in their spatial extent (i.e., added, removed, changed boundaries).
-The proportional allocation mechanism of spatial availability provides flexibility in aggregating parcel-level values at whichever spatial unit is meaningful.
-For this case, the change in $V_i^m$ and $v_i^m$ represented is the subtraction between the value aggregated at 2016 DA for the 2016 and 2011 scenario.
+grid.raster(readTIFF(paste0(here::here(),"/Figures/Fig6.tiff")))
+```
+
+<!--maybe we change i to DA..?-->
+Representing the right column of @fig-Fig6 spatially, @fig-Fig7 displays the spatial availability for each DA for a given mode-using population $m$ (e.g., the sum of school seat availability per parcel in each DA for motorized or non-motorized students).
+@fig-Fig7 visualizes the spatial distribution of motorized and non-motorized school seat availability for 2011, 2016, and 2021.
+As a reminder, the sum of $V_i^m$ values for both motorized and non-motorized populations in a given year equals the city-wide OTGC (i.e., school seats across both the public english HWDSB and public english catholic HWCDSB).
+This is due to $V_i^m$'s proportional allocation mechanism, singly-constraining opportunities through the proportion allocation balancing factors.
+In this way, $V_i^m$ values can be interpreted as a proportion of the total OTGC for that year, as the sum of $V_i^m$ for both modes in a year equals the total OTGC for that year.
+
+Referring to the first two rows of $V_i$ plots in @fig-Fig7 (purples), it is notable that the magnitudes of $V_i$ values for the motorized and non-motorized populations are tremendously different.
+Again, each $V_i$ value reflects the number of school seats spatially available to the mode-using population at that zonal level, so due to the lower non-motorized modal share (refer to @fig-Fig4), non-motorized values will be lower.
+However, both mode-using populations have higher spatial availability values (darker purples) in DAs that are more proximate to schools.
+Hence, all DAs have higher values within Hamilton Central and those more proximate to schools in less-urban communities.
+This trend persists across 2011, 2016 and 2016, aligning with intuition: populations with shorter travel times to with relatively greater schools are calculated to have high school seat spatial availability.
+Motorized populations have shorter travel times at DAs further in distance from schools, while non-motorized populations are only competitive at proximate distances.
+
+The bottom two rows of $v_i$ plots in @fig-Fig7 (yellows to greens) demonstrate the spatial availability per mode-using student population e.g., $V_i^m$ divided by the number of students aged 5-14 using a motorized mode or non-motorized mode.
+It is notable that for motorized populations, their spatial availability per student is above 1 (green) within the center of the city.
+Conversely, this rate is only available to non-motorized populations in DAs that are in less-densely populated rural areas that are school proximate and very few pockets of Hamilton Central in 2011 and 2016.
+In 2021, even fewer DAs within Hamilton Central have non-motorized spatial availability rates above 1.
 
 ```{r}
-#| label: fig-Fig6
-#| fig-cap: "Change in multimodal spatial availability and spatial availability per mode-using student between 2016 and 2011.
-Change is the subtraction of parcel-level results between 2016 and 2011 scenarios summed for each DA in the 2016 Canadian Census DA system.
-Scales are represented in quartiles, with Q4 rounded to 0."
+#| label: fig-Fig7
+#| fig-cap: "Multimodal spatial availability (purples) and spatial availability per mode-using student (diverging reds and greens) aggregated at the DA level for 2011, 2016 and 2021.
+Scales are represented in deciles"
 #| echo: false
@@
 
-grid.raster(readTIFF(paste0(here::here(),"/Figures/Fig6.tiff")))
-```
-
-From @fig-Fig6 (top plots), it is apparent that decreases (pinks and reds) in $V_i^m$ appear across the city for both mode-using populations.
-As the majority of closed schools are within the core of the city: the more urban Hamilton Central and urban-areas of other communities see a concentration of this decreased spatial availability to school seats.
-Evidently, the number of OTGC in the city (equal to the sum of $V_i^m$) decreased between the two years, so some sort of decrease is to be expected.
-However, even when spatial availability is divided by the mode-using population (which also decreased between the years) $v_i^m$ (bottom plots) continues to demonstrate decreases, especially those locally proximate to closed schools.
-
-This decrease follows different patterns depending on the mode.
-$v_i^{non-motorized}$ is most impacted for parcels within a 27 minute walk travel time to schools that closed: these DAs demonstrate the most dramatic decrease.
-Smaller-area DAs (in Hamilton-Central) and larger-area DAs (in rural communities) proximate to closed schools present Q1 decreases (red).
-For $v_i^{motorized}$, the range offered by motorized mode (estimated by shortest-path car travel times) is city-wide, so since the majority of schools were centrally located, the decrease is concentrated within the downtown core of Hamilton Central.
-Some motorized-mode using populations in DAs do experience an increase in $V_i^m$ and $v_i^m$: these DAs are often relatively proximate to schools but more immediately outside walking range.
-Under competitive and constrained calculations, these DAs benefit from the losses in spatial availability that are predominately experienced by the non-motorized populations within Hamilton Central.
-
-The decrease in school seat spatial availability can be further interpreted.
-How does potential non-motorized travel to school decrease?
-How does the increase in potential motorized travel minutes impact GHG emissions?
-@fig-Fig7 visualises these related intermediates for one-way home-to-school travel in a given day.
-The top plot of @fig-Fig7 demonstrates the sum of the change in non-motorized minutes per 2016 DA.
-These travel time minutes are the difference between the aggregated 2016 and 2011 scenarios for non-motorized travel times multiplied by the proportion of walking/cycling mode-use and rate of student population associated with the parcels in each 2016 DA.
-It can be seen that Q1 decreases are proximate to closed schools.
-Smaller decreases (and even a few increases) are apparent in a range further from closed schools, especially those also in proximity to the few opened/expanded schools.
-Intuitively, these trends mirror what is shown within the $V^{non-motorized}$ landscape in @fig-Fig6 (Pearson correlation coefficient of `r cor(SAvail_im_DA_changes |> filter(mode == "non-motorized") |> pull(c_total_tt), SAvail_im_DA_changes |> filter(mode == "non-motorized") |> pull(c_V_im)) |> round(digits=3)`) as travel time is an input into the cost balancing factor $F^{cm}_{ij}$.
-Similarly, the bottom plot (@fig-Fig7) visualises the change in estimated emissions from motorized travel minutes.
-The change in GHG emissions is also highly correlated with $V^{motorized}$ in @fig-Fig6 (Pearson correlation coefficient of `r cor(SAvail_im_DA_changes |> filter(mode == "motorized") |> pull(c_emissions_total_car_tt), SAvail_im_DA_changes |> filter(mode == "motorized") |> pull(c_V_im)) |> round(digits=3)`) as it is also a product of travel time.
+grid.raster(readTIFF(paste0(here::here(),"/Figures/Fig7.tiff")))
+```
+
+To further explore these changes, the difference in $V_i$ (first two rows) and difference in $v_i$ (second two rows) between 2011-2016, 2016-2021 and overall between 2011-2021 are displayed in @fig-Fig8 along with schools that changed during that time period.
+2011-2016 and 2016-2021 add together to equal 2011-2021 changes quite clearly, e.g.
+in the third row, the 2011-16 plots a small decrease within the core of the city, and 2016-2021 demonstrates a more significant decrease in the core and western peripheral communities.
+Together, 2011-2021 plot reflects the changes in these two plots.
+
+A few interesting spatial trends can be observed in @fig-Fig8.
+First, DAs in proximity to schools that closed (square points) see losses in $V_i$ (reds) for both motorized and especially non-motorized populations.
+For non-motorized populations, all closed school resulted in a decrease in $V_i$.
+The majority of school closures took place in the central and eastern parts of Hamilton's urban area (Hamilton Central), as well as in rural areas of western Hamilton (Flamborough).
+Hence, losses characterise those neighbourhoods.
+Secondly, gains in $V_i$ (blues) are seen where schools are opened (diamond points) or expanded (triangle points), especially for motorized populations.
+These gains are most notably in the more recently urbanized southeastern area of Hamilton (Glanbrook), along with a few pockets of DAs in proximity to new or expanded schools in Hamilton Central.
+Thirdly, the concentration of students in proximity to schools and the extent of their OTGC matters from the perspecitve of modal advantage.
+In areas with fewer proximate school options and lower OTGC relative to the amount of students, a change in a school results in a big impact on spatial availability values.
+For instance, in the community of Ancaster, one school expanded in the rural area in this community between 2011-2016.
+The motorized student population, especially in rural DAs, saw gains in spatial availability while non-motorized populations saw losses.
+Students in proximity to the schools in Ancaster have high non-motorized spatial availability, but when the school was expanded (and due to an under-served population) motorized populations saw an increase in spatial availability at the direct expense of non-motorized populations' spatial availability.
+
+Exploring changes in $V_i$ provides a holistic view of spatial availability for a given DA.
+However, if the focus is on the ratio of spatial availability per student, the bottom two rows in @fig-Fig8 offer more insight.
+For instance, a community may have had an initially average rate of school seats per student but is expecting an increase in student population so it expands a school.
+However, due to an increase in population, the spatial availability per student overall still decreases between these two time periods.
+This case appears to occur in Ancaster between 2011 to 2016: while motorized $V_i$ increased, motorized $v_i$ slightly decreased, and non-motorized $v_i$ decreased relatively more than non-motorized $V_i$.
+As shown in @fig-Fig3, the student population and the number of students per residential unit increased more rapidly than the expanded school could accommodate, leading to a decrease in $v_i$ values in DAs proximate to expanded schools.
+
+Similarly, in the southeastern community of Glanbrook, student populations grew from 2011 to 2021, and new schools opened.
+However, $v_i$ rates decreased because OTGC did not keep pace with student population growth.
+As another example, in Hamilton Central and the eastern community of Stoney Creek, changes in $V_i$ are unpatterned, while $v_i$ demonstrate more uniform decreases, with the pattern varying by modal-range for motorized versus non-motorized populations.
+In these communities, the student population remained relatively stable, though OTGC decreased overall.
+$v_i$ is illuminating to consider when comparing changes in spatial availability rates.
 
 ```{r}
-#| label: fig-Fig7
-#| fig-cap: "Change in active travel time minutes (top plot) and home-to-school GHG emissions (bottom plot) between 2011 and 2016 scenarios.
-Change is the subtraction of parcel-level results between 2016 and 2011 scenarios summed for each DA in the 2016 Canadian Census DA system.
-Scales are represented in quartiles, with Q4 rounded to 0."
+#| label: fig-Fig8
+#| fig-cap: "Change in multimodal spatial availability and spatial availability per mode-using student from 2011 to 2016, 2016 to 2021, and together from 2011 to 2021.
+Change is the subtraction of parcel-level results in that given year and summed for each DA in the 2021 Canadian Census DA grid.
+Scales are represented in deciles."
 #| echo: false
+#| out-width: "120%"
+#| fig-pos: "H"
+
+grid.raster(readTIFF(paste0(here::here(),"/Figures/Fig8.tiff")))
+```
+
+Changes in multimodal spatial availability is important: A summary of spatial availability per Hamilton community and associated dimensions in 2021, along with the percentage change between 2011 and 2021, is presented in @tbl-Tab1.
+
+```{r, warning=FALSE}
+#| label: tbl-Tab1
+#| tbl-cap: "Spatial availability and associated dimensions aggregated by communities in 2011 and how much they changed by 2021"
 #| out-width: "90%"
-#| fig-pos: "H"
-
-grid.raster(readTIFF(paste0(here::here(),"/Figures/Fig7.tiff")))
-```
-
-Notably, the mode type of spatial availability loss is important.
-The change in active travel minutes and GHG emissions highlights the negative externalities of _decreased_ non-motorized spatial availability.
-Normatively, our cities should be increasing spatial availability for non-motorized mode users given the climate crisis.
-The wave of school closures in Hamilton increased motorized spatial availability and, hence, potential GHG emissions.
-The policy drastically reduced non-motorized spatial availability, and this is seen in the decrease in potential non-motorized travel minutes as well as increased GHG emissions in rural areas that did benefit from motorized spatial availability gains.
-The proportion of non-motorized mode use is low in both 2011 and 2016, however, the closure of schools that are more proximate to student-population density eliminates the potential of those trips ever becoming active.
-And who is most impacted by the reduction of non-motorized spatial availability?
-A summary of spatial availability per Hamilton community and associated dimensions in 2016, along with the percentage change between 2011 and 2016, is summarized in @tbl-Tab1.
-
-\pagebreak
-
-```{r include=FALSE}
-SAvail_im_COMM_changes_df <- SAvail_im_COMM_changes |> st_drop_geometry()
-
-LIMAT_COMM <- SAvail_im_DA |> st_drop_geometry() |> filter(mode == "motorized" & year == "2016") |>
-  group_by(COMMUNITY) |> 
-  summarise(COMM_prev_LIMAT_under18_2016 = mean(DA_prev_LIMAT_under18, na.rm=T))
-
-LIMAT_COMM_2011 <- SAvail_im_DA |> st_drop_geometry() |> 
-  filter(mode == "motorized" & year == "2011") |>
-  group_by(COMMUNITY) |> 
-  summarise(COMM_prev_LIMAT_under18_2011 = mean(DA_prev_LIMAT_under18, na.rm=T))
-
-LIMAT_COMM <- left_join(LIMAT_COMM, LIMAT_COMM_2011, by = "COMMUNITY") |>  
-  mutate(c_COMM_prev_LIMAT_under18 = ((COMM_prev_LIMAT_under18_2016 - COMM_prev_LIMAT_under18_2011)/COMM_prev_LIMAT_under18_2011) |>
-           scales::percent(accuracy = 2L)) 
-rm(LIMAT_COMM_2011)
-
-SCHOOL_SEATS_COMM <- ALL_SCHOOLS_Elem |> st_drop_geometry() |> group_by(COMMUNITY) |> 
-  summarise(OTGC2011 = sum(OTGC2011, na.rm=T),
-            OTGC2016 = sum(OTGC2016, na.rm=T)) |>
-  mutate(c_OTGC_perc = ((OTGC2016 - OTGC2011)/OTGC2011) |> scales::percent(accuracy = 2L))
-
-SAvail_im_COMM_changes_df <- SAvail_im_COMM_changes_df |> left_join(LIMAT_COMM, by="COMMUNITY") 
-SAvail_im_COMM_changes_df <- SAvail_im_COMM_changes_df |> left_join(SCHOOL_SEATS_COMM, by="COMMUNITY") |> mutate_if(is.numeric, round, 1) 
-```
-
-```{r include=FALSE}
-SAvail_im_COMM_changes_df <- SAvail_im_COMM_changes_df |> 
-  dplyr::select(c("COMMUNITY",
-                  "mode",
-                  "COMM_prev_LIMAT_under18_2016",
-                  "c_COMM_prev_LIMAT_under18",
-                  "kid_2016",
-                  "c_kid_perc",
-                  "OTGC2016",
-                  "c_OTGC_perc",
-
-                  "V_im_2016",
-                  "c_V_im_perc",
-                  "v_im_2016",
-                  "c_v_im_perc",
-
-                  "emissions_2016",
-                  "c_emissions_perc",
-                  "emissions_perkid_2016",
-                  "c_emissions_perkid_perc",
-
-                  "total_tt_2016",
-                  "c_total_tt_perc",
-                  "total_tt_perkid_2016",
-                  "c_total_ttperkid_perc")) |> arrange(desc(COMM_prev_LIMAT_under18_2016))
-```
-
-```{r include=FALSE}
-tdata_1 <- SAvail_im_COMM_changes_df |> 
-  filter(mode == "motorized")|> 
-  dplyr::select(c("COMMUNITY",
-                  "COMM_prev_LIMAT_under18_2016",
-                  "kid_2016",
-                  "OTGC2016"))
-
-tdata_1 <- insertRow(tdata_1 |> as.matrix(),2,
-          SAvail_im_COMM_changes_df |> 
-            filter(mode == "motorized" & COMMUNITY == "HA") |> 
-            dplyr::select(c("COMMUNITY", 
-                            "c_COMM_prev_LIMAT_under18", 
-                            "c_kid_perc",
-                            "c_OTGC_perc")) |> as.matrix())
-
-tdata_1 <- insertRow(tdata_1,4,
-          SAvail_im_COMM_changes_df |> 
-            filter(mode == "motorized" & COMMUNITY == "DU") |> 
-            dplyr::select(c("COMMUNITY", 
-                            "c_COMM_prev_LIMAT_under18", 
-                            "c_kid_perc",
-                            "c_OTGC_perc")) |> as.matrix())
-
-tdata_1 <- insertRow(tdata_1,6,
-          SAvail_im_COMM_changes_df |> 
-            filter(mode == "motorized" & COMMUNITY == "SC") |> 
-            dplyr::select(c("COMMUNITY", 
-                            "c_COMM_prev_LIMAT_under18", 
-                            "c_kid_perc",
-                            "c_OTGC_perc")) |> as.matrix())
-
-tdata_1 <- insertRow(tdata_1,8,
-          SAvail_im_COMM_changes_df |> 
-            filter(mode == "motorized" & COMMUNITY == "AN") |> 
-            dplyr::select(c("COMMUNITY", 
-                            "c_COMM_prev_LIMAT_under18", 
-                            "c_kid_perc",
-                            "c_OTGC_perc")) |> as.matrix())
-
-tdata_1 <- insertRow(tdata_1,10,
-          SAvail_im_COMM_changes_df |> 
-            filter(mode == "motorized" & COMMUNITY == "FL") |> 
-            dplyr::select(c("COMMUNITY", 
-                            "c_COMM_prev_LIMAT_under18", 
-                            "c_kid_perc",
-                            "c_OTGC_perc")) |> as.matrix())
-
-tdata_1 <- insertRow(tdata_1,12,
-          SAvail_im_COMM_changes_df |> 
-            filter(mode == "motorized" & COMMUNITY == "GL") |> 
-            dplyr::select(c("COMMUNITY", 
-                            "c_COMM_prev_LIMAT_under18", 
-                            "c_kid_perc",
-                            "c_OTGC_perc")) |> as.matrix())
-```
-
-```{r}
-tdata_2 <- SAvail_im_COMM_changes_df |> 
-  filter(mode == "motorized")|> 
-  dplyr::select(c("COMMUNITY",
-                  "V_im_2016",
-                  "v_im_2016",
-                  "emissions_2016",
-                  "emissions_perkid_2016"))
-
-tdata_2 <- insertRow(tdata_2 |> as.matrix(),2,
-          SAvail_im_COMM_changes_df |> 
-            filter(mode == "motorized" & COMMUNITY == "HA") |> 
-            dplyr::select(c("COMMUNITY", 
-                            "c_V_im_perc", 
-                            "c_v_im_perc",
-                            "c_emissions_perc",
-                            "c_emissions_perkid_perc")) |> as.matrix())
-
-tdata_2 <- insertRow(tdata_2,4,
-          SAvail_im_COMM_changes_df |> 
-            filter(mode == "motorized" & COMMUNITY == "DU") |> 
-            dplyr::select(c("COMMUNITY", 
-                            "c_V_im_perc", 
-                            "c_v_im_perc",
-                            "c_emissions_perc",
-                            "c_emissions_perkid_perc")) |> as.matrix())
-
-tdata_2 <- insertRow(tdata_2,6,
-          SAvail_im_COMM_changes_df |> 
-            filter(mode == "motorized" & COMMUNITY == "SC") |> 
-            dplyr::select(c("COMMUNITY", 
-                            "c_V_im_perc", 
-                            "c_v_im_perc",
-                            "c_emissions_perc",
-                            "c_emissions_perkid_perc")) |> as.matrix())
-
-tdata_2 <- insertRow(tdata_2,8,
-          SAvail_im_COMM_changes_df |> 
-            filter(mode == "motorized" & COMMUNITY == "AN") |> 
-            dplyr::select(c("COMMUNITY", 
-                            "c_V_im_perc", 
-                            "c_v_im_perc",
-                            "c_emissions_perc",
-                            "c_emissions_perkid_perc")) |> as.matrix())
-
-tdata_2 <- insertRow(tdata_2,10,
-          SAvail_im_COMM_changes_df |> 
-            filter(mode == "motorized" & COMMUNITY == "FL") |> 
-            dplyr::select(c("COMMUNITY", 
-                            "c_V_im_perc", 
-                            "c_v_im_perc",
-                            "c_emissions_perc",
-                            "c_emissions_perkid_perc")) |> as.matrix())
-
-tdata_2 <- insertRow(tdata_2,12,
-          SAvail_im_COMM_changes_df |> 
-            filter(mode == "motorized" & COMMUNITY == "GL") |> 
-            dplyr::select(c("COMMUNITY", 
-                            "c_V_im_perc", 
-                            "c_v_im_perc",
-                            "c_emissions_perc",
-                            "c_emissions_perkid_perc")) |> as.matrix())
-```
-
-```{r}
-tdata_3 <- SAvail_im_COMM_changes_df |> 
-  filter(mode == "non-motorized")|> 
-  dplyr::select(c("COMMUNITY",
-                  "V_im_2016",
-                  "v_im_2016",
-                  "total_tt_2016",
-                  "total_tt_perkid_2016"))
-
-tdata_3 <- insertRow(tdata_3 |> as.matrix(),2,
-          SAvail_im_COMM_changes_df |> 
-            filter(mode == "non-motorized" & COMMUNITY == "HA") |> 
-            dplyr::select(c("COMMUNITY", 
-                            "c_V_im_perc", 
-                            "c_v_im_perc",
-                            "c_total_tt_perc",
-                            "c_total_ttperkid_perc")) |> as.matrix())
-
-tdata_3 <- insertRow(tdata_3,4,
-          SAvail_im_COMM_changes_df |> 
-            filter(mode == "non-motorized" & COMMUNITY == "DU") |> 
-            dplyr::select(c("COMMUNITY", 
-                            "c_V_im_perc", 
-                            "c_v_im_perc",
-                            "c_total_tt_perc",
-                            "c_total_ttperkid_perc")) |> as.matrix())
-
-tdata_3 <- insertRow(tdata_3,6,
-          SAvail_im_COMM_changes_df |> 
-            filter(mode == "non-motorized" & COMMUNITY == "SC") |> 
-            dplyr::select(c("COMMUNITY", 
-                            "c_V_im_perc", 
-                            "c_v_im_perc",
-                            "c_total_tt_perc",
-                            "c_total_ttperkid_perc")) |> as.matrix())
-
-tdata_3 <- insertRow(tdata_3,8,
-          SAvail_im_COMM_changes_df |> 
-            filter(mode == "non-motorized" & COMMUNITY == "AN") |> 
-            dplyr::select(c("COMMUNITY", 
-                            "c_V_im_perc", 
-                            "c_v_im_perc",
-                            "c_total_tt_perc",
-                            "c_total_ttperkid_perc")) |> as.matrix())
-
-tdata_3 <- insertRow(tdata_3,10,
-          SAvail_im_COMM_changes_df |> 
-            filter(mode == "non-motorized" & COMMUNITY == "FL") |> 
-            dplyr::select(c("COMMUNITY", 
-                            "c_V_im_perc", 
-                            "c_v_im_perc",
-                            "c_total_tt_perc",
-                            "c_total_ttperkid_perc")) |> as.matrix())
-
-tdata_3 <- insertRow(tdata_3,12,
-          SAvail_im_COMM_changes_df |> 
-            filter(mode == "non-motorized" & COMMUNITY == "GL") |> 
-            dplyr::select(c("COMMUNITY", 
-                            "c_V_im_perc", 
-                            "c_v_im_perc",
-                            "c_total_tt_perc",
-                            "c_total_ttperkid_perc")) |> as.matrix())
-```
-
-```{r include=FALSE}
-tdata_1 <- t(tdata_1) |> data.frame() 
-tdata_2 <- t(tdata_2) |> data.frame() 
-tdata_3 <- t(tdata_3) |> data.frame()
-
-tdata <- rbind(tdata_1,tdata_2,tdata_3)
-tdata <- tdata[-c(5,10),]
-rownames(tdata) <- c("", "LIM-AT", "Pop.", "OTGC", 
-                     "V (mt)", "v (mt)", "GHG", "GHG/pop",
-                     "V (nmt)", "v (nmt)", "tt", "tt/pop")
-```
-
-```{r}
-#| label: tbl-Tab1
-#| tbl-cap: "Spatial availability and associated dimensions aggregated by communities"
-#| out-width: "90%"
-
-tdata_hux <- as_hux(tdata) %>% set_escape_contents(FALSE)
-# tdata_hux[1,1:12] <- c("Hamilton C.", "", "Dundas" ,"", "Stoney Creek", "", "Ancaster", "", "Flamborough", "", "Glanbrook", "")
-# # tdata_hux[1,1:12] <- c("HA", "", "DU" ,"", "SC", "", "AN", "", "FL", "", "GL", "")
-# tdata_hux[2,1:12] <- c("", "% {\\Delta}", "" , "% {\\Delta}", "", "% {\\Delta}Δ", "", "% ${\\Delta}", "", "% {\\Delta}", "", "% {\\Delta}")
-
-tdata_hux <- set_contents(tdata_hux,
-                          row= c(1),
-                          col = c(1:12),
-                          value = c("Hamilton C.", "", "Dundas" ,"", "Stoney Creek", "", "Ancaster", "", "Flamborough", "", "Glanbrook", ""))
-tdata_hux <- set_contents(tdata_hux,
-                          row= c(2),
-                          col = c(1:12),
-                          value = c("", "% Δ", "", "% Δ", "", "% Δ", "", "% Δ",
-                                    "", "% Δ", "", "% Δ"))
-
-options("huxtable.latex_use_fontspec" = TRUE) #use the same font as the document
-
-tdata_hux %>% 
-  insert_row("Motorized", fill = "", colspan=12, after=5) %>% 
-  insert_row("Non-motorized", fill = "", colspan=12, after=10) %>% 
-  merge_across(1,c(1:2)) %>%
-  merge_across(1,c(3:4)) %>%
-  merge_across(1,c(5:6)) %>%
-  merge_across(1,c(7:8)) %>%
-  merge_across(1,c(9:10)) %>%
-  merge_across(1,c(11:12)) %>%
-  add_rownames() |>
-
-  set_text_color(c(3,10), 3, "darkgreen") |>
-  set_text_color(c(4,5,7,8,9,12,13,14), 3, "darkred") |>
-
-  set_text_color(c(3), 5, "darkgreen") |>
-  set_text_color(c(4,7,9,12,13,14), 5, "darkred") |>
-
-  set_text_color(c(3,4,5,7,9,15), 7, "darkgreen") |>
-  set_text_color(c(12,13,14), 7, "darkred") |>
-
-  set_text_color(c(3,4,5,7,9,10,15), 9, "darkgreen") |>
-  set_text_color(c(8,12,13,14), 9, "darkred") |>
-
-  set_text_color(c(3,5,8,15), 11, "darkgreen") |>
-  set_text_color(c(4,7,9,12,13,14,15), 11, "darkred") |>
-
-  set_text_color(c(4,5,7,9,10), 13, "darkgreen") |>
-  set_text_color(c(3,12,13,14,15), 13, "darkred") |>
-  set_font_size(everywhere, everywhere, 6) |> 
-  set_all_padding(0) |>
-  set_bold(everywhere, 1, value = TRUE) |>
-  set_bold(1:2, everywhere, value = TRUE) |>
-  set_bold(c(6,11), everywhere, value = TRUE) |>
-  set_contents(c(1,6,11), 1, "") |>
-  set_right_border(c(3:15), c(3,5,7,9,11), brdr(1, "solid", "grey")) |>
-  #set_bottom_border(c(2,5,6,10,11), c(2:13), brdr(1, "solid", "grey")) |>
-  set_align(everywhere, everywhere, "left")|>
-  set_align(c(1,2,6,11), everywhere, "center")|>
-  set_caption("Spatial availability and associated dimensions aggregated by communities") |>
-  set_label("tbl-Tab1") |> as_flextable()
+
+pal <- function(x) {
+  f_neg <- scales::col_numeric(
+    palette = c('red', '#ffffff'),
+    domain = c(-1, 0)
+  )
+  f_pos_smaller <- scales::col_numeric(
+    palette = c('#ffffff', 'palegreen3'),
+    domain = c(0.00000001, 1.5)
+  )
+  f_pos <- scales::col_numeric(
+    palette = c('palegreen3', 'palegreen3'),
+    domain = c(1.50000001, 13)
+  )
+  ifelse(x <= 0, f_neg(x), 
+         ifelse(x > 0 & x <= 1.5, 
+                f_pos_smaller(x), f_pos(x)))
+}
+
+#setting default font and size for the table
+set_flextable_defaults(
+  font.family = "Arial", font.size = 10,
+  border.color = "gray", big.mark = "", digits = 1, pct_digits=1)
+
+flextable(tdata) |>
+
+  bold(part = "header") |> #bolding header
+
+  bg(j = c("HA.1", "DU.1", "SC.1", "GL.1","FL.1","AN.1"), bg = pal) |> #adding background colurs to % changes
+
+  mk_par(j = c(3), value = as_paragraph(fmt_pct(HA.1))) |> #formating the % as with the percentage sign
+  mk_par(j = c(5), value = as_paragraph(fmt_pct(DU.1))) |>
+  mk_par(j = c(7), value = as_paragraph(fmt_pct(SC.1))) |>
+  mk_par(j = c(9), value = as_paragraph(fmt_pct(GL.1))) |>
+  mk_par(j = c(11), value = as_paragraph(fmt_pct(FL.1))) |>
+  mk_par(j = c(13), value = as_paragraph(fmt_pct(AN.1))) |>
+
+  set_header_labels(values = c("","Hamilton C.", "% Δ", "Dundas" ,"% Δ",
+                                    "Stoney Creek", "% Δ", "Glanbrook", "% Δ",
+                                    "Flamborough", "% Δ", "Ancaster", "% Δ")) |> #changing names of headers
+
+merge_at(i = 4, j = 1:13) |> #merging all cells at row 4 to create "Motorized" label
+merge_at(i = 7, j = 1:13) |> #merging all cells at row 4 to create "Non-motorized" label
+bold(i=c(4,7), j=1:13, part = "body") |> #bolding labels
+hline(i =c(3,4,6,7), j=1:13) |>  #adding horizontal line above and below the labels
+
+align(align = "center", part = "all")
 
@@
 Due to spatial availability's flexible proportional allocation mechanism and this cases' calculation at the parcel level, results can also be summarized by community (@tbl-Tab1).
-The majority of schools were closed in Hamilton Central.
-This community represents `r SAvail_im_COMM_changes |> st_drop_geometry() |> filter(COMMUNITY == "HA") |> dplyr::select("kid_proportion_2016") |> sum() |> scales::percent()` of the 5-14 year old population in 2016.
-Though the population between 2011 and 2016 decreased, the rate of school-seat spatial availability decreased disproportionately more (population by `r SAvail_im_COMM_changes |> st_drop_geometry() |> filter(COMMUNITY == "HA") |> transmute(calc = ((sum(kid_2016)-sum(kid_2011))/sum(kid_2011)) |> scales::percent()) |> pull() |> first()` and spatial availability by `r SAvail_im_COMM_changes |> st_drop_geometry() |> filter(COMMUNITY == "HA") |> transmute(calc = ((sum(V_im_2016)-sum(V_im_2011))/sum(V_im_2011)) |> scales::percent()) |> pull() |> first()`).
-This disproportionate decrease is greatest in Hamilton Central than any other community.
-This may have equity concerns as Hamilton Central has the highest LIM-AT prevalence, with `r SAvail_im_COMM_changes_df |> filter(COMMUNITY == "HA" & mode == "motorized") |> dplyr::select("COMM_prev_LIMAT_under18_2016") |> paste0("%")` of households LIM-AT in 2016, 3.7 times greater than Glanbrook, a community with the lowest level of LIM-AT prevalence in the city.
-Between 2016 and 2011, Hamilton Central had the lowest levels of emissions per student and saw the highest per student increase in 2016.
-Hamilton Center had the highest non-motorized trip potential, and it was reduced in part by the school closure policies.
-
-Conversely, communities with the lowest LIM-AT benefit the most, notably in gains to motorized spatial availability.
-Glanbrook, Flamborough and Ancaster have the lowest prevalence of LIM-AT of any Hamilton Community and together account for `r SAvail_im_COMM_changes |> st_drop_geometry() |> filter(COMMUNITY == "FL" | COMMUNITY == "GL" | COMMUNITY == "AN") |> dplyr::select("kid_proportion_2016") |> sum() |> scales::percent()` of the 5-14 year old population in 2016.
-These communities experienced a growth in student population (or only a small decrease) and proportionally benefited from an increase in motorized spatial availability.
-
-Across the city, those who commute by non-motorized mode see the most drastic lose in spatial availability.
-In all communities, spatial availability decreases at greater levels than population.
-Notably in Dundas (2nd highest LIM-AT), the only school in the community closed, leaving the population in the community with no other option but to use motorized modes.
-Glanbrook (lowest LIM-AT), which experienced a relatively sharp increase in student population along with non-motorized using population, suffered a lose in non-motorized spatial availability despite seeing the highest increase in motorized spatial availability.
-Overall, the gain in motorized spatial availability resulted in a further erosion of non-motorized spatial availability.
-This erosion is related to a lose in potential active home-to-school travel minutes (`r ((SAvail_im_COMM_changes |> filter(mode == "non-motorized") |> pull(total_tt_2011)  |> sum()) -  (SAvail_im_COMM_changes |> filter(mode == "non-motorized") |> pull(total_tt_2016)  |> sum())) |> round(digit=2)` fewer minutes than in 2011, decreasing the potential active travel time per (all mode using) student from `r ((SAvail_im_COMM_changes |> filter(mode == "non-motorized") |> pull(total_tt_2011)  |> sum()) / (SAvail_im_COMM_changes |> st_drop_geometry() |> pull(kid_2011) |> sum())) |> round(digit=2)` to `r ((SAvail_im_COMM_changes |> filter(mode == "non-motorized") |> pull(total_tt_2016)  |> sum()) / (SAvail_im_COMM_changes |> st_drop_geometry() |> pull(kid_2016) |> sum())) |> round(digit=2)` minutes), as well as an increase in potential GHG emissions city-wide (`r (SAvail_im_COMM_changes |> st_drop_geometry() |> pull(emissions_2016) |> sum()) -  (SAvail_im_COMM_changes |> st_drop_geometry() |> pull(emissions_2011) |> sum()) |> round(digit=2)` kg $CO_2e$ more emission than in 2011, increasing the kg $CO_2e$ per child rate for home-to-school trips from `r ((SAvail_im_COMM_changes |> st_drop_geometry() |> pull(emissions_2011) |> sum()) / (SAvail_im_COMM_changes |> st_drop_geometry() |> pull(kid_2011) |> sum())) |> round(digit=2)` to `r ((SAvail_im_COMM_changes |> st_drop_geometry() |> pull(emissions_2016) |> sum()) / (SAvail_im_COMM_changes |> st_drop_geometry() |> pull(kid_2016) |> sum())) |> round(digit=2)`).
-
-# Conclusion
-
-In this paper, we estimate the change in multimodal spatial availability as a result of school consolidations/closures between the years of 2011 and 2016 in an empirical case study of Hamilton, a mid-size city (\~570,000 pop) in Ontario, Canada [@governmentofcanadaProfileTableCensus2022].
-The spatial availability and spatial availability per student are calculated at the parcel level and aggregated at the Canadian Census DA spatial unit and at the six Hamilton communities for comparison.
-Schools closed and the number of school seats decreased so spatial availability declined overall: we discuss where, for whom, and with what potential impacts.
-
-Overall, the closure of `r ((ALL_SCHOOLS_Elem %>% filter(Year != "2016" & Level == "Elementary") %>% pull(SchoolID) %>% length() - ALL_SCHOOLS_Elem %>% filter(Year != "2011" & Level == "Elementary") %>% pull(SchoolID) %>% length())/ALL_SCHOOLS_Elem %>% filter(Year != "2016" & Level == "Elementary") %>% pull(SchoolID) %>% length()) %>% scales::percent()` of elementary schools in Hamilton between 2011 and 2016 resulted in `r (((SAvail_im_COMM_changes$V_im_2016 |> sum()) - (SAvail_im_COMM_changes$V_im_2011 |> sum()))  / SAvail_im_COMM_changes$V_im_2011 |> sum()) |> scales::percent()` decline in school-seat availability city-wide (`r (ALL_SCHOOLS_Elem %>% st_drop_geometry() %>% filter( (Year == "2011" | Year == "2011 and 2016") & Level == "Elementary") %>% dplyr::select(OTGC2011) %>% sum() - ALL_SCHOOLS_Elem %>% st_drop_geometry() %>% filter( (Year == "2016" | Year == "2011 and 2016") & Level == "Elementary") %>% dplyr::select(OTGC2016) %>% sum()) %>% prettyNum(big.mark=",", digits=1)` fewer school-seats), but specifically a `r (((SAvail_im_COMM_changes |> filter(mode == "non-motorized") |> pull(V_im_2016) |> sum()) - (SAvail_im_COMM_changes  |> filter(mode == "non-motorized") |> pull(V_im_2011) |> sum()))  / SAvail_im_COMM_changes  |> filter(mode == "non-motorized") |> pull(V_im_2011) |> sum()) |> scales::percent()` decline in non-motorized spatial availability.
+The majority of schools were closed in Hamilton Central as seen in the largest reduction in OTGC by percentage and magnitude.
+Hamilton Central represents `r (cSAvail_im_COMM2011_2021_df |> filter(COMMUNITY == "HA" & mode == "c") |> dplyr::select("kid_popadj_2021") |> sum() / cSAvail_im_COMM2011_2021_df |> filter(mode == "c") |> dplyr::select("kid_popadj_2021") |> sum() )|> scales::percent()` of the 5-14 year old population in 2021.
+Though the population decreased a modest amount from 2011 to 2021, the rate of school-seat spatial availability decreased disproportionately more (population by `r cSAvail_im_COMM2011_2021_df |> filter(COMMUNITY == "HA" & mode=="c") |> dplyr::select("ckid_popadj") |> sum() |> scales::percent(0.1)` and spatial availability by `r (((cSAvail_im_COMM2011_2021_df |> filter(COMMUNITY == "HA") |> dplyr::select("V_im_2021") |> sum()) - (cSAvail_im_COMM2011_2021_df |> filter(COMMUNITY == "HA") |> dplyr::select("V_im_2011") |> sum() )) /  ((cSAvail_im_COMM2011_2021_df |> filter(COMMUNITY == "HA") |> dplyr::select("V_im_2011") |> sum() ))) |> scales::percent(0.1)` for both modes).
+Schools and additional OTGC was not sufficiently expanded to provide the same levels of spatial availability as in 2011.
+This disproportionate decrease in spatial availability to students is greater in Hamilton Central by magnitude than any other community.
+
+From the perspective of equity, Hamilton Central has the highest LIM-AT prevalence, with `r ct201621_2011 |> filter(COMMUNITY == "HA") |> dplyr::select("LIMAT_2016") |> pull() |> round() |> paste0("%")` of households LIM-AT in 2016, more than 3 times greater than Ancaster, Flamborough and Glanbrook, communities with the lowest level of LIM-AT prevalence.
+Is the disproportionate decrease in spatial availability within Hamilton Central fair?
+In some ways it is: communities with the lowest LIM-AT prevalence appear to benefit from gains in motorized spatial availability.
+Glanbrook, Flamborough and Ancaster together account for `r ((ct201621_2011 |> filter(COMMUNITY == "FL" | COMMUNITY == "GL" | COMMUNITY == "AN") |> dplyr::select("kid_popadj_2021") |> sum()) / (ct201621_2011|> dplyr::select("kid_popadj_2021") |> sum()) ) |> scales::percent()` of the 5-14 year old population in 2021.
+However, they access `r ((cSAvail_im_COMM2011_2021 |> filter(COMMUNITY == "FL" | COMMUNITY == "GL" | COMMUNITY == "AN") |> dplyr::select("V_im_2021") |> sum()) / (cSAvail_im_COMM2011_2021|> dplyr::select("V_im_2021") |> sum()) ) |> scales::percent()` of the spatial availability of school seats.
+Hamilton Central currently captures `r ((cSAvail_im_COMM2011_2021 |> filter(COMMUNITY == "HA") |> dplyr::select("V_im_2021") |> sum()) / (cSAvail_im_COMM2011_2021|> dplyr::select("V_im_2021") |> sum()) ) |> scales::percent()` of the spatial availability for it's `r ((ct201621_2011 |> filter(COMMUNITY == "HA") |> dplyr::select("kid_popadj_2021") |> sum()) / (ct201621_2011|> dplyr::select("kid_popadj_2021") |> sum() )) |> scales::percent(0.1)` of the population.
+With student aged population growing in other communities, OTGC should be expanded - potentially to levels that Hamilton had in 2011 or beyond (`r (cSAvail_im_COMM2011_2021 |> filter(COMMUNITY == "HA" & mode == "c") |> dplyr::select("v_im_kid_popadj_2011")) |> pull() |> round(1)` school-seats per motorized mode using student and `r (cSAvail_im_COMM2011_2021 |> filter(COMMUNITY == "HA" & mode == "w") |> dplyr::select("v_im_kid_popadj_2011")) |> pull() |> round(1)` school-seats per active mode using student).
+However, should those gains come at a loss to other communities, especially those with a significantly higher proportion of households who are lower-income, a strong determinant of transport poverty?
+
+Furthermore, while there are gains in motorized spatial availability in certain communities by some metrics, there are losses in non-motorized spatial availability per student from 2011 to 2021 in all communities.
+And the losses are drastic.
+Communities with high community averag spatial availability per students with values above 1 school-seat per student saw losses, as well as communities below this value.
+
+# Discussion and conclusion
+
+In this paper, we compared how multimodal spatial availability changed after a wave of school closures and consolidations in Hamilton.
+To do so, we constructed the student-aged population, OTGC of schools and their locations, and their travel behaviour to generate spatial availability landscapes for the three study years.
+We aggregated the resulting values at the DA and community level and descriptively compared differences.
+We demonstrated that city-wide there are decreases in $V_i$ for both mode-using populations: the majority of students in the city have access to fewer school-seats than they would have in 2011.
+Furthermore, as the majority of closed schools are within the core of the city: the more urban Hamilton Central and urban-areas of other communities saw the largest amount of this decrease.
+And at a local-scale, students in residences proximate to schools that closed also saw dramatic reduce in non-motorized spatial availability values.
+
+Evidently, the number of OTGC in the city decreased between 2011 and 2021, so some sort of decrease was to be expected.
+However, should that decrease be felt hardest within neighbourhoods with the highest LIM-AT?
+By students who have the potential to travel actively to school?
+Normatively, our cities should be increasing spatial availability for non-motorized mode users given the benefits of active travel and the climate crisis.
+Hamilton's policies from 2011 through 2021 drastically reduced non-motorized spatial availability, and gains in motorized spatial availability especially in more rural areas.
+The proportion of non-motorized mode use is low in both 2011 and 2021, however, the closure of schools that are more proximate to student-population density eliminates the potential of those trips ever becoming active.
+
+Overall, the closure of `r -((SCHOOLS %>% filter(Year_2 != "2016") %>% pull(SchoolID) %>% length() - SCHOOLS %>% filter(Year != "2016") %>% pull(SchoolID) %>% length())/SCHOOLS %>% filter(Year != "2016") %>% pull(SchoolID) %>% length()) %>% scales::percent()` of elementary schools in Hamilton between 2011 and 2021 resulted in `r -(((cSAvail_im_COMM2011_2021$V_im_2021 |> sum()) - (cSAvail_im_COMM2011_2021$V_im_2011 |> sum()))  / cSAvail_im_COMM2011_2021$V_im_2011|> sum()) |> scales::percent()` decline in school-seat availability city-wide ( `r  -(( ((avail_2021 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2021)) |> select(OTGC) |> sum())) - (avail_2011 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2011)) |> select(OTGC) |> sum())) / (avail_2011 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2011)) |> select(OTGC) |> sum()) ) |> scales::percent()` fewer school-seats), but specifically a `r (((cSAvail_im_COMM2011_2021 |> filter(mode == "w") |> pull(V_im_2021) |> sum()) - (cSAvail_im_COMM2011_2021  |> filter(mode == "non-w") |> pull(V_im_2011) |> sum()))  / cSAvail_im_COMM2011_2021  |> filter(mode == "w") |> pull(V_im_2011) |> sum()) |> scales::percent()` decline in non-motorized spatial availability.
 Though non-motorized modal share was low in both 2011 and 2016, the _potential_ for active transport trips was significantly eroded as schools within a 27 minute walk catchment were closed in the more densely populated and central areas of the city.
@@
 The motivation of operational savings associated with public service consolidation for the school boards did not account for the mobility-related burdens _certain_ families are now saddled with.
-
-In addition to spatial availability impacts, associated mobility measures are assessed.
-City-wide, home-to-school non-motorized minutes (for a one-way trip) reduced by `r SAvail_im_COMM_changes |> st_drop_geometry() |> filter(mode == "non-motorized") |> transmute(total_tt = total_tt_2016-total_tt_2011) |> sum() %>% abs() %>% prettyNum(big.mark=",", digits=1)` and GHG emissions associated with motorized home-to-school travel increased by `r SAvail_im_COMM_changes |> st_drop_geometry() |> filter(mode == "motorized") |> transmute(emissions = emissions_2016-emissions_2011) |> sum() %>% abs() %>% prettyNum(big.mark=",", digits=1)` kg $CO_2e$.
-The decision to consolidate schools, increase the length of motorized travel and decrease the potential for non-motorized impacts will have daily impacts on students and their families [@mandicAdolescentsPerceptionsWalking2022; @pabayoImportanceActiveTransportation2012], the broader community [@pietrabissaSchoolAccessCity2023;@MerrallWhatsASchool2024; @bittencourtEvaluatingAccessibilityAvailability2023] and the environment [@pantelakiImpactHomeschoolCommuting2024; @rongReviewResearchLowcarbon2022].
-
-Like many other localities [@Lee_Lubienski_2017; @Autti_Hyry-Beihammer_2014; @Beuchert_Humlum_Nielsen_Smith_2018], Hamilton is not immune to top-down operational efficiency assessments that determine if community infrastructure is underutilized and closed.
+The decision to consolidate schools, likely has increased the length of motorized travel, though this would need to be further investigated.
+A decrease in non-motorized trips will have daily impacts on students and their families [@mandicAdolescentsPerceptionsWalking2022; @pabayoImportanceActiveTransportation2012], the broader community [@pietrabissaSchoolAccessCity2023;@MerrallWhatsASchool2024; @bittencourtEvaluatingAccessibilityAvailability2023] and the environment [@pantelakiImpactHomeschoolCommuting2024; @rongReviewResearchLowcarbon2022].
+
+Like many other localities [@Lee_Lubienski_2017; @AuttiHyryBeihammer2014; @Beuchert_Humlum_Nielsen_Smith_2018], Hamilton is not immune to top-down operational efficiency assessments that determine if community infrastructure is underutilized and closed.
 From the perspective of the provincial government of Ontario, the school closures which occurred in Hamilton resulted in operational savings-per-student.
-We demonstrate how these decisions resulted in both families and students paying a price through spatial availability, lost potential walk minutes, and increased GHG emissions.
+We demonstrate how these decisions resulted in both families and students paying a price through multimodal spatial availability.
 
@@
 In this way, the spatial availability demonstrates what _potential interaction_ a student could have based on historic and average travel behaviour and shortest network-path travel times if residing in point in space.
-This representation is average, and by design obfuscates considerations that could make non-average and shortest-path travel unrealistic such as safety and built-environment concerns [@leeNeighborhoodEnvironmentsUtilitarian2021; @yumitaSchoolCommutingBarriers2021] nor considers significant non-transportation factors [@palmRolePublicTransit2020].
+This representation is average, and by design obfuscates considerations that could make non-average and shortest-path travel unrealistic such as safety and built-environment concerns [@leeNeighborhoodEnvironmentsUtilitarian2021; @yumitaSchoolCommutingBarriers2021] nor considers significant non-transportation factors that impact fulsome school accessibility like quality [@bittencourtEvaluatingAccessibilityAvailability2023].
 Further, this study does not consider transit, school-bus, or cycling modes which have implications on how the GHG emissions estimation can be interpreted.
@@
 
-\pagebreak
-
-# Appendix
-
-\input{TabA1.tex} 
-<!--made in the hamONdests repo-->
+<!-- BELOW IS REJECTED TEXT / CODE -- WILL DELETE SOON -->
+
+<!-- An intermediate product in spatial availability calculations are travel time, @fig-Fig8 displays the change in _potential_ travel time between 2011 and 2021 for motorized and non-motorized populations.
+-->
+
+<!-- Changes in school seat spatial availability throughout the years can be further interpreted.
+How does something tangibly measurable, such as motorized and non-motorized potential travel minutes change?
+@fig-Fig9 visualises these related intermediates for one-way home-to-school travel in a given day, namely the sum of the change in non-motorized minutes per DA.
+These travel time minutes are the difference between the aggregated 2021 and 2011 scenarios for non-motorized travel times multiplied by the proportion of walking/cycling mode-use and rate of student population associated with the parcels in each DA.
+-->
+
+<!-- Firstly, there are modal disparities.
+For motorized modes - the range in travel time change for students changed between -15 minutes to an increase in 2 minutes.
+As these are potential decreases in motorized minutes, a decrease is good.
+Increases are seen city-wide, as more of the population, on average, must use motorized transprot to access school-seat as fewer school-seats are proximate.
+However, this trend is not seen in Glanbrook.
+Most of the community see decreases in motorozied travel minutes.
+Indeed, schools opened and were expanded exactly where students are.
+But what about other communities?
+-->
+
+<!-- In terms of non-motorized minutes  -->
+
+<!-- From an environmental perspective, increase in motorized minutes is good and increase in non-motorized minutes is bad.
+However, does the built environment support this?
+Captured active mobility.
+-->
+
+<!-- It can be seen that decreases are proximate to closed schools.
+Smaller decreases (and even a few increases) are apparent in a range further from closed schools, especially those also in proximity to the few opened/expanded schools.
+Intuitively, these trends mirror what is shown within the $V^{non-motorized}$ landscape in @fig-Fig7  -->
+
+<!-- This decrease follows different patterns depending on the mode.
+$v_i^{non-motorized}$ is most impacted for parcels within a 27 minute walk travel time to schools that closed: these DAs demonstrate the most dramatic decrease.
+Smaller-area DAs (in Hamilton-Central) and larger-area DAs (in rural communities) proximate to closed schools present Q1 decreases (red).
+For $v_i^{motorized}$, the range offered by motorized mode (estimated by shortest-path car travel times) is city-wide, so since the majority of schools were centrally located, the decrease is concentrated within the downtown core of Hamilton Central.
+Some motorized-mode using populations in DAs do experience an increase in $V_i^m$ and $v_i^m$: these DAs are often relatively proximate to schools but more immediately outside walking range.
+Under competitive and constrained calculations, these DAs benefit from the losses in spatial availability that are predominately experienced by the non-motorized populations within Hamilton Central.
+-->
+
+<!-- ```{r} -->
+<!-- #| label: fig-Fig8 -->
+<!-- #| fig-cap: "Change in active travel minutes (top plot) and motorized travel minutes (bottom plot) between 2011 and 2021. Scales are represented in decibles." -->
+<!-- #| echo: false -->
+<!-- #| out-width: "90%" -->
+<!-- #| fig-pos: "H" -->
+
+<!-- grid.raster(readTIFF(paste0(here::here(),"/Figures/Fig8.tiff"))) -->
+<!-- ``` -->
+
+<!-- The data reviewed spans 10 years -- school closures that occured from 2011 until the 2017 memortorium on school closures, and census data from 2011 to 2021.
+The conversation on school closures is still pertinent.
+Recently, school boards have been calling the current provincial government to lift the momoratorium on closures, as school boards cannot run deficits and additional funding is frozen.
+school boards are forced to balance their budgets, so are cancelling programs as they cannot close under-utilised schools.
+School boards don't have their own agency, they are creatures of the province.
+This mimics the limiting of municipal powers, the munciaplity that is given power by the province over community affairs.
+
+We argue with this paper, that the trip from home to school is inherently a community level one.
+CLosing schools is unpopularly politically, but school boards were forced to do it due to limited budgets before the memorotium.
+Under the memortium and limited budget, school boards must cut program.
+They'd like the ability to close schools to balance budgets, however inthis paper - we argue this is determinental from the perspective of children's walkable access to schools and associatd environmental emissions.
+Furthermore, the trip to school is one that is inherently a community level affair.
+To close schools to sustain programs is a false choice: why not well funded schools in walkable proximities?
+Who is planning proximity if not local powers?
+-->
+
+<!-- School boards are also under increase fiscal pressure as inflation-adjusted per-pupil provincial funding, the main funding source for school boards, has continued to decline since 2018 [@opsbaOntariosEducationFunding2024].
+-->
+
+<!-- As will be later demonstrated in this paper, the increase in student population is spatially uneven.
+Areas with highest children population are characterized by newly developed low-density suburban housing.
+School boards cannot change residential land-use and built form that incentives an increase in student-aged population in neighbourhoods experiencing a decline in student-aged population and/or under-enrolled schools.
+-->
+
+<!-- Our study is focued on travel implications, e.g., proximity.
+Many factors imapct accessibility such as quality.
+Holisitc funding of school system is required.
+Transportation is just a single important dimension.
+-->
+
+<!-- While previous research has investigated existing conditions, such as highlighting that living closer to schools increases the likelihood of children's active travel [@wongModeShiftingSchool2011] and traffic safety implications of motorized modes negative impact on children's accessibility to schools [@tavakoliTrafficDangersPotential2024].
+-->

### 2024-12-12 paper/Manuscript.tex
draft revised manuscript - for other author reviews

@@
 \hypersetup{
-  pdftitle={School closures and consolidations: the active travel and emission implications of reduced accessibility},
+  pdftitle={10 years of school closures and consolidations: the impact on school accessibility},
   pdfauthor={Anastasia Soukhov; Christopher D. Higgins; Antonio Páez; Moataz Mohamed},
@@
 
-\title[School closures and consolidations: the active travel and
-emission implications of reduced accessibility]{School closures and
-consolidations: the active travel and emission implications of reduced
-accessibility}
+\title[10 years of school closures and consolidations: the impact on
+school accessibility]{10 years of school closures and consolidations:
+the impact on school accessibility}
 
@@
 
-\abstract{Reducing emissions and increasing active travel are top
-priorities for many communities. Often, though, these priorities are
+\abstract{Reducing motorized travel and increasing active travel are top
+priorities for many communities. However, these priorities are
 confronted with pressures to reduce the fiscal burden of public
-services, including schools. Public schools have been strained to do
-more with less, and governments are opting to close under-capacity
-schools and expand existing schools to accommodate lost capacity. The
-objective of this study is to quantify the travel burdens of school
-consolidation/closure policies. We do this through the case of Hamilton,
-a mid-sized city in Ontario, Canada. Between the years 2011 and 2015,
-7\% of the elementary schools closed in the city. We study how these
-closures changed the accessibility landscape for the elementary school
-aged population, as well as the active travel and emission implications.
-In terms of methods, we use spatial availability, a singly-constrained
-multimodal measure of spatial accessibility. Thanks to its proportional
-allocation mechanism, spatial availability allows for an interpretable
-comparison between `before' and `after' the implementation of school
-closure/consolidation policy. Analysis is conducted at the parcel-level
-with all school seats being eligible and travel behaviour based on
-observed home-to-school patterns for motorized and non-motorized modes.
-We report estimates of system-wide impacts that result from the
-closures, such as potential growth in GHG emission and reduction of
-walking minutes. Findings demonstrate that overall, spatial availability
-decreased in the city. Though the school consolidation policy may have
-resulted in cost-savings in operation and maintenance of facilities,
-students, families, and communities pay the price in reduced
-opportunities for physical activity and greater motorized
-mobility-related pollution.}
+services, including schools. Public school boards have been strained to
+do more with less, and opting to close under-capacity schools to deliver
+educational services more cost-effectively. However, once a school is
+closed, students' proximity to schools decreases, and those who
+previously lived close enough to walk or cycle can no longer do so. In
+this paper, the City of Hamilton, a mid-sized city in Ontario, Canada,
+which experienced a 10\% decrease in elementary school locations between
+2011 and 2021, serves as a case study. We quanitfiy how these closures
+and consolidations have reshaped accessibility for the elementary
+school-aged population and discuss their implications for active and
+motorized travel. To measure accessibility, spatial availability, a
+singly-constrained multimodal measure of spatial accessibility is used.
+This method's proportional allocation mechanism enables interpretable
+comparison between accessibility `before' and `after' the implementation
+of school closure/consolidation policy considering changes in potential
+travel time and population. The analysis is conducted at the parcel
+level, considering all school seats as eligible and incorporating
+observed home-to-school travel patterns for motorized and non-motorized
+modes. We present system-wide impact estimates resulting from these
+closures. The findings reveal that overall spatial availability in the
+city declined. While the school consolidation policy may have achieved
+cost savings in facility operation and maintenance, the burden falls on
+students, families, and communities through \emph{potentially} reduced
+opportunities for physical activity through an active journey to school
+and increased motorized travel externalities.}
 
@@
 
-Communities are by nature in flux: economic and demographic shifts,
-rural and urban configurations, and political pressures all combine to
-produce change. These forces spur the re-organization and consolidation
-of public facilities in communities around the world where changes to
-service provision may improve the quality of service for some but not
-for all (Christiaanse 2020; Rosik, Puławska-Obiedowska, and Goliszek
-2021). While change is natural, oftentimes top-down consolidation
-decisions are driven by operational cost-savings that do not take into
-consideration the full extent of losses in benefits.
-
-From this perspective, school closure policies have been pursued by
+The amount of services accessible within communities oscillate: economic
+and demographic shifts, rural and urban configurations and political
+in/action all induce change. These forces spur pressures to
+reorganization and consolidation of public facilities that provide
+services, potentially improving access for some but not for all
+(Christiaanse 2020; Rosik, Puławska-Obiedowska, and Goliszek 2021).
+Schools are a compelling case study in this respect. In many communities
+globally, school closure and consolidation policies have been pursued by
 governments to optimize per-student operational cost-efficiency (Rong et
-al. 2022; Dai et al. 2019) but other implications, such as the impact of
-closures on property values (Merrall, Higgins, and Páez 2024), their
-effects on the social infrastructure of communities (Butler, Kane, and
-Cooligan 2019; Irwin and Seasons 2012), and particularly their travel
-implications have often been overlooked by decision makers (Bierbaum,
-Karner, and Barajas 2021; J. Lee and Lubienski 2017). This seems a
-mistake: journey-to-school trips are tremendously important, and travel
-for this purpose is currently the largest contributor of carbon
-emissions from personal travel after the journey-to-work (Rong et al.
-2022; Pantelaki, Claudia Caspani, and Maggi 2024), where it could be
-instead an excellent opportunity for physical activity (Desjardins et
-al. 2024). Alas, in many communities globally, schools are closing
+al. 2022; Dai et al. 2019). Research has highlighted that these polices
+have knock-on implications, including the weakening of communities'
+social infrastructure (Butler, Kane, and Cooligan 2019; Irwin and
+Seasons 2012; Autti and Hyry-Beihammer 2014) and negative
+household-level fiscal impacts such as a reduction of property values
+(Merrall, Higgins, and Páez 2024). Operationally, savings may be
+experienced but they are paid at different levels by students, families,
+communities and society.
+
+A school offers many things to a community: but one dimension is the
+travel to and from school, often overlooked by decision-makers
+considering school closure and consolidation policies (Bierbaum, Karner,
+and Barajas 2021; J. Lee and Lubienski 2017). Schools and residential
+locations often proximately co-located, the journey-to-school is an
+excellent opportunity for physical activity (Desjardins et al. 2024).
+However, closures exchange the physical activity of students' realized
+active trips or ``reserved'' active trips (Morency, Roorda, and Demers
+2009; Morency, Demers, and Lapierre 2007) (e.g., trips that could be
+made active) for motorized travel (Rong et al. 2022). Indeed, motorized
+travel to school is the largest contributor of carbon emissions from
+personal daily travel after the journey-to-work (Rong et al. 2022;
+Pantelaki, Claudia Caspani, and Maggi 2024). Despite these societal
+implications, many communities are closing and/or consolidating schools
 (Sageman 2022). For instance, 65\% of comprehensive schools (e.g.~grades
@@
 primary and lower secondary schools (kindergarten through grade 9) were
-closed in Denmark between 2010-2011 (Beuchert et al. 2018). School
-closures often increase the distance to school for many students and
-thus tend to induce motorized trips (Rong et al. 2022).
-
-We focus on the transportation implications of school closure and
-consolidation policies. While some research has considered the ``steps
-in reserve'', that is, the potential for active travel hidden behind
-motorized trips (Morency, Roorda, and Demers 2009; Morency, Demers, and
-Lapierre 2007), school closures potentially make actual steps vanish,
-sent to the reserve for the foreseeable future. In exchange for less
-physical activity, communities get more motorized travel, with the
-concomitant costs and burdens, including emissions. Quantification of
-losses in potential active travel and associated growth in carbon
-emissions is scarce in the literature, and the focus of this paper. In
-addition to its empirical contributions, this work makes use of a novel
-multimodal and singly-constrained accessibility measure, called spatial
-availability (Soukhov et al. 2023; Soukhov et al. 2024), which we use to
-estimate the changes in accessibility and travel after a recent wave of
-school closures/consolidations in Hamilton, Ontario.
-
-The results of the research indicate that, overall, school seats became
-less spatially available in the city. This change happened unevenly,
-with some communities being disproportionately affected. Though the
-school consolidation policy pursued by the province and implemented by
-the local school board may have resulted in cost-savings in operation
-and maintenance of facilities, families, students, and communities now
-pay the price in terms of lost opportunities for healthier active
-travel, as well as increased emissions from motorized travel.
+closed in Denmark between 2010-2011 (Beuchert et al. 2018).
+
+In this paper, we explore the case study of Hamilton, a mid-sized city
+in Canada that has pursued similar school closure and consolidation
+policies. We detail the Canadian political context that incentived these
+policies and quantify their transportation implications using data that
+constructs where students may have lived and the capacity of schools in
+2011, 2016 and 2021. To quantify transportation implications, we use a
+novel multimodal extension of a singly-constrained accessibility measure
+known as spatial availability (Soukhov et al. 2023; Soukhov et al.
+2024). Spatial availability is used to estimate active (i.e., walking)
+and motorized (i.e., car) accessibility considering the capacity of each
+school, and the residential location and quantity of school-aged
+population, and travel times from residential locations to schools for
+three time periods. In this way, this work contributes a empirical
+findings documenting how government austerity hurts public education,
+along with a methodology that results in interpretable accessibility
+values that can be compared across modes, the City, and three time
+periods.
+
+To foreshadow the paper's conclusions, our findings indicate that school
+seats have became less spatially available in the city. This change
+happened unevenly, with certain communities being disproportionately
+affected. The school closure and consolidation policy pursued may have
+achieved cost savings in facility operation and maintenance. However, it
+has come at a cost to families, students, and communities, resulting in
+lost opportunities for healthier active travel and a potential increase
+in motorized travel. We investigate: were the closures and
+consolidations worth the operational savings? The methods used in this
+paper can be applied by decision-makers to evaluate the spatial and
+transport-related impacts of public service provision policy.
+Furthermore, to increase the diffusion of this analysis, all work is
+transparent and reproducible, following best practices in the spatial
+sciences (Brunsdon and Comber 2021; Antonio Páez 2021). Accordingly, all
+associated code and data is publicly available within the lead author's
+\href{https://github.com/soukhova/School-closures-accessibility-impacts}{GitHub
+repository}.
 
 \section{Background}\label{background}
+
+\subsection{Overview of proximity and equity in the school accessibility
+literature}\label{overview-of-proximity-and-equity-in-the-school-accessibility-literature}
+
+The examination of access to public services is a well established area
+of the transportation literature: from parks (El-Murr et al. 2021) to
+healthcare (Pereira, Braga, et al. 2021) and associated dimensions of
+inequities (Tiznado-Aitken, Munoz, and Hurtubia 2021; Bittencourt and
+Giannotti 2023; Kelobonye et al. 2020), spatial accessibility has been
+explored in cases studies worldwide.
+
+Accessibility to schools has been a public service explored across
+several transportation dimensions. For instance, Marques, Wolf, and
+Feitosa (2020) demonstrates that lower-income neighbourhoods, compounded
+by other intersecting dimensions of disadvantage, tend to have lower
+accessibility to schools compared to more advantaged neighbourhoods in
+Portugal. Similar conclusions are drawn in West Virginia, US by Talen
+(2001), England by Burgess et al. (2011), Baton Rouge, US by Williams
+and Wang (2014), São Paulo and Curitiba, Brazil by Moreno-Monroy,
+Lovelace, and Ramos (2018), Pizzol, Giannotti, and Tomasiello (2021),
+and Bittencourt and Giannotti (2023), and Santiago, Chile by
+Tiznado-Aitken, Munoz, and Hurtubia (2021). Research has also
+demonstrated that when school closures occurs, what schools end up
+closing are not a politically neutral decision. For instance, these
+spatial trends are explored in the context of England by Pinch (1987),
+areas of declining school-aged populations in Dreseden, Germany by
+Müller (2011), rural Finnish communities by Autti and Hyry-Beihammer
+(2014), and Chicago, US by J. Lee and Lubienski (2017).
+
+The methods applied to measure school accessibility have been varied,
+from simple indicators to complex metrics. Beginning from more simple,
+Marques, Wolf, and Feitosa (2020) present indicators of school seats per
+student along with the shortest distance from origin to schools within a
+school catchment, methods also used in Talen (2001) and Burgess et al.
+(2011). Authors make refinements to these methods; Müller (2011)
+considers travel time by public transit from origins to schools and
+Pizzol, Giannotti, and Tomasiello (2021) considers travel times by
+different modes and multiple schools from a single origin along with
+other indicators related to school quality. By contrast, Williams and
+Wang (2014), J. Lee and Lubienski (2017), Moreno-Monroy, Lovelace, and
+Ramos (2018), Tiznado-Aitken, Munoz, and Hurtubia (2021), and
+Bittencourt and Giannotti (2023) use more complex approaches such as the
+two-step floating catchment area (2SFCA). The 2SFCA is popular in
+measure in evaluating access to healthcare services (Antonio Páez,
+Higgins, and Vivona 2019) and yields a supply-to-demand ratio that
+accounts for the population demanding a service and the quantity that is
+supplied within a certain distance range (Shen 1998; Luo and Wang 2003).
+The consideration of the supply of service is pertinent in the case of
+schools and is acknowledged to as a limitation to using simpler
+accessibility measures (Pizzol, Giannotti, and Tomasiello 2021).
+
+In addition to competition for school-seats, theoretically, the
+advantage students using faster modes (e.g., motorized modes) offer in
+terms of a wider range in school-seat choice is also important. Though
+some of the reviewed works using 2SFCA approaches consider multiple
+modes, explicit attention to the modal \emph{advantage} has not been
+discussed. To this end, in this work we use spatial availability, a
+singly-constrained measure of spatial accessibility (Soukhov et al.
+2023; Soukhov et al. 2024). Spatial availability allows for the
+representation of how many `school-seats' are spatially available to
+motorized- and active-trip taking elementary school-aged student
+population at a spatial-level, facilitating interpretable comparison
+across years. Further, spatial availability can be represented as
+school-seats per capita, mathematically equivalent to the
+supply-to-demand ratio of the 2SFCA (see appendix in Soukhov et al.
+(2023)).
+
+\subsection{Hamilton's school closures and
+consolidation}\label{hamiltons-school-closures-and-consolidation}
 
@@
 suburban and urban characteristics in the province of Ontario, Canada
-(Government of Canada 2022). In 2013, the Hamilton-Wentworth District
-School Board (HWDSB), the largest public English school board in the
-city, released its Long Term Facilities Master Plan (HWDSB 2013). To
-ensure ``equitable, affordable and sustainable learning facilities'',
-the Master Plan indicated that 80\% of its elementary schools (attended
-by students aged 5 to 13) would be subjected to accommodation
-evaluations, including an ``Accommodation Review'', the outcome of which
-historically has been school closures/consolidations (HWDSB 2013; Craggs
-2013; Seasons 2014b). These decisions, underhandedly supported by the
-province, were in part motivated by a search for operational savings
-based on projected reductions in student population and underutilized
-school capacity (Craggs 2012). Hamilton was particularly affected, as
-the HWDSB underwent an unprecedented number of Accommodation Reviews
-relative to other school boards (Seasons 2014a).
-
-Though the Accommodation Review guidelines outline a community-focused
-process to assess the value that a school provides to students, the
-community, and the local economy (Seasons 2014b; Ministry of Education
-2006), in practice residents felt that a wholesome breakdown of costs to
-repair/maintain schools was not made consistently available (Kleinhuis
-2013). Further, others pointed to flawed incentive structures motivating
-the HWDSB: ``By closing three schools the HWDSB would have a strong case
-with the Ministry of Education to receive full funding for a new
-school'' (Kleinhuis 2013). Along with insufficient funding to maintain
-schools in a state of good repair (Auditor General of Ontario 2015), the
-operational savings as a case for school closures is a long-standing
-critique of the school ``funding formula'' established by the
-conservative provincial government in 1998 (Mackenzie 2018; Irwin and
-Seasons 2012). Within a backdrop of budget cuts, the funding formula
-pits the financing of school maintenance and student needs against the
-financial compensation of staff and the powers of local school boards,
-which depend on the province for the allocation of resources (Mackenzie
-2018).
-
-In the wave of school accommodation reviews between 2013 and 2016, 12
-elementary schools closed in Hamilton. These closures represented a 7\%
-and 4\% decline in elementary school locations and capacity between 2011
-(before the school closures) and 2016, despite the elementary-aged
-student population in the city slightly increasing between 2011 and
-2016( from 58,265 students to 58,865 students).
-
-While accommodation reviews were undertaken to reduce operational costs,
-how might have the closure of schools impacted local communities? In
-this paper, we hypothesize that the school consolidation policy reduced
-accessibility to schools for many students. By extension, this loss in
-access and the consequent displacement of active trips to school
-increased transportation-related emissions. To test these hypotheses,
-the paper examines the following: how does the (1) spatial availability,
-(2) loss in walk minutes and (3) associated carbon emissions change for
-home-to-school trips after a city closes and consolidates schools?
-Further, this work examines (4) what communities, based on low-income
-prevalence, experienced the greatest losses as a result of this policy.
-
-Spatial availability, a singly-constrained measure of spatial
-accessibility (Soukhov et al. 2023; Soukhov et al. 2024), is used to
-quantify accessibility before- and after- school closures, and the
-associated impacts for the city of Hamilton and the elementary
-school-aged student population as a result of a policy with narrowly
-defined benefits. The methods used in this paper can be applied by
-decision-makers to evaluate the spatial and transport-related impacts of
-public service provision policy. Furthermore, to increase the diffusion
-of this analysis, all work is transparent and reproducible, following
-best practices in the spatial sciences (Brunsdon and Comber 2021;
-Antonio Páez 2021). Accordingly, all associated code and data will be
-publicly available within the lead author's
-\href{https://github.com/soukhova/School-closures-accessibility-impacts}{GitHub
-repository}
+(Government of Canada 2022a). The city's public English education is
+covered by the Hamilton-Wentworth District School Board (HWDSB). In
+2013, the HWDSB released its Long Term Facilities Master Plan (HWDSB
+2013). To ensure ``equitable, affordable and sustainable learning
+facilities'', the Master Plan indicated that 80\% of its elementary
+schools (attended by students aged approximately 5 to 14) would be
+subject to evaluations, including a ``Pupil Accommodation Review''
+(HWDSB 2013). This Master Plan and the slated Pupil Accommodation
+Reviews prompted public outcry, as the outcome of Pupil Accommodation
+Reviews are typically schools closures and consolidations (Craggs 2013;
+Seasons 2014b).
+
+Like all school boards in Ontario, HWDSB relies on funding from the
+province which is distributed through a formula based on pupil enrolment
+(Mackenzie 2018; Irwin and Seasons 2012). However, it has been
+demonstrated that this funding formula has become insufficient to
+maintain schools in a state of good repair (Auditor General of Ontario
+2015). Further, existing ``top-up'' programs to assist in operation and
+maintenance costs (e.g., School Facilities Operation and Renewal Grant)
+were entirely phased out by 2018 (TDSB 2024). Within this funding
+backdrop, the Pupil Accommodation Review guidelines outlined a
+``community-focused'' process to assessing the value that a school
+provides to students, the community, and the local economy (Seasons
+2014b; Ministry of Education 2006). However in practice, critical
+residents felt the outcomes of the Pupil Accommodation Reviews were
+predetermined (Thompson, Collins, and Dean 2024): older schools were
+slated to be closed, and a wholesome breakdown of costs to repair and
+maintain schools were not made available (Kleinhuis 2013). Other critics
+pointed to unbalanced incentives motivating the HWDSB: ``By closing
+three schools, the HWDSB would have a strong case with the Ministry of
+Education to receive full funding for a new school'' (Kleinhuis 2013).
+The Ministry's framework promotes the closing and consolidation of
+schools rather than cost-effective solutions for keeping older schools
+open (Kleinhuis 2013).
+
+The HWDSB and the Hamilton-Wentworth Catholic District School Board
+(HWCDSB), which together provide public English education to the
+majority of school-aged children in Hamilton (FAO 2023), were
+particularly impacted. While Accommodation Reviews were a province-wide
+trend during this period, Hamilton's school boards underwent an
+unprecedented number of reviews compared to other boards in other
+Ontario municipalities. (Seasons 2014a). In the wave of Pupil
+Accommodation Reviews between 2013 and 2016, 12 of the 147 elementary
+schools closed in Hamilton. Likely in part due to the public
+unpopularity of school closures, in 2017, the provincial government
+introduced a moratorium on beginning new Pupil Accommodation Reviews. In
+the following years, the HWDSB addressed its backlog and closed 12 more
+schools between 2017 to 2021 (HWDSB 2023). Though elementary school
+locations declined by 10\% between 2011-2021, HWDSB intends to conduct
+future Pupil Accommodation Reviews as initially scheduled, waiting until
+the end of the moratorium before dates are assigned (HWDSB 2023).
+
+A justification in these Pupil Accommodation Reviews are operational
+savings based on ``projected'' reductions in student population and
+``under-utilized'' school capacity (Craggs 2012; TDSB 2024; OPSBA 2024).
+As later demonstrated in this study, these policies in fact did result
+in school capacity per student reducing overall. Indeed, school seats in
+the city declined 5\%. It should be noted that though closures dominated
+the Pupil Accommodation Review process, some schools were
+``consolidated'' e.g., a school or two were closed and another school
+was expanded or opened to accommodate students from closed schools.
+Taken that school locations decreased more than overall school-seats,
+elementary schools in Hamilton are on average larger than they once
+were.
+
+In 2024, the moratorium on Pupil Accommodation Reviews is still in place
+and school boards are motivated by the same funding conditions. In
+absence of new revenue-generating levers and under the same funding
+formula with even lower inflation adjusted per-pupil funding, many
+school boards are calling for the end of the moratorium (OPSBA 2024;
+TDSB 2024; Draaisma 2024). In the meantime, some school boards are
+reducing specialized and community-wide programming (TDSB 2024). Under
+the status-quo, the end of the moratorium will result in further school
+closures. These pertinent decisions, made under an austere funding
+formula and flawed school board incentives does not create an
+environment where solutions for a sustainable public education system
+can thrive.
+
+Though schools have a plethora of benefits not captured by their
+maintenance and operational costs (Butler, Kane, and Cooligan 2019;
+Irwin and Seasons 2012; Autti and Hyry-Beihammer 2014), this paper's
+scope is limited to their travel implications. For students who live in
+proximity to schools, their closure or consolidation directly eliminate
+the realized or potential active trips that are, or could have been,
+made. Closing a school permanently reduces the potential for active
+trips and further entrenches reliance on motorized travel. Additionally,
+from the perspective of modal equity, school-aged populations with
+access to motorized transport gain a speed-time advantage, providing
+them with greater flexibility and choice in accessing school-seats.
+
+The spatial and population-level distribution of these impacts are of
+question for this paper. Specifically, the paper's aim is to demonstrate
+how these policies impacted the motorized and non-motorized (active)
+school-seat accessibility landscape. We hypothesize that the school
+closure and consolidation policies reduced accessibility to school-seats
+for many students. By extension, this loss in access and the
+displacement of active trips likely resulted in decreased potential
+active travel minutes and increased motorized travel minutes, with these
+impacts distributed inequitably across space and populations. To test
+these hypotheses, the paper examines the following:
+
+\begin{itemize}
+\tightlist
+\item
+  How has spatial availability changed for home-to-school trips
+  following school closures and consolidations?
+\item
+  What is the extent of the loss in walking minutes for these trips?
+\item
+  How has motorized travel increased as a result?
+\item
+  Which communities, particularly households with a high prevalence of
+  low-income and hence more likely experiencing transport poverty,
+  experienced the greatest losses due to this policy?
+\end{itemize}
 
@@
 Our focus is to evaluate the impact of school consolidations between
-2011 and 2016 on multimodal school spatial availability and
-travel-related GHG emission in Hamilton. To facilitate comparison, two
-scenarios are generated. Each scenario reflects the conditions in 2011
-and 2016 respectively, assuming the school configurations in each year,
-as well as average student population and low-income prevalence of
-households, residential parcel locations, and estimated modal travel
-times. The difference between the two scenarios is quantified in terms
-of accessibility, emissions and equity implications and discussed in
-context with the school consolidation policy. Data was retrieved at
-different levels of spatial granularity from the following sources.
+2011, 2016 and 2021 on multimodal school-seat spatial availability in
+Hamilton. To facilitate comparison, the following conditions are
+constructed for each year: 1) school configurations, 2) average student
+population, low-income prevalence of households, and residential parcel
+locations, and 3) estimated multimodal travel behaviour. The following
+sub-sections describe these three types of data.
 
@@
 
-First, schools and associated school catchments were retrieved from the
-City's school boards. In Hamilton, the majority of student-aged
-population attends a school in one of the two English public school
-boards (FAO 2023). One board is referred to as \emph{public} (i.e.,
-Hamilton Wentworth District School Board (HWDSB)) in our work and the
-other is referred to as \emph{public-catholic} (Hamilton Wentworth
-Catholic District School Board (HWCDSB)). These catchments indicate what
-residential property is assigned to what school, by default. Families
-can decide to attend schools out of catchment; in fact, 21\%-23\% of
-motorized school trips are out of catchment according to the regional
-travel survey, the Transportation Tomorrow Survey (TTS) in 2011 and 2016
-(Data Management Group 2018). The elementary school locations and
-catchments were provided by the HWDSB and HWCDSB for the 2010-2011 and
-2015-2016 academic years, and are shown in Figure~\ref{fig-Fig1}.
-
+First, elementary schools locations, how the changed throughout the
+study period and their capacities are required for each studied year.
+
+Concerning school locations, the City's school boards provided the
+authors with locations and associated school catchments for the
+2010-2011 and 2015-2016 academic years. Data for 2021 locations was
+retrieved from the City's open data portal (Hamilton 2024). In Hamilton,
+the majority of student-aged population attend school in one of the two
+English public school boards (FAO 2023). One school board is referred as
+Hamilton Wentworth District School Board (HWDSB) and the other as
+Hamilton Wentworth Catholic District School Board (HWCDSB). School
+catchments for the public and public-catholic catchments indicate what
+residential property is assigned to what school by default. However,
+families can decide to attend schools out of catchment; in fact,
+21\%-23\% of motorized school trips for populations aged 5 through 14
+are out of catchment according to the regional travel survey (the
+Transportation Tomorrow Survey (TTS)) in 2011 and 2016, the most recent
+TTS available at the time of writing (Data Management Group 2018).
+
+In addition to the school's location in 2011, 2016 and 2021, all schools
+were also tagged with a school status regarding their existence and/or
+change in location or school-seat capacity. A summary of the status of
+schools through 2011 to 2021 is illustrated in Figure~\ref{fig-Fig1}.
+Figure~\ref{fig-Fig1} also displays a count of the number of schools
+with the status between years, the number of schools overall present in
+each year, and the symbolic representation of a school status that is
+used in Figures throughout the paper. Overall, a school can have one of
+seven school statuses in any given year. Either: 1) the school did not
+change throughout 2011 to 2021, 2) the school expanded (i.e., increased
+school-seat capacity) or was 3) relocated and expanded (N.B., schools
+expanded only between 2011 through 2016 as they were typically done
+\emph{before} school closures and no new Pupil Accommodation Review was
+commenced after 2017), 4) the school closed sometime after 2011 and
+before 2016 or 5) after 2016 and before 2021, and 6) the school opened
+sometime after 2011 and before 2016 or 7) after 2016 and before 2021.
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=1.2\textwidth,height=\textheight]{../Figures/Fig1.png}
+
+}
+
+\caption{\label{fig-Fig1}Overview of the number of schools in Hamilton
+in 2011, 2016 and 2021 along with the type of school status change..}
+
+\end{figure}%
+
+The capacity of each school is also required for each simulated year.
+This capacity is reflected as the number of `seats' available in the
+school. Enrolment exceeding this capacity indicates over-enrolment,
+while enrolment below it indicates under-enrolment. This value is
+calculated by the provincial Ministry of Education and referred to as
+the on-the-ground capacity (OTGC). The current and historic OTGCs for
+schools are no longer publicly available (Ontario 2017), so we estimated
+them using a multivariate regression model and validated the results
+with a set of OTGC values manually gathered. These values were retrieved
+for schools that had undergone a completed Pupil Accommodation Review
+and had publicly available School Information Profiles.
+
+For schools lacking OTGC values, two regression models were created: one
+for public schools and another for public catholic schools. The
+independent variables in the model were 1) a dummy variable indicating
+the type of school grade instruction (\emph{MidElem}, \emph{JrElem},
+\emph{Elem}), 2) the school's building footprint (\emph{F}) in units of
+(\(m^2\)), and 3) the school's shortest Euclidean distance from the
+centroid of Hamilton's central business district area (\emph{DistCBD})
+defined in metres:
+
+\begin{itemize}
+\tightlist
+\item
   Formally, elementary schools are defined as schools that provide
-instruction to any combination of grades between kindergarten to grade 8
-(i.e.~typically children aged 5 to 14). As such, elementary includes
+  instruction to any combination of grades between kindergarten to grade
+  8 (i.e.~typically children aged 5 to 14). As such, elementary includes
   middle schools that only instruct grades 6 to 8 (\emph{MidElem}),
@@
   (\emph{JrElem}), and all grade elementary schools (\emph{Elem}) which
-instruct all grades from kindergarten to grade 8. The number of `seats'
-available in the school beyond or below which it will be over/under-
-capacity is calculated by the provincial Ministry of Education; this
-value is referred to as the on-the-ground capacity (\emph{OTGC}), and is
-assigned to each school. The current/historic OTGCs are not publicly
-available (Ontario 2017). However, for schools that underwent the full
-Accommodation Review process, OTGCs for a given year could be obtained
-from publicly available School Information Profiles. A regression model
-was used to estimate the OTGCs for schools that lacked OTGC information.
-The independent variables in the model were 1) the school's building
-footprint (\emph{F}) in units of (\(m^2\)); 2) the school's shortest
-Euclidean distance from the centroid of Hamilton's central business
-district area (\emph{DistCBD}) defined in units of (m); and 3)
-coefficients associated with a binary variable indicating the type of
-school grade instruction (\emph{MidElem}, \emph{JrElem}, \emph{Elem}).
-The building footprint from an archived spatial data set (Spatial 2015)
-and the 2016 footprints were retrieved from OSM (OpenStreetMap 2021).
-Because the age of construction for different schools is not available,
-we used each school's distance from the centroid of the Downtown
-Business Improvement Area (Hamilton 2014) as a proxy for school age as
-the oldest buildings are generally located closer to the city centre
-(Merrall 2021). Of note, the models were also used to estimate OTGC of
-secondary schools (\emph{Sec}) though not included in the scope of this
-work; these public institutions provide instruction for grades 9 to 12
-(i.e., children aged 14 to 17).
-
-Equations (\ref{eq:OTGC-public}) and (\ref{eq:OTGC-catholic}) describe
-the estimated OTGC for public schools and catholic schools respectively.
-School status and OTGC summed by catchment is also visualised in
-Figure~\ref{fig-Fig1}. As seen in Table \ref{TabA1-OTGC} in the
-Appendix, the coefficient of determination for these models is
-\(R^2 = 0.999\) and \(R^2 = 0.998\) for the Public \(OTGC_{Public}\) and
-Catholic \(OTGC_{Catholic}\) Boards respectively, and the residual
+  instruct all grades from kindergarten to grade 8. \emph{MidElem} and
+  \emph{JrElem} school grade instruction type is only present in the
+  public school board.
+\item
+  The building footprint \emph{F} was retrieved from an archived spatial
+  data set (Spatial 2015) and footprints from newer schools in 2016 and
+  2021 from OSM (OpenStreetMap 2021).
+\item
+  \emph{DistCBD} was calculated `as the crow flies' from each school to
+  the centroid of the Hamilton CBD (43.256684°N, 79.869039°W). Notably,
+  this variable can also be seen as a proxy for school construction age
+  as older buildings in Hamilton are generally located closer to the
+  CBD, the ``old'' Hamilton community (Merrall 2021).
+\end{itemize}
+
+Equations (\ref{eq:OTGC-public}) and (\ref{eq:OTGC-catholic}) summarise
+the OTGC regression model for public schools and public catholic schools
+respectively. As seen in Table \ref{TabA1-OTGC}, the coefficient of
+determination for these models are fantastic: \(R^2 = 0.999\) and
+\(R^2 = 0.998\) for the public \(OTGC_{Public}\) and public catholic
+\(OTGC_{Public Catholic}\) school boards respectively, and the residual
 standard deviation is \(\sigma = 0.212\) and \(sigma = 0.331\).
@@
 \label{eq:OTGC-catholic}
-OTGC_{Catholic} =F^{0.471}-e^{0.00003*DistCBD}+e^{2.333*Elem}
+OTGC_{Public Catholic} =F^{0.471}-e^{0.00003*DistCBD}+e^{2.333*Elem}
 \end{equation}
 
+\input{Tab1.tex}
+
+Taken together, Figure~\ref{fig-Fig2} displays the city's elementary
+schools locations, status and their estimated or retrieved OTGCs for
+2011, 2016 and 2021. For additional context on the degree of residential
+urbanization in Hamilton, the percentage of residential parcels
+considered `urban' compared to `suburban' or `rural' is reflected at an
+aggregated spatial unit across all six plots in Figure~\ref{fig-Fig2}.
+This land-use classification is available within the 2021 residential
+parcel file described in the following sub-section. In
+Figure~\ref{fig-Fig2}, it can be seen that the majority of schools that
+changed status are in the HWDSB, with OTGC capacity decreasing in 2016
+and 2021 compared to the previous year. Most school closures took place
+in the central and eastern parts of Hamilton's urban area (Hamilton
+Central), as well as in rural areas of western Hamilton (Flamborough).
+When schools did open or expand, it was primarily in the more recently
+urbanized southeastern area of Hamilton (Glanbrook), with a few also in
+Hamilton Central to offset some of the lost capacity.
+
 \begin{figure}
@@
 
-\includegraphics[width=1.2\textwidth,height=\textheight]{Manuscript_files/figure-pdf/fig-Fig1-1.pdf}
+\includegraphics[width=1.2\textwidth,height=\textheight]{Manuscript_files/figure-pdf/fig-Fig2-1.pdf}
 
@@
 
-\caption{\label{fig-Fig1}The school catchments, estimated On The Ground
-Capacity (OTGC) of schools summed per catchment, and school status for
-both public and public catholic school boards for year 2011 and 2016.
-OTGC scale is presented by quartiles.}
+\caption{\label{fig-Fig2}The on-the-ground capacity (OTGC) of schools
+and school status for year 2011, 2016 and 2021 for both the HWDSB and
+HWCDSB. Schools are presented overtop a layer visualising the degree of
+residential urbanisation in 2021.}
 
@@
 
-In addition to OTGC for schools, statistics about the student population
-and proportion of low-income after tax (LIM-AT) households were
-retrieved from the 2011 and 2016 Canadian Census (Canada 2011, 2016)
-using the \{cancensus\} \texttt{R} package (Bergmann, Shkolnik, and
+Secondly, a detailed account of the average student population,
+low-income prevalence of households, and where they may live is required
+for each studied year.
+
+Concerning the student population and poverty, information from the
+2011, 2016 and 2021 Canadian Census (Canada 2011, 2016, 2021) was
+sourced using the \{cancensus\} R package (Bergmann, Shkolnik, and
 Jacobs 2021). The census releases population data by age group category,
 so the population aged between 5-9 years and 10-14 years were retained.
-LIM-AT refers to the proportion of private households that are below the
-median after-tax income (Government of Canada 2017). The LIM-AT
-prevalence for households with children under 18 (there is no LIM-AT
+A common poverty measure across all three census years is the low-income
+after-tax measure (LIM-AT), which reflects the proportion of private
+households that are below the median after-tax income in the region
+(Government of Canada 2017). The LIM-AT prevalence for households with
+children under 18 was retrained for this paper as there is no LIM-AT
 prevalence tabulated for households with exclusively elementary aged
-children) was retrained. Population and LIM-AT variables were taken at
-the Dissemination Area (DA) level, which is the finest geographic unit
-publicly available. DAs are designed by Statistics Canada with the aim
-of population uniformity hence DAs greatly vary in area but represent
+children. Population and LIM-AT variables were taken at the
+Dissemination Area (DA) level, the finest geographic unit publicly
+available. DAs are designed by Statistics Canada with the aim of
+population uniformity hence DAs greatly vary in area but represent
 between approximately 400 and 700 (1st and 3rd quartile) in total
 population. The population aged 5 to 14 and the proportion of LIM-AT are
-visualised in the first and last rows in Figure~\ref{fig-Fig2}.
-
-The third source of data consists of centroids of residential parcels in
-2011 and 2016 that are used to represent residential locations (Teranet
-2009). There are 134,340 and 139,467 residential locations in 2011 and
-2016 respectively, representing 203,806 and 211,596 dwellings (according
-to the Canadian Census (Canada 2011, 2016)). As the number of children
-per parcel are publicly unavailable, the average number of 5 to 14 year
-olds for each DA is divided by the number of residential households in
-that DA. Then the potential number of children at each dwelling is
-calculated by multiplying this DA rate by the number of dwellings in
-each parcel. Due to the proprietary nature of the parcel data, the
-middle row in Figure~\ref{fig-Fig2} visualises an aggregation of the
-information: the average rate of 5-14 year olds per residential
-household at the DA level.
+visualised in the first and last rows in Figure~\ref{fig-Fig3}. LIM-AT
+prevalence is notably more concentrated within the centre of Hamilton
+(Hamilton Central) through 2011 to 2021, overlapping the most urbanised
+land-use and the largest amount of schools closed
+(Figure~\ref{fig-Fig2}). Also of note, LIM-AT prevalence drastically
+decreased in 2021 as a result of Pandemic benefits that reduced income
+inequality (Government of Canada 2022b).
+
+Where students live are estimated at the level of the residential
+parcel. Centroids for parcels in 2011, 2016 and 2021 are retrieved from
+Teranet (2009), representing 134,340, 139,467 and 143,890 unique
+locations in those years, respectively. The datasets do not provide
+information on the number of households or children per parcel. To
+estimate this value, the average number of students per parcel is
+calculated using the number of school-aged population in each DA is
+divided by the number of households in each DA. This rate is assigned to
+all parcels within an associated DA. Due to the proprietary nature of
+the parcel data, the second row in Figure~\ref{fig-Fig3} visualises an
+aggregation of the information: the average rate of 5-14 year old
+population per residential unit (parcel) at the DA level. Notably,
+though the rate of 5-14 year old population per parcel is higher in more
+peripheral (and rural) communities, there are many DAs within Hamilton
+Central that have high rates and populations that are similar to those
+in these more peripheral communities.
 
@@
 
-\includegraphics[width=1.2\textwidth,height=\textheight]{Manuscript_files/figure-pdf/fig-Fig2-1.pdf}
+\includegraphics[width=1.2\textwidth,height=\textheight]{Manuscript_files/figure-pdf/fig-Fig3-1.pdf}
 
@@
 
-\caption{\label{fig-Fig2}The magnitude (top row) and rate per
+\caption{\label{fig-Fig3}The magnitude (top row) and rate per
 residential unit (middle row) of elementary aged student population per
-DA in 2011 and 2016. Bottom row: the proportion of low-income households
-(prevelance of low-income after-tax, LIM-AT, measured by the 2011 and
-2016 Canadian Census) with dependents under 18 years old shown. All
-scales are represented in quartiles.}
+DA in 2011, 2016 and 2021. Bottom row: the proportion of low-income
+households (prevelance of low-income after-tax, LIM-AT, measured by the
+2011, 2016 and 2021 Canadian Census) with dependents under 18 years old
+shown. All scales are represented in quartiles.}
 
@@
 
-\subsection{Estimated travel times and
-emissions}\label{estimated-travel-times-and-emissions}
-
-It is assumed that children can go to any elementary school, but there
-is a preference for facilities that are more proximate. Based on this
-assumption, the fourth source of data is information regarding the
-mode-used and origin-destination locations of home-to-school trips from
-the 2011 and 2016 TTS (Data Management Group 2018). The TTS is a travel
-survey conducted in the Greater Golden Horseshoe Area in Ontario every
-five years. With a target 5\% sampling rate, the survey is expanded to
-be representative at the traffic analysis zone (TAZ) level of geography.
+\subsection{Multimodal travel
+behaviour}\label{multimodal-travel-behaviour}
+
+To reflect multimodal travel behaviour for all studied years, two types
+of data are estimated and complied. First, observed home-to-school trips
+by mode are retained from travel surveys. Next, travel times are
+calculated assuming active and motorized modes at parcel-to-school level
+and at the level matching the travel survey. Then, travel impedance
+functions are calibrated using the observed travel data, and using the
+functions and the parcel-to-school calculated travel times as input, the
+travel impedance values for each residential location to each school
+(origin-destination pair) is estimated. These three steps are described
+as follows:
+
+First, we retrieve information about the mode used and
+origin-destination locations of home-to-school trips from the 2011 and
+2016 Transportation Tomorrow Survey (TTS) (Data Management Group 2018);
+the 2021 TTS survey is unavailable at the time of writing so 2016 flows
+are assumed for the 2021 year. The TTS is a travel survey conducted in
+the Greater Golden Horseshoe Area in Ontario typically every five years.
+With a target 5\% sampling rate, the survey is expanded to be
+representative at the traffic analysis zone (TAZ) level of geography.
 TAZ are spatial units created for the purpose of the TTS and pulled from
@@
 this paper represent 13,715 and 12,878 motorized trips (mode used
-includes private car passenger, school bus, taxi, and transit) and 2,736
-and 1,430 non-motorized trips (mode used includes walk and cycling) for
-5 to 13 year olds from home-to-school in 2011 and 2016 respectively. The
-travel times for motorized and non-motorized travel are estimated using
-\{r5r\} (Pereira et al. 2021), which interfaces with the java-based R5
-routing engine developed separately by Conveyal (Conveyal {[}2015{]}
-2022). As travel times are not associated with the TTS, motorized and
-non-motorized travel times are estimated for all TAZ centroids to all
-TAZ centroids. We estimate motorized travel to be by car and
-non-motorized travel to be by walk as those are the most used modes in
-their respective categories. The travel times and the intensity of modal
-home-to-school flows are visualised in Figure~\ref{fig-Fig3} along with
-the boundaries of the TAZ and DAs. In Figure~\ref{fig-Fig3}, not all
-TAZs capture an elementary school trip by both modes. For this reason,
-the proportion of motorized modal share is aggregated at the community
-level. Hamilton (Ancaster: 85\%-95\%, Dundas: 89\%-100\%, Flamborough:
-91\%-97\%, Glanbrook: 92\%-95\%, Hamilton Central: 80\%-87\%, and Stoney
-Creek: 82\%-88\%).
+includes private car passenger, school bus, taxi, and transit) and 7,432
+and 7,085 non-motorized trips (mode used includes walk and cycling) for
+5 to 14 year olds from home-to-school in 2011 and 2016 respectively. The
+intensity of modal home-to-school flows are visualised in
+Figure~\ref{fig-Fig4} along with the boundaries of the TAZ and DAs. Also
+in Figure~\ref{fig-Fig4}, not all TAZs capture an elementary school trip
+by both modes. For this reason, the proportion of motorized modal share
+is aggregated at the community level. Hamilton (Ancaster - 2011: 78\%and
+2016:94\%, Dundas - 2011: 71\% and 2016: 67\%, Flamborough - 2011: 90\%
+and 2016: 91\%, Glanbrook - 2011: 91\% and 2016: 78\%, Hamilton Central
+- 2011: 55\% and 2016:50\%, and Stoney Creek - 2011: 71\% and 2016:
+79\%).
 
@@
 
-\includegraphics[width=1.2\textwidth,height=\textheight]{Manuscript_files/figure-pdf/fig-Fig3-1.pdf}
+\includegraphics[width=1.2\textwidth,height=\textheight]{Manuscript_files/figure-pdf/fig-Fig4-1.pdf}
 
@@
 
-\caption{\label{fig-Fig3}The origin-destination flows from home-based
-motorized and non-motorized school trips for students between 5-13 years
+\caption{\label{fig-Fig4}The origin-destination flows from home-based
+motorized and non-motorized school trips for students between 5-14 years
 old as retrieved from TTS 2011 and 2016. Flows are mapped atop the
@@
 
-The fifth source of data consists of estimated travel time matrices from
-all residential parcel locations to all elementary schools for both
-motorized and non-motorized travel. The travel times are also estimated
-using the \{r5r\} package (Pereira et al. 2021), with car mode and
-walking representing motorized and non-motorized travel respectively.
-For both motorized and non-motorized modes, the inputs assume a maximum
-travel time threshold of 60 minutes and the free-flow OpenStreetMap road
-network of Hamilton (retrieved using Geofabrik (Geofabrik 2022) and
-manually edited to only include the network within Hamilton). Further,
-only walking trips shorter than 27 min minutes were retained. This
-travel time is approximately equivalent to 1.6 km which is the point
-within-catchment students qualify for transport provided by the school
-board (HWDSB 2019). Estimated trip-weighted motorized travel times are
-on average 16.5 minutes in 2011 and 16.8 minutes in 2016 (summary
-statistics for both years are approximately; min: 0.5, Q1: 11.0, Q2:
-16.0, Q3: 21.0, max: 60.0). Estimated trip-weighted non-motorized travel
-times are on average 18.0 minutes in 2011 and 18.1 minutes in 2016
-(summary statistics for both years are approximately; min: 0.5, Q1:
-13.0, Q2: 19.0, Q3: 24.0, max: 27.0).
-
-These travel time matrices along with the origin-destination flows were
-used to create modal trip length distribution (TLD) functions. Analysis
-of the TLD is a useful technique to calibrate impedance functions as
-they are a representation of the propensity of travel by trip length
+Next, the travel times matrices for motorized and active travel are
+estimated using the `travel\_time\_matrix()' function part of the
+\{r5r\} package (Pereira, Saraiva, et al. 2021). Travel time matrices
+are estimated for all TAZ centroids to TAZ centroids (matrices of size
+for 2011 and 2016: 234 x 234) as they ar enot reported in the TTS.
+Additionally matrices are estimated for all residential parcels to
+schools (matrices of size 134,340 x 147 for 2011, 139,467 x 137 for
+2016, and 143,890 x 132 for 2021) for additional granularity. \{r5r\} is
+a R-based interface to the R5 routing engine developed separately by
+Conveyal (Conveyal {[}2015{]} 2022). For simplicity, we assume motorized
+travel is by car and non-motorized travel is by walking, as these are
+the most common modes in their respective categories. For travel time
+calculations, we set a maximum threshold of 60 minutes and use the
+free-flow OpenStreetMap road network of Hamilton (Geofabrik 2022). While
+all car trips are retained, walking trips over 27 minutes are filtered
+out. This value approximately corresponds to a distance of 1.6 km at a
+walking speed of 3.6 km/h, which is the threshold for qualifying for
+motorized transport provided by the school board (HWDSB 2019).
+
+In terms of the travel impedance function: we assume that children can
+go to any elementary school, however, there is a preference for
+facilities that are more proximate. Based on this assumption, we match
+the associated TAZ-to-TAZ travel times to all observed student-to-school
+travel flows from the TTS; these are visualised as grey bar columns in
+Figure~\ref{fig-Fig5}. Using these observed values, the
+\emph{theoretical} trip length distribution (TLD) functions are
+calibrated. TLDs can be interpreted as travel impedance functions as
+they represent the propensity of realized travel, by trip length
 (Horbachov and Svichynskyi 2018; Batista, Leclercq, and Geroliminis
-2019). The empirical TLD is used to fit a density distribution using
-maximum likelihood and moment matching techniques and the Nelder-Mead
-and Brent methods for direct optimization available within the
-\{fitdistrplus\} package (Delignette-Muller and Dutang 2015). Based on
-goodness-of-fit criteria and diagnostics, the gamma and exponential
-distributions were selected for the motorized and non-motorized modal
-distributions respectively. The gamma distribution is defined by the
-shape (\(\alpha\)) parameter of 1.939 (2011) and 2.046 (2016) and the
-rate (\(\beta\)) of 0.233 (2011) and 0.236 (2016). The exponential
-distribution is defined by the rate (\(\beta\)) parameter of 0.092
-(2011) and 0.1 (2016). The TLDs for the motorized and non-motorized
-trips are shown in Figure~\ref{fig-Fig4}. For reference, the gamma
-distribution and the exponential distribution function are displayed in
-Equations (\ref{eq:exp-dist}) and (\ref{eq:gamma-dist}) where \(x\) is
-\(c_{ij}\):
+2019). The TLDs are calibrated using the maximum likelihood and moment
+matching techniques and the Nelder-Mead and Brent methods for direct
+optimization available within the \{fitdistrplus\} R package
+(Delignette-Muller and Dutang 2015). The theoretical TLD for each mode
+and available study year is visualised in blue in Figure~\ref{fig-Fig5}.
+Based on goodness-of-fit criteria and diagnostics, the gamma and
+exponential distributions were selected for the motorized and
+non-motorized modal distributions respectively. The gamma distribution
+is defined by the shape (\(\alpha\)) parameter of 1.939 (2011) and 2.046
+(2016) and the rate (\(\beta\)) of 0.233 (2011) and 0.236 (2016). The
+exponential distribution is defined by the rate (\(\beta\)) parameter of
+0.092 (2011) and 0.1 (2016). For reference, the gamma distribution and
+the exponential distribution function are displayed in Equations
+(\ref{eq:exp-dist}) and (\ref{eq:gamma-dist}) where \(x\) is \(c_{ij}\):
 
@@
 
-\includegraphics[width=1\textwidth,height=\textheight]{Manuscript_files/figure-pdf/fig-Fig4-1.pdf}
+\includegraphics[width=1\textwidth,height=\textheight]{Manuscript_files/figure-pdf/fig-Fig5-1.pdf}
 
@@
 
-\caption{\label{fig-Fig4}Empirical (grey bars) and theoretical (blue
+\caption{\label{fig-Fig5}Observed (grey bars) and theoretical (blue
 curves) motorized and non-motorized impedance functions. Motorized
@@
 
-The sixth and final data source relates to motorized GHG estimations. It
-assumed that all non-motorized trips produce no emissions and those that
-are motorized take the average emission factor of 0.1 kg \(CO_{2}e\) per
-minute travel. This emission factor is assumed based on an average speed
-of 40 km/h, the 2,380 g \(CO_{2}e\) produced when 1 L of gasoline
-(well-to-wheel) is combusted, and an internal combustion engine
-passenger vehicle fuel efficiency of 7 g/100km as discussed in Soukhov
-and Mohamed (2022).
+Lastly, to achieve greater granularity, the TLDs calibrated using
+TAZ-to-TAZ flows are used to estimate travel impedance values for each
+parcel-to-school flow based on calculated parcel-to-school travel times.
+These values represent the likelihood that students in a parcel will
+travel to a school, informed by observed home-to-school flows from the
+TTS.
 
@@
 Accessibility is defined as the ``potential for spatial interaction''.
-It is classically presented as the gravity-based measure popularized by
-Hansen (1959). It takes the following general multimodal formulation
-(Equation (\ref{eq:multimodal-conventional-accessibility})):
+It is classically presented as the gravity-based measure defined in
+Hansen (1959) and takes the following general multimodal formulation:
 
@@
 
-As indicators of urban structure, Hansen-type measures are informative,
-but the meaning of 10,000 accessible school-seats is harder to pin down:
-how many opportunities must any single student have access to?
-Furthermore, in scenario comparisons, the interpretation of the
-percentage change in accessible school seats is unclear. The
+As indicators of urban structure, Hansen-type accessibility measures
+like Equation (\ref{eq:multimodal-conventional-accessibility}) are
+informative in reflecting the magnitude of access but meaning of the
+value itself is elusive. The significance of 10,000 accessible
+school-seats is hard to pin down: how many opportunities must any single
+student have access to? Furthermore, this opaque interpretation
+especially is complicated when comparing accessibility of school seats
+between years, such as different years as in this work. The
 interpretability of Hansen-type accessibility has been discussed in
@@
 Soukhov et al. (2023) and Soukhov et al. (2024). The interpretation of
-accessibility is suggested to depend on how many people demand the
-opportunity, especially for exclusive opportunity-types like
-schools-seats (i.e., one school-seat is for one student).
-
-In this paper, we benefit from new developments in accessibility
+accessibility depends on how many people demand for the opportunity,
+especially for exclusive opportunity-types like schools-seats (i.e., one
+school-seat is for one student).
+
+In this paper, our work benefits from new developments in accessibility
 research, particularly the multimodal spatial availability measure
@@
 availability \(V_{i}^m\) is defined as given by Equation
-(\ref{eq:spatial-availability}).
+(\ref{eq:spatial-availability}):
 
@@
 (weighted by the friction of distance). Also, as discussed in Soukhov et
-al. (2023), popular competitive accessibility measures such as the 2SFCA
-(Joseph and Bantock 1982; Weibull 1976; Shen 1998; Luo and Wang 2003)
-are internally inconsistent, and the only way it preserves the number of
-opportunities is if the effect of the impedance function is ignored when
-expanding the values of opportunities per capita to obtain the total
-number of opportunities. On the other hand, the proportional allocation
-procedure associated with calculating multimodal spatial availability
-\(V_i^m\) consistently returns a number of opportunities available to
-populations by mode that matches the total number of opportunities in
-the region when summed. By doing this consistently, it is possible to
-define a measure of multimodal spatial availability per capita as
-presented in Equation (\ref{eq:SA-per-capita}) for use as a benchmark to
-compare against the regional opportunities per capita
+al. (2023), popular competitive accessibility measures such as the
+two-step floating catchment area (2SFCA) (Joseph and Bantock 1982;
+Weibull 1976; Shen 1998; Luo and Wang 2003) are internally inconsistent,
+and the only way it preserves the number of opportunities is if the
+effect of the impedance function is ignored when expanding the values of
+opportunities per capita to obtain the total number of opportunities. On
+the other hand, the proportional allocation procedure associated with
+calculating multimodal spatial availability \(V_i^m\) consistently
+returns a number of opportunities available to populations by mode that
+matches the total number of opportunities in the region when summed. By
+doing this consistently, it is possible to define a measure of
+multimodal spatial availability per capita as presented in Equation
+(\ref{eq:SA-per-capita}) for use as a benchmark to compare against the
+regional opportunities per capita
 (\(\frac{\sum_{j} O_j}{\sum_{i} P_i}\)).
@@
 
-This paper uses the multimodal spatial availability \(V_i^m\) and
-spatial availability per student \(v_i^m\) measures to quantify
-accessibility changes for the following reasons.
-
-First, all residential parcels are associated to their respective census
-DA, TTS TAZ, and community boundary based on spatial location
-(aggregated by DA and shown as a per capita rate in the second row of
-Figure~\ref{fig-Fig2}). Each parcel is assigned a number of `potential'
-motorized and non-motorized student aged population (from DA) and modal
-share (from TTS) to calculate a motorized and non-motorized population
-balancing factor \(F_{i}^{pm}\). For each parcel, the following is
-retained: the 2011 and 2016 census average elementary-aged student
-population per parcel (Figure~\ref{fig-Fig2} second row) and the average
+To summarise the methodology and the data as described in Section 3,
+multimodal spatial availability \(V_i^m\) and spatial availability per
+student \(v_i^m\) is calculated for each parcel for each studied year as
+follows:
+
+\begin{itemize}
+\tightlist
+\item
+  First, all residential parcels are associated to their respective
+  census DA, TTS TAZ, and community boundary based on spatial location.
+  Each parcel is assigned a number of `potential' motorized and
+  non-motorized student aged population (from DA) and modal share (from
+  TTS) to calculate a motorized and non-motorized population balancing
+  factor \(F_{i}^{pm}\). For each parcel, the elementary-aged student
+  population per parcel (Figure~\ref{fig-Fig3} second row) and the
   motorized and non-motorized share as informed by the TTS aggregated by
-community (Figure~\ref{fig-Fig3}).
-
+  community (Figure~\ref{fig-Fig4}) is retained.
+\item
   Second, the school capacity \(O_j\) is estimated and associated with
-each school (Figure~\ref{fig-Fig1}). All residential locations are
+  each school (Figure~\ref{fig-Fig2}). All residential locations are
   assumed to be able to access all schools by non-motorized and/or
@@
   schools within a 27 minute non-motorized trip.
-
+\item
   Third, the motorized and non-motorized impedance-balancing factors
-\(F_{ij}^{cm}\) are calculated using a mode-specific impedance function
-(Figure~\ref{fig-Fig4}) based on the respective estimated travel times
-for each origin (parcel) to school combination (OD flows retained from
-the TTS). Using network estimated travel time sensitive to observed
-parcel origin to school destination flows conceptually addresses the
-aggregation error that could result from using less granular zonal units
-to represent origin/destinations (Kane and Kim 2020; Kwan 1998; Hewko,
-Smoyer-Tomic, and Hodgson 2002). Finally, outputs from all three stages
-are joined together. \(F_{i}^{pm}\) joined with \(F_{ij}^{cm}\) yields
-the total balancing factor \(F^{tm}_{ij}\) which serves to
-proportionally allocate the school capacity \(O_j\) to each parcel. The
-value is multimodal spatial availability \(V_i^m\) which can be
-interpreted as the number of school-seats that are available to the
-mode-using population at that parcel. \(V_i^m\) is then summed and
-represented at the DA and community level along with the calculated
-\(v_i^m\) for interpretation.
+  \(F_{ij}^{cm}\) are calculated using a mode-specific impedance
+  function (Figure~\ref{fig-Fig5}) based on the respective estimated
+  travel times for each origin (parcel) to school combination. Using
+  network estimated travel time sensitive to observed parcel origin to
+  school destination flows conceptually addresses the aggregation error
+  that could result from using less granular zonal units to represent
+  origin/destinations (Kane and Kim 2020; Kwan 1998; Hewko,
+  Smoyer-Tomic, and Hodgson 2002).
+\item
+  Finally, outputs from all three stages are joined together.
+  \(F_{i}^{pm}\) joined with \(F_{ij}^{cm}\) yields the total balancing
+  factor \(F^{tm}_{ij}\) which serves to proportionally allocate the
+  school capacity \(O_j\) to each parcel. The value is multimodal
+  spatial availability \(V_i^m\) which can be interpreted as the number
+  of school-seats that are available to the mode-using population at
+  that parcel. \(V_i^m\) is then summed and represented at the DA and
+  community level along with the calculated \(v_i^m\) for
+  interpretation.
+\end{itemize}
 
@@
 
-The motorized and non-motorized spatial availabilities of school seats
-\(V_i^m\) for both 2011 and 2016 are presented in the top half of
-Figure~\ref{fig-Fig5}. The sum of \(V_i^m\) values in a year for both
-motorized and non-motorized population equals the city-wide OTGC (across
-both public and catholic school boards) in that year. This is a result
-of \(V_i^m\)'s proportional allocation mechanism, singly-constraining
-opportunities through the proportion allocation balancing factors. In
-this way, \(V_i^m\) results can be interpreted as a proportion of the
-total OTGC in that year as the sum of \(V_i^m\) for both modes in a year
-is equal to the total OTGC in that year. In Figure~\ref{fig-Fig5}, \(i\)
-is a DA, so \(V_i^m\) is the sum of the school seat availability per
-parcel in each DA. Each parcel is assigned a potential student
-population based on the average rate of school-aged population per
-residential parcel in that DA and the travel impedance weighted travel
-time to all schools. The top four plots of Figure~\ref{fig-Fig5}
-demonstrate different magnitudes in part due to significantly lower
-non-motorized modal share (refer to Figure~\ref{fig-Fig3}), however it
-is apparent that both mode using populations have higher spatial
-availability when they are more proximate to schools. Hence, all DAs
-have higher values within Hamilton Central and those more proximate to
-schools in less-urban communities. Trends appear similar for both 2011
-and 2016: this aligns with intuition, proximate populations using
-motorized modes benefit from the ability to reach more distant
-destinations than non-motorized modes.
-
-When \(V_i^m\) is divided by the number of students by mode in that DA,
-\(v_i^m\) is derived (bottom four plots in Figure~\ref{fig-Fig5}). The
-\(v_i^m\) plots demonstrate different spatial distribution trends than
-\(V_i^m\) plots. If our aim is to provide sufficient availability to
-schools for both motorized and non-motorized populations (e.g., greater
-than 1.0 \(v_i^m\) school-seat availability per student), this rate is
-only achieved in the core of the city for motorized modes (greens and
-yellows). Notably, this rate is only available to non-motorized
-populations in DAs that are in less-densely populated rural areas that
-are school proximate and very few pockets of Hamilton Central. The
-majority of schools closed were within Hamilton Central, so
-non-motorized populations in DAs within Hamilton Central were the most
-negatively impacted by losses in availability while certain motorized
-populations captured these losses as availability gains.
+The probability density plots of the spatial availability values \(V_i\)
+and the per capita spatial availability \(v_i\) values for both modes
+and all three years are displayed in Figure~\ref{fig-Fig6} as a
+city-wide overview. The left column displays parcel-level values, and
+the right reflects parcel-level values aggregated at the 2021 DA grid.
+Both the parcel and DA level values reflect the spatial unit at which
+they were calculated or aggregated. However, the values of the \(V_i\)
+plots are not directly comparable since the x- and y- axes are set at
+different scale limits to highlight between year differences. However,
+overall trends in their distribution can be compared: in all four
+\(V_i\) plots, the 2011 data is more right-skewed than more current
+years, indicating that spatial availability values used to be higher
+(e.g., more school-seat access). School closures and consolidations had
+an impact despite the background dynamics of assumed student population,
+mode use, travel time and parcel location. The DA level \(V_i\) plots
+show the distribution of each DA value, which is the summation of all
+parcel \(V_i\) values within a DA. In 2011, DAs can be observed to also
+have more right-skewed spatial availability values than previous years.
+In contrast, the 2016 and 2021 distributions are more intense and
+heavily skewed toward lower values.
+
+Although \(V_i\) values cannot be directly compared, zonal per capita
+\(v_i\) values can be as the school-age population per spatial unit
+(e.g., parcel or DA) is used to normalize the respective \(V_i\) value.
+\(v_i\) values are presented in the bottom four plots of
+Figure~\ref{fig-Fig6}. It can be observed that the same 2011
+right-skewed trend persists across all modes. Motorized mode-using
+students have access to more spatially available school seats than
+non-motorized mode using students (in 2011, approximately 1.3
+school-seats per student as compared to 0.7 and in 2021 approximately
+1.2 compared to 0.5). Due to school-seats per student at the DA level
+being an interpretable aggregation, values at this level will be the
+discussion focus of the remainder of this paper.
 
@@
 
-\includegraphics[width=1.2\textwidth,height=\textheight]{Manuscript_files/figure-pdf/fig-Fig5-1.pdf}
+\includegraphics[width=1.2\textwidth,height=\textheight]{Manuscript_files/figure-pdf/fig-Fig6-1.pdf}
 
@@
 
-\caption{\label{fig-Fig5}Multimodal spatial availability (purples) and
+\caption{\label{fig-Fig6}Spatial availability (Vi) and per capita
+spatial availability (vi) probability density distributions for the city
+of Hamilton in 2011, 2016 and 2021. Aggregated at the parcel level (left
+column) and DA level (right column).}
+
+\end{figure}%
+
+Representing the right column of Figure~\ref{fig-Fig6} spatially,
+Figure~\ref{fig-Fig7} displays the spatial availability for each DA for
+a given mode-using population \(m\) (e.g., the sum of school seat
+availability per parcel in each DA for motorized or non-motorized
+students). Figure~\ref{fig-Fig7} visualizes the spatial distribution of
+motorized and non-motorized school seat availability for 2011, 2016, and
+2021. As a reminder, the sum of \(V_i^m\) values for both motorized and
+non-motorized populations in a given year equals the city-wide OTGC
+(i.e., school seats across both the public english HWDSB and public
+english catholic HWCDSB). This is due to \(V_i^m\)'s proportional
+allocation mechanism, singly-constraining opportunities through the
+proportion allocation balancing factors. In this way, \(V_i^m\) values
+can be interpreted as a proportion of the total OTGC for that year, as
+the sum of \(V_i^m\) for both modes in a year equals the total OTGC for
+that year.
+
+Referring to the first two rows of \(V_i\) plots in
+Figure~\ref{fig-Fig7} (purples), it is notable that the magnitudes of
+\(V_i\) values for the motorized and non-motorized populations are
+tremendously different. Again, each \(V_i\) value reflects the number of
+school seats spatially available to the mode-using population at that
+zonal level, so due to the lower non-motorized modal share (refer to
+Figure~\ref{fig-Fig4}), non-motorized values will be lower. However,
+both mode-using populations have higher spatial availability values
+(darker purples) in DAs that are more proximate to schools. Hence, all
+DAs have higher values within Hamilton Central and those more proximate
+to schools in less-urban communities. This trend persists across 2011,
+2016 and 2016, aligning with intuition: populations with shorter travel
+times to with relatively greater schools are calculated to have high
+school seat spatial availability. Motorized populations have shorter
+travel times at DAs further in distance from schools, while
+non-motorized populations are only competitive at proximate distances.
+
+The bottom two rows of \(v_i\) plots in Figure~\ref{fig-Fig7} (yellows
+to greens) demonstrate the spatial availability per mode-using student
+population e.g., \(V_i^m\) divided by the number of students aged 5-14
+using a motorized mode or non-motorized mode. It is notable that for
+motorized populations, their spatial availability per student is above 1
+(green) within the center of the city. Conversely, this rate is only
+available to non-motorized populations in DAs that are in less-densely
+populated rural areas that are school proximate and very few pockets of
+Hamilton Central in 2011 and 2016. In 2021, even fewer DAs within
+Hamilton Central have non-motorized spatial availability rates above 1.
+
+\begin{figure}[H]
+
+\centering{
+
+\includegraphics[width=1.2\textwidth,height=\textheight]{Manuscript_files/figure-pdf/fig-Fig7-1.pdf}
+
+}
+
+\caption{\label{fig-Fig7}Multimodal spatial availability (purples) and
 spatial availability per mode-using student (diverging reds and greens)
-aggregated at the DA level for 2011 and 2016. Scales are represented in
-quartiles.}
+aggregated at the DA level for 2011, 2016 and 2021. Scales are
+represented in deciles}
 
@@
 
-To highlight this point, Figure~\ref{fig-Fig6} visualises the change in
-school spatial availability between 2016 and 2011 for both mode where
-\(i\) are 2016 DAs. Though the 2011 and 2016 DAs system are very similar
-(see Figure~\ref{fig-Fig2} and Figure~\ref{fig-Fig3}), they are not
-identical: 16 DAs changed in their spatial extent (i.e., added, removed,
-changed boundaries). The proportional allocation mechanism of spatial
-availability provides flexibility in aggregating parcel-level values at
-whichever spatial unit is meaningful. For this case, the change in
-\(V_i^m\) and \(v_i^m\) represented is the subtraction between the value
-aggregated at 2016 DA for the 2016 and 2011 scenario.
+To further explore these changes, the difference in \(V_i\) (first two
+rows) and difference in \(v_i\) (second two rows) between 2011-2016,
+2016-2021 and overall between 2011-2021 are displayed in
+Figure~\ref{fig-Fig8} along with schools that changed during that time
+period. 2011-2016 and 2016-2021 add together to equal 2011-2021 changes
+quite clearly, e.g.~in the third row, the 2011-16 plots a small decrease
+within the core of the city, and 2016-2021 demonstrates a more
+significant decrease in the core and western peripheral communities.
+Together, 2011-2021 plot reflects the changes in these two plots.
+
+A few interesting spatial trends can be observed in
+Figure~\ref{fig-Fig8}. First, DAs in proximity to schools that closed
+(square points) see losses in \(V_i\) (reds) for both motorized and
+especially non-motorized populations. For non-motorized populations, all
+closed school resulted in a decrease in \(V_i\). The majority of school
+closures took place in the central and eastern parts of Hamilton's urban
+area (Hamilton Central), as well as in rural areas of western Hamilton
+(Flamborough). Hence, losses characterise those neighbourhoods.
+Secondly, gains in \(V_i\) (blues) are seen where schools are opened
+(diamond points) or expanded (triangle points), especially for motorized
+populations. These gains are most notably in the more recently urbanized
+southeastern area of Hamilton (Glanbrook), along with a few pockets of
+DAs in proximity to new or expanded schools in Hamilton Central.
+Thirdly, the concentration of students in proximity to schools and the
+extent of their OTGC matters from the perspecitve of modal advantage. In
+areas with fewer proximate school options and lower OTGC relative to the
+amount of students, a change in a school results in a big impact on
+spatial availability values. For instance, in the community of Ancaster,
+one school expanded in the rural area in this community between
+2011-2016. The motorized student population, especially in rural DAs,
+saw gains in spatial availability while non-motorized populations saw
+losses. Students in proximity to the schools in Ancaster have high
+non-motorized spatial availability, but when the school was expanded
+(and due to an under-served population) motorized populations saw an
+increase in spatial availability at the direct expense of non-motorized
+populations' spatial availability.
+
+Exploring changes in \(V_i\) provides a holistic view of spatial
+availability for a given DA. However, if the focus is on the ratio of
+spatial availability per student, the bottom two rows in
+Figure~\ref{fig-Fig8} offer more insight. For instance, a community may
+have had an initially average rate of school seats per student but is
+expecting an increase in student population so it expands a school.
+However, due to an increase in population, the spatial availability per
+student overall still decreases between these two time periods. This
+case appears to occur in Ancaster between 2011 to 2016: while motorized
+\(V_i\) increased, motorized \(v_i\) slightly decreased, and
+non-motorized \(v_i\) decreased relatively more than non-motorized
+\(V_i\). As shown in Figure~\ref{fig-Fig3}, the student population and
+the number of students per residential unit increased more rapidly than
+the expanded school could accommodate, leading to a decrease in \(v_i\)
+values in DAs proximate to expanded schools.
+
+Similarly, in the southeastern community of Glanbrook, student
+populations grew from 2011 to 2021, and new schools opened. However,
+\(v_i\) rates decreased because OTGC did not keep pace with student
+population growth. As another example, in Hamilton Central and the
+eastern community of Stoney Creek, changes in \(V_i\) are unpatterned,
+while \(v_i\) demonstrate more uniform decreases, with the pattern
+varying by modal-range for motorized versus non-motorized populations.
+In these communities, the student population remained relatively stable,
+though OTGC decreased overall. \(v_i\) is illuminating to consider when
+comparing changes in spatial availability rates.
 
@@
 
-\includegraphics[width=1.2\textwidth,height=\textheight]{Manuscript_files/figure-pdf/fig-Fig6-1.pdf}
+\includegraphics[width=1.2\textwidth,height=\textheight]{Manuscript_files/figure-pdf/fig-Fig8-1.pdf}
 
@@
 
-\caption{\label{fig-Fig6}Change in multimodal spatial availability and
-spatial availability per mode-using student between 2016 and 2011.
-Change is the subtraction of parcel-level results between 2016 and 2011
-scenarios summed for each DA in the 2016 Canadian Census DA system.
-Scales are represented in quartiles, with Q4 rounded to 0.}
+\caption{\label{fig-Fig8}Change in multimodal spatial availability and
+spatial availability per mode-using student from 2011 to 2016, 2016 to
+2021, and together from 2011 to 2021. Change is the subtraction of
+parcel-level results in that given year and summed for each DA in the
+2021 Canadian Census DA grid. Scales are represented in deciles.}
 
@@
 
-From Figure~\ref{fig-Fig6} (top plots), it is apparent that decreases
-(pinks and reds) in \(V_i^m\) appear across the city for both mode-using
-populations. As the majority of closed schools are within the core of
-the city: the more urban Hamilton Central and urban-areas of other
-communities see a concentration of this decreased spatial availability
-to school seats. Evidently, the number of OTGC in the city (equal to the
-sum of \(V_i^m\)) decreased between the two years, so some sort of
-decrease is to be expected. However, even when spatial availability is
-divided by the mode-using population (which also decreased between the
-years) \(v_i^m\) (bottom plots) continues to demonstrate decreases,
-especially those locally proximate to closed schools.
-
-This decrease follows different patterns depending on the mode.
-\(v_i^{non-motorized}\) is most impacted for parcels within a 27 minute
-walk travel time to schools that closed: these DAs demonstrate the most
-dramatic decrease. Smaller-area DAs (in Hamilton-Central) and
-larger-area DAs (in rural communities) proximate to closed schools
-present Q1 decreases (red). For \(v_i^{motorized}\), the range offered
-by motorized mode (estimated by shortest-path car travel times) is
-city-wide, so since the majority of schools were centrally located, the
-decrease is concentrated within the downtown core of Hamilton Central.
-Some motorized-mode using populations in DAs do experience an increase
-in \(V_i^m\) and \(v_i^m\): these DAs are often relatively proximate to
-schools but more immediately outside walking range. Under competitive
-and constrained calculations, these DAs benefit from the losses in
-spatial availability that are predominately experienced by the
-non-motorized populations within Hamilton Central.
-
-The decrease in school seat spatial availability can be further
-interpreted. How does potential non-motorized travel to school decrease?
-How does the increase in potential motorized travel minutes impact GHG
-emissions? Figure~\ref{fig-Fig7} visualises these related intermediates
-for one-way home-to-school travel in a given day. The top plot of
-Figure~\ref{fig-Fig7} demonstrates the sum of the change in
-non-motorized minutes per 2016 DA. These travel time minutes are the
-difference between the aggregated 2016 and 2011 scenarios for
-non-motorized travel times multiplied by the proportion of
-walking/cycling mode-use and rate of student population associated with
-the parcels in each 2016 DA. It can be seen that Q1 decreases are
-proximate to closed schools. Smaller decreases (and even a few
-increases) are apparent in a range further from closed schools,
-especially those also in proximity to the few opened/expanded schools.
-Intuitively, these trends mirror what is shown within the
-\(V^{non-motorized}\) landscape in Figure~\ref{fig-Fig6} (Pearson
-correlation coefficient of 0.796) as travel time is an input into the
-cost balancing factor \(F^{cm}_{ij}\). Similarly, the bottom plot
-(Figure~\ref{fig-Fig7}) visualises the change in estimated emissions
-from motorized travel minutes. The change in GHG emissions is also
-highly correlated with \(V^{motorized}\) in Figure~\ref{fig-Fig6}
-(Pearson correlation coefficient of 0.838) as it is also a product of
-travel time.
-
-\begin{figure}[H]
-
-\centering{
-
-\includegraphics[width=0.9\textwidth,height=\textheight]{Manuscript_files/figure-pdf/fig-Fig7-1.pdf}
-
-}
-
-\caption{\label{fig-Fig7}Change in active travel time minutes (top plot)
-and home-to-school GHG emissions (bottom plot) between 2011 and 2016
-scenarios. Change is the subtraction of parcel-level results between
-2016 and 2011 scenarios summed for each DA in the 2016 Canadian Census
-DA system. Scales are represented in quartiles, with Q4 rounded to 0.}
-
-\end{figure}%
-
-Notably, the mode type of spatial availability loss is important. The
-change in active travel minutes and GHG emissions highlights the
-negative externalities of \emph{decreased} non-motorized spatial
-availability. Normatively, our cities should be increasing spatial
-availability for non-motorized mode users given the climate crisis. The
-wave of school closures in Hamilton increased motorized spatial
-availability and, hence, potential GHG emissions. The policy drastically
-reduced non-motorized spatial availability, and this is seen in the
-decrease in potential non-motorized travel minutes as well as increased
-GHG emissions in rural areas that did benefit from motorized spatial
-availability gains. The proportion of non-motorized mode use is low in
-both 2011 and 2016, however, the closure of schools that are more
-proximate to student-population density eliminates the potential of
-those trips ever becoming active. And who is most impacted by the
-reduction of non-motorized spatial availability? A summary of spatial
-availability per Hamilton community and associated dimensions in 2016,
-along with the percentage change between 2011 and 2016, is summarized in
-Table~\ref{tbl-Tab1}.
-
-\pagebreak
+Changes in multimodal spatial availability is important: A summary of
+spatial availability per Hamilton community and associated dimensions in
+2021, along with the percentage change between 2011 and 2021, is
+presented in Table~\ref{tbl-Tab1}.
 
@@
 
-\begin{longtable}[c]{|p{0.47in}|p{0.40in}|p{0.29in}|p{0.38in}|p{0.34in}|p{0.40in}|p{0.29in}|p{0.38in}|p{0.29in}|p{0.40in}|p{0.29in}|p{0.38in}|p{0.29in}}
+\begin{longtable}[c]{|p{0.75in}|p{0.75in}|p{0.75in}|p{0.75in}|p{0.75in}|p{0.75in}|p{0.75in}|p{0.75in}|p{0.75in}|p{0.75in}|p{0.75in}|p{0.75in}|p{0.75in}}
 
 \caption{\label{tbl-Tab1}Spatial availability and associated dimensions
-aggregated by communities}
+aggregated by communities in 2011 and how much they changed by 2021}
 
@@
 
-\multicolumn{1}{>{\centering}p{\dimexpr 0.47in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{}}}} & \multicolumn{2}{>{\centering}p{\dimexpr 0.7in+2\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{Hamilton\ C.}}}} & \multicolumn{2}{>{\centering}p{\dimexpr 0.72in+2\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{Dundas}}}} & \multicolumn{2}{>{\centering}p{\dimexpr 0.7in+2\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{Stoney\ Creek}}}} & \multicolumn{2}{>{\centering}p{\dimexpr 0.67in+2\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{Ancaster}}}} & \multicolumn{2}{>{\centering}p{\dimexpr 0.7in+2\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{Flamborough}}}} & \multicolumn{2}{>{\centering}p{\dimexpr 0.67in+2\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{Glanbrook}}}} \\
-
-\multicolumn{1}{>{\centering}p{\dimexpr 0.47in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{}}}} & \multicolumn{1}{>{\centering}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{}}}} & \multicolumn{1}{>{\centering}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{\%\ Δ}}}} & \multicolumn{1}{>{\centering}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{}}}} & \multicolumn{1}{>{\centering}p{\dimexpr 0.34in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{\%\ Δ}}}} & \multicolumn{1}{>{\centering}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{}}}} & \multicolumn{1}{>{\centering}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{\%\ Δ}}}} & \multicolumn{1}{>{\centering}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{}}}} & \multicolumn{1}{>{\centering}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{\%\ Δ}}}} & \multicolumn{1}{>{\centering}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{}}}} & \multicolumn{1}{>{\centering}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{\%\ Δ}}}} & \multicolumn{1}{>{\centering}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{}}}} & \multicolumn{1}{>{\centering}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{\%\ Δ}}}} \\
-
-\multicolumn{1}{>{\raggedright}p{\dimexpr 0.47in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{LIM-AT}}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{27.0}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{14\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{12.5}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.34in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{10\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{11.6}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{10\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 8.9}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{24\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 7.6}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{4\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 7.5}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-14\%}}} \\
-
-\multicolumn{1}{>{\raggedright}p{\dimexpr 0.47in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{Pop.}}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{33437.7}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-4\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 2365.3}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.34in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-8\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 8070.3}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{6\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 5387.5}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{12\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 5172.4}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-4\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 4150.0}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{58\%}}} \\
-
-\multicolumn{1}{>{\raggedright}p{\dimexpr 0.47in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{OTGC}}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{36878.3}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-8\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 2377.0}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.34in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{0\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 9387.8}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{4\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 3404.7}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{0\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 4638.2}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{2\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 1601.8}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{22\%}}} \\
-
-\multicolumn{1}{>{\centering}p{\dimexpr 0.47in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{}}}} & \multicolumn{12}{>{\centering}p{\dimexpr 4.15in+22\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{Motorized}}}} \\
-
-\multicolumn{1}{>{\raggedright}p{\dimexpr 0.47in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{V\ (mt)}}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{39626.9}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-8\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 1817.7}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.34in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-8\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 6656.8}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{6\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 4338.0}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{8\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 3290.3}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-2\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 2371.7}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{58\%}}} \\
-
-\multicolumn{1}{>{\raggedright}p{\dimexpr 0.47in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{v\ (mt)}}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{1.2}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-4\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{0.8}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.34in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{0\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{0.8}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{0\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{0.8}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-4\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{0.6}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{2\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{0.6}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{0\%}}} \\
-
-\multicolumn{1}{>{\raggedright}p{\dimexpr 0.47in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{GHG}}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{47788.1}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-2\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 4851.7}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.34in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-8\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{15183.3}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{8\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 9470.9}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{14\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{12839.1}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-4\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 8798.4}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{60\%}}} \\
-
-\multicolumn{1}{>{\raggedright}p{\dimexpr 0.47in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{GHG/pop}}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{1.4}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{2\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{2.1}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.34in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{0\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{1.9}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{0\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{1.8}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{2\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{2.5}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{0\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{2.1}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{2\%}}} \\
-
-\multicolumn{1}{>{\centering}p{\dimexpr 0.47in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{}}}} & \multicolumn{12}{>{\centering}p{\dimexpr 4.15in+22\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{Non-motorized}}}} \\
-
-\multicolumn{1}{>{\raggedright}p{\dimexpr 0.47in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{V\ (nmt)}}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{152.7}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-44\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ \ 0.0}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.34in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-100\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 23.9}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-42\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ \ 4.0}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-68\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ \ 3.9}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-64\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ \ 1.9}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-2\%}}} \\
-
-\multicolumn{1}{>{\raggedright}p{\dimexpr 0.47in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{v\ (nmt)}}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{0.6}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-8\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{0.0}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.34in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-100\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{0.7}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-18\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{0.7}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-18\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{1.7}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-6\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{1.4}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-16\%}}} \\
-
-\multicolumn{1}{>{\raggedright}p{\dimexpr 0.47in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{tt}}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{4365.7}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-38\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ \ \ 0.0}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.34in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-100\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 585.3}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-28\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ 103.3}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-62\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ \ 40.4}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-62\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\ \ 22.5}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{12\%}}} \\
-
-\multicolumn{1}{>{\raggedright}p{\dimexpr 0.47in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{\textbf{tt/pop}}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{18.2}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{0\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.34in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{17.8}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{2\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{18.3}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{006400}{\fontsize{6}{6}\selectfont{2\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.4in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{17.9}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-2\%}}} & \multicolumn{1}{!{\color[HTML]{BEBEBE}\vrule width 1pt}>{\raggedright}p{\dimexpr 0.38in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{6}{6}\selectfont{16.4}}} & \multicolumn{1}{>{\raggedright}p{\dimexpr 0.29in+0\tabcolsep}}{\textcolor[HTML]{8B0000}{\fontsize{6}{6}\selectfont{-4\%}}} \\
+\hhline{>{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}-}
+
+\multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{\textbf{}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{\textbf{Hamilton\ C.}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{\textbf{\%\ Δ}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{\textbf{Dundas}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{\textbf{\%\ Δ}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{\textbf{Stoney\ Creek}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{\textbf{\%\ Δ}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{\textbf{Glanbrook}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{\textbf{\%\ Δ}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{\textbf{Flamborough}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{\textbf{\%\ Δ}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{\textbf{Ancaster}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{\textbf{\%\ Δ}}}}} \\
+
+\noalign{\global\arrayrulewidth 0pt}\arrayrulecolor[HTML]{000000}
+
+\hhline{>{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}-}\endfirsthead
+
+\hhline{>{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}-}
+
+\multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{\textbf{}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{\textbf{Hamilton\ C.}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{\textbf{\%\ Δ}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{\textbf{Dundas}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{\textbf{\%\ Δ}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{\textbf{Stoney\ Creek}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{\textbf{\%\ Δ}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{\textbf{Glanbrook}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{\textbf{\%\ Δ}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{\textbf{Flamborough}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{\textbf{\%\ Δ}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{\textbf{Ancaster}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{\textbf{\%\ Δ}}}}} \\
+
+\noalign{\global\arrayrulewidth 0pt}\arrayrulecolor[HTML]{000000}
+
+\hhline{>{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}-}\endhead
+
+\multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{LIM-AT}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{23.8516970}}}} & \multicolumn{1}{>{\cellcolor[HTML]{F4FBF3}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{13.2\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{11.2879110}}}} & \multicolumn{1}{>{\cellcolor[HTML]{F6FCF5}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{10.6\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{10.8019420}}}} & \multicolumn{1}{>{\cellcolor[HTML]{F7FCF7}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{9.1\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{8.3083330}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFFAF8}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-2.5\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{7.2350900}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FAFDFA}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{5.4\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{7.0989650}}}} & \multicolumn{1}{>{\cellcolor[HTML]{EAF7E9}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{24.9\%}}}} \\
+
+\noalign{\global\arrayrulewidth 0pt}\arrayrulecolor[HTML]{000000}
+
+\multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{Kid\ pop.}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{35310.5960000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFF9F7}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-3.2\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{2590.0000000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFE3D9}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-14.7\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{7514.7510000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{EEF9ED}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{20.0\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{2639.6530000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{A2DB9F}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{108.9\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{5365.0000000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{F8FCF7}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{8.7\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{4845.0000000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{F5FBF4}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{12.4\%}}}} \\
+
+\noalign{\global\arrayrulewidth 0pt}\arrayrulecolor[HTML]{000000}
+
+\multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{OTGC}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{39889.0820000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFECE6}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-9.8\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{2377.0380000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFFFFF}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{0.0\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{9010.1960000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFF1EB}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-7.6\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{1313.5220000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{91D58F}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{127.6\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{4590.7010000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFF1EB}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-7.5\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{3386.8900000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{F6FBF5}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{11.1\%}}}} \\
+
+\noalign{\global\arrayrulewidth 0pt}\arrayrulecolor[HTML]{000000}
+
+\hhline{>{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}-}
+
+\multicolumn{13}{>{\centering}m{\dimexpr 9.75in+24\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{\textbf{Motorized}}}}} \\
+
+\noalign{\global\arrayrulewidth 0pt}\arrayrulecolor[HTML]{000000}
+
+\hhline{>{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}-}
+
+\multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{V\ (mt)}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{42527.7500000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFE9E1}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-11.6\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{1952.2590000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFD4C5}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-22.5\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{6328.0100000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{F7FCF6}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{9.9\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{1531.3080000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{9DD99A}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{114.8\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{3382.5270000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFFBF9}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-2.3\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{4086.6160000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{F9FDF9}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{7.2\%}}}} \\
+
+\noalign{\global\arrayrulewidth 0pt}\arrayrulecolor[HTML]{000000}
+
+\multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{v\ (mt)}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{1.2393060}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFEEE8}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-8.9\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{0.7590198}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFEEE7}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-9.0\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{0.8369173}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFF1EC}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-7.1\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{0.5823797}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FCFEFC}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{3.1\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{0.6312921}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFECE5}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-10.1\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{0.8491653}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFF6F2}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-5.0\%}}}} \\
+
+\noalign{\global\arrayrulewidth 0pt}\arrayrulecolor[HTML]{000000}
+
+\hhline{>{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}-}
+
+\multicolumn{13}{>{\centering}m{\dimexpr 9.75in+24\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{\textbf{Non-motorized}}}}} \\
+
+\noalign{\global\arrayrulewidth 0pt}\arrayrulecolor[HTML]{000000}
+
+\hhline{>{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}-}
+
+\multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{V\ (nmt)}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{631.8817250}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFEAE2}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-11.0\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{28.7632500}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFF1EC}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-7.1\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{65.1592800}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFB39B}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-39.5\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{2.3855420}}}} & \multicolumn{1}{>{\cellcolor[HTML]{7CCD7C}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{1274.9\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{12.2516440}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFFDFC}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-1.0\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{18.5170610}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FF7957}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-67.6\%}}}} \\
+
+\noalign{\global\arrayrulewidth 0pt}\arrayrulecolor[HTML]{000000}
+
+\multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{v\ (nmt)}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{0.7054674}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFE6DD}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-13.0\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{1.5790158}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFDFD4}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-16.6\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{0.8967578}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFCFBE}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-25.3\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{1.6565846}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFD0BF}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-24.8\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{1.8555444}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFE8DF}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-12.2\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{0.8641681}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFDFD3}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-17.0\%}}}} \\
+
+\noalign{\global\arrayrulewidth 0pt}\arrayrulecolor[HTML]{000000}
+
+\hhline{>{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}-}
 
@@
 summarized by community (Table~\ref{tbl-Tab1}). The majority of schools
-were closed in Hamilton Central. This community represents 60\% of the
-5-14 year old population in 2016. Though the population between 2011 and
-2016 decreased, the rate of school-seat spatial availability decreased
-disproportionately more (population by -4\% and spatial availability by
--8\%). This disproportionate decrease is greatest in Hamilton Central
-than any other community. This may have equity concerns as Hamilton
-Central has the highest LIM-AT prevalence, with 27\% of households
-LIM-AT in 2016, 3.7 times greater than Glanbrook, a community with the
-lowest level of LIM-AT prevalence in the city. Between 2016 and 2011,
-Hamilton Central had the lowest levels of emissions per student and saw
-the highest per student increase in 2016. Hamilton Center had the
-highest non-motorized trip potential, and it was reduced in part by the
-school closure policies.
-
-Conversely, communities with the lowest LIM-AT benefit the most, notably
-in gains to motorized spatial availability. Glanbrook, Flamborough and
-Ancaster have the lowest prevalence of LIM-AT of any Hamilton Community
-and together account for 30\% of the 5-14 year old population in 2016.
-These communities experienced a growth in student population (or only a
-small decrease) and proportionally benefited from an increase in
-motorized spatial availability.
-
-Across the city, those who commute by non-motorized mode see the most
-drastic lose in spatial availability. In all communities, spatial
-availability decreases at greater levels than population. Notably in
-Dundas (2nd highest LIM-AT), the only school in the community closed,
-leaving the population in the community with no other option but to use
-motorized modes. Glanbrook (lowest LIM-AT), which experienced a
-relatively sharp increase in student population along with non-motorized
-using population, suffered a lose in non-motorized spatial availability
-despite seeing the highest increase in motorized spatial availability.
-Overall, the gain in motorized spatial availability resulted in a
-further erosion of non-motorized spatial availability. This erosion is
-related to a lose in potential active home-to-school travel minutes
-(3309.3 fewer minutes than in 2011, decreasing the potential active
-travel time per (all mode using) student from 0.14 to 0.09 minutes), as
-well as an increase in potential GHG emissions city-wide (3416 kg
-\(CO_2e\) more emission than in 2011, increasing the kg \(CO_2e\) per
-child rate for home-to-school trips from 1.64 to 1.68).
-
-\section{Conclusion}\label{conclusion}
-
-In this paper, we estimate the change in multimodal spatial availability
-as a result of school consolidations/closures between the years of 2011
-and 2016 in an empirical case study of Hamilton, a mid-size city
-(\textasciitilde570,000 pop) in Ontario, Canada (Government of Canada
-2022). The spatial availability and spatial availability per student are
-calculated at the parcel level and aggregated at the Canadian Census DA
-spatial unit and at the six Hamilton communities for comparison. Schools
-closed and the number of school seats decreased so spatial availability
-declined overall: we discuss where, for whom, and with what potential
-impacts.
-
-Overall, the closure of 7\% of elementary schools in Hamilton between
-2011 and 2016 resulted in -4\% decline in school-seat availability
-city-wide (2,280 fewer school-seats), but specifically a -46\% decline
-in non-motorized spatial availability. Though non-motorized modal share
-was low in both 2011 and 2016, the \emph{potential} for active transport
+were closed in Hamilton Central as seen in the largest reduction in OTGC
+by percentage and magnitude. Hamilton Central represents 55\% of the
+5-14 year old population in 2021. Though the population decreased a
+modest amount from 2011 to 2021, the rate of school-seat spatial
+availability decreased disproportionately more (population by -3.2\% and
+spatial availability by -11.6\% for both modes). Schools and additional
+OTGC was not sufficiently expanded to provide the same levels of spatial
+availability as in 2011. This disproportionate decrease in spatial
+availability to students is greater in Hamilton Central by magnitude
+than any other community.
+
+From the perspective of equity, Hamilton Central has the highest LIM-AT
+prevalence, with 27\% of households LIM-AT in 2016, more than 3 times
+greater than Ancaster, Flamborough and Glanbrook, communities with the
+lowest level of LIM-AT prevalence. Is the disproportionate decrease in
+spatial availability within Hamilton Central fair? In some ways it is:
+communities with the lowest LIM-AT prevalence appear to benefit from
+gains in motorized spatial availability. Glanbrook, Flamborough and
+Ancaster together account for 27\% of the 5-14 year old population in
+2021. However, they access 19\% of the spatial availability of school
+seats. Hamilton Central currently captures 66\% of the spatial
+availability for it's 55.0\% of the population. With student aged
+population growing in other communities, OTGC should be expanded -
+potentially to levels that Hamilton had in 2011 or beyond (1.2
+school-seats per motorized mode using student and 0.7 school-seats per
+active mode using student). However, should those gains come at a loss
+to other communities, especially those with a significantly higher
+proportion of households who are lower-income, a strong determinant of
+transport poverty?
+
+Furthermore, while there are gains in motorized spatial availability in
+certain communities by some metrics, there are losses in non-motorized
+spatial availability per student from 2011 to 2021 in all communities.
+And the losses are drastic. Communities with high community averag
+spatial availability per students with values above 1 school-seat per
+student saw losses, as well as communities below this value.
+
+\section{Discussion and conclusion}\label{discussion-and-conclusion}
+
+In this paper, we compared how multimodal spatial availability changed
+after a wave of school closures and consolidations in Hamilton. To do
+so, we constructed the student-aged population, OTGC of schools and
+their locations, and their travel behaviour to generate spatial
+availability landscapes for the three study years. We aggregated the
+resulting values at the DA and community level and descriptively
+compared differences. We demonstrated that city-wide there are decreases
+in \(V_i\) for both mode-using populations: the majority of students in
+the city have access to fewer school-seats than they would have in 2011.
+Furthermore, as the majority of closed schools are within the core of
+the city: the more urban Hamilton Central and urban-areas of other
+communities saw the largest amount of this decrease. And at a
+local-scale, students in residences proximate to schools that closed
+also saw dramatic reduce in non-motorized spatial availability values.
+
+Evidently, the number of OTGC in the city decreased between 2011 and
+2021, so some sort of decrease was to be expected. However, should that
+decrease be felt hardest within neighbourhoods with the highest LIM-AT?
+By students who have the potential to travel actively to school?
+Normatively, our cities should be increasing spatial availability for
+non-motorized mode users given the benefits of active travel and the
+climate crisis. Hamilton's policies from 2011 through 2021 drastically
+reduced non-motorized spatial availability, and gains in motorized
+spatial availability especially in more rural areas. The proportion of
+non-motorized mode use is low in both 2011 and 2021, however, the
+closure of schools that are more proximate to student-population density
+eliminates the potential of those trips ever becoming active.
+
+Overall, the closure of 10\% of elementary schools in Hamilton between
+2011 and 2021 resulted in 5\% decline in school-seat availability
+city-wide ( 5\% fewer school-seats), but specifically a 90\% decline in
+non-motorized spatial availability. Though non-motorized modal share was
+low in both 2011 and 2016, the \emph{potential} for active transport
 trips was significantly eroded as schools within a 27 minute walk
@@
 school boards did not account for the mobility-related burdens
-\emph{certain} families are now saddled with.
-
-In addition to spatial availability impacts, associated mobility
-measures are assessed. City-wide, home-to-school non-motorized minutes
-(for a one-way trip) reduced by 3,309 and GHG emissions associated with
-motorized home-to-school travel increased by 3,416 kg \(CO_2e\). The
-decision to consolidate schools, increase the length of motorized travel
-and decrease the potential for non-motorized impacts will have daily
-impacts on students and their families (Mandic et al. 2022; Pabayo et
-al. 2012), the broader community (Pietrabissa 2023; Merrall, Higgins,
-and Páez 2024; Bittencourt and Giannotti 2023) and the environment
-(Pantelaki, Claudia Caspani, and Maggi 2024; Rong et al. 2022).
+\emph{certain} families are now saddled with. The decision to
+consolidate schools, likely has increased the length of motorized
+travel, though this would need to be further investigated. A decrease in
+non-motorized trips will have daily impacts on students and their
+families (Mandic et al. 2022; Pabayo et al. 2012), the broader community
+(Pietrabissa 2023; Merrall, Higgins, and Páez 2024; Bittencourt and
+Giannotti 2023) and the environment (Pantelaki, Claudia Caspani, and
+Maggi 2024; Rong et al. 2022).
 
@@
 these decisions resulted in both families and students paying a price
-through spatial availability, lost potential walk minutes, and increased
-GHG emissions.
+through multimodal spatial availability.
 
@@
 2021; Yumita et al. 2021) nor considers significant non-transportation
-factors (Palm and Farber 2020). Further, this study does not consider
+factors that impact fulsome school accessibility like quality
+(Bittencourt and Giannotti 2023). Further, this study does not consider
 transit, school-bus, or cycling modes which have implications on how the
@@
 
-\bibitem[\citeproctext]{ref-Autti_Hyry-Beihammer_2014}
+\bibitem[\citeproctext]{ref-AuttiHyryBeihammer2014}
 Autti, Outi, and Eeva Kaisa Hyry-Beihammer. 2014. {``School Closures in
@@
 
+\bibitem[\citeproctext]{ref-burgess2011parental}
+Burgess, Simon, Ellen Greaves, Anna Vignoles, and Deborah Wilson. 2011.
+{``Parental Choice of Primary School in England: What Types of School Do
+Different Types of Family Really Have Available to Them?''} \emph{Policy
+Studies} 32 (5): 531--47.
+
 \bibitem[\citeproctext]{ref-butlerClosureRideau2019a}
@@
 ---------. 2016. {``Census of Canda Census Profiles: Median After-Tax
+Household Income, 5 to 9 Years, 10 to 14 Years, 15 to 19 Years, for
+Census Metroplian Areas and Dissimination Areas.''}
+
+\bibitem[\citeproctext]{ref-Statistics_Canada_2021}
+---------. 2021. {``Census of Canda Census Profiles: Median After-Tax
 Household Income, 5 to 9 Years, 10 to 14 Years, 15 to 19 Years, for
@@
 
+\bibitem[\citeproctext]{ref-draaismaTDSBConsideringCuts2024}
+Draaisma, Muriel. 2024. {``{TDSB} Considering Cuts to Programming Amid
+\$20M Deficit and Many Trustees Are Concerned.''} \emph{{CBC} News},
+March.
+\url{https://www.cbc.ca/news/canada/toronto/toronto-district-school-board-finance-budget-enrolment-committee-proposed-cuts-1.7149195}.
+
+\bibitem[\citeproctext]{ref-elmurrWalkingAccessibilityParks2021}
+El-Murr, Karl, Arianne Robillard, Owen Waygood, and Geneviève Boisjoly.
+2021. {``Walking Accessibility to Parks: Considering Number of Parks,
+Surface Area and Type of Activities.''} \emph{Findings}, October.
+\url{https://doi.org/10.32866/001c.27479}.
+
 \bibitem[\citeproctext]{ref-faoOntarioSchoolBoards2023}
@@
 \bibitem[\citeproctext]{ref-governmentofcanadaProfileTableCensus2022}
----------. 2022. {``Profile Table, Census Profile, 2021 Census of
+---------. 2022a. {``Profile Table, Census Profile, 2021 Census of
 Population - Hamilton, City (c) {[}Census Subdivision{]}, Ontario.''}
@@
 
-\bibitem[\citeproctext]{ref-Hamilton_2014}
-Hamilton. 2014. {``Map 5: Downtown Hamilton Buisness Improvement
-Area.''} Hamilton.
-\url{https://www.hamilton.ca/sites/default/files/media/browser/2014-11-27/downtown-hamilton-bia-map.pdf}.
+\bibitem[\citeproctext]{ref-governmentofcanadaDailyPandemicBenefits2022}
+---------. 2022b. {``The Daily --- Pandemic Benefits Cushion Losses for
+Low Income Earners and Narrow Income Inequality -- After-Tax Income
+Grows Across Canada Except in Alberta and Newfoundland and Labrador.''}
+July 13, 2022.
+\url{https://www150.statcan.gc.ca/n1/daily-quotidien/220713/dq220713d-eng.htm}.
+
+\bibitem[\citeproctext]{ref-hamiltonEducationalInstitutions2024}
+Hamilton. 2024. {``Educational Institutions.''} Open Hamilton.
+\url{https://open.hamilton.ca/datasets/34794547fe3e46619b3ff92a574fd558_19/explore}.
 
@@
 
+\bibitem[\citeproctext]{ref-hwdsb2023LongTermFacilities2023}
+HWDSB. 2023. {``2023 Long-Term Facilities Master Plan - Section 1.7:
+Accommodation Strategy Schedule.''} Hamilton, Ontario.
+\url{https://www.hwdsb.on.ca/wp-content/uploads/2023/10/LTFMP-1.7-Accommodation-Strategy-Schedule-2023.pdf}.
+
 \bibitem[\citeproctext]{ref-irwinSchoolClosure2012}
@@
 
+\bibitem[\citeproctext]{ref-marquesAccessibilityPrimarySchools2020}
+Marques, João Lourenço, Jan Wolf, and Fillipe Feitosa. 2020.
+{``Accessibility to Primary Schools in Portugal: A Case of Spatial
+Inequity?''} \emph{Regional Science Policy \& Practice}, July,
+rsp3.12303. \url{https://doi.org/10.1111/rsp3.12303}.
+
 \bibitem[\citeproctext]{ref-merlin2017competition}
@@
 
+\bibitem[\citeproctext]{ref-morenomonroyPublicTransportSchool2018}
+Moreno-Monroy, Ana I., Robin Lovelace, and Frederico R. Ramos. 2018.
+{``Public Transport and School Location Impacts on Educational
+Inequalities: Insights from São Paulo.''} \emph{Journal of Transport
+Geography} 67 (February): 110--18.
+\url{https://doi.org/10.1016/j.jtrangeo.2017.08.012}.
+
+\bibitem[\citeproctext]{ref-mullerAssessmentSchoolClosures2011}
+Müller, Sven. 2011. {``Assessment of School Closures in Urban Areas by
+Simple Accessibility Measures.''} \emph{Erdkunde} 65 (4): 401--14.
+\url{https://doi.org/10.3112/erdkunde.2011.04.06}.
+
 \bibitem[\citeproctext]{ref-ontarioSchoolFacilityInventory2017}
@@
 OpenStreetMap. 2021.
+
+\bibitem[\citeproctext]{ref-opsbaRightSchoolsRight2024}
+OPSBA. 2024. {``The Right Schools in the Right Locations: {OPSBA} Asks
+Government to Lift Moratorium on School Reviews.''} 2024.
+\url{https://www.opsba.org/opsba_news/the-right-schools-in-the-right-locations-opsba-asks-government-to-lift-moratorium-on-school-reviews/}.
 
@@
 
+\bibitem[\citeproctext]{ref-paez2019}
+Páez, Antonio, Christopher D. Higgins, and Salvatore F. Vivona. 2019.
+{``Demand and Level of Service Inflation in Floating Catchment Area
+(FCA) Methods.''} Edited by Tayyab Ikram Shah. \emph{PLOS ONE} 14 (6):
+e0218773. \url{https://doi.org/10.1371/journal.pone.0218773}.
+
 \bibitem[\citeproctext]{ref-paez2012measuring}
@@
 25: 141--53. \url{https://doi.org/10.1016/j.jtrangeo.2012.03.016}.
-
-\bibitem[\citeproctext]{ref-palmRolePublicTransit2020}
-Palm, Matthew, and Steven Farber. 2020. {``The Role of Public Transit in
-School Choice and After-School Activity Participation Among Toronto High
-School Students.''} \emph{Travel Behaviour and Society} 19 (April):
-219--30. \url{https://doi.org/10.1016/j.tbs.2020.01.007}.
 
@@
 
+\bibitem[\citeproctext]{ref-pereiraGeographicAccessCOVID192021}
+Pereira, Rafael H. M., Carlos Kauê Vieira Braga, Luciana Mendes Servo,
+Bernardo Serra, Pedro Amaral, Nelson Gouveia, and Antonio Paez. 2021.
+{``Geographic Access to {COVID}-19 Healthcare in Brazil Using a Balanced
+Float Catchment Area Approach.''} \emph{Social Science \& Medicine} 273
+(March): 113773. \url{https://doi.org/10.1016/j.socscimed.2021.113773}.
+
 \bibitem[\citeproctext]{ref-Pereira2021r5r}
@@
 
+\bibitem[\citeproctext]{ref-pinchChangingGeographyPreschool1987}
+Pinch, S. 1987. {``The Changing Geography of Preschool Services in
+England Between 1977 and 1983.''} \emph{Environment and Planning C:
+Government and Policy} 5 (4): 469--80.
+\url{https://doi.org/10.1068/c050469}.
+
+\bibitem[\citeproctext]{ref-pizzolQualifyingAccessibilityEducation2021}
+Pizzol, Bruna, Mariana Giannotti, and Diego Bogado Tomasiello. 2021.
+{``Qualifying Accessibility to Education to Investigate Spatial
+Equity.''} \emph{Journal of Transport Geography} 96 (October): 103199.
+\url{https://doi.org/10.1016/j.jtrangeo.2021.103199}.
+
 \bibitem[\citeproctext]{ref-rongReviewResearchLowcarbon2022}
@@
 
-\bibitem[\citeproctext]{ref-Rosik_Pulawska-Obiedowska_Goliszek_2021}
+\bibitem[\citeproctext]{ref-Rosik_PulawskaObiedowska_Goliszek_2021}
 Rosik, Piotr, Sabina Puławska-Obiedowska, and Sławomir Goliszek. 2021.
@@
 
+\bibitem[\citeproctext]{ref-Talen_2001}
+Talen, Emily. 2001. {``School, Community, and Spatial Equity: An
+Empirical Investigation of Access to Elementary Schools in West
+Virginia.''} \emph{Annals of the Association of American Geographers} 91
+(3): 465--86. \url{https://doi.org/10.1111/0004-5608.00254}.
+
+\bibitem[\citeproctext]{ref-tdsbLiftingMinistryEducations2024}
+TDSB. 2024. {``Lifting the Ministry of Education's Moratorium on School
+Closures to Allow for Modernization Through Consolidation Monday April
+15, 2024.''} April 2024.
+\url{https://www.tdsb.on.ca/home/ctl/Details/mid/43824/itemid/280\#:~:text=At\%20a\%20recent\%20Board\%20meeting,under\%2Denrolled\%20schools\%20through\%20consolidation.}
+
 \bibitem[\citeproctext]{ref-Teranet_2009}
@@
 Economic Development.
+
+\bibitem[\citeproctext]{ref-thompsonItJustGalls2024}
+Thompson, Carise, Patricia A. Collins, and Jennifer Dean. 2024. {``{`It
+Just Galls Me, as a Taxpayer'}: Trust Implications of School Closure
+Decision-Making in Ontario.''} \emph{Administration \& Society},
+November, 00953997241296117.
+\url{https://doi.org/10.1177/00953997241296117}.
+
+\bibitem[\citeproctext]{ref-tiznadoaitkenPublicTransportAccessibility2021}
+Tiznado-Aitken, I, JC Munoz, and R Hurtubia. 2021. {``Public Transport
+Accessibility Accounting for Level of Service and Competition for Urban
+Opportunities: An Equity Analysis for Education in Santiago de Chile.''}
+\emph{{JOURNAL} {OF} {TRANSPORT} {GEOGRAPHY}} 90 (January).
+\url{https://doi.org/10.1016/j.jtrangeo.2020.102919}.
 
@@
 357--79. \url{https://doi.org/10.1016/0166-0462(76)90031-4}.
+
+\bibitem[\citeproctext]{ref-williamsDisparitiesAccessibilityPublic2014}
+Williams, Shaun, and Fahui Wang. 2014. {``Disparities in Accessibility
+of Public High Schools, in Metropolitan Baton Rouge, Louisiana
+1990--2010.''} \emph{Urban Geography} 35 (7): 1066--83.
+\url{https://doi.org/10.1080/02723638.2014.936668}.
 
@@
 
-\pagebreak
-
-\section{Appendix}\label{appendix}
-
-\input{TabA1.tex}
-
 \end{document}

### 2025-01-12 paper/Manuscript.qmd
Proofread revisions

@@
   freeze: auto
-
 author:
@@
   | Reducing motorized travel and increasing active travel are top priorities for many communities.
-However, these priorities are confronted with pressures to reduce the fiscal burden of public services, including schools.
-Public school boards have been strained to do more with less, and opting to close under-capacity schools to deliver educational services more cost-effectively.
-However, once a school is closed, students' proximity to schools decreases, and those who previously lived close enough to walk or cycle can no longer do so.
-In this paper, the City of Hamilton, a mid-sized city in Ontario, Canada, which experienced a 10% decrease in elementary school locations between 2011 and 2021, serves as a case study.
-We quanitfiy how these closures and consolidations have reshaped accessibility for the elementary school-aged population and discuss their implications for active and motorized travel.
-To measure accessibility, spatial availability, a singly-constrained multimodal measure of spatial accessibility is used.
-This method's proportional allocation mechanism enables interpretable comparison between accessibility 'before' and 'after' the implementation of school closure/consolidation policy considering changes in potential travel time and population.
-The analysis is conducted at the parcel level, considering all school seats as eligible and incorporating observed home-to-school travel patterns for motorized and non-motorized modes.
-We present system-wide impact estimates resulting from these closures.
-The findings reveal that overall spatial availability in the city declined.
-While the school consolidation policy may have achieved cost savings in facility operation and maintenance, the burden falls on students, families, and communities through _potentially_ reduced opportunities for physical activity through an active journey to school and increased motorized travel externalities.
+However, these priorities clash against many government's reluctance to invest in public services, including schools.
+Public school boards are increasingly asked to do more with less, a directive often accomplished by closing and consolidating schools.
+Alas, this policy tends to reduce the proximity of schools for many students, thus reducing the chances for active travel.
+In this paper we study the City of Hamilton, a mid-sized city in Ontario, Canada, which experienced numerous elementary schools closures between 2011 and 2021.
+Our analysis evaluates how school closures and consolidations reshaped accessibility for elementary school students by motorized and active travel.
+In this task we use spatial availability, a singly-constrained multimodal measure of spatial accessibility.
+This method's proportional allocation mechanisms allow us to make consistent comparisons in accessibility between time periods.
+Analysis is conducted at the parcel level, considering on-the-ground-capacity (OTGC) of schools and observed home-to-school trip lengths for motorized and active modes.
+Our findings reveal that overall spatial availability in the city declined, and while the school consolidation policy pursued by the province may have achieved cost savings in facility operation and maintenance, it imposes ongoing costs on students, families, and communities through reduced opportunities for physical activity and entrenched reliance on motorized travel.
 
@@
 ```{r, include=FALSE}
-load(paste0(here::here(),file="/data-prep/data-products/cen_data_2011_hamiltononly.Rdata"))
+load(paste0(here::here(),file="/data-prep/data-products/cen_data_2011_hamiltononly.RData"))
 cen_data_2011_hamiltononly <- cen_data_2011_1 |> filter(`Region Name` == "Hamilton") #NOTE: GEOUID 35250242, 35250133, 35250457 do not have any child population
 rm("cen_data_2011_1")
-load(paste0(here::here(),file="/data-prep/data-products/cen_data_2016_hamiltononly.Rdata"))
+load(paste0(here::here(),file="/data-prep/data-products/cen_data_2016_hamiltononly.RData"))
 cen_data_2016_hamiltononly <- cen_data_2016_1 |> filter(`Region Name` == "Hamilton") #NOTE: GEOUID 35250133, 35250398, 35250427 do not have any child population
 rm("cen_data_2016_1")
-load(paste0(here::here(),file="/data-prep/data-products/cen_data_2021_hamiltononly.Rdata"))
+load(paste0(here::here(),file="/data-prep/data-products/cen_data_2021_hamiltononly.RData"))
 cen_data_2021_hamiltononly <- cen_data_2021_1 
@@
 #adding $ URBAN to the 2021 census DA file... retrieved from parcel points. No other year of parcel points provide this information 
-load(paste0(here::here(),file="/data-prep/data-products/DA_2021_URBvsRUR.Rdata"))
+load(paste0(here::here(),file="/data-prep/data-products/DA_2021_URBvsRUR.RData"))
 cen_data_2021_hamiltononly <- cen_data_2021_hamiltononly |> left_join(DA_2021_URBvsRUR, by="GeoUID")
@@
 
-```{r}
+```{r cache=FALSE}
 community_car_perc_2011 <- avail_2011 |>
@@
 
-```{r, eval=FALSE}
+```{r, include=FALSE}
 #check to make sure that V_i is equal to the total capacity in Hamilton. It is!
@@
 #the tts origin-destination trips, by mode
-load(file = paste0(here::here(),"/data-prep/data-inputs/TTS/od_trips_perc_2011.Rdata"))
-load(file = paste0(here::here(),"/data-prep/data-inputs/TTS/od_trips_perc_2016.Rdata"))
+load(file = paste0(here::here(),"/data-prep/data-inputs/TTS/od_trips_perc_2011.RData"))
+load(file = paste0(here::here(),"/data-prep/data-inputs/TTS/od_trips_perc_2016.RData"))
 
@@
   geom_vline(data = q_3, mapping = aes(xintercept = x), colour = "grey60") +
-  geom_line(data = tld_theoretical, aes(x = t, y = f), colour = "blue", size=0.8, alpha=0.6) + 
+  geom_line(data = tld_theoretical, aes(x = t, y = f), colour = "blue", linewidth=0.8, alpha=0.6) + 
   xlab("Estimated travel time associated with TTS OD flows (minutes)") +
@@
 
-The amount of services accessible within communities oscillate: economic and demographic shifts, rural and urban configurations and political in/action all induce change.
-These forces spur pressures to reorganization and consolidation of public facilities that provide services, potentially improving access for some but not for all [@Christiaanse_2020; @Rosik_PulawskaObiedowska_Goliszek_2021].
-Schools are a compelling case study in this respect.
-In many communities globally, school closure and consolidation policies have been pursued by governments to optimize per-student operational cost-efficiency [@rongReviewResearchLowcarbon2022; @Dai_Wang_Zhang_Liao_Liu_2019].
+Public services accessible within communities tend to oscillate: economic and demographic shifts, rural and urban configurations and political in/action all induce change.
+These forces spur pressures to reorganize and consolidate public facilities, potentially improving access for some but not for all residents in a community [@Christiaanse_2020; @Rosik_PulawskaObiedowska_Goliszek_2021].
+Public schools are a compelling case study in this respect.
+Across the globe, school closure and consolidation policies have been pursued by governments to reduce the cost of running public education systems  [@rongReviewResearchLowcarbon2022; @Dai_Wang_Zhang_Liao_Liu_2019].
 Research has highlighted that these polices have knock-on implications, including the weakening of communities' social infrastructure [@butlerClosureRideau2019a; @irwinSchoolClosure2012; @AuttiHyryBeihammer2014] and negative household-level fiscal impacts such as a reduction of property values [@MerrallWhatsASchool2024].
-Operationally, savings may be experienced but they are paid at different levels by students, families, communities and society.
-
-A school offers many things to a community: but one dimension is the travel to and from school, often overlooked by decision-makers considering school closure and consolidation policies [@bierbaumMobilityJusticeLinking2021; @Lee_Lubienski_2017].
-Schools and residential locations often proximately co-located, the journey-to-school is an excellent opportunity for physical activity [@desjardinsFramingActive2024].
-However, closures exchange the physical activity of students' realized active trips or "reserved" active trips [@morencyStepsReserve2009; @morencyHowMany2007] (e.g., trips that could be made active) for motorized travel [@rongReviewResearchLowcarbon2022].
+Operationally, savings may be experienced by governments, but they are paid at different levels by students, families, communities, and ultimately society in general.
+
+A school offers many things to a community, and one relevant dimension is how it shapes travel.
+This dimension is often overlooked by decision-makers tasked with implementing school closure and consolidation policies [@bierbaumMobilityJusticeLinking2021; @Lee_Lubienski_2017].
+When schools and residential locations are co-located, the journey-to-school presents an excellent opportunity for physical activity [@desjardinsFramingActive2024].
+However, school closures exchange the opportunities for active trips by students [@morencyStepsReserve2009; @morencyHowMany2007] for the near certainty of motorized travel [@rongReviewResearchLowcarbon2022].
 Indeed, motorized travel to school is the largest contributor of carbon emissions from personal daily travel after the journey-to-work [@rongReviewResearchLowcarbon2022; @pantelakiImpactHomeschoolCommuting2024].
-Despite these societal implications, many communities are closing and/or consolidating schools [@sageman2022].
+Alas, these impacts are not often considered in policy when schools close and/or consolidate [@sageman2022], a relatively common occurrence globally.
 For instance, 65% of comprehensive schools (e.g.
@@
 
-In this paper, we explore the case study of Hamilton, a mid-sized city in Canada that has pursued similar school closure and consolidation policies.
-We detail the Canadian political context that incentived these policies and quantify their transportation implications using data that constructs where students may have lived and the capacity of schools in 2011, 2016 and 2021.
-To quantify transportation implications, we use a novel multimodal extension of a singly-constrained accessibility measure known as spatial availability [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024].
-Spatial availability is used to estimate active (i.e., walking) and motorized (i.e., car) accessibility considering the capacity of each school, and the residential location and quantity of school-aged population, and travel times from residential locations to schools for three time periods.
-In this way, this work contributes a empirical findings documenting how government austerity hurts public education, along with a methodology that results in interpretable accessibility values that can be compared across modes, the City, and three time periods.
-
-To foreshadow the paper's conclusions, our findings indicate that school seats have became less spatially available in the city.
+In this paper, we investigate the case of Hamilton, a mid-sized city in Canada that has been forced by the province of Ontario to pursue school closure and consolidation policies.
+We <!--detail the Canadian political context that incentivized these policies and--> quantify the transportation implications of these policies using the capacity of schools in 2011, 2016 and 2021, after several waves of school closures and consolidations.
+To this end, we use a novel multimodal extension of a singly-constrained accessibility measure known as spatial availability [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024].
+We use this measure to estimate active (i.e., walking) and motorized (i.e., car) accessibility while constraining the supply of the service to the capacity schools.
+In this way, this work furnishes evidence to document how government austerity hurts public goals (such as physical activity) and locks school districts and families into motorized patterns of travel for the foreseeable future.
+
+To foreshadow the paper's conclusions, our findings indicate that school seats have became less available in the city.
 This change happened unevenly, with certain communities being disproportionately affected.
-The school closure and consolidation policy pursued may have achieved cost savings in facility operation and maintenance.
-However, it has come at a cost to families, students, and communities, resulting in lost opportunities for healthier active travel and a potential increase in motorized travel.
-We investigate: were the closures and consolidations worth the operational savings?
-The methods used in this paper can be applied by decision-makers to evaluate the spatial and transport-related impacts of public service provision policy.
-Furthermore, to increase the diffusion of this analysis, all work is transparent and reproducible, following best practices in the spatial sciences [@brunsdonOpeningPractice2021; @paezOpenSpatial2021c].
-Accordingly, all associated code and data is publicly available within the lead author's [GitHub repository](https://github.com/soukhova/School-closures-accessibility-impacts).
+The school closure and consolidation policy pursued by the government may have achieved cost savings in facility operation and maintenance.
+However, it has come at a cost to families, students, and communities, resulting in lost opportunities for healthier life-styles and an inevitable increase in motorized travel [@hammerWhy2012; @roseKids2013; @buckWhy2019].
+To increase the diffusion of this analysis, all work is transparent and reproducible, following best practices in the spatial sciences [@brunsdonOpeningPractice2021; @paezOpenSpatial2021c].
+For this reason, all associated code and data are publicly available within the lead author's [GitHub repository](https://github.com/soukhova/School-closures-accessibility-impacts).
 
@@
 
-The examination of access to public services is a well established area of the transportation literature: from parks [@elmurrWalkingAccessibilityParks2021] to healthcare [@pereiraGeographicAccessCOVID192021] and associated dimensions of inequities [@tiznadoaitkenPublicTransportAccessibility2021; @bittencourtEvaluatingAccessibilityAvailability2023; @kelobonye2020measuring], spatial accessibility has been explored in cases studies worldwide.
-
-Accessibility to schools has been a public service explored across several transportation dimensions.
-For instance, @marquesAccessibilityPrimarySchools2020 demonstrates that lower-income neighbourhoods, compounded by other intersecting dimensions of disadvantage, tend to have lower accessibility to schools compared to more advantaged neighbourhoods in Portugal.
+The examination of accessibility to public services is well-established, including studies of parks [@reyesalking2014; @elmurrWalkingAccessibilityParks2021], healthcare [@pereiraGeographicAccessCOVID192021], and public transportation [@moniruzzamanAccessibility2012; @tiznadoaitkenPublicTransportAccessibility2021], among many other services [e.g., @bittencourtEvaluatingAccessibilityAvailability2023; @kelobonye2020measuring].
+
+Schools are a service that has been studied across several transportation dimensions.
+For instance, @marquesAccessibilityPrimarySchools2020 demonstrate that lower-income neighbourhoods in Portugal tend to have lower accessibility to schools compared to more advantaged neighbourhoods.
 Similar conclusions are drawn in West Virginia, US by @Talen_2001, England by @burgess2011parental, Baton Rouge, US by @williamsDisparitiesAccessibilityPublic2014, São Paulo and Curitiba, Brazil by @morenomonroyPublicTransportSchool2018, @pizzolQualifyingAccessibilityEducation2021, and @bittencourtEvaluatingAccessibilityAvailability2023, and Santiago, Chile by @tiznadoaitkenPublicTransportAccessibility2021.
-Research has also demonstrated that when school closures occurs, what schools end up closing are not a politically neutral decision.
-For instance, these spatial trends are explored in the context of England by @pinchChangingGeographyPreschool1987, areas of declining school-aged populations in Dreseden, Germany by @mullerAssessmentSchoolClosures2011, rural Finnish communities by @AuttiHyryBeihammer2014, and Chicago, US by @Lee_Lubienski_2017.
-
-The methods applied to measure school accessibility have been varied, from simple indicators to complex metrics.
-Beginning from more simple, @marquesAccessibilityPrimarySchools2020 present indicators of school seats per student along with the shortest distance from origin to schools within a school catchment, methods also used in @Talen_2001 and @burgess2011parental.
-Authors make refinements to these methods; @mullerAssessmentSchoolClosures2011 considers travel time by public transit from origins to schools and @pizzolQualifyingAccessibilityEducation2021 considers travel times by different modes and multiple schools from a single origin along with other indicators related to school quality.
+Research has also demonstrated that school closures are seldom politically neutral decisions.
+For instance, spatial trends were explored in the context of England by @pinchChangingGeographyPreschool1987, areas of declining school-aged populations in Dreseden, Germany by @mullerAssessmentSchoolClosures2011, rural Finnish communities by @AuttiHyryBeihammer2014, and Chicago, US by @Lee_Lubienski_2017.
+
+The methods applied to measure school accessibility have also been varied.
+For example, @marquesAccessibilityPrimarySchools2020 present indicators of school seats per student along with the shortest distance from origin to schools within a school catchment.
+Similar methods are also used in @Talen_2001 and @burgess2011parental.
+Several authors propose refinements to these methods; @mullerAssessmentSchoolClosures2011, for instance, considers travel time by public transit from origins to schools, whereas @pizzolQualifyingAccessibilityEducation2021 considers travel times by different modes and multiple schools from a single origin along with other indicators related to school quality.
 By contrast, @williamsDisparitiesAccessibilityPublic2014, @Lee_Lubienski_2017, @morenomonroyPublicTransportSchool2018, @tiznadoaitkenPublicTransportAccessibility2021, and @bittencourtEvaluatingAccessibilityAvailability2023 use more complex approaches such as the two-step floating catchment area (2SFCA).
@@
 
-In addition to competition for school-seats, theoretically, the advantage students using faster modes (e.g., motorized modes) offer in terms of a wider range in school-seat choice is also important.
-Though some of the reviewed works using 2SFCA approaches consider multiple modes, explicit attention to the modal _advantage_ has not been discussed.
+In addition to competition for school-seats, theoretically, the use of faster modes (e.g., motorized modes) offers an advantage in terms of a wider range in school-seat choice.
+Though some of the works reviewed consider multiple modes, explicit attention to the modal _advantage_ has not been discussed.
 To this end, in this work we use spatial availability, a singly-constrained measure of spatial accessibility [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024].
-Spatial availability allows for the representation of how many 'school-seats' are spatially available to motorized- and active-trip taking elementary school-aged student population at a spatial-level, facilitating interpretable comparison across years.
-Further, spatial availability can be represented as school-seats per capita, mathematically equivalent to the supply-to-demand ratio of the 2SFCA (see appendix in @soukhovIntroducingSpatialAvailability2023).
+Spatial availability allows for the representation of the number of 'school-seats' that are spatially available to students by mode of transportation (i.e., motorized and active travel), considering the school-aged student population by zone.
+Further, spatial availability can be represented as school-seats per capita, mathematically equivalent to the supply-to-demand ratio of the 2SFCA (see appendix in @soukhovIntroducingSpatialAvailability2023), which increases the interpretability of the analysis, and the comparability between years.
 
@@
 
-Hamilton is a mid-size city (\~570,000 pop) with rural, suburban and urban characteristics in the province of Ontario, Canada [@governmentofcanadaProfileTableCensus2022].
-The city's public English education is covered by the Hamilton-Wentworth District School Board (HWDSB).
-In 2013, the HWDSB released its Long Term Facilities Master Plan [@hwdsbLongTermFacilities2013].
-To ensure "equitable, affordable and sustainable learning facilities", the Master Plan indicated that 80% of its elementary schools (attended by students aged approximately 5 to 14) would be subject to evaluations, including a "Pupil Accommodation Review" [@hwdsbLongTermFacilities2013].
-This Master Plan and the slated Pupil Accommodation Reviews prompted public outcry, as the outcome of Pupil Accommodation Reviews are typically schools closures and consolidations [@craggsBoardEvaluatingFuture2013; @seasonsSchoolClosurePolicy2014].
-
-Like all school boards in Ontario, HWDSB relies on funding from the province which is distributed through a formula based on pupil enrolment [@mackenzieBlueprintFixOntario2018; @irwinSchoolClosure2012].
-However, it has been demonstrated that this funding formula has become insufficient to maintain schools in a state of good repair [@auditorgeneralofontarioAnnualReport20152015].
+Hamilton is a mid-size city (\~570,000 pop) in the province of Ontario, Canada, with rural, suburban and urban characteristics [@governmentofcanadaProfileTableCensus2022].
+The city's public English education system is managed by the Hamilton-Wentworth District School Board (HWDSB) and the Hamilton-Wentworth Catholic District School Board (HWCDSB).
+These two boards, like many in the province have faced fiscal pressures from the province.
+For example, in 2013, the HWDSB released its Long Term Facilities Master Plan [@hwdsbLongTermFacilities2013], which nominally meant to ensure "equitable, affordable and sustainable learning facilities".
+This plan indicated that 80% of its elementary schools (attended by students aged approximately 5 to 14) would be subject to evaluations, including a "Pupil Accommodation Review" [@hwdsbLongTermFacilities2013].
+The announcement promptly led to a public outcry, since the outcome of Pupil Accommodation Reviews historically translated into school closures and consolidations [@craggsBoardEvaluatingFuture2013; @seasonsSchoolClosurePolicy2014].
+
+School boards in Ontario rely on funding from the province which is distributed through a formula based on pupil enrollment [@mackenzieBlueprintFixOntario2018; @irwinSchoolClosure2012].
+However, this funding formula has demonstrably become insufficient to maintain schools in a state of good repair [@auditorgeneralofontarioAnnualReport20152015].
 Further, existing "top-up" programs to assist in operation and maintenance costs (e.g., School Facilities Operation and Renewal Grant) were entirely phased out by 2018 [@tdsbLiftingMinistryEducations2024].
-Within this funding backdrop, the Pupil Accommodation Review guidelines outlined a "community-focused" process to assessing the value that a school provides to students, the community, and the local economy [@seasonsSchoolClosurePolicy2014; @ministryofeducationPupilAccommodationReview2006].
-However in practice, critical residents felt the outcomes of the Pupil Accommodation Reviews were predetermined [@thompsonItJustGalls2024]: older schools were slated to be closed, and a wholesome breakdown of costs to repair and maintain schools were not made available [@kleinhuisSeriousConcernsSchool2013].
-Other critics pointed to unbalanced incentives motivating the HWDSB: "By closing three schools, the HWDSB would have a strong case with the Ministry of Education to receive full funding for a new school" [@kleinhuisSeriousConcernsSchool2013].
+Against this backdrop, the Pupil Accommodation Review guidelines outlined a "community-focused" process for assessing the value that a school provides to students, the community, and the local economy [@seasonsSchoolClosurePolicy2014; @ministryofeducationPupilAccommodationReview2006].
+However in practice, critical residents felt the outcomes of the Pupil Accommodation Reviews were predetermined [@thompsonItJustGalls2024]: older schools were slated to be closed, and a wholesome breakdown of costs to repair and maintain schools was not made available [@kleinhuisSeriousConcernsSchool2013].
+The process introduced unbalanced incentives for school boards: as a concerned resident pointed at the time, "[b]y closing three schools, the HWDSB would have a strong case with the Ministry of Education to receive full funding for a new school" [@kleinhuisSeriousConcernsSchool2013].
 The Ministry's framework promotes the closing and consolidation of schools rather than cost-effective solutions for keeping older schools open [@kleinhuisSeriousConcernsSchool2013].
 
-The HWDSB and the Hamilton-Wentworth Catholic District School Board (HWCDSB), which together provide public English education to the majority of school-aged children in Hamilton [@faoOntarioSchoolBoards2023], were particularly impacted.
-While Accommodation Reviews were a province-wide trend during this period, Hamilton's school boards underwent an unprecedented number of reviews compared to other boards in other Ontario municipalities.
+Hamilton's two school boards (HWDSB and HWCDSB), which together provide public English education to the majority of school-aged children in Hamilton [@faoOntarioSchoolBoards2023], were particularly impacted by fiscal pressures.
+While Accommodation Reviews were a province-wide trend during this period, Hamilton's school boards underwent an unprecedented number of reviews compared to boards in other Ontario municipalities.
 [@seasons2014ARCCatalogues2014].
 In the wave of Pupil Accommodation Reviews between 2013 and 2016, `r SCHOOLS %>% filter(Status == "Removed") %>% pull(SchoolID) %>% length()` of the `r SCHOOLS %>% filter(Year == "2011 and 2016" | Year == "2011") %>% pull(SchoolID) %>% length()` elementary schools closed in Hamilton.
-Likely in part due to the public unpopularity of school closures, in 2017, the provincial government introduced a moratorium on beginning new Pupil Accommodation Reviews.
-In the following years, the HWDSB addressed its backlog and closed `r SCHOOLS %>% filter(Status_2 == "Removed") %>% pull(SchoolID) %>% length()` more schools between 2017 to 2021 [@hwdsb2023LongTermFacilities2023].
+Likely in part due to the unpopularity of school closures, in 2017, the provincial government introduced a moratorium, which paused any scheduled Pupil Accommodation Reviews.
+This created a backlog of Accommodation Reviews that HWDSB caught up to in subsequent years, when it closed `r SCHOOLS %>% filter(Status_2 == "Removed") %>% pull(SchoolID) %>% length()` more schools between 2017 to 2021 [@hwdsb2023LongTermFacilities2023].
 Though elementary school locations declined by `r -((SCHOOLS %>% filter(Year_2 != "2016") %>% pull(SchoolID) %>% length() - SCHOOLS %>% filter(Year != "2016") %>% pull(SchoolID) %>% length())/SCHOOLS %>% filter(Year != "2016") %>% pull(SchoolID) %>% length()) %>% scales::percent()` between 2011-2021, HWDSB intends to conduct future Pupil Accommodation Reviews as initially scheduled, waiting until the end of the moratorium before dates are assigned [@hwdsb2023LongTermFacilities2023].
 
-A justification in these Pupil Accommodation Reviews are operational savings based on "projected" reductions in student population and "under-utilized" school capacity [@craggsAxeCouldFall2012; @tdsbLiftingMinistryEducations2024; @opsbaRightSchoolsRight2024].
-As later demonstrated in this study, these policies in fact did result in school capacity per student reducing overall.
-Indeed, school seats in the city declined `r  -(( ((avail_2021 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2021)) |> select(OTGC) |> sum())) - (avail_2011 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2011)) |> select(OTGC) |> sum())) / (avail_2011 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2011)) |> select(OTGC) |> sum()) ) |> scales::percent()`.
-It should be noted that though closures dominated the Pupil Accommodation Review process, some schools were "consolidated" e.g., a school or two were closed and another school was expanded or opened to accommodate students from closed schools.
-Taken that school locations decreased more than overall school-seats, elementary schools in Hamilton are on average larger than they once were.
-
+A justification for these Pupil Accommodation Reviews are operational savings based on "projected" reductions in student population and "under-utilized" school capacity [@craggsAxeCouldFall2012; @tdsbLiftingMinistryEducations2024; @opsbaRightSchoolsRight2024].
+These policies led to overall reductions in school capacity per student: school seats in the city declined `r  -(( ((avail_2021 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2021)) |> select(OTGC) |> sum())) - (avail_2011 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2011)) |> select(OTGC) |> sum())) / (avail_2011 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2011)) |> select(OTGC) |> sum()) ) |> scales::percent()`--but the impacts were felt unevenly across the city.
+It should be noted that although closures dominated the Pupil Accommodation Review process, some schools were "consolidated", that is, a school or two were closed and another school was expanded or opened to accommodate students from closed schools.
+Since school locations decreased proportionally more than school-seats, elementary schools in Hamilton are on average larger than they once were.
+
+<!--
 In 2024, the moratorium on Pupil Accommodation Reviews is still in place and school boards are motivated by the same funding conditions.
-In absence of new revenue-generating levers and under the same funding formula with even lower inflation adjusted per-pupil funding, many school boards are calling for the end of the moratorium [@opsbaRightSchoolsRight2024; @tdsbLiftingMinistryEducations2024; @draaismaTDSBConsideringCuts2024].
-In the meantime, some school boards are reducing specialized and community-wide programming [@tdsbLiftingMinistryEducations2024].
-Under the status-quo, the end of the moratorium will result in further school closures.
-These pertinent decisions, made under an austere funding formula and flawed school board incentives does not create an environment where solutions for a sustainable public education system can thrive.
-
-Though schools have a plethora of benefits not captured by their maintenance and operational costs [@butlerClosureRideau2019a; @irwinSchoolClosure2012; @AuttiHyryBeihammer2014], this paper's scope is limited to their travel implications.
-For students who live in proximity to schools, their closure or consolidation directly eliminate the realized or potential active trips that are, or could have been, made.
+In absence of new revenue-generating levers and under the same funding formula with even lower inflation adjusted per-pupil funding, many school boards are calling for an end to the moratorium [@opsbaRightSchoolsRight2024; @tdsbLiftingMinistryEducations2024; @draaismaTDSBConsideringCuts2024].
+In the meantime, some school boards have reduced specialized and community-wide programming [@tdsbLiftingMinistryEducations2024].
+Under the status-quo, the end of the moratorium is likely to result in further school closures.
+These decisions, made under an austere funding formula and flawed school board incentives does not create an environment where solutions for a sustainable public education system can thrive.
+-->
+
+Existing research indicates that schools have a plethora of benefits not captured by their maintenance and operational costs [@butlerClosureRideau2019a; @irwinSchoolClosure2012; @AuttiHyryBeihammer2014], and in this paper we focus on the transportation implications.
+For students who live in proximity to schools, closures or consolidations directly eliminate the potential for active trips, and consequently physical activity.
 Closing a school permanently reduces the potential for active trips and further entrenches reliance on motorized travel.
@@
 The spatial and population-level distribution of these impacts are of question for this paper.
-Specifically, the paper's aim is to demonstrate how these policies impacted the motorized and non-motorized (active) school-seat accessibility landscape.
-We hypothesize that the school closure and consolidation policies reduced accessibility to school-seats for many students.
-By extension, this loss in access and the displacement of active trips likely resulted in decreased potential active travel minutes and increased motorized travel minutes, with these impacts distributed inequitably across space and populations.
+Specifically, the paper's aim is to investigate how these policies impacted the motorized and non-motorized (active) school-seat accessibility landscape.
+We hypothesize that school closure and consolidation policies reduced the availability of school-seats for many students.
+By extension, this loss in accessibility and the displacement of active trips likely resulted in decreased potential active travel minutes and increased motorized travel minutes, with these impacts distributed inequitably across space and populations.
 To test these hypotheses, the paper examines the following:
 
-- How has spatial availability changed for home-to-school trips following school closures and consolidations?
+- Did spatial availability change for home-to-school trips following school closures and consolidations?
 - What is the extent of the loss in walking minutes for these trips? 
@@
 
-Our focus is to evaluate the impact of school consolidations between 2011, 2016 and 2021 on multimodal school-seat spatial availability in Hamilton.
-To facilitate comparison, the following conditions are constructed for each year: 1) school configurations, 2) average student population, low-income prevalence of households, and residential parcel locations, and 3) estimated multimodal travel behaviour.
-The following sub-sections describe these three types of data.
+We focus on the evaluation of the impacts of school closures and consolidations between 2011, 2016 and 2021.
+The outcome of interest is the multimodal school-seat spatial availability in Hamilton.
+To research this, the following data were prepared for each year: 1) school configurations, 2) student population, prevalence of low-income households, and residential parcel locations, and 3) estimated trip length by mode.
+These data are described next.
 
@@
 
-First, elementary schools locations, how the changed throughout the study period and their capacities are required for each studied year.
+For this study, we use the locations of elementary schools and their capacities at each of three years: 2011, 2016, and 2021.
 
 Concerning school locations, the City's school boards provided the authors with locations and associated school catchments for the 2010-2011 and 2015-2016 academic years.
-Data for 2021 locations was retrieved from the City's open data portal [@hamiltonEducationalInstitutions2024].
+School locations in 2021 were retrieved from the City's open data portal [@hamiltonEducationalInstitutions2024].
 In Hamilton, the majority of student-aged population attend school in one of the two English public school boards [@faoOntarioSchoolBoards2023].
-One school board is referred as Hamilton Wentworth District School Board (HWDSB) and the other as Hamilton Wentworth Catholic District School Board (HWCDSB).
-School catchments for the public and public-catholic catchments indicate what residential property is assigned to what school by default.
-However, families can decide to attend schools out of catchment; in fact, 21%-23% <!--calculated in 03-travel-times-TTS.qmd--> of motorized school trips for populations aged 5 through 14 are out of catchment according to the regional travel survey (the Transportation Tomorrow Survey (TTS)) in 2011 and 2016, the most recent TTS available at the time of writing [@data_management_group_tts_2018].
-
-In addition to the school's location in 2011, 2016 and 2021, all schools were also tagged with a school status regarding their existence and/or change in location or school-seat capacity.
-A summary of the status of schools through 2011 to 2021 is illustrated in @fig-Fig1.
+School catchments for the public and public-catholic catchments indicate the default school for households.
+However, families can decide to attend schools out of catchment, and in fact, 21%-23% <!--calculated in 03-travel-times-TTS.qmd--> of motorized school trips for populations aged 5 through 14 are out of catchment according to the regional travel survey (the Transportation Tomorrow Survey, or TTS) in 2011 and 2016, the most recent TTS available at the time of writing [@data_management_group_tts_2018].
+
+In addition to school locations, all schools were also tagged with a school status regarding their operational status (active or closed) and/or changes in location or school-seat capacity.
+A summary of the status of schools through 2011 to 2021 is shown in @fig-Fig1.
 @fig-Fig1 also displays a count of the number of schools with the status between years, the number of schools overall present in each year, and the symbolic representation of a school status that is used in Figures throughout the paper.
@@
 
-The capacity of each school is also required for each simulated year.
-This capacity is reflected as the number of 'seats' available in the school.
-Enrolment exceeding this capacity indicates over-enrolment, while enrolment below it indicates under-enrolment.
-This value is calculated by the provincial Ministry of Education and referred to as the on-the-ground capacity (OTGC).
-The current and historic OTGCs for schools are no longer publicly available [@ontarioSchoolFacilityInventory2017], so we estimated them using a multivariate regression model and validated the results with a set of OTGC values manually gathered.
-These values were retrieved for schools that had undergone a completed Pupil Accommodation Review and had publicly available School Information Profiles.
-
-For schools lacking OTGC values, two regression models were created: one for public schools and another for public catholic schools.
-The independent variables in the model were 1) a dummy variable indicating the type of school grade instruction (*MidElem*, *JrElem*, *Elem*), 2) the school's building footprint (*F*) in units of ($m^2$), and 3) the school's shortest Euclidean distance from the centroid of Hamilton's central business district area (*DistCBD*) defined in metres:
-
-- Formally, elementary schools are defined as schools that provide instruction to any combination of grades between kindergarten to grade 8 (i.e.
+The capacity of each school is also required for each year.
+This capacity is the number of 'seats' available in the school.
+Capacity is calculated by the provincial Ministry of Education and referred to as the "on-the-ground capacity" (OTGC) of a school, albeit it is important to note that schools rarely operate at their nominal capacity, and can be either over-enrolled or under-enrolled.
+
+Current and historic OTGCs for schools are not publicly available [@ontarioSchoolFacilityInventory2017], so we used a set of known OTGC values gathered from Pupil Accommodation Review documents and multivariate regression to estimate and validate these values.
+Two regression models were created: one for public schools and another for public catholic schools.
+The independent variables in these models were 1) a dummy variable indicating the level of the school (*MidElem*, *JrElem*, *Elem*); 2) the school's building footprint (*F*) in $m^2$; and 3) the school's Euclidean distance from the centroid of Hamilton's central business district area (*DistCBD*) in metres.
+
+Formally, elementary schools are defined as schools that provide instruction to any combination of grades between kindergarten to grade 8 (i.e.
 typically children aged 5 to 14).
@@
 *MidElem* and *JrElem* school grade instruction type is only present in the public school board.
+
 - The building footprint *F* was retrieved from an archived spatial data set [@DMTI_Spatial_2015] and footprints from newer schools in 2016 and 2021 from OSM [@OpenStreetMap_2021].
-- *DistCBD* was calculated 'as the crow flies' from each school to the centroid of the Hamilton CBD (43.256684°N, 79.869039°W).
+
+- *DistCBD* was calculated 'as the crow flies' from each school to the centroid of the Hamilton CBD (43.256684$\circ$N, 79.869039$\circ$W).
 Notably, this variable can also be seen as a proxy for school construction age as older buildings in Hamilton are generally located closer to the CBD, the "old" Hamilton community [@Merrall_2021].
 
-Equations (\ref{eq:OTGC-public}) and (\ref{eq:OTGC-catholic}) summarise the OTGC regression model for public schools and public catholic schools respectively.
-As seen in Table \ref{TabA1-OTGC}, the coefficient of determination for these models are fantastic: $R^2 = 0.999$ and $R^2 = 0.998$ for the public $OTGC_{Public}$ and public catholic $OTGC_{Public Catholic}$ school boards respectively, and the residual standard deviation is $\sigma = 0.212$ and $sigma = 0.331$.
+Equations (\ref{eq:OTGC-public}) and (\ref{eq:OTGC-catholic}) are the OTGC regression model for public schools and public catholic schools respectively.
+As seen in Table \ref{TabA1-OTGC}, the coefficient of determination for these models are quite high, at $R^2 = 0.999$ and $R^2 = 0.998$ for the public and public catholic school boards respectively, and the residual standard deviation is $\sigma = 0.212$ and $sigma = 0.331$.
 
@@
 
-Taken together, @fig-Fig2 displays the city's elementary schools locations, status and their estimated or retrieved OTGCs for 2011, 2016 and 2021.
+Taken together, @fig-Fig2 displays the city's elementary schools locations, status, and their observed or estimated OTGCs for 2011, 2016 and 2021.
 For additional context on the degree of residential urbanization in Hamilton, the percentage of residential parcels considered 'urban' compared to 'suburban' or 'rural' is reflected at an aggregated spatial unit across all six plots in @fig-Fig2.
@@
 
-Where students live are estimated at the level of the residential parcel.
-Centroids for parcels in 2011, 2016 and 2021 are retrieved from @Teranet_2009, representing 134,340, 139,467 and 143,890 unique locations in those years, respectively.
-The datasets do not provide information on the number of households or children per parcel.
-To estimate this value, the average number of students per parcel is calculated using the number of school-aged population in each DA is divided by the number of households in each DA.
-This rate is assigned to all parcels within an associated DA.
-Due to the proprietary nature of the parcel data, the second row in @fig-Fig3 visualises an aggregation of the information: the average rate of 5-14 year old population per residential unit (parcel) at the DA level.
+Student residential locations were approximated at the level of residential parcels.
+Centroids for parcels in 2011, 2016 and 2021 were retrieved from @Teranet_2009, representing $134,340$, $139,467$ and $143,890$ unique locations, respectively.
+Each residential unit in a parcel was populated with the average number of children aged 5-14 in the corresponding dissemination area (DA).
+Due to the proprietary nature of the parcel data, the second row in @fig-Fig3 shows an aggregation of the information: the average rate of 5-14 year old population per residential unit (parcel) at the DA level.
 Notably, though the rate of 5-14 year old population per parcel is higher in more peripheral (and rural) communities, there are many DAs within Hamilton Central that have high rates and populations that are similar to those in these more peripheral communities.
@@
 
-## Multimodal travel behaviour
-
-To reflect multimodal travel behaviour for all studied years, two types of data are estimated and complied.
-First, observed home-to-school trips by mode are retained from travel surveys.
-Next, travel times are calculated assuming active and motorized modes at parcel-to-school level and at the level matching the travel survey.
-Then, travel impedance functions are calibrated using the observed travel data, and using the functions and the parcel-to-school calculated travel times as input, the travel impedance values for each residential location to each school (origin-destination pair) is estimated.
-These three steps are described as follows:
-
-First, we retrieve information about the mode used and origin-destination locations of home-to-school trips from the 2011 and 2016 Transportation Tomorrow Survey (TTS) [@data_management_group_tts_2018]; the 2021 TTS survey is unavailable at the time of writing so 2016 flows are assumed for the 2021 year.
+## Trip length by mode
+
+To reflect multimodal travel behaviour, two types of data were estimated and compiled. The process to do so is explained next.
+
+First, we retrieves information about the mode used and origin-destination locations of home-to-school trips from the 2011 and 2016 Transportation Tomorrow Survey (TTS) [@data_management_group_tts_2018]; the 2021 TTS survey is unavailable at the time of writing so 2016 flows are assumed for the 2021 year.
 The TTS is a travel survey conducted in the Greater Golden Horseshoe Area in Ontario typically every five years.
@@
 
-Next, the travel times matrices for motorized and active travel are estimated using the 'travel_time_matrix()' function part of the {r5r} package [@Pereira2021r5r].
-Travel time matrices are estimated for all TAZ centroids to TAZ centroids (matrices of size for 2011 and 2016: 234 x 234) as they ar enot reported in the TTS.
-Additionally matrices are estimated for all residential parcels to schools (matrices of size 134,340 x 147 for 2011, 139,467 x 137 for 2016, and 143,890 x 132 for 2021) for additional granularity.
-{r5r} is a R-based interface to the R5 routing engine developed separately by Conveyal [@conveyalConveyalR5Routing2022].
+Next, travel times matrices for motorized and active travel were calculated using {r5r} package [@Pereira2021r5r].
+Travel time matrices were estimated for all TAZ centroids to TAZ centroids (matrices of size for 2011 and 2016: 234 x 234) as they are not reported in the TTS.
+Additionally matrices were estimated for all residential parcels to schools (matrices of size 134,340 x 147 for 2011, 139,467 x 137 for 2016, and 143,890 x 132 for 2021) for additional granularity.
+{r5r} is a R-based interface to the R5 routing engine [@conveyalConveyalR5Routing2022].
 For simplicity, we assume motorized travel is by car and non-motorized travel is by walking, as these are the most common modes in their respective categories.
 For travel time calculations, we set a maximum threshold of 60 minutes and use the free-flow OpenStreetMap road network of Hamilton [@geofabrikOntarioOpenStreetMapGeofabrik2022].
-While all car trips are retained, walking trips over 27 minutes are filtered out.
+While all car trips are retained, walking trips over 27 minutes were filtered out.
 This value approximately corresponds to a distance of 1.6 km at a walking speed of 3.6 km/h, which is the threshold for qualifying for motorized transport provided by the school board [@HWDSB_2019].
@@
 In terms of the travel impedance function: we assume that children can go to any elementary school, however, there is a preference for facilities that are more proximate.
-Based on this assumption, we match the associated TAZ-to-TAZ travel times to all observed student-to-school travel flows from the TTS; these are visualised as grey bar columns in @fig-Fig5.
+Based on this assumption, we matched the associated TAZ-to-TAZ travel times to all observed student-to-school travel flows from the TTS; these are visualised as grey bar columns in @fig-Fig5.
 Using these observed values, the  _theoretical_ trip length distribution (TLD) functions are calibrated.
@@
 The interpretability of Hansen-type accessibility has been discussed in numerous studies, including recently by @hu_2019_measuring, @kelobonye2020measuring, and in greater depth by @merlin2017competition along with @soukhovIntroducingSpatialAvailability2023 and @soukhovMultimodalSpatialAvailability2024.
-The interpretation of accessibility depends on how many people demand for the opportunity, especially for exclusive opportunity-types like schools-seats (i.e., one school-seat is for one student).
+The interpretation of accessibility depends on how many people demand the opportunity, especially for exclusive opportunity-types like schools-seats (i.e., one school-seat is for one student).
 
@@
 set_flextable_defaults(
-  font.family = "Arial", font.size = 10,
+  #font.family = "Arial", 
+  font.size = 10,
   border.color = "gray", big.mark = "", digits = 1, pct_digits=1)
@@
 
-Due to spatial availability's flexible proportional allocation mechanism and this cases' calculation at the parcel level, results can also be summarized by community (@tbl-Tab1).
+Due to spatial availability's proportional allocation mechanism and this cases' calculation at the parcel level, results can also be summarized by community (@tbl-Tab1).
 The majority of schools were closed in Hamilton Central as seen in the largest reduction in OTGC by percentage and magnitude.

### 2025-01-12 paper/Manuscript.tex
Proofread revisions

@@
 % Make \paragraph and \subparagraph free-standing
+\makeatletter
 \ifx\paragraph\undefined\else
   \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+  \renewcommand{\paragraph}{
+    \@ifstar
+      \xxxParagraphStar
+      \xxxParagraphNoStar
+  }
+  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
+  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
 \fi
@@
   \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+  \renewcommand{\subparagraph}{
+    \@ifstar
+      \xxxSubParagraphStar
+      \xxxSubParagraphNoStar
+  }
+  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
+  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
 \fi
+\makeatother
 
@@
 \makeatother
+
 \ifLuaTeX
@@
 \abstract{Reducing motorized travel and increasing active travel are top
-priorities for many communities. However, these priorities are
-confronted with pressures to reduce the fiscal burden of public
-services, including schools. Public school boards have been strained to
-do more with less, and opting to close under-capacity schools to deliver
-educational services more cost-effectively. However, once a school is
-closed, students' proximity to schools decreases, and those who
-previously lived close enough to walk or cycle can no longer do so. In
-this paper, the City of Hamilton, a mid-sized city in Ontario, Canada,
-which experienced a 10\% decrease in elementary school locations between
-2011 and 2021, serves as a case study. We quanitfiy how these closures
-and consolidations have reshaped accessibility for the elementary
-school-aged population and discuss their implications for active and
-motorized travel. To measure accessibility, spatial availability, a
-singly-constrained multimodal measure of spatial accessibility is used.
-This method's proportional allocation mechanism enables interpretable
-comparison between accessibility `before' and `after' the implementation
-of school closure/consolidation policy considering changes in potential
-travel time and population. The analysis is conducted at the parcel
-level, considering all school seats as eligible and incorporating
-observed home-to-school travel patterns for motorized and non-motorized
-modes. We present system-wide impact estimates resulting from these
-closures. The findings reveal that overall spatial availability in the
-city declined. While the school consolidation policy may have achieved
-cost savings in facility operation and maintenance, the burden falls on
-students, families, and communities through \emph{potentially} reduced
-opportunities for physical activity through an active journey to school
-and increased motorized travel externalities.}
+priorities for many communities. However, these priorities clash against
+many government's reluctance to invest in public services, including
+schools. Public school boards are increasingly asked to do more with
+less, a directive often accomplished by closing and consolidating
+schools. Alas, this policy tends to reduce the proximity of schools for
+many students, thus reducing the chances for active travel. In this
+paper we study the City of Hamilton, a mid-sized city in Ontario,
+Canada, which experienced numerous elementary schools closures between
+2011 and 2021. Our analysis evaluates how school closures and
+consolidations reshaped accessibility for elementary school students by
+motorized and active travel. In this task we use spatial availability, a
+singly-constrained multimodal measure of spatial accessibility. This
+method's proportional allocation mechanisms allow us to make consistent
+comparisons in accessibility between time periods. Analysis is conducted
+at the parcel level, considering on-the-ground-capacity (OTGC) of
+schools and observed home-to-school trip lengths for motorized and
+active modes. Our findings reveal that overall spatial availability in
+the city declined, and while the school consolidation policy pursued by
+the province may have achieved cost savings in facility operation and
+maintenance, it imposes ongoing costs on students, families, and
+communities through reduced opportunities for physical activity and
+entrenched reliance on motorized travel.}
 
@@
 
-The amount of services accessible within communities oscillate: economic
-and demographic shifts, rural and urban configurations and political
-in/action all induce change. These forces spur pressures to
-reorganization and consolidation of public facilities that provide
-services, potentially improving access for some but not for all
-(Christiaanse 2020; Rosik, Puławska-Obiedowska, and Goliszek 2021).
-Schools are a compelling case study in this respect. In many communities
-globally, school closure and consolidation policies have been pursued by
-governments to optimize per-student operational cost-efficiency (Rong et
-al. 2022; Dai et al. 2019). Research has highlighted that these polices
-have knock-on implications, including the weakening of communities'
-social infrastructure (Butler, Kane, and Cooligan 2019; Irwin and
-Seasons 2012; Autti and Hyry-Beihammer 2014) and negative
-household-level fiscal impacts such as a reduction of property values
-(Merrall, Higgins, and Páez 2024). Operationally, savings may be
-experienced but they are paid at different levels by students, families,
-communities and society.
-
-A school offers many things to a community: but one dimension is the
-travel to and from school, often overlooked by decision-makers
-considering school closure and consolidation policies (Bierbaum, Karner,
-and Barajas 2021; J. Lee and Lubienski 2017). Schools and residential
-locations often proximately co-located, the journey-to-school is an
-excellent opportunity for physical activity (Desjardins et al. 2024).
-However, closures exchange the physical activity of students' realized
-active trips or ``reserved'' active trips (Morency, Roorda, and Demers
-2009; Morency, Demers, and Lapierre 2007) (e.g., trips that could be
-made active) for motorized travel (Rong et al. 2022). Indeed, motorized
-travel to school is the largest contributor of carbon emissions from
-personal daily travel after the journey-to-work (Rong et al. 2022;
-Pantelaki, Claudia Caspani, and Maggi 2024). Despite these societal
-implications, many communities are closing and/or consolidating schools
-(Sageman 2022). For instance, 65\% of comprehensive schools (e.g.~grades
-1 through 9) in rural Finland since 1990 closed (Autti and
-Hyry-Beihammer 2014), 10\% of elementary schools (kindergarten through
-grade 8) in Chicago, USA in 2013 (J. Lee and Lubienski 2017), and 7\% of
-primary and lower secondary schools (kindergarten through grade 9) were
-closed in Denmark between 2010-2011 (Beuchert et al. 2018).
-
-In this paper, we explore the case study of Hamilton, a mid-sized city
-in Canada that has pursued similar school closure and consolidation
-policies. We detail the Canadian political context that incentived these
-policies and quantify their transportation implications using data that
-constructs where students may have lived and the capacity of schools in
-2011, 2016 and 2021. To quantify transportation implications, we use a
-novel multimodal extension of a singly-constrained accessibility measure
-known as spatial availability (Soukhov et al. 2023; Soukhov et al.
-2024). Spatial availability is used to estimate active (i.e., walking)
-and motorized (i.e., car) accessibility considering the capacity of each
-school, and the residential location and quantity of school-aged
-population, and travel times from residential locations to schools for
-three time periods. In this way, this work contributes a empirical
-findings documenting how government austerity hurts public education,
-along with a methodology that results in interpretable accessibility
-values that can be compared across modes, the City, and three time
-periods.
+Public services accessible within communities tend to oscillate:
+economic and demographic shifts, rural and urban configurations and
+political in/action all induce change. These forces spur pressures to
+reorganize and consolidate public facilities, potentially improving
+access for some but not for all residents in a community (Christiaanse
+2020; Rosik, Puławska-Obiedowska, and Goliszek 2021). Public schools are
+a compelling case study in this respect. Across the globe, school
+closure and consolidation policies have been pursued by governments to
+reduce the cost of running public education systems (Rong et al. 2022;
+Dai et al. 2019). Research has highlighted that these polices have
+knock-on implications, including the weakening of communities' social
+infrastructure (Butler, Kane, and Cooligan 2019; Irwin and Seasons 2012;
+Autti and Hyry-Beihammer 2014) and negative household-level fiscal
+impacts such as a reduction of property values (Merrall, Higgins, and
+Páez 2024). Operationally, savings may be experienced by governments,
+but they are paid at different levels by students, families,
+communities, and ultimately society in general.
+
+A school offers many things to a community, and one relevant dimension
+is how it shapes travel. This dimension is often overlooked by
+decision-makers tasked with implementing school closure and
+consolidation policies (Bierbaum, Karner, and Barajas 2021; J. Lee and
+Lubienski 2017). When schools and residential locations are co-located,
+the journey-to-school presents an excellent opportunity for physical
+activity (Desjardins et al. 2024). However, school closures exchange the
+opportunities for active trips by students (Morency, Roorda, and Demers
+2009; Morency, Demers, and Lapierre 2007) for the near certainty of
+motorized travel (Rong et al. 2022). Indeed, motorized travel to school
+is the largest contributor of carbon emissions from personal daily
+travel after the journey-to-work (Rong et al. 2022; Pantelaki, Claudia
+Caspani, and Maggi 2024). Alas, these impacts are not often considered
+in policy when schools close and/or consolidate (Sageman 2022), a
+relatively common occurrence globally. For instance, 65\% of
+comprehensive schools (e.g.~grades 1 through 9) in rural Finland since
+1990 closed (Autti and Hyry-Beihammer 2014), 10\% of elementary schools
+(kindergarten through grade 8) in Chicago, USA in 2013 (J. Lee and
+Lubienski 2017), and 7\% of primary and lower secondary schools
+(kindergarten through grade 9) were closed in Denmark between 2010-2011
+(Beuchert et al. 2018).
+
+In this paper, we investigate the case of Hamilton, a mid-sized city in
+Canada that has been forced by the province of Ontario to pursue school
+closure and consolidation policies. We quantify the transportation
+implications of these policies using the capacity of schools in 2011,
+2016 and 2021, after several waves of school closures and
+consolidations. To this end, we use a novel multimodal extension of a
+singly-constrained accessibility measure known as spatial availability
+(Soukhov et al. 2023; Soukhov et al. 2024). We use this measure to
+estimate active (i.e., walking) and motorized (i.e., car) accessibility
+while constraining the supply of the service to the capacity schools. In
+this way, this work furnishes evidence to document how government
+austerity hurts public goals (such as physical activity) and locks
+school districts and families into motorized patterns of travel for the
+foreseeable future.
 
 To foreshadow the paper's conclusions, our findings indicate that school
-seats have became less spatially available in the city. This change
-happened unevenly, with certain communities being disproportionately
-affected. The school closure and consolidation policy pursued may have
-achieved cost savings in facility operation and maintenance. However, it
-has come at a cost to families, students, and communities, resulting in
-lost opportunities for healthier active travel and a potential increase
-in motorized travel. We investigate: were the closures and
-consolidations worth the operational savings? The methods used in this
-paper can be applied by decision-makers to evaluate the spatial and
-transport-related impacts of public service provision policy.
-Furthermore, to increase the diffusion of this analysis, all work is
-transparent and reproducible, following best practices in the spatial
-sciences (Brunsdon and Comber 2021; Antonio Páez 2021). Accordingly, all
-associated code and data is publicly available within the lead author's
+seats have became less available in the city. This change happened
+unevenly, with certain communities being disproportionately affected.
+The school closure and consolidation policy pursued by the government
+may have achieved cost savings in facility operation and maintenance.
+However, it has come at a cost to families, students, and communities,
+resulting in lost opportunities for healthier life-styles and an
+inevitable increase in motorized travel (Hammer and Baluja 2012; Rose
+2013; BUCK 2019). To increase the diffusion of this analysis, all work
+is transparent and reproducible, following best practices in the spatial
+sciences (Brunsdon and Comber 2021; Antonio Páez 2021). For this reason,
+all associated code and data are publicly available within the lead
+author's
 \href{https://github.com/soukhova/School-closures-accessibility-impacts}{GitHub
@@
 
-The examination of access to public services is a well established area
-of the transportation literature: from parks (El-Murr et al. 2021) to
-healthcare (Pereira, Braga, et al. 2021) and associated dimensions of
-inequities (Tiznado-Aitken, Munoz, and Hurtubia 2021; Bittencourt and
-Giannotti 2023; Kelobonye et al. 2020), spatial accessibility has been
-explored in cases studies worldwide.
-
-Accessibility to schools has been a public service explored across
-several transportation dimensions. For instance, Marques, Wolf, and
-Feitosa (2020) demonstrates that lower-income neighbourhoods, compounded
-by other intersecting dimensions of disadvantage, tend to have lower
-accessibility to schools compared to more advantaged neighbourhoods in
-Portugal. Similar conclusions are drawn in West Virginia, US by Talen
-(2001), England by Burgess et al. (2011), Baton Rouge, US by Williams
-and Wang (2014), São Paulo and Curitiba, Brazil by Moreno-Monroy,
-Lovelace, and Ramos (2018), Pizzol, Giannotti, and Tomasiello (2021),
-and Bittencourt and Giannotti (2023), and Santiago, Chile by
-Tiznado-Aitken, Munoz, and Hurtubia (2021). Research has also
-demonstrated that when school closures occurs, what schools end up
-closing are not a politically neutral decision. For instance, these
-spatial trends are explored in the context of England by Pinch (1987),
-areas of declining school-aged populations in Dreseden, Germany by
-Müller (2011), rural Finnish communities by Autti and Hyry-Beihammer
-(2014), and Chicago, US by J. Lee and Lubienski (2017).
-
-The methods applied to measure school accessibility have been varied,
-from simple indicators to complex metrics. Beginning from more simple,
-Marques, Wolf, and Feitosa (2020) present indicators of school seats per
-student along with the shortest distance from origin to schools within a
-school catchment, methods also used in Talen (2001) and Burgess et al.
-(2011). Authors make refinements to these methods; Müller (2011)
-considers travel time by public transit from origins to schools and
+The examination of accessibility to public services is well-established,
+including studies of parks (Reyes, Paez, and Morency 2014; El-Murr et
+al. 2021), healthcare (Pereira, Braga, et al. 2021), and public
+transportation (Moniruzzaman and Paez 2012; Tiznado-Aitken, Munoz, and
+Hurtubia 2021), among many other services (e.g., Bittencourt and
+Giannotti 2023; Kelobonye et al. 2020).
+
+Schools are a service that has been studied across several
+transportation dimensions. For instance, Marques, Wolf, and Feitosa
+(2020) demonstrate that lower-income neighbourhoods in Portugal tend to
+have lower accessibility to schools compared to more advantaged
+neighbourhoods. Similar conclusions are drawn in West Virginia, US by
+Talen (2001), England by Burgess et al. (2011), Baton Rouge, US by
+Williams and Wang (2014), São Paulo and Curitiba, Brazil by
+Moreno-Monroy, Lovelace, and Ramos (2018), Pizzol, Giannotti, and
+Tomasiello (2021), and Bittencourt and Giannotti (2023), and Santiago,
+Chile by Tiznado-Aitken, Munoz, and Hurtubia (2021). Research has also
+demonstrated that school closures are seldom politically neutral
+decisions. For instance, spatial trends were explored in the context of
+England by Pinch (1987), areas of declining school-aged populations in
+Dreseden, Germany by Müller (2011), rural Finnish communities by Autti
+and Hyry-Beihammer (2014), and Chicago, US by J. Lee and Lubienski
+(2017).
+
+The methods applied to measure school accessibility have also been
+varied. For example, Marques, Wolf, and Feitosa (2020) present
+indicators of school seats per student along with the shortest distance
+from origin to schools within a school catchment. Similar methods are
+also used in Talen (2001) and Burgess et al. (2011). Several authors
+propose refinements to these methods; Müller (2011), for instance,
+considers travel time by public transit from origins to schools, whereas
 Pizzol, Giannotti, and Tomasiello (2021) considers travel times by
@@
 
-In addition to competition for school-seats, theoretically, the
-advantage students using faster modes (e.g., motorized modes) offer in
-terms of a wider range in school-seat choice is also important. Though
-some of the reviewed works using 2SFCA approaches consider multiple
-modes, explicit attention to the modal \emph{advantage} has not been
-discussed. To this end, in this work we use spatial availability, a
-singly-constrained measure of spatial accessibility (Soukhov et al.
-2023; Soukhov et al. 2024). Spatial availability allows for the
-representation of how many `school-seats' are spatially available to
-motorized- and active-trip taking elementary school-aged student
-population at a spatial-level, facilitating interpretable comparison
-across years. Further, spatial availability can be represented as
-school-seats per capita, mathematically equivalent to the
-supply-to-demand ratio of the 2SFCA (see appendix in Soukhov et al.
-(2023)).
+In addition to competition for school-seats, theoretically, the use of
+faster modes (e.g., motorized modes) offers an advantage in terms of a
+wider range in school-seat choice. Though some of the works reviewed
+consider multiple modes, explicit attention to the modal
+\emph{advantage} has not been discussed. To this end, in this work we
+use spatial availability, a singly-constrained measure of spatial
+accessibility (Soukhov et al. 2023; Soukhov et al. 2024). Spatial
+availability allows for the representation of the number of
+`school-seats' that are spatially available to students by mode of
+transportation (i.e., motorized and active travel), considering the
+school-aged student population by zone. Further, spatial availability
+can be represented as school-seats per capita, mathematically equivalent
+to the supply-to-demand ratio of the 2SFCA (see appendix in Soukhov et
+al. (2023)), which increases the interpretability of the analysis, and
+the comparability between years.
 
@@
 
-Hamilton is a mid-size city (\textasciitilde570,000 pop) with rural,
-suburban and urban characteristics in the province of Ontario, Canada
-(Government of Canada 2022a). The city's public English education is
-covered by the Hamilton-Wentworth District School Board (HWDSB). In
-2013, the HWDSB released its Long Term Facilities Master Plan (HWDSB
-2013). To ensure ``equitable, affordable and sustainable learning
-facilities'', the Master Plan indicated that 80\% of its elementary
-schools (attended by students aged approximately 5 to 14) would be
-subject to evaluations, including a ``Pupil Accommodation Review''
-(HWDSB 2013). This Master Plan and the slated Pupil Accommodation
-Reviews prompted public outcry, as the outcome of Pupil Accommodation
-Reviews are typically schools closures and consolidations (Craggs 2013;
-Seasons 2014b).
-
-Like all school boards in Ontario, HWDSB relies on funding from the
-province which is distributed through a formula based on pupil enrolment
-(Mackenzie 2018; Irwin and Seasons 2012). However, it has been
-demonstrated that this funding formula has become insufficient to
-maintain schools in a state of good repair (Auditor General of Ontario
-2015). Further, existing ``top-up'' programs to assist in operation and
-maintenance costs (e.g., School Facilities Operation and Renewal Grant)
-were entirely phased out by 2018 (TDSB 2024). Within this funding
-backdrop, the Pupil Accommodation Review guidelines outlined a
-``community-focused'' process to assessing the value that a school
-provides to students, the community, and the local economy (Seasons
-2014b; Ministry of Education 2006). However in practice, critical
-residents felt the outcomes of the Pupil Accommodation Reviews were
-predetermined (Thompson, Collins, and Dean 2024): older schools were
-slated to be closed, and a wholesome breakdown of costs to repair and
-maintain schools were not made available (Kleinhuis 2013). Other critics
-pointed to unbalanced incentives motivating the HWDSB: ``By closing
-three schools, the HWDSB would have a strong case with the Ministry of
+Hamilton is a mid-size city (\textasciitilde570,000 pop) in the province
+of Ontario, Canada, with rural, suburban and urban characteristics
+(Government of Canada 2022a). The city's public English education system
+is managed by the Hamilton-Wentworth District School Board (HWDSB) and
+the Hamilton-Wentworth Catholic District School Board (HWCDSB). These
+two boards, like many in the province have faced fiscal pressures from
+the province. For example, in 2013, the HWDSB released its Long Term
+Facilities Master Plan (HWDSB 2013), which nominally meant to ensure
+``equitable, affordable and sustainable learning facilities''. This plan
+indicated that 80\% of its elementary schools (attended by students aged
+approximately 5 to 14) would be subject to evaluations, including a
+``Pupil Accommodation Review'' (HWDSB 2013). The announcement promptly
+led to a public outcry, since the outcome of Pupil Accommodation Reviews
+historically translated into school closures and consolidations (Craggs
+2013; Seasons 2014b).
+
+School boards in Ontario rely on funding from the province which is
+distributed through a formula based on pupil enrollment (Mackenzie 2018;
+Irwin and Seasons 2012). However, this funding formula has demonstrably
+become insufficient to maintain schools in a state of good repair
+(Auditor General of Ontario 2015). Further, existing ``top-up'' programs
+to assist in operation and maintenance costs (e.g., School Facilities
+Operation and Renewal Grant) were entirely phased out by 2018 (TDSB
+2024). Against this backdrop, the Pupil Accommodation Review guidelines
+outlined a ``community-focused'' process for assessing the value that a
+school provides to students, the community, and the local economy
+(Seasons 2014b; Ministry of Education 2006). However in practice,
+critical residents felt the outcomes of the Pupil Accommodation Reviews
+were predetermined (Thompson, Collins, and Dean 2024): older schools
+were slated to be closed, and a wholesome breakdown of costs to repair
+and maintain schools was not made available (Kleinhuis 2013). The
+process introduced unbalanced incentives for school boards: as a
+concerned resident pointed at the time, ``{[}b{]}y closing three
+schools, the HWDSB would have a strong case with the Ministry of
 Education to receive full funding for a new school'' (Kleinhuis 2013).
@@
 
-The HWDSB and the Hamilton-Wentworth Catholic District School Board
-(HWCDSB), which together provide public English education to the
-majority of school-aged children in Hamilton (FAO 2023), were
-particularly impacted. While Accommodation Reviews were a province-wide
-trend during this period, Hamilton's school boards underwent an
-unprecedented number of reviews compared to other boards in other
-Ontario municipalities. (Seasons 2014a). In the wave of Pupil
-Accommodation Reviews between 2013 and 2016, 12 of the 147 elementary
-schools closed in Hamilton. Likely in part due to the public
-unpopularity of school closures, in 2017, the provincial government
-introduced a moratorium on beginning new Pupil Accommodation Reviews. In
-the following years, the HWDSB addressed its backlog and closed 12 more
+Hamilton's two school boards (HWDSB and HWCDSB), which together provide
+public English education to the majority of school-aged children in
+Hamilton (FAO 2023), were particularly impacted by fiscal pressures.
+While Accommodation Reviews were a province-wide trend during this
+period, Hamilton's school boards underwent an unprecedented number of
+reviews compared to boards in other Ontario municipalities. (Seasons
+2014a). In the wave of Pupil Accommodation Reviews between 2013 and
+2016, 12 of the 147 elementary schools closed in Hamilton. Likely in
+part due to the unpopularity of school closures, in 2017, the provincial
+government introduced a moratorium, which paused any scheduled Pupil
+Accommodation Reviews. This created a backlog of Accommodation Reviews
+that HWDSB caught up to in subsequent years, when it closed 12 more
 schools between 2017 to 2021 (HWDSB 2023). Though elementary school
@@
 
-A justification in these Pupil Accommodation Reviews are operational
+A justification for these Pupil Accommodation Reviews are operational
 savings based on ``projected'' reductions in student population and
 ``under-utilized'' school capacity (Craggs 2012; TDSB 2024; OPSBA 2024).
-As later demonstrated in this study, these policies in fact did result
-in school capacity per student reducing overall. Indeed, school seats in
-the city declined 5\%. It should be noted that though closures dominated
-the Pupil Accommodation Review process, some schools were
-``consolidated'' e.g., a school or two were closed and another school
-was expanded or opened to accommodate students from closed schools.
-Taken that school locations decreased more than overall school-seats,
-elementary schools in Hamilton are on average larger than they once
-were.
-
-In 2024, the moratorium on Pupil Accommodation Reviews is still in place
-and school boards are motivated by the same funding conditions. In
-absence of new revenue-generating levers and under the same funding
-formula with even lower inflation adjusted per-pupil funding, many
-school boards are calling for the end of the moratorium (OPSBA 2024;
-TDSB 2024; Draaisma 2024). In the meantime, some school boards are
-reducing specialized and community-wide programming (TDSB 2024). Under
-the status-quo, the end of the moratorium will result in further school
-closures. These pertinent decisions, made under an austere funding
-formula and flawed school board incentives does not create an
-environment where solutions for a sustainable public education system
-can thrive.
-
-Though schools have a plethora of benefits not captured by their
-maintenance and operational costs (Butler, Kane, and Cooligan 2019;
-Irwin and Seasons 2012; Autti and Hyry-Beihammer 2014), this paper's
-scope is limited to their travel implications. For students who live in
-proximity to schools, their closure or consolidation directly eliminate
-the realized or potential active trips that are, or could have been,
-made. Closing a school permanently reduces the potential for active
-trips and further entrenches reliance on motorized travel. Additionally,
-from the perspective of modal equity, school-aged populations with
-access to motorized transport gain a speed-time advantage, providing
-them with greater flexibility and choice in accessing school-seats.
+These policies led to overall reductions in school capacity per student:
+school seats in the city declined 5\%--but the impacts were felt
+unevenly across the city. It should be noted that although closures
+dominated the Pupil Accommodation Review process, some schools were
+``consolidated'', that is, a school or two were closed and another
+school was expanded or opened to accommodate students from closed
+schools. Since school locations decreased proportionally more than
+school-seats, elementary schools in Hamilton are on average larger than
+they once were.
+
+Existing research indicates that schools have a plethora of benefits not
+captured by their maintenance and operational costs (Butler, Kane, and
+Cooligan 2019; Irwin and Seasons 2012; Autti and Hyry-Beihammer 2014),
+and in this paper we focus on the transportation implications. For
+students who live in proximity to schools, closures or consolidations
+directly eliminate the potential for active trips, and consequently
+physical activity. Closing a school permanently reduces the potential
+for active trips and further entrenches reliance on motorized travel.
+Additionally, from the perspective of modal equity, school-aged
+populations with access to motorized transport gain a speed-time
+advantage, providing them with greater flexibility and choice in
+accessing school-seats.
 
 The spatial and population-level distribution of these impacts are of
-question for this paper. Specifically, the paper's aim is to demonstrate
+question for this paper. Specifically, the paper's aim is to investigate
 how these policies impacted the motorized and non-motorized (active)
-school-seat accessibility landscape. We hypothesize that the school
-closure and consolidation policies reduced accessibility to school-seats
-for many students. By extension, this loss in access and the
+school-seat accessibility landscape. We hypothesize that school closure
+and consolidation policies reduced the availability of school-seats for
+many students. By extension, this loss in accessibility and the
 displacement of active trips likely resulted in decreased potential
@@
 \item
-  How has spatial availability changed for home-to-school trips
-  following school closures and consolidations?
+  Did spatial availability change for home-to-school trips following
+  school closures and consolidations?
 \item
@@
 
-Our focus is to evaluate the impact of school consolidations between
-2011, 2016 and 2021 on multimodal school-seat spatial availability in
-Hamilton. To facilitate comparison, the following conditions are
-constructed for each year: 1) school configurations, 2) average student
-population, low-income prevalence of households, and residential parcel
-locations, and 3) estimated multimodal travel behaviour. The following
-sub-sections describe these three types of data.
+We focus on the evaluation of the impacts of school closures and
+consolidations between 2011, 2016 and 2021. The outcome of interest is
+the multimodal school-seat spatial availability in Hamilton. To research
+this, the following data were prepared for each year: 1) school
+configurations, 2) student population, prevalence of low-income
+households, and residential parcel locations, and 3) estimated trip
+length by mode. These data are described next.
 
@@
 
-First, elementary schools locations, how the changed throughout the
-study period and their capacities are required for each studied year.
+For this study, we use the locations of elementary schools and their
+capacities at each of three years: 2011, 2016, and 2021.
 
@@
 authors with locations and associated school catchments for the
-2010-2011 and 2015-2016 academic years. Data for 2021 locations was
+2010-2011 and 2015-2016 academic years. School locations in 2021 were
 retrieved from the City's open data portal (Hamilton 2024). In Hamilton,
 the majority of student-aged population attend school in one of the two
-English public school boards (FAO 2023). One school board is referred as
-Hamilton Wentworth District School Board (HWDSB) and the other as
-Hamilton Wentworth Catholic District School Board (HWCDSB). School
-catchments for the public and public-catholic catchments indicate what
-residential property is assigned to what school by default. However,
-families can decide to attend schools out of catchment; in fact,
-21\%-23\% of motorized school trips for populations aged 5 through 14
-are out of catchment according to the regional travel survey (the
-Transportation Tomorrow Survey (TTS)) in 2011 and 2016, the most recent
-TTS available at the time of writing (Data Management Group 2018).
-
-In addition to the school's location in 2011, 2016 and 2021, all schools
-were also tagged with a school status regarding their existence and/or
-change in location or school-seat capacity. A summary of the status of
-schools through 2011 to 2021 is illustrated in Figure~\ref{fig-Fig1}.
-Figure~\ref{fig-Fig1} also displays a count of the number of schools
-with the status between years, the number of schools overall present in
-each year, and the symbolic representation of a school status that is
-used in Figures throughout the paper. Overall, a school can have one of
-seven school statuses in any given year. Either: 1) the school did not
-change throughout 2011 to 2021, 2) the school expanded (i.e., increased
-school-seat capacity) or was 3) relocated and expanded (N.B., schools
-expanded only between 2011 through 2016 as they were typically done
-\emph{before} school closures and no new Pupil Accommodation Review was
-commenced after 2017), 4) the school closed sometime after 2011 and
-before 2016 or 5) after 2016 and before 2021, and 6) the school opened
-sometime after 2011 and before 2016 or 7) after 2016 and before 2021.
+English public school boards (FAO 2023). School catchments for the
+public and public-catholic catchments indicate the default school for
+households. However, families can decide to attend schools out of
+catchment, and in fact, 21\%-23\% of motorized school trips for
+populations aged 5 through 14 are out of catchment according to the
+regional travel survey (the Transportation Tomorrow Survey, or TTS) in
+2011 and 2016, the most recent TTS available at the time of writing
+(Data Management Group 2018).
+
+In addition to school locations, all schools were also tagged with a
+school status regarding their operational status (active or closed)
+and/or changes in location or school-seat capacity. A summary of the
+status of schools through 2011 to 2021 is shown in
+Figure~\ref{fig-Fig1}. Figure~\ref{fig-Fig1} also displays a count of
+the number of schools with the status between years, the number of
+schools overall present in each year, and the symbolic representation of
+a school status that is used in Figures throughout the paper. Overall, a
+school can have one of seven school statuses in any given year. Either:
+1) the school did not change throughout 2011 to 2021, 2) the school
+expanded (i.e., increased school-seat capacity) or was 3) relocated and
+expanded (N.B., schools expanded only between 2011 through 2016 as they
+were typically done \emph{before} school closures and no new Pupil
+Accommodation Review was commenced after 2017), 4) the school closed
+sometime after 2011 and before 2016 or 5) after 2016 and before 2021,
+and 6) the school opened sometime after 2011 and before 2016 or 7) after
+2016 and before 2021.
 
@@
 
-The capacity of each school is also required for each simulated year.
-This capacity is reflected as the number of `seats' available in the
-school. Enrolment exceeding this capacity indicates over-enrolment,
-while enrolment below it indicates under-enrolment. This value is
+The capacity of each school is also required for each year. This
+capacity is the number of `seats' available in the school. Capacity is
 calculated by the provincial Ministry of Education and referred to as
-the on-the-ground capacity (OTGC). The current and historic OTGCs for
-schools are no longer publicly available (Ontario 2017), so we estimated
-them using a multivariate regression model and validated the results
-with a set of OTGC values manually gathered. These values were retrieved
-for schools that had undergone a completed Pupil Accommodation Review
-and had publicly available School Information Profiles.
-
-For schools lacking OTGC values, two regression models were created: one
-for public schools and another for public catholic schools. The
-independent variables in the model were 1) a dummy variable indicating
-the type of school grade instruction (\emph{MidElem}, \emph{JrElem},
-\emph{Elem}), 2) the school's building footprint (\emph{F}) in units of
-(\(m^2\)), and 3) the school's shortest Euclidean distance from the
-centroid of Hamilton's central business district area (\emph{DistCBD})
-defined in metres:
-
-\begin{itemize}
-\tightlist
-\item
+the ``on-the-ground capacity'' (OTGC) of a school, albeit it is
+important to note that schools rarely operate at their nominal capacity,
+and can be either over-enrolled or under-enrolled.
+
+Current and historic OTGCs for schools are not publicly available
+(Ontario 2017), so we used a set of known OTGC values gathered from
+Pupil Accommodation Review documents and multivariate regression to
+estimate and validate these values. Two regression models were created:
+one for public schools and another for public catholic schools. The
+independent variables in these models were 1) a dummy variable
+indicating the level of the school (\emph{MidElem}, \emph{JrElem},
+\emph{Elem}); 2) the school's building footprint (\emph{F}) in \(m^2\);
+and 3) the school's Euclidean distance from the centroid of Hamilton's
+central business district area (\emph{DistCBD}) in metres.
+
 Formally, elementary schools are defined as schools that provide
-  instruction to any combination of grades between kindergarten to grade
-  8 (i.e.~typically children aged 5 to 14). As such, elementary includes
+instruction to any combination of grades between kindergarten to grade 8
+(i.e.~typically children aged 5 to 14). As such, elementary includes
 middle schools that only instruct grades 6 to 8 (\emph{MidElem}),
@@
 public school board.
+
+\begin{itemize}
 \item
@@
   \emph{DistCBD} was calculated `as the crow flies' from each school to
-  the centroid of the Hamilton CBD (43.256684°N, 79.869039°W). Notably,
-  this variable can also be seen as a proxy for school construction age
-  as older buildings in Hamilton are generally located closer to the
-  CBD, the ``old'' Hamilton community (Merrall 2021).
+  the centroid of the Hamilton CBD (43.256684\(\circ\)N,
+  79.869039\(\circ\)W). Notably, this variable can also be seen as a
+  proxy for school construction age as older buildings in Hamilton are
+  generally located closer to the CBD, the ``old'' Hamilton community
+  (Merrall 2021).
 \end{itemize}
 
-Equations (\ref{eq:OTGC-public}) and (\ref{eq:OTGC-catholic}) summarise
-the OTGC regression model for public schools and public catholic schools
+Equations (\ref{eq:OTGC-public}) and (\ref{eq:OTGC-catholic}) are the
+OTGC regression model for public schools and public catholic schools
 respectively. As seen in Table \ref{TabA1-OTGC}, the coefficient of
-determination for these models are fantastic: \(R^2 = 0.999\) and
-\(R^2 = 0.998\) for the public \(OTGC_{Public}\) and public catholic
-\(OTGC_{Public Catholic}\) school boards respectively, and the residual
-standard deviation is \(\sigma = 0.212\) and \(sigma = 0.331\).
+determination for these models are quite high, at \(R^2 = 0.999\) and
+\(R^2 = 0.998\) for the public and public catholic school boards
+respectively, and the residual standard deviation is \(\sigma = 0.212\)
+and \(sigma = 0.331\).
 
@@
 Taken together, Figure~\ref{fig-Fig2} displays the city's elementary
-schools locations, status and their estimated or retrieved OTGCs for
+schools locations, status, and their observed or estimated OTGCs for
 2011, 2016 and 2021. For additional context on the degree of residential
@@
 
-Where students live are estimated at the level of the residential
-parcel. Centroids for parcels in 2011, 2016 and 2021 are retrieved from
-Teranet (2009), representing 134,340, 139,467 and 143,890 unique
-locations in those years, respectively. The datasets do not provide
-information on the number of households or children per parcel. To
-estimate this value, the average number of students per parcel is
-calculated using the number of school-aged population in each DA is
-divided by the number of households in each DA. This rate is assigned to
-all parcels within an associated DA. Due to the proprietary nature of
-the parcel data, the second row in Figure~\ref{fig-Fig3} visualises an
+Student residential locations were approximated at the level of
+residential parcels. Centroids for parcels in 2011, 2016 and 2021 were
+retrieved from Teranet (2009), representing \(134,340\), \(139,467\) and
+\(143,890\) unique locations, respectively. Each residential unit in a
+parcel was populated with the average number of children aged 5-14 in
+the corresponding dissemination area (DA). Due to the proprietary nature
+of the parcel data, the second row in Figure~\ref{fig-Fig3} shows an
 aggregation of the information: the average rate of 5-14 year old
@@
 
-\subsection{Multimodal travel
-behaviour}\label{multimodal-travel-behaviour}
-
-To reflect multimodal travel behaviour for all studied years, two types
-of data are estimated and complied. First, observed home-to-school trips
-by mode are retained from travel surveys. Next, travel times are
-calculated assuming active and motorized modes at parcel-to-school level
-and at the level matching the travel survey. Then, travel impedance
-functions are calibrated using the observed travel data, and using the
-functions and the parcel-to-school calculated travel times as input, the
-travel impedance values for each residential location to each school
-(origin-destination pair) is estimated. These three steps are described
-as follows:
-
-First, we retrieve information about the mode used and
+\subsection{Trip length by mode}\label{trip-length-by-mode}
+
+To reflect multimodal travel behaviour, two types of data were estimated
+and compiled. The process to do so is explained next.
+
+First, we retrieves information about the mode used and
 origin-destination locations of home-to-school trips from the 2011 and
@@
 
-Next, the travel times matrices for motorized and active travel are
-estimated using the `travel\_time\_matrix()' function part of the
-\{r5r\} package (Pereira, Saraiva, et al. 2021). Travel time matrices
-are estimated for all TAZ centroids to TAZ centroids (matrices of size
-for 2011 and 2016: 234 x 234) as they ar enot reported in the TTS.
-Additionally matrices are estimated for all residential parcels to
-schools (matrices of size 134,340 x 147 for 2011, 139,467 x 137 for
-2016, and 143,890 x 132 for 2021) for additional granularity. \{r5r\} is
-a R-based interface to the R5 routing engine developed separately by
-Conveyal (Conveyal {[}2015{]} 2022). For simplicity, we assume motorized
-travel is by car and non-motorized travel is by walking, as these are
-the most common modes in their respective categories. For travel time
-calculations, we set a maximum threshold of 60 minutes and use the
-free-flow OpenStreetMap road network of Hamilton (Geofabrik 2022). While
-all car trips are retained, walking trips over 27 minutes are filtered
-out. This value approximately corresponds to a distance of 1.6 km at a
+Next, travel times matrices for motorized and active travel were
+calculated using \{r5r\} package (Pereira, Saraiva, et al. 2021). Travel
+time matrices were estimated for all TAZ centroids to TAZ centroids
+(matrices of size for 2011 and 2016: 234 x 234) as they are not reported
+in the TTS. Additionally matrices were estimated for all residential
+parcels to schools (matrices of size 134,340 x 147 for 2011, 139,467 x
+137 for 2016, and 143,890 x 132 for 2021) for additional granularity.
+\{r5r\} is a R-based interface to the R5 routing engine (Conveyal
+{[}2015{]} 2022). For simplicity, we assume motorized travel is by car
+and non-motorized travel is by walking, as these are the most common
+modes in their respective categories. For travel time calculations, we
+set a maximum threshold of 60 minutes and use the free-flow
+OpenStreetMap road network of Hamilton (Geofabrik 2022). While all car
+trips are retained, walking trips over 27 minutes were filtered out.
+This value approximately corresponds to a distance of 1.6 km at a
 walking speed of 3.6 km/h, which is the threshold for qualifying for
@@
 go to any elementary school, however, there is a preference for
-facilities that are more proximate. Based on this assumption, we match
+facilities that are more proximate. Based on this assumption, we matched
 the associated TAZ-to-TAZ travel times to all observed student-to-school
@@
 Soukhov et al. (2023) and Soukhov et al. (2024). The interpretation of
-accessibility depends on how many people demand for the opportunity,
+accessibility depends on how many people demand the opportunity,
 especially for exclusive opportunity-types like schools-seats (i.e., one
@@
 
-\multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{\textbf{}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{\textbf{Hamilton\ C.}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{\textbf{\%\ Δ}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{\textbf{Dundas}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{\textbf{\%\ Δ}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{\textbf{Stoney\ Creek}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{\textbf{\%\ Δ}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{\textbf{Glanbrook}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{\textbf{\%\ Δ}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{\textbf{Flamborough}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{\textbf{\%\ Δ}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{\textbf{Ancaster}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{\textbf{\%\ Δ}}}}} \\
+\multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{\textbf{}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{\textbf{Hamilton\ C.}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{\textbf{\%\ Δ}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{\textbf{Dundas}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{\textbf{\%\ Δ}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{\textbf{Stoney\ Creek}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{\textbf{\%\ Δ}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{\textbf{Glanbrook}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{\textbf{\%\ Δ}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{\textbf{Flamborough}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{\textbf{\%\ Δ}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{\textbf{Ancaster}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{\textbf{\%\ Δ}}}}} \\
 
@@
 
-\multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{LIM-AT}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{23.8516970}}}} & \multicolumn{1}{>{\cellcolor[HTML]{F4FBF3}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{13.2\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{11.2879110}}}} & \multicolumn{1}{>{\cellcolor[HTML]{F6FCF5}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{10.6\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{10.8019420}}}} & \multicolumn{1}{>{\cellcolor[HTML]{F7FCF7}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{9.1\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{8.3083330}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFFAF8}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-2.5\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{7.2350900}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FAFDFA}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{5.4\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{7.0989650}}}} & \multicolumn{1}{>{\cellcolor[HTML]{EAF7E9}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{24.9\%}}}} \\
+\multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{LIM-AT}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{23.8516970}}}} & \multicolumn{1}{>{\cellcolor[HTML]{F4FBF3}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{13.2\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{11.2879110}}}} & \multicolumn{1}{>{\cellcolor[HTML]{F6FCF5}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{10.6\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{10.8019420}}}} & \multicolumn{1}{>{\cellcolor[HTML]{F7FCF7}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{9.1\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{8.3083330}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFFAF8}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-2.5\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{7.2350900}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FAFDFA}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{5.4\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{7.0989650}}}} & \multicolumn{1}{>{\cellcolor[HTML]{EAF7E9}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{24.9\%}}}} \\
 
@@
 
-\multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{Kid\ pop.}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{35310.5960000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFF9F7}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-3.2\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{2590.0000000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFE3D9}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-14.7\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{7514.7510000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{EEF9ED}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{20.0\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{2639.6530000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{A2DB9F}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{108.9\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{5365.0000000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{F8FCF7}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{8.7\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{4845.0000000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{F5FBF4}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{12.4\%}}}} \\
+\multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{Kid\ pop.}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{35310.5960000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFF9F7}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-3.2\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{2590.0000000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFE3D9}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-14.7\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{7514.7510000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{EEF9ED}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{20.0\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{2639.6530000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{A2DB9F}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{108.9\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{5365.0000000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{F8FCF7}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{8.7\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{4845.0000000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{F5FBF4}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{12.4\%}}}} \\
 
@@
 
-\multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{OTGC}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{39889.0820000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFECE6}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-9.8\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{2377.0380000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFFFFF}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{0.0\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{9010.1960000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFF1EB}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-7.6\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{1313.5220000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{91D58F}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{127.6\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{4590.7010000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFF1EB}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-7.5\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{3386.8900000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{F6FBF5}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{11.1\%}}}} \\
+\multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{OTGC}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{39889.0820000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFECE6}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-9.8\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{2377.0380000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFFFFF}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{0.0\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{9010.1960000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFF1EB}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-7.6\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{1313.5220000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{91D58F}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{127.6\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{4590.7010000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFF1EB}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-7.5\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{3386.8900000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{F6FBF5}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{11.1\%}}}} \\
 
@@
 
-\multicolumn{13}{>{\centering}m{\dimexpr 9.75in+24\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{\textbf{Motorized}}}}} \\
+\multicolumn{13}{>{\centering}m{\dimexpr 9.75in+24\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{\textbf{Motorized}}}}} \\
 
@@
 
-\multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{V\ (mt)}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{42527.7500000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFE9E1}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-11.6\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{1952.2590000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFD4C5}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-22.5\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{6328.0100000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{F7FCF6}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{9.9\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{1531.3080000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{9DD99A}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{114.8\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{3382.5270000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFFBF9}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-2.3\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{4086.6160000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{F9FDF9}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{7.2\%}}}} \\
+\multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{V\ (mt)}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{42527.7500000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFE9E1}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-11.6\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{1952.2590000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFD4C5}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-22.5\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{6328.0100000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{F7FCF6}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{9.9\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{1531.3080000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{9DD99A}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{114.8\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{3382.5270000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFFBF9}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-2.3\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{4086.6160000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{F9FDF9}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{7.2\%}}}} \\
 
@@
 
-\multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{v\ (mt)}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{1.2393060}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFEEE8}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-8.9\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{0.7590198}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFEEE7}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-9.0\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{0.8369173}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFF1EC}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-7.1\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{0.5823797}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FCFEFC}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{3.1\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{0.6312921}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFECE5}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-10.1\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{0.8491653}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFF6F2}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-5.0\%}}}} \\
+\multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{v\ (mt)}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{1.2393060}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFEEE8}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-8.9\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{0.7590198}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFEEE7}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-9.0\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{0.8369173}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFF1EC}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-7.1\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{0.5823797}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FCFEFC}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{3.1\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{0.6312921}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFECE5}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-10.1\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{0.8491653}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFF6F2}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-5.0\%}}}} \\
 
@@
 
-\multicolumn{13}{>{\centering}m{\dimexpr 9.75in+24\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{\textbf{Non-motorized}}}}} \\
+\multicolumn{13}{>{\centering}m{\dimexpr 9.75in+24\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{\textbf{Non-motorized}}}}} \\
 
@@
 
-\multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{V\ (nmt)}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{631.8817250}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFEAE2}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-11.0\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{28.7632500}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFF1EC}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-7.1\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{65.1592800}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFB39B}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-39.5\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{2.3855420}}}} & \multicolumn{1}{>{\cellcolor[HTML]{7CCD7C}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{1274.9\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{12.2516440}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFFDFC}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-1.0\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{18.5170610}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FF7957}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-67.6\%}}}} \\
+\multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{V\ (nmt)}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{631.8817250}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFEAE2}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-11.0\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{28.7632500}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFF1EC}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-7.1\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{65.1592800}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFB39B}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-39.5\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{2.3855420}}}} & \multicolumn{1}{>{\cellcolor[HTML]{7CCD7C}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{1274.9\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{12.2516440}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFFDFC}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-1.0\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{18.5170610}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FF7957}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-67.6\%}}}} \\
 
@@
 
-\multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{v\ (nmt)}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{0.7054674}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFE6DD}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-13.0\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{1.5790158}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFDFD4}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-16.6\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{0.8967578}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFCFBE}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-25.3\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{1.6565846}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFD0BF}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-24.8\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{1.8555444}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFE8DF}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-12.2\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{0.8641681}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFDFD3}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{Arial}{-17.0\%}}}} \\
+\multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{v\ (nmt)}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{0.7054674}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFE6DD}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-13.0\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{1.5790158}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFDFD4}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-16.6\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{0.8967578}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFCFBE}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-25.3\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{1.6565846}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFD0BF}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-24.8\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{1.8555444}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFE8DF}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-12.2\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{0.8641681}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFDFD3}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-17.0\%}}}} \\
 
@@
 
-Due to spatial availability's flexible proportional allocation mechanism
-and this cases' calculation at the parcel level, results can also be
-summarized by community (Table~\ref{tbl-Tab1}). The majority of schools
-were closed in Hamilton Central as seen in the largest reduction in OTGC
-by percentage and magnitude. Hamilton Central represents 55\% of the
-5-14 year old population in 2021. Though the population decreased a
-modest amount from 2011 to 2021, the rate of school-seat spatial
-availability decreased disproportionately more (population by -3.2\% and
-spatial availability by -11.6\% for both modes). Schools and additional
-OTGC was not sufficiently expanded to provide the same levels of spatial
+Due to spatial availability's proportional allocation mechanism and this
+cases' calculation at the parcel level, results can also be summarized
+by community (Table~\ref{tbl-Tab1}). The majority of schools were closed
+in Hamilton Central as seen in the largest reduction in OTGC by
+percentage and magnitude. Hamilton Central represents 55\% of the 5-14
+year old population in 2021. Though the population decreased a modest
+amount from 2011 to 2021, the rate of school-seat spatial availability
+decreased disproportionately more (population by -3.2\% and spatial
+availability by -11.6\% for both modes). Schools and additional OTGC was
+not sufficiently expanded to provide the same levels of spatial
 availability as in 2011. This disproportionate decrease in spatial
@@
 
+\bibitem[\citeproctext]{ref-buckWhy2019}
+BUCK, NAOMI. 2019. {``Why Did Our Children Stop Walking to School?''}
+\emph{The Globe and Mail}, October, O1.
+\url{http://global.factiva.com/redir/default.aspx?P=sa&an=GLOB000020191019efaj00011&cat=a&ep=ASE}.
+
 \bibitem[\citeproctext]{ref-burgess2011parental}
@@
 
-\bibitem[\citeproctext]{ref-draaismaTDSBConsideringCuts2024}
-Draaisma, Muriel. 2024. {``{TDSB} Considering Cuts to Programming Amid
-\$20M Deficit and Many Trustees Are Concerned.''} \emph{{CBC} News},
-March.
-\url{https://www.cbc.ca/news/canada/toronto/toronto-district-school-board-finance-budget-enrolment-committee-proposed-cuts-1.7149195}.
-
 \bibitem[\citeproctext]{ref-elmurrWalkingAccessibilityParks2021}
@@
 
+\bibitem[\citeproctext]{ref-hammerWhy2012}
+Hammer, Kate, and Tamara Baluja. 2012. {``Why Our Schools Need a Phys-Ed
+Revolution; {An} Inactive Generation with Soaring Obesity Rates Shows
+Schools Are Flunking When It Comes to Fitness.''} \emph{The Globe and
+Mail}, May, A1.
+\url{http://global.factiva.com/redir/default.aspx?P=sa&an=GLOB000020120524e85o0001r&cat=a&ep=ASE}.
+
 \bibitem[\citeproctext]{ref-hansenHowAccessibilityShapes1959}
@@
 
+\bibitem[\citeproctext]{ref-moniruzzamanAccessibility2012}
+Moniruzzaman, M., and A. Paez. 2012. {``Accessibility to Transit, by
+Transit, and Mode Share: Application of a Logistic Model with Spatial
+Filters.''} \emph{Journal of Transport Geography} 24 (September):
+198--205. \url{https://doi.org/10.1016/j.jtrangeo.2012.02.006}.
+
 \bibitem[\citeproctext]{ref-morencyHowMany2007}
@@
 
+\bibitem[\citeproctext]{ref-reyesalking2014}
+Reyes, M., A. Paez, and C. Morency. 2014. {``Walking Accessibility to
+Urban Parks by Children: {A} Case Study of {Montreal}.''}
+\emph{Landscape and Urban Planning} 125 (May): 38--47.
+\url{https://doi.org/10.1016/j.landurbplan.2014.02.002}.
+
 \bibitem[\citeproctext]{ref-rongReviewResearchLowcarbon2022}
@@
 \url{https://doi.org/10.1016/j.jtrangeo.2022.103306}.
+
+\bibitem[\citeproctext]{ref-roseKids2013}
+Rose, Lauren La. 2013. {``Kids Get {D} Minus on Physical Activity Report
+Card as Fewer Walk or Bike to School.''} \emph{The Globe and Mail}, May,
+L12.
+\url{http://global.factiva.com/redir/default.aspx?P=sa&an=GLOB000020130522e95m00009&cat=a&ep=ASE}.
 

### 2025-01-19 paper/Manuscript.qmd
R2 submitted.

@@
 ---
-title: "10 years of school closures and consolidations: the impact on school accessibility"
+title: "Ten years of school closures and consolidations in Hamilton, Canada and the impact on multimodal accessibility"
 bibliography: 
@@
 abstract: | 
-  | Reducing motorized travel and increasing active travel are top priorities for many communities.
-However, these priorities clash against many government's reluctance to invest in public services, including schools.
+  | Reducing motorized travel and increased active travel are top priorities for many communities.
+However, these goals are challenged by pressures that make governments reluctant to re/invest in public services, including schools.
 Public school boards are increasingly asked to do more with less, a directive often accomplished by closing and consolidating schools.
-Alas, this policy tends to reduce the proximity of schools for many students, thus reducing the chances for active travel.
-In this paper we study the City of Hamilton, a mid-sized city in Ontario, Canada, which experienced numerous elementary schools closures between 2011 and 2021.
-Our analysis evaluates how school closures and consolidations reshaped accessibility for elementary school students by motorized and active travel.
-In this task we use spatial availability, a singly-constrained multimodal measure of spatial accessibility.
+Such policies can reduce the proximity of schools for many students, thereby limiting opportunities for active travel.
+In this paper, we examine the City of Hamilton, a mid-sized city in Ontario, Canada, which saw numerous elementary schools closures between 2011 and 2021.
+Our analysis evaluates how school closures and consolidations reshaped accessibility for elementary school students, both by motorized and active travel.
+We use spatial availability, a singly-constrained multimodal measure of spatial accessibility.
 This method's proportional allocation mechanisms allow us to make consistent comparisons in accessibility between time periods.
-Analysis is conducted at the parcel level, considering on-the-ground-capacity (OTGC) of schools and observed home-to-school trip lengths for motorized and active modes.
-Our findings reveal that overall spatial availability in the city declined, and while the school consolidation policy pursued by the province may have achieved cost savings in facility operation and maintenance, it imposes ongoing costs on students, families, and communities through reduced opportunities for physical activity and entrenched reliance on motorized travel.
+The analysis is conducted at the parcel level, taking into account the on-the-ground-capacity (OTGC) of schools and observed home-to-school trip lengths for motorized and active modes.
+Overall, our findings reveal a decline in school spatial availability in the city.
+While the school closure and consolidation policy pursued by the province may have achieved cost savings in facility operation and maintenance, it imposed ongoing costs on students, families, and communities through reduced opportunities for physical activity and entrenched reliance on motorized travel.
 
@@
   tm_polygons(col = "perc_urb", border.alpha=0, 
-              alph=0.8,
+              alph=0.7,
               palette = c("yellow3","yellow4"),
@@
               style="cont",
-              title = "Proportion urban land-use") +
+              title = "% Urbanisation") +
 
@@
             panel.label.bg.color = "white",
-            panel.label.height = c(0.8,0,0)) +
-  # tm_add_legend(type = "line", 
-  #               col = c("grey40","lightblue"),
-  #               labels = c("TAZ boundary","Community boundary"))+
+            panel.label.height = c(0.8,0,0))+
   tm_add_legend(type = "symbol",
                 col = "darkblue",
-                labels = "TAZ centroid")
-
-OD_flow_plot <- tmap_arrange(car_OD_flow_plot, walk_OD_flow, nrow = 2)
-
+                labels = "TAZ centroid",
+                size = 0.1)
+
+OD_flow_plot <- tmap_arrange(car_OD_flow_plot,
+                             walk_OD_flow, nrow = 2)
+OD_flow_plot
 tmap_save(OD_flow_plot, paste0(here::here(),"/Figures/Fig4.tiff"), height=4.5, width=8, dpi=600)
@@
   geom_text(data = q_1, 
-            mapping = aes(x = x, y = y, label = label), colour = "grey20", size=4) +
+            mapping = aes(x = x, y = y, label = label), colour = "grey20", size=3.5) +
   geom_text(data = q_2,
-            mapping = aes(x = x, y = y, label = label), colour = "grey20", size=4) +
+            mapping = aes(x = x, y = y, label = label), colour = "grey20", size=3.5) +
   geom_text(data = q_3,
-            mapping = aes(x = x, y = y, label = label), colour = "grey20", size=4) +
+            mapping = aes(x = x, y = y, label = label), colour = "grey20", size=3.5) +
   theme_classic() +
@@
   tm_lines(col = "COMMUNITY_URB", 
-  palette = c("grey0","grey20","grey40", "grey60",
-                       "grey80","grey100"),
-           title.col = "Hamilton communities", lwd = 1.5) + 
+           palette = c("#365C8DFF", "#277F8EFF", "#1FA187FF", "#4AC16DFF", "#9FDA3AFF", "#FDE725FF"),
+           title.col = "Hamilton communities\n(% urbanisation)", lwd = 1) + 
   tm_shape(ALL_SCHOOLS_Elem_formapping2) +
@@
              size= "OTGC",
-             alpha=0.6,
+             alpha=0.8,
              scale = 0.4,
@@
   tm_lines(col = "COMMUNITY_URB", 
-           palette = c("grey0","grey20","grey40", "grey60",
-                       "grey80","grey100"),
-           title.col = "Hamilton communities", lwd = 1.5, legend.col.show = FALSE) + 
+           palette = c("#365C8DFF", "#277F8EFF", "#1FA187FF", "#4AC16DFF", "#9FDA3AFF", "#FDE725FF"),
+           title.col = "Hamilton communities\n(% urbanisation)", lwd = 1, legend.col.show=FALSE) + 
 tm_shape(ALL_SCHOOLS_Elem_formapping2) +
@@
   tm_lines(col = "COMMUNITY_URB", 
-           palette = c("grey0","grey20","grey40", "grey60","grey80","grey100"),
-           title.col = "Hamilton communities", lwd = 1.5
-           #legend.col.show = FALSE
-           ) + 
+           palette = c("#365C8DFF", "#277F8EFF", "#1FA187FF", "#4AC16DFF", "#9FDA3AFF", "#FDE725FF"),
+           title.col = "Hamilton communities\n(% urbanisation)", lwd = 1, legend.col.show=FALSE) + 
   tm_shape(ALL_SCHOOLS_Elem_formapping3)+
   tm_symbols(shape = "combo_status",
-             alpha=0.6,
+             alpha=0.7,
              scale = 0.3,
@@
   tm_lines(col = "COMMUNITY_URB", 
-           palette = c("grey0","grey20","grey40", "grey60","grey80","grey100"),
-           title.col = "Hamilton communities", lwd = 1.5,
-           legend.col.show = FALSE
-           ) + 
+           palette = c("#365C8DFF", "#277F8EFF", "#1FA187FF", "#4AC16DFF", "#9FDA3AFF", "#FDE725FF"),
+           title.col = "Hamilton communities\n(% urbanisation)", lwd = 1, legend.col.show=FALSE) +  
   tm_shape(ALL_SCHOOLS_Elem_formapping3)+
   tm_symbols(shape = "combo_status",
-             alpha=0.6,
+             alpha=0.7,
              scale = 0.3,
@@
   tm_facets(by = c("year"), nrow=1)+
-  tm_shape(ham_community_bounds %>% st_cast("MULTILINESTRING")) + 
+  tm_shape(ham_community_bounds %>% st_cast("MULTILINESTRING"))+ 
   tm_lines(col = "COMMUNITY_URB", 
-           palette = c("grey0","grey20","grey40", "grey60","grey80","grey100"),
-           title.col = "Hamilton communities", lwd = 1.5,
-           legend.col.show = FALSE
-           ) + 
+           palette = c("#365C8DFF", "#277F8EFF", "#1FA187FF", "#4AC16DFF", "#9FDA3AFF", "#FDE725FF"),
+           title.col = "Hamilton communities\n(% urbanisation)", lwd = 1, legend.col.show=FALSE) + 
   tm_shape(ALL_SCHOOLS_Elem_formapping3)+
   tm_symbols(shape = "combo_status",
-             alpha=0.6,
+             alpha=0.7,
              scale = 0.3,
@@
   tm_lines(col = "COMMUNITY_URB", 
-           palette = c("grey0","grey20","grey40", "grey60","grey80","grey100"),
-           title.col = "Hamilton communities", lwd = 1.5,
-           legend.col.show = FALSE
-           ) + 
+           palette = c("#365C8DFF", "#277F8EFF", "#1FA187FF", "#4AC16DFF", "#9FDA3AFF", "#FDE725FF"),
+           title.col = "Hamilton communities\n(% urbanisation)", lwd = 1) + 
   tm_shape(ALL_SCHOOLS_Elem_formapping3)+
   tm_symbols(shape = "combo_status",
-             alpha=0.6,
+             alpha=0.7,
              scale = 0.3,
@@
 
-tmap_save(Vim_vvim_201120162021_CHANGE_plot, paste0(here::here(),"/Figures/Fig8.tiff"), height=12, width=10, dpi=300)
+tmap_save(Vim_vvim_201120162021_CHANGE_plot, paste0(here::here(),"/Figures/Fig8.tiff"), height=12, width=10, dpi=300) #I manually adjusted the text size of the legend of the first figure to be more similar to the rest of the legends
 ```
@@
              filter(mode == "motorized"), bbox=SAvail_im_DA) +
-  tm_polygons(col = "c_total_ttperkid", border.alpha = 0, border.col = "grey20", lwd=0.2, alpha=0.8,
+  tm_polygons(col = "c_total_ttperkid", border.alpha = 0, border.col = "grey20", lwd=0.2, alpha=0.6,
               palette = "-RdYlGn", midpoint=0, 
               style="quantile", n=10,
-              title = c("Motorized travel time minutes per pop-aged 5 to 14"))+
+              title = c("Motorized travel time minutes\nper pop-aged 5 to 14"))+
   tm_shape(ham_community_bounds %>% st_cast("MULTILINESTRING")) + 
   tm_lines(col = "COMMUNITY_URB", 
-           palette = c("grey0","grey20","grey40", "grey60","grey80","grey100"),
-           title.col = "Hamilton communities", lwd = 1.5,
-           legend.col.show = FALSE
-           ) +
+           palette = c("#365C8DFF", "#277F8EFF", "#1FA187FF", "#4AC16DFF", "#9FDA3AFF", "#FDE725FF"),
+           title.col = "Hamilton communities\n(% urbanisation)", lwd = 1, legend.col.show=FALSE) + 
   tm_shape(ALL_SCHOOLS_Elem_formapping3 |> filter(combo_change_year == "3) 2011-2021"))+
   tm_symbols(shape = "combo_status",
-             alpha=0.6,
+             alpha=0.7,
              scale = 0.3,
@@
              filter(mode == "non-motorized"), bbox=SAvail_im_DA) +
-  tm_polygons(col = "c_total_ttperkid", border.alpha = 0, border.col = "grey20", lwd=0.2, alpha=0.8,
+  tm_polygons(col = "c_total_ttperkid", border.alpha = 0, border.col = "grey20", lwd=0.2, alpha=0.6,
               palette = "RdYlGn", midpoint=0,
               style="quantile", n=10,
-              title = c("Non-motorized travel time minutes per pop-aged 5 to 14"))+
+              title = c("Non-motorized travel time minutes\nper pop-aged 5 to 14"))+
   tm_shape(ham_community_bounds %>% st_cast("MULTILINESTRING")) + 
   tm_lines(col = "COMMUNITY_URB", 
-           palette = c("grey0","grey20","grey40", "grey60","grey80","grey100"),
-           title.col = "Hamilton communities", lwd = 1.5,
-           legend.col.show = FALSE
-           ) +
+           palette = c("#365C8DFF", "#277F8EFF", "#1FA187FF", "#4AC16DFF", "#9FDA3AFF", "#FDE725FF"),
+           title.col = "Hamilton communities\n(% urbanisation)", lwd = 1) + 
   tm_shape(ALL_SCHOOLS_Elem_formapping3 |> filter(combo_change_year == "3) 2011-2021"))+
   tm_symbols(shape = "combo_status",
-             alpha=0.6,
+             alpha=0.7,
              scale = 0.3,
@@
 Public schools are a compelling case study in this respect.
-Across the globe, school closure and consolidation policies have been pursued by governments to reduce the cost of running public education systems  [@rongReviewResearchLowcarbon2022; @Dai_Wang_Zhang_Liao_Liu_2019].
+Across the globe, school closure and consolidation policies have been pursued by governments to reduce the per-student cost of operating public education systems [@rongReviewResearchLowcarbon2022; @Dai_Wang_Zhang_Liao_Liu_2019].
 Research has highlighted that these polices have knock-on implications, including the weakening of communities' social infrastructure [@butlerClosureRideau2019a; @irwinSchoolClosure2012; @AuttiHyryBeihammer2014] and negative household-level fiscal impacts such as a reduction of property values [@MerrallWhatsASchool2024].
-Operationally, savings may be experienced by governments, but they are paid at different levels by students, families, communities, and ultimately society in general.
+Operationally, savings may be experienced by government, but they are paid at different levels by students, families, communities and society.
 
@@
 When schools and residential locations are co-located, the journey-to-school presents an excellent opportunity for physical activity [@desjardinsFramingActive2024].
-However, school closures exchange the opportunities for active trips by students [@morencyStepsReserve2009; @morencyHowMany2007] for the near certainty of motorized travel [@rongReviewResearchLowcarbon2022].
-Indeed, motorized travel to school is the largest contributor of carbon emissions from personal daily travel after the journey-to-work [@rongReviewResearchLowcarbon2022; @pantelakiImpactHomeschoolCommuting2024].
-Alas, these impacts are not often considered in policy when schools close and/or consolidate [@sageman2022], a relatively common occurrence globally.
+However, school closures exchange these opportunities for active transportation with the near-certainty of motorized travel [@morencyStepsReserve2009; @morencyHowMany2007; @rongReviewResearchLowcarbon2022].
+Indeed, motorized travel to school is the second-largest contributor to carbon emissions from personal daily travel, following the journey-to-work [@rongReviewResearchLowcarbon2022; @pantelakiImpactHomeschoolCommuting2024].
+Unfortunately, these environmental impacts are rarely considered when school closure/consolidation policies are implemented [@sageman2022], and these policies are a relatively common occurrence globally.
 For instance, 65% of comprehensive schools (e.g.
@@
 
-In this paper, we investigate the case of Hamilton, a mid-sized city in Canada that has been forced by the province of Ontario to pursue school closure and consolidation policies.
-We <!--detail the Canadian political context that incentivized these policies and--> quantify the transportation implications of these policies using the capacity of schools in 2011, 2016 and 2021, after several waves of school closures and consolidations.
+In this context, we investigate the case of Hamilton, a mid-sized city in Canada that pursed school closure and consolidation policies following changes to educational funding provided by the province of Ontario.
+We quantify the transportation implications of these policies using the capacity of schools in 2011, 2016 and 2021, after several waves of school closures and consolidations.
 To this end, we use a novel multimodal extension of a singly-constrained accessibility measure known as spatial availability [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024].
 We use this measure to estimate active (i.e., walking) and motorized (i.e., car) accessibility while constraining the supply of the service to the capacity schools.
-In this way, this work furnishes evidence to document how government austerity hurts public goals (such as physical activity) and locks school districts and families into motorized patterns of travel for the foreseeable future.
-
-To foreshadow the paper's conclusions, our findings indicate that school seats have became less available in the city.
+In this way, this work furnishes evidence to document how government austerity hurts public goals (such as physical activity) and locks school boards and families into motorized patterns of travel for the foreseeable future.
+
+To foreshadow the paper's conclusions, our findings indicate that school seats have became less spatially available in the city.
 This change happened unevenly, with certain communities being disproportionately affected.
-The school closure and consolidation policy pursued by the government may have achieved cost savings in facility operation and maintenance.
+The school closure and consolidation policy pursued by the school board may have achieved cost savings in facility operation and maintenance.
 However, it has come at a cost to families, students, and communities, resulting in lost opportunities for healthier life-styles and an inevitable increase in motorized travel [@hammerWhy2012; @roseKids2013; @buckWhy2019].
@@
 Schools are a service that has been studied across several transportation dimensions.
-For instance, @marquesAccessibilityPrimarySchools2020 demonstrate that lower-income neighbourhoods in Portugal tend to have lower accessibility to schools compared to more advantaged neighbourhoods.
-Similar conclusions are drawn in West Virginia, US by @Talen_2001, England by @burgess2011parental, Baton Rouge, US by @williamsDisparitiesAccessibilityPublic2014, São Paulo and Curitiba, Brazil by @morenomonroyPublicTransportSchool2018, @pizzolQualifyingAccessibilityEducation2021, and @bittencourtEvaluatingAccessibilityAvailability2023, and Santiago, Chile by @tiznadoaitkenPublicTransportAccessibility2021.
+For instance, @marquesAccessibilityPrimarySchools2020 demonstrates that lower-income neighbourhoods in Portugal tend to have lower accessibility to schools compared to more advantaged neighbourhoods.
+Similar conclusions are drawn in West Virginia, US by @Talen_2001, England by @burgess2011parental, Baton Rouge, US by @williamsDisparitiesAccessibilityPublic2014, São Paulo and Curitiba, Brazil by @morenomonroyPublicTransportSchool2018, @pizzolQualifyingAccessibilityEducation2021, and @bittencourtEvaluatingAccessibilityAvailability2023, and Santiago de Chile by @tiznadoaitkenPublicTransportAccessibility2021.
 Research has also demonstrated that school closures are seldom politically neutral decisions.
@@
 The methods applied to measure school accessibility have also been varied.
-For example, @marquesAccessibilityPrimarySchools2020 present indicators of school seats per student along with the shortest distance from origin to schools within a school catchment.
+For example, @marquesAccessibilityPrimarySchools2020 presents indicators of school seats per student along with the shortest distance from origin to schools within a school catchment.
 Similar methods are also used in @Talen_2001 and @burgess2011parental.
@@
 Spatial availability allows for the representation of the number of 'school-seats' that are spatially available to students by mode of transportation (i.e., motorized and active travel), considering the school-aged student population by zone.
-Further, spatial availability can be represented as school-seats per capita, mathematically equivalent to the supply-to-demand ratio of the 2SFCA (see appendix in @soukhovIntroducingSpatialAvailability2023), which increases the interpretability of the analysis, and the comparability between years.
+Further, spatial availability can be represented as school-seats per capita, mathematically equivalent to the supply-to-demand ratio of the 2SFCA (see appendix in @soukhovIntroducingSpatialAvailability2023).
+Adopting this interpretation enhances the interpretability of the spatial availability analysis and its comparability between years.
 
@@
 The city's public English education system is managed by the Hamilton-Wentworth District School Board (HWDSB) and the Hamilton-Wentworth Catholic District School Board (HWCDSB).
-These two boards, like many in the province have faced fiscal pressures from the province.
-For example, in 2013, the HWDSB released its Long Term Facilities Master Plan [@hwdsbLongTermFacilities2013], which nominally meant to ensure "equitable, affordable and sustainable learning facilities".
-This plan indicated that 80% of its elementary schools (attended by students aged approximately 5 to 14) would be subject to evaluations, including a "Pupil Accommodation Review" [@hwdsbLongTermFacilities2013].
-The announcement promptly led to a public outcry, since the outcome of Pupil Accommodation Reviews historically translated into school closures and consolidations [@craggsBoardEvaluatingFuture2013; @seasonsSchoolClosurePolicy2014].
-
-School boards in Ontario rely on funding from the province which is distributed through a formula based on pupil enrollment [@mackenzieBlueprintFixOntario2018; @irwinSchoolClosure2012].
+Together, HWDSB and HWDCSB provide public English education to the majority of school-aged children in Hamilton [@faoOntarioSchoolBoards2023].
+And like many others school boards in Ontario, HWDSB and HWCDSB faced fiscal pressures from the province.
+For example, in 2013, the HWDSB released its Long Term Facilities Master Plan (LTFMP) [@hwdsbLongTermFacilities2013], which nominally meant to ensure "equitable, affordable and sustainable learning facilities".
+This plan indicated that 80% of its elementary schools (attended by students aged approximately 5 to 14) would be subject to a "Pupil Accommodation Review", an unprecedented amount of Reviews relative to other school boards [@seasons2014ARCCatalogues2014].
+
+The announcement of these Pupil Accommodation Reviews promptly led to public outcry, since the outcome of Pupil Accommodation Reviews historically translated into school closures and consolidations [@craggsBoardEvaluatingFuture2013; @seasonsSchoolClosurePolicy2014].
+The large number of scheduled Reviews was likely driven by the existing funding pressures faced by the school boards.
+The school boards rely almost entirely on provincial funding which is allocated through a pupil-enrolment based formula [@mackenzieBlueprintFixOntario2018; @irwinSchoolClosure2012].
 However, this funding formula has demonstrably become insufficient to maintain schools in a state of good repair [@auditorgeneralofontarioAnnualReport20152015].
-Further, existing "top-up" programs to assist in operation and maintenance costs (e.g., School Facilities Operation and Renewal Grant) were entirely phased out by 2018 [@tdsbLiftingMinistryEducations2024].
-Against this backdrop, the Pupil Accommodation Review guidelines outlined a "community-focused" process for assessing the value that a school provides to students, the community, and the local economy [@seasonsSchoolClosurePolicy2014; @ministryofeducationPupilAccommodationReview2006].
+Additionally, school boards are under intense pressure to justify their spending to the province, and the 2012 Ontario Budget exacerbated matters by calling for school closures as a means of achieving efficiencies [@irwinSchoolClosure2012].
+Furthermore, existing "top-up" programs designed to assist with operation and maintenance costs, such as the School Facilities Operation and Renewal Grant, were gradually phased out and removed by 2018 [@tdsbLiftingMinistryEducations2024].
+Against this backdrop, the Pupil Accommodation Review guidelines technically outlined a "community-focused" process for assessing the value that a school provides to students, the community, and the local economy [@seasonsSchoolClosurePolicy2014; @ministryofeducationPupilAccommodationReview2006].
 However in practice, critical residents felt the outcomes of the Pupil Accommodation Reviews were predetermined [@thompsonItJustGalls2024]: older schools were slated to be closed, and a wholesome breakdown of costs to repair and maintain schools was not made available [@kleinhuisSeriousConcernsSchool2013].
-The process introduced unbalanced incentives for school boards: as a concerned resident pointed at the time, "[b]y closing three schools, the HWDSB would have a strong case with the Ministry of Education to receive full funding for a new school" [@kleinhuisSeriousConcernsSchool2013].
-The Ministry's framework promotes the closing and consolidation of schools rather than cost-effective solutions for keeping older schools open [@kleinhuisSeriousConcernsSchool2013].
-
-Hamilton's two school boards (HWDSB and HWCDSB), which together provide public English education to the majority of school-aged children in Hamilton [@faoOntarioSchoolBoards2023], were particularly impacted by fiscal pressures.
-While Accommodation Reviews were a province-wide trend during this period, Hamilton's school boards underwent an unprecedented number of reviews compared to boards in other Ontario municipalities.
-[@seasons2014ARCCatalogues2014].
-In the wave of Pupil Accommodation Reviews between 2013 and 2016, `r SCHOOLS %>% filter(Status == "Removed") %>% pull(SchoolID) %>% length()` of the `r SCHOOLS %>% filter(Year == "2011 and 2016" | Year == "2011") %>% pull(SchoolID) %>% length()` elementary schools closed in Hamilton.
-Likely in part due to the unpopularity of school closures, in 2017, the provincial government introduced a moratorium, which paused any scheduled Pupil Accommodation Reviews.
-This created a backlog of Accommodation Reviews that HWDSB caught up to in subsequent years, when it closed `r SCHOOLS %>% filter(Status_2 == "Removed") %>% pull(SchoolID) %>% length()` more schools between 2017 to 2021 [@hwdsb2023LongTermFacilities2023].
-Though elementary school locations declined by `r -((SCHOOLS %>% filter(Year_2 != "2016") %>% pull(SchoolID) %>% length() - SCHOOLS %>% filter(Year != "2016") %>% pull(SchoolID) %>% length())/SCHOOLS %>% filter(Year != "2016") %>% pull(SchoolID) %>% length()) %>% scales::percent()` between 2011-2021, HWDSB intends to conduct future Pupil Accommodation Reviews as initially scheduled, waiting until the end of the moratorium before dates are assigned [@hwdsb2023LongTermFacilities2023].
-
-A justification for these Pupil Accommodation Reviews are operational savings based on "projected" reductions in student population and "under-utilized" school capacity [@craggsAxeCouldFall2012; @tdsbLiftingMinistryEducations2024; @opsbaRightSchoolsRight2024].
-These policies led to overall reductions in school capacity per student: school seats in the city declined `r  -(( ((avail_2021 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2021)) |> select(OTGC) |> sum())) - (avail_2011 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2011)) |> select(OTGC) |> sum())) / (avail_2011 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2011)) |> select(OTGC) |> sum()) ) |> scales::percent()`--but the impacts were felt unevenly across the city.
+The process introduced unbalanced incentives for school boards: as a concerned resident summarised at the time, "[b]y closing three schools, the HWDSB would have a strong case with the Ministry of Education to receive full funding for a new school" [@kleinhuisSeriousConcernsSchool2013].
+Ontario's Ministry of Education's framework promotes the closing and consolidation of schools rather than cost-effective solutions for keeping older schools open [@kleinhuisSeriousConcernsSchool2013].
+
+As forewarned, Pupil Accommodation Reviews led to school closures and consolidations.
+Between 2013 and 2016, `r SCHOOLS %>% filter(Status == "Removed") %>% pull(SchoolID) %>% length()` of the `r SCHOOLS %>% filter(Year == "2011 and 2016" | Year == "2011") %>% pull(SchoolID) %>% length()` elementary schools in Hamilton were closed.
+Likely influenced by the unpopularity of school closures, in 2017, the provincial government introduced a moratorium on starting any _new_ Pupil Accommodation Reviews.
+In the subsequent years, HWDSB worked to clear its backlog by completing the Accommodation Reviews that had already been initiated, resulting in the closure of `r SCHOOLS %>% filter(Status_2 == "Removed") %>% pull(SchoolID) %>% length()` additional schools between 2017 to 2021 [@hwdsb2023LongTermFacilities2023].
+While the number of elementary school locations declined by `r -((SCHOOLS %>% filter(Year_2 != "2016") %>% pull(SchoolID) %>% length() - SCHOOLS %>% filter(Year != "2016") %>% pull(SchoolID) %>% length())/SCHOOLS %>% filter(Year != "2016") %>% pull(SchoolID) %>% length()) %>% scales::percent()` between 2011-2021, HWDSB intends to conduct future Pupil Accommodation Reviews as initially scheduled, awaiting the end of the moratorium before scheduling new dates [@hwdsb2023LongTermFacilities2023].
+
+A justification for closures and consolidations are estimated operational savings based on "projected" reductions in student population and "under-utilized" school capacity [@craggsAxeCouldFall2012; @tdsbLiftingMinistryEducations2024; @opsbaRightSchoolsRight2024].
+However, city-wide, these policies led to overall reductions in school capacity per student.
+While many schools where closed and a few expanded (e.g., consolidated), the overall number of school seats in the city declined `r  -(( ((avail_2021 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2021)) |> select(OTGC) |> sum())) - (avail_2011 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2011)) |> select(OTGC) |> sum())) / (avail_2011 |> group_by(SchoolID) |> summarize(OTGC = first(OTGC2011)) |> select(OTGC) |> sum()) ) |> scales::percent()` between 2011 and 2021, and the impacts were felt unevenly across the city.
+
+<!--
 It should be noted that although closures dominated the Pupil Accommodation Review process, some schools were "consolidated", that is, a school or two were closed and another school was expanded or opened to accommodate students from closed schools.
-Since school locations decreased proportionally more than school-seats, elementary schools in Hamilton are on average larger than they once were.
-
-<!--
+Since school locations decreased proportionally more than school-seats, elementary schools in Hamilton are on average larger in 2021 than they were in 2012.
+
 In 2024, the moratorium on Pupil Accommodation Reviews is still in place and school boards are motivated by the same funding conditions.
@@
 
-Existing research indicates that schools have a plethora of benefits not captured by their maintenance and operational costs [@butlerClosureRideau2019a; @irwinSchoolClosure2012; @AuttiHyryBeihammer2014], and in this paper we focus on the transportation implications.
+Furthermore, existing research indicates that schools have a plethora of benefits not captured by their maintenance and operational costs [@butlerClosureRideau2019a; @irwinSchoolClosure2012; @AuttiHyryBeihammer2014], and in this paper we focus on the transportation implications.
 For students who live in proximity to schools, closures or consolidations directly eliminate the potential for active trips, and consequently physical activity.
@@
 
-- Did spatial availability change for home-to-school trips following school closures and consolidations?
-- What is the extent of the loss in walking minutes for these trips? 
-- How has motorized travel increased as a result? 
-- Which communities, particularly households with a high prevalence of low-income and hence more likely experiencing transport poverty, experienced the greatest losses due to this policy?
+- Did accessibility, namely spatial availability, change for home-to-school trips following school closures and consolidations?
+- Which communities, particularly those with a higher prevalence of low-income households, experience the greatest non-motorized (active) and motorized spatial availability losses?
 
@@
 
-We focus on the evaluation of the impacts of school closures and consolidations between 2011, 2016 and 2021.
-The outcome of interest is the multimodal school-seat spatial availability in Hamilton.
-To research this, the following data were prepared for each year: 1) school configurations, 2) student population, prevalence of low-income households, and residential parcel locations, and 3) estimated trip length by mode.
-These data are described next.
+This work focuses on the evaluation of the impacts of school closures and consolidations in Hamilton for the years 2011, 2016 and 2021.
+The primary outcome of interest is the multimodal school-seat spatial availability.
+To investigate this, the following data is prepared for each studied year: 1) school configurations, 2) student population, prevalence of low-income households, and residential parcel locations, and 3) estimated trip length by mode of transportation.
+The following three sub-sections describe this data.
 
@@
 
-For this study, we use the locations of elementary schools and their capacities at each of three years: 2011, 2016, and 2021.
+For this study, we use the locations of elementary schools and their capacities for each of three studied years: 2011, 2016, and 2021.
 
@@
 School catchments for the public and public-catholic catchments indicate the default school for households.
-However, families can decide to attend schools out of catchment, and in fact, 21%-23% <!--calculated in 03-travel-times-TTS.qmd--> of motorized school trips for populations aged 5 through 14 are out of catchment according to the regional travel survey (the Transportation Tomorrow Survey, or TTS) in 2011 and 2016, the most recent TTS available at the time of writing [@data_management_group_tts_2018].
-
-In addition to school locations, all schools were also tagged with a school status regarding their operational status (active or closed) and/or changes in location or school-seat capacity.
+However, families can opt for their children to attend schools out of catchment.
+In fact, according to the regional travel survey (the Transportation Tomorrow Survey, or TTS), 21%-23% <!--calculated in 03-travel-times-TTS.qmd--> of motorized school trips for populations aged 5 through 14 were made to out-of-catchment schools in 2011 and 2016, the most recent TTS available at the time of writing [@data_management_group_tts_2018].
+
+In addition to school locations, all schools were also tagged with a school status regarding their operational status (i.e., open or closed) and/or changes in location or school-seat capacity.
 A summary of the status of schools through 2011 to 2021 is shown in @fig-Fig1.
-@fig-Fig1 also displays a count of the number of schools with the status between years, the number of schools overall present in each year, and the symbolic representation of a school status that is used in Figures throughout the paper.
+@fig-Fig1 also displays a count of the number of schools with the status between years, the number of schools overall present in each year, and the symbolic representation of a school status that are used in Figures throughout the paper.
 Overall, a school can have one of seven school statuses in any given year.
@@
 #| label: fig-Fig1
-#| fig-cap: "Overview of the number of schools in Hamilton in 2011, 2016 and 2021 along with the type of school status change.."
+#| fig-cap: "Overview of the number of schools in Hamilton in 2011, 2016 and 2021 along with the type of school status change."
 #| echo: false
-#| out-width: "120%"
-
-knitr::include_graphics(paste0(here::here(),file="/Figures/Fig1.png"))
+#| out-width: "100%"
+
+knitr::include_graphics(paste0(here::here(),file="/Figures/Fig1.jpg"))
 #grid.raster(readTIFF(paste0(here::here(),"/Figures/Fig1.tiff")))
@@
 
-The capacity of each school is also required for each year.
-This capacity is the number of 'seats' available in the school.
-Capacity is calculated by the provincial Ministry of Education and referred to as the "on-the-ground capacity" (OTGC) of a school, albeit it is important to note that schools rarely operate at their nominal capacity, and can be either over-enrolled or under-enrolled.
+For the spatial availability analysis, the capacity of each school is also required for each year.
+This capacity is the number of 'seats' available at the school as calculated by the provincial Ministry of Education.
+It is referred to as the on-the-ground capacity (OTGC) of a school.
+However, it is important to note that schools rarely operate at exactly their OTGC, they are typically either over-enrolled or under-enrolled.
 
@@
 Equations (\ref{eq:OTGC-public}) and (\ref{eq:OTGC-catholic}) are the OTGC regression model for public schools and public catholic schools respectively.
-As seen in Table \ref{TabA1-OTGC}, the coefficient of determination for these models are quite high, at $R^2 = 0.999$ and $R^2 = 0.998$ for the public and public catholic school boards respectively, and the residual standard deviation is $\sigma = 0.212$ and $sigma = 0.331$.
+As seen in Table \ref{Tab1-OTGC}, the coefficient of determination for these models are quite high, at $R^2 = 0.999$ and $R^2 = 0.998$ for the public and public catholic school boards respectively, and the residual standard deviation is $\sigma = 0.212$ and $sigma = 0.331$.
 
@@
 This land-use classification is available within the 2021 residential parcel file described in the following sub-section.
-In @fig-Fig2, it can be seen that the majority of schools that changed status are in the HWDSB, with OTGC capacity decreasing in 2016 and 2021 compared to the previous year.
+In @fig-Fig2, it can be seen that the majority of schools that changed status are in the HWDSB, with OTGC decreasing in 2016 and 2021 compared to the previous year.
 Most school closures took place in the central and eastern parts of Hamilton's urban area (Hamilton Central), as well as in rural areas of western Hamilton (Flamborough).
@@
 A common poverty measure across all three census years is the low-income after-tax measure (LIM-AT), which reflects the proportion of private households that are below the median after-tax income in the region [@governmentofcanadaDictionaryCensusPopulation2017].
-The LIM-AT prevalence for households with children under 18 was retrained for this paper as there is no LIM-AT prevalence tabulated for households with exclusively elementary aged children.
+The LIM-AT prevalence for households with children under 18 was retrained for this paper as there is no LIM-AT prevalence tabulated for households with exclusively elementary-aged children.
 Population and LIM-AT variables were taken at the Dissemination Area (DA) level, the finest geographic unit publicly available.
@@
 
-To reflect multimodal travel behaviour, two types of data were estimated and compiled. The process to do so is explained next.
-
-First, we retrieves information about the mode used and origin-destination locations of home-to-school trips from the 2011 and 2016 Transportation Tomorrow Survey (TTS) [@data_management_group_tts_2018]; the 2021 TTS survey is unavailable at the time of writing so 2016 flows are assumed for the 2021 year.
+To reflect multimodal travel behaviour, two types of data was estimated and compiled. The process for obtaining and preparing this data is described as follows.
+
+First, we retrieved information about the mode used and origin-destination locations of home-to-school trips from the 2011 and 2016 TTS [@data_management_group_tts_2018]; the 2021 TTS survey is unavailable at the time of writing so 2016 flows are assumed for the 2021 year.
 The TTS is a travel survey conducted in the Greater Golden Horseshoe Area in Ontario typically every five years.
@@
 Travel time matrices were estimated for all TAZ centroids to TAZ centroids (matrices of size for 2011 and 2016: 234 x 234) as they are not reported in the TTS.
-Additionally matrices were estimated for all residential parcels to schools (matrices of size 134,340 x 147 for 2011, 139,467 x 137 for 2016, and 143,890 x 132 for 2021) for additional granularity.
+Additionally, matrices were estimated for all residential parcels to schools (matrices of size 134,340 x 147 for 2011, 139,467 x 137 for 2016, and 143,890 x 132 for 2021) for additional granularity.
 {r5r} is a R-based interface to the R5 routing engine [@conveyalConveyalR5Routing2022].
@@
 While all car trips are retained, walking trips over 27 minutes were filtered out.
-This value approximately corresponds to a distance of 1.6 km at a walking speed of 3.6 km/h, which is the threshold for qualifying for motorized transport provided by the school board [@HWDSB_2019].
+This duration corresponds to a distance of 1.6 km (assuming a walking speed of 3.6 km/h), which is the home-to-school distance threshold for qualifying for motorized transport provided by the school board [@HWDSB_2019].
+We use this threshold to assume that trips longer than 27 minutes are highly unlikely to be made on foot.
 
@@
 
-Accessibility is defined as the "potential for spatial interaction".
-It is classically presented as the gravity-based measure defined in @hansenHowAccessibilityShapes1959 and takes the following general multimodal formulation:
+Accessibility can be defined as the "potential for spatial interaction".
+It is classically presented as the measure defined in @hansenHowAccessibilityShapes1959 and takes the following general multimodal formulation:
 
@@
 In this paper, our work benefits from new developments in accessibility research, particularly the multimodal spatial availability measure [@soukhovIntroducingSpatialAvailability2023; @soukhovMultimodalSpatialAvailability2024].
-Spatial availability is a singly-constrained accessibility measure that accounts for competition by students using different modes for exclusive opportunities, such as school seats.
-The measure’s single constraint ensures that the marginals at the destination are met and thus the number of estimated school seats (opportunities) are preserved and allocated proportionally to the mode-using student population.
+Spatial availability is a singly-constrained accessibility measure that can account for competition of students using different modes potentially interacting with scarce opportunities, such as school seats.
+The measure’s sinlge constraint ensures that the marginals at the destination are met and thus the number of estimated school seats (opportunities) are preserved and allocated proportionally to the mode-using student populations.
 This proportional allocation of opportunities yields an interpretable and meaningful measure of opportunity access, particularly when comparing across modes, at the spatial resolution of a residential parcel, and multiple time periods.
-See @soukhovMultimodalSpatialAvailability2024 for further discussion, multimodal spatial availability $V_{i}^m$ is defined as given by Equation (\ref{eq:spatial-availability}):
+See @soukhovMultimodalSpatialAvailability2024 for further discussion.
+
+The multimodal spatial availability $V_{i}^m$ is defined as given by Equation (\ref{eq:spatial-availability}):
 
@@
 
-$F^{tm}_{ij}$ can be understood as the joint probability of allocating opportunities, where $F^{pm}_{i}$ is the population-based balancing factor that grants a larger share of opportunities to larger $m$ population spatial units and $F^{cm}_{ij}$ is the impedance-based balancing factor that grants a larger share of the opportunities to less $m$-travel costly centers.
+$F^{tm}_{ij}$ can also be understood as the joint probability of allocating opportunities, where $F^{pm}_{i}$ is the population-based balancing factor that grants a larger share of opportunities to larger $m$ population spatial units and $F^{cm}_{ij}$ is the impedance-based balancing factor that grants a larger share of the opportunities to less $m$-travel costly centers.
 Together $F^{tm}_{ij}$ ensures proportional allocation such that opportunities $O$ (like school-seats) are preserved for the whole region (i.e., $O = \sum_{j} O_j = \sum_{i} V_i = \sum_{m}\sum_{i} V_i^m$ ) and is reflected in Equation (\ref{eq:balancing-factors}):
@@
 
-To summarise the methodology and the data as described in Section 3, multimodal spatial availability $V_i^m$ and spatial availability per student $v_i^m$ is calculated for each parcel for each studied year as follows:
+To summarise the methodology and the data as described in Section 3, multimodal spatial availability $V_i^m$ and multimodal spatial availability per student $v_i^m$ is calculated for each parcel for each studied year as follows:
 
@@
 
-The probability density plots of the spatial availability values $V_i$ and the per capita spatial availability $v_i$ values for both modes and all three years are displayed in @fig-Fig6 as a city-wide overview.
+To clarify terminology for readers, $V_i^m$ is the number of school-seats available to populations by mode (either motorized or non-motorized) that matches the total number of opportunities in the region as a result of the proportional allocation mechanism.
+In other words, we describe $V_i^m$ as the number of school-seats available to a mode-using population.
+Each $i$ (parcel, DA, or community) maintains that singly-constrained relationship.
+$v_i^m$ is the per capita spatial availability to mode-using populations.
+
+The probability density plots of the spatial availability values $V_i^m$ and the per capita spatial availability $v_i^m$ values for both modes (motorized and non-motorized) and all three years are displayed in @fig-Fig6 as a city-wide overview.
 The left column displays parcel-level values, and the right reflects parcel-level values aggregated at the 2021 DA grid.
 Both the parcel and DA level values reflect the spatial unit at which they were calculated or aggregated.
-However, the values of the $V_i$ plots are not directly comparable since the x- and y- axes are set at different scale limits to highlight between year differences.
-However, overall trends in their distribution can be compared: in all four $V_i$ plots, the 2011 data is more right-skewed than more current years, indicating that spatial availability values used to be higher (e.g., more school-seat access).
+However, the values of the $V_i^m$ plots are not directly comparable since the x- and y- axes are set at different scale limits to highlight between year differences.
+However, overall trends in their distribution can be compared: in all four $V_i^m$ plots, the 2011 data is more right-skewed than more current years, indicating that spatial availability values used to be higher (e.g., more school-seat access).
 School closures and consolidations had an impact despite the background dynamics of assumed student population, mode use, travel time and parcel location.
-The DA level $V_i$ plots show the distribution of each DA value, which is the summation of all parcel $V_i$ values within a DA.
+The DA level $V_i^m$ plots show the distribution of each DA value, which is the summation of all parcel $V_i^m$ values within a DA.
 In 2011, DAs can be observed to also have more right-skewed spatial availability values than previous years.
@@
 
-Although $V_i$ values cannot be directly compared, zonal per capita $v_i$ values can be as the school-age population per spatial unit (e.g., parcel or DA) is used to normalize the respective $V_i$ value.
-$v_i$ values are presented in the bottom four plots of @fig-Fig6.
+Zonal per capita $v_i^m$ values can be compared across the years as the school-age population per spatial unit (e.g., parcel or DA) is used to normalize the respective $V_i^m$ value.
+$v_i^m$ values are presented in the bottom four plots of @fig-Fig6.
 It can be observed that the same 2011 right-skewed trend persists across all modes.
@@
 #| echo: false
-#| out-width: "120%"
+#| out-width: "100%"
 #| fig-pos: "H"
@@
 
-<!--maybe we change i to DA..?-->
 Representing the right column of @fig-Fig6 spatially, @fig-Fig7 displays the spatial availability for each DA for a given mode-using population $m$ (e.g., the sum of school seat availability per parcel in each DA for motorized or non-motorized students).
 @fig-Fig7 visualizes the spatial distribution of motorized and non-motorized school seat availability for 2011, 2016, and 2021.
-As a reminder, the sum of $V_i^m$ values for both motorized and non-motorized populations in a given year equals the city-wide OTGC (i.e., school seats across both the public english HWDSB and public english catholic HWCDSB).
+As a reminder, the sum of $V_i^m$ values for both motorized and non-motorized populations in a given year equals the city-wide OTGC (i.e., school seats across both the HWDSB and HWCDSB).
 This is due to $V_i^m$'s proportional allocation mechanism, singly-constraining opportunities through the proportion allocation balancing factors.
@@
 
-Referring to the first two rows of $V_i$ plots in @fig-Fig7 (purples), it is notable that the magnitudes of $V_i$ values for the motorized and non-motorized populations are tremendously different.
-Again, each $V_i$ value reflects the number of school seats spatially available to the mode-using population at that zonal level, so due to the lower non-motorized modal share (refer to @fig-Fig4), non-motorized values will be lower.
+Referring to the first two rows of $V_i^m$ plots in @fig-Fig7 (purples), it is notable that the magnitudes of $V_i^m$ values for the motorized and non-motorized populations are tremendously different.
+Again, each $V_i^m$ value reflects the number of school seats spatially available to the mode-using population at that zonal level, so due to the lower non-motorized modal share (refer to @fig-Fig4), non-motorized values will be lower.
 However, both mode-using populations have higher spatial availability values (darker purples) in DAs that are more proximate to schools.
 Hence, all DAs have higher values within Hamilton Central and those more proximate to schools in less-urban communities.
-This trend persists across 2011, 2016 and 2016, aligning with intuition: populations with shorter travel times to with relatively greater schools are calculated to have high school seat spatial availability.
+This trend persists across 2011, 2016 and 2021, aligning with intuition: populations with shorter travel times to with relatively greater schools are calculated to have high school seat spatial availability.
 Motorized populations have shorter travel times at DAs further in distance from schools, while non-motorized populations are only competitive at proximate distances.
 
-The bottom two rows of $v_i$ plots in @fig-Fig7 (yellows to greens) demonstrate the spatial availability per mode-using student population e.g., $V_i^m$ divided by the number of students aged 5-14 using a motorized mode or non-motorized mode.
+The bottom two rows of $v_i^m$ plots in @fig-Fig7 (yellows to greens) demonstrate the spatial availability per mode-using student population e.g., $V_i^m$ divided by the number of students aged 5-14 using a motorized mode or non-motorized mode.
 It is notable that for motorized populations, their spatial availability per student is above 1 (green) within the center of the city.
@@
 #| fig-cap: "Multimodal spatial availability (purples) and spatial availability per mode-using student (diverging reds and greens) aggregated at the DA level for 2011, 2016 and 2021.
-Scales are represented in deciles"
+Scales are represented in deciles."
 #| echo: false
@@
 
-To further explore these changes, the difference in $V_i$ (first two rows) and difference in $v_i$ (second two rows) between 2011-2016, 2016-2021 and overall between 2011-2021 are displayed in @fig-Fig8 along with schools that changed during that time period.
+To further explore these changes, the difference in $V_i^m$ (first two rows) and difference in $v_i^m$ (second two rows) between 2011-2016, 2016-2021 and overall between 2011-2021 are displayed in @fig-Fig8 along with schools that changed during that time period.
 2011-2016 and 2016-2021 add together to equal 2011-2021 changes quite clearly, e.g.
@@
 
-A few interesting spatial trends can be observed in @fig-Fig8.
-First, DAs in proximity to schools that closed (square points) see losses in $V_i$ (reds) for both motorized and especially non-motorized populations.
-For non-motorized populations, all closed school resulted in a decrease in $V_i$.
+A few notable spatial trends can be observed in @fig-Fig8.
+First, DAs in proximity to schools that closed (square points) see losses in $V_i^m$ (reds) for both motorized and especially non-motorized populations.
+For non-motorized populations, all closed school resulted in a decrease in $V_i^m$.
 The majority of school closures took place in the central and eastern parts of Hamilton's urban area (Hamilton Central), as well as in rural areas of western Hamilton (Flamborough).
 Hence, losses characterise those neighbourhoods.
-Secondly, gains in $V_i$ (blues) are seen where schools are opened (diamond points) or expanded (triangle points), especially for motorized populations.
+Secondly, gains in $V_i^m$ (blues) are seen where schools are opened (diamond points) or expanded (triangle points), especially for motorized populations.
 These gains are most notably in the more recently urbanized southeastern area of Hamilton (Glanbrook), along with a few pockets of DAs in proximity to new or expanded schools in Hamilton Central.
@@
 
-Exploring changes in $V_i$ provides a holistic view of spatial availability for a given DA.
+Exploring changes in $V_i^m$ provides a holistic view of spatial availability for a given DA.
 However, if the focus is on the ratio of spatial availability per student, the bottom two rows in @fig-Fig8 offer more insight.
@@
 However, due to an increase in population, the spatial availability per student overall still decreases between these two time periods.
-This case appears to occur in Ancaster between 2011 to 2016: while motorized $V_i$ increased, motorized $v_i$ slightly decreased, and non-motorized $v_i$ decreased relatively more than non-motorized $V_i$.
-As shown in @fig-Fig3, the student population and the number of students per residential unit increased more rapidly than the expanded school could accommodate, leading to a decrease in $v_i$ values in DAs proximate to expanded schools.
+This case appears to occur in Ancaster between 2011 to 2016: while motorized $V_i^m$ increased, motorized $v_i^m$ slightly decreased, and non-motorized $v_i^m$ decreased relatively more than non-motorized $V_i^m$.
+As shown in @fig-Fig3, the student population and the number of students per residential unit increased more rapidly than the expanded school could accommodate, leading to a decrease in $v_i^m$ values in DAs proximate to expanded schools.
 
 Similarly, in the southeastern community of Glanbrook, student populations grew from 2011 to 2021, and new schools opened.
-However, $v_i$ rates decreased because OTGC did not keep pace with student population growth.
-As another example, in Hamilton Central and the eastern community of Stoney Creek, changes in $V_i$ are unpatterned, while $v_i$ demonstrate more uniform decreases, with the pattern varying by modal-range for motorized versus non-motorized populations.
+However, $v_i^m$ rates decreased because OTGC did not keep pace with student population growth.
+As another example, in Hamilton Central and the eastern community of Stoney Creek, changes in $V_i^m$ are unpatterned, while $v_i^m$ demonstrate more uniform decreases, with the pattern varying by modal-range for motorized versus non-motorized populations.
 In these communities, the student population remained relatively stable, though OTGC decreased overall.
-$v_i$ is illuminating to consider when comparing changes in spatial availability rates.
+$v_i^m$ is illuminating to consider when comparing changes in spatial availability rates.
 
@@
 
-Changes in multimodal spatial availability is important: A summary of spatial availability per Hamilton community and associated dimensions in 2021, along with the percentage change between 2011 and 2021, is presented in @tbl-Tab1.
+Changes in multimodal spatial availability is important: A summary of spatial availability per Hamilton community and associated dimensions in 2021, along with the percentage change between 2011 and 2021, is presented in @tbl-Tab2.
 
 ```{r, warning=FALSE}
-#| label: tbl-Tab1
-#| tbl-cap: "Spatial availability and associated dimensions aggregated by communities in 2011 and how much they changed by 2021"
-#| out-width: "90%"
+#| label: tbl-Tab2
+#| tbl-cap: "Spatial availability and associated dimensions aggregated by communities in 2011 and percentage change by 2021.
+Motorized and non-motorized values are labeled with mt and nmt respectively."
 
@@
 
-#setting default font and size for the table
-set_flextable_defaults(
-  #font.family = "Arial", 
-  font.size = 10,
-  border.color = "gray", big.mark = "", digits = 1, pct_digits=1)
+tdata_1 <- tdata |>
+  select(c("HA.1","DU.1","SC.1","GL.1","FL.1","AN.1"))|> #rounding
+  mutate_if(is.numeric, round, digits=3)
+
+tdata_2 <- tdata |>
+  select(c("type","HA","DU","SC","GL","FL","AN"))|> #rounding
+  mutate_if(is.numeric, round, digits=1)
+
+tdata <- data.frame(type = tdata_2$type, 
+           HA = tdata_2$HA,
+           HA.1 = tdata_1$HA.1,
+           DU = tdata_2$DU,
+           DU.1 = tdata_1$DU.1,
+           SC = tdata_2$SC,
+           SC.1 = tdata_1$SC.1,
+           GL = tdata_2$GL,
+           GL.1 = tdata_1$GL.1,
+           FL = tdata_2$FL,
+           FL.1 = tdata_1$FL.1,
+           AN = tdata_2$AN,
+           AN.1 = tdata_1$AN.1)
 
@@
 
-  mk_par(j = c(3), value = as_paragraph(fmt_pct(HA.1))) |> #formating the % as with the percentage sign
+  mk_par(j = c(3), value = as_paragraph(fmt_pct(HA.1))) |> #formatting the % as with the percentage sign
   mk_par(j = c(5), value = as_paragraph(fmt_pct(DU.1))) |>
@@
 
-align(align = "center", part = "all")
-
-```
-
-Due to spatial availability's proportional allocation mechanism and this cases' calculation at the parcel level, results can also be summarized by community (@tbl-Tab1).
+align(align = "center", part = "all") |>
+fontsize(size = 7, part = "all") |>
+  set_table_properties(layout = "autofit")
+```
+
+Due to spatial availability's proportional allocation mechanism and this cases' calculation at the parcel level, results can also be summarized by community (@tbl-Tab2).
 The majority of schools were closed in Hamilton Central as seen in the largest reduction in OTGC by percentage and magnitude.
@@
 However, they access `r ((cSAvail_im_COMM2011_2021 |> filter(COMMUNITY == "FL" | COMMUNITY == "GL" | COMMUNITY == "AN") |> dplyr::select("V_im_2021") |> sum()) / (cSAvail_im_COMM2011_2021|> dplyr::select("V_im_2021") |> sum()) ) |> scales::percent()` of the spatial availability of school seats.
-Hamilton Central currently captures `r ((cSAvail_im_COMM2011_2021 |> filter(COMMUNITY == "HA") |> dplyr::select("V_im_2021") |> sum()) / (cSAvail_im_COMM2011_2021|> dplyr::select("V_im_2021") |> sum()) ) |> scales::percent()` of the spatial availability for it's `r ((ct201621_2011 |> filter(COMMUNITY == "HA") |> dplyr::select("kid_popadj_2021") |> sum()) / (ct201621_2011|> dplyr::select("kid_popadj_2021") |> sum() )) |> scales::percent(0.1)` of the population.
+Hamilton Central currently captures `r ((cSAvail_im_COMM2011_2021 |> filter(COMMUNITY == "HA") |> dplyr::select("V_im_2021") |> sum()) / (cSAvail_im_COMM2011_2021|> dplyr::select("V_im_2021") |> sum()) ) |> scales::percent()` of the spatial availability for its `r ((ct201621_2011 |> filter(COMMUNITY == "HA") |> dplyr::select("kid_popadj_2021") |> sum()) / (ct201621_2011|> dplyr::select("kid_popadj_2021") |> sum() )) |> scales::percent(0.1)` of the population.
 With student aged population growing in other communities, OTGC should be expanded - potentially to levels that Hamilton had in 2011 or beyond (`r (cSAvail_im_COMM2011_2021 |> filter(COMMUNITY == "HA" & mode == "c") |> dplyr::select("v_im_kid_popadj_2011")) |> pull() |> round(1)` school-seats per motorized mode using student and `r (cSAvail_im_COMM2011_2021 |> filter(COMMUNITY == "HA" & mode == "w") |> dplyr::select("v_im_kid_popadj_2011")) |> pull() |> round(1)` school-seats per active mode using student).
@@
 And the losses are drastic.
-Communities with high community averag spatial availability per students with values above 1 school-seat per student saw losses, as well as communities below this value.
-
-# Discussion and conclusion
-
-In this paper, we compared how multimodal spatial availability changed after a wave of school closures and consolidations in Hamilton.
-To do so, we constructed the student-aged population, OTGC of schools and their locations, and their travel behaviour to generate spatial availability landscapes for the three study years.
+Communities with high community average spatial availability per students with values above 1 school-seat per student saw losses, as well as communities below this value.
+
+# Discussion and conclusions
+
+In this paper, we compared how multimodal spatial availability changed after waves of school closures and consolidations in Hamilton between 2011 to 2021.
+To do so, we constructed the student-aged population, OTGC of schools and their locations, and their travel behaviour to generate spatial availability landscapes for the three study years 2011, 2016 and 2021.
 We aggregated the resulting values at the DA and community level and descriptively compared differences.
-We demonstrated that city-wide there are decreases in $V_i$ for both mode-using populations: the majority of students in the city have access to fewer school-seats than they would have in 2011.
+We demonstrated that city-wide there are decreases in $V_i^m$ for both mode-using populations: the majority of students in the city have access to fewer school-seats than they would have in 2011.
 Furthermore, as the majority of closed schools are within the core of the city: the more urban Hamilton Central and urban-areas of other communities saw the largest amount of this decrease.
@@
 Evidently, the number of OTGC in the city decreased between 2011 and 2021, so some sort of decrease was to be expected.
-However, should that decrease be felt hardest within neighbourhoods with the highest LIM-AT?
+However, should that decrease be felt most intensely within neighbourhoods with the highest LIM-AT?
 By students who have the potential to travel actively to school?
@@
 
-By quantifying the spatial availability, environmental, and active-travel implications of school closure/consolidation policies in Hamilton, our paper offers a methodology for spatial policy analysis scenario.
+By quantifying the multimodal spatial availability, namely motorized and non-motorized (active) travel, we explore the environmental and active-travel implications of school closure/consolidation policies in Hamilton.
+Our paper offers a methodology for spatial policy analysis scenario.
 The presented methodology can be used by researchers and decision-makers to plan and evaluate equitable and sustainable urban planning policies from a spatial perspective.
@@
 We urge policy-makers to view the 'spatial availability' of public resources from a per-capita perspective, plan capacity and location of service provision with the cost of travel (and associated implications such as GHG emissions, active transportation mode, safety of active travel modes, etc.) that sufficiently serves the target population.
-In this work, we suggest and assume a benchmark of 1.0 school-seats per student which accounts for sufficiency.
+In this work, we also suggest and assume a benchmark of 1.0 school-seats per student which accounts for sufficiency.
 Spatial availability can be used to obtain this benchmark since the total number of opportunities (i.e., school capacity) is preserved in the region of analysis unlike unconstrained forms of accessibility measurement e.g., Hansen-type measure [@hansenHowAccessibilityShapes1959].
@@
 The calculated spatial availability assumes any residential parcel can have a student (at the DA rate of students per residential unit) so all parcels are assigned a DA rate weighted by how many residential units are contained within the parcel.
-Each parcel can also visit any school, though travel times based on observed origin-destination behaviour (from the TTS) are more likely through the use of the empirical travel impedance functions.
+Each parcel is also assumed to visit any school, though travel times based on observed origin-destination behaviour (from the TTS) are more likely through the use of the empirical travel impedance functions.
 In this way, the spatial availability demonstrates what _potential interaction_ a student could have based on historic and average travel behaviour and shortest network-path travel times if residing in point in space.
-This representation is average, and by design obfuscates considerations that could make non-average and shortest-path travel unrealistic such as safety and built-environment concerns [@leeNeighborhoodEnvironmentsUtilitarian2021; @yumitaSchoolCommutingBarriers2021] nor considers significant non-transportation factors that impact fulsome school accessibility like quality [@bittencourtEvaluatingAccessibilityAvailability2023].
-Further, this study does not consider transit, school-bus, or cycling modes which have implications on how the GHG emissions estimation can be interpreted.
-Transit and school bus modes often taken more circuitous routes [@yumitaSchoolCommutingBarriers2021] but pool students thereby reducing emissions.
+This representation is average, and by design obfuscates considerations that could make non-average and shortest-path travel unrealistic such as safety and built-environment concerns [@leeNeighborhoodEnvironmentsUtilitarian2021; @yumitaSchoolCommutingBarriers2021] nor considers significant non-transportation factors that impact fulsome school accessibility like school quality [@bittencourtEvaluatingAccessibilityAvailability2023].
+An examination of these factors, be it through qualitative, mixed or quantitative methods, could be explored in future research, further building on the understanding of how school closures and consolidations impact communities.
+
+Furthermore, this study does not consider transit, school-bus, or cycling modes which have implications on how the environmental and active transport implications can be interpreted.
+For instance, transit and school bus modes often taken more circuitous routes [@yumitaSchoolCommutingBarriers2021] but pool students thereby reducing emissions.
 Cycling reduces travel time over the same walking distance at no extra emitted GHG.
-Lastly, this study also ascribes a static emission factor to represent motorized travel minutes based on an average diesel passenger car.
-In reality, GHG emissions emitted per minute of driving varies extensively depending on operating conditions and vehicle characteristics [@soukhovOccupancyGHGEmissions2022].
-In these ways, the assumptions made are to illustrate _potential_ and _average_ changes that resulted in a backdrop of school consolidation/closures to illustrate accessibility-related impacts for the city of Hamilton.
+Furthermore, GHG emissions emitted per minute of driving varies extensively depending on operating conditions and vehicle characteristics [@soukhovOccupancyGHGEmissions2022].
+Though this work focuses on changes in accessibility, an examination of dimensions of environmental impacts such as the _realized_ change in active travel minutes and emissions due to school consolidation/closures can be examined.
 

### 2025-01-19 paper/Manuscript.tex
R2 submitted.

@@
 % Make \paragraph and \subparagraph free-standing
-\makeatletter
 \ifx\paragraph\undefined\else
   \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}{
-    \@ifstar
-      \xxxParagraphStar
-      \xxxParagraphNoStar
-  }
-  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
-  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
+  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
 \fi
@@
   \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}{
-    \@ifstar
-      \xxxSubParagraphStar
-      \xxxSubParagraphNoStar
-  }
-  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
-  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
+  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
 \fi
-\makeatother
 
@@
 \makeatother
-
 \ifLuaTeX
@@
 \hypersetup{
-  pdftitle={10 years of school closures and consolidations: the impact on school accessibility},
+  pdftitle={Ten years of school closures and consolidations in Hamilton, Canada and the impact on multimodal accessibility},
   pdfauthor={Anastasia Soukhov; Christopher D. Higgins; Antonio Páez; Moataz Mohamed},
@@
 
-\title[10 years of school closures and consolidations: the impact on
-school accessibility]{10 years of school closures and consolidations:
-the impact on school accessibility}
+\title[Ten years of school closures and consolidations in Hamilton,
+Canada and the impact on multimodal accessibility]{Ten years of school
+closures and consolidations in Hamilton, Canada and the impact on
+multimodal accessibility}
 
@@
 
-\abstract{Reducing motorized travel and increasing active travel are top
-priorities for many communities. However, these priorities clash against
-many government's reluctance to invest in public services, including
-schools. Public school boards are increasingly asked to do more with
-less, a directive often accomplished by closing and consolidating
-schools. Alas, this policy tends to reduce the proximity of schools for
-many students, thus reducing the chances for active travel. In this
-paper we study the City of Hamilton, a mid-sized city in Ontario,
-Canada, which experienced numerous elementary schools closures between
+\abstract{Reducing motorized travel and increased active travel are top
+priorities for many communities. However, these goals are challenged by
+pressures that make governments reluctant to re/invest in public
+services, including schools. Public school boards are increasingly asked
+to do more with less, a directive often accomplished by closing and
+consolidating schools. Such policies can reduce the proximity of schools
+for many students, thereby limiting opportunities for active travel. In
+this paper, we examine the City of Hamilton, a mid-sized city in
+Ontario, Canada, which saw numerous elementary schools closures between
 2011 and 2021. Our analysis evaluates how school closures and
-consolidations reshaped accessibility for elementary school students by
-motorized and active travel. In this task we use spatial availability, a
+consolidations reshaped accessibility for elementary school students,
+both by motorized and active travel. We use spatial availability, a
 singly-constrained multimodal measure of spatial accessibility. This
 method's proportional allocation mechanisms allow us to make consistent
-comparisons in accessibility between time periods. Analysis is conducted
-at the parcel level, considering on-the-ground-capacity (OTGC) of
-schools and observed home-to-school trip lengths for motorized and
-active modes. Our findings reveal that overall spatial availability in
-the city declined, and while the school consolidation policy pursued by
-the province may have achieved cost savings in facility operation and
-maintenance, it imposes ongoing costs on students, families, and
-communities through reduced opportunities for physical activity and
-entrenched reliance on motorized travel.}
+comparisons in accessibility between time periods. The analysis is
+conducted at the parcel level, taking into account the
+on-the-ground-capacity (OTGC) of schools and observed home-to-school
+trip lengths for motorized and active modes. Overall, our findings
+reveal a decline in school spatial availability in the city. While the
+school closure and consolidation policy pursued by the province may have
+achieved cost savings in facility operation and maintenance, it imposed
+ongoing costs on students, families, and communities through reduced
+opportunities for physical activity and entrenched reliance on motorized
+travel.}
 
@@
 closure and consolidation policies have been pursued by governments to
-reduce the cost of running public education systems (Rong et al. 2022;
-Dai et al. 2019). Research has highlighted that these polices have
-knock-on implications, including the weakening of communities' social
-infrastructure (Butler, Kane, and Cooligan 2019; Irwin and Seasons 2012;
-Autti and Hyry-Beihammer 2014) and negative household-level fiscal
-impacts such as a reduction of property values (Merrall, Higgins, and
-Páez 2024). Operationally, savings may be experienced by governments,
-but they are paid at different levels by students, families,
-communities, and ultimately society in general.
+reduce the per-student cost of operating public education systems (Rong
+et al. 2022; Dai et al. 2019). Research has highlighted that these
+polices have knock-on implications, including the weakening of
+communities' social infrastructure (Butler, Kane, and Cooligan 2019;
+Irwin and Seasons 2012; Autti and Hyry-Beihammer 2014) and negative
+household-level fiscal impacts such as a reduction of property values
+(Merrall, Higgins, and Páez 2024). Operationally, savings may be
+experienced by government, but they are paid at different levels by
+students, families, communities and society.
 
@@
 the journey-to-school presents an excellent opportunity for physical
-activity (Desjardins et al. 2024). However, school closures exchange the
-opportunities for active trips by students (Morency, Roorda, and Demers
-2009; Morency, Demers, and Lapierre 2007) for the near certainty of
-motorized travel (Rong et al. 2022). Indeed, motorized travel to school
-is the largest contributor of carbon emissions from personal daily
-travel after the journey-to-work (Rong et al. 2022; Pantelaki, Claudia
-Caspani, and Maggi 2024). Alas, these impacts are not often considered
-in policy when schools close and/or consolidate (Sageman 2022), a
-relatively common occurrence globally. For instance, 65\% of
-comprehensive schools (e.g.~grades 1 through 9) in rural Finland since
-1990 closed (Autti and Hyry-Beihammer 2014), 10\% of elementary schools
-(kindergarten through grade 8) in Chicago, USA in 2013 (J. Lee and
-Lubienski 2017), and 7\% of primary and lower secondary schools
-(kindergarten through grade 9) were closed in Denmark between 2010-2011
-(Beuchert et al. 2018).
-
-In this paper, we investigate the case of Hamilton, a mid-sized city in
-Canada that has been forced by the province of Ontario to pursue school
-closure and consolidation policies. We quantify the transportation
-implications of these policies using the capacity of schools in 2011,
-2016 and 2021, after several waves of school closures and
-consolidations. To this end, we use a novel multimodal extension of a
-singly-constrained accessibility measure known as spatial availability
-(Soukhov et al. 2023; Soukhov et al. 2024). We use this measure to
-estimate active (i.e., walking) and motorized (i.e., car) accessibility
-while constraining the supply of the service to the capacity schools. In
-this way, this work furnishes evidence to document how government
-austerity hurts public goals (such as physical activity) and locks
-school districts and families into motorized patterns of travel for the
-foreseeable future.
+activity (Desjardins et al. 2024). However, school closures exchange
+these opportunities for active transportation with the near-certainty of
+motorized travel (Morency, Roorda, and Demers 2009; Morency, Demers, and
+Lapierre 2007; Rong et al. 2022). Indeed, motorized travel to school is
+the second-largest contributor to carbon emissions from personal daily
+travel, following the journey-to-work (Rong et al. 2022; Pantelaki,
+Claudia Caspani, and Maggi 2024). Unfortunately, these environmental
+impacts are rarely considered when school closure/consolidation policies
+are implemented (Sageman 2022), and these policies are a relatively
+common occurrence globally. For instance, 65\% of comprehensive schools
+(e.g.~grades 1 through 9) in rural Finland since 1990 closed (Autti and
+Hyry-Beihammer 2014), 10\% of elementary schools (kindergarten through
+grade 8) in Chicago, USA in 2013 (J. Lee and Lubienski 2017), and 7\% of
+primary and lower secondary schools (kindergarten through grade 9) were
+closed in Denmark between 2010-2011 (Beuchert et al. 2018).
+
+In this context, we investigate the case of Hamilton, a mid-sized city
+in Canada that pursed school closure and consolidation policies
+following changes to educational funding provided by the province of
+Ontario. We quantify the transportation implications of these policies
+using the capacity of schools in 2011, 2016 and 2021, after several
+waves of school closures and consolidations. To this end, we use a novel
+multimodal extension of a singly-constrained accessibility measure known
+as spatial availability (Soukhov et al. 2023; Soukhov et al. 2024). We
+use this measure to estimate active (i.e., walking) and motorized (i.e.,
+car) accessibility while constraining the supply of the service to the
+capacity schools. In this way, this work furnishes evidence to document
+how government austerity hurts public goals (such as physical activity)
+and locks school boards and families into motorized patterns of travel
+for the foreseeable future.
 
 To foreshadow the paper's conclusions, our findings indicate that school
-seats have became less available in the city. This change happened
-unevenly, with certain communities being disproportionately affected.
-The school closure and consolidation policy pursued by the government
-may have achieved cost savings in facility operation and maintenance.
-However, it has come at a cost to families, students, and communities,
-resulting in lost opportunities for healthier life-styles and an
-inevitable increase in motorized travel (Hammer and Baluja 2012; Rose
-2013; BUCK 2019). To increase the diffusion of this analysis, all work
-is transparent and reproducible, following best practices in the spatial
-sciences (Brunsdon and Comber 2021; Antonio Páez 2021). For this reason,
-all associated code and data are publicly available within the lead
-author's
+seats have became less spatially available in the city. This change
+happened unevenly, with certain communities being disproportionately
+affected. The school closure and consolidation policy pursued by the
+school board may have achieved cost savings in facility operation and
+maintenance. However, it has come at a cost to families, students, and
+communities, resulting in lost opportunities for healthier life-styles
+and an inevitable increase in motorized travel (Hammer and Baluja 2012;
+Rose 2013; BUCK 2019). To increase the diffusion of this analysis, all
+work is transparent and reproducible, following best practices in the
+spatial sciences (Brunsdon and Comber 2021; Antonio Páez 2021). For this
+reason, all associated code and data are publicly available within the
+lead author's
 \href{https://github.com/soukhova/School-closures-accessibility-impacts}{GitHub
@@
 transportation dimensions. For instance, Marques, Wolf, and Feitosa
-(2020) demonstrate that lower-income neighbourhoods in Portugal tend to
+(2020) demonstrates that lower-income neighbourhoods in Portugal tend to
 have lower accessibility to schools compared to more advantaged
@@
 Moreno-Monroy, Lovelace, and Ramos (2018), Pizzol, Giannotti, and
-Tomasiello (2021), and Bittencourt and Giannotti (2023), and Santiago,
+Tomasiello (2021), and Bittencourt and Giannotti (2023), and Santiago de
 Chile by Tiznado-Aitken, Munoz, and Hurtubia (2021). Research has also
@@
 The methods applied to measure school accessibility have also been
-varied. For example, Marques, Wolf, and Feitosa (2020) present
+varied. For example, Marques, Wolf, and Feitosa (2020) presents
 indicators of school seats per student along with the shortest distance
@@
 to the supply-to-demand ratio of the 2SFCA (see appendix in Soukhov et
-al. (2023)), which increases the interpretability of the analysis, and
-the comparability between years.
+al. (2023)). Adopting this interpretation enhances the interpretability
+of the spatial availability analysis and its comparability between
+years.
 
@@
 is managed by the Hamilton-Wentworth District School Board (HWDSB) and
-the Hamilton-Wentworth Catholic District School Board (HWCDSB). These
-two boards, like many in the province have faced fiscal pressures from
-the province. For example, in 2013, the HWDSB released its Long Term
-Facilities Master Plan (HWDSB 2013), which nominally meant to ensure
-``equitable, affordable and sustainable learning facilities''. This plan
-indicated that 80\% of its elementary schools (attended by students aged
-approximately 5 to 14) would be subject to evaluations, including a
-``Pupil Accommodation Review'' (HWDSB 2013). The announcement promptly
-led to a public outcry, since the outcome of Pupil Accommodation Reviews
+the Hamilton-Wentworth Catholic District School Board (HWCDSB).
+Together, HWDSB and HWDCSB provide public English education to the
+majority of school-aged children in Hamilton (FAO 2023). And like many
+others school boards in Ontario, HWDSB and HWCDSB faced fiscal pressures
+from the province. For example, in 2013, the HWDSB released its Long
+Term Facilities Master Plan (LTFMP) (HWDSB 2013), which nominally meant
+to ensure ``equitable, affordable and sustainable learning facilities''.
+This plan indicated that 80\% of its elementary schools (attended by
+students aged approximately 5 to 14) would be subject to a ``Pupil
+Accommodation Review'', an unprecedented amount of Reviews relative to
+other school boards (Seasons 2014a).
+
+The announcement of these Pupil Accommodation Reviews promptly led to
+public outcry, since the outcome of Pupil Accommodation Reviews
 historically translated into school closures and consolidations (Craggs
-2013; Seasons 2014b).
-
-School boards in Ontario rely on funding from the province which is
-distributed through a formula based on pupil enrollment (Mackenzie 2018;
-Irwin and Seasons 2012). However, this funding formula has demonstrably
-become insufficient to maintain schools in a state of good repair
-(Auditor General of Ontario 2015). Further, existing ``top-up'' programs
-to assist in operation and maintenance costs (e.g., School Facilities
-Operation and Renewal Grant) were entirely phased out by 2018 (TDSB
-2024). Against this backdrop, the Pupil Accommodation Review guidelines
-outlined a ``community-focused'' process for assessing the value that a
-school provides to students, the community, and the local economy
-(Seasons 2014b; Ministry of Education 2006). However in practice,
-critical residents felt the outcomes of the Pupil Accommodation Reviews
-were predetermined (Thompson, Collins, and Dean 2024): older schools
-were slated to be closed, and a wholesome breakdown of costs to repair
-and maintain schools was not made available (Kleinhuis 2013). The
-process introduced unbalanced incentives for school boards: as a
-concerned resident pointed at the time, ``{[}b{]}y closing three
-schools, the HWDSB would have a strong case with the Ministry of
-Education to receive full funding for a new school'' (Kleinhuis 2013).
-The Ministry's framework promotes the closing and consolidation of
-schools rather than cost-effective solutions for keeping older schools
-open (Kleinhuis 2013).
-
-Hamilton's two school boards (HWDSB and HWCDSB), which together provide
-public English education to the majority of school-aged children in
-Hamilton (FAO 2023), were particularly impacted by fiscal pressures.
-While Accommodation Reviews were a province-wide trend during this
-period, Hamilton's school boards underwent an unprecedented number of
-reviews compared to boards in other Ontario municipalities. (Seasons
-2014a). In the wave of Pupil Accommodation Reviews between 2013 and
-2016, 12 of the 147 elementary schools closed in Hamilton. Likely in
-part due to the unpopularity of school closures, in 2017, the provincial
-government introduced a moratorium, which paused any scheduled Pupil
-Accommodation Reviews. This created a backlog of Accommodation Reviews
-that HWDSB caught up to in subsequent years, when it closed 12 more
-schools between 2017 to 2021 (HWDSB 2023). Though elementary school
-locations declined by 10\% between 2011-2021, HWDSB intends to conduct
-future Pupil Accommodation Reviews as initially scheduled, waiting until
-the end of the moratorium before dates are assigned (HWDSB 2023).
-
-A justification for these Pupil Accommodation Reviews are operational
-savings based on ``projected'' reductions in student population and
-``under-utilized'' school capacity (Craggs 2012; TDSB 2024; OPSBA 2024).
-These policies led to overall reductions in school capacity per student:
-school seats in the city declined 5\%--but the impacts were felt
-unevenly across the city. It should be noted that although closures
-dominated the Pupil Accommodation Review process, some schools were
-``consolidated'', that is, a school or two were closed and another
-school was expanded or opened to accommodate students from closed
-schools. Since school locations decreased proportionally more than
-school-seats, elementary schools in Hamilton are on average larger than
-they once were.
-
-Existing research indicates that schools have a plethora of benefits not
-captured by their maintenance and operational costs (Butler, Kane, and
-Cooligan 2019; Irwin and Seasons 2012; Autti and Hyry-Beihammer 2014),
-and in this paper we focus on the transportation implications. For
-students who live in proximity to schools, closures or consolidations
-directly eliminate the potential for active trips, and consequently
-physical activity. Closing a school permanently reduces the potential
-for active trips and further entrenches reliance on motorized travel.
-Additionally, from the perspective of modal equity, school-aged
+2013; Seasons 2014b). The large number of scheduled Reviews was likely
+driven by the existing funding pressures faced by the school boards. The
+school boards rely almost entirely on provincial funding which is
+allocated through a pupil-enrolment based formula (Mackenzie 2018; Irwin
+and Seasons 2012). However, this funding formula has demonstrably become
+insufficient to maintain schools in a state of good repair (Auditor
+General of Ontario 2015). Additionally, school boards are under intense
+pressure to justify their spending to the province, and the 2012 Ontario
+Budget exacerbated matters by calling for school closures as a means of
+achieving efficiencies (Irwin and Seasons 2012). Furthermore, existing
+``top-up'' programs designed to assist with operation and maintenance
+costs, such as the School Facilities Operation and Renewal Grant, were
+gradually phased out and removed by 2018 (TDSB 2024). Against this
+backdrop, the Pupil Accommodation Review guidelines technically outlined
+a ``community-focused'' process for assessing the value that a school
+provides to students, the community, and the local economy (Seasons
+2014b; Ministry of Education 2006). However in practice, critical
+residents felt the outcomes of the Pupil Accommodation Reviews were
+predetermined (Thompson, Collins, and Dean 2024): older schools were
+slated to be closed, and a wholesome breakdown of costs to repair and
+maintain schools was not made available (Kleinhuis 2013). The process
+introduced unbalanced incentives for school boards: as a concerned
+resident summarised at the time, ``{[}b{]}y closing three schools, the
+HWDSB would have a strong case with the Ministry of Education to receive
+full funding for a new school'' (Kleinhuis 2013). Ontario's Ministry of
+Education's framework promotes the closing and consolidation of schools
+rather than cost-effective solutions for keeping older schools open
+(Kleinhuis 2013).
+
+As forewarned, Pupil Accommodation Reviews led to school closures and
+consolidations. Between 2013 and 2016, 12 of the 147 elementary schools
+in Hamilton were closed. Likely influenced by the unpopularity of school
+closures, in 2017, the provincial government introduced a moratorium on
+starting any \emph{new} Pupil Accommodation Reviews. In the subsequent
+years, HWDSB worked to clear its backlog by completing the Accommodation
+Reviews that had already been initiated, resulting in the closure of 12
+additional schools between 2017 to 2021 (HWDSB 2023). While the number
+of elementary school locations declined by 10\% between 2011-2021, HWDSB
+intends to conduct future Pupil Accommodation Reviews as initially
+scheduled, awaiting the end of the moratorium before scheduling new
+dates (HWDSB 2023).
+
+A justification for closures and consolidations are estimated
+operational savings based on ``projected'' reductions in student
+population and ``under-utilized'' school capacity (Craggs 2012; TDSB
+2024; OPSBA 2024). However, city-wide, these policies led to overall
+reductions in school capacity per student. While many schools where
+closed and a few expanded (e.g., consolidated), the overall number of
+school seats in the city declined 5\% between 2011 and 2021, and the
+impacts were felt unevenly across the city.
+
+Furthermore, existing research indicates that schools have a plethora of
+benefits not captured by their maintenance and operational costs
+(Butler, Kane, and Cooligan 2019; Irwin and Seasons 2012; Autti and
+Hyry-Beihammer 2014), and in this paper we focus on the transportation
+implications. For students who live in proximity to schools, closures or
+consolidations directly eliminate the potential for active trips, and
+consequently physical activity. Closing a school permanently reduces the
+potential for active trips and further entrenches reliance on motorized
+travel. Additionally, from the perspective of modal equity, school-aged
 populations with access to motorized transport gain a speed-time
@@
 \item
-  Did spatial availability change for home-to-school trips following
-  school closures and consolidations?
+  Did accessibility, namely spatial availability, change for
+  home-to-school trips following school closures and consolidations?
 \item
-  What is the extent of the loss in walking minutes for these trips?
-\item
-  How has motorized travel increased as a result?
-\item
-  Which communities, particularly households with a high prevalence of
-  low-income and hence more likely experiencing transport poverty,
-  experienced the greatest losses due to this policy?
+  Which communities, particularly those with a higher prevalence of
+  low-income households, experience the greatest non-motorized (active)
+  and motorized spatial availability losses?
 \end{itemize}
@@
 
-We focus on the evaluation of the impacts of school closures and
-consolidations between 2011, 2016 and 2021. The outcome of interest is
-the multimodal school-seat spatial availability in Hamilton. To research
-this, the following data were prepared for each year: 1) school
-configurations, 2) student population, prevalence of low-income
-households, and residential parcel locations, and 3) estimated trip
-length by mode. These data are described next.
+This work focuses on the evaluation of the impacts of school closures
+and consolidations in Hamilton for the years 2011, 2016 and 2021. The
+primary outcome of interest is the multimodal school-seat spatial
+availability. To investigate this, the following data is prepared for
+each studied year: 1) school configurations, 2) student population,
+prevalence of low-income households, and residential parcel locations,
+and 3) estimated trip length by mode of transportation. The following
+three sub-sections describe this data.
 
@@
 For this study, we use the locations of elementary schools and their
-capacities at each of three years: 2011, 2016, and 2021.
+capacities for each of three studied years: 2011, 2016, and 2021.
 
@@
 public and public-catholic catchments indicate the default school for
-households. However, families can decide to attend schools out of
-catchment, and in fact, 21\%-23\% of motorized school trips for
-populations aged 5 through 14 are out of catchment according to the
-regional travel survey (the Transportation Tomorrow Survey, or TTS) in
-2011 and 2016, the most recent TTS available at the time of writing
-(Data Management Group 2018).
+households. However, families can opt for their children to attend
+schools out of catchment. In fact, according to the regional travel
+survey (the Transportation Tomorrow Survey, or TTS), 21\%-23\% of
+motorized school trips for populations aged 5 through 14 were made to
+out-of-catchment schools in 2011 and 2016, the most recent TTS available
+at the time of writing (Data Management Group 2018).
 
 In addition to school locations, all schools were also tagged with a
-school status regarding their operational status (active or closed)
+school status regarding their operational status (i.e., open or closed)
 and/or changes in location or school-seat capacity. A summary of the
@@
 schools overall present in each year, and the symbolic representation of
-a school status that is used in Figures throughout the paper. Overall, a
-school can have one of seven school statuses in any given year. Either:
-1) the school did not change throughout 2011 to 2021, 2) the school
-expanded (i.e., increased school-seat capacity) or was 3) relocated and
-expanded (N.B., schools expanded only between 2011 through 2016 as they
-were typically done \emph{before} school closures and no new Pupil
-Accommodation Review was commenced after 2017), 4) the school closed
-sometime after 2011 and before 2016 or 5) after 2016 and before 2021,
-and 6) the school opened sometime after 2011 and before 2016 or 7) after
-2016 and before 2021.
+a school status that are used in Figures throughout the paper. Overall,
+a school can have one of seven school statuses in any given year.
+Either: 1) the school did not change throughout 2011 to 2021, 2) the
+school expanded (i.e., increased school-seat capacity) or was 3)
+relocated and expanded (N.B., schools expanded only between 2011 through
+2016 as they were typically done \emph{before} school closures and no
+new Pupil Accommodation Review was commenced after 2017), 4) the school
+closed sometime after 2011 and before 2016 or 5) after 2016 and before
+2021, and 6) the school opened sometime after 2011 and before 2016 or 7)
+after 2016 and before 2021.
 
@@
 
-\includegraphics[width=1.2\textwidth,height=\textheight]{../Figures/Fig1.png}
+\includegraphics[width=1\textwidth,height=\textheight]{../Figures/Fig1.jpg}
 
@@
 \caption{\label{fig-Fig1}Overview of the number of schools in Hamilton
-in 2011, 2016 and 2021 along with the type of school status change..}
+in 2011, 2016 and 2021 along with the type of school status change.}
 
@@
 
-The capacity of each school is also required for each year. This
-capacity is the number of `seats' available in the school. Capacity is
-calculated by the provincial Ministry of Education and referred to as
-the ``on-the-ground capacity'' (OTGC) of a school, albeit it is
-important to note that schools rarely operate at their nominal capacity,
-and can be either over-enrolled or under-enrolled.
+For the spatial availability analysis, the capacity of each school is
+also required for each year. This capacity is the number of `seats'
+available at the school as calculated by the provincial Ministry of
+Education. It is referred to as the on-the-ground capacity (OTGC) of a
+school. However, it is important to note that schools rarely operate at
+exactly their OTGC, they are typically either over-enrolled or
+under-enrolled.
 
@@
 OTGC regression model for public schools and public catholic schools
-respectively. As seen in Table \ref{TabA1-OTGC}, the coefficient of
+respectively. As seen in Table \ref{Tab1-OTGC}, the coefficient of
 determination for these models are quite high, at \(R^2 = 0.999\) and
@@
 Figure~\ref{fig-Fig2}, it can be seen that the majority of schools that
-changed status are in the HWDSB, with OTGC capacity decreasing in 2016
-and 2021 compared to the previous year. Most school closures took place
-in the central and eastern parts of Hamilton's urban area (Hamilton
-Central), as well as in rural areas of western Hamilton (Flamborough).
-When schools did open or expand, it was primarily in the more recently
+changed status are in the HWDSB, with OTGC decreasing in 2016 and 2021
+compared to the previous year. Most school closures took place in the
+central and eastern parts of Hamilton's urban area (Hamilton Central),
+as well as in rural areas of western Hamilton (Flamborough). When
+schools did open or expand, it was primarily in the more recently
 urbanized southeastern area of Hamilton (Glanbrook), with a few also in
@@
 children under 18 was retrained for this paper as there is no LIM-AT
-prevalence tabulated for households with exclusively elementary aged
+prevalence tabulated for households with exclusively elementary-aged
 children. Population and LIM-AT variables were taken at the
@@
 
-To reflect multimodal travel behaviour, two types of data were estimated
-and compiled. The process to do so is explained next.
-
-First, we retrieves information about the mode used and
+To reflect multimodal travel behaviour, two types of data was estimated
+and compiled. The process for obtaining and preparing this data is
+described as follows.
+
+First, we retrieved information about the mode used and
 origin-destination locations of home-to-school trips from the 2011 and
-2016 Transportation Tomorrow Survey (TTS) (Data Management Group 2018);
-the 2021 TTS survey is unavailable at the time of writing so 2016 flows
-are assumed for the 2021 year. The TTS is a travel survey conducted in
-the Greater Golden Horseshoe Area in Ontario typically every five years.
-With a target 5\% sampling rate, the survey is expanded to be
-representative at the traffic analysis zone (TAZ) level of geography.
-TAZ are spatial units created for the purpose of the TTS and pulled from
-the R data package \{TTS2016R\} (Soukhov and Páez 2023); a few DAs
-typically nest within each TAZ. The trip-level travel data extracted for
-this paper represent 13,715 and 12,878 motorized trips (mode used
-includes private car passenger, school bus, taxi, and transit) and 7,432
-and 7,085 non-motorized trips (mode used includes walk and cycling) for
-5 to 14 year olds from home-to-school in 2011 and 2016 respectively. The
+2016 TTS (Data Management Group 2018); the 2021 TTS survey is
+unavailable at the time of writing so 2016 flows are assumed for the
+2021 year. The TTS is a travel survey conducted in the Greater Golden
+Horseshoe Area in Ontario typically every five years. With a target 5\%
+sampling rate, the survey is expanded to be representative at the
+traffic analysis zone (TAZ) level of geography. TAZ are spatial units
+created for the purpose of the TTS and pulled from the R data package
+\{TTS2016R\} (Soukhov and Páez 2023); a few DAs typically nest within
+each TAZ. The trip-level travel data extracted for this paper represent
+13,715 and 12,878 motorized trips (mode used includes private car
+passenger, school bus, taxi, and transit) and 7,432 and 7,085
+non-motorized trips (mode used includes walk and cycling) for 5 to 14
+year olds from home-to-school in 2011 and 2016 respectively. The
 intensity of modal home-to-school flows are visualised in
@@
 (matrices of size for 2011 and 2016: 234 x 234) as they are not reported
-in the TTS. Additionally matrices were estimated for all residential
+in the TTS. Additionally, matrices were estimated for all residential
 parcels to schools (matrices of size 134,340 x 147 for 2011, 139,467 x
@@
 trips are retained, walking trips over 27 minutes were filtered out.
-This value approximately corresponds to a distance of 1.6 km at a
-walking speed of 3.6 km/h, which is the threshold for qualifying for
-motorized transport provided by the school board (HWDSB 2019).
+This duration corresponds to a distance of 1.6 km (assuming a walking
+speed of 3.6 km/h), which is the home-to-school distance threshold for
+qualifying for motorized transport provided by the school board (HWDSB
+2019). We use this threshold to assume that trips longer than 27 minutes
+are highly unlikely to be made on foot.
 
@@
 
-Accessibility is defined as the ``potential for spatial interaction''.
-It is classically presented as the gravity-based measure defined in
+Accessibility can be defined as the ``potential for spatial
+interaction''. It is classically presented as the measure defined in
 Hansen (1959) and takes the following general multimodal formulation:
@@
 (Soukhov et al. 2023; Soukhov et al. 2024). Spatial availability is a
-singly-constrained accessibility measure that accounts for competition
-by students using different modes for exclusive opportunities, such as
-school seats. The measure's single constraint ensures that the marginals
-at the destination are met and thus the number of estimated school seats
-(opportunities) are preserved and allocated proportionally to the
-mode-using student population. This proportional allocation of
-opportunities yields an interpretable and meaningful measure of
-opportunity access, particularly when comparing across modes, at the
-spatial resolution of a residential parcel, and multiple time periods.
-See Soukhov et al. (2024) for further discussion, multimodal spatial
-availability \(V_{i}^m\) is defined as given by Equation
-(\ref{eq:spatial-availability}):
+singly-constrained accessibility measure that can account for
+competition of students using different modes potentially interacting
+with scarce opportunities, such as school seats. The measure's sinlge
+constraint ensures that the marginals at the destination are met and
+thus the number of estimated school seats (opportunities) are preserved
+and allocated proportionally to the mode-using student populations. This
+proportional allocation of opportunities yields an interpretable and
+meaningful measure of opportunity access, particularly when comparing
+across modes, at the spatial resolution of a residential parcel, and
+multiple time periods. See Soukhov et al. (2024) for further discussion.
+
+The multimodal spatial availability \(V_{i}^m\) is defined as given by
+Equation (\ref{eq:spatial-availability}):
 
@@
 
-\(F^{tm}_{ij}\) can be understood as the joint probability of allocating
-opportunities, where \(F^{pm}_{i}\) is the population-based balancing
-factor that grants a larger share of opportunities to larger \(m\)
-population spatial units and \(F^{cm}_{ij}\) is the impedance-based
-balancing factor that grants a larger share of the opportunities to less
-\(m\)-travel costly centers. Together \(F^{tm}_{ij}\) ensures
-proportional allocation such that opportunities \(O\) (like
-school-seats) are preserved for the whole region (i.e.,
+\(F^{tm}_{ij}\) can also be understood as the joint probability of
+allocating opportunities, where \(F^{pm}_{i}\) is the population-based
+balancing factor that grants a larger share of opportunities to larger
+\(m\) population spatial units and \(F^{cm}_{ij}\) is the
+impedance-based balancing factor that grants a larger share of the
+opportunities to less \(m\)-travel costly centers. Together
+\(F^{tm}_{ij}\) ensures proportional allocation such that opportunities
+\(O\) (like school-seats) are preserved for the whole region (i.e.,
 \(O = \sum_{j} O_j = \sum_{i} V_i = \sum_{m}\sum_{i} V_i^m\) ) and is
@@
 To summarise the methodology and the data as described in Section 3,
-multimodal spatial availability \(V_i^m\) and spatial availability per
-student \(v_i^m\) is calculated for each parcel for each studied year as
-follows:
+multimodal spatial availability \(V_i^m\) and multimodal spatial
+availability per student \(v_i^m\) is calculated for each parcel for
+each studied year as follows:
 
@@
 
-The probability density plots of the spatial availability values \(V_i\)
-and the per capita spatial availability \(v_i\) values for both modes
-and all three years are displayed in Figure~\ref{fig-Fig6} as a
-city-wide overview. The left column displays parcel-level values, and
-the right reflects parcel-level values aggregated at the 2021 DA grid.
-Both the parcel and DA level values reflect the spatial unit at which
-they were calculated or aggregated. However, the values of the \(V_i\)
-plots are not directly comparable since the x- and y- axes are set at
-different scale limits to highlight between year differences. However,
-overall trends in their distribution can be compared: in all four
-\(V_i\) plots, the 2011 data is more right-skewed than more current
-years, indicating that spatial availability values used to be higher
-(e.g., more school-seat access). School closures and consolidations had
-an impact despite the background dynamics of assumed student population,
-mode use, travel time and parcel location. The DA level \(V_i\) plots
-show the distribution of each DA value, which is the summation of all
-parcel \(V_i\) values within a DA. In 2011, DAs can be observed to also
-have more right-skewed spatial availability values than previous years.
-In contrast, the 2016 and 2021 distributions are more intense and
-heavily skewed toward lower values.
-
-Although \(V_i\) values cannot be directly compared, zonal per capita
-\(v_i\) values can be as the school-age population per spatial unit
-(e.g., parcel or DA) is used to normalize the respective \(V_i\) value.
-\(v_i\) values are presented in the bottom four plots of
-Figure~\ref{fig-Fig6}. It can be observed that the same 2011
-right-skewed trend persists across all modes. Motorized mode-using
-students have access to more spatially available school seats than
-non-motorized mode using students (in 2011, approximately 1.3
-school-seats per student as compared to 0.7 and in 2021 approximately
-1.2 compared to 0.5). Due to school-seats per student at the DA level
-being an interpretable aggregation, values at this level will be the
-discussion focus of the remainder of this paper.
+To clarify terminology for readers, \(V_i^m\) is the number of
+school-seats available to populations by mode (either motorized or
+non-motorized) that matches the total number of opportunities in the
+region as a result of the proportional allocation mechanism. In other
+words, we describe \(V_i^m\) as the number of school-seats available to
+a mode-using population. Each \(i\) (parcel, DA, or community) maintains
+that singly-constrained relationship. \(v_i^m\) is the per capita
+spatial availability to mode-using populations.
+
+The probability density plots of the spatial availability values
+\(V_i^m\) and the per capita spatial availability \(v_i^m\) values for
+both modes (motorized and non-motorized) and all three years are
+displayed in Figure~\ref{fig-Fig6} as a city-wide overview. The left
+column displays parcel-level values, and the right reflects parcel-level
+values aggregated at the 2021 DA grid. Both the parcel and DA level
+values reflect the spatial unit at which they were calculated or
+aggregated. However, the values of the \(V_i^m\) plots are not directly
+comparable since the x- and y- axes are set at different scale limits to
+highlight between year differences. However, overall trends in their
+distribution can be compared: in all four \(V_i^m\) plots, the 2011 data
+is more right-skewed than more current years, indicating that spatial
+availability values used to be higher (e.g., more school-seat access).
+School closures and consolidations had an impact despite the background
+dynamics of assumed student population, mode use, travel time and parcel
+location. The DA level \(V_i^m\) plots show the distribution of each DA
+value, which is the summation of all parcel \(V_i^m\) values within a
+DA. In 2011, DAs can be observed to also have more right-skewed spatial
+availability values than previous years. In contrast, the 2016 and 2021
+distributions are more intense and heavily skewed toward lower values.
+
+Zonal per capita \(v_i^m\) values can be compared across the years as
+the school-age population per spatial unit (e.g., parcel or DA) is used
+to normalize the respective \(V_i^m\) value. \(v_i^m\) values are
+presented in the bottom four plots of Figure~\ref{fig-Fig6}. It can be
+observed that the same 2011 right-skewed trend persists across all
+modes. Motorized mode-using students have access to more spatially
+available school seats than non-motorized mode using students (in 2011,
+approximately 1.3 school-seats per student as compared to 0.7 and in
+2021 approximately 1.2 compared to 0.5). Due to school-seats per student
+at the DA level being an interpretable aggregation, values at this level
+will be the discussion focus of the remainder of this paper.
 
@@
 
-\includegraphics[width=1.2\textwidth,height=\textheight]{Manuscript_files/figure-pdf/fig-Fig6-1.pdf}
+\includegraphics[width=1\textwidth,height=\textheight]{Manuscript_files/figure-pdf/fig-Fig6-1.pdf}
 
@@
 non-motorized populations in a given year equals the city-wide OTGC
-(i.e., school seats across both the public english HWDSB and public
-english catholic HWCDSB). This is due to \(V_i^m\)'s proportional
-allocation mechanism, singly-constraining opportunities through the
-proportion allocation balancing factors. In this way, \(V_i^m\) values
-can be interpreted as a proportion of the total OTGC for that year, as
-the sum of \(V_i^m\) for both modes in a year equals the total OTGC for
-that year.
-
-Referring to the first two rows of \(V_i\) plots in
+(i.e., school seats across both the HWDSB and HWCDSB). This is due to
+\(V_i^m\)'s proportional allocation mechanism, singly-constraining
+opportunities through the proportion allocation balancing factors. In
+this way, \(V_i^m\) values can be interpreted as a proportion of the
+total OTGC for that year, as the sum of \(V_i^m\) for both modes in a
+year equals the total OTGC for that year.
+
+Referring to the first two rows of \(V_i^m\) plots in
 Figure~\ref{fig-Fig7} (purples), it is notable that the magnitudes of
-\(V_i\) values for the motorized and non-motorized populations are
-tremendously different. Again, each \(V_i\) value reflects the number of
-school seats spatially available to the mode-using population at that
+\(V_i^m\) values for the motorized and non-motorized populations are
+tremendously different. Again, each \(V_i^m\) value reflects the number
+of school seats spatially available to the mode-using population at that
 zonal level, so due to the lower non-motorized modal share (refer to
@@
 to schools in less-urban communities. This trend persists across 2011,
-2016 and 2016, aligning with intuition: populations with shorter travel
+2016 and 2021, aligning with intuition: populations with shorter travel
 times to with relatively greater schools are calculated to have high
@@
 
-The bottom two rows of \(v_i\) plots in Figure~\ref{fig-Fig7} (yellows
+The bottom two rows of \(v_i^m\) plots in Figure~\ref{fig-Fig7} (yellows
 to greens) demonstrate the spatial availability per mode-using student
@@
 aggregated at the DA level for 2011, 2016 and 2021. Scales are
-represented in deciles}
+represented in deciles.}
 
@@
 
-To further explore these changes, the difference in \(V_i\) (first two
-rows) and difference in \(v_i\) (second two rows) between 2011-2016,
+To further explore these changes, the difference in \(V_i^m\) (first two
+rows) and difference in \(v_i^m\) (second two rows) between 2011-2016,
 2016-2021 and overall between 2011-2021 are displayed in
@@
 
-A few interesting spatial trends can be observed in
-Figure~\ref{fig-Fig8}. First, DAs in proximity to schools that closed
-(square points) see losses in \(V_i\) (reds) for both motorized and
-especially non-motorized populations. For non-motorized populations, all
-closed school resulted in a decrease in \(V_i\). The majority of school
+A few notable spatial trends can be observed in Figure~\ref{fig-Fig8}.
+First, DAs in proximity to schools that closed (square points) see
+losses in \(V_i^m\) (reds) for both motorized and especially
+non-motorized populations. For non-motorized populations, all closed
+school resulted in a decrease in \(V_i^m\). The majority of school
 closures took place in the central and eastern parts of Hamilton's urban
@@
 (Flamborough). Hence, losses characterise those neighbourhoods.
-Secondly, gains in \(V_i\) (blues) are seen where schools are opened
+Secondly, gains in \(V_i^m\) (blues) are seen where schools are opened
 (diamond points) or expanded (triangle points), especially for motorized
@@
 
-Exploring changes in \(V_i\) provides a holistic view of spatial
+Exploring changes in \(V_i^m\) provides a holistic view of spatial
 availability for a given DA. However, if the focus is on the ratio of
@@
 case appears to occur in Ancaster between 2011 to 2016: while motorized
-\(V_i\) increased, motorized \(v_i\) slightly decreased, and
-non-motorized \(v_i\) decreased relatively more than non-motorized
-\(V_i\). As shown in Figure~\ref{fig-Fig3}, the student population and
+\(V_i^m\) increased, motorized \(v_i^m\) slightly decreased, and
+non-motorized \(v_i^m\) decreased relatively more than non-motorized
+\(V_i^m\). As shown in Figure~\ref{fig-Fig3}, the student population and
 the number of students per residential unit increased more rapidly than
-the expanded school could accommodate, leading to a decrease in \(v_i\)
-values in DAs proximate to expanded schools.
+the expanded school could accommodate, leading to a decrease in
+\(v_i^m\) values in DAs proximate to expanded schools.
 
@@
 populations grew from 2011 to 2021, and new schools opened. However,
-\(v_i\) rates decreased because OTGC did not keep pace with student
+\(v_i^m\) rates decreased because OTGC did not keep pace with student
 population growth. As another example, in Hamilton Central and the
-eastern community of Stoney Creek, changes in \(V_i\) are unpatterned,
-while \(v_i\) demonstrate more uniform decreases, with the pattern
+eastern community of Stoney Creek, changes in \(V_i^m\) are unpatterned,
+while \(v_i^m\) demonstrate more uniform decreases, with the pattern
 varying by modal-range for motorized versus non-motorized populations.
 In these communities, the student population remained relatively stable,
-though OTGC decreased overall. \(v_i\) is illuminating to consider when
-comparing changes in spatial availability rates.
+though OTGC decreased overall. \(v_i^m\) is illuminating to consider
+when comparing changes in spatial availability rates.
 
@@
 2021, along with the percentage change between 2011 and 2021, is
-presented in Table~\ref{tbl-Tab1}.
+presented in Table~\ref{tbl-Tab2}.
 
@@
 
-\begin{longtable}[c]{|p{0.75in}|p{0.75in}|p{0.75in}|p{0.75in}|p{0.75in}|p{0.75in}|p{0.75in}|p{0.75in}|p{0.75in}|p{0.75in}|p{0.75in}|p{0.75in}|p{0.75in}}
-
-\caption{\label{tbl-Tab1}Spatial availability and associated dimensions
-aggregated by communities in 2011 and how much they changed by 2021}
+\begin{longtable}[c]{ccccccccccccc}
+
+\caption{\label{tbl-Tab2}Spatial availability and associated dimensions
+aggregated by communities in 2011 and percentage change by 2021.
+Motorized and non-motorized values are labeled with mt and nmt
+respectively.}
 
@@
 
-\hhline{>{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}-}
-
-\multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{\textbf{}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{\textbf{Hamilton\ C.}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{\textbf{\%\ Δ}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{\textbf{Dundas}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{\textbf{\%\ Δ}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{\textbf{Stoney\ Creek}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{\textbf{\%\ Δ}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{\textbf{Glanbrook}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{\textbf{\%\ Δ}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{\textbf{Flamborough}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{\textbf{\%\ Δ}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{\textbf{Ancaster}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{\textbf{\%\ Δ}}}}} \\
+\hhline{>{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}-}
+
+\multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{\textbf{}}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{\textbf{Hamilton\ C.}}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{\textbf{\%\ Δ}}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{\textbf{Dundas}}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{\textbf{\%\ Δ}}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{\textbf{Stoney\ Creek}}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{\textbf{\%\ Δ}}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{\textbf{Glanbrook}}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{\textbf{\%\ Δ}}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{\textbf{Flamborough}}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{\textbf{\%\ Δ}}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{\textbf{Ancaster}}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{\textbf{\%\ Δ}}}}} \\
 
@@
 
-\hhline{>{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}-}\endfirsthead
-
-\hhline{>{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}-}
-
-\multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{\textbf{}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{\textbf{Hamilton\ C.}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{\textbf{\%\ Δ}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{\textbf{Dundas}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{\textbf{\%\ Δ}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{\textbf{Stoney\ Creek}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{\textbf{\%\ Δ}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{\textbf{Glanbrook}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{\textbf{\%\ Δ}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{\textbf{Flamborough}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{\textbf{\%\ Δ}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{\textbf{Ancaster}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{\textbf{\%\ Δ}}}}} \\
+\hhline{>{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}-}\endfirsthead
+
+\hhline{>{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}-}
+
+\multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{\textbf{}}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{\textbf{Hamilton\ C.}}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{\textbf{\%\ Δ}}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{\textbf{Dundas}}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{\textbf{\%\ Δ}}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{\textbf{Stoney\ Creek}}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{\textbf{\%\ Δ}}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{\textbf{Glanbrook}}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{\textbf{\%\ Δ}}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{\textbf{Flamborough}}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{\textbf{\%\ Δ}}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{\textbf{Ancaster}}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{\textbf{\%\ Δ}}}}} \\
 
@@
 
-\hhline{>{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}-}\endhead
-
-\multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{LIM-AT}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{23.8516970}}}} & \multicolumn{1}{>{\cellcolor[HTML]{F4FBF3}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{13.2\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{11.2879110}}}} & \multicolumn{1}{>{\cellcolor[HTML]{F6FCF5}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{10.6\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{10.8019420}}}} & \multicolumn{1}{>{\cellcolor[HTML]{F7FCF7}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{9.1\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{8.3083330}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFFAF8}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-2.5\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{7.2350900}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FAFDFA}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{5.4\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{7.0989650}}}} & \multicolumn{1}{>{\cellcolor[HTML]{EAF7E9}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{24.9\%}}}} \\
+\hhline{>{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}-}\endhead
+
+\multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{LIM-AT}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{23.9}}}} & \multicolumn{1}{>{\cellcolor[HTML]{F4FBF3}}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{13.2\%}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{11.3}}}} & \multicolumn{1}{>{\cellcolor[HTML]{F6FCF5}}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{10.6\%}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{10.8}}}} & \multicolumn{1}{>{\cellcolor[HTML]{F7FCF7}}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{9.1\%}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{8.3}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFFAF8}}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{-2.5\%}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{7.2}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FAFDFA}}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{5.4\%}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{7.1}}}} & \multicolumn{1}{>{\cellcolor[HTML]{EAF7E9}}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{24.9\%}}}} \\
 
@@
 
-\multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{Kid\ pop.}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{35310.5960000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFF9F7}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-3.2\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{2590.0000000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFE3D9}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-14.7\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{7514.7510000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{EEF9ED}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{20.0\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{2639.6530000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{A2DB9F}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{108.9\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{5365.0000000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{F8FCF7}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{8.7\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{4845.0000000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{F5FBF4}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{12.4\%}}}} \\
+\multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{Kid\ pop.}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{35,310.6}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFF9F7}}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{-3.2\%}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{2,590.0}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFE3D9}}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{-14.7\%}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{7,514.8}}}} & \multicolumn{1}{>{\cellcolor[HTML]{EEF9ED}}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{20.0\%}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{2,639.7}}}} & \multicolumn{1}{>{\cellcolor[HTML]{A2DB9F}}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{108.9\%}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{5,365.0}}}} & \multicolumn{1}{>{\cellcolor[HTML]{F8FCF7}}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{8.7\%}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{4,845.0}}}} & \multicolumn{1}{>{\cellcolor[HTML]{F5FBF4}}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{12.4\%}}}} \\
 
@@
 
-\multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{OTGC}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{39889.0820000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFECE6}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-9.8\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{2377.0380000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFFFFF}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{0.0\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{9010.1960000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFF1EB}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-7.6\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{1313.5220000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{91D58F}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{127.6\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{4590.7010000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFF1EB}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-7.5\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{3386.8900000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{F6FBF5}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{11.1\%}}}} \\
+\multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{OTGC}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{39,889.1}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFECE5}}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{-9.8\%}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{2,377.0}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFFFFF}}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{0.0\%}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{9,010.2}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFF1EB}}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{-7.6\%}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{1,313.5}}}} & \multicolumn{1}{>{\cellcolor[HTML]{91D58F}}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{127.6\%}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{4,590.7}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFF1EB}}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{-7.5\%}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{3,386.9}}}} & \multicolumn{1}{>{\cellcolor[HTML]{F6FBF5}}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{11.1\%}}}} \\
 
@@
 
-\hhline{>{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}-}
-
-\multicolumn{13}{>{\centering}m{\dimexpr 9.75in+24\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{\textbf{Motorized}}}}} \\
+\hhline{>{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}-}
+
+\multicolumn{13}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{\textbf{Motorized}}}}} \\
 
@@
 
-\hhline{>{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}-}
-
-\multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{V\ (mt)}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{42527.7500000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFE9E1}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-11.6\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{1952.2590000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFD4C5}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-22.5\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{6328.0100000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{F7FCF6}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{9.9\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{1531.3080000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{9DD99A}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{114.8\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{3382.5270000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFFBF9}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-2.3\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{4086.6160000}}}} & \multicolumn{1}{>{\cellcolor[HTML]{F9FDF9}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{7.2\%}}}} \\
+\hhline{>{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}-}
+
+\multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{V\ (mt)}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{42,527.8}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFE9E1}}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{-11.6\%}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{1,952.3}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFD4C5}}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{-22.5\%}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{6,328.0}}}} & \multicolumn{1}{>{\cellcolor[HTML]{F7FCF6}}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{9.9\%}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{1,531.3}}}} & \multicolumn{1}{>{\cellcolor[HTML]{9DD99A}}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{114.8\%}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{3,382.5}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFFBF9}}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{-2.3\%}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{4,086.6}}}} & \multicolumn{1}{>{\cellcolor[HTML]{F9FDF9}}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{7.2\%}}}} \\
 
@@
 
-\multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{v\ (mt)}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{1.2393060}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFEEE8}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-8.9\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{0.7590198}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFEEE7}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-9.0\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{0.8369173}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFF1EC}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-7.1\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{0.5823797}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FCFEFC}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{3.1\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{0.6312921}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFECE5}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-10.1\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{0.8491653}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFF6F2}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-5.0\%}}}} \\
+\multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{v\ (mt)}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{1.2}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFEEE8}}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{-8.9\%}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{0.8}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFEEE8}}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{-9.0\%}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{0.8}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFF1EC}}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{-7.1\%}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{0.6}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FCFEFC}}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{3.1\%}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{0.6}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFECE5}}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{-10.1\%}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{0.8}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFF5F2}}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{-5.0\%}}}} \\
 
@@
 
-\hhline{>{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}-}
-
-\multicolumn{13}{>{\centering}m{\dimexpr 9.75in+24\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{\textbf{Non-motorized}}}}} \\
+\hhline{>{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}-}
+
+\multicolumn{13}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{\textbf{Non-motorized}}}}} \\
 
@@
 
-\hhline{>{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1pt}-}
-
-\multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{V\ (nmt)}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{631.8817250}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFEAE2}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-11.0\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{28.7632500}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFF1EC}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-7.1\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{65.1592800}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFB39B}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-39.5\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{2.3855420}}}} & \multicolumn{1}{>{\cellcolor[HTML]{7CCD7C}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{1274.9\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{12.2516440}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFFDFC}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-1.0\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{18.5170610}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FF7957}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-67.6\%}}}} \\
+\hhline{>{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1pt}-}
+
+\multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{V\ (nmt)}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{631.9}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFEAE2}}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{-11.0\%}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{28.8}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFF1EC}}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{-7.1\%}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{65.2}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFB39B}}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{-39.5\%}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{2.4}}}} & \multicolumn{1}{>{\cellcolor[HTML]{7CCD7C}}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{1,274.9\%}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{12.3}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFFDFC}}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{-1.0\%}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{18.5}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FF7958}}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{-67.6\%}}}} \\
 
@@
 
-\multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{v\ (nmt)}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{0.7054674}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFE6DD}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-13.0\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{1.5790158}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFDFD4}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-16.6\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{0.8967578}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFCFBE}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-25.3\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{1.6565846}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFD0BF}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-24.8\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{1.8555444}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFE8DF}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-12.2\%}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{0.8641681}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFDFD3}\centering}m{\dimexpr 0.75in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{10}\selectfont{\global\setmainfont{DejaVu Sans}{-17.0\%}}}} \\
+\multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{v\ (nmt)}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{0.7}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFE6DD}}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{-13.0\%}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{1.6}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFDFD4}}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{-16.6\%}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{0.9}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFCFBE}}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{-25.3\%}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{1.7}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFD0BF}}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{-24.8\%}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{1.9}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFE8DF}}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{-12.2\%}}}} & \multicolumn{1}{>{}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{0.9}}}} & \multicolumn{1}{>{\cellcolor[HTML]{FFDFD3}}c}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\global\setmainfont{Arial}{-17.0\%}}}} \\
 
@@
 
-\hhline{>{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{BEBEBE}\global\arrayrulewidth=1.5pt}-}
+\hhline{>{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}->{\arrayrulecolor[HTML]{666666}\global\arrayrulewidth=1.5pt}-}
 
@@
 cases' calculation at the parcel level, results can also be summarized
-by community (Table~\ref{tbl-Tab1}). The majority of schools were closed
+by community (Table~\ref{tbl-Tab2}). The majority of schools were closed
 in Hamilton Central as seen in the largest reduction in OTGC by
@@
 seats. Hamilton Central currently captures 66\% of the spatial
-availability for it's 55.0\% of the population. With student aged
+availability for its 55.0\% of the population. With student aged
 population growing in other communities, OTGC should be expanded -
@@
 spatial availability per student from 2011 to 2021 in all communities.
-And the losses are drastic. Communities with high community averag
+And the losses are drastic. Communities with high community average
 spatial availability per students with values above 1 school-seat per
@@
 
-\section{Discussion and conclusion}\label{discussion-and-conclusion}
+\section{Discussion and conclusions}\label{discussion-and-conclusions}
 
 In this paper, we compared how multimodal spatial availability changed
-after a wave of school closures and consolidations in Hamilton. To do
-so, we constructed the student-aged population, OTGC of schools and
-their locations, and their travel behaviour to generate spatial
-availability landscapes for the three study years. We aggregated the
-resulting values at the DA and community level and descriptively
-compared differences. We demonstrated that city-wide there are decreases
-in \(V_i\) for both mode-using populations: the majority of students in
-the city have access to fewer school-seats than they would have in 2011.
-Furthermore, as the majority of closed schools are within the core of
-the city: the more urban Hamilton Central and urban-areas of other
-communities saw the largest amount of this decrease. And at a
-local-scale, students in residences proximate to schools that closed
-also saw dramatic reduce in non-motorized spatial availability values.
+after waves of school closures and consolidations in Hamilton between
+2011 to 2021. To do so, we constructed the student-aged population, OTGC
+of schools and their locations, and their travel behaviour to generate
+spatial availability landscapes for the three study years 2011, 2016 and
+2021. We aggregated the resulting values at the DA and community level
+and descriptively compared differences. We demonstrated that city-wide
+there are decreases in \(V_i^m\) for both mode-using populations: the
+majority of students in the city have access to fewer school-seats than
+they would have in 2011. Furthermore, as the majority of closed schools
+are within the core of the city: the more urban Hamilton Central and
+urban-areas of other communities saw the largest amount of this
+decrease. And at a local-scale, students in residences proximate to
+schools that closed also saw dramatic reduce in non-motorized spatial
+availability values.
 
@@
 2021, so some sort of decrease was to be expected. However, should that
-decrease be felt hardest within neighbourhoods with the highest LIM-AT?
-By students who have the potential to travel actively to school?
+decrease be felt most intensely within neighbourhoods with the highest
+LIM-AT? By students who have the potential to travel actively to school?
 Normatively, our cities should be increasing spatial availability for
@@
 
-By quantifying the spatial availability, environmental, and
+By quantifying the multimodal spatial availability, namely motorized and
+non-motorized (active) travel, we explore the environmental and
 active-travel implications of school closure/consolidation policies in
-Hamilton, our paper offers a methodology for spatial policy analysis
+Hamilton. Our paper offers a methodology for spatial policy analysis
 scenario. The presented methodology can be used by researchers and
@@
 modes, etc.) that sufficiently serves the target population. In this
-work, we suggest and assume a benchmark of 1.0 school-seats per student
-which accounts for sufficiency. Spatial availability can be used to
-obtain this benchmark since the total number of opportunities (i.e.,
+work, we also suggest and assume a benchmark of 1.0 school-seats per
+student which accounts for sufficiency. Spatial availability can be used
+to obtain this benchmark since the total number of opportunities (i.e.,
 school capacity) is preserved in the region of analysis unlike
@@
 weighted by how many residential units are contained within the parcel.
-Each parcel can also visit any school, though travel times based on
-observed origin-destination behaviour (from the TTS) are more likely
-through the use of the empirical travel impedance functions. In this
-way, the spatial availability demonstrates what \emph{potential
+Each parcel is also assumed to visit any school, though travel times
+based on observed origin-destination behaviour (from the TTS) are more
+likely through the use of the empirical travel impedance functions. In
+this way, the spatial availability demonstrates what \emph{potential
 interaction} a student could have based on historic and average travel
@@
 2021; Yumita et al. 2021) nor considers significant non-transportation
-factors that impact fulsome school accessibility like quality
-(Bittencourt and Giannotti 2023). Further, this study does not consider
-transit, school-bus, or cycling modes which have implications on how the
-GHG emissions estimation can be interpreted. Transit and school bus
-modes often taken more circuitous routes (Yumita et al. 2021) but pool
-students thereby reducing emissions. Cycling reduces travel time over
-the same walking distance at no extra emitted GHG. Lastly, this study
-also ascribes a static emission factor to represent motorized travel
-minutes based on an average diesel passenger car. In reality, GHG
-emissions emitted per minute of driving varies extensively depending on
-operating conditions and vehicle characteristics (Soukhov and Mohamed
-2022). In these ways, the assumptions made are to illustrate
-\emph{potential} and \emph{average} changes that resulted in a backdrop
-of school consolidation/closures to illustrate accessibility-related
-impacts for the city of Hamilton.
+factors that impact fulsome school accessibility like school quality
+(Bittencourt and Giannotti 2023). An examination of these factors, be it
+through qualitative, mixed or quantitative methods, could be explored in
+future research, further building on the understanding of how school
+closures and consolidations impact communities.
+
+Furthermore, this study does not consider transit, school-bus, or
+cycling modes which have implications on how the environmental and
+active transport implications can be interpreted. For instance, transit
+and school bus modes often taken more circuitous routes (Yumita et al.
+2021) but pool students thereby reducing emissions. Cycling reduces
+travel time over the same walking distance at no extra emitted GHG.
+Furthermore, GHG emissions emitted per minute of driving varies
+extensively depending on operating conditions and vehicle
+characteristics (Soukhov and Mohamed 2022). Though this work focuses on
+changes in accessibility, an examination of dimensions of environmental
+impacts such as the \emph{realized} change in active travel minutes and
+emissions due to school consolidation/closures can be examined.
 
